### 2014-10-14 paper/broom.tex
Added skeleleton of the paper, with abstract

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames]{color}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{stmaryrd}
\usepackage{comment}
\usepackage{float}
\usepackage{verbatimbox}
\usepackage{breakurl}
\usepackage[breaklinks]{hyperref}

%Formatting Commands
%-------------------
\newcommand{\C}[1]{\code{#1}}
\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
\newcommand{\code}[1]{\,{\tt #1}\,}
\newcommand{\conj}{\wedge}
\newcommand{\disj}{\vee}
\newcommand{\seqn}[1]{\overline {#1}}
\newcommand{\tuplee}[1]{\langle #1 \rangle}
\newcommand{\ALT}{~\mid~}
\newcommand{\lamb}[2]{\lambda #1.\,#2}
\newcommand{\spc}[0]{\quad}
\newcommand{\U}[0]{\cup}
\newcommand{\X}[0]{\times}
\newcommand{\xn}[0]{\cap}
\newcommand{\redsto}[0]{\longrightarrow}
\newcommand{\rredsto}[0]{\rightsquigarrow_r }
\newcommand{\evalsto}[0]{\longrightarrow }
\newcommand{\env}[0]{\Gamma}
\newcommand{\finmaparrow}[0]{\overset { Fin }{ \longrightarrow } }
\newcommand{\envwf}[1]{\vdash\,#1}
\newcommand{\tywf}[2]{#1\,\vdash\,#2}
\newcommand{\hastyp}[3]{#1\,\vdash\,#2\,:\,#3}
\newcommand{\hassort}[3]{#1\,\vdash\,#2\,::\,#3}
\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
\newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
\newcommand{\msfolsig}[0]{{\Lambda \equiv}}
%\newcommand{\msentails}[2]{#1\,\models_L\,#2}
\newcommand{\msentails}[2]{#1\,\models\,#2}
\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
\newcommand{\defeq}[0]{ \triangleq }
\newcommand{\tymap}[0]{\mathcal{F}}
\newcommand{\etaeq}[0]{\eta_{eq}}
\newcommand{\semof}[1]{\thesemof{#1}}
\newcommand{\mssemof}[1]{\thesemof{#1}}
\newcommand{\rwsto}[0]{=}
\newcommand{\rrwsto}[0]{\hookrightarrow}
\newcommand{\rwbindfwd}[0]{\gamma_{\Rightarrow}}
\newcommand{\rwbindbwd}[0]{\gamma_{\Leftarrow}}
\newcommand{\eqhlt}[1]{\underline{#1}}
\newcommand{\tfr}[0]{F_R}
\newcommand{\ground}[1]{\underline{#1}}
\newcommand{\semeq}[0]{\equiv}
\newcommand{\elabsto}[0]{\looparrowright }
\newcommand{\arr}[0]{$\rightarrow$}
\newcommand{\eef}[0]{$\Leftrightarrow$}
\newcommand{\imp}[0]{$\Rightarrow$}
\newcommand{\RNull}[0]{\emptyset}
\newcommand{\cty}[0]{ty}
\newcommand{\hsp}[0]{\hspace*{-0.1in}}
\newcommand{\hspa}[0]{\hspace*{-0.08in}}
% removable:
\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
\newcommand{\absof}[1]{\left\| #1 \right\| }

%Rules
%-----
\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
\newcommand{\ilrulelabel}[1]{{\sc #1}}
\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
                           {\begin{array}{c}#2\end{array}}}

% Math mode
%-----------
\newenvironment{nop}{}{}
\newenvironment{smathpar}{
\begin{nop}\small\begin{mathpar}}{
\end{mathpar}\end{nop}\ignorespacesafterend}

% Theorems Environment
%----------------------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\qed}{\nobreak \ifvmode \relax \else
      \ifdim\lastskip<1.5em \hskip-\lastskip
      \hskip1.5em plus0em minus0.5em \fi \nobreak
      \vrule height0.75em width0.5em depth0.25em\fi}

\usepackage{listings}
\usepackage{subcaption}
\usepackage{stmaryrd}
\usepackage[skip=5pt]{caption}

\lstset{ %
language=Java, % choose the language of the code
basicstyle=\scriptsize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={relation, not, : , /\\}
}
\lstdefinestyle{hs} {
  language=Haskell,
  basicstyle=\scriptsize\ttfamily
}

\begin{document}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

% \titlebanner{banner above paper title}        % These are ignored unless
% \preprintfooter{short description of paper}   % 'preprint' option specified.

%The Paper
%---------------------
\title{Broom: Safe Region-Based Memory Management for Dataflow
Programs}

\authorinfo{}
           {}
           {}
\maketitle

\begin{abstract}

Distributed execution engines, such as MapReduce and Dryad, represent
a data-parallel computation as a dataflow graph with computational
vertices connected by communication edges. The computation is run by
executing the vertices of this graph on a set of available computers,
while communication edges are materialized as either network
connections, or appropriate inter-process communication channels.
Although MapReduce and Dryad can perform optimal distribution the
computation across available computers, they nevertheless rely on
language runtimes, such as JVM or C\#.Net, to manage an atomic
computation on a given machine.

In this paper, we demonstrate that the general purpose garbage
collection scheme employed by C\#.Net runtime is not very effective
for dataflow programs insofar as the latency due to GC is intolerably
high in many practical examples. We propose a memory management scheme
that combines user-managed memory regions with automatic garbage
collection, and lets programmers exploit specific memory behaviours
exhibited by dataflow programs. To protect programmers from the
dangers of manual memory management, we supplement our memory
management scheme with a region type system that statically ensures
the memory safety of region-annotated C\# programs, even in presence
of higher-order stateful functions. Further, we describe a type
inference algorithm that infers principal region types for first-order
programs thereby saving programmers the need to write region
annotations on types. Our inference algorithm is also very effective,
albeit incomplete, for higher-order programs. As a testament to
efficacy of our approach, we demonstrate performance improvements over
a range of realistic dataflow programs that were rewritten to use
region-based manual memory management, but are nevertheless judged to
be safe by our type system.

\end{abstract}

% general terms are not compulsory anymore, 
% you may leave them out
%% \terms
%% term1, term2

\input{introduction}

% \appendix
% \section{Appendix Title}

% This is the text of the appendix, if you need one.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{catalyst}

\end{document}

### 2015-08-27 paper/introduction.tex
Contributions added.

\section{Introduction}
\label{sec:introduction}

The paper makes the following contributions:

\begin{itemize}
  \item We present \name, a Java-like typed object-oriented language
  that eschews garbage collection in favour of programmer-managed
  memory regions . \name extends its core language, which includes
  \emph{lambdas} (higher-order functions) and \emph{generics}
  (parametric polymorphism), with constructs to create, manage and
  destroy static and dynamic memory regions. Dynamic regions are
  first-class values in \name; they manifest as objects of type
  \C{Region}, and are treated on par with other objects. \name is
  particularly suited for implementing distributed dataflow operators
  (e.g: \c{SELECT} and \c{JOIN}), whose memory behavior exhibits
  certain characteristics that admits efficient and \emph{mostly} safe
  region-based memory management. Experiments demonstrate that
  \naiad~\cite{naiad} dataflow operators using programmer-managed
  memory regions outperform their GC counterparts by a margin of upto
  59\%.

  \item \name is equipped with a region type system that statically
  guarantees safety for a well-defined subset of all memory accesses
  in a well-typed program. This subset includes all memory accesses,
  including accesses that happen across different regions, except
  dereferences of \C{Region} objects themselves. For the later kind,
  we rely on \GK{reference counting?} to ensure safety at run-time.
  Our experiments with \naiad workloads show that \GK{x\%} of all
  memory memory accesses fall under the statically safe subset, while
  only \GK{100-x\%} require run-time mechanisms to ensure safety.

  \item We define an operational semantics for \name, and a type
  safety result that formalizes and proves safety guarantees described
  above.

  \item We describe a region type inference algorithm for \name that
  (a). completely eliminates the need to annotate \name programs with
  region types, and (b). enables seamless interoperability between
  region-aware \name programs and legacy standard library code that is
  region-oblivious. The cornerstone of our inference algorithm is
  \csolve, a novel constraint solving algorithm that performs
  abduction and deduction in a Herbrand constraint system to solve the
  constraints generated by the type inference algorithm.

  \item We prove that region type inference is sound with respect to
  its type system, and also complete for the set of first-order
  programs (i.e., programs without lambdas).

  \item We describe an implementation of \name frontend in OCaml,
  along with experimental results, and case studies where the region
  type system was able to identify unsafe memory accesses statically.

\end{itemize}

The rest of the paper is organized as follows. The next section
presents an informal overview of \name and its region type system.
\S~\ref{sec:type-system} formalizes the type system and its safety
guarantees. The type inference algorithm is described in
\S~\ref{sec:type-inference}. \S~\ref{sec:csolve} focuses on \csolve,
the constraint solving algorithm, and its correctness guarantees.
Implementations details and practical extensions to the type system
are described in \S~\ref{sec:implementation}. \S~\ref{sec:evaluation}
presents experimental evaluation and case studies.
\S~\ref{sec:related} discusses the related work, and
\S~\ref{sec:conclusion} concludes.

### 2015-09-12 paper/broom.tex
Contributions added.
Sec 2 full content.

@@
 
+\usepackage{courier}            % standard fixed width font
+\usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
+\usepackage{listings}          % format code
+\usepackage{enumitem}      % adjust spacing in enums
+\usepackage{xspace}
 \usepackage{mathtools}
@@
 \newcommand{\absof}[1]{\left\| #1 \right\| }
+\newcommand{\broom}{{\sc Broom}\xspace}
+\newcommand{\name}{{\sc Broom}\xspace}
+\newcommand{\naiad}{{\sc Naiad}\xspace}
+\newcommand{\csolve}{{\sc Csolve}\xspace}
+\newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
+\newcommand{\csharp}{C\#~}
+\newcommand{\outlives}{\succeq}
 
@@
 
-\lstset{ %
+\newcommand{\lstjava}{\lstset{ %
 language=Java, % choose the language of the code
-basicstyle=\scriptsize\ttfamily,       % the size of the fonts that are used for the code
+basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
 keywordstyle=\color{Bittersweet},
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={relation, not, : , /\\}
-}
-\lstdefinestyle{hs} {
-  language=Haskell,
-  basicstyle=\scriptsize\ttfamily
-}
+morekeywords={letregion, openalloc, open, withroot, not, : , /\\}
+}}
+\lstnewenvironment{codejava}
+    { % \centering
+			\lstjava
+      \lstset{}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
+
+\lstnewenvironment{numcodejava}
+    { % \centering
+			\lstjava
+      \lstset{numbers=left,firstnumber=1}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
 
@@
 
+\input{overview}
+
 % \appendix

### 2015-09-12 paper/introduction.tex
Sec 2 full content.

@@
 \begin{itemize}
-  \item We present \name, a Java-like typed object-oriented language
+  \item We present \name, a \csharp-like typed object-oriented language
   that eschews garbage collection in favour of programmer-managed
@@
 The rest of the paper is organized as follows. The next section
-presents an informal overview of \name and its region type system.
-\S~\ref{sec:type-system} formalizes the type system and its safety
-guarantees. The type inference algorithm is described in
-\S~\ref{sec:type-inference}. \S~\ref{sec:csolve} focuses on \csolve,
-the constraint solving algorithm, and its correctness guarantees.
-Implementations details and practical extensions to the type system
-are described in \S~\ref{sec:implementation}. \S~\ref{sec:evaluation}
-presents experimental evaluation and case studies.
-\S~\ref{sec:related} discusses the related work, and
+presents an informal overview of \name, and motivates the need for a
+region type system.  \S~\ref{sec:type-system} formalizes the type
+system and its safety guarantees. The type inference algorithm is
+described in \S~\ref{sec:type-inference}. \S~\ref{sec:csolve} focuses
+on \csolve, the constraint solving algorithm, and its correctness
+guarantees.  Implementations details and practical extensions to the
+type system are described in \S~\ref{sec:implementation}.
+\S~\ref{sec:evaluation} presents experimental evaluation and case
+studies.  \S~\ref{sec:related} discusses the related work, and
 \S~\ref{sec:conclusion} concludes.

### 2015-09-12 paper/overview.tex
Sec 2 full content.

\section{Overview of \name}
\label{sec:overview}

Stripped of its region-specific constructs, the core language of \name
is a familar subset of \csharp that includes features such as
class-based inheritance, dynamic dispatch, parametric polymorphism,
and lambdas.  We therefore elide a discussion of the core language,
focusing instead on the concepts specific to region-based memory
management in \name.

\subsection{Stack Regions and Outlives Relation}

Simplest of the region-specific constructs in \name is a \C{letregion}
lexical block that creates a new labeled memory region. The semantics
of the \C{letregion} block is similar to Tofte and
Talpin~\cite{ttpopl94}'s \C{letregion} expression; the newly
introduced region is live only inside the block, where its label is in
scope for \C{new} expressions to make allocations inside the region.
For example, in the following code, object \C{a} is allocated in the
region \C{R0} introduced by the \C{letregion}:
\begin{center}
\begin{codejava}
  letregion R0 {
    Object a = new@R0 Object();
  }
\end{codejava}
\end{center}
The \C{letregion} blocks can be nested, leading to a stack of regions
that are deallocated in the reverse order in which they are allocated.
Following ~\cite{cyclonepldi02}, we therefore call regions introduced
by \C{letregion} expressions as \emph{stack regions}. The stack
discipline induces \emph{outlives} relationship among regions created
by nested \C{letregion}s, where the region introduced by the outer
\C{letregion} is guaranteed to outlive the one introduced by the inner
\C{letregion}. It is therefore safe to refer to an object allocated in
outer region from the inner region, but the converse is not true. For
example, consider the following code (assume that class \C{A} has
field \C{x} of type \C{Object}):
\begin{center}
\begin{codejava}
  letregion R0 {
    A a0 = new@R0 A();
    letregion R1 {
      A a1 = new@R1 A();
      a1.x = new@R0 Object();
      a0.x = new@R1 Object();
      ...
    }
    ...
  }
\end{codejava}
\end{center}
The code creates two stack regions - \C{R0} and \C{R1}, where \C{R0}
outlives \C{R1} (denoted as $\C{R0 \outlives R1}$). Objects \C{a0} and
\C{a1} are allocated in regions \C{R0} and \C{R1}, respectively. The
first assignment statement assigns to \C{a1.x} an object allocated in
outer region (\C{R0}). This assignment is absolutely safe as \C{a1.x}
refers to a longer living object, hence is guaranteed to be a valid
reference throughout the lifetime of \C{a1}. In contrast, the second
assignment is unsafe, as it assigns to \C{a0.x} an object, whose
lifetime is shorter than the lifetime of \C{a0}, making it unsafe to
dereference \C{a0.x} outside the inner block. 
% Unsafe assignments can also happen indirectly via a function call..
Preempting such unsafe assignments is the \emph{raison d'etre} of the
region type system.

\subsection{Allocation Context}

As evident from the examples shown above, the suffix \C{@R} in
\C{new@R} instruct the \C{new} keyword to allocate the newly created
object in the region labeled \C{R}. However, \name does not require
the allocation region to be specified following a \C{new}. In the
absence of explicit specification, \name allocates the newly created
object in its \emph{allocation context}, which is the region
introduced by the closest enclosing \C{letregion} block. Besides being
a syntactic convenience, the facility to elide allocation region
specification is imperative to interface region-oblivious standard
library code with region-aware application code. For instance, the
following \name code, which makes explicit use of regions, calls
the standard library's iterator methods, which are region-oblivious:
\begin{center}
\begin{codejava}
  int findMin(List<int> l) {
    assert(!l.isEmpty());
    int min = l.get(0);
    letregion R0 {
      ListIterator<int> iter = l.listIterator();
      while(iter.hasNext()) {
        int n = iter.getNext();
        if (n<min) min = n;
      }
    }
    return min;
  }
\end{codejava}
\end{center}
Due to the absence of new region creations, and explicit region
annotations with \C{new} expressions, all new objects created by a
standard library method are allocated in the allocation context for
the method call. In the above example, the allocation context for the
call to \C{List} library's \C{listIterator} method is the region
\C{R1}. Therefore, all objects created by the method, including the
iterator object, are allocated in \C{R0}. This fact should be captured
appropriately by the region type of \C{listIterator} method if the
type system were to prove the safety of the above code.

\subsection{Boxed and Unboxed Values}

In \name, like in \csharp, values of some primitive types, such as
integers and bools, are unboxed, whereas others, such as strings are
arrays, are boxed. Unlike unboxed values, boxed values are objects,
hence passed by reference. It is therefore important to be mindful
of allocation regions for strings and arrays, but not so much for
integers and bools. For instance, in the following code, the
assignment to integer variable \C{x} is safe, but the assignment to
the string variable \C{y} is unsafe, as the allocation region (\C{R0})
of the string constant has lesser lifetime than \C{y}.
\begin{center}
\begin{codejava}
  int x;
  string y;
  letregion R0 {
    x = 2;
    y = "Hello, World!";
  }
  ...
\end{codejava}
\end{center}

\subsection{Dynamic (Transferable) Regions}

\name's stack regions have lexically scoped lifetimes, hence are not
suitable for applications that need to transfer data between
autonomous actors, such as \naiad dataflow operators. To support these
applications, it is imperative to support regions whose lifetimes
transcend the boundaries of lexical blocks, functions, or even
programs. Regions that fit this description in \name are called
\emph{dynamic regions}, or \emph{transferable regions}.

Transferable regions, like stack regions, are memory regions
Transferable regions are first class values of \name; they are
considered objects of \C{Region} class, and, like objects of other
classes, they are created using the \C{new} keyword, can be passed as
arguments, and returned from methods. Like a stack region, a
transferable region facilitates arbitrary object allocations, but
unlike a stack region, it has a well-defined \emph{root} object
denoting the root of the data structure being transferred. In a
typical dataflow computation, an upstream actor (e.g., a \C{SELECT}
operator) constructs a transferable region, sets its root to the data
structure supposed to contain intermediate results, populates the data
structure with intermediate results, and then transfers it to a
downstream actor (e.g., a \C{COUNT} operator), which performs further
processing. Since a transferable region escapes the lifetime of the
sender, there must be no references from inside of the transferable
region to objects allocated in other memory regions of the sender.
Such references, if exist, may become invalid references in the
context of recipient's address space, jeopardizing memory safety.
\name relies on its region type system to prevent such unsafe
references from being created.

\begin{figure}
\begin{codejava}
  public class Region<T> {
    public Region(Func<void,T> mkRoot);
    public void free();
    public void transfer();
  }
\end{codejava}
\caption{The type signature of the \C{Region} class}
\label{fig:region-class}
\end{figure}
The type signature of \C{Region} class is described in
Fig.~\ref{fig:region-class}. The class is parametric over the type of
the root object (i.e., the \emph{root type}) of the transferable
region it creates. The \C{Region} constructor expects a \C{Func}
object - a higher-order function argument that returns an object of
root type.  The idea is to execute this function with the newly
created transferable region as its allocation context. The object thus
created inside the transferable region will serve as its root object.
For instance, the following code creates a transferable region
(\C{rgn}) and sets its root to the string object returned by its
\C{Func} argument (a lambda expression):
\begin{codejava}
  Region<string> rgn = new Region<string>
                        (() => "Hello");
\end{codejava}
Among the methods of the \C{Region} class, \C{free} deallocates
region's memory, while \C{transfer} transfers the region to a
downstream actor as determined by the run-time. The precise semantics
of \C{transfer} are unimportant in the context of the region type
system; it suffices to understand \C{transfer} as an alias of
\C{free}.

A transferable region is said to be in \emph{closed} state when
created. \name provides \C{openalloc} lexical construct to \emph{open}
a transferable region as an allocation context, and to access its root
object:
\begin{codejava}
  Region<string> rgn = new Region<string>
                        (() => "Hello");
  openalloc rgn withroot str {
    print str; //prints "Hello"
    str = str + " World!"; 
  }
\end{codejava}
In the above code, the \C{openalloc} lexical block opens \C{rgn} for
allocation, and sets the newly introduced local variable (\C{str}) to
refer to its root object.  The assignment statement within the block
updates the root. Since \C{rgn} is the allocation context within the
block, the new string is allocated inside \C{rgn}, hence the
assignment is safe. In some cases, it is desirable to open a
transferable region only to read its contents, but not to perform any
allocations. For this purpose, \name provides \C{open} lexical
construct that, like \C{openalloc}, opens a transferable region and
names its root object, but unlike \C{openalloc}, does not make it an
allocation context. A transferable region can be \C{openalloc}'d or
\C{open}'d many times, before it is eventually \C{free}'d or
\C{transfer}'d.

The design decisions reflected above, \emph{viz.}, to ascribe a root
object to every transferable region, to make use of a higher-order
argument to initialize the root object, and to require that a region
be explictly opened in a lexical block for allocations, were made to
reflect the message semantics of transferable regions, to promote
uniform style of programming with different kinds of regions, and to
precisely state and enforce the safety guarantees offered by \name's
region type system (\S~\ref{sec:type-system}).

\input{naiad-ex}

Fig~\ref{fig:naiad-ex} shows \C{SelectVertex} class, which is a
straightforward implementation of the \C{SELECT} dataflow operator.
\C{SelectVertex} implements \C{onReceive} method, which is called by
the run-time when a message from an upstream actor is ready to be
delivered. A message is usually a sequence of tuples (e.g., key-value
pairs) wrapped inside a transferable region (\C{inRgn}).
\C{SelectVertex} opens its input region \C{inRgn} (line 16), iterates
through the sequence, filtering the input tuples that meet the
selection criterion of the \C{whereFn} (line 19), and applying the
\C{mapFn} to map the selected input tuples to output (line 22).
Observe that the iterator is allocated in the temporary (stack) region
\C{R0} that lives just long enough.  Functions \C{mapFn} and
\C{whereFn} are defined by the users of the dataflow system, hence
they are called \emph{user-defined functions}.  Unlike dataflow system
builders, users are oblivious of regions, hence user-defined functions
are expected to not contain explicit region annotations. While
\C{whereFn} defines the selection criterion that the user is
interested in (e.g: \C{age $\ge$ 18}), \C{mapFn} can either project
attributes of its input tuple amounting to type \C{TOut}, or can
create and return a new \C{TOut} object. Note that, unlike \C{inRgn},
the output transferable region (\C{outRgn}) was opened for allocation
and the \C{mapFn} is applied under this allocation context. Also note
that the input tuple is cloned under this allocation context.
Consequently, the result (\C{outp}) of applying the \C{mapFn} (line
22), which is subsequently added to \C{outMsg}, is guaranteed to be in
the \C{outRgn} regardless of the semantics of \C{mapFn}. On the other
hand, if the input tuple is not cloned (i.e., \C{inp} instead of
\C{inp.clone()} on line 22) and the \C{mapFn} simply projects its
input, then \C{outp} refers to an object inside \C{inp}, which no
longer exists when the \C{outRgn} is transferred to a downstream
actor. Therefore, if \C{mapFn} is applied over \C{inp} instead of
\C{inp.clone()}, then it better be a function that creates and returns
a new \C{TOut} object. Under such scenario, the region type of
\C{SelectVertex.mapFn} should be precise enough to prevent it from
being assigned a function (line 6) that simply returns its input.
Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
capable of capturing such nuances in the type of
\C{SelectVertex.mapFn}. Furthermore, the type can be automatically
inferred by \name's region type inference
(\S~\ref{sec:type-inference}), which can peform the above non-trivial
reasoning on behalf of the programmer.

### 2015-09-15 paper/type-system.tex
Sec 3. First commit.

\section{The Region Type System}
\label{sec:type-system}

\input{fb-syntax}
\input{fb-staticsem}

We build on the Featherweight Generic Java (FGJ)~\cite{fgj} formalism
to formalize \name and its region type system.
Fig~\ref{fig:fb-syntax} describes the syntax of the formal language,
which we call {\sc Featherweight \name} (FB).

### 2015-09-24 paper/broom.tex
Sec 3. First commit.
Added more type rules.
First commit with *all* type rules.
Bugfixes in rules.
Getting rid of stmts - first commit.
Complete type system with lambdas.
Bugfixes. Added "new Region" type rule.

@@
 
-\documentclass{sigplanconf}
+\documentclass[pldi]{sigplanconf}
 
@@
 \newcommand{\envwf}[1]{\vdash\,#1}
-\newcommand{\tywf}[2]{#1\,\vdash\,#2}
-\newcommand{\hastyp}[3]{#1\,\vdash\,#2\,:\,#3}
+\newcommand{\tywf}[2]{#1\,\vdash\,#2 \; \texttt{ok}}
+\newcommand{\fgjtywf}[2]{#1\,\Vdash\,#2 \; \texttt{ok}}
+\newcommand{\isvalid}[2]{#1\,\vdash\,#2}
+\newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
+\newcommand{\hasfgjtyp}[3]{#1\,\Vdash^{}\,#2\,:\,#3}
 \newcommand{\hassort}[3]{#1\,\vdash\,#2\,::\,#3}
 \newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
+\newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
 \newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
@@
 \newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
-\newcommand{\rel}[1]{{R}_{\mathit{#1}}}
 \newcommand{\absof}[1]{\left\| #1 \right\| }
@@
 \newcommand{\name}{{\sc Broom}\xspace}
+\newcommand{\fbname}{{\sc Featherweight Broom}\xspace}
+\newcommand{\FB}{{\sc FB}\xspace}
 \newcommand{\naiad}{{\sc Naiad}\xspace}
@@
 \newcommand{\outlives}{\succeq}
+\newcommand{\coloneqq}{::=}
+\newcommand{\R}[1]{\textrm{#1}}
+\newcommand{\N}[1]{{\normalfont #1}}
+\newcommand{\M}[1]{\mathtt{#1}}
+\newcommand{\ObjZ}{\C{Object}}
+\newcommand{\RgnZ}{\C{Region}}
+\newcommand{\thisZ}{\C{this}}
+\newcommand{\superZ}{\C{super}}
+\newcommand{\unitZ}{\C{unit}}
+\newcommand{\FuncZ}{\C{Func}}
+\newcommand{\unitval}{\C{()}}
+\newcommand{\inang}[1]{\langle #1 \rangle}
+\newcommand{\rhoalloc}{\rho^a}
+\newcommand{\rhobar}{\bar{\rho}}
+\newcommand{\rhoallocm}{\rho^a_m}
+\newcommand{\rhobarm}{\bar{\rho_m}}
+\newcommand{\ralloc}{\pi^a}
+\newcommand{\rgn}{\pi}
+\newcommand{\rbar}{\bar{\pi}}
+\newcommand{\tbar}{\bar{T}}
+\newcommand{\taubar}{\bar{\tau}}
+\newcommand{\xbar}{\bar{x}}
+\newcommand{\vbar}{\bar{v}}
+\newcommand{\ebar}{\bar{e}}
+\newcommand{\extends}{\triangleleft}
+\newcommand{\rextends}{\blacktriangleleft}
+\renewcommand{\bar}[1]{\overline{#1}}
+\newcommand{\angR}{\inang{\rhoalloc\rhobar \,|\, \phi}}
+\newcommand{\angT}{\inang{\tbar}}
+\newcommand{\tyvar}{a}
+\newcommand{\tyvarb}{b}
+\newcommand{\angAlpha}{\inang{\bar{\tyvar} \extends \bar{\fgjN}}}
+\newcommand{\fgjN}{K}
+\newcommand{\fbN}{N}
+\newcommand{\headerOf}[1]{\C{class}\; #1\angR\angAlpha \extends \fbN}
+\newcommand{\substFn}{\mathcal{S}}
+\newcommand{\mang}{\inang{\rhoalloc_m\bar{\rho_m} \,|\, \phi_m}}
+\newcommand{\rhoset}{\Sigma}
+\newcommand{\rhoenv}{\Delta}
+\newcommand{\aenv}{\Theta}
+\newcommand{\phicx}{\Phi}
+\newcommand{\subtypcx}{\rhoset,\rhoenv,\aenv,\phicx}
+\newcommand{\A}{{\mathcal{A}}}
+\newcommand{\exptycx}[2]{\A,#1,#2}
+\newcommand{\void}{\C{void}}
+\newcommand{\functy}[3]{\FuncZ\inang{#1}\inang{#2,#3}}
+\newcommand{\stmtsemcx}{\A,\env}
+\newcommand{\stmtsemcxp}{\A',\env'}
+\newcommand{\stmtsemcxpp}{\A'',\env''}
+\newcommand{\sredsto}{\Rightarrow}
+\newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
+\newcommand{\okin}[2]{#1 \; \texttt{ok \; in} \; #2}
+\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2).#3}
+\newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
+\newcommand{\letexp}[3]{\C{let}\;#1\,=\,#2\;\C{in}\;#3}
+\newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
+\newcommand{\open}[3]{\C{open}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
+\newcommand{\openalloc}[3]{\C{openalloc}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
+\newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
 
@@
 collection scheme employed by C\#.Net runtime is not very effective
-for dataflow programs insofar as the latency due to GC is intolerably
-high in many practical examples. We propose a memory management scheme
-that combines user-managed memory regions with automatic garbage
-collection, and lets programmers exploit specific memory behaviours
-exhibited by dataflow programs. To protect programmers from the
-dangers of manual memory management, we supplement our memory
-management scheme with a region type system that statically ensures
-the memory safety of region-annotated C\# programs, even in presence
-of higher-order stateful functions. Further, we describe a type
-inference algorithm that infers principal region types for first-order
-programs thereby saving programmers the need to write region
-annotations on types. Our inference algorithm is also very effective,
-albeit incomplete, for higher-order programs. As a testament to
-efficacy of our approach, we demonstrate performance improvements over
-a range of realistic dataflow programs that were rewritten to use
-region-based manual memory management, but are nevertheless judged to
-be safe by our type system.
+for dataflow programs to the extent that the latency due to GC is
+intolerably high in many practical examples. We propose a memory
+management scheme that combines user-managed memory regions with
+automatic garbage collection, and lets programmers exploit specific
+memory behaviours exhibited by dataflow programs. To protect
+programmers from the dangers of manual memory management, we
+supplement our memory management scheme with a region type system that
+statically ensures the memory safety of region-annotated C\# programs,
+even in presence of higher-order stateful functions. Further, we
+describe a type inference algorithm that infers principal region types
+for first-order programs thereby saving programmers the need to write
+region annotations on types. Our inference algorithm is also very
+effective, albeit incomplete, for higher-order programs. As a
+testament to efficacy of our approach, we demonstrate performance
+improvements over a range of realistic dataflow programs that were
+rewritten to use region-based manual memory management, but are
+nevertheless judged to be safe by our type system.
 
@@
 
+\input{type-system}
+
 % \appendix

### 2015-10-14 paper/overview.tex
Text added to Sec.3

@@
 Simplest of the region-specific constructs in \name is a \C{letregion}
-lexical block that creates a new labeled memory region. The semantics
-of the \C{letregion} block is similar to Tofte and
-Talpin~\cite{ttpopl94}'s \C{letregion} expression; the newly
-introduced region is live only inside the block, where its label is in
-scope for \C{new} expressions to make allocations inside the region.
-For example, in the following code, object \C{a} is allocated in the
-region \C{R0} introduced by the \C{letregion}:
+lexical block that creates a new memory region adressable via its
+static identifier. The semantics of the \C{letregion} block is similar
+to Tofte and Talpin~\cite{ttpopl94}'s \C{letregion} expression; the
+newly introduced region is live only inside the block, where its
+identifier is in scope for \C{new} expressions to make allocations
+inside the region. For example, in the following code, object \C{a} is
+allocated in the region with identifier \C{R0} introduced by the
+\C{letregion}:
 \begin{center}
@@
 \end{center}
-The code creates two stack regions - \C{R0} and \C{R1}, where \C{R0}
-outlives \C{R1} (denoted as $\C{R0 \outlives R1}$). Objects \C{a0} and
-\C{a1} are allocated in regions \C{R0} and \C{R1}, respectively. The
-first assignment statement assigns to \C{a1.x} an object allocated in
-outer region (\C{R0}). This assignment is absolutely safe as \C{a1.x}
-refers to a longer living object, hence is guaranteed to be a valid
-reference throughout the lifetime of \C{a1}. In contrast, the second
-assignment is unsafe, as it assigns to \C{a0.x} an object, whose
-lifetime is shorter than the lifetime of \C{a0}, making it unsafe to
-dereference \C{a0.x} outside the inner block. 
+The code creates two stack regions with identifiers \C{R0} and \C{R1},
+where the region \C{R0} outlives the region \C{R1} (denoted as $\C{R0}
+\outlives \C{R1}$).  Objects \C{a0} and \C{a1} are allocated in regions
+\C{R0} and \C{R1}, respectively. The first assignment statement
+assigns to \C{a1.x} an object allocated in outer region (\C{R0}). This
+assignment is safe as \C{a1.x} refers to a longer living object, hence
+is guaranteed to be a valid reference throughout the lifetime of
+\C{a1}.  In contrast, the second assignment is unsafe, as it assigns
+to \C{a0.x} an object, whose lifetime is shorter than the lifetime of
+\C{a0}, making it unsafe to dereference \C{a0.x} outside the inner
+block. 
 % Unsafe assignments can also happen indirectly via a function call..
@@
 
+\paragraph{A Note on region identifiers} In \name, every \C{letregion}
+expression introduces a new stack region with a static identifier that
+uniquely identifies the region within its scope. At runtime, a
+\C{letregion} expression could be evaluated multiple times if, for
+instance, it is wrapped inside a \C{while} loop or a recursive method,
+introducing multiple stack regions with the same static identifier.
+Which among these regions the identifier is bound to under a given
+runtime context is then determined by the scope of the identifier.
+Furthermore, any proposition involving static region identifiers is
+true if and only if the proposition is true in \emph{all} possible
+runtime contexts containing bindings for those identifiers. For
+instance, consider the following example:
+%\begin{minipage}{1.55in}
+\begin{codejava}
+for (int i=0; i<=10; i++) {
+  letregion R0 {
+    ...
+    letregion R1 {
+      A a1 = new@R1 A();
+    }
+  }
+}
+\end{codejava}
+%\end{minipage}
+%\begin{minipage}{1.4in}
+%\begin{codejava}
+%void m(int i) {
+%  letregion R0 {
+%    A a = new@R0 A();
+%  }
+%  this.m(++i);
+%}
+%...
+%this.m(1);
+%\end{codejava}
+%\end{minipage}
+When \C{i=2}, the \C{new} expression refers to the \C{R1} (and the
+corresponding stack region) introduced in the 2nd iteration of the
+loop, as it is the only \C{R1} in scope. Furthermore, the proposition
+\emph{\C{a1} refers to an object of type \C{A} contained in region
+\C{R1}} (denoted succinctly as $\C{a1}:\C{A@R1}$) is true because it
+is true for all stack regions identified by \C{R1} at runtime.
+Likewise, the propostion $\C{R0} \outlives \C{R1}$ is true because it
+is true in every runtime context that binds \C{R0} and \C{R1} to
+regions .
+
+Note that it is possible for multiple stack regions introduced by
+evaluating a \C{letregion} multiple times in a recursive definition to
+be in outlives relationship. For instance, in the following example:
+\begin{codejava}
+void m(A a) {
+  letregion R0 {
+    B b = new@R0 B();
+    b.x = a;
+    this.m(new@R0 A());
+  }
+}
+\end{codejava}
+stack regions identified by \C{R0} introduced in successive recursive
+calls are in outlives relationship. Consequently, the assignment to
+\C{b.x} should be considered safe. Region type system has to allow for
+this possibility.
+
 \subsection{Allocation Context}
+\label{sec:alloc-ctxt}
 

### 2015-10-14 paper/type-system.tex
Added more type rules.
First commit with *all* type rules.
Bugfixes. Added "new Region" type rule.
Text added to Sec.3

@@
 \input{fb-syntax}
+\input{fb-auxdef}
 \input{fb-staticsem}
+\input{fb-morewfrules}
 
@@
 Fig~\ref{fig:fb-syntax} describes the syntax of the formal language,
-which we call {\sc Featherweight \name} (FB).
+which we call \fbname (\FB). The language is seeded with \ObjZ and
+\RgnZ classes. More classes can be defined using the \C{class}
+keyword. Class types in \FB are region-annotated variants of class
+types in FGJ (also called \emph{core types}). This correspondence is
+reflected in the $T@\rgn$ syntax of the region type of an object,
+whose core type is $T$, and which is contained in the region $\rgn$.
+We let $\rgn$ range over static identifiers of regions in \FB. Note
+that all \FB objects are boxed values, hence their \FB types are
+always region-annotated. The only unboxed value in \FB is $\unitval$
+of type \unitZ.
+
+\FB class definitions can be understood as extensions of FGJ class
+definitions with region annotations and region parameters. More
+precisely, a \fbname definition of a class is obtained from its
+Featherweight Generic Java definition by (a). lifting all core types
+in the body of the class (i.e, fields, constructors and methods) to
+their region-annotated versions, and (b). adding region parameters to
+its header. Note that region parameterization is independent of its
+parameterization over core types; they are not conflated to make the
+class definition parameteric over region types. We let $a, b$ range
+over (core) type parameters and $\rho$ range over region parameters of
+a class. The following definition of \C{Pair} class is an example of
+type and region parametric definition in \FB (The symbol $\extends$
+should be read \emph{extends}):
+\begin{codejava}[mathescape=true]
+class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
+                      \conj \rho_2 \outlives \rho^a$>
+          <a $\extends$ Object, b $\extends$ Object> {
+  a@$\rho_1$ fst; 
+  b@$\rho_2$ snd;
+  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
+    super(); 
+    this.fst = fst; 
+    this.snd = snd;
+  }
+  a@$\rho_1$ getFst() {
+    return this.fst;
+  }
+}
+\end{codejava}
+Every class definition in \FB is necessarily parameterized over the
+allocation region of its objects. As a rule, allocation region
+parameter (denoted $\rho^a$) is the first region parameter of a class.
+Besides the allocation region of its objects, \C{Pair} class is also
+parametric over the regions its first and second elements are
+allocated in. References between objects allocated in different
+regions are only allowed if the referred object is guaranteed to
+outlive the referring object. In case of \C{Pair} class, this means
+that allocation regions ($\rho_1$ and $\rho_2$) of both objects that
+make up the pair must outlive the allocation region ($\rho^a$) of the
+\C{Pair} object. Such conditions over region parameters of a class
+need to be recorded in its header as region constraints ($\phi$) in
+order for the class to be judged well-formed by the type system
+(Fig.~\ref{fig:fb-morewfrules}).
+
+To construct objects of the \C{Pair} class, its type and region
+parameters need to be instantiated with core types ($T$) and concrete
+region identifiers ($\rgn$), respectively. Region instantiation has to
+satisfy the stated constraints over region parameters. For example,
+the following code snippet appropriately instantiates region
+parameters of the \C{Pair} class to construct a \C{Pair} of
+{\ObjZ}{\!}s, each allocated in different regions:
+\begin{codejava}
+letregion $\rgn_0$ {
+  Object<$\rgn_0$> fst = new Object<$\rgn_0$>();
+  letregion $\rgn_1$ {
+    Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
+    Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$> p = 
+       new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>(fst,snd);
+  }
+}
+\end{codejava}
+Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
+region constraints of the \C{Pair} class are satisfied. Observe that
+the region type of pair object \C{p} conveys the fact that (a). it is
+allocated in region $\rgn_1$, and (b). it holds references to
+object(s) allocated in region $\rgn_0$. On the other hand, if we
+choose to allocate the \C{fst} object also in $\rgn_1$, the region
+type of \C{p} would be
+\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
+written simply as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, the
+\C{@} notation in a region type of an object \C{x} highlights that
+\C{x}, and all the objects reachable from \C{x} via references are
+allocated in a single region. We say that \C{x} is \emph{contained} in
+the region. The objects of $\ObjZ$ class, by the virtue of
+containing no references to other objects, are always contained in
+their allocation region. Hence the type $\ObjZ@\rgn$ is equivalent to
+the type $\ObjZ\inang{\rgn}$.
+%%An object \emph{allocated} in a region $\rgn$ contains its spine in
+%%$\rgn$, but can refer to objects allocated in other regions.  On the
+%%other hand,
+
+A method definitions ($d$) in \fbname can be region polymorphic with
+respect to (a). its allocation context (Sec.~\ref{sec:alloc-ctxt}),
+and (b). the regions occuring in the region types of its arguments.
+Region parameters on the methods, like those on classes, are
+accompanied by constraints ($\phi$) capturing the conditions that the
+parameters need to satisfy for the method to be considered well-formed
+(Fig.~\ref{fig:fb-morewfrules}). Allocation context ($\rho^a$ or
+$\rho^a_m$) is the first and inevitable region parameter of every
+method in \FB. If a method is not polymorphic with respect to its
+allocation context (for example, if its allocation context needs to be
+same as the allocation region of its class), then the monomorphism
+needs to be encoded as an explicit equality constraint in $\phi$.
+
+\fbname supports first-class functions via lambda expressions and
+arrow types. A lambda expression defines a region-polymorphic
+multi-argument function closure parameterized over function's
+allocation context parameter. The arrow type of a function closure is
+prenex-quantified over its constrained region arguments
+($\inang{\rho^a\rhobar \,|\, \phi}$). Since closures can escape the
+context in which they are created, it is important to keep track of
+the regions where closures are allocated in order to prevent unsafe
+dereferences. The $\rgn^a$ annotation above the arrow in the arrow
+type denotes the allocation region of the corresponding closure. Note
+that it is important to distinguish between the allocation context
+argument of a function and the allocation region of its closure. For
+instance, in the following example:
+\begin{codejava}
+letregion $\rgn$ {
+  let f = $\lambda{\inang{\rho^a}}$()$\,\Rightarrow\,$new Object$\inang{\rho^a}$() 
+  in f
+}
+\end{codejava}
+The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
+\ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
+in $\rgn$, and (b). when executed under an allocation context
+$\rho^a$, the closure returns an object allocated in $\rho^a$.

### 2015-10-15 paper/overview.tex
Edits.

@@
 
-\paragraph{A Note on region identifiers} In \name, every \C{letregion}
-expression introduces a new stack region with a static identifier that
-uniquely identifies the region within its scope. At runtime, a
-\C{letregion} expression could be evaluated multiple times if, for
-instance, it is wrapped inside a \C{while} loop or a recursive method,
-introducing multiple stack regions with the same static identifier.
-Which among these regions the identifier is bound to under a given
-runtime context is then determined by the scope of the identifier.
-Furthermore, any proposition involving static region identifiers is
-true if and only if the proposition is true in \emph{all} possible
-runtime contexts containing bindings for those identifiers. For
-instance, consider the following example:
+\paragraph{A Note on region identifiers} Every stack region in \name
+is associated with a static identifier that uniquely identifies the
+region within its scope. At runtime, if a \C{letregion} expression is
+evaluated multiple times in a loop or a recursive method, the
+corresponding identifier is bound to a new stack region each time. 
+Any proposition involving static region identifiers is
+considered true at a program location if and only if the proposition
+is true under all possible evaluation contexts of that program
+location. For instance, consider the following example:
 %\begin{minipage}{1.55in}
-\begin{codejava}
+\begin{numcodejava}
 for (int i=0; i<=10; i++) {
   letregion R0 {
-    ...
     letregion R1 {
       A a1 = new@R1 A();
+      ...
     }
@@
 }
-\end{codejava}
+\end{numcodejava}
 %\end{minipage}
@@
 %\end{minipage}
-When \C{i=2}, the \C{new} expression refers to the \C{R1} (and the
-corresponding stack region) introduced in the 2nd iteration of the
-loop, as it is the only \C{R1} in scope. Furthermore, the proposition
-\emph{\C{a1} refers to an object of type \C{A} contained in region
-\C{R1}} (denoted succinctly as $\C{a1}:\C{A@R1}$) is true because it
-is true for all stack regions identified by \C{R1} at runtime.
-Likewise, the propostion $\C{R0} \outlives \C{R1}$ is true because it
-is true in every runtime context that binds \C{R0} and \C{R1} to
-regions .
+The identifiers \C{R0} and \C{R1} are bound to new stack regions each
+time the loop is evaulated. Nonetheless, the propositions (a). $\C{R0}
+\outlives \C{R1}$, and (b). $\C{a1}:\C{A@R1}$ (read as \emph{\C{a1}
+refers to an object of type \C{A} contained in region \C{R1}}) are
+true at line 5, as they are true under all possible bindings of \C{R0}
+and \C{R1} at line 5.
 
-Note that it is possible for multiple stack regions introduced by
-evaluating a \C{letregion} multiple times in a recursive definition to
-be in outlives relationship. For instance, in the following example:
-\begin{codejava}
-void m(A a) {
-  letregion R0 {
-    B b = new@R0 B();
-    b.x = a;
-    this.m(new@R0 A());
-  }
-}
-\end{codejava}
-stack regions identified by \C{R0} introduced in successive recursive
-calls are in outlives relationship. Consequently, the assignment to
-\C{b.x} should be considered safe. Region type system has to allow for
-this possibility.
+% Interestingly, it is possible for multiple stack regions introduced by
+% evaluating a the same \C{letregion} expression to be in outlives
+% relationship. For instance, in the following example:
+% \begin{codejava}
+% void m(A a) {
+%   letregion R0 {
+%     B b = new@R0 B();
+%     b.x = a;
+%     this.m(new@R0 A());
+%   }
+% }
+% \end{codejava}
+% Stack regions identified by \C{R0} introduced in successive recursive
+% calls are in outlives relationship. Consequently, the assignment to
+% \C{b.x} should be considered safe. Region type system has to allow for
+% this possibility.
 
-\subsection{Allocation Context}
+\subsection{Allocation Context and Qualified Region Polymorphism}
 \label{sec:alloc-ctxt}
@@
 As evident from the examples shown above, the suffix \C{@R} in
-\C{new@R} instruct the \C{new} keyword to allocate the newly created
+\C{new@R} instructs the \C{new} keyword to allocate the newly created
 object in the region labeled \C{R}. However, \name does not require
@@
 absence of explicit specification, \name allocates the newly created
-object in its \emph{allocation context}, which is the region
-introduced by the closest enclosing \C{letregion} block. Besides being
-a syntactic convenience, the facility to elide allocation region
-specification is imperative to interface region-oblivious standard
-library code with region-aware application code. For instance, the
-following \name code, which makes explicit use of regions, calls
-the standard library's iterator methods, which are region-oblivious:
+object in its \emph{allocation context}, which is the region at the
+top of the stack when \C{new} is executed. For instance, the
+\C{findMin} function shown below, which makes explicit use of regions,
+calls the standard \C{List} library's \C{listIterator} method:
 \begin{center}
@@
 \end{center}
-Due to the absence of new region creations, and explicit region
-annotations with \C{new} expressions, all new objects created by a
-standard library method are allocated in the allocation context for
-the method call. In the above example, the allocation context for the
-call to \C{List} library's \C{listIterator} method is the region
-\C{R1}. Therefore, all objects created by the method, including the
-iterator object, are allocated in \C{R0}. This fact should be captured
-appropriately by the region type of \C{listIterator} method if the
-type system were to prove the safety of the above code.
+The \C{listIterator} method neither creates new regions, nor does it
+annotate \C{new} expressions with region specifications. Consequently,
+all new objects created by the method are allocated in the allocation
+context for the call to \C{listIterator}, which in this case is
+\C{R0}.
+
+Observe that \C{listIterator} can be called under multiple different
+allocation contexts, and each time it returns a \C{ListIterator}
+allocated in its allocation context. The iterator object might hold
+references to the list, requiring the list to be allocated in a region
+that outlives \C{listIterator}\!'s allocation context. However, modulo
+this constraint, \C{listIterator} is not concerned about where the
+list is allocated. As such, \C{listIterator} is
+\emph{region-polymorphic} with respect to (a). its allocation context
+argument, and (b). the allocation region of the list, subject to the
+constraint that the later outlives the former. We call such region
+polymorphism with constraints in \name as \emph{qualified region
+polymorphism}. The provision to elide allocation region
+specifications, and the ability to infer qualified region-region
+polymorphic types are pivotal to interface region-oblivious standard
+library code with region-aware application code in \name.
 
@@
 \end{center}
+Accordingly, \name's region type system lets the former assignment
+pass by, but flags the later as an error.
 
@@
 suitable for applications that need to transfer data between
-autonomous actors, such as \naiad dataflow operators. To support these
-applications, it is imperative to support regions whose lifetimes
-transcend the boundaries of lexical blocks, functions, or even
-programs. Regions that fit this description in \name are called
+autonomous entities, such as \naiad dataflow operators. To support
+these applications, it is imperative to support regions whose
+lifetimes transcend the boundaries of lexical blocks, functions, or
+even programs. Regions that fit this description in \name are called
 \emph{dynamic regions}, or \emph{transferable regions}.
 
-Transferable regions, like stack regions, are memory regions
-Transferable regions are first class values of \name; they are
+Transferable regions, like stack regions, are memory regions, but
+unlike stack regions, they are first class values of \name; they are
 considered objects of \C{Region} class, and, like objects of other
 classes, they are created using the \C{new} keyword, can be passed as
-arguments, and returned from methods. Like a stack region, a
-transferable region facilitates arbitrary object allocations, but
-unlike a stack region, it has a well-defined \emph{root} object
-denoting the root of the data structure being transferred. In a
-typical dataflow computation, an upstream actor (e.g., a \C{SELECT}
-operator) constructs a transferable region, sets its root to the data
-structure supposed to contain intermediate results, populates the data
-structure with intermediate results, and then transfers it to a
-downstream actor (e.g., a \C{COUNT} operator), which performs further
-processing. Since a transferable region escapes the lifetime of the
-sender, there must be no references from inside of the transferable
-region to objects allocated in other memory regions of the sender.
-Such references, if exist, may become invalid references in the
-context of recipient's address space, jeopardizing memory safety.
-\name relies on its region type system to prevent such unsafe
-references from being created.
+arguments, stored in data structures, and returned from methods. Like
+a stack region, a transferable region facilitates arbitrary object
+allocations, but unlike a stack region, it has a well-defined
+\emph{root} object denoting the root of the data structure being
+transferred. In a typical dataflow computation, an upstream actor
+(e.g., a \C{SELECT} operator) constructs a transferable region, sets
+its root to the data structure containing intermediate results, and
+then transfers it to a downstream actor (e.g., a \C{COUNT} operator),
+which performs further processing. Since a transferable region escapes
+the lifetime of the sender, there must be no references from inside of
+the transferable region to objects allocated in other memory regions
+of the sender. Such references, if exist, may become invalid
+references in the context of recipient's address space, jeopardizing
+memory safety. \name relies on its region type system to prevent such
+unsafe references from being created.
 

### 2015-10-15 paper/type-system.tex
Edits.

@@
 parameterization over core types; they are not conflated to make the
-class definition parameteric over region types. We let $a, b$ range
+class definition parametric over region types. We let $a, b$ range
 over (core) type parameters and $\rho$ range over region parameters of
@@
 $\rho^a$, the closure returns an object allocated in $\rho^a$.
+
+The expression language admits expressions to instantiate
+($e\inang{\ralloc\rbar}$) region parameters, and to apply
+($e(\bar{e})$) functions. For a method call, however, region parameter
+instantiation and method application are captured in a single
+expression ($e.mn\inang{\ralloc \rbar}(\ebar)$). The language also
+admits a big-lambda expression, which, when used in conjunction with
+the instantiation expression, serves the bureaucratic purpose of
+renaming bound region variables in the type of a function closure.
+
+The

### 2015-10-16 paper/overview.tex
Edits.

@@
 
+\subsection{Memory Safety}
+\label{sec:memory-safety}
+
+Memory safety in \name is predominantly due to its region type system.
+However, the type system does not ensure the integrity of a small
+subset of pointers, called the \emph{unsafe set}, for which runtime
+verification is needed to guarantee safety. This is more a design
+decision rather than an unfortunate inevitability. Rest of this
+subsection presents an overview of the region type system, discusses
+the unsafe set, and briefly describes the runtime mechanisms used to
+ensure safety of references in the unsafe set.
+
+At a fundamental level, the region type system's approach to memory
+safety involves approximating (a). \emph{liveness}: the set of memory
+regions that are live at any given program location, and (b).
+\emph{outlives} relationships between them, and using this information
+to preempt potentially unsafe references from being created. These
+include:
+\begin{itemize}
+\item References into dead/non-existing regions,
+\item References from longer-living stack regions to shorter-living ones, and 
+\item References that escape transferable regions.
+\end{itemize}
+Unsurprisingly, the type system's approximation of liveness and
+outlives information for stack regions is precise. However, for
+transferable regions, statically determining liveness is equivalent to
+determining if a C pointer is null in presence of dynamic memory
+allocation, which is undecidable in general. Therefore, the region
+type system has to be necessarily conservative in approximating the
+liveness of transferable regions in order to ensure full memory
+safety. Maintaining a good conservative approximation requires the
+type system to employ techniques such as linear types~\cite{wadlerlt}
+or unique pointers~\cite{rust}, which impose restrictions that are
+often too hard to program around~\footnote{Internet has many tutorials
+on how to \emph{fight} the borrow checker in Rust.}. This is
+particularly unacceptable in the context of \name, as dataflow
+programs are often run offline, and the concern for performance and
+developer productivity outweighs the concern for safety.
+
+\name's region type system therefore overapproximates the liveness
+information for transferable regions, i.e., it might assume that a
+transferable region is live when, in fact, it might have already been
+transferred or freed. This necessarily leads to cases where creation of
+unsafe references, and their subsquent dereferencing can go unnoticed
+by the type system. Considering that stack regions can contain
+pointers into transferable regions, the worst case cost of
+overapproximating liveness is therefore the integrity of \emph{every}
+pointer on stack.
+
+Fortunately, \name avoids this worst case scenario. It does so by
+restricting the unsafe set to only include $\RgnZ$ objects. In
+particular, pointers \emph{into} transferable regions are excluded
+from this set (hence safe). \name achieves this by imposing simple
+restrictions on how $\RgnZ$ objects can be used to operate on
+transferable regions. The following describes the $\RgnZ$ class, and
+how $\RgnZ$ objects should be used to make allocations into, and to
+read from transferable regions.
+
 \begin{figure}
@@
 
+\subsection{An Example}
+
 Fig~\ref{fig:naiad-ex} shows \C{SelectVertex} class, which is a

### 2015-10-17 paper/introduction.tex
The type system ensures complete memory safety.

@@
   \item \name is equipped with a region type system that statically
-  guarantees safety for a well-defined subset of all memory accesses
-  in a well-typed program. This subset includes all memory accesses,
-  including accesses that happen across different regions, except
-  dereferences of \C{Region} objects themselves. For the later kind,
-  we rely on \GK{reference counting?} to ensure safety at run-time.
-  Our experiments with \naiad workloads show that \GK{x\%} of all
-  memory memory accesses fall under the statically safe subset, while
-  only \GK{100-x\%} require run-time mechanisms to ensure safety.
+  guarantees safety all memory accesses in a well-typed program,
+  subject to some conditions that can be checked efficiently at
+  run-time. The overhead of checking these conditions is less than
+  \GK{x\%} in our experiments with \naiad workloads.
 

### 2015-10-17 paper/overview.tex
The type system ensures complete memory safety.

--- 2015-10-15 paper/overview.tex
@@
 
-A transferable region is said to be in \emph{closed} state when
-created. \name provides \C{openalloc} lexical construct to \emph{open}
-a transferable region as an allocation context, and to access its root
-object:
+\name imposes a simple discipline on how $\RgnZ$ objects can be used
+to perform operations on transferable regions. This discipline is
+imposed in the interest of memory safety, which is discussed in the
+following subsection.
+
+\subsection{Memory Safety}
+\label{sec:memory-safety}
+
+Memory safety in \name is primarily due to its region type system,
+although few runtime checks are needed.
+% However, the type system does not ensure the integrity of a small
+% subset of pointers, called the \emph{unsafe set}, for which runtime
+% verification is needed to guarantee safety. This is more a design
+% decision rather than an unfortunate inevitability, as the following
+% discussion elaborates.
+At a fundamental level, the region type system's approach to memory
+safety involves approximating (a). the set of memory regions that are
+live at any given program location\footnote{We say the type system
+tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
+relationships between them, and using this information to preempt
+potentially unsafe references from being created. These include:
+\begin{itemize}
+\item References into dead/non-existing regions,
+\item References from longer-living stack regions to shorter-living ones, and 
+\item References that escape transferable regions.
+\end{itemize}
+Maintaining precise approximation of liveness and outlives information
+for stack regions is easy. However, for dynamic (transferable) regions,
+statically determining liveness is equivalent to determining if a C
+pointer is null in presence of dynamic memory allocation, which is
+undecidable in general. Systems that attempt to ensure full memory
+safety in presence of dynamic memory (e.g., ~\cite{henglein},
+~\cite{rust}) do so by conservatively approximating liveness
+information for memory (regions). Maintaining a good conservative
+approximation requires them to employ techniques such as linear
+types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
+restrictions that are often too hard to program
+around\footnote{Internet has many tutorials on how to \emph{fight} the
+borrow checker in Rust.}. This is particularly unacceptable in the
+context of \name, as dataflow programs are often run offline, and the
+concern for performance and developer productivity outweighs the
+concern for safety.
+
+\name's region type system therefore adopts the novel approach of
+overapproximating liveness information for transferable regions, i.e.,
+it might assume that a transferable region is live when, in fact, it
+might have already been transferred or freed. This imprecision should
+be a cause of concern as it can lead to cases where creation of unsafe
+references, and their subsquent dereferencing go unnoticed by the type
+system. Considering that stack regions can contain pointers into
+transferable regions, the worst case cost of overapproximating
+liveness is therefore the integrity of \emph{every} pointer on stack.
+
+Fortunately, \name avoids this worst case scenario.  \name achieves
+this by imposing a simple state transition discipline on the lifetime
+of a transferable region.  The relevant finite state machine is shown
+in Fig.~\ref{fig:region-fsm}.
+
+\begin{figure}
+\includegraphics[scale=0.45]{region-fsm.png}
+\caption{The lifetime of a dynamic (transferable) region in \name}
+\label{fig:region-fsm}
+\end{figure}
+
+A transferable region starts its lifetime in a \emph{closed} state,
+when it is created through \C{new Region} expresion. \name provides
+\C{open} lexical construct to \emph{open} a transferable region as an
+allocation context, and to access its root object:
 \begin{codejava}
-  Region<string> rgn = new Region<string>
-                        (() => "Hello");
-  openalloc rgn withroot str {
-    print str; //prints "Hello"
-    str = str + " World!"; 
+  Region<string> rgn = new Region<List<String>>
+                        (() => new List<String>());
+  open rgn as R0 withroot strList {
+    strList.addAtHead("World");
+    strList.addAtHead("Hello");
+  }
+  rgn.transfer();
+\end{codejava}
+In the above code, the \C{open} lexical block performs the
+following operations: (a). It opens the transferable region handled by
+\C{rgn} for allocation (i.e., makes it the current allocation
+context), (b). assigns the identifier \C{R0} to the open region, and
+(c). sets the newly introduced local variable (\C{strList}) to refer
+to its root object (a list of strings). As Fig.~\ref{fig:region-fsm}
+indicates, it is not possible to open a region that is already
+transferred/freed. The fact that \C{rgn} is open within the block
+therefore guarantees that it is not yet transferred/freed, and that
+dereferencing \C{strList} within the block is safe. The end of
+\C{open} block marks the return of transferable region to the closed
+state.  While closed, the region is eligible to be transferred to a
+downstream actor, or to be freed.  An actor that receives the
+transferable region, receives it in the closed state. It can then
+reopen the received region to read its contents, possibly add more
+data and transfer it to another actor, or free the region.
+
+\name lets stack regions to be used as working memory while operating
+with the data stored in a transferable region. Consider the following
+code, for instance\footnote{For brevity, we drop \C{as} part of the
+\C{open} expression whenever there is no need to assign a static
+identifier for the open region.}:
+\begin{codejava}
+void onReceive(Region<List<String>> rgn)
+  open rgn withroot strList {
+    letregion R1 {
+      String s = "";
+      ListIterator<String> i = strList.listIterator();
+      while(i.hasNext()) {
+        s += i.getNext();
+      }
+      print s; //prints "HelloWorld"
+    }
+  }
+  rgn.free();
 }
 \end{codejava}
-In the above code, the \C{openalloc} lexical block opens \C{rgn} for
-allocation, and sets the newly introduced local variable (\C{str}) to
-refer to its root object.  The assignment statement within the block
-updates the root. Since \C{rgn} is the allocation context within the
-block, the new string is allocated inside \C{rgn}, hence the
-assignment is safe. In some cases, it is desirable to open a
-transferable region only to read its contents, but not to perform any
-allocations. For this purpose, \name provides \C{open} lexical
-construct that, like \C{openalloc}, opens a transferable region and
-names its root object, but unlike \C{openalloc}, does not make it an
-allocation context. A transferable region can be \C{openalloc}'d or
-\C{open}'d many times, before it is eventually \C{free}'d or
-\C{transfer}'d.
+The stack region \C{R1} is being used in the above code to provide
+working memory to work with the objects of transferable region
+\C{rgn}. Since a transferable region cannot be transferred/freed while
+it is still open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to
+outlive the stack region \C{R1} in the above code, making it safe for
+the later to contain references to the former. \name therefore allows
+such references.
 
-The design decisions reflected above, \emph{viz.}, to ascribe a root
-object to every transferable region, to make use of a higher-order
-argument to initialize the root object, and to require that a region
-be explictly opened in a lexical block for allocations, were made to
-reflect the message semantics of transferable regions, to promote
-uniform style of programming with different kinds of regions, and to
-precisely state and enforce the safety guarantees offered by \name's
-region type system (\S~\ref{sec:type-system}).
+% The design decisions reflected above, \emph{viz.}, to ascribe a root
+% object to every transferable region, to make use of a higher-order
+% argument to initialize the root object, and to require that a region
+% be explictly opened in a lexical block for allocations, were made to
+% reflect the message semantics of transferable regions, to promote
+% uniform style of programming with different kinds of regions, and to
+% precisely state and enforce the safety guarantees offered by \name's
+% region type system (\S~\ref{sec:type-system}).
+
+As evident in the above examples, the safety of memory accesses in
+\name is now subject to the condition that the lifetime of every
+transferable region aligns with the state transition discipline shown
+in Fig.~\ref{fig:region-fsm}. If the transition discipline is somehow
+guaranteed, then \name's region type system statically guarantees the
+safety of all memory accesses. In other words, the type system reduces
+the problem of ensuring memory safety in \name programs to the problem
+of enforcing state transition discipline on the lifetimes of
+transferable regions. In \name, this enforcement is done at runtime by
+explicitly keeping track \emph{current state} for $\RgnZ$ objects, and
+preventing them from going through invalid transitions.
 
 \input{naiad-ex}
+
+\subsection{An Example}
 

### 2015-10-18 paper/type-system.tex
Edits.

@@
-\section{The Region Type System}
+\section{The Core Language}
 \label{sec:type-system}
@@
 
+\subsection{Syntax}
+\label{sec:fb-syntax}
+
 We build on the Featherweight Generic Java (FGJ)~\cite{fgj} formalism
-to formalize \name and its region type system.
-Fig~\ref{fig:fb-syntax} describes the syntax of the formal language,
+to formalize \name and its region type system. Our development reuses
+notations from~\cite{fgj}, and relies on several of its definitions,
+such as the definition of type well-formedness for the core
+(region-free) language. Due to space constraints, we are unable to
+reproduce them here, and instead encourage the reader to refer
+to~\cite{fgj}.
+
+Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
 which we call \fbname (\FB). The language is seeded with \ObjZ and
@@
 
-\FB class definitions can be understood as extensions of FGJ class
-definitions with region annotations and region parameters. More
-precisely, a \fbname definition of a class is obtained from its
-Featherweight Generic Java definition by (a). lifting all core types
-in the body of the class (i.e, fields, constructors and methods) to
-their region-annotated versions, and (b). adding region parameters to
-its header. Note that region parameterization is independent of its
-parameterization over core types; they are not conflated to make the
-class definition parametric over region types. We let $a, b$ range
-over (core) type parameters and $\rho$ range over region parameters of
-a class. The following definition of \C{Pair} class is an example of
-type and region parametric definition in \FB (The symbol $\extends$
-should be read \emph{extends}):
+The expression language of FGJ is impoverished, lacking, among others,
+an assignment expression. Note that it is possible to \emph{encode}
+assignments to instance variables (object fields) in FGJ by creating a
+new object with its fields set to new values. Unfortunately, such
+encoding is not semantics preserving in case of \name, as the new
+object could be allocated in a region different from the allocation
+region of the original object. To avoid such complications, we extend
+the expression language with assignments and local variable
+declarations (via \C{let} expressions). Instead of distinguishing
+between statements (e.g., assignments) and expressions (e.g., method
+calls) in \FB, we consider statements as expressions of \unitZ type.
+Statements that introduce lexical blocks in \name, such as
+\C{letregion} and \C{open} are, are converted to their expression form
+using the \C{in} keyword.
+
+The expression language has been equipped with lambda abstraction
+($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
+expressions to define and apply anonymous functions.  Modulo the angle
+braces containing $\rho$'s and $\phi$'s, whose presence will be
+justified later, these expressions are uncurried variants of
+corresponding expressions from simply typed lambda calculus. Note that
+FGJ formalism does not include higher-order functions, hence the
+syntactic class (\C{T}) of FGJ types needs to be extended with an
+(uncurried) arrow type.
+
+Informally, \FB class definitions are extensions of FGJ class
+definitions with regions. A class definition in FGJ can be lifted to
+its \FB definition by (a). lifting all core types in the body of the
+class (i.e, fields, constructors and methods) to their
+region-annotated versions, and (b). parameterizing the class over
+region variables that occur free in the region-annotated body of the
+class. Note that in \FB, regions, and their static identifiers
+($\rgn$) are confined to the expression language. Consequently, the
+region annotated types of class fields, constructors and methods do
+not refer to identifiers of concrete regions, but rather their place
+holders called \emph{region variables} ($\rho$). Informally, region
+variables ($\rho$) are to region identifiers ($\rgn$) as type
+variables ($a$) are to types ($T$).
+
+Note that when a generic class definition in FGJ is lifted to a
+region-polymorphic definition in \FB, it nonetheless remains generic
+with respect to core types. We say that the class is now a
+region-polymorphic generic class. Region parameterization of a class
+in \FB is independent of its parameterization over core types; they
+are not conflated to make the class parametric over region types. For
+instance, following is the definition of region-polymorphic generic
+\C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
 \begin{codejava}[mathescape=true]
@@
 \end{codejava}
-Every class definition in \FB is necessarily parameterized over the
-allocation region of its objects. As a rule, allocation region
+Every class definition in \FB is necessarily polymorphic with respect
+to the allocation region of its objects. As a rule, allocation region
 parameter (denoted $\rho^a$) is the first region parameter of a class.

### 2015-10-18 paper/type-system.tex
Edits.

@@
 types in FGJ (also called \emph{core types}). This correspondence is
-reflected in the $T@\rgn$ syntax of the region type of an object,
-whose core type is $T$, and which is contained in the region $\rgn$.
-We let $\rgn$ range over static identifiers of regions in \FB. Note
-that all \FB objects are boxed values, hence their \FB types are
-always region-annotated. The only unboxed value in \FB is $\unitval$
-of type \unitZ.
+reflected in the $T@\rgn$ syntax of a region type, which is the
+simplest form of a region type describing an object, of core type $T$
+contained in a region $\rgn$. We let $\rgn$ range over static
+identifiers of regions in \FB. Note that all \FB objects are boxed
+values, hence their \FB types are always region-annotated. The only
+unboxed value in \FB is $\unitval$ of type \unitZ.
 
 The expression language of FGJ is impoverished, lacking, among others,
-an assignment expression. Note that it is possible to \emph{encode}
+an assignment expression. While it is possible to encode
 assignments to instance variables (object fields) in FGJ by creating a
-new object with its fields set to new values. Unfortunately, such
-encoding is not semantics preserving in case of \name, as the new
-object could be allocated in a region different from the allocation
-region of the original object. To avoid such complications, we extend
-the expression language with assignments and local variable
-declarations (via \C{let} expressions). Instead of distinguishing
-between statements (e.g., assignments) and expressions (e.g., method
-calls) in \FB, we consider statements as expressions of \unitZ type.
-Statements that introduce lexical blocks in \name, such as
-\C{letregion} and \C{open} are, are converted to their expression form
-using the \C{in} keyword.
+new object with its fields set to new values, such encoding is not
+semantics preserving in case of \name, as the new object could be
+allocated in a region different from the allocation region of the
+original object. To avoid such complications, we extend the expression
+language with assignments and local variable declarations (via \C{let}
+expressions). Instead of distinguishing between statements (e.g.,
+assignments) and expressions (e.g., method calls) in \FB, we consider
+statements as expressions of \unitZ type.  Statements that introduce
+lexical blocks in \name, such as \C{letregion} and \C{open} are, are
+converted to their expression form using the \C{in} keyword. We let
+$\rgn$ range over static region identifiers introduced by these
+expressions.
 
@@
 expressions to define and apply anonymous functions. Modulo the angle
-braces containing $\rho$'s and $\phi$'s, whose presence will be
-justified later, these expressions are uncurried variants of
-corresponding expressions from simply typed lambda calculus. Note that
-FGJ formalism does not include higher-order functions, hence the
-syntactic class (\C{T}) of FGJ types needs to be extended with an
-(uncurried) arrow type.
+braces, these expressions are uncurried variants of corresponding
+expressions from System F. Note that FGJ formalism does not include
+higher-order functions, hence the syntactic class (\C{T}) of FGJ types
+needs to be extended with an (uncurried) arrow type.
 
@@
 class (i.e, fields, constructors and methods) to their
-region-annotated versions, and (b). parameterizing the class over
+region-annotated versions, (b). parameterizing the class over
 region variables that occur free in the region-annotated body of the
-class. Note that in \FB, regions, and their static identifiers
+class, and (c). recording any outlives constraints on the region
+parameters of the class. The later two steps effectively make the
+class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
+polymorphism). Note that in \FB, regions, and their static identifiers
 ($\rgn$) are confined to the expression language. Consequently, the
@@
 are not conflated to make the class parametric over region types. For
-instance, following is the definition of region-polymorphic generic
+instance, following is the region-polymorphic definition of a generic
 \C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
@@
 \end{codejava}
-Every class definition in \FB is necessarily polymorphic with respect
-to the allocation region of its objects. As a rule, allocation region
-parameter (denoted $\rho^a$) is the first region parameter of a class.
-Besides the allocation region of its objects, \C{Pair} class is also
-parametric over the regions its first and second elements are
-allocated in. References between objects allocated in different
-regions are only allowed if the referred object is guaranteed to
-outlive the referring object. In case of \C{Pair} class, this means
-that allocation regions ($\rho_1$ and $\rho_2$) of both objects that
-make up the pair must outlive the allocation region ($\rho^a$) of the
-\C{Pair} object. Such conditions over region parameters of a class
-need to be recorded in its header as region constraints ($\phi$) in
-order for the class to be judged well-formed by the type system
-(Fig.~\ref{fig:fb-morewfrules}).
+When objects of a class are allocated in a region $\rgn$, it means
+that the class's constructor is run with $\rgn$ as its allocation
+context. Every class definition in \FB is necessarily polymorphic with
+respect to the allocation region of its objects, i.e., the allocation
+context of its constructor. We adopt a convention that requires the
+allocation region parameter (denoted $\rho^a$) to be the first region
+parameter of a class.  Besides the allocation region of its objects,
+\C{Pair} class is also parametric over the regions its first and
+second elements are allocated in. References between objects allocated
+in different regions are only allowed if the referred object is
+guaranteed to outlive the referring object. In case of \C{Pair} class,
+this means that allocation regions ($\rho_1$ and $\rho_2$) of both
+objects that make up the pair must outlive the allocation region
+($\rho^a$) of the \C{Pair} object. Such conditions over region
+parameters of a class need to be recorded in its header as region
+constraints ($\phi$) in order for the class to be judged well-formed
+by the type system (Fig.~\ref{fig:fb-morewfrules}).
 
@@
     Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
-    Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$> p = 
-       new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>(fst,snd);
+    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
+                  (fst,snd);
   }
@@
 Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
-region constraints of the \C{Pair} class are satisfied. Observe that
-the region type of pair object \C{p} conveys the fact that (a). it is
-allocated in region $\rgn_1$, and (b). it holds references to
-object(s) allocated in region $\rgn_0$. On the other hand, if we
-choose to allocate the \C{fst} object also in $\rgn_1$, the region
-type of \C{p} would be
+region constraints of the \C{Pair} class are satisfied, hence the
+instantiation is valid. Observe that the region type of \C{p} conveys
+the fact that (a). it is allocated in region $\rgn_1$, and (b). it
+holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
+In contrast, if we choose to allocate the \C{fst} object also in
+$\rgn_1$, the region type of \C{p} would be
 \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
@@
 allocated in a single region. We say that \C{x} is \emph{contained} in
-the region. The objects of $\ObjZ$ class, by the virtue of
-containing no references to other objects, are always contained in
-their allocation region. Hence the type $\ObjZ@\rgn$ is equivalent to
-the type $\ObjZ\inang{\rgn}$.
+the region. The $\ObjZ$ class in \FB contains no references to other
+objects, hence its objects are always contained in their allocation
+region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
+$\ObjZ@\rgn$), for some region $\rgn$.
 %%An object \emph{allocated} in a region $\rgn$ contains its spine in
@@
 %%other hand,
+
+\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
+$\RgnZ\inang{\rgn}$. However, $\rgn$ cannot be any region. Recall
+that $\RgnZ$ objects have special semantics in \name - they act as
+handlers to transferable regions. Constructing a new $\RgnZ$ object
+entails the creation of a new transferable region, and it is in this
+region that the new object is allocated in. It follows that $\rgn$ in
+$\RgnZ\inang{\rgn}$ should be the static identifier of the new
+transferable region. But, static identifiers for transferable
+regions are introduced only when such regions are opened for
+allocation via \C{open} expression. What, then, should be the region
+type of a \C{new Region} expression?
+
+\FB resolves this problem by existentially quantifying the allocation
+region of $\RgnZ$ objects when they are created. In other words, the
+type of \C{new Region} expression in \FB is
+$\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
+type captures the fact that there now exists a transferable region
+containing the newly constructed $\RgnZ$ object. Elimination of
+existential quantification is facilitated by the \C{open} expression,
+which opens the transferable region and assigns it an identifier
+($\rgn$).  Within the scope of \C{open}, the transferable region is
+identified with $\rgn$, allowing its handler to instantiate the
+existentially bound region variable with $\rgn$, and assume the type
+of $\RgnZ\inang{\rgn}$.
+
+Existentially quantifying the allocation region in the type of $\RgnZ$
+objects is also useful to assign appropriate types to data structures
+of transferable regions. For example, a list of transferable regions
+has the type $\C{List}\inang{??}$ in \FB.
 

### 2015-10-20 paper/type-system.tex
Edits.

@@
 ($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
-expressions to define and apply anonymous functions. Modulo the angle
-braces, these expressions are uncurried variants of corresponding
-expressions from System F. Note that FGJ formalism does not include
-higher-order functions, hence the syntactic class (\C{T}) of FGJ types
-needs to be extended with an (uncurried) arrow type.
+expressions to define and apply anonymous functions. These expressions
+are uncurried variants of the corresponding expressions from System F.
+Note that FGJ formalism does not include higher-order functions, hence
+the syntactic class (\C{T}) of FGJ types needs to be extended with an
+(uncurried) arrow type.
 
@@
 variables ($\rho$) are to region identifiers ($\rgn$) as type
-variables ($a$) are to types ($T$).
+variables ($a$) are to core types ($T$).
 
@@
 \FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
-$\RgnZ\inang{\rgn}$. However, $\rgn$ cannot be any region. Recall
-that $\RgnZ$ objects have special semantics in \name - they act as
-handlers to transferable regions. Constructing a new $\RgnZ$ object
-entails the creation of a new transferable region, and it is in this
-region that the new object is allocated in. It follows that $\rgn$ in
-$\RgnZ\inang{\rgn}$ should be the static identifier of the new
-transferable region. But, static identifiers for transferable
+$\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
+$\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
+region. Recall that $\RgnZ$ objects have special semantics in \name -
+they act as handlers to transferable regions.  Constructing a new
+$\RgnZ$ object entails the creation of a new transferable region, and
+it is in this region that the new object is allocated in. It follows
+that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
+the new transferable region. But, static identifiers for transferable
 regions are introduced only when such regions are opened for
@@
 
-Existentially quantifying the allocation region in the type of $\RgnZ$
-objects is also useful to assign appropriate types to data structures
-of transferable regions. For example, a list of transferable regions
-has the type $\C{List}\inang{??}$ in \FB.
+A method definition ($d$) can be region polymorphic with respect to
+(a). its allocation context (Sec.~\ref{sec:alloc-ctxt}), and (b). the
+regions occuring in the region types of its arguments.  Region
+parameters on the methods, like those on classes, are accompanied by
+constraints ($\phi$) capturing the conditions that the parameters need
+to satisfy for the method to be considered well-formed
+(Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
+or $\rho^a_m$) is the first and inevitable region parameter of every
+method in \FB. If a method is not intended to be polymorphic with
+respect to its allocation context (for example, if its allocation
+context needs to be same as the allocation region of its object), then
+the monomorphism needs to be encoded as an explicit equality
+constraint in $\phi$.
 
-A method definitions ($d$) in \fbname can be region polymorphic with
-respect to (a). its allocation context (Sec.~\ref{sec:alloc-ctxt}),
-and (b). the regions occuring in the region types of its arguments.
-Region parameters on the methods, like those on classes, are
-accompanied by constraints ($\phi$) capturing the conditions that the
-parameters need to satisfy for the method to be considered well-formed
-(Fig.~\ref{fig:fb-morewfrules}). Allocation context ($\rho^a$ or
-$\rho^a_m$) is the first and inevitable region parameter of every
-method in \FB. If a method is not polymorphic with respect to its
-allocation context (for example, if its allocation context needs to be
-same as the allocation region of its class), then the monomorphism
-needs to be encoded as an explicit equality constraint in $\phi$.
+Like methods, anonymous functions can also be region-polymorphic. 
+% A lambda expression defines a region-polymorphic
+% multi-argument function closure parameterized over function's
+% allocation context parameter. 
+The angle braces in the lambda expression ($\inang{\rho^a\rhobar \,|\,
+\phi}$) serve the same purpose as they do in a method definition - to
+capture region parameters along with their constraints. Region
+parameters also appear in the arrow type of the lambda expression 
+($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn} 
+\tau$) at the prenex position, similar to ML type schemes. However,
+unlike ML, we don't distinguish between region types and region type
+schemes; any of the $\tau$'s in the arrow type can themselves be
+region-parametric arrow types. In this respect, our region type system
+is more like System F's type system, which admits higher-rank
+parametric polymorphism. Like System F, \FB provides a region
+instantiation expression ($e\inang{\ralloc\rbar}$), and a region
+generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
+\phi}{e}$) to instantiate and generalize region variables.
 
-\fbname supports first-class functions via lambda expressions and
-arrow types. A lambda expression defines a region-polymorphic
-multi-argument function closure parameterized over function's
-allocation context parameter. The arrow type of a function closure is
-prenex-quantified over its constrained region arguments
-($\inang{\rho^a\rhobar \,|\, \phi}$). Since closures can escape the
-context in which they are created, it is important to keep track of
-the regions where closures are allocated in order to prevent unsafe
-dereferences. The $\rgn^a$ annotation above the arrow in the arrow
+Note that a lambda expression creates a closure, which can escape the
+context in which it is created. It is therefore important to keep track of
+the region in which a closure is allocated in order to avoid unsafe
+dereferences. The $\rgn$ annotation above the arrow in the arrow
 type denotes the allocation region of the corresponding closure. Note
 that it is important to distinguish between the allocation context
-argument of a function and the allocation region of its closure. For
-instance, in the following example:
-\begin{codejava}
-letregion $\rgn$ {
-  let f = $\lambda{\inang{\rho^a}}$()$\,\Rightarrow\,$new Object$\inang{\rho^a}$() 
-  in f
-}
-\end{codejava}
-The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
-\ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
-in $\rgn$, and (b). when executed under an allocation context
-$\rho^a$, the closure returns an object allocated in $\rho^a$.
-
-The expression language admits expressions to instantiate
-($e\inang{\ralloc\rbar}$) region parameters, and to apply
-($e(\bar{e})$) functions. For a method call, however, region parameter
-instantiation and method application are captured in a single
-expression ($e.mn\inang{\ralloc \rbar}(\ebar)$). The language also
-admits a big-lambda expression, which, when used in conjunction with
-the instantiation expression, serves the bureaucratic purpose of
-renaming bound region variables in the type of a function closure.
-
-The
+argument ($\rhoalloc$) of a function and the allocation region
+($\rgn$) of its closure. In \name, the later corresponds to the region where
+a \C{Func} object is allocated, while the former corresponds to the
+region where it is applied. 
+% For instance, in the following example:
+% \begin{codejava}
+% letregion $\rgn$ {
+%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
+%   in f
+% }
+% \end{codejava}
+% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
+% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
+% in $\rgn$, and (b). when executed under an allocation context
+% $\rho^a$, the closure returns an object allocated in $\rho^a$.

### 2015-10-21 paper/type-system.tex
Touch techrep.
Reduced rules to use in main paper.

@@
 \input{fb-syntax}
-\input{fb-auxdef}
-\input{fb-staticsem}
-\input{fb-morewfrules}
 
@@
 ($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
-expressions to define and apply anonymous functions. These expressions
-are uncurried variants of the corresponding expressions from System F.
+expressions to define and apply anonymous functions. Module the angle
+braces (whose presence will be justified later), these expressions are
+uncurried variants of the corresponding expressions from System F.
 Note that FGJ formalism does not include higher-order functions, hence
@@
 % $\rho^a$, the closure returns an object allocated in $\rho^a$.
+
+\input{fb-staticsem}
+\subsection{Types and Well-formedness}
+
+Static semantics of \fbname is defined by the rules to establish
+well-formedness of region types and region constraints, check
+subtyping between a pair of region types, and type-check expressions
+against region types. Fig.~\ref{fig:fb-staticsem} contains an
+illustrative subset of such rules\footnote{The full formal development
+can be found in our technical report~\cite{techrep}}, along with some
+auxiliary definitions. The judgments defined by these rules make use
+of a context ($\A$), which is a tuple of:
+\begin{itemize}
+\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
+that are estimated to be live,
+\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
+variables to their bounds\footnote{A bound of a type variable in
+FGJ~\cite{fgj} is the class that type variable was declared to extend.
+The map from type variables to their bounds is denoted as $\Delta$
+in FGJ formalism}.
+\item A constraint formula ($\phicx$) that captures the outlives
+constraints on regions in $\rhoenv$.
+\end{itemize}

### 2015-10-22 paper/type-system.tex
Text on static semantics.

@@
 \tau$) at the prenex position, similar to ML type schemes. However,
-unlike ML, we don't distinguish between region types and region type
-schemes; any of the $\tau$'s in the arrow type can themselves be
-region-parametric arrow types. In this respect, our region type system
-is more like System F's type system, which admits higher-rank
-parametric polymorphism. Like System F, \FB provides a region
-instantiation expression ($e\inang{\ralloc\rbar}$), and a region
-generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
-\phi}{e}$) to instantiate and generalize region variables.
+unlike ML, we don't distinguish between types and type schemes; any of
+the $\tau$'s in the arrow type can themselves be region-parametric
+arrow types. In this respect, our region type system is more like
+System F's type system, which admits higher-rank parametric
+polymorphism. Like System F, \FB provides a region instantiation
+expression ($e\inang{\ralloc\rbar}$), and a region generalization
+expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
+instantiate and generalize region variables.
 
@@
 \item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
-variables to their bounds\footnote{A bound of a type variable in
-FGJ~\cite{fgj} is the class that type variable was declared to extend.
-The map from type variables to their bounds is denoted as $\Delta$
-in FGJ formalism}.
+variables to their bounds\footnote{A bound of a type variable
+($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
+was declared to extend.  The map $\tyvar \mapsto \fgjN$ is denoted
+$\Delta$ in FGJ formalism}, and
 \item A constraint formula ($\phicx$) that captures the outlives
@@
 \end{itemize}
+The context for the expression typing judgment also includes:
+\begin{itemize}
+\item A type environment ($\env \in x \mapsto \tau$) that contains the
+type bindings for variables in scope, and 
+\item The static identifier ($\rgn^a$) of the allocation context for
+the expression is being type-checked.  
+\end{itemize}
+
+\noindent The well-formedness judgment on region types
+($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
+judgments on core types \footnote{We use a double-piped turnstile
+($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
+($\vdash$) for those in \FB.}.  An existentially quantified type of a
+region handler is well-formed if, effectively, $\RgnZ\inang{T}$ is a
+well-formed FGJ type. A region type
+($B\inang{\ralloc\rbar}\inang{\tbar}$) of an object of class $B$ is
+well-formed if and only if (a). its region-erased variant
+($B\inang{\tbar}$) is well-formed, and (b). its regions arguments
+($\ralloc\rbar$) are live, and satisfy the outlives constraints
+($\phi$) required by the class $B$. Well-formedness of region
+constraints is straightforward.
+
+The subtype relation between region types of objects
+($B\inang{\ralloc\rbar}\inang{\tbar}$) follows straightforwardly from
+the subclass relation between their class types. Conspicuous by its
+absence among the rules that define subtype relation is a rule that
+lifts outlives relation between regions to subtype relation between
+types of objects allocated in those regions: 
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.1}
+\RULE
+  {
+    \isvalid{\A.\phicx}{\rgn_1 \outlives \rgn_2}
+  }
+  {
+    \subtyp{\A}{T@{\rgn_1}}{T@{\rgn_2}}
+  }
+\end{array}
+\end{smathpar}
+Such a rule is unsound for it allows an object in a region ($\rgn_1$)
+to be treated as an object in a shorter living region ($\rgn_2
+\underlives \rgn_1$), thus allowing it to refer to shorter living
+objects in $\rgn_2$. The type system should prevent such unsafe
+references by restricting subtype relation to only those region types
+that are invariant in their allocation region. For instance, assuming
+that class $A$ is a subclass of $\ObjZ$ (as per FGJ), the subtype
+relation $A@\rgn <: \ObjZ@\rgn$ should be allowed, but not the
+relation $A@\rgn <: \ObjZ@\rgn'$, for any $\rgn' \neq \rgn$. \FB's
+type system enforces this not by restricting the subtype rule, but by
+constraining the subclass relation between region-annotated classes.
+In particular, it lets a class $A\inang{\rhoalloc}$ to be declared to
+extend $\ObjZ\inang{\rhoalloc}$, but never to extend
+$\ObjZ\inang{\rhoalloc_1}$, where $\rhoalloc_1 \neq \rhoalloc$. The
+well-formedness rule on class definitions (not shown) captures this
+restriction.
+
+As expected, subtyping between arrow types is contravariant in
+argument types and covariant in return types. It is also contravariant
+in the region constraints that the region parameters have to satisfy.
+For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
+\unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
+A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
+weaker than the later's. Note that, like other types, the subtype
+relation between function types is invariant with respect to the
+allocation region ($\rgn$) of their closures. This subtype rule
+assumes that both arrow types have same set of region parameters.
+However, this is not restrictive as (a). it is possible to add more
+region parameters to a function type and declare them to be equal to
+existing ones (as the arrow type above does), and (b). it is possible
+to rename region parameters by composing instantiation and
+generalization (big-lambda) expressions together.
+
+The type rule for \C{letregion} expression requires that the static
+identifier introduced by the expression be unique under the current
+context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
+order to prevent the new region from incorrectly assuming any existing
+outlives relationships on an eponymus region. Provided this is
+satisfied, the expression ($e$) under \C{letregion} is then
+type-checked against $\unitZ$ assuming that the new region is live
+($\rgn \in \rhoset$) and that it underlives all existing live regions
+($\rhoset \outlives \rgn$).
+
+The \C{new} $\RgnZ\inang{T}$ expression creates a new transferable
+region, whose root object is of (core) type $T$, and constructs a new
+$\RgnZ$ object to serve as a handler to the region. The corresponding
+type rule requires that the type $\RgnZ\inang{T}$ be well-formed as
+per FGJ semantics. Furthermore, the constructor is expected to be
+passed a thunk that returns an object of type $T$ contained in its
+allocation context. For the reasons described earlier, the allocation
+region of a new $\RgnZ$ object is existentially quantified.
+
+Like the rule for \C{letregion}, the type rule for the \C{open}
+expression requires that the static identifier ($\rgn$) introduced to
+refer to the transferable region being opened be unique. However,
+unlike the \C{letregion} rule, the rule for \C{open} does not
+introduce any outlives relationship between the newly opened region
+and any pre-existing region while checking the type of the expression
+($e$) under \C{open}. This effectively preempts the possibility of any
+references from objects inside the transferable region to those
+outside. Environment ($\env$) is extended with binding for the type of
+root object while type-checking $e$. The type of region handler ($x$)
+is also updated to reflect the elimination of existentially quantified
+region variable ($\rho$) in favor of concrete identifier ($\rgn$).
+
+The type rule for lamdba expression requires that its region
+parameters ($\rhoalloc\rhobar$) be unique under the current context,
+and that the constraints ($\phi$) over region parameters be
+well-formed. The lambda-bound expression ($e$) is checked under an
+extended type environment containing bindings for function's
+arguments, assuming that region parameters are live, and that declared
+constraints over region parameters hold. Note that the closure is
+always allocated in the current allocation context ($\rgn^a$). This
+prevents the closure from escaping the context in which it is created
+(cf. the following paragraph), thus trivially ensuring the safety of
+any dereferences inside the closure.
+
+The type rule for assignment expression ($e_1 := e_2$) ensures that a
+(local or instance) variable is only assigned objects of either the
+same type or a subtype. Recall that subtyping between region types is
+invariant with respect to the allocation regions of their objects. It
+follows that if a variable is declared to refer to objects in a region
+$\rgn$, then it is only ever assigned objects that allocated in
+$\rgn$. This prevents, for example, a temporary object allocated
+inside a stack region within a method from escaping the region's scope
+via an assignment to an instance variable.
+
+\subsection{Operational Semantics and Type Safety}
+
+Operational semantics of \fbname defines an evaluation relation
+($\redsto$) over its expression language. In particular, the
+evaluation gets stuck if an expression attempts to dereference a
+pointer into a region that has either been deallocated, freed or
+transferred. On the other hand, if expression attempts to commit an
+operation on a transferable region handler (a $\RgnZ$ object) that is
+not sanctioned by the state transition discipline in
+Fig.~\ref{fig:region-fsm}, then it is reduced to a value called
+\invalidexn. The accompanying technical report~\cite{techrep} contains
+full operational semantics of \fbname.

### 2015-10-22 paper/type-system.tex
Sec 3 complete.

@@
 full operational semantics of \fbname.
+
+To help state the type safety theorem, we define the syntactic class
+of values in \FB as following:
+\begin{smathpar}
+\begin{array}{lclcl}
+v & \in & \mathtt{values} & \coloneqq & \C{new}\; cn\inang{\tbar}
+      \inang{\ralloc\rbar} \ALT \invalidexn \\
+\end{array}
+\end{smathpar}
+Type safety theorem is stated thus:
+\begin{theorem}
+\emph{(\textbf{Type Safety})}
+\label{thm:core-safety}
+Let $\A_{\emptyset} = (\emptyset,\cdot,true)$ denote an empty context.
+If $\hastyp{\A_{\emptyset}, \toprgn, \cdot}{e}{\tau}$, then either $e$
+is a value, or there exists an $e'$ such that $e \redsto
+e'$ and $\hastyp{\A_{\emptyset}, \toprgn, \cdot}{e'}{\tau}$.
+\end{theorem}

### 2015-10-22 paper/type-system.tex
Major simplification of type rules.

--- 2015-10-21 paper/type-system.tex
@@
 \input{fb-syntax}
+\input{fb-auxdef}
+\input{fb-staticsem}
+\input{fb-morewfrules}
 
@@
 reflected in the $T@\rgn$ syntax of a region type, which is the
-simplest form of a region type describing an object, of core type $T$
+simplest form of a region type describing an object of core type $T$
 contained in a region $\rgn$. We let $\rgn$ range over static
@@
 ($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
-expressions to define and apply anonymous functions. Module the angle
-braces (whose presence will be justified later), these expressions are
-uncurried variants of the corresponding expressions from System F.
+expressions to define and apply anonymous functions. These expressions
+are uncurried variants of the corresponding expressions from System F.
 Note that FGJ formalism does not include higher-order functions, hence
@@
 
-\input{fb-staticsem}
-\subsection{Types and Well-formedness}
-
-Static semantics of \fbname is defined by the rules to establish
-well-formedness of region types and region constraints, check
-subtyping between a pair of region types, and type-check expressions
-against region types. Fig.~\ref{fig:fb-staticsem} contains an
-illustrative subset of such rules\footnote{The full formal development
-can be found in our technical report~\cite{techrep}}, along with some
-auxiliary definitions. The judgments defined by these rules make use
-of a context ($\A$), which is a tuple of:
-\begin{itemize}
-\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
-that are estimated to be live,
-\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
-variables to their bounds\footnote{A bound of a type variable in
-FGJ~\cite{fgj} is the class that type variable was declared to extend.
-The map from type variables to their bounds is denoted as $\Delta$
-in FGJ formalism}.
-\item A constraint formula ($\phicx$) that captures the outlives
-constraints on regions in $\rhoenv$.
-\end{itemize}
+\subsection{Static Semantics}

### 2015-10-24 paper/type-system.tex
Sec 3. complete.

@@
 such as the definition of type well-formedness for the core
-(region-free) language. Due to space constraints, we are unable to
-reproduce them here, and instead encourage the reader to refer
-to~\cite{fgj}.
+(region-free) language. We refrain from reproducing them here, and
+instead encourage the reader to refer to~\cite{fgj,techrep}.
 
 Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
-which we call \fbname (\FB). The language is seeded with \ObjZ and
-\RgnZ classes. More classes can be defined using the \C{class}
-keyword. Class types in \FB are region-annotated variants of class
-types in FGJ (also called \emph{core types}). This correspondence is
-reflected in the $T@\rgn$ syntax of a region type, which is the
-simplest form of a region type describing an object of core type $T$
-contained in a region $\rgn$. We let $\rgn$ range over static
-identifiers of regions in \FB. Note that all \FB objects are boxed
-values, hence their \FB types are always region-annotated. The only
-unboxed value in \FB is $\unitval$ of type \unitZ.
+which we call \fbname (\FB). Class types in \FB are region-annotated
+variants of class types in FGJ (also called \emph{core types}). This
+correspondence is reflected in the $T@\rgn$ syntax of a region type,
+which is the simplest form of a region type describing an object of
+core type $T$ contained in a region $\rgn$. We let $\rgn$ range over
+static identifiers of regions in \FB. The only unboxed value in \FB is
+$\unitval$ of type \unitZ. Rest are objects.
 
-The expression language of FGJ is impoverished, lacking, among others,
-an assignment expression. While it is possible to encode
-assignments to instance variables (object fields) in FGJ by creating a
-new object with its fields set to new values, such encoding is not
-semantics preserving in case of \name, as the new object could be
-allocated in a region different from the allocation region of the
-original object. To avoid such complications, we extend the expression
-language with assignments and local variable declarations (via \C{let}
-expressions). Instead of distinguishing between statements (e.g.,
-assignments) and expressions (e.g., method calls) in \FB, we consider
-statements as expressions of \unitZ type.  Statements that introduce
-lexical blocks in \name, such as \C{letregion} and \C{open} are, are
-converted to their expression form using the \C{in} keyword. We let
-$\rgn$ range over static region identifiers introduced by these
-expressions.
+\FB extends FGJ's expression language with assignments, local variable
+declarations (via \C{let} expressions), region constructs
+(\C{letregion} and \C{open}), lambda abstraction
+($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
+($e\inang{...}(\bar{e})$) expressions. Assignments and region
+constructs, which are statements in \name, are treated as effectful
+expressions in \FB. Lambda abstraction and application expressions are
+to define and apply anonymous functions. Modulo the angle braces
+(whose presence will be justified later), these expressions are
+uncurried variants of the corresponding expressions from System F.
+Since FGJ formalism does not include higher-order functions, we extend
+the syntactic class (\C{T}) of FGJ types with an (uncurried) arrow
+type.
 
-The expression language has been equipped with lambda abstraction
-($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
-expressions to define and apply anonymous functions. These expressions
-are uncurried variants of the corresponding expressions from System F.
-Note that FGJ formalism does not include higher-order functions, hence
-the syntactic class (\C{T}) of FGJ types needs to be extended with an
-(uncurried) arrow type.
-
-Informally, \FB class definitions are extensions of FGJ class
-definitions with regions. A class definition in FGJ can be lifted to
-its \FB definition by (a). lifting all core types in the body of the
-class (i.e, fields, constructors and methods) to their
-region-annotated versions, (b). parameterizing the class over
-region variables that occur free in the region-annotated body of the
-class, and (c). recording any outlives constraints on the region
-parameters of the class. The later two steps effectively make the
-class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
-polymorphism). Note that in \FB, regions, and their static identifiers
-($\rgn$) are confined to the expression language. Consequently, the
-region annotated types of class fields, constructors and methods do
-not refer to identifiers of concrete regions, but rather their place
-holders called \emph{region variables} ($\rho$). Informally, region
-variables ($\rho$) are to region identifiers ($\rgn$) as type
-variables ($a$) are to core types ($T$).
-
-Note that when a generic class definition in FGJ is lifted to a
-region-polymorphic definition in \FB, it nonetheless remains generic
-with respect to core types. We say that the class is now a
-region-polymorphic generic class. Region parameterization of a class
-in \FB is independent of its parameterization over core types; they
-are not conflated to make the class parametric over region types. For
-instance, following is the region-polymorphic definition of a generic
-\C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
+Class definitions in \FB extend Java's with region parameters
+($\rho$). Informally, region variables ($\rho$) are to region
+identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
+they can be instantiated with region identifiers, as needed. A class
+in \FB is necessarily parameterized over the allocation region of its
+objects. We use $\rho^a$ to denote this parameter, and often refer to
+it as \emph{class}'s allocation region, although it is really the
+allocation region of its objects. An Object in \FB can contain fields
+referring to objects allocated in regions ($\rhobar$) other than its
+own allocation region ($\rhoalloc$), provided that the former outlive
+the later (i.e., $\rhobar \outlives \rho$). In such cases, the
+definition of object's class needs to be parametric over allocation
+regions of its fields (i.e., their classes). Furthermore, the
+constraint that such regions must outlive the allocation region of the
+class needs to be made explicit in the definition. For instance, a
+\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
+component coming from ${\rho_1}$ and second from $\rho_2$, where all
+the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
+$\rhoalloc$. The corresponding definition of the \C{Pair} class is as
+follows(The symbol $\extends$ should be read \emph{extends}):  
 \begin{codejava}[mathescape=true]
-class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
-                      \conj \rho_2 \outlives \rho^a$>
-          <a $\extends$ Object, b $\extends$ Object> {
+class Pair<a $\extends$ Object, b $\extends$ Object>
+          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$>
+           {
   a@$\rho_1$ fst; 
@@
 \end{codejava}
-When objects of a class are allocated in a region $\rgn$, it means
-that the class's constructor is run with $\rgn$ as its allocation
-context. Every class definition in \FB is necessarily polymorphic with
-respect to the allocation region of its objects, i.e., the allocation
-context of its constructor. We adopt a convention that requires the
-allocation region parameter (denoted $\rho^a$) to be the first region
-parameter of a class.  Besides the allocation region of its objects,
-\C{Pair} class is also parametric over the regions its first and
-second elements are allocated in. References between objects allocated
-in different regions are only allowed if the referred object is
-guaranteed to outlive the referring object. In case of \C{Pair} class,
-this means that allocation regions ($\rho_1$ and $\rho_2$) of both
-objects that make up the pair must outlive the allocation region
-($\rho^a$) of the \C{Pair} object. Such conditions over region
-parameters of a class need to be recorded in its header as region
-constraints ($\phi$) in order for the class to be judged well-formed
-by the type system (Fig.~\ref{fig:fb-morewfrules}).
-
 To construct objects of the \C{Pair} class, its type and region
 parameters need to be instantiated with core types ($T$) and concrete
-region identifiers ($\rgn$), respectively. Region instantiation has to
-satisfy the stated constraints over region parameters. For example,
-the following code snippet appropriately instantiates region
-parameters of the \C{Pair} class to construct a \C{Pair} of
-{\ObjZ}{\!}s, each allocated in different regions:
+region identifiers ($\rgn$), respectively. For example:
 \begin{codejava}
 letregion $\rgn_0$ {
-  Object<$\rgn_0$> fst = new Object<$\rgn_0$>();
+  Object<$\rgn_0$> snd = new Object<$\rgn_0$>();
   letregion $\rgn_1$ {
-    Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
+    Object<$\rgn_1$> fst = new Object<$\rgn_1$>();
     let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
@@
 \end{codejava}
-Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
-region constraints of the \C{Pair} class are satisfied, hence the
-instantiation is valid. Observe that the region type of \C{p} conveys
+In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
+$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
+and $\rgn_1$ are live during the instantation, and (b). $\rgn_0
+\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
+reflexive). Observe that the region type of \C{p} conveys
 the fact that (a). it is allocated in region $\rgn_1$, and (b). it
 holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
-In contrast, if we choose to allocate the \C{fst} object also in
+In contrast, if we choose to allocate the \C{snd} object also in
 $\rgn_1$, the region type of \C{p} would be
-\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
-written simply as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, the
-\C{@} notation in a region type of an object \C{x} highlights that
-\C{x}, and all the objects reachable from \C{x} via references are
-allocated in a single region. We say that \C{x} is \emph{contained} in
-the region. The $\ObjZ$ class in \FB contains no references to other
-objects, hence its objects are always contained in their allocation
-region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
-$\ObjZ@\rgn$), for some region $\rgn$.
+\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which we abbreviate
+as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}\footnote{In general, we treat
+$B\inang{\tbar}@\rgn$ as an abbreviation of
+$B\inang{\tbar}\inang{\rbar}$}. The $\ObjZ$ class in \FB has no
+fields, hence its region type is of form $\ObjZ@\rgn$.
+% In general, the \C{@} notation in a
+% region type of an object \C{x} highlights that
+% \C{x}, and all the objects reachable from \C{x} via references are
+% allocated in a single region. We say that \C{x} is \emph{contained} in
+% the region. The $\ObjZ$ class in \FB contains no references to other
+% objects, hence its objects are always contained in their allocation
+% region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
+% $\ObjZ@\rgn$), for some region $\rgn$.
+
+% Note that when a generic class definition in FGJ is lifted to a
+% region-polymorphic definition in \FB, it nonetheless remains generic
+% with respect to core types. We say that the class is now a
+% region-polymorphic generic class. Region parameterization of a class
+% in \FB is independent of its parameterization over core types; they
+% are not conflated to make the class parametric over region types. For
+% instance, following is the region-polymorphic definition of a generic
+% \C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
+
+% When objects of a class are allocated in a region $\rgn$, it means
+% that the class's constructor is run with $\rgn$ as its allocation
+% context. Every class definition in \FB is necessarily polymorphic with
+% respect to the allocation region of its objects, i.e., the allocation
+% context of its constructor. We adopt a convention that requires the
+% allocation region parameter (denoted $\rho^a$) to be the first region
+% parameter of a class.  Besides the allocation region of its objects,
+% \C{Pair} class is also parametric over the regions its first and
+% second elements are allocated in. References between objects allocated
+% in different regions are only allowed if the referred object is
+% guaranteed to outlive the referring object. In case of \C{Pair} class,
+% this means that allocation regions ($\rho_1$ and $\rho_2$) of both
+% objects that make up the pair must outlive the allocation region
+% ($\rho^a$) of the \C{Pair} object. Such conditions over region
+% parameters of a class need to be recorded in its header as region
+% constraints ($\phi$) in order for the class to be judged well-formed
+% by the type system (Fig.~\ref{fig:fb-morewfrules}).
+
 %%An object \emph{allocated} in a region $\rgn$ contains its spine in
@@
 
-\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
-$\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
-$\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
-region. Recall that $\RgnZ$ objects have special semantics in \name -
-they act as handlers to transferable regions.  Constructing a new
-$\RgnZ$ object entails the creation of a new transferable region, and
-it is in this region that the new object is allocated in. It follows
-that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
-the new transferable region. But, static identifiers for transferable
-regions are introduced only when such regions are opened for
-allocation via \C{open} expression. What, then, should be the region
-type of a \C{new Region} expression?
+% \FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
+% $\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
+% $\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
+% region. Recall that $\RgnZ$ objects have special semantics in \name -
+% they act as handlers to transferable regions.  Constructing a new
+% $\RgnZ$ object entails the creation of a new transferable region, and
+% it is in this region that the new object is allocated in. It follows
+% that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
+% the new transferable region. But, static identifiers for transferable
+% regions are introduced only when such regions are opened for
+% allocation via \C{open} expression. What, then, should be the region
+% type of a \C{new Region} expression?
 
-\FB resolves this problem by existentially quantifying the allocation
-region of $\RgnZ$ objects when they are created. In other words, the
-type of \C{new Region} expression in \FB is
-$\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
-type captures the fact that there now exists a transferable region
-containing the newly constructed $\RgnZ$ object. Elimination of
-existential quantification is facilitated by the \C{open} expression,
-which opens the transferable region and assigns it an identifier
-($\rgn$).  Within the scope of \C{open}, the transferable region is
-identified with $\rgn$, allowing its handler to instantiate the
-existentially bound region variable with $\rgn$, and assume the type
-of $\RgnZ\inang{\rgn}$.
+% \FB resolves this problem by existentially quantifying the allocation
+% region of $\RgnZ$ objects when they are created. In other words, the
+% type of \C{new Region} expression in \FB is
+% $\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
+% type captures the fact that there now exists a transferable region
+% containing the newly constructed $\RgnZ$ object. Elimination of
+% existential quantification is facilitated by the \C{open} expression,
+% which opens the transferable region and assigns it an identifier
+% ($\rgn$).  Within the scope of \C{open}, the transferable region is
+% identified with $\rgn$, allowing its handler to instantiate the
+% existentially bound region variable with $\rgn$, and assume the type
+% of $\RgnZ\inang{\rgn}$.
 
-A method definition ($d$) can be region polymorphic with respect to
-(a). its allocation context (Sec.~\ref{sec:alloc-ctxt}), and (b). the
-regions occuring in the region types of its arguments.  Region
-parameters on the methods, like those on classes, are accompanied by
-constraints ($\phi$) capturing the conditions that the parameters need
-to satisfy for the method to be considered well-formed
-(Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
-or $\rho^a_m$) is the first and inevitable region parameter of every
-method in \FB. If a method is not intended to be polymorphic with
-respect to its allocation context (for example, if its allocation
-context needs to be same as the allocation region of its object), then
-the monomorphism needs to be encoded as an explicit equality
-constraint in $\phi$.
+Like classes, methods can also exhibit qualified region polymorphism.
+A method definition in \FB is necessarily polymorphic over its
+allocation context (Sec.~\ref{sec:alloc-ctxt}), and optionally
+polymorphic with respect to the regions containing its arguments.
+Region parameters, like those on classes, are qualified with
+constraints ($\phi$).
+% Region parameters on the methods, like those on classes, are
+% accompanied by constraints ($\phi$) capturing the conditions that the
+% parameters need to satisfy for the method to be considered well-formed
+% (Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
+% or $\rho^a_m$) is the first and inevitable region parameter of every
+% method in \FB. 
+Note that if a method is not intended to be polymorphic with respect
+to its allocation context (for example, if its allocation context
+needs to be same as the allocation region of its \emph{this}
+argument), then the required monomorphism can be captured as an
+equality constraint in $\phi$.
 
-Like methods, anonymous functions can also be region-polymorphic. 
+% Like methods, functions can also be region-polymorphic. 
 % A lambda expression defines a region-polymorphic
@@
 \tau$) at the prenex position, similar to ML type schemes. However,
-unlike ML, we don't distinguish between region types and region type
-schemes; any of the $\tau$'s in the arrow type can themselves be
-region-parametric arrow types. In this respect, our region type system
-is more like System F's type system, which admits higher-rank
-parametric polymorphism. Like System F, \FB provides a region
-instantiation expression ($e\inang{\ralloc\rbar}$), and a region
-generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
-\phi}{e}$) to instantiate and generalize region variables.
+unlike ML, we don't distinguish between types and type schemes; any of
+the $\tau$'s in the arrow type can themselves be region-parametric
+arrow types. Like with methods, region parameters on functions can be
+instantiated when they are applied.
+% In this respect, our region type system is more like
+% System F's type system, which admits higher-rank parametric
+% polymorphism. Like System F, \FB provides a region instantiation
+% expression ($e\inang{\ralloc\rbar}$), and a region generalization
+% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
+% instantiate and generalize region variables.
 
-Note that a lambda expression creates a closure, which can escape the
+A lambda expression creates a closure, which can escape the
 context in which it is created. It is therefore important to keep track of
@@
 a \C{Func} object is allocated, while the former corresponds to the
-region where it is applied. 
+region where it is applied ($e\inang{\ralloc\rbar}(\bar{e})$). 
 % For instance, in the following example:
@@
 
-\subsection{Static Semantics}
+\input{fb-staticsem}
+\subsection{Types and Well-formedness}
+
+Static semantics of \fbname is defined by the rules to establish
+well-formedness of region types and region constraints, check
+subtyping between a pair of region types, and type-check expressions
+against region types. Fig.~\ref{fig:fb-staticsem} contains an
+illustrative subset of such rules\footnote{The full formal development
+can be found in our technical report~\cite{techrep}}, along with some
+auxiliary definitions. The judgments defined by these rules make use
+of a context ($\A$), which is a tuple of:
+\begin{itemize}
+\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
+that are estimated to be live,
+\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
+variables to their bounds\footnote{A bound of a type variable
+($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
+was declared to extend.  The map $\tyvar \mapsto \fgjN$ is denoted
+$\Delta$ in FGJ formalism}, and
+\item A constraint formula ($\phicx$) that captures the outlives
+constraints on regions in $\rhoenv$.
+\end{itemize}
+The context for the expression typing judgment also includes:
+\begin{itemize}
+\item A type environment ($\env \in x \mapsto \tau$) that contains the
+type bindings for variables in scope, and 
+\item The static identifier ($\rgn^a$) of the allocation context for
+the expression is being type-checked.  
+\end{itemize}
+
+\begin{figure}
+
+\begin{codejava}
+class Region<a $\extends$ Object><${\rhoalloc}$> {
+  <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
+  Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
+    super(); 
+    this.thunk = thunk;
+  }
+  unit transfer$\inang{\rho}$(unit u) {
+    return u;
+  }
+}
+\end{codejava}
+
+\caption{The stub used for $\RgnZ$ class}
+\label{fig:region-stub}
+\end{figure}
+
+\noindent The well-formedness judgment on region types
+($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
+judgments on core types \footnote{We use a double-piped turnstile
+($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
+($\vdash$) for those in \FB.}. All judgments are implicitly
+parameterized on a class table ($CT \in cn \mapsto C$) that maps class
+names to their definitions in \FB. For judgments in FGJ, we use a
+class table that maps class names to their region-erased definitions.
+We denote the region-erased class table as ($\absof{CT}$). The precise
+semantics of region erasure ($\absof{\cdot}$) are defined
+in~\cite{techrep}, but it can be understood as an operation that
+erases region annotations on types. The class table is assumed to bind
+the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
+This mostly eliminates the need to treat $\RgnZ$ objects, which double
+up as transferable region handlers, as a special case in semantics.
+
+The well-formedness rule for region instantiation in class types
+($B\inang{\tbar}\inang{\ralloc\rbar}$) requires that the region-erased
+(core) type be well-formed, instantiated regions be live, and that
+they satisfy the constraints ($\phi$) imposed by the class on its
+region parameters. The later is enforced by checking the validity of
+$\phi$, with actual region arguments substituted\footnote{The notation
+$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
+region parameters, under the conditions ($\A.\phicx$) guaranteed by
+the context. The semantics of this sequent is straightforward, and
+follows directly from the properties of outlives and equality
+relations. For any well-formed core type $T$, $T@\rgn$ is a
+well-formed region type if $\rgn$ is a valid region. Well-formedness
+of region constraints requires that the regions referred by the
+constraints be live.
+
+The subtype relation between region types of objects
+($B\inang{\ralloc\rbar}\inang{\tbar}$) follows directly from the
+subclass relation between their class types. Note that, in general, we
+don't lift outlives relation between regions to subtype relation
+between objects (of same core type) allocated in those regions. Such
+lifting is unsound for the same reason as covariant/contravariant
+subtyping between mutable references is unsound. However, we make a
+special case for $\RgnZ$ objects: we allow $\RgnZ$ objects allocated
+in any region to be considered as objects allocated in an eternal
+region ($\toprgn$) that outlives every other region. This lets $\RgnZ$
+objects escape the context they are created in (as required by the
+\C{transfer} operation), and also facilitates data structures of
+regions created in different contexts (for e.g., a list of $\RgnZ$
+objects, whose root objects are of (core) type $T$, can be typed
+$\C{List}\inang{\RgnZ\inang{T}}\inang{\rgn,\toprgn}$, where $\rgn$ is
+the region where the spine of the list is allocated). As described in
+\S\ref{sec:memory-safety}, runtime checks are used to prevent the
+unsound subtyping rule on region objects from effecting the memory safety.
+% \begin{smathpar}
+% \begin{array}{c}
+% \renewcommand*{\arraystretch}{1.1}
+% \RULE
+%   {
+%     \isvalid{\A.\phicx}{\rgn_1 \outlives \rgn_2}
+%   }
+%   {
+%     \subtyp{\A}{T@{\rgn_1}}{T@{\rgn_2}}
+%   }
+% \end{array}
+% \end{smathpar}
+
+The actual implementations of region-based memory management for \name
+may choose to realize the $\toprgn$ as a garbage-collected region, but
+this is not required. Our operational semantics
+(\S\ref{sec:fb-opsem}), for example, does not even associate a region
+with $\toprgn$, instead opting to allocate $\RgnZ$ objects in the same
+transferable regions that they represent. Our type safety result
+nonetheless guarantees memory safety in this implementation.
+
+% Such a rule is unsound for it allows an object in a region ($\rgn_1$)
+% to be treated as an object in a shorter living region ($\rgn_2
+% \underlives \rgn_1$), thus allowing it to refer to shorter living
+% objects in $\rgn_2$. The type system should prevent such unsafe
+% references by restricting subtype relation to only those region types
+% that are invariant in their allocation region. For instance, assuming
+% that class $A$ is a subclass of $\ObjZ$ (as per FGJ), the subtype
+% relation $A@\rgn <: \ObjZ@\rgn$ should be allowed, but not the
+% relation $A@\rgn <: \ObjZ@\rgn'$, for any $\rgn' \neq \rgn$. \FB's
+% type system enforces this not by restricting the subtype rule, but by
+% constraining the subclass relation between region-annotated classes.
+% In particular, it lets a class $A\inang{\rhoalloc}$ to be declared to
+% extend $\ObjZ\inang{\rhoalloc}$, but never to extend
+% $\ObjZ\inang{\rhoalloc_1}$, where $\rhoalloc_1 \neq \rhoalloc$. The
+% well-formedness rule on class definitions (not shown) captures this
+% restriction.
+
+As expected, subtyping between arrow types is contravariant in
+argument types and covariant in return types. It is also contravariant
+in the region constraints that the region parameters have to satisfy.
+For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
+\unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
+A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
+weaker than the later's. Finally, the subtyping is invariant with
+respect to the allocation region ($\rgn$) of closures. 
+% This subtype rule assumes that both arrow types have same set of
+% region parameters.  However, this is not restrictive as (a). it is
+% possible to add more region parameters to a function type and declare
+% them to be equal to existing ones (as the arrow type above does), and
+% (b). it is possible to rename region parameters by composing
+% instantiation and generalization (big-lambda) expressions together.
+
+The type rule for \C{letregion} expression requires that the static
+identifier introduced by the expression be unique under the current
+context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
+order to prevent the new region from incorrectly assuming any existing
+outlives relationships on an eponymus region. Provided this is
+satisfied, the expression ($e$) under \C{letregion} is then
+type-checked against $\unitZ$ assuming that the new region is live
+($\rgn \in \rhoenv$) and that it is outlived by all existing live
+regions ($\rhoenv \outlives \rgn$).
+
+Like the rule for \C{letregion}, the type rule for the \C{open}
+expression requires that the static identifier ($\rgn$) introduced to
+refer to the transferable region being opened be unique. However,
+unlike the \C{letregion} rule, the rule for \C{open} does not
+introduce any outlives relationship between the newly opened region
+and any pre-existing region while checking the type of the expression
+($e$) under \C{open}. This effectively preempts the possibility of any
+references from objects inside the transferable region to those
+outside. Environment ($\env$) is extended with binding for the type of
+root object while type-checking $e$.
+
+The type rule for lamdba expression requires that its region
+parameters ($\rhoalloc\rhobar$) be unique under the current context,
+and that the constraints ($\phi$) over region parameters be
+well-formed. The lambda-bound expression ($e$) is checked under an
+extended type environment containing bindings for function's
+arguments, assuming that region parameters are live, and that declared
+constraints over region parameters hold. Note that the closure is
+always allocated in the current allocation context ($\rgn^a$). This
+prevents the closure from escaping the context in which it is created,
+thus trivially ensuring the safety of any dereferences inside the
+closure.
+
+The type rule for the \C{new} expression is a straightforward
+adaptation of corresponding FGJ rule to the the region type setting.
+
+% The type rule for assignment expression ($e_1 := e_2$) ensures that a
+% (local or instance) variable is only assigned objects of either the
+% same type or a subtype. Recall that subtyping between region types is
+% invariant with respect to the allocation regions of their objects. It
+% follows that if a variable is declared to refer to objects in a region
+% $\rgn$, then it is only ever assigned objects that allocated in
+% $\rgn$. This prevents, for example, a temporary object allocated
+% inside a stack region within a method from escaping the region's scope
+% via an assignment to an instance variable.
+
+\subsection{Operational Semantics and Type Safety}
+\label{sec:fb-opsem}
+
+Operational semantics of \fbname defines an evaluation relation
+($\redsto$) over its expression language. In particular, the
+evaluation gets stuck if an expression attempts to dereference a
+pointer into a region that has either been deallocated or transferred.
+On the other hand, if expression attempts to commit an operation on a
+transferable region handler (a $\RgnZ$ object) that is not sanctioned
+by the state transition discipline in Fig.~\ref{fig:region-fsm}, then
+it is reduced to a value called $\invalidexn$. The accompanying
+technical report~\cite{techrep} contains full operational semantics of
+\fbname.
+
+To help state the type safety theorem, we extend \FB's expressions
+with $\invalidexn$, and define the syntactic class of values:
+\begin{smathpar}
+\begin{array}{lclcl}
+v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \\
+\end{array}
+\end{smathpar}
+Type safety theorem is now stated thus:
+\begin{theorem}
+\emph{(\textbf{Type Safety})}
+\label{thm:core-safety}
+Let $\A_{\emptyset} = (\emptyset,\cdot,true)$ denote an empty context.
+If $\hastyp{\A_{\emptyset}, \toprgn, \cdot}{e}{\tau}$, then either $e$
+is a value, or $e \redsto \invalidexn$, or there exists an $e'$ such
+that $e \redsto e'$ and $\hastyp{\A_{\emptyset}, \toprgn,
+\cdot}{e'}{\tau}$.
+\end{theorem}

### 2015-10-27 paper/broom.tex
Text added to Sec.3
The type system ensures complete memory safety.
Edits.
Reduced rules to use in main paper.
Text on static semantics.
Sec 3 complete.
Major simplification of type rules.
Simplifications applied from techrep to master.
Sec 3. complete.
Type inference first commit.

@@
 \newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
+\newcommand{\eqtyp}[3]{#1\,\vdash\,#2 \equiv #3}
 \newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
@@
 \newcommand{\outlives}{\succeq}
+\newcommand{\underlives}{\preceq}
 \newcommand{\coloneqq}{::=}
@@
 \newcommand{\fbN}{N}
-\newcommand{\headerOf}[1]{\C{class}\; #1\angR\angAlpha \extends \fbN}
+\newcommand{\headerOf}[1]{\C{class}\; #1\angAlpha\angR \extends \fbN}
 \newcommand{\substFn}{\mathcal{S}}
 \newcommand{\mang}{\inang{\rhoalloc_m\bar{\rho_m} \,|\, \phi_m}}
-\newcommand{\rhoset}{\Sigma}
+\newcommand{\rhoset}{\rhoenv}
 \newcommand{\rhoenv}{\Delta}
@@
 \newcommand{\phicx}{\Phi}
-\newcommand{\subtypcx}{\rhoset,\rhoenv,\aenv,\phicx}
+\newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
 \newcommand{\A}{{\mathcal{A}}}
@@
 \newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
-\newcommand{\okin}[2]{#1 \; \texttt{ok \; in} \; #2}
-\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2).#3}
+\newcommand{\okin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
+\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2). #3}
 \newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
@@
 \newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
-\newcommand{\open}[3]{\C{open}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
+\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\;#2\;\C{withroot}\;#3\;\C{in}\;#4}
 \newcommand{\openalloc}[3]{\C{openalloc}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
 \newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
+\newcommand{\invalidexn}{\C{InvalidOpnException}}
+\newcommand{\toprgn}{\rgn_{\top}}
+\newcommand{\csolvestar}{{\sc CSolve*}\xspace}
+\newcommand{\elabExpr}{{\sf elabExpr}}
+\newcommand{\fields}{{\sf fields}}
+\newcommand{\mtype}{{\sf mtype}}
+\newcommand{\bound}{{\sf bound}}
+\newcommand{\fresh}{{\sf fresh}}
+\newcommand{\typeOk}{{\sf typeOk}}
+\newcommand{\subtypeOk}{{\sf subtypeOk}}
+\newcommand{\templateTy}{{\sf templateTy}}
+\newcommand{\allocRgn}{{\sf allocRgn}}
+%\newcommand{\smallskip}{\vspace*{0.1in}}
 
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={letregion, openalloc, open, withroot, not, : , /\\}
+morekeywords={letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
 }}
@@
 			\lstjava
-      \lstset{}%
+      \lstset{mathescape=true}%
       \csname lst@setfirstlabel\endcsname}
@@
       \lstset{numbers=left,firstnumber=1}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
+
+\newcommand{\lstml}{\lstset{ %
+language=[Objective]Caml, % choose the language of the code
+basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+keywordstyle=\color{Bittersweet},
+% numbers=left,                   % where to put the line-numbers
+numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
+stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
+numbersep=5pt,                  % how far the line-numbers are from the code
+showspaces=false,               % show spaces adding particular underscores
+showstringspaces=false,         % underline spaces within strings
+showtabs=false,                 % show tabs within strings adding particular underscores
+% frame=single,                   % adds a frame around the code
+tabsize=2,                      % sets default tabsize to 2 spaces
+captionpos=b,                   % sets the caption-position to bottom
+breaklines=true,                % sets automatic line breaking
+breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+commentstyle=\itshape\color{MidnightBlue},
+escapeinside={\%*}{*)},         % if you want to add a comment within your code
+morekeywords={relation, not, : , /\\}
+}}
+\lstnewenvironment{codeml}
+    { % \centering
+			\lstml
+      \lstset{mathescape=true}%
       \csname lst@setfirstlabel\endcsname}
@@
 
+\input{type-inference}
+
 % \appendix

### 2015-10-27 paper/type-inference.tex
Type inference first commit.

\section{Type Inference}
\label{sec:type-inference}

We formalize region type inference as an elaboration function from
programs in $\absof{\FB}$ (i.e., region-erased \FB) to programs in
\FB.  The type inference algorithm alternates between a constraint
generation phase and a constraint solving phase. During the generation
phase, inference algorithm introduces region variables ($\rho$) for
unknown region annotations on types, and constraint variables (denoted
$\varphi$) for unknown region constraints ($\phi$) on methods and
functions. Subsequently, the algorithm makes use of static semantics
in Fig.~\ref{fig:fb-staticsem} to generate constraints on region and
constraint variables. Due to the presence of region-polymorphic
recursion in \FB, constraints generated by the algorithm can be
circular. More precisely, constraints generated can assume the form
$\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
non-idempotent substitution function for region variables in
$\varphi$. In the constraint solving phase, the algorithm then relies
on a fixpoint constraint solving algorithm called \csolvestar to solve
the constraints and determine assignments for unknown region and
constraint variables.

\input{fb-constraintgen}

### 2015-10-30 paper/type-inference.tex
elabMeth added.
Edits.

@@
 programs in $\absof{\FB}$ (i.e., region-erased \FB) to programs in
-\FB.  The type inference algorithm alternates between a constraint
-generation phase and a constraint solving phase. During the generation
-phase, inference algorithm introduces region variables ($\rho$) for
-unknown region annotations on types, and constraint variables (denoted
-$\varphi$) for unknown region constraints ($\phi$) on methods and
-functions. Subsequently, the algorithm makes use of static semantics
-in Fig.~\ref{fig:fb-staticsem} to generate constraints on region and
-constraint variables. Due to the presence of region-polymorphic
-recursion in \FB, constraints generated by the algorithm can be
-circular. More precisely, constraints generated can assume the form
-$\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
-non-idempotent substitution function for region variables in
-$\varphi$. In the constraint solving phase, the algorithm then relies
-on a fixpoint constraint solving algorithm called \csolvestar to solve
-the constraints and determine assignments for unknown region and
-constraint variables.
+\FB.  The elaboration proceeds in three steps. In the first step we
+make use of the observation that region types are \emph{refinements}
+of FGJ types with region annotations and, possibly, constraints over
+such region annotations, and generate region type templates for
+expressions, methods and classes from their FGJ types. The templates
+contain region variables ($\rho$) to denote unknown region
+annotations, and predicate variables ($\varphi$) to denote unknown
+constraints over such region annotations. We also introduce region
+variables to denote unknown region arguments in \C{new} expressions,
+method calls and function applications. Next, we make use of the
+static semantics in Fig.~\ref{fig:fb-staticsem} to build a system of
+constraints over region and predicate variables that capture
+well-formedness requirements and subtyping relationships between type
+templates that must hold for the region elaboration to be valid.
+Finally, we solve the constraints by making use of our fixpoint
+constraint solving algorithm \csolvestar, which reduces the
+constraint solving problem to an abduction problem in a Herbrand
+constraint system, and then relies on \csolve, our abduction
+solver for that domain. 
+% Due to the presence of region-polymorphic
+% recursion in \FB, constraints generated by the algorithm can be
+% circular. More precisely, constraints generated can assume the form
+% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
+% non-idempotent substitution function for region variables in
+% $\varphi$. In the constraint solving phase, the algorithm then relies
+% on a fixpoint constraint solving algorithm called \csolvestar to solve
+% the constraints and determine assignments for unknown region and
+% predicate variables.
 
-\input{fb-constraintgen}
+\subsection{Region Type Templates}
+
+Region type templates are FGJ types extended with fresh region
+variables ($\rho$) and predicate variables ($\varphi$) to denote
+unknown region annotations and region constraints, respectively. For
+instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
+template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
+region variable. Likewise, given the region-annotated definition of
+\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for
+the method \C{mkSnd} shown below is \footnote{In our
+exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
+subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
+accept one region parameter - the allocation region of their
+objects.}\footnote{We abuse arrow notation to also represent types of
+methods, but unlike function types, there is no allocation region
+annotation atop the arrow in a method type.}
+$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
+\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
+\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
+variables, and $\varphi_0$ is a fresh predicate variable denoting
+unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  
+\begin{codejava}
+  unit mkSnd(Pair<A,B> p) {
+    return (p.snd = new B());
+  }
+\end{codejava}
+Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
+$\rho$ is fresh. For a class, the template is derived by generalizing
+free region variables in the templates of its fields as region
+parameters, and introducing a new predicate variable to denote
+unknown constraints on region parameters. A fresh region variable
+denoting allocation region argument is also added as a parameter. For
+example, given the standard FGJ definition of 
+$\C{Pair}\inang{\C{a},\C{b}}$ class, if $\C{a}$
+and $\C{b}$, the types of $\C{fst}$ and $\C{snd}$ fields (resp.), are
+templatized as $\C{a}@\rho_1$ and $\C{b}@\rho_2$ (resp.), then
+the template for the class in \FB is:
+\begin{center}
+$\C{Pair}\inang{\C{a}\extends\ObjZ,\C{b}\extends\ObjZ}
+\inang{\rhoalloc_0,\rho_0,\rho_1\,|\,\varphi_0} \extends \ObjZ \{...\}$
+\end{center}
+Fresh region variable $\rhoalloc_0$ is added to denote the allocation
+region argument. The body of the class, which is elided with the
+ellipsis above, contains region type templates of all its fields,
+methods and constructors.
+
+The type template for a recursively defined class is computed in two
+steps. First, all recursive occurances of the class among the types of
+its fields are ignored and the class is templatized as if it is a
+non-recursive class. Next, all the recursive occurances are
+templatized with respect to the class template computed in the first
+step, such that their region annotations are same as the region
+parameters of the class. For example, consider a generic
+$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
+type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.  where
+$\C{a}$ is the type of the data stored in the linked list node.  To
+templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
+recursive occurance in the type of $\C{next}$ field, and templatize
+the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
+type template of $\C{data}$ field, we compute the class's template
+as following:
+\begin{center}
+$\C{ListNode}\inang{\C{a}\extends\ObjZ}
+\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
+\end{center}
+Next, we trivially templatize the type of $\C{next}$ field in the body
+of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
+The resulting class represents a linked list with spine in the region
+$\rhoalloc_0$ and data objects in the region $\rho_0$.
+
+% Observe that our strategy of trivially templatizing the recursive
+% occurances of a class type in its definition ensures that no new
+% region variables are introduced.
+
+\input{fb-elabexpr}
+\input{fb-elabmeth}
+
+\subsection{Constraint Generation}
+
+Next, we describe our constraint generation algorithm that analyzes a
+templatized class definition and generates constraints over region and
+predicate variables occuring in the definition, such that the
+definition is well-formed if and only if there exists a satisfying
+assignment to region and predicate variables.
+
+% Recall that the overarching purpose of region type inference is to
+% elaborate programs in $\absof{\FB}$ to region-annotated programs in
+% $\FB$. A templatized class definition is only a partially elaborated 
+At the heart of our constraint generation algorithm is a function
+$\elabExpr$ that elaborates expressions in $\absof{\FB}$ to
+region-annotated expressions in $\FB$. Fig.~\ref{fig:fb-elabexpr}
+shows the definition of $\elabExpr$ for a subset of expressions in
+$\absof{\FB}$, whose corresponding $\FB$ expressions have been
+ascribed static semantics in Fig.~\ref{fig:fb-staticsem}. The function
+traverses expressions in a syntax-directed manner of a type checker,
+introducing fresh region type templates for unknown region types,
+while generating constraints that capture the relationship between the
+types of various subexpressions and their well-formedness. $\elabExpr$
+is defined under the same context as the expression typing judgment in
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
+retaining their meaning. Functions $\typeOk$ and $\subtypeOk$
+(definitions not shown) referred by $\elabExpr$ generate constraints
+making use of the syntax-directed well-formedness and subtype
+judgments from Fig.~\ref{fig:fb-staticsem}.
+
+The $\elabMeth$ function shown in Fig.~\ref{fig:fb-elabmeth} lifts the
+definition of elaboration from expressions to methods. We let methods
+to be defined recursively, but for the sake of presentation, we assume
+that there is no mutual recursion. This means that there exists a
+total order of all the classes in a program, and a total order of all
+the method definitions within each class, where definitions occuring
+prior in the order do not refer to those that occur later. We assume
+that classes and methods are elaborated in that order. Consequently,
+
+\subsection{Constraint Solving}
+
+Constraints generated by the method elaboration algorithm
+($\elabMeth$) are of two kinds: 
+\begin{itemize}
+\item Well-formedness constraints over region variables ($\rho$) of
+form $\rho \in \rhoenv$, where $\rhoenv$ is a set of region
+identifiers, and 
+\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
+$\phicx$ and $\phi$ are conjunctions of (a). outlives and equality
+constraints over region variables and region identifiers, and (b).
+a predicate variable ($\varphi$) denoting the unknown pre-condition
+of the method being elaborated.
+\end{itemize}
+\noindent The task of a constraint solving algorithm is to assign
+regions to region

### 2015-11-02 paper/broom.tex
elabMeth added.
Edits.

@@
 \usepackage[breaklinks]{hyperref}
+\usepackage[multiple]{footmisc}
 
@@
 \newcommand{\naiad}{{\sc Naiad}\xspace}
-\newcommand{\csolve}{{\sc Csolve}\xspace}
+\newcommand{\csolve}{{\sc CSolve}\xspace}
 \newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
@@
 %\newcommand{\smallskip}{\vspace*{0.1in}}
+\newcommand{\elabMeth}{{\sf elabMeth}}
+\newcommand{\frv}{{\sf frv}}
+\newcommand{\solve}{{\sf solve}}
+\newcommand{\elabClass}{{\sf elabClass}}
+\newcommand{\elabCons}{{\sf elabCons}}
+\newcommand{\elabClassTable}{{\sf elabClassTable}}
 
@@
 escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={relation, not, : , /\\}
+morekeywords={foreach, do, done, not, : , /\\}
 }}

### 2015-11-02 paper/type-inference.tex
Edits.

@@
 programs in $\absof{\FB}$ (i.e., region-erased \FB) to programs in
-\FB.  The elaboration proceeds in three steps. In the first step we
+\FB.  The elaboration proceeds in four steps. In the first step we
 make use of the observation that region types are \emph{refinements}
-of FGJ types with region annotations and, possibly, constraints over
-such region annotations, and generate region type templates for
-expressions, methods and classes from their FGJ types. The templates
-contain region variables ($\rho$) to denote unknown region
-annotations, and predicate variables ($\varphi$) to denote unknown
-constraints over such region annotations. We also introduce region
-variables to denote unknown region arguments in \C{new} expressions,
-method calls and function applications. Next, we make use of the
-static semantics in Fig.~\ref{fig:fb-staticsem} to build a system of
-constraints over region and predicate variables that capture
-well-formedness requirements and subtyping relationships between type
-templates that must hold for the region elaboration to be valid.
-Finally, we solve the constraints by making use of our fixpoint
-constraint solving algorithm \csolvestar, which reduces the
-constraint solving problem to an abduction problem in a Herbrand
-constraint system, and then relies on \csolve, our abduction
-solver for that domain. 
+of FGJ types with region annotations and constraints over such region
+annotations, and compute polymorphic region type templates for methods
+and classes from their FGJ types. The templates contain region
+variables ($\rho$) to denote unknown region annotations, and predicate
+variables ($\varphi$) to denote unknown constraints over such region
+annotations. Free region variables are generalized in types (hence,
+polymorphic). Second, we make use of the computed region type
+templates to elaborate expressions by introducing region variables to
+denote unknown region arguments in \C{new} expressions, method calls
+and function applications. While elaborating expressions, we also
+build a system of constraints that capture well-formedness
+requirements and subtyping relationships between type templates that
+must hold (as per the static semantics in Fig.~\ref{fb-staticsem}) for
+the elaboration to be valid. Third, we lift expression elaboration and
+constraint generation to methods and classes. Finally, we solve the
+constraints by making use of our fixpoint constraint solving algorithm
+\csolvestar, which reduces the constraint solving problem to an
+abduction problem in a Herbrand constraint system, and then relies on
+\csolve, our abduction solver for that domain. 
 % Due to the presence of region-polymorphic
@@
 \subsection{Region Type Templates}
+\label{sec:fb-templatization}
 
@@
 templatized with respect to the class template computed in the first
-step, such that their region annotations are same as the region
-parameters of the class. For example, consider a generic
+step, such that their region annotations are exactly same as the
+region parameters of the class. For example, consider a generic
 $\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
-type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.  where
-$\C{a}$ is the type of the data stored in the linked list node.  To
-templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
+type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
+where $\C{a}$ is the type of the data stored in the linked list node.
+To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
 recursive occurance in the type of $\C{next}$ field, and templatize
 the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
-type template of $\C{data}$ field, we compute the class's template
-as following:
+type template of $\C{data}$ field, we compute the class's template as
+following:
 \begin{center}
@@
 % region variables are introduced.
+The templatization technique we described above for recursive class
+definitions can be extended to mutually recursive definitions in a
+straightforward manner, by simultaneously templatizing them. Using the
+techniques outlined above, we compute region type templates for all
+classes bound in the class table of the $\absof{\FB}$ program before
+we proceed to elaborate expressions.
 
@@
 
-\subsection{Constraint Generation}
+\subsection{Expression Elaboration}
 
-Next, we describe our constraint generation algorithm that analyzes a
-templatized class definition and generates constraints over region and
-predicate variables occuring in the definition, such that the
-definition is well-formed if and only if there exists a satisfying
-assignment to region and predicate variables.
+Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
+(a). replacing core types in variable declarations and \C{new}
+expressions with fresh region type templates, and (b). explicitly
+instantiating region parameters of methods with fresh region variables
+in method calls and function applications. This elaboration is
+performed with respect to the polymorphic type templates of classes
+and methods computed as per \S\ref{sec:fb-templatization}.
 
-% Recall that the overarching purpose of region type inference is to
-% elaborate programs in $\absof{\FB}$ to region-annotated programs in
-% $\FB$. A templatized class definition is only a partially elaborated 
-At the heart of our constraint generation algorithm is a function
-$\elabExpr$ that elaborates expressions in $\absof{\FB}$ to
-region-annotated expressions in $\FB$. Fig.~\ref{fig:fb-elabexpr}
-shows the definition of $\elabExpr$ for a subset of expressions in
-$\absof{\FB}$, whose corresponding $\FB$ expressions have been
-ascribed static semantics in Fig.~\ref{fig:fb-staticsem}. The function
-traverses expressions in a syntax-directed manner of a type checker,
-introducing fresh region type templates for unknown region types,
-while generating constraints that capture the relationship between the
-types of various subexpressions and their well-formedness. $\elabExpr$
-is defined under the same context as the expression typing judgment in
+Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+this elaboration for a subset of expressions in $\absof{\FB}$, whose
+corresponding $\FB$ expressions have been ascribed static semantics in
+Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
+context as the expression typing judgment in
 Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning. Functions $\typeOk$ and $\subtypeOk$
-(definitions not shown) referred by $\elabExpr$ generate constraints
-making use of the syntax-directed well-formedness and subtype
-judgments from Fig.~\ref{fig:fb-staticsem}.
+retaining their meaning. The function traverses expressions in a
+syntax-directed manner of a type checker, introducing fresh region
+type templates for unknown region types, while generating constraints
+over region and predicate variables. The precise nature of these
+constraints will be clarified shortly, but their essential nature is
+that they capture the relationship between the type templates of
+various subexpressions and their well-formedness. Note that
+$\elabExpr$ returns the region type template of the subexpression,
+which is used to generate constraints for the expression. Functions
+$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
+implement type well-formedness and subtype judgments from
+Fig.~\ref{fig:fb-staticsem}, respectively.
 
-The $\elabMeth$ function shown in Fig.~\ref{fig:fb-elabmeth} lifts the
-definition of elaboration from expressions to methods. We let methods
-to be defined recursively, but for the sake of presentation, we assume
-that there is no mutual recursion. This means that there exists a
-total order of all the classes in a program, and a total order of all
-the method definitions within each class, where definitions occuring
-prior in the order do not refer to those that occur later. We assume
-that classes and methods are elaborated in that order. Consequently,
+\subsection{Method and Class Elaboration}
+
+Functions $\elabMeth$ and $\elabClass$ shown in
+Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+class definitions, respectively. Both functions first build a context
+($\A$) containing a set ($ \rhoenv$) of region variables denoting
+regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ denoting the fact that such constraints are yet
+to be inferred.
+
+Function $\elabMeth$ elaborates a method definition of class $B$. It
+calls $\elabExpr$ with the context $\A$, its allocation context
+parameter ($\rhoallocm$), and a type environment ($\env$) that
+contains region type bindings for all the arguments of the method,
+including the implicit $\C{this}$ argument. The region type template
+returned by $\elabExpr$ for the method body is checked against its
+expected type (derived from the type template of the method)
+generating more constraints. The function then returns the elaborated
+method definition and the set of constraints.
+
+$\elabClass$ elaborates a class's constructor ($k$) and method
+definitions ($\bar{d}$) using $\elabCons$\footnote{The definition of
+$\elabCons$ is straightforward, hence not shown.} and $\elabMeth$
+functions, respectively. To the set of constraints returned by these
+functions, $\elabClass$ adds constraints generated by checking the
+well-formedness of the type templates of its superclass and fields,
+and also a new constraint capturing the safety condition that
+allocation regions of objects referred by the instance variables
+should outlive the allocation region of the instance itself.
+
+Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
+every definition in the class table $CT$, while accumulating
+constraints. The constraints are finally solved by $\solve$
+(\S\ref{sec:fb-constraintsolving}), which returns substitution
+functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
+predicate variables, respectively, introduced during templatization
+and elaboration stages. The substitutions are applied to the class
+table (and to the artifacts that make up the class table, recursively)
+to compute a class table that maps classes to their fully
+region-annotated definitions in $\FB$.
+
+Note that if the original program in $\absof{\FB}$ contains unsafe
+references, for example, a reference from a transferable region to a
+stack regions, then the constraints generated during the elaboration
+are not satisfiable. In such case, $\solve$ fails to solve
+constraints, causing the program to be rejected.
 
 \subsection{Constraint Solving}
+\label{sec:fb-constraintsolving}
 

### 2015-11-02 paper/type-system.tex
Edits.

@@
 
-Class definitions in \FB extend Java's with region parameters
+Class definitions in \FB extend FGJ's with region parameters
 ($\rho$). Informally, region variables ($\rho$) are to region
@@
 ($\vdash$) for those in \FB.}. All judgments are implicitly
-parameterized on a class table ($CT \in cn \mapsto C$) that maps class
+parameterized on a class table ($CT \in cn \mapsto D$) that maps class
 names to their definitions in \FB. For judgments in FGJ, we use a

### 2015-11-03 paper/type-inference.tex
Edits.

@@
 Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
-$\rho$ is fresh. For a class, the template is derived by generalizing
-free region variables in the templates of its fields as region
-parameters, and introducing a new predicate variable to denote
-unknown constraints on region parameters. A fresh region variable
-denoting allocation region argument is also added as a parameter. For
-example, given the standard FGJ definition of 
-$\C{Pair}\inang{\C{a},\C{b}}$ class, if $\C{a}$
-and $\C{b}$, the types of $\C{fst}$ and $\C{snd}$ fields (resp.), are
-templatized as $\C{a}@\rho_1$ and $\C{b}@\rho_2$ (resp.), then
-the template for the class in \FB is:
+$\rho$ is fresh. For a class, a template is computed in two steps. In
+the first step, we templatize the types of all its fields, constructor
+arguments, and arguments and return values of all its methods, along
+with the type of its superclass. In the second step, we generalize all
+the free region variables occuring in the templates computed in the
+first step as region parameters of the class. Finally, we add a fresh
+allocation region parameter ($\rhoalloc$) to the list of parameters,
+and introduce a new predicate variable ($\varphi$) to denote unknown
+constraints on region parameters. For example, consider the standard
+FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
+and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
+respectively. It can be templatized as following:
 \begin{center}
-$\C{Pair}\inang{\C{a}\extends\ObjZ,\C{b}\extends\ObjZ}
-\inang{\rhoalloc_0,\rho_0,\rho_1\,|\,\varphi_0} \extends \ObjZ \{...\}$
+\begin{codejava}
+class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
+          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
+  $a@\rho_0$ fst;
+  $b@\rho_1$ snd;
+  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
+  $...$
+}
+\end{codejava}
 \end{center}
-Fresh region variable $\rhoalloc_0$ is added to denote the allocation
-region argument. The body of the class, which is elided with the
-ellipsis above, contains region type templates of all its fields,
-methods and constructors.
+% \footnote{In reality, the generalized region variables only come
+% from the type templates of class's fields and superclass. Method
+% types do not contain any free region variables.}
+The template elides $\C{Pair}$'s methods, whose region type templates
+contain no free variables. Among the region parameters of the class
+template, $\rho_{0-4}$ are obtained by generalizing free region
+variables in the types of its class's fields, constructor arguments,
+and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
+fresh region variable denoting the allocation region argument. Fresh
+predicate variable $\varphi_0$ denotes unknown constraints over
+$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
+well-formed region-annotated class definition in $\FB$.
 
@@
 type templates for unknown region types, while generating constraints
-over region and predicate variables. The precise nature of these
-constraints will be clarified shortly, but their essential nature is
-that they capture the relationship between the type templates of
+over region and predicate variables. The precise nature of generated
+constraints is explained in \S\ref{sec:fb-constraintsem}, but in
+summary, they capture the relationships between the type templates of
 various subexpressions and their well-formedness. Note that
@@
 
-$\elabClass$ elaborates a class's constructor ($k$) and method
-definitions ($\bar{d}$) using $\elabCons$\footnote{The definition of
-$\elabCons$ is straightforward, hence not shown.} and $\elabMeth$
-functions, respectively. To the set of constraints returned by these
-functions, $\elabClass$ adds constraints generated by checking the
-well-formedness of the type templates of its superclass and fields,
-and also a new constraint capturing the safety condition that
-allocation regions of objects referred by the instance variables
-should outlive the allocation region of the instance itself.
+$\elabClass$ elaborates the definition of a class $B$. It relies on
+$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
+hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
+constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+the set of constraints returned by these functions, $\elabClass$ adds
+constraints generated by checking the well-formedness of the type
+templates of its superclass and fields, and also a new constraint
+capturing a couple of safety conditions: first, the allocation regions
+of objects referred by the instance variables should outlive the
+allocation region of the instance itself, and second, the allocation
+regions of a class type and its superclass type must be the same.
 
@@
 
+\subsection{The Constraint Satisfaction Problem}
+\label{sec:fb-constraintsem}
+
+Our constraint generation algorithm generates three kinds of
+constraints:
+\begin{itemize}
+\item Well-formedness constraints of form $\rho \in \rhoenv$,
+restricting the domain of unification for a region variable ($\rho$)
+to the set ($\rhoenv$) of regions in scope,
+\item Well-formedness constraints of form
+$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
+variable ($\varphi$) to the set of all possible constraint formulas
+over region variables ($\rhoenv$) in scope, and
+\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
+$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
+extended with predicate variables and \emph{pending
+substitutions}\footnote{we borrowed this terminology
+from~\cite{ltpldi08}}:
+\begin{smathpar}
+\begin{array}{lcl}
+\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
+    \ALT F(\varphi) \ALT \phi \conj \phi\\
+F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
+\end{array}
+\end{smathpar}
+\end{itemize}
+A pending substitution ($F$) is a substitution function over region
+variables/identifiers. They represent the substitutions that need to
+be carried out when a predicate variable ($\varphi$) is replaced by a
+concrete formula in a validity constraint. For instance, in the
+validity constraint $\isvalid{\rgn_1 \outlives
+\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
+is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
+$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
+$\varphi$ if and only the formula obtained by substituting $\rgn_1$
+and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
+deducible from $\rgn_1 \outlives \rgn_2$.
+
+% Validity constraints primarily result from checking if a region type
+% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
+% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
+% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
+% formal region parameters with actual region arguments satisfy its
+% precondition, where actuals are subsituted for formals. Since $B$'s
+% precondition could be unknown, the requirement results in a in a
+% validity constraint with predicate variables and pending
+% substitutions. Other sources of validity constraints are calls to
+% region-polymorphic methods, where similar situation arises. 
+In general, validity constraints generated by our algorithm are of the
+form:
+\begin{center}
+\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+           {\phicstr \conj \bigwedge_{j}F_j(\varphi_j)}\)
+\end{center}
+Where $\varphi_i$'s denote the unknown preconditions of the class or
+the method under elaboration. If the constraint is generated while
+checking the well-formedness of a type or elaborating an expression,
+then $\varphi_j$'s denote the unknown preconditions of classes and
+methods that were used in that type or expression.  Each use of a
+(region-polymorphic) class or a method may instantiate region
+parameters differently, resulting in a different pending substitution
+($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
+of predicate variables and pending substitutions.  They capture the
+concrete relationships between regions that are \emph{known} to hold,
+and \emph{required} to hold, respectively, under the context in which
+the constraint was generated.
+
+\paragraph{Constraints Example 1} The original source of all
+outlives constraints is the validity constraint $C_3$ in $\elabClass$
+function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
+class template from \S\ref{sec:fb-templatization}, the constraint is
+as following:
+\begin{center}
+  \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
+  \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
+\end{center}
+
+\paragraph{Constraints Example 2} Consider a contrived method
+$\C{alt}$ that accepts an argument $\C{p}$ of type
+$\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
+of type $A$. It then assigns one object to $\C{p.fst}$ and other to
+$\C{p.snd}$, and calls itself recursively without ever terminating.
+Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
+alternated between recursive calls. The region-annotated definition of
+$\C{alt}$ is shown below:
+\begin{codejava}
+unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
+                          A<$\rho_3$> x, A<$\rho_4$> y) {
+  return p.fst = x; p.snd = y; 
+         this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
+}
+\end{codejava}
+Note that, to avoid clutter, we have already resolved appropriate
+region arguments to the recursive call, instead of introducing new
+region variables and generating equality constraints on them. Rest of
+the validity constraints generated while elaborating $\C{alt}$ to the
+above region-annotated definition are shown below:
+\begin{center}
+\(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
+$\quad$
+\(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
+$\quad$
+\(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
+\(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
+\end{center}
+The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
+Second and third are generated by the the assignment expressions, and
+the last constraint is generated from the recursive call.
+
 \subsection{Constraint Solving}
 \label{sec:fb-constraintsolving}
-
-Constraints generated by the method elaboration algorithm
-($\elabMeth$) are of two kinds: 
-\begin{itemize}
-\item Well-formedness constraints over region variables ($\rho$) of
-form $\rho \in \rhoenv$, where $\rhoenv$ is a set of region
-identifiers, and 
-\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
-$\phicx$ and $\phi$ are conjunctions of (a). outlives and equality
-constraints over region variables and region identifiers, and (b).
-a predicate variable ($\varphi$) denoting the unknown pre-condition
-of the method being elaborated.
-\end{itemize}
-\noindent The task of a constraint solving algorithm is to assign
-regions to region

### 2015-11-03 paper/type-system.tex
Edits.

@@
 class Pair<a $\extends$ Object, b $\extends$ Object>
-          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$>
+          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$>
            {
@@
 \begin{codejava}
-class Region<a $\extends$ Object><${\rhoalloc}$> {
+class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
   <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;

### 2015-11-04 paper/type-inference.tex
Edits.

@@
 
-\subsection{The Constraint Satisfaction Problem}
+\subsection{Constraints}
 \label{sec:fb-constraintsem}
@@
 \label{sec:fb-constraintsolving}
+
+Our constraint generation algorithm traverses the entire program,
+performing elaboration and collecting constraints, which are
+subsequently solved enmasse. The motivation behind the whole-program
+approach to constraint generation is twofold: it simplifies
+elaboration functions and makes presentation easier, and second, it
+naturally generalizes to mutual recursion. Nonetheless, we do not
+intend our type inference to be a whole-program analysis for (a). it
+preempts opportunities for separate compilation and dynamic linking,
+and (b). it is expensive and an overkill in most practical cases. We
+therefore reclaim the compositionality of type inference by solving
+the constraints in a compositional fashion. In more practical terms
+this means that our constraint solving algorithm visits and solves
+every constraint (or, every set of mutually recursive constraints)
+only once. It composes computed solutions to solve other constraints
+that depend on the solved constraints. Importantly, the failure to
+solve a dependent constraint does not result in backtracking. We now
+describe our compositional constraint solving algorithm in detail.
+
+\paragraph{Notation and terminology} In a validity constraint, a
+predicate variable occuring on the left side of the turnstile is said
+to occur negatively. In contrast, a predicate variable occuring on the
+right side is said to occur positively.
+
+Given a set of constraints, we first build a dependency graph ($G_c$)
+with validity constraints as nodes, and dependencies between them
+captured as edges. In particular, there exists an edge from a
+constraint $c_2$ to a constraint $c_1$ in the graph if any of the
+following conditions hold:
+\begin{itemize}
+\item $c_2$ and $c_1$ have atleast one common predicate variable
+($\varphi$) that occurs positively in $c_2$ and negatively in $c_1$.
+\item The set of predicate variables that occur negatively in $c_2$ is
+a (non-strict) subset of the corresponding set for $c_1$.
+\end{itemize}
+The first condition intuitively corresponds to a case, where
+expression or type, whose region elaboration is constrained by $c_2$
+refers to a method or a class, whose unknown precondition is
+constrained by $c_1$. The common predicate variable ($\varphi$)
+represents the unknown precondition in this case. The dependency from
+$c_2$ to $c_1$ means that $c_2$ must be solved to compute $\varphi$
+before $c_1$ is solved, thus enforcing the rule that the precondition
+of a method must not depend on its calling context. Since mutually
+recursive definitions refer each other, the first condition also
+results in bidirectional dependencies between mutually recursive
+constraints (and self-loops over recursive constraints). The second
+condition captures two kinds of dependencies. First, the dependency
+from the constraints over a method precondition to the constraints
+over the precondition of the class containing the method. This
+captures our preference that the constraints over a class's region
+parameters should not depend on the idiosyncracies of its methods. Any
+additional constraints required by any of its methods must be captured
+in the precondition of the method itself (well-formedness rules allow
+this possibility). The second condition also adds bidirectional
+dependencies between two validity constraints over same set of
+predicate variables, thus requiring them to be solved simultaneously.
+
+In the second step, we convert the dependency graph over constraints
+into a dependency DAG ($G_C$) over sets of constraints, where each set
+represents a strongly connected component in the dependency graph.
+Constraints in each set are mutually dependent, and need to be solved
+simultaneously, whereas constraints in different sets can be solved as
+per any valid topological ordering of the graph. Accordingly, we
+obtain a topological ordering of nodes in the graph $G_{{C}}$, and
+solve the sets of constraints in that order. The solutions obtained
+after solving a constraint set are applied to the constraints in
+subsequent sets before attempting to solve them.
+
+\subsection{The Constraint Satisfaction Problem}
+
+We now formally define the constraint satisfaction problem over sets
+of constraints. For simplicity, we ignore mutual recursion and assume
+that a a set only contains (possibly recursive) constraints over a
+single predicate variable. The general form of such a set of
+constraints is as following:
+\begin{center}
+\(
+  \bigcup_{i=1}^n \{
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {\phicstr^{i} \conj \bigwedge_{k=1}^n F_k(\varphi)}
+  \}
+\)
+\end{center}
+The solution to the above set of constraints is a tuple $(\substFn,)$
+problem is to find a map $$
+
+% After solving all the dependencies and applying solutions, the most
+% general form of a set of constraints is the following:
+% \begin{center}
+% \(
+%   \bigcup_{i=1}^n \{
+%     \bigcup_{j=1}^m \{ 
+%       \isvalid{\phictxt^{ij} \conj \varphi_i}{\phicstr^{ij} \conj
+%       \bigwedge_{k=1}^n F_k(\varphi_k)}
+%       \}
+%     \}
+% \)
+% \end{center}
+% Our technical report~\cite{techrep} contains formal proofs
+% establishing the equivalence of the above constraint set and the one
+% below, with respect to the constraint satisfaction problem:
+% \begin{center}
+% \(
+%   \bigcup_{i=1}^n \{
+%     \isvalid{\bigwedge_{j=1}^m \phictxt^{ij} \conj \varphi_i}
+%             {\bigwedge_{j=1}^m \phicstr^{ij} \conj \bigwedge_{k=1}^n F_k(\varphi_k)}
+%   \}
+% \)
+% \end{center}
+% Our technical report~\cite{techrep} contains formal proofs
+
+\subsubsection{Solving Non-Recursive Constraints}
+
+Non-recursive constraints are of the form $\isvalid{\phictxt \wedge
+\varphi}{\phicstr}$, where the predicate variable $\varphi$ has an
+associated well-formedness constraint $\tywf{\rhoenv}{\varphi}$.  The
+formula $\phicstr$ may contain free region variables ($\rho$), each
+with its own well-formedness constraint ($\rho \in \rhoenv$)
+specifying its domain of unification. Solving the constraint entails
+generating well-formed assignments to $\varphi$ and all free region
+variables in $\phicstr$.
+
+\begin{definition}
+\emph{(\textbf{\csolve})}
+\label{def:csolve}
+Let $L(S,R_1,R_2)$ be the following constraint language:
+\begin{smathpar}
+\begin{array}{lcl}
+x,y & \in & S\\
+\phi & \coloneqq & true \ALT R_1(x,y) \ALT R_2(x,y) \ALT \phi \conj \phi \\
+\end{array}
+\end{smathpar}
+\csolve is a function that accepts:
+\begin{itemize}
+\item A set $C$ of constants, and a set $V$ of variables
+\item A constraint $\phi$
+\end{itemize}
+
+\end{definition}

### 2015-11-04 paper/type-inference.tex
Edits.

@@
 represents the unknown precondition in this case. The dependency from
-$c_2$ to $c_1$ means that $c_2$ must be solved to compute $\varphi$
-before $c_1$ is solved, thus enforcing the rule that the precondition
+$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
+before $c_2$ is solved, thus enforcing the rule that the precondition
 of a method must not depend on its calling context. Since mutually
@@
 
-In the second step, we convert the dependency graph over constraints
-into a dependency DAG ($G_C$) over sets of constraints, where each set
+Next, we convert the dependency graph over constraints into a
+dependency DAG ($G_C$) over sets of constraints, where each set
 represents a strongly connected component in the dependency graph.
@@
 \end{center}
-The solution to the above set of constraints is a tuple $(\substFn,)$
-problem is to find a map $$
+% The solution to the above set of constraints is a tuple $(\substFn,)$
+% problem is to find a map
 

### 2015-11-06 paper/broom.tex
Edits.
Operational Semantics v1.

@@
 \usepackage[breaklinks]{hyperref}
-\usepackage[multiple]{footmisc}
 
@@
 \newcommand{\xn}[0]{\cap}
-\newcommand{\redsto}[0]{\longrightarrow}
+\newcommand{\redsto}[3]{#1 \vdash #2 \longrightarrow #3}
 \newcommand{\rredsto}[0]{\rightsquigarrow_r }
@@
 \newcommand{\naiad}{{\sc Naiad}\xspace}
-\newcommand{\csolve}{{\sc CSolve}\xspace}
+\newcommand{\csolve}{{\sc Csolve}\xspace}
 \newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
@@
 \newcommand{\outlives}{\succeq}
-\newcommand{\underlives}{\preceq}
 \newcommand{\coloneqq}{::=}
@@
 \newcommand{\mang}{\inang{\rhoalloc_m\bar{\rho_m} \,|\, \phi_m}}
-\newcommand{\rhoset}{\rhoenv}
+\newcommand{\rhoset}{\Sigma}
+\newcommand{\rhomap}{\Sigma}
 \newcommand{\rhoenv}{\Delta}
@@
 \newcommand{\okin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
-\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2). #3}
+\newcommand{\lambdaexp}[4]{\lambda@#1\inang{#2}(#3). #4}
 \newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
@@
 \newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
-\newcommand{\invalidexn}{\C{InvalidOpnException}}
+\newcommand{\invalidexn}{\bot}
 \newcommand{\toprgn}{\rgn_{\top}}
@@
 \newcommand{\mtype}{{\sf mtype}}
+\newcommand{\mbody}{{\sf mbody}}
 \newcommand{\bound}{{\sf bound}}
@@
 \newcommand{\allocRgn}{{\sf allocRgn}}
+\newcommand{\shape}{{\sf shape}}
 %\newcommand{\smallskip}{\vspace*{0.1in}}
@@
 \newcommand{\elabClassTable}{{\sf elabClassTable}}
+\newcommand{\CLOSED}{\mathtt{C}}
+\newcommand{\OPEN}{\mathtt{O}}
+\newcommand{\XFERRED}{\mathtt{X}}
+\newcommand{\transfer}{{\sf transfer}}
 
@@
 
-\newcommand{\lstml}{\lstset{ %
-language=[Objective]Caml, % choose the language of the code
-basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
-keywordstyle=\color{Bittersweet},
-% numbers=left,                   % where to put the line-numbers
-numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
-stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-numbersep=5pt,                  % how far the line-numbers are from the code
-showspaces=false,               % show spaces adding particular underscores
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-% frame=single,                   % adds a frame around the code
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to bottom
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-commentstyle=\itshape\color{MidnightBlue},
-escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={foreach, do, done, not, : , /\\}
-}}
-\lstnewenvironment{codeml}
-    { % \centering
-			\lstml
-      \lstset{mathescape=true}%
-      \csname lst@setfirstlabel\endcsname}
-    { %\centering
-      \csname lst@savefirstlabel\endcsname}
-
 \begin{document}
@@
 
-\input{type-inference}
-
 % \appendix

### 2015-11-06 paper/type-system.tex
Operational Semantics v1.

--- 2015-11-02 paper/type-system.tex
@@
 such as the definition of type well-formedness for the core
-(region-free) language. We refrain from reproducing them here, and
-instead encourage the reader to refer to~\cite{fgj,techrep}.
+(region-free) language. Due to space constraints, we are unable to
+reproduce them here, and instead encourage the reader to refer
+to~\cite{fgj}.
 
 Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
-which we call \fbname (\FB). Class types in \FB are region-annotated
-variants of class types in FGJ (also called \emph{core types}). This
-correspondence is reflected in the $T@\rgn$ syntax of a region type,
-which is the simplest form of a region type describing an object of
-core type $T$ contained in a region $\rgn$. We let $\rgn$ range over
-static identifiers of regions in \FB. The only unboxed value in \FB is
-$\unitval$ of type \unitZ. Rest are objects.
+which we call \fbname (\FB). The language is seeded with \ObjZ and
+\RgnZ classes. More classes can be defined using the \C{class}
+keyword. Class types in \FB are region-annotated variants of class
+types in FGJ (also called \emph{core types}). This correspondence is
+reflected in the $T@\rgn$ syntax of a region type, which is the
+simplest form of a region type describing an object of core type $T$
+contained in a region $\rgn$. We let $\rgn$ range over static
+identifiers of regions in \FB. Note that all \FB objects are boxed
+values, hence their \FB types are always region-annotated. The only
+unboxed value in \FB is $\unitval$ of type \unitZ.
 
-\FB extends FGJ's expression language with assignments, local variable
-declarations (via \C{let} expressions), region constructs
-(\C{letregion} and \C{open}), lambda abstraction
-($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
-($e\inang{...}(\bar{e})$) expressions. Assignments and region
-constructs, which are statements in \name, are treated as effectful
-expressions in \FB. Lambda abstraction and application expressions are
-to define and apply anonymous functions. Modulo the angle braces
-(whose presence will be justified later), these expressions are
-uncurried variants of the corresponding expressions from System F.
-Since FGJ formalism does not include higher-order functions, we extend
-the syntactic class (\C{T}) of FGJ types with an (uncurried) arrow
-type.
+The expression language of FGJ is impoverished, lacking, among others,
+an assignment expression. While it is possible to encode
+assignments to instance variables (object fields) in FGJ by creating a
+new object with its fields set to new values, such encoding is not
+semantics preserving in case of \name, as the new object could be
+allocated in a region different from the allocation region of the
+original object. To avoid such complications, we extend the expression
+language with assignments and local variable declarations (via \C{let}
+expressions). Instead of distinguishing between statements (e.g.,
+assignments) and expressions (e.g., method calls) in \FB, we consider
+statements as expressions of \unitZ type.  Statements that introduce
+lexical blocks in \name, such as \C{letregion} and \C{open} are, are
+converted to their expression form using the \C{in} keyword. We let
+$\rgn$ range over static region identifiers introduced by these
+expressions.
 
-Class definitions in \FB extend FGJ's with region parameters
-($\rho$). Informally, region variables ($\rho$) are to region
-identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
-they can be instantiated with region identifiers, as needed. A class
-in \FB is necessarily parameterized over the allocation region of its
-objects. We use $\rho^a$ to denote this parameter, and often refer to
-it as \emph{class}'s allocation region, although it is really the
-allocation region of its objects. An Object in \FB can contain fields
-referring to objects allocated in regions ($\rhobar$) other than its
-own allocation region ($\rhoalloc$), provided that the former outlive
-the later (i.e., $\rhobar \outlives \rho$). In such cases, the
-definition of object's class needs to be parametric over allocation
-regions of its fields (i.e., their classes). Furthermore, the
-constraint that such regions must outlive the allocation region of the
-class needs to be made explicit in the definition. For instance, a
-\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
-component coming from ${\rho_1}$ and second from $\rho_2$, where all
-the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
-$\rhoalloc$. The corresponding definition of the \C{Pair} class is as
-follows(The symbol $\extends$ should be read \emph{extends}):  
+The expression language has been equipped with lambda abstraction
+($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
+expressions to define and apply anonymous functions. These expressions
+are uncurried variants of the corresponding expressions from System F.
+Note that FGJ formalism does not include higher-order functions, hence
+the syntactic class (\C{T}) of FGJ types needs to be extended with an
+(uncurried) arrow type.
+
+Informally, \FB class definitions are extensions of FGJ class
+definitions with regions. A class definition in FGJ can be lifted to
+its \FB definition by (a). lifting all core types in the body of the
+class (i.e, fields, constructors and methods) to their
+region-annotated versions, (b). parameterizing the class over
+region variables that occur free in the region-annotated body of the
+class, and (c). recording any outlives constraints on the region
+parameters of the class. The later two steps effectively make the
+class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
+polymorphism). Note that in \FB, regions, and their static identifiers
+($\rgn$) are confined to the expression language. Consequently, the
+region annotated types of class fields, constructors and methods do
+not refer to identifiers of concrete regions, but rather their place
+holders called \emph{region variables} ($\rho$). Informally, region
+variables ($\rho$) are to region identifiers ($\rgn$) as type
+variables ($a$) are to core types ($T$).
+
+Note that when a generic class definition in FGJ is lifted to a
+region-polymorphic definition in \FB, it nonetheless remains generic
+with respect to core types. We say that the class is now a
+region-polymorphic generic class. Region parameterization of a class
+in \FB is independent of its parameterization over core types; they
+are not conflated to make the class parametric over region types. For
+instance, following is the region-polymorphic definition of a generic
+\C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
 \begin{codejava}[mathescape=true]
-class Pair<a $\extends$ Object, b $\extends$ Object>
-          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$>
-           {
+class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
+                      \conj \rho_2 \outlives \rho^a$>
+          <a $\extends$ Object, b $\extends$ Object> {
   a@$\rho_1$ fst; 
@@
 \end{codejava}
+When objects of a class are allocated in a region $\rgn$, it means
+that the class's constructor is run with $\rgn$ as its allocation
+context. Every class definition in \FB is necessarily polymorphic with
+respect to the allocation region of its objects, i.e., the allocation
+context of its constructor. We adopt a convention that requires the
+allocation region parameter (denoted $\rho^a$) to be the first region
+parameter of a class.  Besides the allocation region of its objects,
+\C{Pair} class is also parametric over the regions its first and
+second elements are allocated in. References between objects allocated
+in different regions are only allowed if the referred object is
+guaranteed to outlive the referring object. In case of \C{Pair} class,
+this means that allocation regions ($\rho_1$ and $\rho_2$) of both
+objects that make up the pair must outlive the allocation region
+($\rho^a$) of the \C{Pair} object. Such conditions over region
+parameters of a class need to be recorded in its header as region
+constraints ($\phi$) in order for the class to be judged well-formed
+by the type system (Fig.~\ref{fig:fb-morewfrules}).
+
 To construct objects of the \C{Pair} class, its type and region
 parameters need to be instantiated with core types ($T$) and concrete
-region identifiers ($\rgn$), respectively. For example:
+region identifiers ($\rgn$), respectively. Region instantiation has to
+satisfy the stated constraints over region parameters. For example,
+the following code snippet appropriately instantiates region
+parameters of the \C{Pair} class to construct a \C{Pair} of
+{\ObjZ}{\!}s, each allocated in different regions:
 \begin{codejava}
 letregion $\rgn_0$ {
-  Object<$\rgn_0$> snd = new Object<$\rgn_0$>();
+  Object<$\rgn_0$> fst = new Object<$\rgn_0$>();
   letregion $\rgn_1$ {
-    Object<$\rgn_1$> fst = new Object<$\rgn_1$>();
+    Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
     let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
@@
 \end{codejava}
-In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
-$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
-and $\rgn_1$ are live during the instantation, and (b). $\rgn_0
-\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
-reflexive). Observe that the region type of \C{p} conveys
+Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
+region constraints of the \C{Pair} class are satisfied, hence the
+instantiation is valid. Observe that the region type of \C{p} conveys
 the fact that (a). it is allocated in region $\rgn_1$, and (b). it
 holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
-In contrast, if we choose to allocate the \C{snd} object also in
+In contrast, if we choose to allocate the \C{fst} object also in
 $\rgn_1$, the region type of \C{p} would be
-\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which we abbreviate
-as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}\footnote{In general, we treat
-$B\inang{\tbar}@\rgn$ as an abbreviation of
-$B\inang{\tbar}\inang{\rbar}$}. The $\ObjZ$ class in \FB has no
-fields, hence its region type is of form $\ObjZ@\rgn$.
-% In general, the \C{@} notation in a
-% region type of an object \C{x} highlights that
-% \C{x}, and all the objects reachable from \C{x} via references are
-% allocated in a single region. We say that \C{x} is \emph{contained} in
-% the region. The $\ObjZ$ class in \FB contains no references to other
-% objects, hence its objects are always contained in their allocation
-% region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
-% $\ObjZ@\rgn$), for some region $\rgn$.
-
-% Note that when a generic class definition in FGJ is lifted to a
-% region-polymorphic definition in \FB, it nonetheless remains generic
-% with respect to core types. We say that the class is now a
-% region-polymorphic generic class. Region parameterization of a class
-% in \FB is independent of its parameterization over core types; they
-% are not conflated to make the class parametric over region types. For
-% instance, following is the region-polymorphic definition of a generic
-% \C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
-
-% When objects of a class are allocated in a region $\rgn$, it means
-% that the class's constructor is run with $\rgn$ as its allocation
-% context. Every class definition in \FB is necessarily polymorphic with
-% respect to the allocation region of its objects, i.e., the allocation
-% context of its constructor. We adopt a convention that requires the
-% allocation region parameter (denoted $\rho^a$) to be the first region
-% parameter of a class.  Besides the allocation region of its objects,
-% \C{Pair} class is also parametric over the regions its first and
-% second elements are allocated in. References between objects allocated
-% in different regions are only allowed if the referred object is
-% guaranteed to outlive the referring object. In case of \C{Pair} class,
-% this means that allocation regions ($\rho_1$ and $\rho_2$) of both
-% objects that make up the pair must outlive the allocation region
-% ($\rho^a$) of the \C{Pair} object. Such conditions over region
-% parameters of a class need to be recorded in its header as region
-% constraints ($\phi$) in order for the class to be judged well-formed
-% by the type system (Fig.~\ref{fig:fb-morewfrules}).
-
+\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
+written simply as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, the
+\C{@} notation in a region type of an object \C{x} highlights that
+\C{x}, and all the objects reachable from \C{x} via references are
+allocated in a single region. We say that \C{x} is \emph{contained} in
+the region. The $\ObjZ$ class in \FB contains no references to other
+objects, hence its objects are always contained in their allocation
+region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
+$\ObjZ@\rgn$), for some region $\rgn$.
 %%An object \emph{allocated} in a region $\rgn$ contains its spine in
@@
 
-% \FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
-% $\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
-% $\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
-% region. Recall that $\RgnZ$ objects have special semantics in \name -
-% they act as handlers to transferable regions.  Constructing a new
-% $\RgnZ$ object entails the creation of a new transferable region, and
-% it is in this region that the new object is allocated in. It follows
-% that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
-% the new transferable region. But, static identifiers for transferable
-% regions are introduced only when such regions are opened for
-% allocation via \C{open} expression. What, then, should be the region
-% type of a \C{new Region} expression?
+\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
+$\RgnZ\inang{\rgn}$. However, unlike the $\rgn$ in
+$\ObjZ\inang{\rgn}$, $\rgn$ in $\RgnZ\inang{\rgn}$ cannot be any
+region. Recall that $\RgnZ$ objects have special semantics in \name -
+they act as handlers to transferable regions.  Constructing a new
+$\RgnZ$ object entails the creation of a new transferable region, and
+it is in this region that the new object is allocated in. It follows
+that $\rgn$ in $\RgnZ\inang{\rgn}$ should be the static identifier of
+the new transferable region. But, static identifiers for transferable
+regions are introduced only when such regions are opened for
+allocation via \C{open} expression. What, then, should be the region
+type of a \C{new Region} expression?
 
-% \FB resolves this problem by existentially quantifying the allocation
-% region of $\RgnZ$ objects when they are created. In other words, the
-% type of \C{new Region} expression in \FB is
-% $\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
-% type captures the fact that there now exists a transferable region
-% containing the newly constructed $\RgnZ$ object. Elimination of
-% existential quantification is facilitated by the \C{open} expression,
-% which opens the transferable region and assigns it an identifier
-% ($\rgn$).  Within the scope of \C{open}, the transferable region is
-% identified with $\rgn$, allowing its handler to instantiate the
-% existentially bound region variable with $\rgn$, and assume the type
-% of $\RgnZ\inang{\rgn}$.
+\FB resolves this problem by existentially quantifying the allocation
+region of $\RgnZ$ objects when they are created. In other words, the
+type of \C{new Region} expression in \FB is
+$\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
+type captures the fact that there now exists a transferable region
+containing the newly constructed $\RgnZ$ object. Elimination of
+existential quantification is facilitated by the \C{open} expression,
+which opens the transferable region and assigns it an identifier
+($\rgn$).  Within the scope of \C{open}, the transferable region is
+identified with $\rgn$, allowing its handler to instantiate the
+existentially bound region variable with $\rgn$, and assume the type
+of $\RgnZ\inang{\rgn}$.
 
-Like classes, methods can also exhibit qualified region polymorphism.
-A method definition in \FB is necessarily polymorphic over its
-allocation context (Sec.~\ref{sec:alloc-ctxt}), and optionally
-polymorphic with respect to the regions containing its arguments.
-Region parameters, like those on classes, are qualified with
-constraints ($\phi$).
-% Region parameters on the methods, like those on classes, are
-% accompanied by constraints ($\phi$) capturing the conditions that the
-% parameters need to satisfy for the method to be considered well-formed
-% (Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
-% or $\rho^a_m$) is the first and inevitable region parameter of every
-% method in \FB. 
-Note that if a method is not intended to be polymorphic with respect
-to its allocation context (for example, if its allocation context
-needs to be same as the allocation region of its \emph{this}
-argument), then the required monomorphism can be captured as an
-equality constraint in $\phi$.
+A method definition ($d$) can be region polymorphic with respect to
+(a). its allocation context (Sec.~\ref{sec:alloc-ctxt}), and (b). the
+regions occuring in the region types of its arguments.  Region
+parameters on the methods, like those on classes, are accompanied by
+constraints ($\phi$) capturing the conditions that the parameters need
+to satisfy for the method to be considered well-formed
+(Fig.~\ref{fig:fb-morewfrules}). Allocation context (usually $\rho^a$
+or $\rho^a_m$) is the first and inevitable region parameter of every
+method in \FB. If a method is not intended to be polymorphic with
+respect to its allocation context (for example, if its allocation
+context needs to be same as the allocation region of its object), then
+the monomorphism needs to be encoded as an explicit equality
+constraint in $\phi$.
 
-% Like methods, functions can also be region-polymorphic. 
+Like methods, anonymous functions can also be region-polymorphic. 
 % A lambda expression defines a region-polymorphic
@@
 \tau$) at the prenex position, similar to ML type schemes. However,
-unlike ML, we don't distinguish between types and type schemes; any of
-the $\tau$'s in the arrow type can themselves be region-parametric
-arrow types. Like with methods, region parameters on functions can be
-instantiated when they are applied.
-% In this respect, our region type system is more like
-% System F's type system, which admits higher-rank parametric
-% polymorphism. Like System F, \FB provides a region instantiation
-% expression ($e\inang{\ralloc\rbar}$), and a region generalization
-% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
-% instantiate and generalize region variables.
+unlike ML, we don't distinguish between region types and region type
+schemes; any of the $\tau$'s in the arrow type can themselves be
+region-parametric arrow types. In this respect, our region type system
+is more like System F's type system, which admits higher-rank
+parametric polymorphism. Like System F, \FB provides a region
+instantiation expression ($e\inang{\ralloc\rbar}$), and a region
+generalization expression ($\Lambdaexp{\rhoalloc\rhobar \,|\,
+\phi}{e}$) to instantiate and generalize region variables.
 
-A lambda expression creates a closure, which can escape the
+Note that a lambda expression creates a closure, which can escape the
 context in which it is created. It is therefore important to keep track of
@@
 a \C{Func} object is allocated, while the former corresponds to the
-region where it is applied ($e\inang{\ralloc\rbar}(\bar{e})$). 
+region where it is applied. 
 % For instance, in the following example:
@@
 
-\input{fb-staticsem}
-\subsection{Types and Well-formedness}
+\subsection{Static Semantics}
 
-Static semantics of \fbname is defined by the rules to establish
-well-formedness of region types and region constraints, check
-subtyping between a pair of region types, and type-check expressions
-against region types. Fig.~\ref{fig:fb-staticsem} contains an
-illustrative subset of such rules\footnote{The full formal development
-can be found in our technical report~\cite{techrep}}, along with some
-auxiliary definitions. The judgments defined by these rules make use
-of a context ($\A$), which is a tuple of:
-\begin{itemize}
-\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
-that are estimated to be live,
-\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
-variables to their bounds\footnote{A bound of a type variable
-($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
-was declared to extend.  The map $\tyvar \mapsto \fgjN$ is denoted
-$\Delta$ in FGJ formalism}, and
-\item A constraint formula ($\phicx$) that captures the outlives
-constraints on regions in $\rhoenv$.
-\end{itemize}
-The context for the expression typing judgment also includes:
-\begin{itemize}
-\item A type environment ($\env \in x \mapsto \tau$) that contains the
-type bindings for variables in scope, and 
-\item The static identifier ($\rgn^a$) of the allocation context for
-the expression is being type-checked.  
-\end{itemize}
-
-\begin{figure}
-
-\begin{codejava}
-class Region<a $\extends$ Object><${\rhoalloc}$> {
-  <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
-  Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
-    super(); 
-    this.thunk = thunk;
-  }
-  unit transfer$\inang{\rho}$(unit u) {
-    return u;
-  }
-}
-\end{codejava}
-
-\caption{The stub used for $\RgnZ$ class}
-\label{fig:region-stub}
-\end{figure}
-
-\noindent The well-formedness judgment on region types
-($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
-judgments on core types \footnote{We use a double-piped turnstile
-($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
-($\vdash$) for those in \FB.}. All judgments are implicitly
-parameterized on a class table ($CT \in cn \mapsto D$) that maps class
-names to their definitions in \FB. For judgments in FGJ, we use a
-class table that maps class names to their region-erased definitions.
-We denote the region-erased class table as ($\absof{CT}$). The precise
-semantics of region erasure ($\absof{\cdot}$) are defined
-in~\cite{techrep}, but it can be understood as an operation that
-erases region annotations on types. The class table is assumed to bind
-the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
-This mostly eliminates the need to treat $\RgnZ$ objects, which double
-up as transferable region handlers, as a special case in semantics.
-
-The well-formedness rule for region instantiation in class types
-($B\inang{\tbar}\inang{\ralloc\rbar}$) requires that the region-erased
-(core) type be well-formed, instantiated regions be live, and that
-they satisfy the constraints ($\phi$) imposed by the class on its
-region parameters. The later is enforced by checking the validity of
-$\phi$, with actual region arguments substituted\footnote{The notation
-$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
-region parameters, under the conditions ($\A.\phicx$) guaranteed by
-the context. The semantics of this sequent is straightforward, and
-follows directly from the properties of outlives and equality
-relations. For any well-formed core type $T$, $T@\rgn$ is a
-well-formed region type if $\rgn$ is a valid region. Well-formedness
-of region constraints requires that the regions referred by the
-constraints be live.
-
-The subtype relation between region types of objects
-($B\inang{\ralloc\rbar}\inang{\tbar}$) follows directly from the
-subclass relation between their class types. Note that, in general, we
-don't lift outlives relation between regions to subtype relation
-between objects (of same core type) allocated in those regions. Such
-lifting is unsound for the same reason as covariant/contravariant
-subtyping between mutable references is unsound. However, we make a
-special case for $\RgnZ$ objects: we allow $\RgnZ$ objects allocated
-in any region to be considered as objects allocated in an eternal
-region ($\toprgn$) that outlives every other region. This lets $\RgnZ$
-objects escape the context they are created in (as required by the
-\C{transfer} operation), and also facilitates data structures of
-regions created in different contexts (for e.g., a list of $\RgnZ$
-objects, whose root objects are of (core) type $T$, can be typed
-$\C{List}\inang{\RgnZ\inang{T}}\inang{\rgn,\toprgn}$, where $\rgn$ is
-the region where the spine of the list is allocated). As described in
-\S\ref{sec:memory-safety}, runtime checks are used to prevent the
-unsound subtyping rule on region objects from effecting the memory safety.
-% \begin{smathpar}
-% \begin{array}{c}
-% \renewcommand*{\arraystretch}{1.1}
-% \RULE
-%   {
-%     \isvalid{\A.\phicx}{\rgn_1 \outlives \rgn_2}
-%   }
-%   {
-%     \subtyp{\A}{T@{\rgn_1}}{T@{\rgn_2}}
-%   }
-% \end{array}
-% \end{smathpar}
-
-The actual implementations of region-based memory management for \name
-may choose to realize the $\toprgn$ as a garbage-collected region, but
-this is not required. Our operational semantics
-(\S\ref{sec:fb-opsem}), for example, does not even associate a region
-with $\toprgn$, instead opting to allocate $\RgnZ$ objects in the same
-transferable regions that they represent. Our type safety result
-nonetheless guarantees memory safety in this implementation.
-
-% Such a rule is unsound for it allows an object in a region ($\rgn_1$)
-% to be treated as an object in a shorter living region ($\rgn_2
-% \underlives \rgn_1$), thus allowing it to refer to shorter living
-% objects in $\rgn_2$. The type system should prevent such unsafe
-% references by restricting subtype relation to only those region types
-% that are invariant in their allocation region. For instance, assuming
-% that class $A$ is a subclass of $\ObjZ$ (as per FGJ), the subtype
-% relation $A@\rgn <: \ObjZ@\rgn$ should be allowed, but not the
-% relation $A@\rgn <: \ObjZ@\rgn'$, for any $\rgn' \neq \rgn$. \FB's
-% type system enforces this not by restricting the subtype rule, but by
-% constraining the subclass relation between region-annotated classes.
-% In particular, it lets a class $A\inang{\rhoalloc}$ to be declared to
-% extend $\ObjZ\inang{\rhoalloc}$, but never to extend
-% $\ObjZ\inang{\rhoalloc_1}$, where $\rhoalloc_1 \neq \rhoalloc$. The
-% well-formedness rule on class definitions (not shown) captures this
-% restriction.
-
-As expected, subtyping between arrow types is contravariant in
-argument types and covariant in return types. It is also contravariant
-in the region constraints that the region parameters have to satisfy.
-For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
-\unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
-A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
-weaker than the later's. Finally, the subtyping is invariant with
-respect to the allocation region ($\rgn$) of closures. 
-% This subtype rule assumes that both arrow types have same set of
-% region parameters.  However, this is not restrictive as (a). it is
-% possible to add more region parameters to a function type and declare
-% them to be equal to existing ones (as the arrow type above does), and
-% (b). it is possible to rename region parameters by composing
-% instantiation and generalization (big-lambda) expressions together.
-
-The type rule for \C{letregion} expression requires that the static
-identifier introduced by the expression be unique under the current
-context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
-order to prevent the new region from incorrectly assuming any existing
-outlives relationships on an eponymus region. Provided this is
-satisfied, the expression ($e$) under \C{letregion} is then
-type-checked against $\unitZ$ assuming that the new region is live
-($\rgn \in \rhoenv$) and that it is outlived by all existing live
-regions ($\rhoenv \outlives \rgn$).
-
-Like the rule for \C{letregion}, the type rule for the \C{open}
-expression requires that the static identifier ($\rgn$) introduced to
-refer to the transferable region being opened be unique. However,
-unlike the \C{letregion} rule, the rule for \C{open} does not
-introduce any outlives relationship between the newly opened region
-and any pre-existing region while checking the type of the expression
-($e$) under \C{open}. This effectively preempts the possibility of any
-references from objects inside the transferable region to those
-outside. Environment ($\env$) is extended with binding for the type of
-root object while type-checking $e$.
-
-The type rule for lamdba expression requires that its region
-parameters ($\rhoalloc\rhobar$) be unique under the current context,
-and that the constraints ($\phi$) over region parameters be
-well-formed. The lambda-bound expression ($e$) is checked under an
-extended type environment containing bindings for function's
-arguments, assuming that region parameters are live, and that declared
-constraints over region parameters hold. Note that the closure is
-always allocated in the current allocation context ($\rgn^a$). This
-prevents the closure from escaping the context in which it is created,
-thus trivially ensuring the safety of any dereferences inside the
-closure.
-
-The type rule for the \C{new} expression is a straightforward
-adaptation of corresponding FGJ rule to the the region type setting.
-
-% The type rule for assignment expression ($e_1 := e_2$) ensures that a
-% (local or instance) variable is only assigned objects of either the
-% same type or a subtype. Recall that subtyping between region types is
-% invariant with respect to the allocation regions of their objects. It
-% follows that if a variable is declared to refer to objects in a region
-% $\rgn$, then it is only ever assigned objects that allocated in
-% $\rgn$. This prevents, for example, a temporary object allocated
-% inside a stack region within a method from escaping the region's scope
-% via an assignment to an instance variable.
-
-\subsection{Operational Semantics and Type Safety}
-\label{sec:fb-opsem}
-
-Operational semantics of \fbname defines an evaluation relation
-($\redsto$) over its expression language. In particular, the
-evaluation gets stuck if an expression attempts to dereference a
-pointer into a region that has either been deallocated or transferred.
-On the other hand, if expression attempts to commit an operation on a
-transferable region handler (a $\RgnZ$ object) that is not sanctioned
-by the state transition discipline in Fig.~\ref{fig:region-fsm}, then
-it is reduced to a value called $\invalidexn$. The accompanying
-technical report~\cite{techrep} contains full operational semantics of
-\fbname.
-
-To help state the type safety theorem, we extend \FB's expressions
-with $\invalidexn$, and define the syntactic class of values:
-\begin{smathpar}
-\begin{array}{lclcl}
-v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \\
-\end{array}
-\end{smathpar}
-Type safety theorem is now stated thus:
-\begin{theorem}
-\emph{(\textbf{Type Safety})}
-\label{thm:core-safety}
-Let $\A_{\emptyset} = (\emptyset,\cdot,true)$ denote an empty context.
-If $\hastyp{\A_{\emptyset}, \toprgn, \cdot}{e}{\tau}$, then either $e$
-is a value, or $e \redsto \invalidexn$, or there exists an $e'$ such
-that $e \redsto e'$ and $\hastyp{\A_{\emptyset}, \toprgn,
-\cdot}{e'}{\tau}$.
-\end{theorem}
+\include{fb-opsem}

### 2015-11-08 paper/related.tex
Related work.

\section{Related Work}
\label{sec:related-work}

% Tofte and Talpin, A Theory of Stack Allocation in Polymorphically
%     Typed Languages, 
% Tofte and Talpin, Implementation of the Typed Call-by-Value
%     λ-calculus using a Stack of Regions, POPL'94
% Tofte and Talpin, Region-Based Memory Management, IC'97
% Tofte, Birkedal, Elsman, Hallenberg, A Retrospective on Region-based
%     Memory Management, HOSC'2004.
% Yates, A Type-And-Effect System for Encapsulating Memory In Java, 1999
% Salcianu et al, Ownership Types for Safe Region-Based Memory Management in 
%     Real-Time Java, PLDI'03
% Bacchino et al, A Type and Effect System for Deterministic Parallel Java,
%     OOPSLA'09.
% Calcagno et al, Stratified operational semantics for safety and correctness
%     of the region calculus, POPL'01
% Hicks et al, Experience With Safe Manual Memory-Management in
%     Cyclone, ISMM'04
% Grossman et al, Region-based Memory Management in Cyclone, PLDI'02
% Henglein, Makholm, & Niss, A direct approach to control-flow sensitive
%     region-based memory management, PPDP'01.
% Talpin & Jouvelot, Polymorphic Type, Region and Effect Inference, JFP'92.
% Tofte & Birkedal, A Region Inference Algorithm, TOPLAS'98.

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
system and lightweight runtime checks. The type system reduces the
problem of ensuring safety of all memory accesses in a program into a
problem of enforcing a state-transition discipline on memory accesses
over a small subset of objects in the program. The later can be done
efficiently at runtime. his lets us guarantee safety in presence of
dynamic memory allocations and deallocations without requiring either
restrictive static mechanisms (e.g., linear types and borrow checker)
or expensive runtime mechanisms (e.g., garbage collection and
reference counting). Lastly, our region type system comes equipped
with full type inference that completely eliminates the need to write
region annotations on types to convince the type checker that the
program is safe.

% In contrast, ~\cite{tofte97} proposes to
% extend Standard ML, a higher-order functional language, with
% lexically-scoped stack regions. Cyclone~\cite{cyclone02}

% and defines an elaboration from
% Standard ML to the extended version of Standard ML. The aim of the
% elaboration is to introduce stack regions and do away with GC in a
% transparent fashion without jeopardizing memory safety. We too define
% an elaboration, but our focus is on introducing region annotations
% necessary to prove the safety of an object-oriented program with
% (stack and dynamic) regions. Cyclone

### 2015-11-09 paper/related.tex
Edits.

@@
 over a small subset of objects in the program. The later can be done
-efficiently at runtime. his lets us guarantee safety in presence of
+efficiently at runtime. This lets us guarantee safety in presence of
 dynamic memory allocations and deallocations without requiring either
-restrictive static mechanisms (e.g., linear types and borrow checker)
+restrictive static mechanisms (e.g., linear types and unique pointers)
 or expensive runtime mechanisms (e.g., garbage collection and
@@
 
-% In contrast, ~\cite{tofte97} proposes to
-% extend Standard ML, a higher-order functional language, with
-% lexically-scoped stack regions. Cyclone~\cite{cyclone02}
+~\cite{tofte97} proposes to extend Standard ML, a higher-order
+functional language, with lexically-scoped stack regions, and defines
+an elaboration from Standard ML to the region-annotated version of
+Standard ML. The aim of the elaboration is to introduce stack regions
+and do away with GC in a transparent fashion without jeopardizing
+memory safety. We too define an elaboration, but our focus is on
+introducing region annotations necessary to prove the safety of an
+object-oriented program that already uses (stack and dynamic) regions.
+Similar to their region inference algorithm, our region type inference
+algorithm can deal with region-polymorphic recursion by computing
+fixed points for recursive constraints. While their inference
+algorithm only ever generates equality constraints, which can be
+solved via unification, our type inference algorithm also generates
+partial order outlives constraints, which are required to capture
+subtle relationships between lifetimes of transferable regions and
+stack regions. Consequently, our constraint solving algorithm is more
+sophisticated, and is capable of inferring unknown outlives
+constraints over region arguments of polymorphic recursive functions.
 
-% and defines an elaboration from
-% Standard ML to the extended version of Standard ML. The aim of the
-% elaboration is to introduce stack regions and do away with GC in a
-% transparent fashion without jeopardizing memory safety. We too define
-% an elaboration, but our focus is on introducing region annotations
-% necessary to prove the safety of an object-oriented program with
-% (stack and dynamic) regions. Cyclone
+Cyclone~\cite{cyclone02} equips C with programmer-managed stack
+regions, and a typing discipline that statically guarantees the safety
+of all pointer dereferences. However, types need to be written
+explicitly. While some region annotations can be inferred, type
+inference is purely intraprocedural, and cannot infer region types for
+functions. Later proposal~\cite{cyclone04} extends Cyclone with
+dynamic regions, but has to impose unique pointer restrictions to
+guarantee safety. Objects that need to be shared within data
+structures are allowed to have multiple references, and safety for
+such objects was ensured via runtime reference counting.
+
+An ownership type system for safe region-based memory management in
+real-time Java has been proposed by~\cite{MIT03}. Like us, they too
+assume a source language with programmer-managed memory regions, and
+focus on proving safety of programs written in that language. Their
+source language admits various kinds of regions in order to support
+shared-memory concurrency. However, all their regions have
+lexically-scoped lifetimes. In contrast, we admit regions with
+dynamically determined lifetimes in order to support message-passing
+concurrency. We borrow outlives relation from their formal
+development, and our type system bears some similarities to their's.
+However, our language also admits parametric polymorphism (generics)
+and higher-order functions, whose interaction is non-trivial in
+context of a region type system. On the other hand, our type system
+does not admit region kinds and ownership, and is considerably simpler
+than theirs. Moreover, our type system comes with a type
+safety result that formalizes its guarantees with respect to a
+well-defined operational semantics. Like Cyclone, ~\cite{MIT03}'s
+language is explicitly typed, with some support for local type
+inference. Consequently, region types for methods and classes need to
+be written explicitly. In contrast, we support full type inference
+that eliminates such need.
+
+~\cite{HMN01} proposes a flow-sensitive region-based memory management
+for first-order programs that proposes to overcome some of the
+drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
+to regions with dynamic lifetimes. However, dynamic regions are still
+not first-class values of the language. Nonetheless, due to aliasing
+among region names, ~\cite{HMN01} requires reference counting for
+regions to ensure memory safety.

### 2015-11-09 paper/related.tex
Edits.

@@
 does not admit region kinds and ownership, and is considerably simpler
-than theirs. Moreover, our type system comes with a type
+than theirs. Furthermore, our type system comes with a type
 safety result that formalizes its guarantees with respect to a
 well-defined operational semantics. Like Cyclone, ~\cite{MIT03}'s
-language is explicitly typed, with some support for local type
-inference. Consequently, region types for methods and classes need to
-be written explicitly. In contrast, we support full type inference
-that eliminates such need.
+language is explicitly typed. Although there is some support for local
+type inference, region types for methods and classes need to be
+written explicitly. In contrast, we support full type inference that
+eliminates any such need.
 
@@
 to regions with dynamic lifetimes. However, dynamic regions are still
-not first-class values of the language. Nonetheless, due to aliasing
-among region names, ~\cite{HMN01} requires reference counting for
-regions to ensure memory safety.
+not first-class values of the language, and reference counting is
+nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
+calculus with first-class regions with dynamic lifetimes, and imposes
+restrictive linear typing to control accesses to regions. All type
+annotations need to be written explicitly. Moreover, invariants needed
+to ensure safety under region transfers, such as the absence of
+references that escape a transferable region, cannot be enforced
+under~\cite{WW01}'s type system.

### 2015-11-09 paper/introduction.tex

@@
 \label{sec:introduction}
+
+\subsection*{Outline}
+
+\begin{itemize}
+\item
+Setting: Consider a producer that creates a data-structure and passes it to a consumer to process.
+Suppose that the data-structure is very large, and it takes some time to produce it.
+A GC that occurs during this phase will [potentially, depending on generations] scan the partially constructed data-structure, but this scan is useless since the entire data-structure is alive.
+In a system, where most of the work takes this form, with producers producing really large data-structures, the GC overheads become significant and better performance can be achieved using alternative memory management techniques.
+(We could present the Ci\# version of one of motivating Naiad examples here.
+E.g., it could be the Select example.
+But when I look at this example, it lacks one of the complicating aspects, namely that a dynamic region may be created in one call of the OnReceive method, updated in several subsequent invocations of the OnReceive method, and finally be transferred in some invocation.)
+
+\item
+The goal of this work is to enable transferable regions: essentially a data-structure such as the above.
+It is best to consider a transferable region as representing a single data-structure (that is, a composite object), and not a collection of unrelated individual objects.
+
+\item
+Each such region represents a data-structure D1 that can be used independent of any other data-structure D2, from a memory management perspective: in other words, we should be able to deallocate D2, without affecting our ability to use D1 and vice-versa.
+This more-or-less reduces to the restriction that objects in D1 cannot point to objects in D2 and vice-versa.
+\item
+In our system, a transferable region with this strong property [invariant] is referred to as a closed transferable region.
+
+\item
+We cannot always ensure that a transferable region D1 has this property.
+For example, while working with such a region (either during its production stage or consumption stage), we may need to create temporary objects (constituting working storage, so to speak) that point to internal objects of D1.
+We use the following idiom to handle such situations: we open a transferable region D1, to indicate that we are going to be working D1 and that D1 cannot be freed during this period; when we are done with this processing, we close the region D1.
+At this point we should regain the strong invariant that no pointer crosses the boundary of D1.
+[The goal of our type-system is to ensure this.]
+
+\item
+Re-establishing the desired invariant when a region D1 is closed requires that all temporary objects created while processing the region [that contain pointers into the region D1] be necessarily freed by the time D1 is closed.
+We ensure this by allocating such objects in a static-region S1 that is guaranteed to be deallocated before D1 is closed.
+We ensure this by statically ensuring that S1's scope is statically nested within the open-close scope for D1.
+
+\item
+The same technique also ensures that stack variables that may point to objects inside D1 are out-of-scope by the time D1 is closed.
+
+\item
+Note that in this section we introduce dynamic regions first, which are our primary goal (and which are the primary novel aspect of this paper).
+Static regions arise as a necessary and useful (but not novel) ingredient to support working with dynamic regions.
+
+\item
+This reduces the problem of ensuring memory safety to two sub-problems: the first one, to ensure the inter-region pointer invariants about which objects in which regions can point to which objects in which regions, and the second one, ensuring the correct usage protocol for pointers (handles) to regions themselves.
+In particular, the second one includes the constraints such as: do not free an open region, do not open a freed region, etc.
+
+\item
+We can elaborate on these two categories on invariants.
+
+\item
+We then explain that our type-system statically ensures that the program will satisfy the first set of constraints.
+
+\item
+We explain that we verify that the second set of constraints are satisfied dynamically.
+
+\item
+We justify our choice to do in this fashion: the first set of constraints are expensive to check dynamically, while the second set of constraints is not.
+In contrast, a static typechecker that rules out the second category of violations would be very restrict from an expressiveness perspective (requiring a linear type system).
+Finally, the human effort to ensure the second set of invariants is not as burdensome (as these are coarse-grained objects manipulated infrequently), unlike in the case for the first set.
+
+\item
+We can perhaps illustrate the need to create handles/pointers to regions and store them in dynamic data-structures like dictionaries to illustrate why restricting the usage of such pointers would be too constraining.
+
+\end{itemize}
+
+\subsection*{Contributions}
 

### 2015-11-09 paper/introduction.tex
Partial updates to introduction

@@
 
+Consider the example shown in Fig.~\ref{fig:motivating-eg}.
+\textbf{TODO: Replace by more complex example where data-structure is updated
+over multiple invocations of method and finally transferred.}
+This code represents the logic for a streaming query operator (also known
+as a continuous query). The operator receives a continuous stream of
+input messages, which are batched into time-windows associated with a logical
+timestamp. Messages may be delivered out-of-order and multiple messages with
+the same timestamp may be delivered. A timing-message, periodically indicates that
+no more input messages with a timestamp less than a given value $t$ will be subsequently
+delivered. At this point, the operator can create the output message corresponding
+to time $t$ and deliver it to its successors.
+
+\input{motivating-eg}
+
+This example is an instance of a general pattern, where a producer creates a data-structure
+and passes it to a consumer.
+In a system where most of the computation takes this form and these data-structures are
+very large, as is the case with many streaming big-data analysis systems,
+the garbage collection overheads become significant.
+% as the collector scans these large data-structures
+However, these data-structures have a well-defined lifetime, which can be exploited to
+achieve better performance 
+% using alternative memory management techniques.
+
+Region-based memory management is a good fit for such programs.
+A region is a block of memory (consisting of potentially more than one contiguous
+range of memory locations, permitting a region to grow) that is used to realize
+a data-structure with a well-defined lifetime. In particular, the entire region
+is typically freed in one shot.
+
+Region-based memory management, both manual as well as automatic, has been known for
+a long time. Manual region-based memory management suffers from the usual challenges,
+namely the potential for dangling references and lack of memory safety. Automatic
+region-based memory managements guarantee memory safety, but impose various restrictions.
+Tofte and Talpin's system, for example, requires that the set of all regions (existing
+at a point in time) forms a stack. Specifically, this requires that the lifetimes of
+all regions must be well-nested: it is not possible to have two regions whose lifetimes
+overlap, with neither one's lifetime contained within the other.
+Unfortunately, the data-structures in the above example do not satisfy this restriction.
+We refer to such regions as \emph{dynamic} regions.
+
+The goal of this work is an automatic (memory-safe) region-based memory management technique
+that supports dynamic regions and higher-order functions.
+
 \begin{itemize}
-\item
-Setting: Consider a producer that creates a data-structure and passes it to a consumer to process.
-Suppose that the data-structure is very large, and it takes some time to produce it.
-A GC that occurs during this phase will [potentially, depending on generations] scan the partially constructed data-structure, but this scan is useless since the entire data-structure is alive.
-In a system, where most of the work takes this form, with producers producing really large data-structures, the GC overheads become significant and better performance can be achieved using alternative memory management techniques.
-(We could present the Ci\# version of one of motivating Naiad examples here.
-E.g., it could be the Select example.
-But when I look at this example, it lacks one of the complicating aspects, namely that a dynamic region may be created in one call of the OnReceive method, updated in several subsequent invocations of the OnReceive method, and finally be transferred in some invocation.)
-
-\item
-The goal of this work is to enable transferable regions: essentially a data-structure such as the above.
-It is best to consider a transferable region as representing a single data-structure (that is, a composite object), and not a collection of unrelated individual objects.
 

### 2015-11-09 paper/introduction.tex
updatess to introduction

@@
+\newcommand{\TODO}[1]{\textbf{TODO: #1}}
+
 \section{Introduction}
@@
 Consider the example shown in Fig.~\ref{fig:motivating-eg}.
-\textbf{TODO: Replace by more complex example where data-structure is updated
+\TODO{Replace by more complex example where data-structure is updated
 over multiple invocations of method and finally transferred.}
@@
 that supports dynamic regions and higher-order functions.
+Though our focus is on memory-safety, dynamic regions also have other side-benefits.
+When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
+provides an ownership discipline that ensures that the producer cannot subsequently modify
+the data-structure. When the same transfer happens in a distributed setting, the system guarantees
+that the data-structure is self-contained and can be safely transmitted across address spaces.
+\TODO{How accurate is this claim?}
+
+The key memory safety property we wish to ensure is that there are no dangling references:
+i.e., a reference to an object that has been freed.
+\TODO{Or, should we say: no dangling reference is dereferenced}
+A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
+region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
+(In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
+This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
+Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
+when a garbage collector promotes an object from a lower generation to a higher generation.
+From this perspective, a region may be seen as a user-controlled generation.)
 

### 2015-11-09 paper/introduction.tex
updates to introduction

@@
 \newcommand{\TODO}[1]{\textbf{TODO: #1}}
+\newcommand{\eg}{\emph{e.g.}}
 
@@
 
+This ensures that we can safely free one dynamic region (\eg, the input-message in our example)
+without affecting the validity of pointers in another dynamic region (\eg, the output-message).
+However, this is not sufficient! When processing the input-message $D$, we necessarily will have to create
+pointers that point to objects that are internal to $D$. Such pointers may be stack-allocated or even reside
+in the heap (\eg, consider the iterator object used to iterate over the list in the input-message).
+How do we ensure the safety of these pointers?
+
+We combine \emph{static} regions with the following protocol to ensure safety in these cases.
+To work with a dynamic region $D$, we must first \emph{open} region D1,
+to indicate that $D$ should not be freed during this period.
+When the processing of the region is complete, we close the region $D$,
+to indicate that it is safe to free $D$.
+Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
+but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
+Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
+guaranteed to be contained within the interval when $D$ is open.
+As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
+no pointer from outside $D$ points to an internal object of $D$.
+
 \begin{itemize}
-
-\item
-Each such region represents a data-structure D1 that can be used independent of any other data-structure D2, from a memory management perspective: in other words, we should be able to deallocate D2, without affecting our ability to use D1 and vice-versa.
-This more-or-less reduces to the restriction that objects in D1 cannot point to objects in D2 and vice-versa.
-\item
-In our system, a transferable region with this strong property [invariant] is referred to as a closed transferable region.
-
-\item
-We cannot always ensure that a transferable region D1 has this property.
-For example, while working with such a region (either during its production stage or consumption stage), we may need to create temporary objects (constituting working storage, so to speak) that point to internal objects of D1.
-We use the following idiom to handle such situations: we open a transferable region D1, to indicate that we are going to be working D1 and that D1 cannot be freed during this period; when we are done with this processing, we close the region D1.
-At this point we should regain the strong invariant that no pointer crosses the boundary of D1.
-[The goal of our type-system is to ensure this.]
-
-\item
-Re-establishing the desired invariant when a region D1 is closed requires that all temporary objects created while processing the region [that contain pointers into the region D1] be necessarily freed by the time D1 is closed.
-We ensure this by allocating such objects in a static-region S1 that is guaranteed to be deallocated before D1 is closed.
-We ensure this by statically ensuring that S1's scope is statically nested within the open-close scope for D1.
-
-\item
-The same technique also ensures that stack variables that may point to objects inside D1 are out-of-scope by the time D1 is closed.
-
-\item
-Note that in this section we introduce dynamic regions first, which are our primary goal (and which are the primary novel aspect of this paper).
-Static regions arise as a necessary and useful (but not novel) ingredient to support working with dynamic regions.
 

### 2015-11-09 paper/introduction.tex
updates to introduction

@@
 \label{sec:introduction}
-
-\subsection*{Outline}
 
@@
 The goal of this work is an automatic (memory-safe) region-based memory management technique
-that supports dynamic regions and higher-order functions.
+that supports dynamic regions (that are first-class objects) and higher-order functions.
 Though our focus is on memory-safety, dynamic regions also have other side-benefits.
@@
 
-\begin{itemize}
+The final piece required to ensure memory-safety is ensuring that the program correctly follows the above-mentioned
+protocol for dynamic regions. The protocol is essentially a typestate specification: an open region cannot be freed and,
+dually, a freed region cannot be opened.
 
-\item
-This reduces the problem of ensuring memory safety to two sub-problems: the first one, to ensure the inter-region pointer invariants about which objects in which regions can point to which objects in which regions, and the second one, ensuring the correct usage protocol for pointers (handles) to regions themselves.
-In particular, the second one includes the constraints such as: do not free an open region, do not open a freed region, etc.
+Dynamic regions are first-class objects in our system: e.g., in our running example,
+we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
+aliases (pointers) to the same region, which, in turn, means that statically checking if the program correctly follows the
+dynamic region protocol is hard (undecidable, in fact). Safety can be ensured by using, \eg, linear typing to prevent the
+creation of aliases, but this would be quite restrictive. Hence, in our system, we dynamically check for adherence to
+this protocol.
 
-\item
-We can elaborate on these two categories on invariants.
-
-\item
-We then explain that our type-system statically ensures that the program will satisfy the first set of constraints.
-
-\item
-We explain that we verify that the second set of constraints are satisfied dynamically.
-
-\item
-We justify our choice to do in this fashion: the first set of constraints are expensive to check dynamically, while the second set of constraints is not.
-In contrast, a static typechecker that rules out the second category of violations would be very restrict from an expressiveness perspective (requiring a linear type system).
-Finally, the human effort to ensure the second set of invariants is not as burdensome (as these are coarse-grained objects manipulated infrequently), unlike in the case for the first set.
-
-\item
-We can perhaps illustrate the need to create handles/pointers to regions and store them in dynamic data-structures like dictionaries to illustrate why restricting the usage of such pointers would be too constraining.
-
-\end{itemize}
+In summary, our approach reduces memory safety to two sub-problems:
+(a) Ensuring  low-level invariants about pointers to (fine-grained) objects inside regions and 
+(b) Ensuring higher-level (typestate) invariants about the regions themselves. 
+We use a type system to statically ensure the first set of invariants. We use a dynamic check to ensure that the
+second set of invariants are satisfied.
+%
+This is a novel aspect of our system. We believe that this is a reasonable choice:
+dynamically checking for the second category of invariants is easy and does not pose a performance concern (unlike, \eg, the first category of constraints);
+a static typechecker that rules out the second category of violations would be very restrictive from an expressiveness perspective;
+finally, the human effort required to ensure the second set of invariants is not as burdensome (as regions are coarse-grained objects
+manipulated infrequently), \eg, as it would be for the first set of invariants.
 

### 2015-11-09 paper/introduction.tex
Motivating examples added.

@@
 Consider the example shown in Fig.~\ref{fig:motivating-eg}.
-\TODO{Replace by more complex example where data-structure is updated
-over multiple invocations of method and finally transferred.}
-This code represents the logic for a streaming query operator (also known
-as a continuous query). The operator receives a continuous stream of
-input messages, which are batched into time-windows associated with a logical
-timestamp. Messages may be delivered out-of-order and multiple messages with
-the same timestamp may be delivered. A timing-message, periodically indicates that
-no more input messages with a timestamp less than a given value $t$ will be subsequently
-delivered. At this point, the operator can create the output message corresponding
-to time $t$ and deliver it to its successors.
+This code represents the logic for a streaming query operator (also
+known as a continuous query). The operator receives a continuous
+stream of input messages, which are processed via a user-defined
+function, and batched into time-windows associated with a logical
+timestamp. Messages may be delivered out-of-order and multiple
+messages with the same timestamp may be delivered. A timing-message,
+periodically indicates that no more input messages with a timestamp
+less than or equalt to a given value $t$ will be subsequently
+delivered. At this point, the operator can create the output message
+corresponding to time $t$ and deliver it to its successors.
 
@@
 
-This example is an instance of a general pattern, where a producer creates a data-structure
-and passes it to a consumer.
-In a system where most of the computation takes this form and these data-structures are
-very large, as is the case with many streaming big-data analysis systems,
-the garbage collection overheads become significant.
+This example is an instance of a general pattern, where a producer
+creates a data-structure and passes it to a consumer.  In a system
+where most of the computation takes this form, and these
+data-structures are very large, as is the case with many streaming
+big-data analysis systems, the garbage collection overheads become
+significant.
 % as the collector scans these large data-structures
-However, these data-structures have a well-defined lifetime, which can be exploited to
-achieve better performance 
+However, these data-structures have a well-defined lifetime, which can
+be exploited to achieve better performance 
 % using alternative memory management techniques.

### 2015-11-09 paper/related.tex
Motivating examples added.

@@
 memory safety that relies on a combination of a simple static type
-system and lightweight runtime checks. The type system reduces the
-problem of ensuring safety of all memory accesses in a program into a
-problem of enforcing a state-transition discipline on memory accesses
-over a small subset of objects in the program. The later can be done
-efficiently at runtime. This lets us guarantee safety in presence of
-dynamic memory allocations and deallocations without requiring either
-restrictive static mechanisms (e.g., linear types and unique pointers)
-or expensive runtime mechanisms (e.g., garbage collection and
-reference counting). Lastly, our region type system comes equipped
-with full type inference that completely eliminates the need to write
-region annotations on types to convince the type checker that the
-program is safe.
+discipline and lightweight runtime checks. 
+% The type system reduces the problem of ensuring safety of all memory
+% accesses in a program into a problem of enforcing a state-transition
+% discipline on memory accesses over a small subset of objects in the
+% program. The later can be done efficiently at runtime. This lets us
+% guarantee safety in presence of
+In particular, our approach requires neither restrictive static
+mechanisms (e.g., linear types and unique pointers) nor expensive
+runtime mechanisms (e.g., garbage collection and reference counting)
+in order to guarantee safety. Lastly, our region type system comes
+equipped with full type inference that completely eliminates the need
+to write region annotations on types to convince the type checker that
+the program is safe.
 

### 2015-11-09 paper/introduction.tex
Edits.

@@
 periodically indicates that no more input messages with a timestamp
-less than or equalt to a given value $t$ will be subsequently
+less than or equal to a given value $t$ will be subsequently
 delivered. At this point, the operator can create the output message

### 2015-11-09 paper/related.tex
Edits.

@@
 settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
-WW01, rust}. Our work differs from the existing proposals in a number
-of ways. Firstly, our problem setting includes lexically scoped stack
-regions and dynamic transferable regions (both programmer-managed) in
-context of an object-oriented programming language equipped with
-higher-order functions. Second, we adopt a two-pronged approach to
-memory safety that relies on a combination of a simple static type
-discipline and lightweight runtime checks. 
+WW01, rust, gpu14}. Our work differs from the existing proposals in a
+number of ways. Firstly, our problem setting includes lexically scoped
+stack regions and dynamic transferable regions (both
+programmer-managed) in context of an object-oriented programming
+language equipped with higher-order functions. Second, we adopt a
+two-pronged approach to memory safety that relies on a combination of
+a simple static type discipline and lightweight runtime checks. 
 % The type system reduces the problem of ensuring safety of all memory
@@
 context of a region type system. On the other hand, our type system
-does not admit region kinds and ownership, and is considerably simpler
-than theirs. Furthermore, our type system comes with a type
-safety result that formalizes its guarantees with respect to a
-well-defined operational semantics. Like Cyclone, ~\cite{MIT03}'s
-language is explicitly typed. Although there is some support for local
-type inference, region types for methods and classes need to be
-written explicitly. In contrast, we support full type inference that
+eschews the notions of region kinds and ownership, leading to a more
+succinct formalization. Furthermore, we establish a type safety result
+that formalizes its guarantees with respect to a well-defined
+operational semantics. Like Cyclone, ~\cite{MIT03}'s language is
+explicitly typed. Although there is some support for local type
+inference, region types for methods and classes need to be written
+explicitly. In contrast, we support full type inference that
 eliminates any such need.
@@
 calculus with first-class regions with dynamic lifetimes, and imposes
-restrictive linear typing to control accesses to regions. All type
-annotations need to be written explicitly. Moreover, invariants needed
-to ensure safety under region transfers, such as the absence of
-references that escape a transferable region, cannot be enforced
-under~\cite{WW01}'s type system.
+linear typing to control accesses to regions. All type annotations
+need to be written explicitly. Moreover, invariants needed to ensure
+safety under region transfers, such as the absence of references that
+escape a transferable region, cannot be enforced under~\cite{WW01}'s
+type system.
+
+The idea of using region-based memory management to facilitate the
+safe transfer of rich data structures between computational nodes has
+been previously explored by~\cite{gpu14} in the context of Scheme
+language. However, their setting only includes lexically-scoped
+regions for which Tofte and Talpin~\cite{tofte97} style analysis
+suffices. In contrast, our language provides first-class support for
+transferable regions with dynamic lifetimes. We require this
+generality in order to support streaming query operators, such as the
+one shown in Fig.~\ref{fig:motivating-eg}.

### 2015-11-10 paper/introduction.tex
updates to intro

@@
 Region-based memory management is a good fit for such programs.
-A region is a block of memory (consisting of potentially more than one contiguous
+A region is a block of memory (consisting of one or more contiguous
 range of memory locations, permitting a region to grow) that is used to realize
-a data-structure with a well-defined lifetime. In particular, the entire region
-is typically freed in one shot.
+a data-structure with a well-defined lifetime. 
+Once a region is created, individual objects may be dynamically allocated within
+the region over time. Finally, the entire region is typically freed in one shot.
 
@@
 namely the potential for dangling references and lack of memory safety. Automatic
-region-based memory managements guarantee memory safety, but impose various restrictions.
+region-based memory management systems guarantee memory safety, but impose various restrictions.
 Tofte and Talpin's system, for example, requires that the set of all regions (existing
@@
 that the data-structure is self-contained and can be safely transmitted across address spaces.
-\TODO{How accurate is this claim?}
 
@@
 i.e., a reference to an object that has been freed.
-\TODO{Or, should we say: no dangling reference is dereferenced}
 A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
@@
 
-We combine \emph{static} regions with the following protocol to ensure safety in these cases.
-To work with a dynamic region $D$, we must first \emph{open} region D1,
+We use \emph{static regions} (regions with lexically-scoped lifetimes), as well as the stack,
+for such temporary pointers (to objects inside a dynamic region) and use the following protocol
+to ensure safety of such pointers.
+To work with a dynamic region $D$, we must first \emph{open} region $D$,
 to indicate that $D$ should not be freed during this period.
@@
 The final piece required to ensure memory-safety is ensuring that the program correctly follows the above-mentioned
-protocol for dynamic regions. The protocol is essentially a typestate specification: an open region cannot be freed and,
-dually, a freed region cannot be opened.
+protocol for dynamic regions: \eg, an open region should not be freed and, dually, a freed region should not be opened.
+(A complete typestate specification of the protocol is presented later.)
 
-Dynamic regions are first-class objects in our system: e.g., in our running example,
+Dynamic regions are first-class objects in our system: \eg, in our running example,
 we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
@@
 We use a type system to statically ensure the first set of invariants. We use a dynamic check to ensure that the
-second set of invariants are satisfied.
+second set of invariants are satisfied at runtime.
 %

### 2015-11-10 paper/abstract.tex
Updates to abstract and introduction

Recent work in high-performance systems written in managed languages (such as Java or C\#) has shown
that garbage-collection can become a significant performance bottleneck in such systems.
A class of these systems, which we refer to as dataflow programs, create huge data-structures with well-defined
lifetimes that open up the possibility for significant performance improvements by using alternative memory
management techniques that can exploit knowledge of these data-structures' lifetimes.

In this paper, we present a memory management scheme that uses user-managed
memory regions that allow programmers to exploit the specific memory behaviours exhibited by dataflow programs.
To protect programmers from the dangers of manual memory management, we
supplement our memory management scheme with a region type system that
statically ensures the memory safety of region-annotated C\# programs,
even in presence of higher-order stateful functions. Further, we
describe a type inference algorithm that infers principal region types
for first-order programs thereby saving programmers the need to write
region annotations on types. Our inference algorithm is also very
effective, albeit incomplete, for higher-order programs.

%As a testament to efficacy of our approach, we demonstrate performance
%improvements over a range of realistic dataflow programs that were
%rewritten to use region-based manual memory management, but are
%nevertheless judged to be safe by our type system.

### 2015-11-10 paper/broom.tex
Related work.
Motivating examples added.
Added bib file.
Updates to abstract and introduction

--- 2015-11-02 paper/broom.tex
@@
 \newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
-\newcommand{\absof}[1]{\left\| #1 \right\| }
+%\newcommand{\absof}[1]{\left\| #1 \right\| }
+\newcommand{\absof}[1]{\thesemof{#1}}
 \newcommand{\broom}{{\sc Broom}\xspace}
@@
 \renewcommand{\bar}[1]{\overline{#1}}
-\newcommand{\angR}{\inang{\rhoalloc\rhobar \,|\, \phi}}
+\newcommand{\angR}{\inang{\rhoalloc,\rhobar \,|\, \phi}}
 \newcommand{\angT}{\inang{\tbar}}
@@
 \newcommand{\substFn}{\mathcal{S}}
-\newcommand{\mang}{\inang{\rhoalloc_m\bar{\rho_m} \,|\, \phi_m}}
+\newcommand{\mang}{\inang{\rhoalloc_m,\bar{\rho_m} \,|\, \phi_m}}
 \newcommand{\rhoset}{\rhoenv}
@@
 \newcommand{\phicx}{\Phi}
+\newcommand{\phictxt}{\phi_{cx}}
+\newcommand{\phicstr}{\phi_{cs}}
+\newcommand{\phisol}{\phi_{sol}}
 \newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
+morekeywords={var, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
 }}
@@
 \begin{abstract}
-
-Distributed execution engines, such as MapReduce and Dryad, represent
-a data-parallel computation as a dataflow graph with computational
-vertices connected by communication edges. The computation is run by
-executing the vertices of this graph on a set of available computers,
-while communication edges are materialized as either network
-connections, or appropriate inter-process communication channels.
-Although MapReduce and Dryad can perform optimal distribution the
-computation across available computers, they nevertheless rely on
-language runtimes, such as JVM or C\#.Net, to manage an atomic
-computation on a given machine.
-
-In this paper, we demonstrate that the general purpose garbage
-collection scheme employed by C\#.Net runtime is not very effective
-for dataflow programs to the extent that the latency due to GC is
-intolerably high in many practical examples. We propose a memory
-management scheme that combines user-managed memory regions with
-automatic garbage collection, and lets programmers exploit specific
-memory behaviours exhibited by dataflow programs. To protect
-programmers from the dangers of manual memory management, we
-supplement our memory management scheme with a region type system that
-statically ensures the memory safety of region-annotated C\# programs,
-even in presence of higher-order stateful functions. Further, we
-describe a type inference algorithm that infers principal region types
-for first-order programs thereby saving programmers the need to write
-region annotations on types. Our inference algorithm is also very
-effective, albeit incomplete, for higher-order programs. As a
-testament to efficacy of our approach, we demonstrate performance
-improvements over a range of realistic dataflow programs that were
-rewritten to use region-based manual memory management, but are
-nevertheless judged to be safe by our type system.
-
+\input{abstract}
 \end{abstract}
@@
 
+\input{related}
 % \appendix
@@
 \bibliographystyle{abbrvnat}
-\bibliography{catalyst}
+\bibliography{broom}
 

### 2015-11-10 paper/introduction.tex
Updates to abstract and introduction

@@
   \C{Region}, and are treated on par with other objects. \name is
-  particularly suited for implementing distributed dataflow operators
-  (e.g: \c{SELECT} and \c{JOIN}), whose memory behavior exhibits
-  certain characteristics that admits efficient and \emph{mostly} safe
-  region-based memory management. Experiments demonstrate that
-  \naiad~\cite{naiad} dataflow operators using programmer-managed
+  particularly suited for dataflow programs, such as~\naiad,
+  whose memory behavior exhibits
+  certain characteristics that admits efficient region-based memory management.
+  Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
   memory regions outperform their GC counterparts by a margin of upto
@@
   \item \name is equipped with a region type system that statically
-  guarantees safety all memory accesses in a well-typed program,
+  guarantees safety of all memory accesses in a well-typed program,
   subject to some conditions that can be checked efficiently at
-  run-time. The overhead of checking these conditions is less than
+  run-time.
+  The overhead of checking these conditions is less than
   \GK{x\%} in our experiments with \naiad workloads.

### 2015-11-10 paper/type-inference.tex

@@
 this means that our constraint solving algorithm visits and solves
-every constraint (or, every set of mutually recursive constraints)
+every constraint (or, every set of mutually dependent constraints)
 only once. It composes computed solutions to solve other constraints
@@
 solve a dependent constraint does not result in backtracking. We now
-describe our compositional constraint solving algorithm in detail.
+describe our compositional constraint solving algorithm in detail. To
+simplify the presentation of our algorithm, we assume that are no
+mutually recursive definitions in the source program. Recursive
+definitions are nonetheless allowed.
 
-\paragraph{Notation and terminology} In a validity constraint, a
-predicate variable occuring on the left side of the turnstile is said
-to occur negatively. In contrast, a predicate variable occuring on the
-right side is said to occur positively.
+\paragraph{Terminology} In a validity constraint, a predicate variable
+occuring on the left side of the turnstile is said to occur
+negatively, or with \emph{negative polarity}. In contrast, a predicate
+variable occuring on the right side is said to occur positively, or
+with \emph{positive polarity}. A validity constraint
+\emph{constrains} the set of predicate variables that occur
+negatively in the constraint, while it \emph{uses} the set of
+predicate variables that occur positively. A constraint is said to be
+\emph{recursive} if it constrains and uses a predicate variable.
+% A pair of validity
+% constraints are said to be \emph{mutually dependent} if there exists a
+% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
+% with opposite polarities in both the constraints. In such case, the
+% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
+% be mutually dependent. We call the transitive closure of mutual
+% dependency relation as \emph{transitive dependency} relation. Like
+% mutual dependency, transitive dependency is also extended to predicate
+% variables.
 
-Given a set of constraints, we first build a dependency graph ($G_c$)
-with validity constraints as nodes, and dependencies between them
-captured as edges. In particular, there exists an edge from a
-constraint $c_2$ to a constraint $c_1$ in the graph if any of the
-following conditions hold:
+Given a set of validity constraints, we first build a dependency graph
+($G_c$) with constraints as nodes, and dependencies between them
+captured as edges. There exists an edge from a constraint $c_2$ to a
+constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
+any of the following conditions hold:
 \begin{itemize}
-\item $c_2$ and $c_1$ have atleast one common predicate variable
-($\varphi$) that occurs positively in $c_2$ and negatively in $c_1$.
-\item The set of predicate variables that occur negatively in $c_2$ is
-a (non-strict) subset of the corresponding set for $c_1$.
+\item $c_1$ constrains a predicate variable that $c_2$ uses. 
+\item $c_1$ constrains a (non-strict) subset of predicate variables
+that $c_2$ constraint.
 \end{itemize}
@@
 before $c_2$ is solved, thus enforcing the rule that the precondition
-of a method must not depend on its calling context. Since mutually
-recursive definitions refer each other, the first condition also
-results in bidirectional dependencies between mutually recursive
-constraints (and self-loops over recursive constraints). The second
-condition captures two kinds of dependencies. First, the dependency
-from the constraints over a method precondition to the constraints
-over the precondition of the class containing the method. This
-captures our preference that the constraints over a class's region
-parameters should not depend on the idiosyncracies of its methods. Any
-additional constraints required by any of its methods must be captured
-in the precondition of the method itself (well-formedness rules allow
-this possibility). The second condition also adds bidirectional
-dependencies between two validity constraints over same set of
-predicate variables, thus requiring them to be solved simultaneously.
+of a method must not depend on its calling context. 
+% Note that the first condition results in self-loops over recursive
+% constraints in the dependency graph.  Since mutually recursive
+% definitions refer each other, the first condition also results in
+% bidirectional dependencies between mutually recursive constraints
+% (and self-loops over recursive constraints). 
+The second condition captures two kinds of dependencies. First, the
+dependency from the constraints over a method precondition to the
+constraints over the precondition of the class containing the method.
+This captures our preference that the constraints over a class's
+region parameters should not depend on the idiosyncracies of its
+methods. Any additional constraints required by any of its methods
+must be captured in the precondition of the method itself
+(well-formedness rules allow this possibility). The second condition
+adds bidirectional dependencies between validity constraints that
+constraint the same set of predicate variables.
 
@@
 after solving a constraint set are applied to the constraints in
-subsequent sets before attempting to solve them.
+subsequent sets before attempting to solve them. Consequently, when
+the turn of a constraint set ($C$) arrives during the constraint solving
+process, it satisfies certain properties:
+\begin{itemize}
+\item There exists only one predicate variable that is either
+constrained or used by the constraints in the set ($C$). The variable is
+called set's \emph{subject}. This property follows from (a). the fact
+that all the dependency constraints have already been solved (and
+solutions applied), and (b). the assumption that there are no mutually
+recursive definitions. 
+\item All the constraints that constrain the set's subject are present
+in the set. This follows from our definition of the dependency relation.
+\end{itemize}
+
+Solving the set ($C$) of constraints entails finding an assignment for
+$C$'s subject, and also all the region variables that occur free in
+$C$, such that the solutions satisfy the well-formedness constraints.
+We now formalize the constraint satisfaction problem, and its
+solution.
+
+\begin{definition}
+\emph{(\textbf{Constraint Satisfaction Problem})}
+A constraint satisfaction problem is a tuple
+$(C,\rhoenv,\varphi,\bar{\rho} \in \bar{\rhoenv})$, where $C$, the set
+of validity constraints, is of the form:
+\begin{center}
+\(
+  \bigcup_{i=1}^n \{
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {\phicstr^{i} \conj F_i(\varphi)}
+  \}
+\)
+\end{center}
+The subject of the set is $\varphi$, and its free region variables
+range over $\rhobar$. The solution to the constraint satisfaction
+problem is a pair $(\substFn,\phisol)$, where $\substFn$ is a map from
+region variables to region identifiers such that $\substFn(\rho_j) \in
+\rhoenv_j$, for every $j$, and $\phisol$ is a constraint formula such
+that:
+\begin{itemize}
+\item $\phisol$ is well-formed under $\rhoenv$ (i.e.,
+$\tywf{\rhoenv}{\phisol}$).
+\item $\phisol$ satisfies every validity constraint in $\substFn(C)$.
+That is, the following sequent is valid for every $i$:
+\begin{center}
+\(
+  \isvalid{\substFn(\phictxt^{i}) \conj \phisol}
+          {\substFn(\phicstr^{i}) \conj (\substFn \circ F_i)(\phisol))}
+\)
+\end{center}
+\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
+that $\phisol'$ satisfies every validity constraint in $\substFn(C)$,
+and $\phisol'$ is strictly weaker than $\phisol$ (i.e.,
+$\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
+$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
+\end{itemize}
+\end{definition}
 

### 2015-11-11 paper/introduction.tex
Updates to introduction

@@
 Consider the example shown in Fig.~\ref{fig:motivating-eg}.
-This code represents the logic for a streaming query operator (also
-known as a continuous query). The operator receives a continuous
-stream of input messages, which are processed via a user-defined
-function, and batched into time-windows associated with a logical
-timestamp. Messages may be delivered out-of-order and multiple
-messages with the same timestamp may be delivered. A timing-message,
-periodically indicates that no more input messages with a timestamp
-less than or equal to a given value $t$ will be subsequently
-delivered. At this point, the operator can create the output message
-corresponding to time $t$ and deliver it to its successors.
+This code represents the logic for a streaming query operator.
+The operator receives a stream of input messages, each associated
+with a time (window) $t$, processed by method \texttt{onReceive}.
+Each input message contains a list of inputs, each of which is processed
+by applying a user-defined function to create a corresponding output.
+The operator may receive multiple messages with the same timestamp
+(and messages with different timestamps may be delivered out of order).
+A timing-message (an invocation of method \texttt{OnNotify}) indicates that
+no more input messages with a timestamp $t$ will be subsequently delivered.
+At this point, the operator completes the processing for time window $t$ and sends
+a corresponding output message to its successor.
 
@@
 data-structures are very large, as is the case with many streaming
-big-data analysis systems, the garbage collection overheads become
+big-data analysis systems, the garbage collection overhead becomes
 significant.
@@
 However, these data-structures have a well-defined lifetime, which can
-be exploited to achieve better performance 
+be exploited to achieve better performance.
 % using alternative memory management techniques.
@@
 the region over time. Finally, the entire region is typically freed in one shot.
+In this example, the output to be constructed for each time window $t$ can be
+implemented using a separate region.
 
@@
 region-based memory management systems guarantee memory safety, but impose various restrictions.
-Tofte and Talpin's system, for example, requires that the set of all regions (existing
-at a point in time) forms a stack. Specifically, this requires that the lifetimes of
+Tofte and Talpin's system~\cite{tofte94,tofte97} , for example, uses lexically scoped regions.
+At runtime, the set of all regions (existing at a point in time) forms a stack. Thus, the lifetimes of
 all regions must be well-nested: it is not possible to have two regions whose lifetimes
 overlap, with neither one's lifetime contained within the other.
-Unfortunately, the data-structures in the above example do not satisfy this restriction.
-We refer to such regions as \emph{dynamic} regions.
+Unfortunately, the data-structures in the above example do not satisfy this restriction
+(as the output messages for multiple time windows may be simultaneously live, without
+any containment relation between their lifetimes).
+We refer to regions with lexically scoped lifetimes as \emph{stack regions} and to regions 
+that do not have such a lexically scoped region as \emph{dynamic} regions.
 
@@
 
-We use \emph{static regions} (regions with lexically-scoped lifetimes), as well as the stack,
+We use \emph{stack regions} (regions with lexically-scoped lifetimes), as well as the stack,
 for such temporary pointers (to objects inside a dynamic region) and use the following protocol
@@
 to indicate that it is safe to free $D$.
+In general, a dynamic region may be opened and closed multiple times before it is finally freed.
+
 Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,

### 2015-11-11 paper/type-inference.tex
Edits.

@@
 % region-polymorphic methods, where similar situation arises. 
-In general, validity constraints generated by our algorithm are of the
-form:
+In general, validity constraints generated by our algorithm assume one
+of the following two forms:
 \begin{center}
 \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
-           {\phicstr \conj \bigwedge_{j}F_j(\varphi_j)}\)
+           {\phicstr} \qquad
+   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+           {F_j(\varphi_j)}
+\)
 \end{center}
-Where $\varphi_i$'s denote the unknown preconditions of the class or
+Where $\varphi_i$'s denote the unknown preconditions of the class and
 the method under elaboration. If the constraint is generated while
@@
 ($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-of predicate variables and pending substitutions.  They capture the
-concrete relationships between regions that are \emph{known} to hold,
-and \emph{required} to hold, respectively, under the context in which
-the constraint was generated.
+of predicate variables and pending substitutions. While $\phictxt$
+captures relationships that are \emph{known} to hold between concrete
+region identifiers (i.e., $\rgn$'s) when the constraint was generated,
+$\phicstr$ captures relationships that are \emph{required} to hold
+among region varibles (i.e., $\rho$'s), or relationships between
+region variables and identifiers. Each region variable occuring in
+$\phicstr$ has an associated well-formedness constraint, which
+specifies its unification domain. The unification domain of a
+constraint is the union of unification domains of all region variables
+occuring in the constraint.
 
@@
 \begin{itemize}
-\item There exists only one predicate variable that is either
+\item There exists only one predicate variable ($\varphi$) that is either
 constrained or used by the constraints in the set ($C$). The variable is
@@
 Solving the set ($C$) of constraints entails finding an assignment for
-$C$'s subject, and also all the region variables that occur free in
-$C$, such that the solutions satisfy the well-formedness constraints.
-We now formalize the constraint satisfaction problem, and its
-solution.
+$C$'s subject ($\varphi$), and also all the region variables
+($\rhobar$) that occur free in $C$, such that the solution satisfies
+well-formedness constraints on $\varphi$ and $\rhobar$. To simplify
+presentation, we think of $C$ as being parameterized on
+$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
+\rbrack$. We now formalize the constraint satisfaction problem, and
+its solution.
 
 \begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem})}
-A constraint satisfaction problem is a tuple
-$(C,\rhoenv,\varphi,\bar{\rho} \in \bar{\rhoenv})$, where $C$, the set
-of validity constraints, is of the form:
+\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
+satisfaction problem is a tuple $\CSP$, where $C\lbrack
+\varphi,\rhobar \rbrack$ is a set of validity constraints, where
+$i$'th validity constraint assumes one of the following forms:
 \begin{center}
 \(
-  \bigcup_{i=1}^n \{
     \isvalid{\phictxt^{i} \conj \varphi}
-            {\phicstr^{i} \conj F_i(\varphi)}
-  \}
+            {\phicstr^{i}}\qquad
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {F_i(\varphi)}
 \)
 \end{center}
-The subject of the set is $\varphi$, and its free region variables
-range over $\rhobar$. The solution to the constraint satisfaction
-problem is a pair $(\substFn,\phisol)$, where $\substFn$ is a map from
-region variables to region identifiers such that $\substFn(\rho_j) \in
-\rhoenv_j$, for every $j$, and $\phisol$ is a constraint formula such
-that:
+$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
+union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
+unification domain ($\rhoenv$) of the CSP.  The solution to the
+constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
+$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
+$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
+constraint formula such that:
 \begin{itemize}
-\item $\phisol$ is well-formed under $\rhoenv$ (i.e.,
-$\tywf{\rhoenv}{\phisol}$).
-\item $\phisol$ satisfies every validity constraint in $\substFn(C)$.
-That is, the following sequent is valid for every $i$:
-\begin{center}
-\(
-  \isvalid{\substFn(\phictxt^{i}) \conj \phisol}
-          {\substFn(\phicstr^{i}) \conj (\substFn \circ F_i)(\phisol))}
-\)
-\end{center}
+\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
+$\tywf{\rhoenv_{\varphi}}{\phisol}$).
+\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
+is valid.
+% \begin{center}
+% \(
+%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
+%           {\phicstr^{i} \conj  F_i(\phisol))}
+% \)
+% \end{center}
 \item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
-that $\phisol'$ satisfies every validity constraint in $\substFn(C)$,
-and $\phisol'$ is strictly weaker than $\phisol$ (i.e.,
-$\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
+that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
+\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
+(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
 $\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
@@
 
-\subsection{The Constraint Satisfaction Problem}
+\subsection{Solving the CSP}
 
-We now formally define the constraint satisfaction problem over sets
-of constraints. For simplicity, we ignore mutual recursion and assume
-that a a set only contains (possibly recursive) constraints over a
-single predicate variable. The general form of such a set of
-constraints is as following:
+The first step of solving the CSP $\CSP$ is to cast it as an
+equivalent problem $\csp$ involving a single validity constraint. The
+single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
 \begin{center}
 \(
-  \bigcup_{i=1}^n \{
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {\phicstr^{i} \conj \bigwedge_{k=1}^n F_k(\varphi)}
-  \}
+  \isvalid{ \Phicx \conj \varphi}
+        {\Phics \conj  \bigwedge_i F_i(\varphi)}
 \)
 \end{center}
-% The solution to the above set of constraints is a tuple $(\substFn,)$
-% problem is to find a map
+Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
+\phicstr^{i}$. To see why both CSPs are equivalent, consider two
+distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
+$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
+\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
+generated before $c_j$ by the constraint generation algorithm. Observe
+that when our constraint generation algorithm generates a constraint,
+all relationships between concrete region identifiers referred by the
+consequent of constraint are already present in its antecedent (this
+includes the unification domains of region variables referred by the
+consequent).  Since no new relationships between existing regions are
+added when a new constraints is generated, $\phictxt^j$ either
+describes the same relationships that are already described by
+$\phictxt^i$, or describes relationships among region identifiers that
+are new, and not relevant to $c_i$. Therefore, strengthening the
+context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
+nor strengthens $c_i$ (resp.  $c_j$).
 
-% After solving all the dependencies and applying solutions, the most
-% general form of a set of constraints is the following:
+\subsubsection{Non-Recursive Constraints}
+\label{sec:csolve}
+
+We first describe how we solve a non-recursive CSP $\csp$ where
+c$\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
+\conj \varphi}{\Phics}$.
+
+Our approach is based on the observation that $\FB$ does not admit
+null values or uninitialized variables, thus forcing every region
+variable to be unified with some concrete region. The constraint
+formula $\Phics$ captures all such unification constraints on
+$\rhobar$. Since the set of all concrete region identifiers is the
+unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
+with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
+exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
+\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
+in which case the well-formedness constraint on $\rho_i$ is not
+satisfied, and constraint solving must fail.  Therefore, there exists
+a unique assignment $\substFn$ such that $c\lbrack
+\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
+$\phi_{sol}$.
+
+To obtain $\phisol$, we make use of another observation. Let $c\lbrack
+\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
+\begin{center}
+\(
+  \isvalid{ \Phicx \conj \varphi}
+        {\Phics'}
+\)
+\end{center}
+Consider a maximally weak formula $\phi$ such that
+$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
+% Since $\phi$ is
+% maximally weak, there does not exist a $\phi'$ satisfying all the
+% below three conditions:
+% \begin{itemize}
+% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
+% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
+% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
+% \end{itemize}
+Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
+\subseteq \rhoenv$, we have two cases:
+\begin{itemize}
+
+\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
+a solution to $\varphi$. 
+% Furthermore, it is the weakest solution, because it is equivalent to
+% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
+% \substFn(\rho_j))$, and any other solution has to stronger than
+% $\Phics$ under the same context.
+
+\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
+contains at least one equality or outlives constraint on two region
+identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
+\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
+\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
+constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
+is not maximally weak). Let us denote such constraint on $\rgn_i$ and
+$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
+$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
+Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
+$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
+$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
+\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
+this is impossible. To see why, recall that all the constraints on
+identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
+by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
+Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
+only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
+solution $\phisol$ cannot exist.
+\end{itemize}
+The above discussion hints at an algorithm to compute a solution to
+$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
+{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
+{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
+solution to $\varphi$.
+
+\subsubsection{Solving Recursive Constraints}
+
+We now extend constraint solving to recursive CSP $\csp$, where
+$c\lbrack \varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx
+\conj \varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake
+of brevity, we define $G(\varphi) = \Phicx \conj \bigwedge_i
+F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
+\begin{center}
+\(
+  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
+\)
+\end{center}
+To solve the above recursive constraint, we start with the observation
+that the set of all constaints ($\phi$) over $\rhoenv \cup
+\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
+\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
+monotone over the lattice:
+\begin{center}
+  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
+  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
+\end{center}
+From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
+point ($\phi_f$), with following properties:
+\begin{itemize}
+\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
+$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
+\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
+G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
+{\cdot} {\phi_f' \Rightarrow \phi_f}$.
+\end{itemize}
+We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
+convert the recursive constraint to the following non-recursive
+constraint:
+\begin{center}
+\(
+  \isvalid{\Phicx \conj \varphi}{\phi_f}
+\)
+\end{center}
+The technique described in \S~\ref{sec:csolve} now suffices to solve
+the above constraint.
+
+% Note that $F$ is a non-idempotent substitution function with following
+% properties:
+% \begin{itemize}
+% \item $F$ is a monotone: 
 % \begin{center}
-% \(
-%   \bigcup_{i=1}^n \{
-%     \bigcup_{j=1}^m \{ 
-%       \isvalid{\phictxt^{ij} \conj \varphi_i}{\phicstr^{ij} \conj
-%       \bigwedge_{k=1}^n F_k(\varphi_k)}
-%       \}
-%     \}
-% \)
+%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
+%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
 % \end{center}
-% Our technical report~\cite{techrep} contains formal proofs
-% establishing the equivalence of the above constraint set and the one
-% below, with respect to the constraint satisfaction problem:
+
+% \item $F$ distributes over conjunction (mind the syntactic equality):
 % \begin{center}
-% \(
-%   \bigcup_{i=1}^n \{
-%     \isvalid{\bigwedge_{j=1}^m \phictxt^{ij} \conj \varphi_i}
-%             {\bigwedge_{j=1}^m \phicstr^{ij} \conj \bigwedge_{k=1}^n F_k(\varphi_k)}
-%   \}
-% \)
+%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
+%   F(\phi_2)$
 % \end{center}
-% Our technical report~\cite{techrep} contains formal proofs
+% \end{itemize}
 
-\subsubsection{Solving Non-Recursive Constraints}
+% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
+% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
+% $\phi_f$ be the GFP of $G$. Following properties ensue:
+% \begin{itemize}
+% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
+% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
+% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
+% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
+% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
+% \end{itemize}
 
-Non-recursive constraints are of the form $\isvalid{\phictxt \wedge
-\varphi}{\phicstr}$, where the predicate variable $\varphi$ has an
-associated well-formedness constraint $\tywf{\rhoenv}{\varphi}$.  The
-formula $\phicstr$ may contain free region variables ($\rho$), each
-with its own well-formedness constraint ($\rho \in \rhoenv$)
-specifying its domain of unification. Solving the constraint entails
-generating well-formed assignments to $\varphi$ and all free region
-variables in $\phicstr$.
+% \begin{theorem}
+% $\phi_f$ is a solution.
+% \end{theorem}
+% \begin{proof}
+% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
+% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
+% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
+% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
+% solution for $\varphi$.
+% \end{proof}
 
-\begin{definition}
-\emph{(\textbf{\csolve})}
-\label{def:csolve}
-Let $L(S,R_1,R_2)$ be the following constraint language:
-\begin{smathpar}
-\begin{array}{lcl}
-x,y & \in & S\\
-\phi & \coloneqq & true \ALT R_1(x,y) \ALT R_2(x,y) \ALT \phi \conj \phi \\
-\end{array}
-\end{smathpar}
-\csolve is a function that accepts:
-\begin{itemize}
-\item A set $C$ of constants, and a set $V$ of variables
-\item A constraint $\phi$
-\end{itemize}
+% \begin{theorem}
+% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
+% \phisol}{\phi_f}$
+% \end{theorem}
+% \begin{proof}
+%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
+%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
+%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
+%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$
 
-\end{definition}
+% \end{proof}

### 2015-11-11 paper/overview.tex

@@
 by \C{letregion} expressions as \emph{stack regions}. The stack
-discipline induces \emph{outlives} relationship among regions created
+discipline induces an \emph{outlives} relationship among regions created
 by nested \C{letregion}s, where the region introduced by the outer
@@
       A a1 = new@R1 A();
-      a1.x = new@R0 Object();
-      a0.x = new@R1 Object();
+      a1.x = new@R0 Object(); // safe & legal
+      a0.x = new@R1 Object(); // unsafe & illegal
       ...
@@
 
-As evident from the examples shown above, the suffix \C{@R} in
-\C{new@R} instructs the \C{new} keyword to allocate the newly created
-object in the region labeled \C{R}. However, \name does not require
-the allocation region to be specified following a \C{new}. In the
+As the preceding examples illustrate, the suffix \C{@R} in
+\C{new@R} indicates that the newly created object should be allocated in the region labeled \C{R}.
+However, \name does not require the allocation region to be specified following a \C{new}.
+This is critical, as it enables \name applications to use existing region-oblivious C\# libraries.
+In the
 absence of explicit specification, \name allocates the newly created
@@
 
-\subsection{Boxed and Unboxed Values}
-
-In \name, like in \csharp, values of some primitive types, such as
-integers and bools, are unboxed, whereas others, such as strings are
-arrays, are boxed. Unlike unboxed values, boxed values are objects,
-hence passed by reference. It is therefore important to be mindful
-of allocation regions for strings and arrays, but not so much for
-integers and bools. For instance, in the following code, the
-assignment to integer variable \C{x} is safe, but the assignment to
-the string variable \C{y} is unsafe, as the allocation region (\C{R0})
-of the string constant has lesser lifetime than \C{y}.
-\begin{center}
-\begin{codejava}
-  int x;
-  string y;
-  letregion R0 {
-    x = 2;
-    y = "Hello, World!";
-  }
-  ...
-\end{codejava}
-\end{center}
-Accordingly, \name's region type system lets the former assignment
-pass by, but flags the later as an error.
+%\subsection{Boxed and Unboxed Values}
+%
+%In \name, like in \csharp, values of some primitive types, such as
+%integers and bools, are unboxed, whereas others, such as strings are
+%arrays, are boxed. Unlike unboxed values, boxed values are objects,
+%hence passed by reference. It is therefore important to be mindful
+%of allocation regions for strings and arrays, but not so much for
+%integers and bools. For instance, in the following code, the
+%assignment to integer variable \C{x} is safe, but the assignment to
+%the string variable \C{y} is unsafe, as the allocation region (\C{R0})
+%of the string constant has lesser lifetime than \C{y}.
+%\begin{center}
+%\begin{codejava}
+%  int x;
+%  string y;
+%  letregion R0 {
+%    x = 2;
+%    y = "Hello, World!";
+%  }
+%  ...
+%\end{codejava}
+%\end{center}
+%Accordingly, \name's region type system lets the former assignment
+%pass by, but flags the later as an error.
 
@@
 
-Memory safety in \name is primarily due to its region type system,
-although few runtime checks are needed.
-% However, the type system does not ensure the integrity of a small
-% subset of pointers, called the \emph{unsafe set}, for which runtime
-% verification is needed to guarantee safety. This is more a design
-% decision rather than an unfortunate inevitability, as the following
-% discussion elaborates.
-At a fundamental level, the region type system's approach to memory
-safety involves approximating (a). the set of memory regions that are
-live at any given program location\footnote{We say the type system
-tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
-relationships between them, and using this information to preempt
-potentially unsafe references from being created. These include:
-\begin{itemize}
-\item References into dead/non-existing regions,
-\item References from longer-living stack regions to shorter-living ones, and 
-\item References that escape transferable regions.
-\end{itemize}
-Maintaining precise approximation of liveness and outlives information
-for stack regions is easy. However, for dynamic (transferable) regions,
-statically determining liveness is equivalent to determining if a C
-pointer is null in presence of dynamic memory allocation, which is
-undecidable in general. Systems that attempt to ensure full memory
-safety in presence of dynamic memory (e.g., ~\cite{henglein},
-~\cite{rust}) do so by conservatively approximating liveness
-information for memory (regions). Maintaining a good conservative
-approximation requires them to employ techniques such as linear
-types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
-restrictions that are often too hard to program
-around\footnote{Internet has many tutorials on how to \emph{fight} the
-borrow checker in Rust.}. This is particularly unacceptable in the
-context of \name, as dataflow programs are often run offline, and the
-concern for performance and developer productivity outweighs the
-concern for safety.
+The key to memory safety in \name is the following restriction:
+an object $o_1$ in a region $R_1$ is allowed to store a pointer to
+an object $o_2$ in a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.
+(A similar restriction applies in the case where $o_1$ is a stack-allocated variable.)
 
-\name's region type system therefore adopts the novel approach of
-overapproximating liveness information for transferable regions, i.e.,
-it might assume that a transferable region is live when, in fact, it
-might have already been transferred or freed. This imprecision should
-be a cause of concern as it can lead to cases where creation of unsafe
-references, and their subsquent dereferencing go unnoticed by the type
-system. Considering that stack regions can contain pointers into
-transferable regions, the worst case cost of overapproximating
-liveness is therefore the integrity of \emph{every} pointer on stack.
+Enforcing this restriction is simple in the case of stack regions since the outlives relation
+between stack regions can be inferred from their lexical nesting. Unfortunately,
+inferring outlives relations between transferable regions is not easy.
+\name imposes the following protocol on the use of transferable regions to help simplify
+this check.
 
-Fortunately, \name avoids this worst case scenario.  \name achieves
-this by imposing a simple state transition discipline on the lifetime
-of a transferable region.  The relevant finite state machine is shown
-in Fig.~\ref{fig:region-fsm}.
+A transferable region (that has not been freed or transferred) can be in one of two possible
+states, \emph{open} or \emph{closed}. A newly created region is in the closed state.
+A region must be opened, using the open construct (explained in detail below), in order
+to read or update or allocate an object within that region.
+An open region cannot be freed or transferred. 
+In particular, an open region is guaranteed to be live for the entire duration of the open construct.
+This allows the type system to infer a valid outlives relation between the opened region
+and any stack region that is nested within the open construct.
+
+%Memory safety in \name is primarily due to its region type system,
+%although few runtime checks are needed.
+%% However, the type system does not ensure the integrity of a small
+%% subset of pointers, called the \emph{unsafe set}, for which runtime
+%% verification is needed to guarantee safety. This is more a design
+%% decision rather than an unfortunate inevitability, as the following
+%% discussion elaborates.
+%At a fundamental level, the region type system's approach to memory
+%safety involves approximating (a). the set of memory regions that are
+%live at any given program location\footnote{We say the type system
+%tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
+%relationships between them, and using this information to preempt
+%potentially unsafe references from being created. These include:
+%\begin{itemize}
+%\item References into dead/non-existing regions,
+%\item References from longer-living stack regions to shorter-living ones, and 
+%\item References that escape transferable regions.
+%\end{itemize}
+%Maintaining precise approximation of liveness and outlives information
+%for stack regions is easy. However, for dynamic (transferable) regions,
+%statically determining liveness is equivalent to determining if a C
+%pointer is null in presence of dynamic memory allocation, which is
+%undecidable in general. Systems that attempt to ensure full memory
+%safety in presence of dynamic memory (e.g., ~\cite{henglein},
+%~\cite{rust}) do so by conservatively approximating liveness
+%information for memory (regions). Maintaining a good conservative
+%approximation requires them to employ techniques such as linear
+%types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
+%restrictions that are often too hard to program
+%around\footnote{Internet has many tutorials on how to \emph{fight} the
+%borrow checker in Rust.}. This is particularly unacceptable in the
+%context of \name, as dataflow programs are often run offline, and the
+%concern for performance and developer productivity outweighs the
+%concern for safety. 
+%
+%\name's region type system therefore adopts the novel approach of
+%overapproximating liveness information for transferable regions, i.e.,
+%it might assume that a transferable region is live when, in fact, it
+%might have already been transferred or freed. This imprecision should
+%be a cause of concern as it can lead to cases where creation of unsafe
+%references, and their subsquent dereferencing go unnoticed by the type
+%system. Considering that stack regions can contain pointers into
+%transferable regions, the worst case cost of overapproximating
+%liveness is therefore the integrity of \emph{every} pointer on stack.
+
+%Fortunately, \name avoids this worst case scenario.  \name achieves
+%this by imposing a simple state transition discipline on the lifetime
+%of a transferable region.  The relevant finite state machine is shown
+%in Fig.~\ref{fig:region-fsm}.
+
+The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
 
@@
 As evident in the above examples, the safety of memory accesses in
-\name is now subject to the condition that the lifetime of every
-transferable region aligns with the state transition discipline shown
-in Fig.~\ref{fig:region-fsm}. If the transition discipline is somehow
+\name is now subject to the condition that every transferable region
+correctly follows the state transition discipline shown
+in Fig.~\ref{fig:region-fsm}. If this is
 guaranteed, then \name's region type system statically guarantees the
@@
 the problem of ensuring memory safety in \name programs to the problem
-of enforcing state transition discipline on the lifetimes of
-transferable regions. In \name, this enforcement is done at runtime by
-explicitly keeping track \emph{current state} for $\RgnZ$ objects, and
-preventing them from going through invalid transitions.
+of enforcing the state transition discipline for transferable regions.
+
+In \name, this enforcement is done at runtime by explicitly keeping
+track of the \emph{current state} for $\RgnZ$ objects, and
+checking the validity of every open, transfer, or free operation
+and throwing an exception if it is invalid.
+The challenge in enforcing this discipline statically is that transferable regions
+are first-class objects. Hence, the program can create multiple aliases for
+the same region, \eg, open it via one alias and free it via another.
+Typestate verification in the presence of aliases is hard.
+The checking can be done statically by preventing the creation of aliases using, \eg, linear types
+or unique types. However, this would be quite restrictive, in terms of expressiveness.
+
+\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
+in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
+as well as the runtime overhead of checking the region's state transition discipline is acceptable.
 

### 2015-11-11 paper/type-inference.tex
Type Inference.
Edits.

@@
 region variable. Likewise, given the region-annotated definition of
-\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for
-the method \C{mkSnd} shown below is \footnote{In our
-exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
-subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
-accept one region parameter - the allocation region of their
-objects.}\footnote{We abuse arrow notation to also represent types of
-methods, but unlike function types, there is no allocation region
-annotation atop the arrow in a method type.}
+\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
+method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
+is \footnote{In our exposition, we assume that classes $\C{A}$ and
+$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
+$\ObjZ$, they accept one region parameter - the allocation region of
+their objects.}\footnote{We abuse arrow notation to also represent
+types of methods, but unlike function types, there is no allocation
+region annotation atop the arrow in a method type.}
 $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
@@
 variables, and $\varphi_0$ is a fresh predicate variable denoting
-unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  
-\begin{codejava}
-  unit mkSnd(Pair<A,B> p) {
-    return (p.snd = new B());
-  }
-\end{codejava}
-Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
-$\rho$ is fresh. For a class, a template is computed in two steps. In
-the first step, we templatize the types of all its fields, constructor
+unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
+template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
+fresh. For a class, a template is computed in two steps. In the first
+step, we templatize the types of all its fields, constructor
 arguments, and arguments and return values of all its methods, along
@@
 We first describe how we solve a non-recursive CSP $\csp$ where
-c$\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
+$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
 \conj \varphi}{\Phics}$.

### 2015-11-11 paper/type-inference.tex
Edits.

@@
 region annotation atop the arrow in a method type.}
-$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
-\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
-\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
+$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
+\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
+\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
 variables, and $\varphi_0$ is a fresh predicate variable denoting
@@
 and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
-respectively. It can be templatized as following:
+respectively. It can be templatized as following (method definitions
+are left intact while templatizing, hence they are elided below):
+\vspace*{-0.1in}
 \begin{center}
@@
           $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
-  $a@\rho_0$ fst;
-  $b@\rho_1$ snd;
+  $a@\rho_0$ fst; $b@\rho_1$ snd;
   Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
-  $...$
+  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5}$() { $...$ }
 }
@@
 \end{center}
+\vspace*{-0.1in}
 % \footnote{In reality, the generalized region variables only come
@@
 % types do not contain any free region variables.}
-The template elides $\C{Pair}$'s methods, whose region type templates
-contain no free variables. Among the region parameters of the class
-template, $\rho_{0-4}$ are obtained by generalizing free region
-variables in the types of its class's fields, constructor arguments,
-and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
-fresh region variable denoting the allocation region argument. Fresh
-predicate variable $\varphi_0$ denotes unknown constraints over
-$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
-well-formed region-annotated class definition in $\FB$.
+% The template elides $\C{Pair}$'s methods, whose region type
+% templates contain no free variables. Among the region parameters of
+% the class template, $\rho_{0-4}$ are obtained by generalizing free
+% region variables in the types of its class's fields, constructor
+% arguments, and its superclass type.  The remaining parameter
+% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
+% region argument. Fresh predicate variable $\varphi_0$ denotes
+% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
+% hold for template to be a well-formed region-annotated class
+% definition in $\FB$.
 
@@
 templatized with respect to the class template computed in the first
-step, such that their region annotations are exactly same as the
-region parameters of the class. For example, consider a generic
-$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
-type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
-where $\C{a}$ is the type of the data stored in the linked list node.
-To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
-recursive occurance in the type of $\C{next}$ field, and templatize
-the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
-type template of $\C{data}$ field, we compute the class's template as
-following:
-\begin{center}
-$\C{ListNode}\inang{\C{a}\extends\ObjZ}
-\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
-\end{center}
-Next, we trivially templatize the type of $\C{next}$ field in the body
-of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
-The resulting class represents a linked list with spine in the region
-$\rhoalloc_0$ and data objects in the region $\rho_0$.
-
+step, such that their region annotations are \emph{exactly} same as
+the region parameters of the class. Applying this technique on a
+recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
+example, results in a template that uses same region variable to
+denote allocation regions of the current node, and the node that the
+current node refers to. 
+% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
+% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
+% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
+% data stored in the linked list node.  To templatize the
+% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
+% occurance in the type of $\C{next}$ field, and templatize the type
+% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
+% template of $\C{data}$ field, we compute the class's template as
+% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
+% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
+% \end{center} Next, we trivially templatize the type of $\C{next}$
+% field in the body of the class as
+% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
+% resulting class represents a linked list with spine in the region
+% $\rhoalloc_0$ and data objects in the region $\rho_0$.
 % Observe that our strategy of trivially templatizing the recursive
@@
 % region variables are introduced.
-The templatization technique we described above for recursive class
-definitions can be extended to mutually recursive definitions in a
+This technique can be extended to mutually recursive definitions in a
 straightforward manner, by simultaneously templatizing them. Using the
@@
 Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-this elaboration for a subset of expressions in $\absof{\FB}$, whose
-corresponding $\FB$ expressions have been ascribed static semantics in
-Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
-context as the expression typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning. The function traverses expressions in a
-syntax-directed manner of a type checker, introducing fresh region
-type templates for unknown region types, while generating constraints
-over region and predicate variables. The precise nature of generated
-constraints is explained in \S\ref{sec:fb-constraintsem}, but in
-summary, they capture the relationships between the type templates of
-various subexpressions and their well-formedness. Note that
-$\elabExpr$ returns the region type template of the subexpression,
-which is used to generate constraints for the expression. Functions
-$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
-implement type well-formedness and subtype judgments from
-Fig.~\ref{fig:fb-staticsem}, respectively.
+this elaboration for a subset of expressions in $\absof{\FB}$
+%, whose corresponding $\FB$ expressions have been ascribed static
+%semantics in Fig.~\ref{fig:fb-staticsem}. 
+$\elabExpr$ is defined under the same context as the expression typing
+judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
+and $\env$ retaining their meaning. The function traverses expressions
+in a syntax-directed manner of a type checker, introducing fresh
+region type templates for unknown region types, while generating
+constraints over region and predicate variables. The precise nature of
+generated constraints is explained in \S\ref{sec:fb-constraintsem},
+but in summary, they capture the relationships between the type
+templates of various subexpressions and their well-formedness. Note
+that $\elabExpr$ returns the region type template of the
+subexpression, which is used to generate constraints for the
+expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
+shown) used by $\elabExpr$ implement type well-formedness and subtype
+judgments from Fig.~\ref{fig:fb-staticsem}, respectively.
 
@@
 class definitions, respectively. Both functions first build a context
-($\A$) containing a set ($ \rhoenv$) of region variables denoting
-regions that are currently live, a map ($\aenv$) mapping type
-variables to their bounds, and a constraint formula ($\phicx$)
-capturing constraints over live region variables. We use predicate
-variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
+($\A$) containing a set ($ \rhoenv$) of regions that are currently
+live, a map ($\aenv$) mapping type variables to their bounds, and a
+constraint formula ($\phicx$) capturing constraints over live region
+variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
+capture constraints over variables in $\rhoenv$ denoting the fact that
+such constraints are yet to be inferred. Both elaboration functions
+generate new constraints by checking well-formedness and subtype
+relationships among types of their constituents. A fresh constraint
+($C_4$) generated by $\elabClass$ captures a couple of safety
+conditions: first, the allocation regions of objects referred by the
+instance variables should outlive the allocation region of the
+instance itself, and second, the allocation regions of a class type
+and its superclass type must be the same.
 
-Function $\elabMeth$ elaborates a method definition of class $B$. It
-calls $\elabExpr$ with the context $\A$, its allocation context
-parameter ($\rhoallocm$), and a type environment ($\env$) that
-contains region type bindings for all the arguments of the method,
-including the implicit $\C{this}$ argument. The region type template
-returned by $\elabExpr$ for the method body is checked against its
-expected type (derived from the type template of the method)
-generating more constraints. The function then returns the elaborated
-method definition and the set of constraints.
+% Function $\elabMeth$ elaborates a method definition of class $B$. It
+% calls $\elabExpr$ with the context $\A$, its allocation context
+% parameter ($\rhoallocm$), and a type environment ($\env$) that
+% contains region type bindings for all the arguments of the method,
+% including the implicit $\C{this}$ argument. The region type template
+% returned by $\elabExpr$ for the method body is checked against its
+% expected type (derived from the type template of the method)
+% generating more constraints. The function then returns the elaborated
+% method definition and the set of constraints.
 
-$\elabClass$ elaborates the definition of a class $B$. It relies on
-$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
-hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
-constructor ($k$) and method definitions ($\bar{d}$), respectively. To
-the set of constraints returned by these functions, $\elabClass$ adds
-constraints generated by checking the well-formedness of the type
-templates of its superclass and fields, and also a new constraint
-capturing a couple of safety conditions: first, the allocation regions
-of objects referred by the instance variables should outlive the
-allocation region of the instance itself, and second, the allocation
-regions of a class type and its superclass type must be the same.
+% $\elabClass$ elaborates the definition of a class $B$. It relies on
+% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
+% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+% the set of constraints returned by these functions, $\elabClass$ adds
+% constraints generated by checking the well-formedness of the type
+% templates of its superclass and fields, and also a new constraint
+% capturing a couple of safety conditions: first, the allocation regions
+% of objects referred by the instance variables should outlive the
+% allocation region of the instance itself, and second, the allocation
+% regions of a class type and its superclass type must be the same.
 
@@
 
-Note that if the original program in $\absof{\FB}$ contains unsafe
-references, for example, a reference from a transferable region to a
-stack regions, then the constraints generated during the elaboration
-are not satisfiable. In such case, $\solve$ fails to solve
-constraints, causing the program to be rejected.
+% Note that if the original program in $\absof{\FB}$ contains unsafe
+% references, for example, a reference from a transferable region to a
+% stack regions, then the constraints generated during the elaboration
+% are not satisfiable. In such case, $\solve$ fails to solve
+% constraints, causing the program to be rejected.
 
@@
 methods that were used in that type or expression. Each use of a
-(region-polymorphic) class or a method may instantiate region
-parameters differently, resulting in a different pending substitution
-($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-of predicate variables and pending substitutions. While $\phictxt$
-captures relationships that are \emph{known} to hold between concrete
-region identifiers (i.e., $\rgn$'s) when the constraint was generated,
+(region-polymorphic) class or a method has to instantiate region
+parameters, resulting in pending substitutions ($F_j$). Formulas
+$\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
+variables and pending substitutions. While $\phictxt$ captures
+relationships that are \emph{known} to hold between concrete region
+identifiers (i.e., $\rgn$'s) when the constraint was generated,
 $\phicstr$ captures relationships that are \emph{required} to hold

### 2015-11-11 paper/type-inference.tex
Edits to type inference.

@@
 and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
-respectively. It can be templatized as following (method definitions
-are left intact while templatizing, hence they are elided below):
+respectively. It can be templatized as following :
 \vspace*{-0.1in}
@@
   $a@\rho_0$ fst; $b@\rho_1$ snd;
-  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
-  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5}$() { $...$ }
+  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
+    super(); this.fst=fst; this.snd=snd; 
+  }
+  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
 }
@@
 deducible from $\rgn_1 \outlives \rgn_2$.
-
 % Validity constraints primarily result from checking if a region type
@@
 % region-polymorphic methods, where similar situation arises. 
-In general, validity constraints generated by our algorithm assume one
-of the following two forms:
-\begin{center}
-\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
-           {\phicstr} \qquad
-   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
-           {F_j(\varphi_j)}
-\)
-\end{center}
-Where $\varphi_i$'s denote the unknown preconditions of the class and
-the method under elaboration. If the constraint is generated while
-checking the well-formedness of a type or elaborating an expression,
-then $\varphi_j$'s denote the unknown preconditions of classes and
-methods that were used in that type or expression. Each use of a
-(region-polymorphic) class or a method has to instantiate region
-parameters, resulting in pending substitutions ($F_j$). Formulas
-$\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
-variables and pending substitutions. While $\phictxt$ captures
-relationships that are \emph{known} to hold between concrete region
-identifiers (i.e., $\rgn$'s) when the constraint was generated,
-$\phicstr$ captures relationships that are \emph{required} to hold
-among region varibles (i.e., $\rho$'s), or relationships between
-region variables and identifiers. Each region variable occuring in
-$\phicstr$ has an associated well-formedness constraint, which
-specifies its unification domain. The unification domain of a
-constraint is the union of unification domains of all region variables
-occuring in the constraint.
+% In general, validity constraints generated by our algorithm assume one
+% of the following two forms:
+% \begin{center}
+% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+%            {\phicstr} \qquad
+%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+%            {F_j(\varphi_j)}
+% \)
+% \end{center}
+% Where $\varphi_i$'s denote the unknown preconditions of the class and
+% the method under elaboration. If the constraint is generated while
+% checking the well-formedness of a type or elaborating an expression,
+% then $\varphi_j$'s denote the unknown preconditions of classes and
+% methods that were used in that type or expression. Each use of a
+% (region-polymorphic) class or a method has to instantiate region
+% parameters, resulting in pending substitutions ($F_j$). Formulas
+% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
+% variables and pending substitutions. While $\phictxt$ captures
+% relationships that are \emph{known} to hold between concrete region
+% identifiers (i.e., $\rgn$'s) when the constraint was generated,
+% $\phicstr$ captures relationships that are \emph{required} to hold
+% among region varibles (i.e., $\rho$'s), or relationships between
+% region variables and identifiers. Each region variable occuring in
+% $\phicstr$ has an associated well-formedness constraint, which
+% specifies its unification domain. The unification domain of a
+% constraint is the union of unification domains of all region variables
+% occuring in the constraint.
 
-\paragraph{Constraints Example 1} The original source of all
-outlives constraints is the validity constraint $C_3$ in $\elabClass$
-function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
-class template from \S\ref{sec:fb-templatization}, the constraint is
-as following:
-\begin{center}
-  \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
-  \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
-\end{center}
+\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
+template from \S\ref{sec:fb-templatization}. Following constraints are
+generated during its elaboration (Constrains are identified with
+$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
+\rhoenv_0$, where $\rhoenv_0 = \{\rhoalloc_0,\rho_{0-4}\}$, have been
+elided): 
+\begin{smathpar}
+\begin{array}{l}
+  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
+  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
+     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
+  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
+  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \spc
+  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
+\end{array}
+\end{smathpar}
 
-\paragraph{Constraints Example 2} Consider a contrived method
-$\C{alt}$ that accepts an argument $\C{p}$ of type
+\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
+contrived method $\C{alt}$ that accepts an argument $\C{p}$ of type
 $\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
@@
 Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
-alternated between recursive calls. The region-annotated definition of
-$\C{alt}$ is shown below:
+alternated between recursive calls. Elaboration phase elaborates
+$\C{alt}$ to the following region-annotated definition\footnote{The
+original definition of \C{alt} can be obtained by erasing all the
+region annotations from the elaborated version}\footnote{The
+elaboration actually uses new region variables as parameters to the
+recursive call, and then generates constraints that unify them with
+the ones shown. We coalesced both steps to simplify presentation.}:
 \begin{codejava}
-unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
-                          A<$\rho_3$> x, A<$\rho_4$> y) {
-  return p.fst = x; p.snd = y; 
-         this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
+unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
+                          A<$\rho_9$> x, A<$\rho_{10}$> y) {
+  p.fst = x; p.snd = y; 
+  this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
 }
 \end{codejava}
-Note that, to avoid clutter, we have already resolved appropriate
-region arguments to the recursive call, instead of introducing new
-region variables and generating equality constraints on them. Rest of
-the validity constraints generated while elaborating $\C{alt}$ to the
-above region-annotated definition are shown below:
-\begin{center}
-\(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
-$\quad$
-\(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
-$\quad$
-\(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
-\(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
-\end{center}
-The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
-Second and third are generated by the the assignment expressions, and
-the last constraint is generated from the recursive call.
+% Note that, to avoid clutter, we have already resolved appropriate
+% region arguments to the recursive call, instead of introducing new
+% region variables and generating equality constraints on them. 
+Constraints generated during the elaboration are shown below
+(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-10}\}$ ):
+\begin{smathpar}
+\begin{array}{l}
+\csid{6} \tywf{\rhoenv_2}{\varphi_2}\qquad
+\csid{7} \isvalid{\varphi_2}{\rho_7 \outlives \rho_6 \conj \rho_8
+  \outlives \rho_6} \\
+\csid{8} \isvalid{\varphi_2}{\rho_9 = \rho_7} \qquad
+\csid{9} \isvalid{\varphi_2}{\rho_{10} = \rho_8} \\
+\csid{10} \isvalid{\varphi_2}{[\rho_9/\rho_{10}][\rho_{10}/\rho_9]\varphi_2}
+\end{array}
+\end{smathpar}
+% The first constraint is generated by $\typeOk$ on the type of
+% $\C{p}$.  Second and third are generated by the the assignment
+% expressions, and the last constraint is generated from the recursive
+% call.
 

### 2015-11-12 paper/naiad-ex-discussion.tex
Discussion of motivating example in BROOM

--- 2015-10-16 paper/overview.tex
@@
-\section{Overview of \name}
-\label{sec:overview}
-
-Stripped of its region-specific constructs, the core language of \name
-is a familar subset of \csharp that includes features such as
-class-based inheritance, dynamic dispatch, parametric polymorphism,
-and lambdas.  We therefore elide a discussion of the core language,
-focusing instead on the concepts specific to region-based memory
-management in \name.
-
-\subsection{Stack Regions and Outlives Relation}
-
-Simplest of the region-specific constructs in \name is a \C{letregion}
-lexical block that creates a new memory region adressable via its
-static identifier. The semantics of the \C{letregion} block is similar
-to Tofte and Talpin~\cite{ttpopl94}'s \C{letregion} expression; the
-newly introduced region is live only inside the block, where its
-identifier is in scope for \C{new} expressions to make allocations
-inside the region. For example, in the following code, object \C{a} is
-allocated in the region with identifier \C{R0} introduced by the
-\C{letregion}:
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    Object a = new@R0 Object();
-  }
-\end{codejava}
-\end{center}
-The \C{letregion} blocks can be nested, leading to a stack of regions
-that are deallocated in the reverse order in which they are allocated.
-Following ~\cite{cyclonepldi02}, we therefore call regions introduced
-by \C{letregion} expressions as \emph{stack regions}. The stack
-discipline induces \emph{outlives} relationship among regions created
-by nested \C{letregion}s, where the region introduced by the outer
-\C{letregion} is guaranteed to outlive the one introduced by the inner
-\C{letregion}. It is therefore safe to refer to an object allocated in
-outer region from the inner region, but the converse is not true. For
-example, consider the following code (assume that class \C{A} has
-field \C{x} of type \C{Object}):
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    A a0 = new@R0 A();
-    letregion R1 {
-      A a1 = new@R1 A();
-      a1.x = new@R0 Object();
-      a0.x = new@R1 Object();
-      ...
-    }
-    ...
-  }
-\end{codejava}
-\end{center}
-The code creates two stack regions with identifiers \C{R0} and \C{R1},
-where the region \C{R0} outlives the region \C{R1} (denoted as $\C{R0}
-\outlives \C{R1}$).  Objects \C{a0} and \C{a1} are allocated in regions
-\C{R0} and \C{R1}, respectively. The first assignment statement
-assigns to \C{a1.x} an object allocated in outer region (\C{R0}). This
-assignment is safe as \C{a1.x} refers to a longer living object, hence
-is guaranteed to be a valid reference throughout the lifetime of
-\C{a1}.  In contrast, the second assignment is unsafe, as it assigns
-to \C{a0.x} an object, whose lifetime is shorter than the lifetime of
-\C{a0}, making it unsafe to dereference \C{a0.x} outside the inner
-block. 
-% Unsafe assignments can also happen indirectly via a function call..
-Preempting such unsafe assignments is the \emph{raison d'etre} of the
-region type system.
-
-\paragraph{A Note on region identifiers} Every stack region in \name
-is associated with a static identifier that uniquely identifies the
-region within its scope. At runtime, if a \C{letregion} expression is
-evaluated multiple times in a loop or a recursive method, the
-corresponding identifier is bound to a new stack region each time. 
-Any proposition involving static region identifiers is
-considered true at a program location if and only if the proposition
-is true under all possible evaluation contexts of that program
-location. For instance, consider the following example:
-%\begin{minipage}{1.55in}
-\begin{numcodejava}
-for (int i=0; i<=10; i++) {
-  letregion R0 {
-    letregion R1 {
-      A a1 = new@R1 A();
-      ...
-    }
-  }
-}
-\end{numcodejava}
-%\end{minipage}
-%\begin{minipage}{1.4in}
-%\begin{codejava}
-%void m(int i) {
-%  letregion R0 {
-%    A a = new@R0 A();
-%  }
-%  this.m(++i);
-%}
-%...
-%this.m(1);
-%\end{codejava}
-%\end{minipage}
-The identifiers \C{R0} and \C{R1} are bound to new stack regions each
-time the loop is evaulated. Nonetheless, the propositions (a). $\C{R0}
-\outlives \C{R1}$, and (b). $\C{a1}:\C{A@R1}$ (read as \emph{\C{a1}
-refers to an object of type \C{A} contained in region \C{R1}}) are
-true at line 5, as they are true under all possible bindings of \C{R0}
-and \C{R1} at line 5.
-
-% Interestingly, it is possible for multiple stack regions introduced by
-% evaluating a the same \C{letregion} expression to be in outlives
-% relationship. For instance, in the following example:
-% \begin{codejava}
-% void m(A a) {
-%   letregion R0 {
-%     B b = new@R0 B();
-%     b.x = a;
-%     this.m(new@R0 A());
-%   }
-% }
-% \end{codejava}
-% Stack regions identified by \C{R0} introduced in successive recursive
-% calls are in outlives relationship. Consequently, the assignment to
-% \C{b.x} should be considered safe. Region type system has to allow for
-% this possibility.
-
-\subsection{Allocation Context and Qualified Region Polymorphism}
-\label{sec:alloc-ctxt}
-
-As evident from the examples shown above, the suffix \C{@R} in
-\C{new@R} instructs the \C{new} keyword to allocate the newly created
-object in the region labeled \C{R}. However, \name does not require
-the allocation region to be specified following a \C{new}. In the
-absence of explicit specification, \name allocates the newly created
-object in its \emph{allocation context}, which is the region at the
-top of the stack when \C{new} is executed. For instance, the
-\C{findMin} function shown below, which makes explicit use of regions,
-calls the standard \C{List} library's \C{listIterator} method:
-\begin{center}
-\begin{codejava}
-  int findMin(List<int> l) {
-    assert(!l.isEmpty());
-    int min = l.get(0);
-    letregion R0 {
-      ListIterator<int> iter = l.listIterator();
-      while(iter.hasNext()) {
-        int n = iter.getNext();
-        if (n<min) min = n;
-      }
-    }
-    return min;
-  }
-\end{codejava}
-\end{center}
-The \C{listIterator} method neither creates new regions, nor does it
-annotate \C{new} expressions with region specifications. Consequently,
-all new objects created by the method are allocated in the allocation
-context for the call to \C{listIterator}, which in this case is
-\C{R0}.
-
-Observe that \C{listIterator} can be called under multiple different
-allocation contexts, and each time it returns a \C{ListIterator}
-allocated in its allocation context. The iterator object might hold
-references to the list, requiring the list to be allocated in a region
-that outlives \C{listIterator}\!'s allocation context. However, modulo
-this constraint, \C{listIterator} is not concerned about where the
-list is allocated. As such, \C{listIterator} is
-\emph{region-polymorphic} with respect to (a). its allocation context
-argument, and (b). the allocation region of the list, subject to the
-constraint that the later outlives the former. We call such region
-polymorphism with constraints in \name as \emph{qualified region
-polymorphism}. The provision to elide allocation region
-specifications, and the ability to infer qualified region-region
-polymorphic types are pivotal to interface region-oblivious standard
-library code with region-aware application code in \name.
-
-\subsection{Boxed and Unboxed Values}
-
-In \name, like in \csharp, values of some primitive types, such as
-integers and bools, are unboxed, whereas others, such as strings are
-arrays, are boxed. Unlike unboxed values, boxed values are objects,
-hence passed by reference. It is therefore important to be mindful
-of allocation regions for strings and arrays, but not so much for
-integers and bools. For instance, in the following code, the
-assignment to integer variable \C{x} is safe, but the assignment to
-the string variable \C{y} is unsafe, as the allocation region (\C{R0})
-of the string constant has lesser lifetime than \C{y}.
-\begin{center}
-\begin{codejava}
-  int x;
-  string y;
-  letregion R0 {
-    x = 2;
-    y = "Hello, World!";
-  }
-  ...
-\end{codejava}
-\end{center}
-Accordingly, \name's region type system lets the former assignment
-pass by, but flags the later as an error.
-
-\subsection{Dynamic (Transferable) Regions}
-
-\name's stack regions have lexically scoped lifetimes, hence are not
-suitable for applications that need to transfer data between
-autonomous entities, such as \naiad dataflow operators. To support
-these applications, it is imperative to support regions whose
-lifetimes transcend the boundaries of lexical blocks, functions, or
-even programs. Regions that fit this description in \name are called
-\emph{dynamic regions}, or \emph{transferable regions}.
-
-Transferable regions, like stack regions, are memory regions, but
-unlike stack regions, they are first class values of \name; they are
-considered objects of \C{Region} class, and, like objects of other
-classes, they are created using the \C{new} keyword, can be passed as
-arguments, stored in data structures, and returned from methods. Like
-a stack region, a transferable region facilitates arbitrary object
-allocations, but unlike a stack region, it has a well-defined
-\emph{root} object denoting the root of the data structure being
-transferred. In a typical dataflow computation, an upstream actor
-(e.g., a \C{SELECT} operator) constructs a transferable region, sets
-its root to the data structure containing intermediate results, and
-then transfers it to a downstream actor (e.g., a \C{COUNT} operator),
-which performs further processing. Since a transferable region escapes
-the lifetime of the sender, there must be no references from inside of
-the transferable region to objects allocated in other memory regions
-of the sender. Such references, if exist, may become invalid
-references in the context of recipient's address space, jeopardizing
-memory safety. \name relies on its region type system to prevent such
-unsafe references from being created.
-
-\subsection{Memory Safety}
-\label{sec:memory-safety}
-
-Memory safety in \name is predominantly due to its region type system.
-However, the type system does not ensure the integrity of a small
-subset of pointers, called the \emph{unsafe set}, for which runtime
-verification is needed to guarantee safety. This is more a design
-decision rather than an unfortunate inevitability. Rest of this
-subsection presents an overview of the region type system, discusses
-the unsafe set, and briefly describes the runtime mechanisms used to
-ensure safety of references in the unsafe set.
-
-At a fundamental level, the region type system's approach to memory
-safety involves approximating (a). \emph{liveness}: the set of memory
-regions that are live at any given program location, and (b).
-\emph{outlives} relationships between them, and using this information
-to preempt potentially unsafe references from being created. These
-include:
-\begin{itemize}
-\item References into dead/non-existing regions,
-\item References from longer-living stack regions to shorter-living ones, and 
-\item References that escape transferable regions.
-\end{itemize}
-Unsurprisingly, the type system's approximation of liveness and
-outlives information for stack regions is precise. However, for
-transferable regions, statically determining liveness is equivalent to
-determining if a C pointer is null in presence of dynamic memory
-allocation, which is undecidable in general. Therefore, the region
-type system has to be necessarily conservative in approximating the
-liveness of transferable regions in order to ensure full memory
-safety. Maintaining a good conservative approximation requires the
-type system to employ techniques such as linear types~\cite{wadlerlt}
-or unique pointers~\cite{rust}, which impose restrictions that are
-often too hard to program around~\footnote{Internet has many tutorials
-on how to \emph{fight} the borrow checker in Rust.}. This is
-particularly unacceptable in the context of \name, as dataflow
-programs are often run offline, and the concern for performance and
-developer productivity outweighs the concern for safety.
-
-\name's region type system therefore overapproximates the liveness
-information for transferable regions, i.e., it might assume that a
-transferable region is live when, in fact, it might have already been
-transferred or freed. This necessarily leads to cases where creation of
-unsafe references, and their subsquent dereferencing can go unnoticed
-by the type system. Considering that stack regions can contain
-pointers into transferable regions, the worst case cost of
-overapproximating liveness is therefore the integrity of \emph{every}
-pointer on stack.
-
-Fortunately, \name avoids this worst case scenario. It does so by
-restricting the unsafe set to only include $\RgnZ$ objects. In
-particular, pointers \emph{into} transferable regions are excluded
-from this set (hence safe). \name achieves this by imposing simple
-restrictions on how $\RgnZ$ objects can be used to operate on
-transferable regions. The following describes the $\RgnZ$ class, and
-how $\RgnZ$ objects should be used to make allocations into, and to
-read from transferable regions.
-
-\begin{figure}
-\begin{codejava}
-  public class Region<T> {
-    public Region(Func<void,T> mkRoot);
-    public void free();
-    public void transfer();
-  }
-\end{codejava}
-\caption{The type signature of the \C{Region} class}
-\label{fig:region-class}
-\end{figure}
-The type signature of \C{Region} class is described in
-Fig.~\ref{fig:region-class}. The class is parametric over the type of
-the root object (i.e., the \emph{root type}) of the transferable
-region it creates. The \C{Region} constructor expects a \C{Func}
-object - a higher-order function argument that returns an object of
-root type.  The idea is to execute this function with the newly
-created transferable region as its allocation context. The object thus
-created inside the transferable region will serve as its root object.
-For instance, the following code creates a transferable region
-(\C{rgn}) and sets its root to the string object returned by its
-\C{Func} argument (a lambda expression):
-\begin{codejava}
-  Region<string> rgn = new Region<string>
-                        (() => "Hello");
-\end{codejava}
-Among the methods of the \C{Region} class, \C{free} deallocates
-region's memory, while \C{transfer} transfers the region to a
-downstream actor as determined by the run-time. The precise semantics
-of \C{transfer} are unimportant in the context of the region type
-system; it suffices to understand \C{transfer} as an alias of
-\C{free}.
-
-A transferable region is said to be in \emph{closed} state when
-created. \name provides \C{openalloc} lexical construct to \emph{open}
-a transferable region as an allocation context, and to access its root
-object:
-\begin{codejava}
-  Region<string> rgn = new Region<string>
-                        (() => "Hello");
-  openalloc rgn withroot str {
-    print str; //prints "Hello"
-    str = str + " World!"; 
-  }
-\end{codejava}
-In the above code, the \C{openalloc} lexical block opens \C{rgn} for
-allocation, and sets the newly introduced local variable (\C{str}) to
-refer to its root object.  The assignment statement within the block
-updates the root. Since \C{rgn} is the allocation context within the
-block, the new string is allocated inside \C{rgn}, hence the
-assignment is safe. In some cases, it is desirable to open a
-transferable region only to read its contents, but not to perform any
-allocations. For this purpose, \name provides \C{open} lexical
-construct that, like \C{openalloc}, opens a transferable region and
-names its root object, but unlike \C{openalloc}, does not make it an
-allocation context. A transferable region can be \C{openalloc}'d or
-\C{open}'d many times, before it is eventually \C{free}'d or
-\C{transfer}'d.
-
-The design decisions reflected above, \emph{viz.}, to ascribe a root
-object to every transferable region, to make use of a higher-order
-argument to initialize the root object, and to require that a region
-be explictly opened in a lexical block for allocations, were made to
-reflect the message semantics of transferable regions, to promote
-uniform style of programming with different kinds of regions, and to
-precisely state and enforce the safety guarantees offered by \name's
-region type system (\S~\ref{sec:type-system}).
-
-\input{naiad-ex}
-
-\subsection{An Example}
-
 Fig~\ref{fig:naiad-ex} shows \C{SelectVertex} class, which is a

### 2015-11-12 paper/overview.tex
Discussion of motivating example in BROOM

@@
 
-\input{naiad-ex}
+\input{motivating-eg-in-broom}
+
+% \input{naiad-ex}
 
@@
 
-Fig~\ref{fig:naiad-ex} shows \C{SelectVertex} class, which is a
-straightforward implementation of the \C{SELECT} dataflow operator.
-\C{SelectVertex} implements \C{onReceive} method, which is called by
-the run-time when a message from an upstream actor is ready to be
-delivered. A message is usually a sequence of tuples (e.g., key-value
-pairs) wrapped inside a transferable region (\C{inRgn}).
-\C{SelectVertex} opens its input region \C{inRgn} (line 16), iterates
-through the sequence, filtering the input tuples that meet the
-selection criterion of the \C{whereFn} (line 19), and applying the
-\C{mapFn} to map the selected input tuples to output (line 22).
-Observe that the iterator is allocated in the temporary (stack) region
-\C{R0} that lives just long enough.  Functions \C{mapFn} and
-\C{whereFn} are defined by the users of the dataflow system, hence
-they are called \emph{user-defined functions}.  Unlike dataflow system
-builders, users are oblivious of regions, hence user-defined functions
-are expected to not contain explicit region annotations. While
-\C{whereFn} defines the selection criterion that the user is
-interested in (e.g: \C{age $\ge$ 18}), \C{mapFn} can either project
-attributes of its input tuple amounting to type \C{TOut}, or can
-create and return a new \C{TOut} object. Note that, unlike \C{inRgn},
-the output transferable region (\C{outRgn}) was opened for allocation
-and the \C{mapFn} is applied under this allocation context. Also note
-that the input tuple is cloned under this allocation context.
-Consequently, the result (\C{outp}) of applying the \C{mapFn} (line
-22), which is subsequently added to \C{outMsg}, is guaranteed to be in
-the \C{outRgn} regardless of the semantics of \C{mapFn}. On the other
-hand, if the input tuple is not cloned (i.e., \C{inp} instead of
-\C{inp.clone()} on line 22) and the \C{mapFn} simply projects its
-input, then \C{outp} refers to an object inside \C{inp}, which no
-longer exists when the \C{outRgn} is transferred to a downstream
-actor. Therefore, if \C{mapFn} is applied over \C{inp} instead of
-\C{inp.clone()}, then it better be a function that creates and returns
-a new \C{TOut} object. Under such scenario, the region type of
-\C{SelectVertex.mapFn} should be precise enough to prevent it from
-being assigned a function (line 6) that simply returns its input.
+Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
+of Fig.~\ref{fig:motivating-eg} can be written in \name.
+The \C{onReceive} method receives its input message in a \emph{transferred}
+region (\ie, a \emph{closed} region whose ownership is transferred to the
+recipient).
+Line 7 creates a new region to store the output for time \C{t},
+initializing it to contain an empty list.
+Line 9 opens the input region to process it.
+Line 10 creates a stack region \C{R0}.
+Thus, the temporary objects created by the iteration in line 11,
+for example, will be allocated in this stack region that lives just
+long enough.
+We open the desired output region in line 12, so that the new output
+objects created by the invocation of \C{userDefinedFunction} in line 13
+are allocated in the output region.
+Finally, the input region is freed in line 19.
+
+Note that this approach works correctly only if \C{userDefinedFunction}
+creates an output object that does not share any subobjects with its input
+object (which resides in the input region and will be freed at the end of
+the method).
+If desired, such subobjects must be cloned (to copy them from the input
+region to the output region).
 Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of
-\C{SelectVertex.mapFn}. Furthermore, the type can be automatically
-inferred by \name's region type inference
-(\S~\ref{sec:type-inference}), which can peform the above non-trivial
-reasoning on behalf of the programmer.
+capable of capturing such nuances in the type of \C{userDefinedFunction}
+and the type checker will ensure correctness.
+Furthermore, the type can be automatically inferred by \name's region type
+inference (\S~\ref{sec:type-inference}), which can peform the above
+non-trivial reasoning on behalf of the programmer.

### 2015-11-12 paper/overview.tex
overview

@@
                         (() => new List<String>());
-  open rgn as R0 withroot strList {
+  open rgn as  strList@R0 {
     strList.addAtHead("World");
@@
 with the data stored in a transferable region. Consider the following
-code, for instance\footnote{For brevity, we drop \C{as} part of the
-\C{open} expression whenever there is no need to assign a static
-identifier for the open region.}:
+code, for instance\footnote{For brevity, we drop the region identifier binding part of the
+\C{open} expression whenever the identifier is not used.}:
 \begin{codejava}
 void onReceive(Region<List<String>> rgn)
-  open rgn withroot strList {
+  open rgn as strList {
     letregion R1 {
@@
 
-% \input{naiad-ex}
-
 \subsection{An Example}
@@
 We open the desired output region in line 12, so that the new output
-objects created by the invocation of \C{userDefinedFunction} in line 13
+objects created by the invocation of \C{selector} in line 13
 are allocated in the output region.
@@
 
-Note that this approach works correctly only if \C{userDefinedFunction}
+Note that this approach works correctly only if \C{selector}
 creates an output object that does not share any subobjects with its input
@@
 Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of \C{userDefinedFunction}
+capable of capturing such nuances in the type of \C{selector}
 and the type checker will ensure correctness.

### 2015-11-12 paper/type-inference.tex
Type Inference Edits.

@@
 template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
-region variable. Likewise, given the region-annotated definition of
-\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
-method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
-is \footnote{In our exposition, we assume that classes $\C{A}$ and
-$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
-$\ObjZ$, they accept one region parameter - the allocation region of
-their objects.}\footnote{We abuse arrow notation to also represent
-types of methods, but unlike function types, there is no allocation
-region annotation atop the arrow in a method type.}
+region variable. FGJ type $\RgnZ\inang{T}$ is special; it is
+always templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
+region-annotated definition of \C{Pair} class from
+\S\ref{sec:fb-syntax} a region type template for a method with FGJ
+type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
+our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
+subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
+accept one region parameter - the allocation region of their
+objects.}\footnote{We abuse arrow notation to also represent types of
+methods, but unlike function types, there is no allocation region
+annotation atop the arrow in a method type.}
 $\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
 \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
-\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
-variables, and $\varphi_0$ is a fresh predicate variable denoting
-unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
-template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
-fresh. For a class, a template is computed in two steps. In the first
-step, we templatize the types of all its fields, constructor
+\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
+region variables, and $\varphi_0$ is a fresh predicate variable
+denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
+$\rho$ is fresh. For a class, a template is computed in two steps. In
+the first step, we templatize the types of all its fields, constructor
 arguments, and arguments and return values of all its methods, along
@@
 
-Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
+Function $\elabClassTable$ (not shown) elaborates
 every definition in the class table $CT$, while accumulating
-constraints. The constraints are finally solved by $\solve$
-(\S\ref{sec:fb-constraintsolving}), which returns substitution
-functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
-predicate variables, respectively, introduced during templatization
-and elaboration stages. The substitutions are applied to the class
-table (and to the artifacts that make up the class table, recursively)
-to compute a class table that maps classes to their fully
-region-annotated definitions in $\FB$.
+constraints. The constraints are finally solved for free region and
+predicate variables introduced during templatization and elaboration
+stages. The solutions are applied to the class table (and to the
+artifacts that make up the class table, recursively) to compute a
+class table that maps classes to their fully region-annotated
+definitions in $\FB$.
 
@@
 \subsection{Constraints}
-\label{sec:fb-constraintsem}
+\label{sec:constraints}
 
@@
 $\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
-\rhoenv_0$, where $\rhoenv_0 = \{\rhoalloc_0,\rho_{0-4}\}$, have been
-elided): 
+\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
+\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
+\{\rho_5\}$, have been elided): 
 \begin{smathpar}
@@
   \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
-  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \spc
-  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
+  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
+  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
+  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
 \end{array}
@@
 \paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
-contrived method $\C{alt}$ that accepts an argument $\C{p}$ of type
-$\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
-of type $A$. It then assigns one object to $\C{p.fst}$ and other to
-$\C{p.snd}$, and calls itself recursively without ever terminating.
-Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
-alternated between recursive calls. Elaboration phase elaborates
-$\C{alt}$ to the following region-annotated definition\footnote{The
-original definition of \C{alt} can be obtained by erasing all the
-region annotations from the elaborated version}\footnote{The
-elaboration actually uses new region variables as parameters to the
-recursive call, and then generates constraints that unify them with
-the ones shown. We coalesced both steps to simplify presentation.}:
+contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
+\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
+\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
+recursively with the same region, a new \C{Pair} object allocated in a
+local region, and an \C{A} object referred by the \C{snd} field of the
+pair inside the region. \C{alt} never terminates.  Elaboration phase
+elaborates the method to the following region-annotated
+definition\footnote{In reality, elaboration uses new region variables
+as parameters to the constructor and method calls, and then generates
+constraints that unify them with actuals. In our examples, to avoid
+clutter due to trivial constraints,we coalesced both steps and show
+the actuals instead.}(The original definition of \C{alt} can be
+obtained by erasing all the region annotations from the elaborated
+version):
+% \begin{codejava}
+% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
+%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
+%   p.fst = x; p.snd = y; 
+%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
+% }
+% \end{codejava}
 \begin{codejava}
-unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
-                          A<$\rho_9$> x, A<$\rho_{10}$> y) {
-  p.fst = x; p.snd = y; 
-  this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
+unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
+                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
+  q.fst := y; q.snd := y; 
+  open r as $\rgn_0$ withroot p in
+    letregion $\rgn_1$ in
+      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
+                      (p.fst,p.fst) in
+        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
 }
@@
 Constraints generated during the elaboration are shown below
-(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-10}\}$ ):
+(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
 \begin{smathpar}
 \begin{array}{l}
-\csid{6} \tywf{\rhoenv_2}{\varphi_2}\qquad
-\csid{7} \isvalid{\varphi_2}{\rho_7 \outlives \rho_6 \conj \rho_8
-  \outlives \rho_6} \\
-\csid{8} \isvalid{\varphi_2}{\rho_9 = \rho_7} \qquad
-\csid{9} \isvalid{\varphi_2}{\rho_{10} = \rho_8} \\
-\csid{10} \isvalid{\varphi_2}{[\rho_9/\rho_{10}][\rho_{10}/\rho_9]\varphi_2}
+\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
+\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
+    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
+\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
+\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
+\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
+\rgn_1} 
+    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
 \end{array}
@@
 Our constraint generation algorithm traverses the entire program,
-performing elaboration and collecting constraints, which are
-subsequently solved enmasse. The motivation behind the whole-program
-approach to constraint generation is twofold: it simplifies
-elaboration functions and makes presentation easier, and second, it
-naturally generalizes to mutual recursion. Nonetheless, we do not
-intend our type inference to be a whole-program analysis for (a). it
-preempts opportunities for separate compilation and dynamic linking,
-and (b). it is expensive and an overkill in most practical cases. We
-therefore reclaim the compositionality of type inference by solving
-the constraints in a compositional fashion. In more practical terms
-this means that our constraint solving algorithm visits and solves
-every constraint (or, every set of mutually dependent constraints)
-only once. It composes computed solutions to solve other constraints
-that depend on the solved constraints. Importantly, the failure to
-solve a dependent constraint does not result in backtracking. We now
-describe our compositional constraint solving algorithm in detail. To
-simplify the presentation of our algorithm, we assume that are no
-mutually recursive definitions in the source program. Recursive
-definitions are nonetheless allowed.
+performing elaboration and collecting constraints. However, we do not
+require region type inference to be a whole-program analysis.
+Compositionality of type inference is reclaimed by solving the
+constraints in a compositional fashion. Concretely, this means that
+our constraint solving algorithm visits and solves every constraint
+(or, every set of mutually dependent constraints) only once. It
+composes computed solutions to solve other constraints that depend on
+the solved constraints. Importantly, the failure to solve a dependent
+constraint does not result in backtracking. We now describe our
+constraint solving algorithm by demonstrating its peformance over the
+example constraints shown in the previous section. A formal account of
+the algorithm can be found in our technical report~\cite{techrep}.
+% To simplify the presentation of our algorithm, we assume that are no
+% mutually recursive definitions in the source program. Recursive
+% definitions are nonetheless allowed.
 
 \paragraph{Terminology} In a validity constraint, a predicate variable
-occuring on the left side of the turnstile is said to occur
-negatively, or with \emph{negative polarity}. In contrast, a predicate
-variable occuring on the right side is said to occur positively, or
-with \emph{positive polarity}. A validity constraint
-\emph{constrains} the set of predicate variables that occur
-negatively in the constraint, while it \emph{uses} the set of
-predicate variables that occur positively. A constraint is said to be
-\emph{recursive} if it constrains and uses a predicate variable.
-% A pair of validity
-% constraints are said to be \emph{mutually dependent} if there exists a
-% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
-% with opposite polarities in both the constraints. In such case, the
-% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
-% be mutually dependent. We call the transitive closure of mutual
-% dependency relation as \emph{transitive dependency} relation. Like
-% mutual dependency, transitive dependency is also extended to predicate
-% variables.
+occuring on the left side (resp. right side) of the turnstile is said
+to occur negatively (resp. positively). A validity constraint
+\emph{constrains} the set of predicate variables that occur negatively
+in the constraint, while it \emph{uses} the set of predicate variables
+that occur positively. A constraint is said to be \emph{recursive} if
+it constrains and uses a predicate variable.
+% A pair of validity constraints are said to be \emph{mutually
+% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
+% predicate variables that occur with opposite polarities in both the
+% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
+% predicate variables are also said to be mutually dependent. We call
+% the transitive closure of mutual dependency relation as
+% \emph{transitive dependency} relation. Like mutual dependency,
+% transitive dependency is also extended to predicate variables.
 
@@
 \begin{itemize}
-\item $c_1$ constrains a predicate variable that $c_2$ uses. 
-\item $c_1$ constrains a (non-strict) subset of predicate variables
-that $c_2$ constraint.
+\item \textbf{Type 1} $c_1$ constrains a predicate variable that $c_2$ uses. 
+\item \textbf{Type 2} $c_1$ constrains a (non-strict) subset of predicate variables
+that $c_2$ constrains.
 \end{itemize}
-The first condition intuitively corresponds to a case, where
-expression or type, whose region elaboration is constrained by $c_2$
-refers to a method or a class, whose unknown precondition is
-constrained by $c_1$. The common predicate variable ($\varphi$)
-represents the unknown precondition in this case. The dependency from
-$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
-before $c_2$ is solved, thus enforcing the rule that the precondition
-of a method must not depend on its calling context. 
+
+% The first condition intuitively corresponds to a case, where
+% expression or type, whose region elaboration is constrained by $c_2$
+% refers to a method or a class, whose unknown precondition is
+% constrained by $c_1$. The common predicate variable ($\varphi$)
+% represents the unknown precondition in this case. The dependency from
+% $c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
+% before $c_2$ is solved, thus enforcing the rule that the precondition
+% of a method must not depend on its calling context. 
 % Note that the first condition results in self-loops over recursive
@@
 % (and self-loops over recursive constraints). 
-The second condition captures two kinds of dependencies. First, the
-dependency from the constraints over a method precondition to the
-constraints over the precondition of the class containing the method.
-This captures our preference that the constraints over a class's
-region parameters should not depend on the idiosyncracies of its
-methods. Any additional constraints required by any of its methods
-must be captured in the precondition of the method itself
-(well-formedness rules allow this possibility). The second condition
-adds bidirectional dependencies between validity constraints that
-constraint the same set of predicate variables.
-
+% The second condition captures two kinds of dependencies. First, the
+% dependency from the constraints over a method precondition to the
+% constraints over the precondition of the class containing the method.
+% This captures our preference that the constraints over a class's
+% region parameters should not depend on the idiosyncracies of its
+% methods. Any additional constraints required by any of its methods
+% must be captured in the precondition of the method itself
+% (well-formedness rules allow this possibility). The second condition
+% adds bidirectional dependencies between validity constraints that
+% constraint the same set of predicate variables. 
 Next, we convert the dependency graph over constraints into a
@@
 represents a strongly connected component in the dependency graph.
+
+\paragraph{Example} The dependency DAG ($G_C$) over validity
+constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
+shown in Fig.~\ref{fig:pair-deps}. Each node (labeled $C_i$) is a set
+of constraints that belong to a strongly connected component in the
+dependency graph ($G_c$), hence are mutually dependent. All
+dependencies, except the self-dependency on $\mathbf{c_{10}}$, are
+type-2 dependencies.
+
+A dependency DAG makes the dependencies between constraints explicit.
 Constraints in each set are mutually dependent, and need to be solved
@@
 after solving a constraint set are applied to the constraints in
-subsequent sets before attempting to solve them. Consequently, when
-the turn of a constraint set ($C$) arrives during the constraint solving
-process, it satisfies certain properties:
-\begin{itemize}
-\item There exists only one predicate variable ($\varphi$) that is either
-constrained or used by the constraints in the set ($C$). The variable is
-called set's \emph{subject}. This property follows from (a). the fact
-that all the dependency constraints have already been solved (and
-solutions applied), and (b). the assumption that there are no mutually
-recursive definitions. 
-\item All the constraints that constrain the set's subject are present
-in the set. This follows from our definition of the dependency relation.
-\end{itemize}
+subsequent sets before attempting to solve them. For the DAG in
+Fig.~\ref{fig:pair-deps}, we consider the topological order $[C_1,
+C_2, C_3]$, and solve the sets of constraints in that order.
 
-Solving the set ($C$) of constraints entails finding an assignment for
-$C$'s subject ($\varphi$), and also all the region variables
-($\rhobar$) that occur free in $C$, such that the solution satisfies
-well-formedness constraints on $\varphi$ and $\rhobar$. To simplify
-presentation, we think of $C$ as being parameterized on
-$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-\rbrack$. We now formalize the constraint satisfaction problem, and
-its solution.
+% Consequently, when the turn of a constraint set ($C$) arrives during
+% the constraint solving process, it satisfies certain properties:
+% \begin{itemize}
+% \item There exists only one predicate variable ($\varphi$) that is either
+% constrained or used by the constraints in the set ($C$). The variable is
+% called set's \emph{subject}. This property follows from (a). the fact
+% that all the dependency constraints have already been solved (and
+% solutions applied), and (b). the assumption that there are no mutually
+% recursive definitions. 
+% \item All the constraints that constrain the set's subject are present
+% in the set. This follows from our definition of the dependency relation.
+% \end{itemize}
 
-\begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
-satisfaction problem is a tuple $\CSP$, where $C\lbrack
-\varphi,\rhobar \rbrack$ is a set of validity constraints, where
-$i$'th validity constraint assumes one of the following forms:
+Solving a set ($C$) of constraints involves finding assignments for
+free region ($\rho$) and predicate ($\varphi$) varibles in the set.
+Assignments to region variables can be found via unification (for the
+examples in \S~\ref{sec:constraints}, we have already performed the
+unification, so $C_{1-3}$ do not contain free region variables). Once
+the region variables have been unified, we solve $C$ to find an
+assignment for free predicate variables. The first step is to reduce
+the set of constraints into an equivalent single constraint by
+conjoining antecedents and consequents of all constraints in the
+set\footnote{\cite{techrep} formally justifies the soundness of this
+step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
+constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
+below:
+\begin{smathpar}
+\begin{array}{l}
+  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
+     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
+  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
+  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
+    {
+        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
+        \rho_7 = \rho_9 \conj 
+    }\\
+  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
+        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
+\end{array}
+\end{smathpar}
+The constraint $c_{12}$ is a non-recursive constraint of the form
+$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
+$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
+this case). Such constraints are solved for $\varphi$ by computing a
+maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
+\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
+well-formedness condition on $\varphi$, then it is clearly a solution
+for $\varphi$, because there cannot exist a weaker solution for
+$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
+maximally weak $\phisol$ fails the well-formedness condition, then we
+establish~\cite{techrep} that it is impossible to satisfy the
+constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
+that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
+and check if it satisfies the well-formedness condition on $\varphi$.
+The former problem can be cast as an instance of graph augmentation
+problem, as described below.
+\vspace*{-0.05in}
+\paragraph{Definitions} Given a constraint formula $\phi$, we define
+its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
+vertices ($V(\phi)$) are free region variables and identifiers in
+$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
+in\footnote{We alternate between viewing a constraint formula ($\phi$)
+as a set of constraints and a conjunction of constraints in this
+description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
+\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
+$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
+conjunction of symmetric outlives constraints for the purpose of graph
+encoding. Conversely, given a digraph $G$, we define its constraint
+encoding $\Phi(G)$ in a straightforward manner. We say that a graph
+$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
+\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
+exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
+exist a path between same pair of vertices in $G_1$.
+
+A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
+\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
+graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
+connected as $G(\phicstr)$. The problem of computing such a $G$ is
+equivalent to the problem finding of finding minimum number of edges
+to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
+Efficient algorithms to solve the later problem are known to
+exist~\cite{graph}, and we rely on them for the solution.
+
+Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
+graph augmentation problems, as described above, results in the
+following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
+outlives constraints are replaced with equalities):
 \begin{center}
 \(
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {\phicstr^{i}}\qquad
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {F_i(\varphi)}
+  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
+\)\\
+\(
+  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
 \)
 \end{center}
-$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
-union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
-unification domain ($\rhoenv$) of the CSP.  The solution to the
-constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
-$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
-$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
-constraint formula such that:
-\begin{itemize}
-\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
-$\tywf{\rhoenv_{\varphi}}{\phisol}$).
-\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
-is valid.
-% \begin{center}
-% \(
-%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
-%           {\phicstr^{i} \conj  F_i(\phisol))}
-% \)
-% \end{center}
-\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
-that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
-\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
-(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
-$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
-\end{itemize}
-\end{definition}
-
-\subsection{Solving the CSP}
-
-The first step of solving the CSP $\CSP$ is to cast it as an
-equivalent problem $\csp$ involving a single validity constraint. The
-single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
+Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
+$c_{14}$ gives us a recursive constraint of the following form:
 \begin{center}
 \(
-  \isvalid{ \Phicx \conj \varphi}
-        {\Phics \conj  \bigwedge_i F_i(\varphi)}
+  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
 \)
 \end{center}
-Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
-\phicstr^{i}$. To see why both CSPs are equivalent, consider two
-distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
-$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
-\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
-generated before $c_j$ by the constraint generation algorithm. Observe
-that when our constraint generation algorithm generates a constraint,
-all relationships between concrete region identifiers referred by the
-consequent of constraint are already present in its antecedent (this
-includes the unification domains of region variables referred by the
-consequent).  Since no new relationships between existing regions are
-added when a new constraints is generated, $\phictxt^j$ either
-describes the same relationships that are already described by
-$\phictxt^i$, or describes relationships among region identifiers that
-are new, and not relevant to $c_i$. Therefore, strengthening the
-context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
-nor strengthens $c_i$ (resp.  $c_j$).
-
-\subsubsection{Non-Recursive Constraints}
-\label{sec:csolve}
-
-We first describe how we solve a non-recursive CSP $\csp$ where
-$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
-\conj \varphi}{\Phics}$.
-
-Our approach is based on the observation that $\FB$ does not admit
-null values or uninitialized variables, thus forcing every region
-variable to be unified with some concrete region. The constraint
-formula $\Phics$ captures all such unification constraints on
-$\rhobar$. Since the set of all concrete region identifiers is the
-unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
-with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
-exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
-\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
-in which case the well-formedness constraint on $\rho_i$ is not
-satisfied, and constraint solving must fail.  Therefore, there exists
-a unique assignment $\substFn$ such that $c\lbrack
-\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
-$\phi_{sol}$.
-
-To obtain $\phisol$, we make use of another observation. Let $c\lbrack
-\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
+Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
+$F$ is a substitution function, not necessarily idempotent. Let
+$G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
+$\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
+recursive constraints  is based on some observations. Firstly, the set
+of all possible constraints over a finite set of region
+variables/identifiers is a complete meet semi-lattice, where (a).
+meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
+\conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
+\triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
+function $H$ is a monotone over the lattice since $\forall
+\phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
+\Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
+theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
+that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
+the recursive constraint shown above, we first compute the greatest
+fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
+following non-recursive constraint:
 \begin{center}
 \(
-  \isvalid{ \Phicx \conj \varphi}
-        {\Phics'}
+  \isvalid{\phictxt \conj \varphi}{\phi_f}
 \)
 \end{center}
-Consider a maximally weak formula $\phi$ such that
-$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
-% Since $\phi$ is
-% maximally weak, there does not exist a $\phi'$ satisfying all the
-% below three conditions:
-% \begin{itemize}
-% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
-% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
-% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
-% \end{itemize}
-Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
-\subseteq \rhoenv$, we have two cases:
-\begin{itemize}
+Which can then be solved via the graph-based approach described
+before.
 
-\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
-a solution to $\varphi$. 
-% Furthermore, it is the weakest solution, because it is equivalent to
-% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
-% \substFn(\rho_j))$, and any other solution has to stronger than
-% $\Phics$ under the same context.
-
-\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
-contains at least one equality or outlives constraint on two region
-identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
-\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
-\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
-constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
-is not maximally weak). Let us denote such constraint on $\rgn_i$ and
-$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
-$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
-Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
-$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
-$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
-\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
-this is impossible. To see why, recall that all the constraints on
-identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
-by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
-Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
-only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
-solution $\phisol$ cannot exist.
-\end{itemize}
-The above discussion hints at an algorithm to compute a solution to
-$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
-{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
-{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
-solution to $\varphi$.
-
-\subsubsection{Solving Recursive Constraints}
-
-We now extend constraint solving to recursive CSP $\csp$, where
-$c\lbrack \varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx
-\conj \varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake
-of brevity, we define $G(\varphi) = \Phicx \conj \bigwedge_i
-F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
+Let $H(\varphi_2)$ denote the consequent of the recursive constraint
+$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
 \begin{center}
 \(
-  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
+  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
+  \rgn_0 = \rgn_0
 \)
 \end{center}
-To solve the above recursive constraint, we start with the observation
-that the set of all constaints ($\phi$) over $\rhoenv \cup
-\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
-\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
-monotone over the lattice:
-\begin{center}
-  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
-  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
-\end{center}
-From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
-point ($\phi_f$), with following properties:
-\begin{itemize}
-\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
-$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
-\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
-G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
-{\cdot} {\phi_f' \Rightarrow \phi_f}$.
-\end{itemize}
-We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
-convert the recursive constraint to the following non-recursive
-constraint:
+Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
+\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
+results in the following solution for $\varphi_2$ that meets its
+well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
+\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
 \begin{center}
 \(
-  \isvalid{\Phicx \conj \varphi}{\phi_f}
+  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+  \rho_9 \conj \rho_8 = \rho_9
 \)
 \end{center}
-The technique described in \S~\ref{sec:csolve} now suffices to solve
-the above constraint.
-
-% Note that $F$ is a non-idempotent substitution function with following
-% properties:
-% \begin{itemize}
-% \item $F$ is a monotone: 
-% \begin{center}
-%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
-%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
-% \end{center}
-
-% \item $F$ distributes over conjunction (mind the syntactic equality):
-% \begin{center}
-%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
-%   F(\phi_2)$
-% \end{center}
-% \end{itemize}
-
-% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
-% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
-% $\phi_f$ be the GFP of $G$. Following properties ensue:
-% \begin{itemize}
-% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
-% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
-% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
-% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
-% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
-% \end{itemize}
-
-% \begin{theorem}
-% $\phi_f$ is a solution.
-% \end{theorem}
-% \begin{proof}
-% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
-% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
-% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
-% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
-% solution for $\varphi$.
-% \end{proof}
-
-% \begin{theorem}
-% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
-% \phisol}{\phi_f}$
-% \end{theorem}
-% \begin{proof}
-%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
-%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
-%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
-%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$
-
-% \end{proof}

### 2015-11-12 paper/type-inference.tex
Type Inference complete.

@@
 \end{center}
+
+\subsection{Higher-Order Type Inference}
+
+We have so far ignored function types in the context of type
+inference. The region type system for $\FB$ admits function types with
+higher-rank region polymorphism, allowing region variables to be
+quantified in the types of higher-order arguments. The type system
+clearly does not have the principal typing property, as demonstrated
+by the \C{app2} method below:
+\begin{codejava}
+bool app2(A x, A y, (A $\rightarrow$ bool) f) 
+    { return (f x && f y); }
+\end{codejava}
+There are two possibilities for the region type of \C{app2}: one,
+where it allows \C{x} and \C{y} to be allocated in different regions,
+but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
+arguments from any region), and second, where it requires \C{x} and
+\C{y} to be allocated in the same region, and allows \C{f} to be
+monomorphic. None of the two types are more general than the other,
+and it is not clear which one is more preferable. The concrete
+manifestation of this dilemma is a constraint of form 
+$\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
+no additional constriants to help determine $\varphi_1$ or $\varphi_2$
+before solving this constraint. Since it is always possible to
+strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
+does not exist unique solution to such constraints. Nonetheless, it is
+possible to bias the constraint-solving algorithm to consistently
+prefer one solution over the other.
+
+In context of \name, however, we adopt a trivial solution of
+assigning the same region to the arguments and return values in
+(higher-order) function types. For user-defined functions (e.g.,
+predicate in the \C{WHERE} clause of the \C{SELECT} dataflow query)
+and lambda arguments to the $\RgnZ$ class constructor, which are the
+prominent uses of higher-order functions in \name, the simplistic
+approach seems to suffice in practice. However, it is always possible
+that this approach makes type checker reject a program that is
+otherwise valid. In such cases, programmer can help the type checker
+by writing region annotations for only the higher-order functions.
+
+% \begin{center}
+%   \C{app2} : $\inang{\rho_0,\rho_1}$(\C{A}$\inang{\rho_0}$, 
+%               \C{A}$\inang{\rho_0}$,
+%               $\inang{\rho_2}$\C{A}$\inang{\rho_2}$ $\rightarrow$
+%               \C{bool}) $\rightarrow$ \C{bool}
+% \end{center}

### 2015-11-13 paper/overview2.tex
Updated overview

--- 2015-11-12 paper/overview.tex
@@
+\newcommand{\COMMENT}[1]{}
+
 \section{Overview of \name}
@@
 
-Stripped of its region-specific constructs, the core language of \name
-is a familar subset of \csharp that includes features such as
-class-based inheritance, dynamic dispatch, parametric polymorphism,
-and lambdas.  We therefore elide a discussion of the core language,
-focusing instead on the concepts specific to region-based memory
-management in \name.
+\name enriches a simple object-oriented language (supporting
+parametric polymorphism and lambdas) with a set of region-specific constructs. 
+In this section, we present an informal overview of these region-specific constructs.
 
+\subsection{Using Regions in \name}
+
+\paragraph{Stack Regions}
+The ``\C{letregion R S}'' construct creates a new stack region,
+with a static identifier \C{R}, whose scope is restricted to the
+statement \C{S}. The semantics of  \C{letregion} is similar
+to Tofte and Talpin~\cite{tofte94}'s \C{letregion} expression:
+new objects can be allocated in the newly created region within
+the scope of \C{S}, but the region and all objects allocated within it
+are freed at the end of \C{S}.
+
+\paragraph{Object Allocation}
+The ``\C{new@R T()}'' construct creates a new object of type \C{T} in
+the region \C{R}. The specification of the allocation region \C{R} in
+this construct is optional.  At runtime, \name maintains a stack of \emph{active}
+regions, and we refer to the region at the top of the stack as the \emph{allocation
+context}. The statement \C{new T()} allocates the newly created object in
+the current allocation context.
+%
+This is important as it enables \name applications to use existing region-oblivious
+C\# libraries. In particular, given a  C\# library function \C{F} (that makes no use of
+\name's region constructs), the statement ``\C{letregion R F()}'' invokes \C{F},
+but has the effect that all objects allocated by this invocation are allocated in the
+new region \C{R}.
+
+\paragraph{Transferable Regions}
+\name's \emph{transferable regions} are an encapsulation of a data-structure
+that can be transferred between autonomous entities
+(\eg, between two concurrently executing threads or actors).
+% Transferable regions are similar to stack regions in that they are memory regions
+% that support arbitrary object allocations, but they differ from stack regions in the
+% following ways.
+Hence, unlike stack regions, transferable regions are not constrained to have a
+lexically scoped lifetime.
+(Hence, we also refer to them as \emph{dynamic} regions.)
+% regions whose lifetimes transcend the boundaries of lexical blocks, functions, or
+% even programs.
+
+Furthermore, transferable regions, unlike stack regions, are first class values of \name:
+they are considered objects of \C{Region} class, and, like objects of other
+classes, they are created using the \C{new} keyword, can be passed as
+arguments, stored in data structures, and returned from methods.
+As mentioned above, a transferable region is intended to encapsulate a single
+data-structure, consisting of a collection of objects with a distinguished root
+object of some type \C{T}, which we refer to as the region's \emph{root} object.
+The class \C{Region} is parametric over the type \C{T} of this root object.
+
+The \C{Region} constructor expects a \C{Func}
+object - a higher-order function argument that returns an object of
+root type.  The idea is to execute this function with the newly
+created transferable region as its allocation context. The object thus
+created inside the transferable region will serve as its root object.
+This is illustrated by the following code that creates a transferable region
+(\C{r}) of type \C{T}:
+\begin{codejava}
+  Region<T> r = new Region<T>(() => new T())
+\end{codejava}
+
+\begin{figure}
+\begin{codejava}
+  public class Region<T> {
+    public Region(Func<void,T> mkRoot);
+    public void free();
+    public void transfer();
+  }
+\end{codejava}
+\caption{The type signature of the \C{Region} class}
+\label{fig:region-class}
+\end{figure}
+
+The type signature of the class \C{Region} appears in Fig.~\ref{fig:region-class}.
+The method \C{free} deallocates the region (including all objects allocated within it).
+The method \C{transfer} transfers the region to a downstream actor as determined by the run-time.
+It is an abstraction of two possible forms of  transfer: a transfer between two actors in a shared
+memory setting or a transfer between two actors in a distributed, message-passing, setting. 
+The precise semantics of \C{transfer} are unimportant in the context of the region type
+system and we will not discuss them further.
+
+\paragraph{Open and Closed Regions}
+A transferable region must be explicitly \emph{opened} using \name's \C{open} construct
+in order to either read or update or allocate objects in the region.
+Specifically, the construct ``\C{open rgn as v@R S}'' does the following:
+(a). It opens the transferable region handled by \C{rgn} for allocation
+(i.e., makes it the current allocation context),
+(b). binds the identifier \C{R} to this open region, and
+(c). initializes the newly introduced local variable \C{v} to refer
+to the root object of the region.
+The \C{@R} part of the statement is optional and may be omitted.
+The \C{open} construct is intended to simplify the problem of ensuring memory safety,
+as will be explained soon.
+We refer to a transferable region that has not been opened as a \emph{closed} region.
+
+\paragraph{Motivating Example}
+% We now illustrate how the features of \name can be used to 
+Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
+of Fig.~\ref{fig:motivating-eg} can be written in \name.
+The \C{onReceive} method receives its input message in a \emph{transferred}
+region (\ie, a \emph{closed} region whose ownership is transferred to the
+recipient).
+Line 7 creates a new region to store the output for time \C{t},
+initializing it to contain an empty list.
+Line 9 opens the input region to process it.
+Line 10 creates a stack region \C{R0}.
+Thus, the temporary objects created by the iteration in line 11,
+for example, will be allocated in this stack region that lives just
+long enough.
+We open the desired output region in line 12, so that the new output
+objects created by the invocation of \C{selector} in line 13
+are allocated in the output region.
+Finally, the input region is freed in line 19.
+
+\input{motivating-eg-in-broom}
+
+\subsection{Memory Safety}
+\label{sec:memory-safety}
+
+The key to memory safety in \name is the following restriction:
+an object $o_1$ in a region $R_1$ is allowed to store a pointer to
+an object $o_2$ in a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.
+(A similar restriction applies in the case where $o_1$ is a stack-allocated variable.)
+
+Enforcing this restriction is simple in the case of stack regions since the outlives relation
+between stack regions can be inferred from their lexical nesting. Unfortunately,
+inferring outlives relations between transferable regions is not easy.
+\name imposes the following protocol on the use of transferable regions to help simplify
+this check.
+
+A transferable region (that has not been freed or transferred) can be in one of two possible
+states, \emph{open} or \emph{closed}. A newly created region is in the closed state.
+A region must be opened, using the open construct (as explained previously), in order
+to read or update or allocate an object within that region.
+An open region cannot be freed or transferred. 
+In particular, an open region is guaranteed to be live for the entire duration of the open construct.
+This allows the type system to infer a valid outlives relation between the opened region
+and any stack region that is nested within the open construct.
+
+\begin{figure}
+\includegraphics[scale=0.45]{region-fsm.png}
+\caption{The lifetime of a dynamic (transferable) region in \name}
+\label{fig:region-fsm}
+\end{figure}
+
+The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
+A transferable region starts its lifetime in a \emph{closed} state,
+when it is created through \C{new Region} expresion. \name provides
+
+As evident in the above examples, the safety of memory accesses in
+\name is now subject to the condition that every transferable region
+correctly follows the state transition discipline shown
+in Fig.~\ref{fig:region-fsm}. If this is
+guaranteed, then \name's region type system statically guarantees the
+safety of all memory accesses. In other words, the type system reduces
+the problem of ensuring memory safety in \name programs to the problem
+of enforcing the state transition discipline for transferable regions.
+
+In \name, this enforcement is done at runtime by explicitly keeping
+track of the \emph{current state} for $\RgnZ$ objects, and
+checking the validity of every open, transfer, or free operation
+and throwing an exception if it is invalid.
+The challenge in enforcing this discipline statically is that transferable regions
+are first-class objects. Hence, the program can create multiple aliases for
+the same region, \eg, open it via one alias and free it via another.
+Typestate verification in the presence of aliases is hard.
+The checking can be done statically by preventing the creation of aliases using, \eg, linear types
+or unique types. However, this would be quite restrictive, in terms of expressiveness.
+
+\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
+in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
+as well as the runtime overhead of checking the region's state transition discipline is acceptable.
+
+\paragraph{Cloning}
+Note that the example in Fig.~\ref{fig:motivating-eg-in-broom} works correctly only if \C{selector}
+creates an output object that does not share any subobjects with its input object (which resides in
+the input region and will be freed at the end of the method).
+If there is a need for the output object to point to subobjects of the input object, such subobjects
+must be cloned (to copy them from the input region to the output region).
+Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
+capable of capturing such nuances in the type of \C{selector}
+and the type checker will ensure correctness.
+Furthermore, the type can be automatically inferred by \name's region type
+inference (\S~\ref{sec:type-inference}), which can peform the above
+non-trivial reasoning on behalf of the programmer.
+
+%\begin{codejava}
+%  Region<string> rgn = new Region<List<String>>
+%                        (() => new List<String>());
+%  open rgn as  strList@R0 {
+%    strList.addAtHead("World");
+%    strList.addAtHead("Hello");
+%  }
+%  rgn.transfer();
+%\end{codejava}
+
+\COMMENT{
 \subsection{Stack Regions and Outlives Relation}
 
-Simplest of the region-specific constructs in \name is a \C{letregion}
-lexical block that creates a new memory region adressable via its
-static identifier. The semantics of the \C{letregion} block is similar
-to Tofte and Talpin~\cite{ttpopl94}'s \C{letregion} expression; the
-newly introduced region is live only inside the block, where its
-identifier is in scope for \C{new} expressions to make allocations
-inside the region. For example, in the following code, object \C{a} is
-allocated in the region with identifier \C{R0} introduced by the
-\C{letregion}:
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    Object a = new@R0 Object();
-  }
-\end{codejava}
-\end{center}
 The \C{letregion} blocks can be nested, leading to a stack of regions
@@
 
-\paragraph{A Note on region identifiers} Every stack region in \name
-is associated with a static identifier that uniquely identifies the
-region within its scope. At runtime, if a \C{letregion} expression is
-evaluated multiple times in a loop or a recursive method, the
-corresponding identifier is bound to a new stack region each time. 
-Any proposition involving static region identifiers is
-considered true at a program location if and only if the proposition
-is true under all possible evaluation contexts of that program
-location. For instance, consider the following example:
-%\begin{minipage}{1.55in}
-\begin{numcodejava}
-for (int i=0; i<=10; i++) {
-  letregion R0 {
-    letregion R1 {
-      A a1 = new@R1 A();
-      ...
-    }
-  }
-}
-\end{numcodejava}
-%\end{minipage}
-%\begin{minipage}{1.4in}
-%\begin{codejava}
-%void m(int i) {
-%  letregion R0 {
-%    A a = new@R0 A();
-%  }
-%  this.m(++i);
-%}
-%...
-%this.m(1);
-%\end{codejava}
-%\end{minipage}
-The identifiers \C{R0} and \C{R1} are bound to new stack regions each
-time the loop is evaulated. Nonetheless, the propositions (a). $\C{R0}
-\outlives \C{R1}$, and (b). $\C{a1}:\C{A@R1}$ (read as \emph{\C{a1}
-refers to an object of type \C{A} contained in region \C{R1}}) are
-true at line 5, as they are true under all possible bindings of \C{R0}
-and \C{R1} at line 5.
-
-% Interestingly, it is possible for multiple stack regions introduced by
-% evaluating a the same \C{letregion} expression to be in outlives
-% relationship. For instance, in the following example:
-% \begin{codejava}
-% void m(A a) {
-%   letregion R0 {
-%     B b = new@R0 B();
-%     b.x = a;
-%     this.m(new@R0 A());
-%   }
-% }
-% \end{codejava}
-% Stack regions identified by \C{R0} introduced in successive recursive
-% calls are in outlives relationship. Consequently, the assignment to
-% \C{b.x} should be considered safe. Region type system has to allow for
-% this possibility.
-
 \subsection{Allocation Context and Qualified Region Polymorphism}
 \label{sec:alloc-ctxt}
-
-As the preceding examples illustrate, the suffix \C{@R} in
-\C{new@R} indicates that the newly created object should be allocated in the region labeled \C{R}.
-However, \name does not require the allocation region to be specified following a \C{new}.
-This is critical, as it enables \name applications to use existing region-oblivious C\# libraries.
-In the
-absence of explicit specification, \name allocates the newly created
-object in its \emph{allocation context}, which is the region at the
-top of the stack when \C{new} is executed. For instance, the
-\C{findMin} function shown below, which makes explicit use of regions,
-calls the standard \C{List} library's \C{listIterator} method:
-\begin{center}
-\begin{codejava}
-  int findMin(List<int> l) {
-    assert(!l.isEmpty());
-    int min = l.get(0);
-    letregion R0 {
-      ListIterator<int> iter = l.listIterator();
-      while(iter.hasNext()) {
-        int n = iter.getNext();
-        if (n<min) min = n;
-      }
-    }
-    return min;
-  }
-\end{codejava}
-\end{center}
-The \C{listIterator} method neither creates new regions, nor does it
-annotate \C{new} expressions with region specifications. Consequently,
-all new objects created by the method are allocated in the allocation
-context for the call to \C{listIterator}, which in this case is
-\C{R0}.
 
@@
 
-%\subsection{Boxed and Unboxed Values}
-%
-%In \name, like in \csharp, values of some primitive types, such as
-%integers and bools, are unboxed, whereas others, such as strings are
-%arrays, are boxed. Unlike unboxed values, boxed values are objects,
-%hence passed by reference. It is therefore important to be mindful
-%of allocation regions for strings and arrays, but not so much for
-%integers and bools. For instance, in the following code, the
-%assignment to integer variable \C{x} is safe, but the assignment to
-%the string variable \C{y} is unsafe, as the allocation region (\C{R0})
-%of the string constant has lesser lifetime than \C{y}.
-%\begin{center}
-%\begin{codejava}
-%  int x;
-%  string y;
-%  letregion R0 {
-%    x = 2;
-%    y = "Hello, World!";
-%  }
-%  ...
-%\end{codejava}
-%\end{center}
-%Accordingly, \name's region type system lets the former assignment
-%pass by, but flags the later as an error.
-
 \subsection{Dynamic (Transferable) Regions}
 
-\name's stack regions have lexically scoped lifetimes, hence are not
-suitable for applications that need to transfer data between
-autonomous entities, such as \naiad dataflow operators. To support
-these applications, it is imperative to support regions whose
-lifetimes transcend the boundaries of lexical blocks, functions, or
-even programs. Regions that fit this description in \name are called
-\emph{dynamic regions}, or \emph{transferable regions}.
-
-Transferable regions, like stack regions, are memory regions, but
-unlike stack regions, they are first class values of \name; they are
-considered objects of \C{Region} class, and, like objects of other
-classes, they are created using the \C{new} keyword, can be passed as
-arguments, stored in data structures, and returned from methods. Like
-a stack region, a transferable region facilitates arbitrary object
-allocations, but unlike a stack region, it has a well-defined
-\emph{root} object denoting the root of the data structure being
-transferred. In a typical dataflow computation, an upstream actor
+In a typical dataflow computation, an upstream actor
 (e.g., a \C{SELECT} operator) constructs a transferable region, sets
@@
 
-\begin{figure}
-\begin{codejava}
-  public class Region<T> {
-    public Region(Func<void,T> mkRoot);
-    public void free();
-    public void transfer();
-  }
-\end{codejava}
-\caption{The type signature of the \C{Region} class}
-\label{fig:region-class}
-\end{figure}
-The type signature of \C{Region} class is described in
-Fig.~\ref{fig:region-class}. The class is parametric over the type of
-the root object (i.e., the \emph{root type}) of the transferable
-region it creates. The \C{Region} constructor expects a \C{Func}
-object - a higher-order function argument that returns an object of
-root type.  The idea is to execute this function with the newly
-created transferable region as its allocation context. The object thus
-created inside the transferable region will serve as its root object.
-For instance, the following code creates a transferable region
-(\C{rgn}) and sets its root to the string object returned by its
-\C{Func} argument (a lambda expression):
-\begin{codejava}
-  Region<string> rgn = new Region<string>
-                        (() => "Hello");
-\end{codejava}
-Among the methods of the \C{Region} class, \C{free} deallocates
-region's memory, while \C{transfer} transfers the region to a
-downstream actor as determined by the run-time. The precise semantics
-of \C{transfer} are unimportant in the context of the region type
-system; it suffices to understand \C{transfer} as an alias of
-\C{free}.
-
-\name imposes a simple discipline on how $\RgnZ$ objects can be used
-to perform operations on transferable regions. This discipline is
-imposed in the interest of memory safety, which is discussed in the
-following subsection.
-
-\subsection{Memory Safety}
-\label{sec:memory-safety}
-
-The key to memory safety in \name is the following restriction:
-an object $o_1$ in a region $R_1$ is allowed to store a pointer to
-an object $o_2$ in a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.
-(A similar restriction applies in the case where $o_1$ is a stack-allocated variable.)
-
-Enforcing this restriction is simple in the case of stack regions since the outlives relation
-between stack regions can be inferred from their lexical nesting. Unfortunately,
-inferring outlives relations between transferable regions is not easy.
-\name imposes the following protocol on the use of transferable regions to help simplify
-this check.
-
-A transferable region (that has not been freed or transferred) can be in one of two possible
-states, \emph{open} or \emph{closed}. A newly created region is in the closed state.
-A region must be opened, using the open construct (explained in detail below), in order
-to read or update or allocate an object within that region.
-An open region cannot be freed or transferred. 
-In particular, an open region is guaranteed to be live for the entire duration of the open construct.
-This allows the type system to infer a valid outlives relation between the opened region
-and any stack region that is nested within the open construct.
-
-%Memory safety in \name is primarily due to its region type system,
-%although few runtime checks are needed.
-%% However, the type system does not ensure the integrity of a small
-%% subset of pointers, called the \emph{unsafe set}, for which runtime
-%% verification is needed to guarantee safety. This is more a design
-%% decision rather than an unfortunate inevitability, as the following
-%% discussion elaborates.
-%At a fundamental level, the region type system's approach to memory
-%safety involves approximating (a). the set of memory regions that are
-%live at any given program location\footnote{We say the type system
-%tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
-%relationships between them, and using this information to preempt
-%potentially unsafe references from being created. These include:
-%\begin{itemize}
-%\item References into dead/non-existing regions,
-%\item References from longer-living stack regions to shorter-living ones, and 
-%\item References that escape transferable regions.
-%\end{itemize}
-%Maintaining precise approximation of liveness and outlives information
-%for stack regions is easy. However, for dynamic (transferable) regions,
-%statically determining liveness is equivalent to determining if a C
-%pointer is null in presence of dynamic memory allocation, which is
-%undecidable in general. Systems that attempt to ensure full memory
-%safety in presence of dynamic memory (e.g., ~\cite{henglein},
-%~\cite{rust}) do so by conservatively approximating liveness
-%information for memory (regions). Maintaining a good conservative
-%approximation requires them to employ techniques such as linear
-%types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
-%restrictions that are often too hard to program
-%around\footnote{Internet has many tutorials on how to \emph{fight} the
-%borrow checker in Rust.}. This is particularly unacceptable in the
-%context of \name, as dataflow programs are often run offline, and the
-%concern for performance and developer productivity outweighs the
-%concern for safety. 
-%
-%\name's region type system therefore adopts the novel approach of
-%overapproximating liveness information for transferable regions, i.e.,
-%it might assume that a transferable region is live when, in fact, it
-%might have already been transferred or freed. This imprecision should
-%be a cause of concern as it can lead to cases where creation of unsafe
-%references, and their subsquent dereferencing go unnoticed by the type
-%system. Considering that stack regions can contain pointers into
-%transferable regions, the worst case cost of overapproximating
-%liveness is therefore the integrity of \emph{every} pointer on stack.
-
-%Fortunately, \name avoids this worst case scenario.  \name achieves
-%this by imposing a simple state transition discipline on the lifetime
-%of a transferable region.  The relevant finite state machine is shown
-%in Fig.~\ref{fig:region-fsm}.
-
-The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
-
-\begin{figure}
-\includegraphics[scale=0.45]{region-fsm.png}
-\caption{The lifetime of a dynamic (transferable) region in \name}
-\label{fig:region-fsm}
-\end{figure}
-
-A transferable region starts its lifetime in a \emph{closed} state,
-when it is created through \C{new Region} expresion. \name provides
-\C{open} lexical construct to \emph{open} a transferable region as an
-allocation context, and to access its root object:
-\begin{codejava}
-  Region<string> rgn = new Region<List<String>>
-                        (() => new List<String>());
-  open rgn as  strList@R0 {
-    strList.addAtHead("World");
-    strList.addAtHead("Hello");
-  }
-  rgn.transfer();
-\end{codejava}
-In the above code, the \C{open} lexical block performs the
-following operations: (a). It opens the transferable region handled by
-\C{rgn} for allocation (i.e., makes it the current allocation
-context), (b). assigns the identifier \C{R0} to the open region, and
-(c). sets the newly introduced local variable (\C{strList}) to refer
-to its root object (a list of strings). As Fig.~\ref{fig:region-fsm}
-indicates, it is not possible to open a region that is already
-transferred/freed. The fact that \C{rgn} is open within the block
-therefore guarantees that it is not yet transferred/freed, and that
-dereferencing \C{strList} within the block is safe. The end of
-\C{open} block marks the return of transferable region to the closed
-state.  While closed, the region is eligible to be transferred to a
-downstream actor, or to be freed.  An actor that receives the
-transferable region, receives it in the closed state. It can then
-reopen the received region to read its contents, possibly add more
-data and transfer it to another actor, or free the region.
+%% USING TEMPORARY STACK REGIONS AND ITS CORRECTNESS
 
@@
   rgn.free();
-}
 \end{codejava}
@@
 
-% The design decisions reflected above, \emph{viz.}, to ascribe a root
-% object to every transferable region, to make use of a higher-order
-% argument to initialize the root object, and to require that a region
-% be explictly opened in a lexical block for allocations, were made to
-% reflect the message semantics of transferable regions, to promote
-% uniform style of programming with different kinds of regions, and to
-% precisely state and enforce the safety guarantees offered by \name's
-% region type system (\S~\ref{sec:type-system}).
+As Fig.~\ref{fig:region-fsm}
+indicates, it is not possible to open a region that is already
+transferred/freed. The fact that \C{rgn} is open within the block
+therefore guarantees that it is not yet transferred/freed, and that
+dereferencing \C{strList} within the block is safe. The end of
+\C{open} block marks the return of transferable region to the closed
+state.  While closed, the region is eligible to be transferred to a
+downstream actor, or to be freed.  An actor that receives the
+transferable region, receives it in the closed state. It can then
+reopen the received region to read its contents, possibly add more
+data and transfer it to another actor, or free the region.
 
-As evident in the above examples, the safety of memory accesses in
-\name is now subject to the condition that every transferable region
-correctly follows the state transition discipline shown
-in Fig.~\ref{fig:region-fsm}. If this is
-guaranteed, then \name's region type system statically guarantees the
-safety of all memory accesses. In other words, the type system reduces
-the problem of ensuring memory safety in \name programs to the problem
-of enforcing the state transition discipline for transferable regions.
-
-In \name, this enforcement is done at runtime by explicitly keeping
-track of the \emph{current state} for $\RgnZ$ objects, and
-checking the validity of every open, transfer, or free operation
-and throwing an exception if it is invalid.
-The challenge in enforcing this discipline statically is that transferable regions
-are first-class objects. Hence, the program can create multiple aliases for
-the same region, \eg, open it via one alias and free it via another.
-Typestate verification in the presence of aliases is hard.
-The checking can be done statically by preventing the creation of aliases using, \eg, linear types
-or unique types. However, this would be quite restrictive, in terms of expressiveness.
-
-\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
-in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
-as well as the runtime overhead of checking the region's state transition discipline is acceptable.
-
-\input{motivating-eg-in-broom}
-
-\subsection{An Example}
-
-Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
-of Fig.~\ref{fig:motivating-eg} can be written in \name.
-The \C{onReceive} method receives its input message in a \emph{transferred}
-region (\ie, a \emph{closed} region whose ownership is transferred to the
-recipient).
-Line 7 creates a new region to store the output for time \C{t},
-initializing it to contain an empty list.
-Line 9 opens the input region to process it.
-Line 10 creates a stack region \C{R0}.
-Thus, the temporary objects created by the iteration in line 11,
-for example, will be allocated in this stack region that lives just
-long enough.
-We open the desired output region in line 12, so that the new output
-objects created by the invocation of \C{selector} in line 13
-are allocated in the output region.
-Finally, the input region is freed in line 19.
-
-Note that this approach works correctly only if \C{selector}
-creates an output object that does not share any subobjects with its input
-object (which resides in the input region and will be freed at the end of
-the method).
-If desired, such subobjects must be cloned (to copy them from the input
-region to the output region).
-Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of \C{selector}
-and the type checker will ensure correctness.
-Furthermore, the type can be automatically inferred by \name's region type
-inference (\S~\ref{sec:type-inference}), which can peform the above
-non-trivial reasoning on behalf of the programmer.
+}

### 2015-11-13 paper/abstract.tex
Updates to abstract

@@
 Recent work in high-performance systems written in managed languages (such as Java or C\#) has shown
-that garbage-collection can become a significant performance bottleneck in such systems.
-A class of these systems, which we refer to as dataflow programs, create huge data-structures with well-defined
-lifetimes that open up the possibility for significant performance improvements by using alternative memory
-management techniques that can exploit knowledge of these data-structures' lifetimes.
+that garbage-collection can be a significant performance bottleneck in such systems.
+A class of these systems, focused on big-data, create huge data-structures with well-defined lifetimes.
+In this paper, we present a language and memory management scheme based on user-managed
+memory regions (called \emph{transferable regions}) that allow programmers to exploit
+knowledge of data-structures' lifetimes to achieve significant performance improvements.
 
-In this paper, we present a memory management scheme that uses user-managed
-memory regions that allow programmers to exploit the specific memory behaviours exhibited by dataflow programs.
-To protect programmers from the dangers of manual memory management, we
-supplement our memory management scheme with a region type system that
-statically ensures the memory safety of region-annotated C\# programs,
-even in presence of higher-order stateful functions. Further, we
-describe a type inference algorithm that infers principal region types
-for first-order programs thereby saving programmers the need to write
+Manual memory management is susceptible to the usual perils of dangling pointers.
+A key contribution of this paper is a region type system that
+ensures the memory safety of region-annotated C\# programs,
+even in presence of higher-order stateful functions. We
+also present a type inference algorithm that infers principal region types
+for first-order programs, eliminating the need for programmers to write
 region annotations on types. Our inference algorithm is also very

### 2015-11-13 paper/introduction.tex
Discussion of motivating example in BROOM
updates to intro

@@
 \newcommand{\eg}{\emph{e.g.}}
+\newcommand{\ie}{\emph{i.e.}}
 
@@
 
-Consider the example shown in Fig.~\ref{fig:motivating-eg}.
+Consider the example, from~\cite{Broom:HotOS}, shown in Fig.~\ref{fig:motivating-eg}.
 This code represents the logic for a streaming query operator.
@@
 
-The goal of this work is an automatic (memory-safe) region-based memory management technique
+The goal of this work is a \emph{memory-safe} region-based memory management technique
 that supports dynamic regions (that are first-class objects) and higher-order functions.
-Though our focus is on memory-safety, dynamic regions also have other side-benefits.
+Though our focus, in this paper, is on memory-safety, our approach offers other side-benefits.
 When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
@@
 that the data-structure is self-contained and can be safely transmitted across address spaces.
+We refer to dynamic regions that provide these guarantees as \emph{transferable} regions.
 
@@
 
-The rest of the paper is organized as follows. The next section
-presents an informal overview of \name, and motivates the need for a
-region type system.  \S~\ref{sec:type-system} formalizes the type
-system and its safety guarantees. The type inference algorithm is
-described in \S~\ref{sec:type-inference}. \S~\ref{sec:csolve} focuses
-on \csolve, the constraint solving algorithm, and its correctness
-guarantees.  Implementations details and practical extensions to the
-type system are described in \S~\ref{sec:implementation}.
-\S~\ref{sec:evaluation} presents experimental evaluation and case
-studies.  \S~\ref{sec:related} discusses the related work, and
-\S~\ref{sec:conclusion} concludes.
+%The rest of the paper is organized as follows. The next section
+%presents an informal overview of \name, and motivates the need for a
+%region type system.  \S~\ref{sec:type-system} formalizes the type
+%system and its safety guarantees. The type inference algorithm is
+%described in \S~\ref{sec:type-inference}. \S~\ref{sec:csolve} focuses
+%on \csolve, the constraint solving algorithm, and its correctness
+%guarantees.  Implementations details and practical extensions to the
+%type system are described in \S~\ref{sec:implementation}.
+%\S~\ref{sec:evaluation} presents experimental evaluation and case
+%studies.  \S~\ref{sec:related} discusses the related work, and
+%\S~\ref{sec:conclusion} concludes.

### 2015-11-13 paper/overview2.tex
updates to motivating example
updated overview

@@
   public class Region<T> {
-    public Region(Func<void,T> mkRoot);
+    public Region(Func<T> mkRoot);
     public void free();
@@
 
+Our goal is a type system that can ensure the memory safety of programs that use the region
+constructs described above.
 The key to memory safety in \name is the following restriction:

### 2015-11-13 paper/broom.tex
Edits.
Edits to type inference.
Updated overview
small to footnotesize code. Fixed bugs in type rules.

@@
 
-\documentclass[pldi]{sigplanconf}
+\documentclass[10pt,pldi]{sigplanconf}
 
@@
 \newcommand{\isvalid}[2]{#1\,\vdash\,#2}
+\newcommand{\isnotvalid}[2]{#1\,\nvdash\,#2}
 \newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
@@
 \newcommand{\elabClassTable}{{\sf elabClassTable}}
+\newcommand{\CSP}{(C\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi}, \bar{\rhoenv})}
+\newcommand{\csp}{(c\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi},\bar{\rhoenv})}
+\newcommand{\Phicx}{\Phi_{cx}}
+\newcommand{\Phics}{\Phi_{cs}}
+\newcommand{\tynwf}[2]{#1\,\nvdash\,#2 \; \texttt{ok}}
+\newcommand{\csid}[1]{\lbrack \mathbf{c_{#1}} \rbrack:\;}
 
@@
 language=Java, % choose the language of the code
-basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
 keywordstyle=\color{Bittersweet},
@@
 language=[Objective]Caml, % choose the language of the code
-basicstyle=\small\ttfamily,       % the size of the fonts that are used for the code
+basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
 keywordstyle=\color{Bittersweet},
@@
 %---------------------
-\title{Broom: Safe Region-Based Memory Management for Dataflow
-Programs}
+\title{Safe Transferable Regions}
 
@@
 
-\input{overview}
+\input{overview2}
 

### 2015-11-16 paper/introduction.tex
updated intro

@@
 A region is a block of memory (consisting of one or more contiguous
-range of memory locations, permitting a region to grow) that is used to realize
-a data-structure with a well-defined lifetime. 
+range of memory locations, permitting a region to grow).
+% that is used to realize a data-structure with a well-defined lifetime. 
 Once a region is created, individual objects may be dynamically allocated within
-the region over time. Finally, the entire region is typically freed in one shot.
+the region over time. Finally, the entire region is freed in one shot.
 In this example, the output to be constructed for each time window $t$ can be
@@
 that supports dynamic regions (that are first-class objects) and higher-order functions.
-Though our focus, in this paper, is on memory-safety, our approach offers other side-benefits.
+Though we focus, in this paper, on memory-safety, our approach offers other benefits.
 When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
@@
 region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
-(In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
+In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
 This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
@@
 when a garbage collector promotes an object from a lower generation to a higher generation.
-From this perspective, a region may be seen as a user-controlled generation.)
+From this perspective, a region may be seen as a user-controlled generation.
 
@@
   destroy static and dynamic memory regions. Dynamic regions are
-  first-class values in \name; they manifest as objects of type
-  \C{Region}, and are treated on par with other objects. \name is
-  particularly suited for dataflow programs, such as~\naiad,
-  whose memory behavior exhibits
-  certain characteristics that admits efficient region-based memory management.
-  Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
-  memory regions outperform their GC counterparts by a margin of upto
-  59\%.
+  first-class values in \name.
+  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
+  % memory regions outperform their GC counterparts by a margin of upto
+  % 59\%.
 
@@
   guarantees safety of all memory accesses in a well-typed program,
-  subject to some conditions that can be checked efficiently at
-  run-time.
+  provided that the region usage correctly follows a simple discipline.
+  The correctness of region usage is checked efficiently at run-time.
   The overhead of checking these conditions is less than

### 2015-11-16 paper/overview2.tex
updated overview

@@
 (\eg, between two concurrently executing threads or actors).
-% Transferable regions are similar to stack regions in that they are memory regions
-% that support arbitrary object allocations, but they differ from stack regions in the
-% following ways.
 Hence, unlike stack regions, transferable regions are not constrained to have a
@@
 (Hence, we also refer to them as \emph{dynamic} regions.)
-% regions whose lifetimes transcend the boundaries of lexical blocks, functions, or
-% even programs.
 
 Furthermore, transferable regions, unlike stack regions, are first class values of \name:
-they are considered objects of \C{Region} class, and, like objects of other
-classes, they are created using the \C{new} keyword, can be passed as
-arguments, stored in data structures, and returned from methods.
-As mentioned above, a transferable region is intended to encapsulate a single
-data-structure, consisting of a collection of objects with a distinguished root
-object of some type \C{T}, which we refer to as the region's \emph{root} object.
+they are objects of the class \C{Region}, 
+they are created using the \C{new} keyword,
+and can be passed as arguments, stored in data structures, and returned from methods.
+A transferable region is intended to encapsulate a single data-structure,
+consisting of a collection of objects with a distinguished root object of some type \C{T},
+which we refer to as the region's \emph{root} object.
 The class \C{Region} is parametric over the type \C{T} of this root object.
 
-The \C{Region} constructor expects a \C{Func}
-object - a higher-order function argument that returns an object of
-root type.  The idea is to execute this function with the newly
-created transferable region as its allocation context. The object thus
-created inside the transferable region will serve as its root object.
-This is illustrated by the following code that creates a transferable region
-(\C{r}) of type \C{T}:
+The \C{Region} constructor takes as a parameter a function that constructs
+the root object: it creates a new region and invokes this function, with the new
+region as the allocation context, to create the root object of the region.
+This following code illustrates the creation of a transferable region:
+% (\C{r}) of type \C{T}:
 \begin{codejava}
@@
 The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
-A transferable region starts its lifetime in a \emph{closed} state,
-when it is created through \C{new Region} expresion. \name provides
-
-As evident in the above examples, the safety of memory accesses in
-\name is now subject to the condition that every transferable region
-correctly follows the state transition discipline shown
-in Fig.~\ref{fig:region-fsm}. If this is
-guaranteed, then \name's region type system statically guarantees the
-safety of all memory accesses. In other words, the type system reduces
-the problem of ensuring memory safety in \name programs to the problem
-of enforcing the state transition discipline for transferable regions.
+A transferable region starts its lifetime in a \emph{closed} state, when it is created.
+%
+The safety of memory accesses in \name is now subject to the condition that
+every transferable region correctly follows the state transition discipline shown
+in Fig.~\ref{fig:region-fsm}. If this is guaranteed, then \name's region type system
+statically guarantees the safety of all memory accesses.
+% In other words, the type system reduces
+% the problem of ensuring memory safety in \name programs to the problem
+% of enforcing the state transition discipline for transferable regions.
 

### 2015-11-16 paper/type-system.tex
Progress lemma.
Progress.
Preservation.
Preservation done.
Proofs proofs.
updates to section 3

--- 2015-11-03 paper/type-system.tex
@@
-\section{The Core Language}
+\section{\fbname}
 \label{sec:type-system}
 
-\input{fb-syntax}
+In this section, we formally develop \fbname, a language that incorporates
+the features introduced in the previous section, as well as its type system,
+building on the Featherweight Generic Java (FGJ)~\cite{fgj} formalism.
+Our development reuses notations and various definitions from~\cite{fgj},
+such as the definition of type well-formedness for the core (region-free) language,
+which are elaborated further in the supplement~\cite{techrep}.
 
@@
 
-We build on the Featherweight Generic Java (FGJ)~\cite{fgj} formalism
-to formalize \name and its region type system. Our development reuses
-notations from~\cite{fgj}, and relies on several of its definitions,
-such as the definition of type well-formedness for the core
-(region-free) language. We refrain from reproducing them here, and
-instead encourage the reader to refer to~\cite{fgj,techrep}.
+\input{fb-syntax}
 

### 2015-11-17 paper/abstract.tex
Edits to Sec.1

@@
-Recent work in high-performance systems written in managed languages (such as Java or C\#) has shown
-that garbage-collection can be a significant performance bottleneck in such systems.
-A class of these systems, focused on big-data, create huge data-structures with well-defined lifetimes.
-In this paper, we present a language and memory management scheme based on user-managed
-memory regions (called \emph{transferable regions}) that allow programmers to exploit
-knowledge of data-structures' lifetimes to achieve significant performance improvements.
+Recent work in high-performance systems written in managed languages
+(such as Java or C\#) has shown that garbage-collection can be a
+significant performance bottleneck in such systems.  A class of these
+systems, focused on big-data, create huge data-structures with
+well-defined lifetimes.  In this paper, we present a language and
+memory management scheme based on user-managed memory regions (called
+\emph{transferable regions}) that allow programmers to exploit
+knowledge of data-structures' lifetimes to achieve significant
+performance improvements.
 
-Manual memory management is susceptible to the usual perils of dangling pointers.
-A key contribution of this paper is a region type system that
-ensures the memory safety of region-annotated C\# programs,
-even in presence of higher-order stateful functions. We
-also present a type inference algorithm that infers principal region types
-for first-order programs, eliminating the need for programmers to write
-region annotations on types. Our inference algorithm is also very
-effective, albeit incomplete, for higher-order programs.
+Manual memory management is susceptible to the usual perils of
+dangling pointers.  A key contribution of this paper is a region type
+system that ensures the memory safety of C\# programs in the presence
+of transferable regions. We compliment our type system with a type
+inference algorithm that infers principal region types for first-order
+programs. This eliminates the need for programmers to write region
+annotations on types, while facilitating the reuse of existing C\#
+libraries with no modifications. Various case studies demonstrate the
+practical utility of our approach.
 

### 2015-11-17 paper/introduction.tex
Edits to Sec.1

@@
-\newcommand{\TODO}[1]{\textbf{TODO: #1}}
-\newcommand{\eg}{\emph{e.g.}}
+\newcommand{\TODO}[1]{\textbf{TODO: #1}} \newcommand{\eg}{\emph{e.g.}}
 \newcommand{\ie}{\emph{i.e.}}
 
-\section{Introduction}
-\label{sec:introduction}
+\section{Introduction} \label{sec:introduction}
 
-Consider the example, from~\cite{Broom:HotOS}, shown in Fig.~\ref{fig:motivating-eg}.
-This code represents the logic for a streaming query operator.
-The operator receives a stream of input messages, each associated
-with a time (window) $t$, processed by method \texttt{onReceive}.
-Each input message contains a list of inputs, each of which is processed
-by applying a user-defined function to create a corresponding output.
-The operator may receive multiple messages with the same timestamp
-(and messages with different timestamps may be delivered out of order).
-A timing-message (an invocation of method \texttt{OnNotify}) indicates that
-no more input messages with a timestamp $t$ will be subsequently delivered.
-At this point, the operator completes the processing for time window $t$ and sends
-a corresponding output message to its successor.
+Consider the example, from~\cite{Broom:HotOS}, shown in
+Fig.~\ref{fig:motivating-eg}.  This code represents the logic for a
+streaming query operator.  The operator receives a stream of input
+messages, each associated with a time (window) $t$, processed by
+method \texttt{onReceive}.  Each input message contains a list of
+inputs, each of which is processed by applying a user-defined function
+to create a corresponding output.  The operator may receive multiple
+messages with the same timestamp (and messages with different
+timestamps may be delivered out of order).  A timing-message (an
+invocation of method \texttt{OnNotify}) indicates that no more input
+messages with a timestamp $t$ will be subsequently delivered.  At this
+point, the operator completes the processing for time window $t$ and
+sends a corresponding output message to its successor.
 
@@
 data-structures are very large, as is the case with many streaming
-big-data analysis systems, the garbage collection overhead becomes
-significant.
-% as the collector scans these large data-structures
-However, these data-structures have a well-defined lifetime, which can
-be exploited to achieve better performance.
-% using alternative memory management techniques.
+big-data analysis systems, garbage collection overhead becomes
+significant. Importantly, much of this overhead results from the
+collector performing avoidable or unproductive
+work~\cite{Broom:HotOS}. For example, a consumer process may trigger a
+GC as soon as it receives a large data-structure from a producer
+process, which has just finished its GC. In such case, consumer's GC
+traverses the entire data-structure performing little or no work.
+Another example is the process executing the code in
+Fig.~\ref{fig:motivating-eg}, whose GC might repeatedly traverse the
+\C{map} data-structure, although its objects cannot be collected until
+a suitable timing message arrives.
+% as the collector scans these large data-structures For instance, in
+% the above example, the list at \C{map[t]} cannot be collected until
+% the the timing message arrives, and the list is transferred.
+% \C{transfer} operation is complete.  using alternative memory
+% management techniques.
 
-Region-based memory management is a good fit for such programs.
-A region is a block of memory (consisting of one or more contiguous
-range of memory locations, permitting a region to grow).
-% that is used to realize a data-structure with a well-defined lifetime. 
-Once a region is created, individual objects may be dynamically allocated within
-the region over time. Finally, the entire region is freed in one shot.
-In this example, the output to be constructed for each time window $t$ can be
-implemented using a separate region.
+An important observation, in the context of processes of kind
+described above, is that the data-structures exchanged between them
+have can be partitioned into sets of fate-sharing objects with common
+lifetimes, which makes them good candidates for a region-based memory
+management discipline. A region is a block of memory that is allocated
+and freed in one shot, consuming constant time. A region may contain
+one or more contiguous range of memory locations, and individual
+objects may be dynamically allocated within the region over time,
+while they are deallocated en masse when the region is freed.  Thus, a
+region is a good fit for a set of fate-sharing objects.
+% that is used to realize a data-structure with a well-defined
+% lifetime. 
+In this example, the output to be constructed for each time window $t$
+can be implemented using a separate region.
 
-Region-based memory management, both manual as well as automatic, has been known for
-a long time. Manual region-based memory management suffers from the usual challenges,
-namely the potential for dangling references and lack of memory safety. Automatic
-region-based memory management systems guarantee memory safety, but impose various restrictions.
-Tofte and Talpin's system~\cite{tofte94,tofte97} , for example, uses lexically scoped regions.
-At runtime, the set of all regions (existing at a point in time) forms a stack. Thus, the lifetimes of
-all regions must be well-nested: it is not possible to have two regions whose lifetimes
-overlap, with neither one's lifetime contained within the other.
-Unfortunately, the data-structures in the above example do not satisfy this restriction
-(as the output messages for multiple time windows may be simultaneously live, without
-any containment relation between their lifetimes).
-We refer to regions with lexically scoped lifetimes as \emph{stack regions} and to regions 
-that do not have such a lexically scoped region as \emph{dynamic} regions.
+Region-based memory management, both manual as well as automatic, has
+been known for a long time. Manual region-based memory management
+suffers from the usual challenges, namely the potential for dangling
+references and lack of memory safety. Automatic region-based memory
+management systems guarantee memory safety, but impose various
+restrictions.  Tofte and Talpin's system~\cite{tofte94,tofte97} , for
+example, uses lexically scoped regions.  At runtime, the set of all
+regions (existing at a point in time) forms a stack. Thus, the
+lifetimes of all regions must be well-nested: it is not possible to
+have two regions whose lifetimes overlap, with neither one's lifetime
+contained within the other.  Unfortunately, the data-structures in the
+above example do not satisfy this restriction (as the output messages
+for multiple time windows may be simultaneously live, without any
+containment relation between their lifetimes).  We refer to regions
+with lexically scoped lifetimes as \emph{stack regions} and to regions
+that do not have such a lexically scoped region as \emph{dynamic}
+regions.
 
-The goal of this work is a \emph{memory-safe} region-based memory management technique
-that supports dynamic regions (that are first-class objects) and higher-order functions.
-Though we focus, in this paper, on memory-safety, our approach offers other benefits.
-When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
-provides an ownership discipline that ensures that the producer cannot subsequently modify
-the data-structure. When the same transfer happens in a distributed setting, the system guarantees
-that the data-structure is self-contained and can be safely transmitted across address spaces.
-We refer to dynamic regions that provide these guarantees as \emph{transferable} regions.
+The goal of this work is a \emph{memory-safe} region-based memory
+management technique that supports dynamic regions as first-class
+objects. Though we focus, in this paper, on memory-safety, our
+approach offers other benefits.  When a producer transfers a
+data-structure to a consumer, in a shared-memory context , our
+approach provides an ownership discipline that ensures that the
+producer cannot subsequently modify the data-structure. When the same
+transfer happens in a distributed setting, the system guarantees that
+the data-structure is self-contained and can be safely transmitted
+across address spaces.  We refer to dynamic regions that provide these
+guarantees as \emph{transferable} regions.
 
-The key memory safety property we wish to ensure is that there are no dangling references:
-i.e., a reference to an object that has been freed.
-A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
-region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
-In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
-This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
-Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
-when a garbage collector promotes an object from a lower generation to a higher generation.
-From this perspective, a region may be seen as a user-controlled generation.
+The key memory safety property we wish to ensure is that there are no
+dangling references: i.e., a reference to an object that has been
+freed.  A key invariant and restriction that we utilize to achieve
+this is that no object $o_1$ in one dynamic region $D_1$ can contain a
+pointer to an object $o_2$ in another dynamic region $D_2$.  In cases
+where this is too restrictive, the user has to resort to (deep)
+copying $o_2$ to $D_1$.  This is conceptually quite reasonable given
+the perspective that $D_1$ is a self-contained data-structure.
+Furthermore, this is quite analogous, from a performance perspective,
+to the copying that happens when a garbage collector promotes an
+object from a lower generation to a higher generation.  From this
+perspective, a region may be seen as a user-controlled generation.
 
-This ensures that we can safely free one dynamic region (\eg, the input-message in our example)
-without affecting the validity of pointers in another dynamic region (\eg, the output-message).
-However, this is not sufficient! When processing the input-message $D$, we necessarily will have to create
-pointers that point to objects that are internal to $D$. Such pointers may be stack-allocated or even reside
-in the heap (\eg, consider the iterator object used to iterate over the list in the input-message).
-How do we ensure the safety of these pointers?
+This ensures that we can safely free one dynamic region (\eg, the
+input-message in our example) without affecting the validity of
+pointers in another dynamic region (\eg, the output-message).
+However, this is not sufficient! When processing the input-message
+$D$, we necessarily will have to create pointers that point to objects
+that are internal to $D$. Such pointers may be stack-allocated or even
+reside in the heap (\eg, consider the iterator object used to iterate
+over the list in the input-message).  How do we ensure the safety of
+these pointers?
 
-We use \emph{stack regions} (regions with lexically-scoped lifetimes), as well as the stack,
-for such temporary pointers (to objects inside a dynamic region) and use the following protocol
-to ensure safety of such pointers.
-To work with a dynamic region $D$, we must first \emph{open} region $D$,
-to indicate that $D$ should not be freed during this period.
-When the processing of the region is complete, we close the region $D$,
-to indicate that it is safe to free $D$.
-In general, a dynamic region may be opened and closed multiple times before it is finally freed.
+We use \emph{stack regions} (regions with lexically-scoped lifetimes),
+as well as the stack, for such temporary pointers (to objects inside a
+dynamic region) and use the following protocol to ensure safety of
+such pointers.  To work with a dynamic region $D$, we must first
+\emph{open} region $D$, to indicate that $D$ should not be freed
+during this period.  When the processing of the region is complete, we
+\emph{close} the region $D$, to indicate that it is safe to free $D$.
+In general, a dynamic region may be opened and closed multiple times
+before it is finally freed.
 
-Our type-system permits the creation of pointers to internal objects of \emph{open dynamic} regions,
-but uses lexical scoping to ensure that such pointers are no longer live when the region is closed.
-Thus, such pointers may be either stack-allocated or reside in a region whose lifetime is statically
-guaranteed to be contained within the interval when $D$ is open.
-As a consequence, the system ensures the strong invariant that for a \emph{closed dynamic} region $D$
-no pointer from outside $D$ points to an internal object of $D$.
+Our type-system permits the creation of pointers to internal objects
+of \emph{open dynamic} regions, but uses lexical scoping to ensure
+that such pointers are no longer live when the region is closed.
+Thus, such pointers may be either stack-allocated or reside in a
+region whose lifetime is statically guaranteed to be contained within
+the interval when $D$ is open.  As a consequence, the system ensures
+the strong invariant that for a \emph{closed dynamic} region $D$ no
+pointer from outside $D$ points to an internal object of $D$.
 
-The final piece required to ensure memory-safety is ensuring that the program correctly follows the above-mentioned
-protocol for dynamic regions: \eg, an open region should not be freed and, dually, a freed region should not be opened.
-(A complete typestate specification of the protocol is presented later.)
+The final piece required to ensure memory-safety is ensuring that the
+program correctly follows the above-mentioned protocol for dynamic
+regions: \eg, an open region should not be freed and, dually, a freed
+region should not be opened.  (A complete typestate specification of
+the protocol is presented later.)
 
-Dynamic regions are first-class objects in our system: \eg, in our running example,
-we would like to use a dictionary that stores dynamic regions. However, this means that the program may create multiple
-aliases (pointers) to the same region, which, in turn, means that statically checking if the program correctly follows the
-dynamic region protocol is hard (undecidable, in fact). Safety can be ensured by using, \eg, linear typing to prevent the
-creation of aliases, but this would be quite restrictive. Hence, in our system, we dynamically check for adherence to
-this protocol.
+Dynamic regions are first-class objects in our system. For instance,
+in our running example, we would like to use a dictionary that stores
+dynamic regions. However, this means that the program may create
+multiple aliases (pointers) to the same region, which, in turn, means
+that statically checking if the program correctly follows the dynamic
+region protocol is hard (undecidable, in fact). Safety can be ensured
+by using, \eg, linear typing to prevent the creation of aliases, but
+this would be quite restrictive. Hence, in our system, we dynamically
+check for adherence to this protocol.
 
 In summary, our approach reduces memory safety to two sub-problems:
-(a) Ensuring  low-level invariants about pointers to (fine-grained) objects inside regions and 
-(b) Ensuring higher-level (typestate) invariants about the regions themselves. 
-We use a type system to statically ensure the first set of invariants. We use a dynamic check to ensure that the
-second set of invariants are satisfied at runtime.
+(a) Ensuring  low-level invariants about pointers to (fine-grained)
+objects inside regions and (b) Ensuring higher-level (typestate)
+invariants about the regions themselves.  We use a type system to
+statically ensure the first set of invariants. We use a dynamic check
+to ensure that the second set of invariants are satisfied at runtime.
 %
-This is a novel aspect of our system. We believe that this is a reasonable choice:
-dynamically checking for the second category of invariants is easy and does not pose a performance concern (unlike, \eg, the first category of constraints);
-a static typechecker that rules out the second category of violations would be very restrictive from an expressiveness perspective;
-finally, the human effort required to ensure the second set of invariants is not as burdensome (as regions are coarse-grained objects
-manipulated infrequently), \eg, as it would be for the first set of invariants.
+This is a novel aspect of our system. We believe that this is a
+reasonable choice: dynamically checking for the second category of
+invariants is easy and does not pose a performance concern (unlike,
+\eg, the first category of constraints); a static typechecker that
+rules out the second category of violations would be very restrictive
+from an expressiveness perspective; finally, the human effort required
+to ensure the second set of invariants is not as burdensome (as
+regions are coarse-grained objects manipulated infrequently), \eg, as
+it would be for the first set of invariants.
 
@@
 
-\begin{itemize}
-  \item We present \name, a \csharp-like typed object-oriented language
-  that eschews garbage collection in favour of programmer-managed
-  memory regions . \name extends its core language, which includes
-  \emph{lambdas} (higher-order functions) and \emph{generics}
-  (parametric polymorphism), with constructs to create, manage and
-  destroy static and dynamic memory regions. Dynamic regions are
-  first-class values in \name.
-  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
-  % memory regions outperform their GC counterparts by a margin of upto
-  % 59\%.
+\begin{itemize} \item We present \name, a \csharp-like typed
+object-oriented language that eschews garbage collection in favour of
+programmer-managed memory regions . \name extends its core language,
+which includes \emph{lambdas} (higher-order functions) and
+\emph{generics} (parametric polymorphism), with constructs to create,
+manage and destroy static and dynamic memory regions. Dynamic regions
+are first-class values in \name.
+  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
+  % program using programmer-managed memory regions outperform their
+  % GC counterparts by a margin of upto 59\%.
 
@@
   guarantees safety of all memory accesses in a well-typed program,
-  provided that the region usage correctly follows a simple discipline.
-  The correctness of region usage is checked efficiently at run-time.
-  The overhead of checking these conditions is less than
-  \GK{x\%} in our experiments with \naiad workloads.
+  provided that certain typestate invariants on regions hold.  The
+  later invariants are enforced via simple runtime checks.
+% The correctness of region usage is checked efficiently at run-time.
+% The overhead of checking these conditions is less than \GK{x\%} in
+% our experiments with \naiad workloads.
 
   \item We define an operational semantics for \name, and a type
-  safety result that formalizes and proves safety guarantees described
-  above.
+  safety result that clearly defines and proves safety guarantees
+  described above.
 
@@
   region-aware \name programs and legacy standard library code that is
-  region-oblivious. The cornerstone of our inference algorithm is
-  \csolve, a novel constraint solving algorithm that performs
-  abduction and deduction in a Herbrand constraint system to solve the
-  constraints generated by the type inference algorithm.
-
-  \item We prove that region type inference is sound with respect to
-  its type system, and also complete for the set of first-order
-  programs (i.e., programs without lambdas).
+  region-oblivious. The cornerstone of our inference algorithm is a
+  novel constraint solver that performs abduction in a partial-order
+  constraint domain to infer weakest solutions to recursive
+  constraints.
 

### 2015-11-17 paper/overview2.tex
updates to section 3
Edits to Sec.1

@@
 
-\section{Overview of \name}
+\section{An Informal Overview of \name}
 \label{sec:overview}
@@
 \paragraph{Stack Regions}
-The ``\C{letregion R S}'' construct creates a new stack region,
+The ``\C{letregion R \{ S \}}'' construct creates a new stack region,
 with a static identifier \C{R}, whose scope is restricted to the
@@
 %
-This is important as it enables \name applications to use existing region-oblivious
-C\# libraries. In particular, given a  C\# library function \C{F} (that makes no use of
-\name's region constructs), the statement ``\C{letregion R F()}'' invokes \C{F},
-but has the effect that all objects allocated by this invocation are allocated in the
-new region \C{R}.
+This is important as it enables \name applications to use existing
+region-oblivious C\# libraries. In particular, given a  C\# library
+function \C{f} (that makes no use of \name's region constructs), the
+statement ``\C{letregion R \{ f(); \}}'' invokes \C{f}, but has the
+effect that all objects allocated by this invocation are allocated in
+the new region \C{R}.
 

### 2015-11-17 paper/overview2.tex
Edits to Sec.2.

@@
 
-\section{An Informal Overview of \name}
-\label{sec:overview}
+\section{An Informal Overview of \name} \label{sec:overview}
 
 \name enriches a simple object-oriented language (supporting
-parametric polymorphism and lambdas) with a set of region-specific constructs. 
-In this section, we present an informal overview of these region-specific constructs.
+parametric polymorphism and lambdas) with a set of region-specific
+constructs.  In this section, we present an informal overview of these
+region-specific constructs.
 
@@
 
-\paragraph{Stack Regions}
-The ``\C{letregion R \{ S \}}'' construct creates a new stack region,
-with a static identifier \C{R}, whose scope is restricted to the
-statement \C{S}. The semantics of  \C{letregion} is similar
-to Tofte and Talpin~\cite{tofte94}'s \C{letregion} expression:
-new objects can be allocated in the newly created region within
-the scope of \C{S}, but the region and all objects allocated within it
-are freed at the end of \C{S}.
+\paragraph{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
+creates a new stack region, with a static identifier \C{R}, whose
+scope is restricted to the statement \C{S}. The semantics of
+\C{letregion} is similar to Tofte and Talpin~\cite{tofte94}'s
+\C{letregion} expression: objects can be allocated by \C{S} in the
+newly created region while \C{R} is in scope, but the region and all
+objects allocated within it are freed at the end of \C{S}.
 
-\paragraph{Object Allocation}
-The ``\C{new@R T()}'' construct creates a new object of type \C{T} in
-the region \C{R}. The specification of the allocation region \C{R} in
-this construct is optional.  At runtime, \name maintains a stack of \emph{active}
-regions, and we refer to the region at the top of the stack as the \emph{allocation
-context}. The statement \C{new T()} allocates the newly created object in
-the current allocation context.
+\paragraph{Object Allocation} The ``\C{new@R T()}'' construct creates
+a new object of type \C{T} in the region \C{R}. The specification of
+the allocation region \C{R} in this construct is optional.  At
+runtime, \name maintains a stack of \emph{active} regions, and we
+refer to the region at the top of the stack as the \emph{allocation
+context}. The statement \C{new T()} allocates the newly created object
+in the current allocation context.
 %
@@
 
-\paragraph{Transferable Regions}
-\name's \emph{transferable regions} are an encapsulation of a data-structure
-that can be transferred between autonomous entities
-(\eg, between two concurrently executing threads or actors).
-Hence, unlike stack regions, transferable regions are not constrained to have a
-lexically scoped lifetime.
-(Hence, we also refer to them as \emph{dynamic} regions.)
+\paragraph{Transferable Regions} \name's \emph{transferable regions}
+are an encapsulation of a data-structure that can be transferred
+between autonomous entities (\eg, between two concurrently executing
+threads or actors).  Hence, unlike stack regions, transferable regions
+are not constrained to have a lexically scoped lifetime.  (Hence, we
+also refer to them as \emph{dynamic} regions.)
 
-Furthermore, transferable regions, unlike stack regions, are first class values of \name:
-they are objects of the class \C{Region}, 
-they are created using the \C{new} keyword,
-and can be passed as arguments, stored in data structures, and returned from methods.
-A transferable region is intended to encapsulate a single data-structure,
-consisting of a collection of objects with a distinguished root object of some type \C{T},
-which we refer to as the region's \emph{root} object.
-The class \C{Region} is parametric over the type \C{T} of this root object.
+Furthermore, transferable regions, unlike stack regions, are first
+class values of \name: they are objects of the class \C{Region}, they
+are created using the \C{new} keyword, and can be passed as arguments,
+stored in data structures, and returned from methods.  A transferable
+region is intended to encapsulate a single data-structure, consisting
+of a collection of objects with a distinguished root object of some
+type \C{T}, which we refer to as the region's \emph{root} object.  The
+class \C{Region} is parametric over the type \C{T} of this root
+object.
 
-The \C{Region} constructor takes as a parameter a function that constructs
-the root object: it creates a new region and invokes this function, with the new
-region as the allocation context, to create the root object of the region.
-This following code illustrates the creation of a transferable region:
+The \C{Region} constructor takes as a parameter a function that
+constructs the root object: it creates a new region and invokes this
+function, with the new region as the allocation context, to create the
+root object of the region. The following code illustrates the
+creation of a transferable region, whose root is an object of type
+\C{A}.
 % (\C{r}) of type \C{T}:
 \begin{codejava} 
-  Region<T> r = new Region<T>(() => new T())
+  Region<A> rgn = new Region<T>(() => new A())
 \end{codejava} 
+In the above code, \C{rgn} is called the \emph{handler} to the newly
+created region, and is required to read the contents of the region, or
+change its state. The class \C{Region} offers two methods: a \C{free}
+method that deallocates the region (and all the objects allocated
+within it), and \C{transfer} method that transfers the region to a
+consumer process. It is an abstraction of two possible forms of
+transfer: a transfer between two processes in a shared memory setting
+or a transfer between two processes in a distributed, message-passing,
+setting. The precise semantics of \C{transfer} are unimportant in the
+context of the region type system and we will not discuss them
+further.
 
-\begin{figure}
-\begin{codejava}
-  public class Region<T> {
-    public Region(Func<T> mkRoot);
-    public void free();
-    public void transfer();
-  }
-\end{codejava}
-\caption{The type signature of the \C{Region} class}
-\label{fig:region-class}
-\end{figure}
+% \begin{figure} \begin{codejava} public class Region<T> { public
+% Region(Func<T> mkRoot); public void free(); public void transfer();
+% } \end{codejava} \caption{The type signature of the \C{Region}
+% class} \label{fig:region-class} \end{figure}
 
-The type signature of the class \C{Region} appears in Fig.~\ref{fig:region-class}.
-The method \C{free} deallocates the region (including all objects allocated within it).
-The method \C{transfer} transfers the region to a downstream actor as determined by the run-time.
-It is an abstraction of two possible forms of  transfer: a transfer between two actors in a shared
-memory setting or a transfer between two actors in a distributed, message-passing, setting. 
-The precise semantics of \C{transfer} are unimportant in the context of the region type
-system and we will not discuss them further.
+% The type signature of the class \C{Region} appears in
+% Fig.~\ref{fig:region-class}.  The method \C{free} deallocates the
+% region (including all objects allocated within it).  The method
+% \C{transfer} transfers the region to a downstream actor as
+% determined by the run-time.  It is an abstraction of two possible
+% forms of  transfer: a transfer between two actors in a shared memory
+% setting or a transfer between two actors in a distributed,
+% message-passing, setting.  The precise semantics of \C{transfer} are
+% unimportant in the context of the region type system and we will not
+% discuss them further.
 
-\paragraph{Open and Closed Regions}
-A transferable region must be explicitly \emph{opened} using \name's \C{open} construct
-in order to either read or update or allocate objects in the region.
-Specifically, the construct ``\C{open rgn as v@R S}'' does the following:
-(a). It opens the transferable region handled by \C{rgn} for allocation
-(i.e., makes it the current allocation context),
-(b). binds the identifier \C{R} to this open region, and
-(c). initializes the newly introduced local variable \C{v} to refer
-to the root object of the region.
-The \C{@R} part of the statement is optional and may be omitted.
-The \C{open} construct is intended to simplify the problem of ensuring memory safety,
-as will be explained soon.
-We refer to a transferable region that has not been opened as a \emph{closed} region.
+\paragraph{Open and Closed Regions} A transferable region must be
+explicitly \emph{opened} using \name's \C{open} construct in order to
+either read or update or allocate objects in the region.
+Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
+following: (a). It opens the transferable region handled by \C{rgn}
+for allocation (i.e., makes it the current allocation context), (b).
+binds the identifier \C{R} to this open region, and (c). initializes
+the newly introduced local variable \C{v} to refer to the root object
+of the region.  The \C{@R} part of the statement is optional and may
+be omitted.  The \C{open} construct is intended to simplify the
+problem of ensuring memory safety, as will be explained soon.  We
+refer to a transferable region that has not been opened as a
+\emph{closed} region.
 
@@
 Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
-of Fig.~\ref{fig:motivating-eg} can be written in \name.
-The \C{onReceive} method receives its input message in a \emph{transferred}
-region (\ie, a \emph{closed} region whose ownership is transferred to the
-recipient).
-Line 7 creates a new region to store the output for time \C{t},
-initializing it to contain an empty list.
-Line 9 opens the input region to process it.
-Line 10 creates a stack region \C{R0}.
-Thus, the temporary objects created by the iteration in line 11,
-for example, will be allocated in this stack region that lives just
-long enough.
-We open the desired output region in line 12, so that the new output
-objects created by the invocation of \C{selector} in line 13
-are allocated in the output region.
-Finally, the input region is freed in line 19.
+of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
+\C{onReceive} method receives its input message in a
+\emph{transferred} region (\ie, a \emph{closed} region whose ownership
+is transferred to the recipient).  Line 7 creates a new region to
+store the output for time \C{t}, initializing it to contain an empty
+list.  Line 9 opens the input region to process it.  Line 10 creates a
+stack region \C{R0}.  Thus, the temporary objects created by the
+iteration in line 11, for example, will be allocated in this stack
+region that lives just long enough.  We open the desired output region
+in line 12, so that the new output objects created by the invocation
+of \C{selector} in line 13 are allocated in the output region.
+Finally, the input region is freed in line 15. The output region at
+\C{map[t]} stays as along as input messages with timestamp \C{t} keep
+arriving. When the timing message for \C{t} arrives, the \C{onNotify}
+method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
+After the transfer, \C{SelectVertex} is no longer allowed to access
+the \C{outRgn}, facilitating its deallocation.
 
@@
 
-\subsection{Memory Safety}
-\label{sec:memory-safety}
+\subsection{Memory Safety} \label{sec:memory-safety}
 
-Our goal is a type system that can ensure the memory safety of programs that use the region
-constructs described above.
-The key to memory safety in \name is the following restriction:
-an object $o_1$ in a region $R_1$ is allowed to store a pointer to
-an object $o_2$ in a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.
-(A similar restriction applies in the case where $o_1$ is a stack-allocated variable.)
+Our goal is a type system that can ensure the memory safety of
+programs that use the region constructs described above.  The key to
+memory safety in \name is the following restriction: an object $o_1$
+in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
+a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
+similar restriction applies in the case where $o_1$ is a
+stack-allocated variable.)
 
-Enforcing this restriction is simple in the case of stack regions since the outlives relation
-between stack regions can be inferred from their lexical nesting. Unfortunately,
-inferring outlives relations between transferable regions is not easy.
-\name imposes the following protocol on the use of transferable regions to help simplify
-this check.
+Enforcing this restriction is simple in the case of stack regions
+since the outlives relation between stack regions can be inferred from
+their lexical nesting. Unfortunately, inferring outlives relations
+in presence of transferable regions is not easy.  \name imposes the following
+protocol on the use of transferable regions to help simplify this
+check.
 
-A transferable region (that has not been freed or transferred) can be in one of two possible
-states, \emph{open} or \emph{closed}. A newly created region is in the closed state.
-A region must be opened, using the open construct (as explained previously), in order
-to read or update or allocate an object within that region.
-An open region cannot be freed or transferred. 
-In particular, an open region is guaranteed to be live for the entire duration of the open construct.
-This allows the type system to infer a valid outlives relation between the opened region
-and any stack region that is nested within the open construct.
+A transferable region (that has not been freed or transferred) can be
+in one of two possible states, \emph{open} or \emph{closed}. A newly
+created region is in the closed state.  A region must be opened, using
+the open construct (as explained previously), in order to read or
+update or allocate an object within that region.  An open region
+cannot be freed or transferred.  In particular, an open region is
+guaranteed to be live for the entire duration of the open construct.
+This allows the type system to infer a valid outlives relation between
+the opened region and any stack region that is nested within the open
+construct.
 
@@
 \label{fig:region-fsm} 
+\vspace*{-0.25in} 
 \end{figure}
 
-The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
-A transferable region starts its lifetime in a \emph{closed} state, when it is created.
+The protocol for transferable regions is presented as a finite state
+machine in Fig.~\ref{fig:region-fsm}.  
+% A transferable region starts
+% its lifetime in a \emph{closed} state, when it is created.
 %
-The safety of memory accesses in \name is now subject to the condition that
-every transferable region correctly follows the state transition discipline shown
-in Fig.~\ref{fig:region-fsm}. If this is guaranteed, then \name's region type system
-statically guarantees the safety of all memory accesses.
-% In other words, the type system reduces
-% the problem of ensuring memory safety in \name programs to the problem
-% of enforcing the state transition discipline for transferable regions.
+The safety of memory accesses in \name is now subject to the condition
+that every transferable region correctly follows the state transition
+discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
+region type system statically guarantees the safety of all memory
+accesses.
+% In other words, the type system reduces the problem of ensuring
+% memory safety in \name programs to the problem of enforcing the
+% state transition discipline for transferable regions.
 
 In \name, this enforcement is done at runtime by explicitly keeping
-track of the \emph{current state} for $\RgnZ$ objects, and
-checking the validity of every open, transfer, or free operation
-and throwing an exception if it is invalid.
-The challenge in enforcing this discipline statically is that transferable regions
-are first-class objects. Hence, the program can create multiple aliases for
-the same region, \eg, open it via one alias and free it via another.
-Typestate verification in the presence of aliases is hard.
-The checking can be done statically by preventing the creation of aliases using, \eg, linear types
-or unique types. However, this would be quite restrictive, in terms of expressiveness.
+track of the \emph{current state} for $\RgnZ$ objects, and checking
+the validity of every open, transfer, or free operation and throwing
+an exception if it is invalid.  
+% The challenge in enforcing this
+% discipline statically is that transferable regions are first-class
+% objects. Hence, the program can create multiple aliases for the same
+% region, \eg, open it via one alias and free it via another.  Typestate
+% verification in the presence of aliases is hard.  The checking can be
+% done statically by preventing the creation of aliases using, \eg,
+% linear types or unique types. However, this would be quite
+% restrictive, in terms of expressiveness.
+As explained previously, this is a reasonable tradeoff in the context
+of \name, as regions are coarse-grained objects, which are manipulated
+infrequently, when compared to fine-grained objects that reside inside
+these regions. Therefore, runtime overhead of checking the region's
+state transition discipline is acceptable.
 
-\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
-in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
-as well as the runtime overhead of checking the region's state transition discipline is acceptable.
+\paragraph{Cloning} Note that in the example from
+Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
+\C{selector} (on Line 13) should not contain any references to the
+input object, since the input region, where the object resides, will
+be freed at the end of the method. If there is a need for the output
+object to point to subobjects of the input object, such subobjects
+must be cloned (to copy them from the input region to the output
+region).  Fortunately, \name's region type system
+(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
+type of \C{selector} and the type checker will ensure correctness.
+Furthermore, the type can be automatically inferred by \name's region
+type inference (\S~\ref{sec:type-inference}), which can peform the
+above reasoning on behalf of the programmer.
 
-\paragraph{Cloning}
-Note that the example in Fig.~\ref{fig:motivating-eg-in-broom} works correctly only if \C{selector}
-creates an output object that does not share any subobjects with its input object (which resides in
-the input region and will be freed at the end of the method).
-If there is a need for the output object to point to subobjects of the input object, such subobjects
-must be cloned (to copy them from the input region to the output region).
-Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of \C{selector}
-and the type checker will ensure correctness.
-Furthermore, the type can be automatically inferred by \name's region type
-inference (\S~\ref{sec:type-inference}), which can peform the above
-non-trivial reasoning on behalf of the programmer.
+%\begin{codejava} Region<string> rgn = new Region<List<String>> (() =>
+%new List<String>()); open rgn as  strList@R0 {
+%strList.addAtHead("World"); strList.addAtHead("Hello"); }
+%rgn.transfer(); \end{codejava}
 
-%\begin{codejava}
-%  Region<string> rgn = new Region<List<String>>
-%                        (() => new List<String>());
-%  open rgn as  strList@R0 {
-%    strList.addAtHead("World");
-%    strList.addAtHead("Hello");
-%  }
-%  rgn.transfer();
-%\end{codejava}
-
-\COMMENT{
-\subsection{Stack Regions and Outlives Relation}
+\COMMENT{ \subsection{Stack Regions and Outlives Relation}
 
@@
 by \C{letregion} expressions as \emph{stack regions}. The stack
-discipline induces an \emph{outlives} relationship among regions created
-by nested \C{letregion}s, where the region introduced by the outer
-\C{letregion} is guaranteed to outlive the one introduced by the inner
-\C{letregion}. It is therefore safe to refer to an object allocated in
-outer region from the inner region, but the converse is not true. For
-example, consider the following code (assume that class \C{A} has
-field \C{x} of type \C{Object}):
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    A a0 = new@R0 A();
-    letregion R1 {
-      A a1 = new@R1 A();
-      a1.x = new@R0 Object(); // safe & legal
-      a0.x = new@R1 Object(); // unsafe & illegal
-      ...
-    }
-    ...
-  }
-\end{codejava}
-\end{center}
-The code creates two stack regions with identifiers \C{R0} and \C{R1},
-where the region \C{R0} outlives the region \C{R1} (denoted as $\C{R0}
-\outlives \C{R1}$).  Objects \C{a0} and \C{a1} are allocated in regions
-\C{R0} and \C{R1}, respectively. The first assignment statement
-assigns to \C{a1.x} an object allocated in outer region (\C{R0}). This
-assignment is safe as \C{a1.x} refers to a longer living object, hence
-is guaranteed to be a valid reference throughout the lifetime of
-\C{a1}.  In contrast, the second assignment is unsafe, as it assigns
-to \C{a0.x} an object, whose lifetime is shorter than the lifetime of
-\C{a0}, making it unsafe to dereference \C{a0.x} outside the inner
-block. 
+discipline induces an \emph{outlives} relationship among regions
+created by nested \C{letregion}s, where the region introduced by the
+outer \C{letregion} is guaranteed to outlive the one introduced by the
+inner \C{letregion}. It is therefore safe to refer to an object
+allocated in outer region from the inner region, but the converse is
+not true. For example, consider the following code (assume that class
+\C{A} has field \C{x} of type \C{Object}): \begin{center}
+\begin{codejava} letregion R0 { A a0 = new@R0 A(); letregion R1 { A a1
+= new@R1 A(); a1.x = new@R0 Object(); // safe & legal a0.x = new@R1
+Object(); // unsafe & illegal ...  } ...  } \end{codejava}
+\end{center} The code creates two stack regions with identifiers
+\C{R0} and \C{R1}, where the region \C{R0} outlives the region \C{R1}
+(denoted as $\C{R0} \outlives \C{R1}$).  Objects \C{a0} and \C{a1} are
+allocated in regions \C{R0} and \C{R1}, respectively. The first
+assignment statement assigns to \C{a1.x} an object allocated in outer
+region (\C{R0}). This assignment is safe as \C{a1.x} refers to a
+longer living object, hence is guaranteed to be a valid reference
+throughout the lifetime of \C{a1}.  In contrast, the second assignment
+is unsafe, as it assigns to \C{a0.x} an object, whose lifetime is
+shorter than the lifetime of \C{a0}, making it unsafe to dereference
+\C{a0.x} outside the inner block. 
 % Unsafe assignments can also happen indirectly via a function call..
@@
 
-In a typical dataflow computation, an upstream actor
-(e.g., a \C{SELECT} operator) constructs a transferable region, sets
-its root to the data structure containing intermediate results, and
-then transfers it to a downstream actor (e.g., a \C{COUNT} operator),
-which performs further processing. Since a transferable region escapes
-the lifetime of the sender, there must be no references from inside of
-the transferable region to objects allocated in other memory regions
-of the sender. Such references, if exist, may become invalid
-references in the context of recipient's address space, jeopardizing
-memory safety. \name relies on its region type system to prevent such
-unsafe references from being created.
+In a typical dataflow computation, an upstream actor (e.g., a
+\C{SELECT} operator) constructs a transferable region, sets its root
+to the data structure containing intermediate results, and then
+transfers it to a downstream actor (e.g., a \C{COUNT} operator), which
+performs further processing. Since a transferable region escapes the
+lifetime of the sender, there must be no references from inside of the
+transferable region to objects allocated in other memory regions of
+the sender. Such references, if exist, may become invalid references
+in the context of recipient's address space, jeopardizing memory
+safety. \name relies on its region type system to prevent such unsafe
+references from being created.
 
@@
 with the data stored in a transferable region. Consider the following
-code, for instance\footnote{For brevity, we drop the region identifier binding part of the
-\C{open} expression whenever the identifier is not used.}:
-\begin{codejava}
-void onReceive(Region<List<String>> rgn)
-  open rgn as strList {
-    letregion R1 {
-      String s = "";
-      ListIterator<String> i = strList.listIterator();
-      while(i.hasNext()) {
-        s += i.getNext();
-      }
-      print s; //prints "HelloWorld"
-    }
-  }
-  rgn.free();
-\end{codejava}
-The stack region \C{R1} is being used in the above code to provide
-working memory to work with the objects of transferable region
-\C{rgn}. Since a transferable region cannot be transferred/freed while
-it is still open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to
-outlive the stack region \C{R1} in the above code, making it safe for
-the later to contain references to the former. \name therefore allows
-such references.
+code, for instance\footnote{For brevity, we drop the region identifier
+binding part of the \C{open} expression whenever the identifier is not
+used.}: \begin{codejava} void onReceive(Region<List<String>> rgn) open
+rgn as strList { letregion R1 { String s = ""; ListIterator<String> i
+= strList.listIterator(); while(i.hasNext()) { s += i.getNext(); }
+print s; //prints "HelloWorld" } } rgn.free(); \end{codejava} The
+stack region \C{R1} is being used in the above code to provide working
+memory to work with the objects of transferable region \C{rgn}. Since
+a transferable region cannot be transferred/freed while it is still
+open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to outlive the
+stack region \C{R1} in the above code, making it safe for the later to
+contain references to the former. \name therefore allows such
+references.
 
-As Fig.~\ref{fig:region-fsm}
-indicates, it is not possible to open a region that is already
-transferred/freed. The fact that \C{rgn} is open within the block
-therefore guarantees that it is not yet transferred/freed, and that
-dereferencing \C{strList} within the block is safe. The end of
-\C{open} block marks the return of transferable region to the closed
-state.  While closed, the region is eligible to be transferred to a
-downstream actor, or to be freed.  An actor that receives the
-transferable region, receives it in the closed state. It can then
-reopen the received region to read its contents, possibly add more
-data and transfer it to another actor, or free the region.
+As Fig.~\ref{fig:region-fsm} indicates, it is not possible to open a
+region that is already transferred/freed. The fact that \C{rgn} is
+open within the block therefore guarantees that it is not yet
+transferred/freed, and that dereferencing \C{strList} within the block
+is safe. The end of \C{open} block marks the return of transferable
+region to the closed state.  While closed, the region is eligible to
+be transferred to a downstream actor, or to be freed.  An actor that
+receives the transferable region, receives it in the closed state. It
+can then reopen the received region to read its contents, possibly add
+more data and transfer it to another actor, or free the region.
 

### 2015-11-17 paper/type-system.tex
Edits.
Made "fresh" fix to the opsem.
Some edits to Sec.3.

--- 2015-11-03 paper/type-system.tex
@@
-\section{The Core Language}
+\section{\fbname}
 \label{sec:type-system}
 
-\input{fb-syntax}
+The purpose of \name's region type system is to enforce the key
+invariant required for memory safety, namely that an object $o_1$ in a
+region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
+$R_2$ is guaranteed to outlive $R_1$. Intuitively, the invariant can
+be enforced by (a). tracking \emph{outlives} relationships between
+various regions in the program (b). tagging the C\# type of every
+object in the program with its allocation region, and (c).  ensuring
+that, when a reference is created, its target object ($o_2$) is
+allocated in a region that is known to be in outlives relationship
+with the object containing the reference. We now formally develop this
+intuition via \fbname, our explicitly typed core language (with region
+types) that incorporates the features introduced in the previous
+section. \fbname builds on the Featherweight Generic Java
+(FGJ)~\cite{fgj} formalism, and reuses notations and various
+definitions from~\cite{fgj}, such as the definition of type
+well-formedness for the core (region-free) language.
+
+% In this section, we formally develop \fbname, a language that
+% incorporates the features introduced in the previous section, as well
+% as its type system, building on the Featherweight Generic Java
+% (FGJ)~\cite{fgj} formalism.  Our development reuses notations and
+% various definitions from~\cite{fgj}, such as the definition of type
+% well-formedness for the core (region-free) language, which are
+% elaborated further in the supplement~\cite{techrep}.
 
@@
 
-We build on the Featherweight Generic Java (FGJ)~\cite{fgj} formalism
-to formalize \name and its region type system. Our development reuses
-notations from~\cite{fgj}, and relies on several of its definitions,
-such as the definition of type well-formedness for the core
-(region-free) language. We refrain from reproducing them here, and
-instead encourage the reader to refer to~\cite{fgj,techrep}.
+\input{fb-syntax}
 
-Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
-which we call \fbname (\FB). Class types in \FB are region-annotated
-variants of class types in FGJ (also called \emph{core types}). This
-correspondence is reflected in the $T@\rgn$ syntax of a region type,
-which is the simplest form of a region type describing an object of
-core type $T$ contained in a region $\rgn$. We let $\rgn$ range over
-static identifiers of regions in \FB. The only unboxed value in \FB is
-$\unitval$ of type \unitZ. Rest are objects.
+Fig~\ref{fig:fb-syntax} describes the syntax of \fbname (\FB). Class
+types in \FB are region-annotated variants of class types in FGJ (also
+called \emph{core types}). This correspondence is reflected in the
+$T@\rgn$ syntax of a region type, which is the simplest form of a
+region type describing an object of core type $T$ contained in a
+region $\rgn$. We let $\rgn$ range over static identifiers of regions
+in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
+Rest are objects.
 
@@
 \begin{array}{lclcl}
-v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \\
+v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
+\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
 \end{array}

### 2015-11-17 paper/type-system.tex
section 3

@@
 
-The actual implementations of region-based memory management for \name
-may choose to realize the $\toprgn$ as a garbage-collected region, but
-this is not required. Our operational semantics
-(\S\ref{sec:fb-opsem}), for example, does not even associate a region
-with $\toprgn$, instead opting to allocate $\RgnZ$ objects in the same
-transferable regions that they represent. Our type safety result
-nonetheless guarantees memory safety in this implementation.
+%The actual implementations of region-based memory management for \name
+%may choose to realize the $\toprgn$ as a garbage-collected region, but
+%this is not required. Our operational semantics
+%(\S\ref{sec:fb-opsem}), for example, does not even associate a region
+%with $\toprgn$, instead opting to allocate $\RgnZ$ objects in the same
+%transferable regions that they represent. Our type safety result
+%nonetheless guarantees memory safety in this implementation.
 

### 2015-11-17 paper/type-system.tex
updated section 3

@@
 with the object containing the reference. We now formally develop this
-intuition via \fbname, our explicitly typed core language (with region
+intuition via \fbname (\FB), our explicitly typed core language (with region
 types) that incorporates the features introduced in the previous
@@
 
-Fig~\ref{fig:fb-syntax} describes the syntax of \fbname (\FB). Class
-types in \FB are region-annotated variants of class types in FGJ (also
-called \emph{core types}). This correspondence is reflected in the
-$T@\rgn$ syntax of a region type, which is the simplest form of a
-region type describing an object of core type $T$ contained in a
-region $\rgn$. We let $\rgn$ range over static identifiers of regions
-in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
-Rest are objects.
-
-\FB extends FGJ's expression language with assignments, local variable
-declarations (via \C{let} expressions), region constructs
-(\C{letregion} and \C{open}), lambda abstraction
-($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
-($e\inang{...}(\bar{e})$) expressions. Assignments and region
-constructs, which are statements in \name, are treated as effectful
-expressions in \FB. Lambda abstraction and application expressions are
-to define and apply anonymous functions. Modulo the angle braces
-(whose presence will be justified later), these expressions are
-uncurried variants of the corresponding expressions from System F.
-Since FGJ formalism does not include higher-order functions, we extend
-the syntactic class (\C{T}) of FGJ types with an (uncurried) arrow
-type.
-
-Class definitions in \FB extend FGJ's with region parameters
-($\rho$). Informally, region variables ($\rho$) are to region
-identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
-they can be instantiated with region identifiers, as needed. A class
-in \FB is necessarily parameterized over the allocation region of its
-objects. We use $\rho^a$ to denote this parameter, and often refer to
-it as \emph{class}'s allocation region, although it is really the
-allocation region of its objects. An Object in \FB can contain fields
-referring to objects allocated in regions ($\rhobar$) other than its
-own allocation region ($\rhoalloc$), provided that the former outlive
-the later (i.e., $\rhobar \outlives \rho$). In such cases, the
-definition of object's class needs to be parametric over allocation
-regions of its fields (i.e., their classes). Furthermore, the
-constraint that such regions must outlive the allocation region of the
-class needs to be made explicit in the definition. For instance, a
-\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
-component coming from ${\rho_1}$ and second from $\rho_2$, where all
-the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
-$\rhoalloc$. The corresponding definition of the \C{Pair} class is as
-follows(The symbol $\extends$ should be read \emph{extends}):  
+Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
+We refer to the class types of FGJ as \emph{core types}.
+%
+The following example illustrates some  key elements of \FB.
 \begin{codejava}[mathescape=true]
@@
 \end{codejava}
+
+A class in \FB is parametric over zero or more type variables (as in FGJ) as well as
+one or more region variables $\rho$.
+(The symbol $\extends$ should be read \emph{extends}.)
+% In the above example, the class \C{Pair} is parameterized over $\rho^a, \rho_1, \rho_2$.
+We refer to the first region parameter, usually denoted $\rho^a$, as the \emph{allocation
+region} of the class: it serves to identify the region where an instance of the class is allocated.
+%
+An object in \FB can contain fields referring to objects allocated in regions ($\rhobar$) other than its
+own allocation region ($\rhoalloc$), provided that the former outlive
+the later (i.e., $\rhobar \outlives \rho$). In such cases, the
+definition of object's class needs to be parametric over allocation
+regions of its fields (i.e., their classes). Furthermore, the
+constraint that such regions must outlive the allocation region of the
+class needs to be made explicit in the definition. 
+%
+For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
+component coming from ${\rho_1}$ and second from $\rho_2$, as long as
+$\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.
+
 To construct objects of the \C{Pair} class, its type and region
@@
 fields, hence its region type is of form $\ObjZ@\rgn$.
+
+%Class definitions in \FB extend FGJ's with region parameters
+%($\rho$). Informally, region variables ($\rho$) are to region
+%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
+%they can be instantiated with region identifiers, as needed. A class
+%in \FB is necessarily parameterized over the allocation region of its
+%objects. We use $\rho^a$ to denote this parameter, and often refer to
+%it as \emph{class}'s allocation region, although it is really the
+%allocation region of its objects. An Object in \FB can contain fields
+%referring to objects allocated in regions ($\rhobar$) other than its
+%own allocation region ($\rhoalloc$), provided that the former outlive
+%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
+%definition of object's class needs to be parametric over allocation
+%regions of its fields (i.e., their classes). Furthermore, the
+%constraint that such regions must outlive the allocation region of the
+%class needs to be made explicit in the definition. For instance, a
+%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
+%component coming from ${\rho_1}$ and second from $\rho_2$, where all
+%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
+%$\rhoalloc$. T
+
+Class types in \FB are region-annotated variants of class types in FGJ (also
+called \emph{core types}). This correspondence is reflected in the
+$T@\rgn$ syntax of a region type, which is the simplest form of a
+region type describing an object of core type $T$ contained in a
+region $\rgn$. We let $\rgn$ range over static identifiers of regions
+in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
+Rest are objects.
+
+\FB extends FGJ's expression language with assignments, local variable
+declarations (via \C{let} expressions), region constructs
+(\C{letregion} and \C{open}), lambda abstraction
+($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
+($e\inang{...}(\bar{e})$) expressions. Assignments and region
+constructs, which are statements in \name, are treated as effectful
+expressions in \FB. Lambda abstraction and application expressions are
+to define and apply anonymous functions. Modulo the angle braces
+(whose presence will be justified later), these expressions are
+uncurried variants of the corresponding expressions from System F.
+Since FGJ formalism does not include higher-order functions, we extend
+the syntactic class (\C{T}) of FGJ types with an (uncurried) arrow
+type.
+
 % In general, the \C{@} notation in a

### 2015-11-17 paper/type-system.tex
Edits to Sec.3.

@@
 %
-The following example illustrates some  key elements of \FB.
+The following definition of \C{Pair} class in \FB illustrates some
+of the key elements of the formal language\footnote{The symbol
+$\extends$ should be read \emph{extends}, and the symbol $\outlives$
+stands for \emph{outlives}}:
 \begin{codejava}[mathescape=true]
 class Pair<a $\extends$ Object, b $\extends$ Object>
-          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$>
-           {
+          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$> {
   a@$\rho_1$ fst; 
@@
   Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
-    super(); 
-    this.fst = fst; 
-    this.snd = snd;
+    super(); this.fst = fst; this.snd = snd;
   }
@@
 
-A class in \FB is parametric over zero or more type variables (as in FGJ) as well as
-one or more region variables $\rho$.
-(The symbol $\extends$ should be read \emph{extends}.)
-% In the above example, the class \C{Pair} is parameterized over $\rho^a, \rho_1, \rho_2$.
-We refer to the first region parameter, usually denoted $\rho^a$, as the \emph{allocation
-region} of the class: it serves to identify the region where an instance of the class is allocated.
+A class in \FB is parametric over zero or more type variables (as in
+FGJ) as well as one or more region variables $\rho$.
+% In the above example, the class \C{Pair} is parameterized over
+% $\rho^a, \rho_1, \rho_2$.
+We refer to the first region parameter, usually denoted $\rho^a$, as
+the \emph{allocation region} of the class: it serves to identify the
+region where an instance of the class is allocated.
 %
@@
 own allocation region ($\rhoalloc$), provided that the former outlive
-the later (i.e., $\rhobar \outlives \rho$). In such cases, the
+the later (i.e., $\rhobar \outlives \rho$). In such case, the
 definition of object's class needs to be parametric over allocation
@@
 constraint that such regions must outlive the allocation region of the
-class needs to be made explicit in the definition. 
+class needs to be made explicit in the definition, as the \C{Pair}
+class does in the above definition. We say that the \C{Pair} class
+exhibits \emph{constrained region polymorphism}.
 %
-For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
-component coming from ${\rho_1}$ and second from $\rho_2$, as long as
-$\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.
+% For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
+% component coming from ${\rho_1}$ and second from $\rho_2$, as long as
+% $\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.
 
@@
 parameters need to be instantiated with core types ($T$) and concrete
-region identifiers ($\rgn$), respectively. For example:
+region identifiers\footnote{Region variables ($\rho$) and region
+identifiers ($\rgn$) belong to the same syntactic class. Region
+identifiers are to region variables, as types ($T$) are to type
+variables ($a$)} ($\rgn$), respectively. For example:
 \begin{codejava}
-letregion $\rgn_0$ {
-  Object<$\rgn_0$> snd = new Object<$\rgn_0$>();
-  letregion $\rgn_1$ {
-    Object<$\rgn_1$> fst = new Object<$\rgn_1$>();
+letregion $\rgn_0$ in
+  let snd = new Object<$\rgn_0$>() in
+  letregion $\rgn_1$ in
+    let fst = new Object<$\rgn_1$>() in
     let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                   (fst,snd);
-  }
-}
 \end{codejava}
@@
 In contrast, if we choose to allocate the \C{snd} object also in
-$\rgn_1$, the region type of \C{p} would be
-\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which we abbreviate
-as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}\footnote{In general, we treat
-$B\inang{\tbar}@\rgn$ as an abbreviation of
-$B\inang{\tbar}\inang{\rbar}$}. The $\ObjZ$ class in \FB has no
-fields, hence its region type is of form $\ObjZ@\rgn$.
-
+$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
+type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
+we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
+$B\inang{\tbar}@\rgn$ as being equivalent to
+$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
+$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
+instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
+\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
+\FB, hence it has no region annotations.
+% The region type of $\ObjZ$ class in \FB has no fields, hence its
+% region type is of form $\ObjZ@\rgn$. 
 %Class definitions in \FB extend FGJ's with region parameters
@@
 
-Class types in \FB are region-annotated variants of class types in FGJ (also
-called \emph{core types}). This correspondence is reflected in the
-$T@\rgn$ syntax of a region type, which is the simplest form of a
-region type describing an object of core type $T$ contained in a
-region $\rgn$. We let $\rgn$ range over static identifiers of regions
-in \FB. The only unboxed value in \FB is $\unitval$ of type \unitZ.
-Rest are objects.
+% Class types in \FB are region-annotated variants of class types in
+% FGJ (also called \emph{core types}). This correspondence is
+% reflected in the $T@\rgn$ syntax of a region type, which is the
+% simplest form of a region type describing an object of core type $T$
+% contained in a region $\rgn$. We let $\rgn$ range over static
+% identifiers of regions in \FB. The only unboxed value in \FB is
+% $\unitval$ of type \unitZ.  Rest are objects.
 
-\FB extends FGJ's expression language with assignments, local variable
-declarations (via \C{let} expressions), region constructs
-(\C{letregion} and \C{open}), lambda abstraction
-($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
-($e\inang{...}(\bar{e})$) expressions. Assignments and region
-constructs, which are statements in \name, are treated as effectful
-expressions in \FB. Lambda abstraction and application expressions are
-to define and apply anonymous functions. Modulo the angle braces
-(whose presence will be justified later), these expressions are
-uncurried variants of the corresponding expressions from System F.
-Since FGJ formalism does not include higher-order functions, we extend
-the syntactic class (\C{T}) of FGJ types with an (uncurried) arrow
-type.
+% \FB extends FGJ's expression language with local variable
+% declarations (via \C{let} expressions), region constructs
+% (\C{letregion} and \C{open}), lambda abstraction
+% ($\lambdaexp{...}{\xbar : \taubar}{e}$) and application
+% ($e\inang{...}(\bar{e})$) expressions. Lambda abstraction and
+% application expressions are to define and apply anonymous functions.
+% Modulo the angle braces (whose presence will be justified later),
+% these expressions are uncurried variants of the corresponding
+% expressions from System F.  Since FGJ formalism does not include
+% higher-order functions, we extend the syntactic class (\C{T}) of FGJ
+% types with an (uncurried) arrow type.
 
@@
 
-Like classes, methods can also exhibit qualified region polymorphism.
-A method definition in \FB is necessarily polymorphic over its
-allocation context (Sec.~\ref{sec:alloc-ctxt}), and optionally
-polymorphic with respect to the regions containing its arguments.
-Region parameters, like those on classes, are qualified with
-constraints ($\phi$).
+Like classes, methods can also exhibit constrained region
+polymorphism.  A method definition in \FB is necessarily polymorphic
+over its allocation context (\S~\ref{sec:alloc-ctxt}), and
+optionally polymorphic with respect to the regions containing its
+arguments.  Region parameters, like those on classes, are qualified
+with constraints ($\phi$).
 % Region parameters on the methods, like those on classes, are
@@
 % allocation context parameter. 
-The angle braces in the lambda expression ($\inang{\rho^a\rhobar \,|\,
-\phi}$) serve the same purpose as they do in a method definition - to
-capture region parameters along with their constraints. Region
-parameters also appear in the arrow type of the lambda expression 
+\FB extends FGJ's expression language with a lambda expression and an
+application expression ($e\inang{\ralloc\rbar}(\bar{e})$) to define
+and apply functions\footnote{We distinguish between functions and
+methods. The formed result from lambda expressions, whereas the latter
+come from class definitions}. Functions, like methods, exhibit
+constrained region polymorphism, as evident in their arrow region type
 ($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
-\tau$) at the prenex position, similar to ML type schemes. However,
-unlike ML, we don't distinguish between types and type schemes; any of
-the $\tau$'s in the arrow type can themselves be region-parametric
-arrow types. Like with methods, region parameters on functions can be
-instantiated when they are applied.
+\tau$).
+% at the prenex position, similar to ML type schemes. However,
+% unlike ML, we don't distinguish between types and type schemes; any of
+% the $\tau$'s in the arrow type can themselves be region-parametric
+% arrow types. Like with methods, region parameters on functions can be
+% instantiated when they are applied.
 % In this respect, our region type system is more like
@@
 % instantiate and generalize region variables. 
-
-A lambda expression creates a closure, which can escape the
-context in which it is created. It is therefore important to keep track of
-the region in which a closure is allocated in order to avoid unsafe
-dereferences. The $\rgn$ annotation above the arrow in the arrow
-type denotes the allocation region of the corresponding closure. Note
-that it is important to distinguish between the allocation context
-argument ($\rhoalloc$) of a function and the allocation region
-($\rgn$) of its closure. In \name, the later corresponds to the region where
-a \C{Func} object is allocated, while the former corresponds to the
-region where it is applied ($e\inang{\ralloc\rbar}(\bar{e})$). 
+Since a function closure can escape the context in which it is
+created, it is important to keep track of the region in which it is
+created in order to avoid unsafe dereferences. The $\rgn$ annotation
+above the arrow in the arrow type denotes the allocation region of the
+corresponding closure. 
+% Note that it is important to distinguish
+% between the allocation context argument ($\rhoalloc$) of a function
+% and the allocation region ($\rgn$) of its closure. In \name, the later
+% corresponds to the region where a \C{Func} object is allocated, while
+% the former corresponds to the region where it is applied
+. 
 % For instance, in the following example:

### 2015-11-17 paper/type-system.tex
Sec 3 full pass.

@@
 \label{sec:fb-syntax}
-
-\input{fb-syntax}
 
@@
 
+\input{fb-staticsem}
+
 Like classes, methods can also exhibit constrained region
@@
 
-\input{fb-staticsem}
 \subsection{Types and Well-formedness}
 
-Static semantics of \fbname is defined by the rules to establish
-well-formedness of region types and region constraints, check
-subtyping between a pair of region types, and type-check expressions
-against region types. Fig.~\ref{fig:fb-staticsem} contains an
-illustrative subset of such rules\footnote{The full formal development
-can be found in our technical report~\cite{techrep}}, along with some
-auxiliary definitions. The judgments defined by these rules make use
-of a context ($\A$), which is a tuple of:
+Well-formedness and typing rules of \fbname establish the conditions
+under which a region type is considered well-formed, and an expression
+is considered to have a certain region type, respectively.
+Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
+rules\footnote{The full formal development can be found in the
+supplement}. The rules refer to a context ($\A$), which is a tuple of:
 \begin{itemize}
@@
 \end{itemize}
-The context for the expression typing judgment also includes:
+$\A$ also contains $\rhomap$, 
+% a map ($\rhomap \in \rgn \mapsto
+% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
+% transferable regions to their current state in the FSM from
+% Fig.~\ref{fig:region-fsm}.
+which is primarily an artifact to facilitate the type safety proof,
+and can be ignored while type checking user-written programs
+in \FB. The context for the expression typing judgment also includes:
 \begin{itemize}
@@
 \item The static identifier ($\rgn^a$) of the allocation context for
-the expression is being type-checked.  
+the expression is being typechecked.  
 \end{itemize}
+Like the judgments in FGJ~\ref{fgj}, all the judgments defined by the
+rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
+class table ($CT \in cn \mapsto D$) that maps class names to their
+definitions in \FB.
+% \begin{figure}
 
-\begin{figure}
+% \begin{codejava}
+% class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
+%   <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
+%   Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
+%     super(); 
+%     this.thunk = thunk;
+%   }
+%   unit transfer$\inang{\rho}$(unit u) {
+%     return u;
+%   }
+% }
+% \end{codejava}
 
-\begin{codejava}
-class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
-  <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
-  Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
-    super(); 
-    this.thunk = thunk;
-  }
-  unit transfer$\inang{\rho}$(unit u) {
-    return u;
-  }
-}
-\end{codejava}
-
-\caption{The stub used for $\RgnZ$ class}
-\label{fig:region-stub}
-\end{figure}
+% \caption{The stub used for $\RgnZ$ class}
+% \label{fig:region-stub}
+% \end{figure}
 
@@
 ($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
-($\vdash$) for those in \FB.}. All judgments are implicitly
-parameterized on a class table ($CT \in cn \mapsto D$) that maps class
-names to their definitions in \FB. For judgments in FGJ, we use a
-class table that maps class names to their region-erased definitions.
-We denote the region-erased class table as ($\absof{CT}$). The precise
-semantics of region erasure ($\absof{\cdot}$) are defined
-in~\cite{techrep}, but it can be understood as an operation that
-erases region annotations on types. The class table is assumed to bind
-the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
-This mostly eliminates the need to treat $\RgnZ$ objects, which double
-up as transferable region handlers, as a special case in semantics.
-
-The well-formedness rule for region instantiation in class types
-($B\inang{\tbar}\inang{\ralloc\rbar}$) requires that the region-erased
-(core) type be well-formed, instantiated regions be live, and that
-they satisfy the constraints ($\phi$) imposed by the class on its
-region parameters. The later is enforced by checking the validity of
-$\phi$, with actual region arguments substituted\footnote{The notation
+($\vdash$) for those in \FB.}. The class table ($\absof{CT}$) for such
+judgments is derived from \FB's class table ($CT$) by erasing all
+region annotations on types, and region arguments in expressions
+($\absof{\cdot}$ denotes the region erasure operation). 
+% For judgments in FGJ, we use a
+% class table that maps class names to their region-erased definitions.
+% We denote the region-erased class table as ($\absof{CT}$). The precise
+% semantics of region erasure ($\absof{\cdot}$) are defined
+% in~\cite{techrep}, but it can be understood as an operation that
+% erases region annotations on types. The class table is assumed to bind
+% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
+% This mostly eliminates the need to treat $\RgnZ$ objects, which double
+% up as transferable region handlers, as a special case in semantics.
+The well-formedness rule for class types
+($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
+the safety property that prevents objects from containing unsafe
+references. It does so by insisting that regions $\ralloc\rbar$
+satisfy the constraints ($\phi$) imposed by the class on its region
+parameters. The later is enforced by checking the validity of $\phi$,
+with actual region arguments substituted\footnote{The notation
 $[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
@@
 relations. For any well-formed core type $T$, $T@\rgn$ is a
-well-formed region type if $\rgn$ is a valid region. Well-formedness
-of region constraints requires that the regions referred by the
-constraints be live.
+well-formed region type if $\rgn$ is a valid region. The type
+$\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
+$\toprgn$ is a special immortal region that outlives every other live
+region. This arrangement allows \C{\RgnZ} handlers to be aliased and
+referenced freely from objects in various regions, regardless of their
+lifetimes. On the flip side, this also opens up the possibility of
+references between transferable regions, which become unsafe in
+context of the receipient's address space. Fortunately, such
+references are explictly prohibited by the type rule of \C{\RgnZ}
+objects, as described below.
 
-The subtype relation between region types of objects
-($B\inang{\ralloc\rbar}\inang{\tbar}$) follows directly from the
-subclass relation between their class types. Note that, in general, we
-don't lift outlives relation between regions to subtype relation
-between objects (of same core type) allocated in those regions. Such
-lifting is unsound for the same reason as covariant/contravariant
-subtyping between mutable references is unsound. However, we make a
-special case for $\RgnZ$ objects: we allow $\RgnZ$ objects allocated
-in any region to be considered as objects allocated in an eternal
-region ($\toprgn$) that outlives every other region. This lets $\RgnZ$
-objects escape the context they are created in (as required by the
-\C{transfer} operation), and also facilitates data structures of
-regions created in different contexts (for e.g., a list of $\RgnZ$
-objects, whose root objects are of (core) type $T$, can be typed
-$\C{List}\inang{\RgnZ\inang{T}}\inang{\rgn,\toprgn}$, where $\rgn$ is
-the region where the spine of the list is allocated). As described in
-\S\ref{sec:memory-safety}, runtime checks are used to prevent the
-unsound subtyping rule on region objects from effecting the memory safety.
+% Recall that a class definition in \FB is annotated with constraints
+% over its region arguments, which require the regions referred by its
+% fields to outlive its own allocation region. The well-formedness rule
+% requires that (core) type be well-formed, instantiated regions be
+% live, and that they   Well-formedness of region constraints requires
+% that the regions referred by the constraints be live.
+
+% The subtype relation between region types of objects
+% ($B\inang{\ralloc\rbar}\inang{\tbar}$) follows directly from the
+% subclass relation between their class types. Note that, in general, we
+% don't lift outlives relation between regions to subtype relation
+% between objects (of same core type) allocated in those regions. Such
+% lifting is unsound for the same reason as covariant/contravariant
+% subtyping between mutable references is unsound. However, we make a
+% special case for $\RgnZ$ objects: we allow $\RgnZ$ objects allocated
+% in any region to be considered as objects allocated in an eternal
+% region ($\toprgn$) that outlives every other region. This lets $\RgnZ$
+% objects escape the context they are created in (as required by the
+% \C{transfer} operation), and also facilitates data structures of
+% regions created in different contexts (for e.g., a list of $\RgnZ$
+% objects, whose root objects are of (core) type $T$, can be typed
+% $\C{List}\inang{\RgnZ\inang{T}}\inang{\rgn,\toprgn}$, where $\rgn$ is
+% the region where the spine of the list is allocated). As described in
+% \S\ref{sec:memory-safety}, runtime checks are used to prevent the
+% unsound subtyping rule on region objects from effecting the memory safety.
 % \begin{smathpar}
@@
 
-As expected, subtyping between arrow types is contravariant in
-argument types and covariant in return types. It is also contravariant
-in the region constraints that the region parameters have to satisfy.
-For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
-\unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
-A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
-weaker than the later's. Finally, the subtyping is invariant with
-respect to the allocation region ($\rgn$) of closures. 
+% As expected, subtyping between arrow types is contravariant in
+% argument types and covariant in return types. It is also contravariant
+% in the region constraints that the region parameters have to satisfy.
+% For example, the type $\inang{\rhoalloc\rho} A@\rho \xrightarrow{\rgn}
+% \unitZ$ is a subtype of $\inang{\rhoalloc\rho \,|\, \rho =  \rhoalloc}
+% A@\rho \xrightarrow{\rgn} \unitZ$, for the former's precondition is
+% weaker than the later's. Finally, the subtyping is invariant with
+% respect to the allocation region ($\rgn$) of closures. 
 % This subtype rule assumes that both arrow types have same set of
@@
 
+The type rules distinguish between the \C{new} expressions that create
+objects of  the \RgnZ class, and \C{new} expressions that create
+objects of other classes. The rule for the later relies on an
+auxiliary definition called $\fields$ that returns the sequence of
+type bindings for fields (instance variables) of a given class type.
+Like in FGJ, the names and types of a constructor's arguments in \FB
+are same as the names and types of its class's fields, and the type
+rule relies on this fact to typecheck constructor applications. Note
+that this rule does not apply to \C{new} expressions involving \RgnZ
+class, as we do not define $\fields$ for \RgnZ.
+
+The type rule for \C{new \RgnZ} expressions expects the \RgnZ class's
+constructor to be called with a nullary function that returns a value
+in its allocation context. It enforces this by typechecking the body
+($e$) of the function under an empty context containing nothing but
+the allocation context of the function. This step ensures that the
+value returned by the function stores no references to objects
+allocated elsewhere, including the top region ($\toprgn$), thus
+preventing cross-region references originating from transferable
+regions\footnote{The body of the function ($e$) might, however, create
+new (transferable) regions while execution, but that is fine as long
+as such regions, and objects allocated in them, don't find their way
+into the result of evaluating $e$}.
+
 The type rule for \C{letregion} expression requires that the static
@@
 satisfied, the expression ($e$) under \C{letregion} is then
-type-checked against $\unitZ$ assuming that the new region is live
-($\rgn \in \rhoenv$) and that it is outlived by all existing live
-regions ($\rhoenv \outlives \rgn$).
+typechecked assuming that the new region is live ($\rgn \in \rhoenv$)
+and that it is outlived by all existing live regions ($\rhoenv
+\outlives \rgn$). The result of a \C{letregion} expression must have a
+type that is well-formed under a context not containing the new
+region. This ensures that the value obtained by evaluating a
+\C{letregion} expression contains no references to the temporary
+objects inside the region\footnote{The supplement contains a formal
+proof of this claim}.
 
-Like the rule for \C{letregion}, the type rule for the \C{open}
-expression requires that the static identifier ($\rgn$) introduced to
-refer to the transferable region being opened be unique. However,
-unlike the \C{letregion} rule, the rule for \C{open} does not
-introduce any outlives relationship between the newly opened region
-and any pre-existing region while checking the type of the expression
-($e$) under \C{open}. This effectively preempts the possibility of any
-references from objects inside the transferable region to those
-outside. Environment ($\env$) is extended with binding for the type of
-root object while type-checking $e$.
+The rule for \C{open} expression, unlike the rule for \C{letregion},
+does not introduce any outlives relationship between the newly opened
+region and any pre-existing region while checking the type of the
+expression ($e$) under \C{open}. This prevents new objects allocated
+inside the transferable region from storing references to those
+outside.  Environment ($\env$) is extended with binding for the type
+of root object while typechecking $e$.
 
-The type rule for lamdba expression requires that its region
-parameters ($\rhoalloc\rhobar$) be unique under the current context,
-and that the constraints ($\phi$) over region parameters be
-well-formed. The lambda-bound expression ($e$) is checked under an
-extended type environment containing bindings for function's
-arguments, assuming that region parameters are live, and that declared
-constraints over region parameters hold. Note that the closure is
-always allocated in the current allocation context ($\rgn^a$). This
+The type rule for lamdba expression typechecks
+% requires that its region
+% parameters ($\rhoalloc\rhobar$) be unique under the current context,
+% and that the constraints ($\phi$) over region parameters be
+% well-formed. 
+the lambda-bound expression ($e$) under an extended type environment
+containing bindings for function's arguments, assuming that region
+parameters are live, and that declared constraints over region
+parameters hold. The constraints ($\phi$) are required to be
+well-formed under $\rhoenv'$, which means that $\phi$ must only refer
+to the region variables in the set $\rhoenv'$. Note that the closure
+is always allocated in the current allocation context ($\rgn^a$). This
 prevents the closure from escaping the context in which it is created,
@@
 
-The type rule for the \C{new} expression is a straightforward
-adaptation of corresponding FGJ rule to the the region type setting.
+% The type rule for the \C{new} expression is a straightforward
+% adaptation of corresponding FGJ rule to the the region type setting.
 
@@
 
-Operational semantics of \fbname defines an evaluation relation
-($\redsto$) over its expression language. In particular, the
-evaluation gets stuck if an expression attempts to dereference a
-pointer into a region that has either been deallocated or transferred.
-On the other hand, if expression attempts to commit an operation on a
-transferable region handler (a $\RgnZ$ object) that is not sanctioned
-by the state transition discipline in Fig.~\ref{fig:region-fsm}, then
-it is reduced to a value called $\invalidexn$. The accompanying
-technical report~\cite{techrep} contains full operational semantics of
-\fbname.
+The supplement defines a small-step operational semantics for \fbname
+via a five-place reduction relation:
+\begin{smathpar}
+  \redstoo{(e,\rhomap)}{(e',\rhomap')}
+\end{smathpar}
+The reduction judgment should be read as following: given a set
+($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
+from unique identifiers of transferable regions to their current
+states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
+$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
+``stuck'' if $e$ attempts to access an object whose allocation region
+is not present in $\rhoenv$, or if $e$ tries to \C{open} a
+transferable region, whose identifier is not mapped to a state by
+$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
+on a $\RgnZ$ object that is not sanctioned by the transition
+discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
+value ($\invalidexn$).
 
@@
 \lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
+  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
 \end{array}
 \end{smathpar}
+Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
+$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
+\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
+$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
+identifier ($\rgn$) for this region. A binding is also added to
+$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
+defines a type rule for such values, allowing them to be typed.
 Type safety theorem is now stated thus:
@@
 \emph{(\textbf{Type Safety})}
-\label{thm:core-safety}
-Let $\A_{\emptyset} = (\emptyset,\cdot,true)$ denote an empty context.
-If $\hastyp{\A_{\emptyset}, \toprgn, \cdot}{e}{\tau}$, then either $e$
-is a value, or $e \redsto \invalidexn$, or there exists an $e'$ such
-that $e \redsto e'$ and $\hastyp{\A_{\emptyset}, \toprgn,
-\cdot}{e'}{\tau}$.
+\label{thm:fb-type-safety}
+$\forall e, \tau, \rhoenv, \rhomap, \rgn$, if $\rgn \in \rhoenv$ and
+$\hastyp{\emptyA,\rgn, \cdot}{e}{\tau}$, then either $e$ is a value,
+or $e$ raises an exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or
+there exists an $e'$ and a $\rhomap'$ such that
+$\redstoo{(e,\rhomap)}{(e',\rhomap')}$ and $\hastyp{\emptyASigp,\rgn,
+\cdot}{e}{\tau}$.
 \end{theorem}
+
+Furthermore, we prove the following theorem about \FB, which, in
+conjunction with the type safety theorem, implies the safety of region
+transfers across adress spaces:
+\begin{theorem}
+\emph{(\textbf{Transfer Safety})}
+\label{thm:fb-type-safety}
+$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
+$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
+$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
+\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
+$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
+{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
+\end{theorem}
+The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
+is well-typed under one context, then it is also well-typed under
+every other context, whose $\rhomap$ maps $\rgn_i$ to closed
+($\CLOSED$) state.  Thus, a receipient of a transferable region only
+needs to add a binding for the region to its $\rhomap$ in order to
+preserve its type safety. Notably, this result could not have been
+established if it were possible for a transferable region to contain
+references to objects outside the region.

### 2015-11-17 paper/type-system.tex
Edits.

@@
 \label{thm:fb-type-safety}
-$\forall e, \tau, \rhoenv, \rhomap, \rgn$, if $\rgn \in \rhoenv$ and
-$\hastyp{\emptyA,\rgn, \cdot}{e}{\tau}$, then either $e$ is a value,
-or $e$ raises an exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or
-there exists an $e'$ and a $\rhomap'$ such that
-$\redstoo{(e,\rhomap)}{(e',\rhomap')}$ and $\hastyp{\emptyASigp,\rgn,
-\cdot}{e}{\tau}$.
+$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
+\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
+\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
+exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
+$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
+and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
 \end{theorem}

### 2015-11-18 paper/overview2.tex
Edits to Sec.3.
Sec 3 full pass.
Edits to Sec.4. Formatting changes.

@@
 \subsection{Using Regions in \name}
+\label{sec:alloc-ctxt}
 
@@
 % restrictive, in terms of expressiveness.
-As explained previously, this is a reasonable tradeoff in the context
+As explained previously, this is a reasonable trade-off in the context
 of \name, as regions are coarse-grained objects, which are manipulated
@@
 state transition discipline is acceptable.
+
+\input{fb-syntax}
 
@@
 Furthermore, the type can be automatically inferred by \name's region
-type inference (\S~\ref{sec:type-inference}), which can peform the
+type inference (\S~\ref{sec:type-inference}), which can perform the
 above reasoning on behalf of the programmer.

### 2015-11-18 paper/related.tex
Progress lemma.
Edits to Sec.4. Formatting changes.

@@
 concurrency. We borrow outlives relation from their formal
-development, and our type system bears some similarities to their's.
+development, and our type system bears some similarities to theirs.
 However, our language also admits parametric polymorphism (generics)
@@
 language. However, their setting only includes lexically-scoped
-regions for which Tofte and Talpin~\cite{tofte97} style analysis
+regions for which Tofte and Talpin-style analysis~\cite{tofte97}
 suffices. In contrast, our language provides first-class support for

### 2015-11-18 paper/type-inference.tex
Edits to Sec.4. Formatting changes.

@@
 
-We formalize region type inference as an elaboration function from
-programs in $\absof{\FB}$ (i.e., region-erased \FB) to programs in
-\FB.  The elaboration proceeds in four steps. In the first step we
-make use of the observation that region types are \emph{refinements}
-of FGJ types with region annotations and constraints over such region
-annotations, and compute polymorphic region type templates for methods
-and classes from their FGJ types. The templates contain region
-variables ($\rho$) to denote unknown region annotations, and predicate
-variables ($\varphi$) to denote unknown constraints over such region
-annotations. Free region variables are generalized in types (hence,
-polymorphic). Second, we make use of the computed region type
-templates to elaborate expressions by introducing region variables to
-denote unknown region arguments in \C{new} expressions, method calls
-and function applications. While elaborating expressions, we also
-build a system of constraints that capture well-formedness
-requirements and subtyping relationships between type templates that
-must hold (as per the static semantics in Fig.~\ref{fb-staticsem}) for
-the elaboration to be valid. Third, we lift expression elaboration and
-constraint generation to methods and classes. Finally, we solve the
-constraints by making use of our fixpoint constraint solving algorithm
-\csolvestar, which reduces the constraint solving problem to an
-abduction problem in a Herbrand constraint system, and then relies on
-\csolve, our abduction solver for that domain. 
+\name's region type system imposes a heavy annotation burden, and the
+cost of manually annotating C\# standard libraries with region types
+is prohibitive. We now present our region type inference algorithm
+that completely eliminates the need to write region type annotations,
+except on some higher-order functions. Formally, the type inference
+algorithm is an elaboration function from programs in $\absof{\FB}$
+(i.e., region annotation-erased \FB) to programs in \FB. The key
+aspects of this elaboration are described below.
+
+% We formalize region type inference as an elaboration function from
+% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
+% programs in \FB.  The elaboration proceeds in four steps. In the first
+% step we make use of the observation that region types are
+% \emph{refinements} of FGJ types with region annotations and
+% constraints over such region annotations, and compute polymorphic
+% region type templates for methods and classes from their FGJ types.
+% The templates contain region variables ($\rho$) to denote unknown
+% region annotations, and predicate variables ($\varphi$) to denote
+% unknown constraints over such region annotations. Free region
+% variables are generalized in types (hence, polymorphic). Second, we
+% elaborate expressions by introducing region variables to denote
+% unknown region arguments in \C{new} expressions, method calls and
+% function applications. While elaborating expressions, we make use of
+% the computed region type templates to build a system of constraints
+% that capture well-formedness requirements and subtyping relationships
+% between type templates that must hold (as per the static semantics in
+% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
+% lift expression elaboration and constraint generation to methods and
+% classes. Finally, we solve the constraints to compute solutions to
+% region ($\rho$) and predicate ($\varphi$) variables.
+%
 % Due to the presence of region-polymorphic
@@
 template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
-region variable. FGJ type $\RgnZ\inang{T}$ is special; it is
-always templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
+region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
+templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
 region-annotated definition of \C{Pair} class from
@@
 Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
-$\rho$ is fresh. For a class, a template is computed in two steps. In
-the first step, we templatize the types of all its fields, constructor
-arguments, and arguments and return values of all its methods, along
-with the type of its superclass. In the second step, we generalize all
-the free region variables occuring in the templates computed in the
-first step as region parameters of the class. Finally, we add a fresh
-allocation region parameter ($\rhoalloc$) to the list of parameters,
-and introduce a new predicate variable ($\varphi$) to denote unknown
-constraints on region parameters. For example, consider the standard
-FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
-and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
-respectively. It can be templatized as following :
+$\rho$ is fresh. For a class, a template is computed based on the 
+type templates of its fields, constructors and methods. Additional
+region and predicate variables are added as necessary. Following template 
+of the \C{Pair} class illustrates this point:
+% For a class, a template is computed in two steps. In
+% the first step, we templatize the types of all its fields, constructor
+% arguments, and arguments and return values of all its methods, along
+% with the type of its superclass. In the second step, we generalize all
+% the free region variables occuring in the templates computed in the
+% first step as region parameters of the class. Finally, we add a fresh
+% allocation region parameter ($\rhoalloc$) to the list of parameters,
+% and introduce a new predicate variable ($\varphi$) to denote unknown
+% constraints on region parameters. For example, consider the standard
+% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
+% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
+% respectively. It can be templatized as following :
 \vspace*{-0.1in}
@@
 % definition in $\FB$. 
-
-The type template for a recursively defined class is computed in two
-steps. First, all recursive occurances of the class among the types of
-its fields are ignored and the class is templatized as if it is a
-non-recursive class. Next, all the recursive occurances are
-templatized with respect to the class template computed in the first
-step, such that their region annotations are \emph{exactly} same as
-the region parameters of the class. Applying this technique on a
-recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
-example, results in a template that uses same region variable to
-denote allocation regions of the current node, and the node that the
-current node refers to. 
+%
+% The type template for a recursively defined class is computed in two
+% steps. First, all recursive occurances of the class among the types of
+% its fields are ignored and the class is templatized as if it is a
+% non-recursive class. Next, all the recursive occurances are
+% templatized with respect to the class template computed in the first
+% step, such that their region annotations are \emph{exactly} same as
+% the region parameters of the class. Applying this technique on a
+% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
+% example, results in a template that uses same region variable to
+% denote allocation regions of the current node, and the node that the
+% current node refers to. 
+%
 % For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
@@
 % region variables are introduced.
-This technique can be extended to mutually recursive definitions in a
-straightforward manner, by simultaneously templatizing them. Using the
-techniques outlined above, we compute region type templates for all
-classes bound in the class table of the $\absof{\FB}$ program before
-we proceed to elaborate expressions.
+The type template for for a recursively defined class is computed in
+two steps. First, a template is computed by ignoring the recursive
+occurrences. Next, all the recursive occurrences are templatized based
+on the template computed in first step.  This approach can be extended
+to mutually recursive definitions in a straightforward manner. Using
+the techniques outlined above, we compute region type templates for
+all classes bound in the class table of the $\absof{\FB}$ program
+before we proceed to elaborate expressions.
 
@@
 Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
-(a). replacing core types in variable declarations and \C{new}
-expressions with fresh region type templates, and (b). explicitly
-instantiating region parameters of methods with fresh region variables
-in method calls and function applications. This elaboration is
-performed with respect to the polymorphic type templates of classes
-and methods computed as per \S\ref{sec:fb-templatization}.
-
-Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-this elaboration for a subset of expressions in $\absof{\FB}$
-%, whose corresponding $\FB$ expressions have been ascribed static
-%semantics in Fig.~\ref{fig:fb-staticsem}. 
-$\elabExpr$ is defined under the same context as the expression typing
-judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
-and $\env$ retaining their meaning. The function traverses expressions
-in a syntax-directed manner of a type checker, introducing fresh
-region type templates for unknown region types, while generating
+(a). replacing core types occurring in expressions with fresh region
+type templates, and (b). explicitly instantiating region parameters
+with fresh region variables in method calls and function applications.
+This elaboration is performed with respect to the polymorphic type
+templates of classes and methods computed as per
+\S\ref{sec:fb-templatization}. The elaboration function traverses
+expressions in a syntax-directed manner of a type checker, introducing
+fresh region type templates for unknown region types, while generating
 constraints over region and predicate variables. The precise nature of
@@
 but in summary, they capture the relationships between the type
-templates of various subexpressions and their well-formedness. Note
-that $\elabExpr$ returns the region type template of the
-subexpression, which is used to generate constraints for the
-expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
-shown) used by $\elabExpr$ implement type well-formedness and subtype
-judgments from Fig.~\ref{fig:fb-staticsem}, respectively.
+templates of various subexpressions and their well-formedness.
+
+% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+% this elaboration for a subset of expressions in $\absof{\FB}$
+% %, whose corresponding $\FB$ expressions have been ascribed static
+% %semantics in Fig.~\ref{fig:fb-staticsem}. 
+% $\elabExpr$ is defined under the same context as the expression typing
+% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
+% and $\env$ retaining their meaning. The function traverses expressions
+% in a syntax-directed manner of a type checker, introducing fresh
+% region type templates for unknown region types, while generating
+% constraints over region and predicate variables. The precise nature of
+% generated constraints is explained in \S\ref{sec:fb-constraintsem},
+% but in summary, they capture the relationships between the type
+% templates of various subexpressions and their well-formedness. Note
+% that $\elabExpr$ returns the region type template of the
+% subexpression, which is used to generate constraints for the
+% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
+% shown) used by $\elabExpr$ implement type well-formedness and subtype
+% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.
 
@@
 constraint does not result in backtracking. We now describe our
-constraint solving algorithm by demonstrating its peformance over the
+constraint solving algorithm by demonstrating its performance over the
 example constraints shown in the previous section. A formal account of
@@
 \paragraph{Terminology} In a validity constraint, a predicate variable
-occuring on the left side (resp. right side) of the turnstile is said
+occurring on the left side (resp. right side) of the turnstile is said
 to occur negatively (resp. positively). A validity constraint
@@
 Solving a set ($C$) of constraints involves finding assignments for
-free region ($\rho$) and predicate ($\varphi$) varibles in the set.
+free region ($\rho$) and predicate ($\varphi$) variables in the set.
 Assignments to region variables can be found via unification (for the
@@
 $\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
-no additional constriants to help determine $\varphi_1$ or $\varphi_2$
+no additional constraints to help determine $\varphi_1$ or $\varphi_2$
 before solving this constraint. Since it is always possible to

### 2015-11-18 paper/type-system.tex
Edits to Sec.4. Formatting changes.

@@
 \end{codejava}
+\vspace*{-0.15in}
 
@@
 \end{codejava}
+%\vspace*{-0.1in}
 In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
 $\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
-and $\rgn_1$ are live during the instantation, and (b). $\rgn_0
+and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
 \outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
@@
 references between transferable regions, which become unsafe in
-context of the receipient's address space. Fortunately, such
-references are explictly prohibited by the type rule of \C{\RgnZ}
+context of the recipient's address space. Fortunately, such
+references are explicitly prohibited by the type rule of \C{\RgnZ}
 objects, as described below.
@@
 order to prevent the new region from incorrectly assuming any existing
-outlives relationships on an eponymus region. Provided this is
+outlives relationships on an eponymous region. Provided this is
 satisfied, the expression ($e$) under \C{letregion} is then
@@
 
-The type rule for lamdba expression typechecks
+The type rule for lambda expression typechecks
 % requires that its region
@@
 via a five-place reduction relation:
+\vspace*{-0.15in}
 \begin{smathpar}
@@
 conjunction with the type safety theorem, implies the safety of region
-transfers across adress spaces:
+transfers across address spaces:
 \begin{theorem}
@@
 every other context, whose $\rhomap$ maps $\rgn_i$ to closed
-($\CLOSED$) state.  Thus, a receipient of a transferable region only
+($\CLOSED$) state.  Thus, a recipient of a transferable region only
 needs to add a binding for the region to its $\rhomap$ in order to

### 2015-11-18 paper/type-inference.tex
Sec 4. done.

@@
 algorithm is an elaboration function from programs in $\absof{\FB}$
-(i.e., region annotation-erased \FB) to programs in \FB. The key
-aspects of this elaboration are described below.
+(i.e., \FB without region types, but with \C{letregion} and \C{open}
+expressions) to programs in \FB. The key aspects of this elaboration
+are described below.
 
@@
 methods, but unlike function types, there is no allocation region
-annotation atop the arrow in a method type.}
+annotation atop the arrow.}
 $\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
@@
 
-\input{fb-elabexpr}
-\input{fb-elabmeth}
+%\input{fb-elabexpr}
+%\input{fb-elabmeth}
 
-\subsection{Expression Elaboration}
+\subsection{Elaboration and Constraint Generation}
 
@@
 templates of classes and methods computed as per
-\S\ref{sec:fb-templatization}. The elaboration function traverses
-expressions in a syntax-directed manner of a type checker, introducing
-fresh region type templates for unknown region types, while generating
-constraints over region and predicate variables. The precise nature of
-generated constraints is explained in \S\ref{sec:fb-constraintsem},
-but in summary, they capture the relationships between the type
-templates of various subexpressions and their well-formedness.
+\S\ref{sec:fb-templatization}. Our expression elaboration
+function\footnote{Supplement contains the pseudocode for all the
+elaboration functions.} traverses expressions in a syntax-directed
+manner of a type checker, introducing fresh region type templates for
+unknown region types, while generating constraints over region and
+predicate variables. The precise nature of generated constraints is
+explained in \S\ref{sec:constraints}, but in summary, they
+capture the relationships between the type templates of various
+subexpressions and their well-formedness.
 
@@
 
-\subsection{Method and Class Elaboration}
+Expression elaboration can be trivially lifted to the level of a
+method, and method elaboration to the level of a class, while
+accumulating constraints. Besides such constraints, elaboration
+functions for methods and classes also generate new constraints by
+checking well-formedness and subtype relationships among the types of
+their constituents. For instance, method elaboration generates a new
+constraint that captures the requirement that, under the guarantees
+provided by the context, the region type of a method body must be a
+subtype\footnote{Subtyping is type equality in \FB. In \name,
+subtyping between region types is a straightforward extension of
+subtyping between core types.} of method's return type.  Likewise, the
+elaboration function for a class generates a fresh constraint that
+captures a couple of safety conditions: first, the allocation regions
+of objects referred by the instance variables should outlive the
+allocation region of the instance itself, and second, the allocation
+regions of a class type and its superclass type must be the same.
 
-Functions $\elabMeth$ and $\elabClass$ shown in
-Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
-class definitions, respectively. Both functions first build a context
-($\A$) containing a set ($ \rhoenv$) of regions that are currently
-live, a map ($\aenv$) mapping type variables to their bounds, and a
-constraint formula ($\phicx$) capturing constraints over live region
-variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
-capture constraints over variables in $\rhoenv$ denoting the fact that
-such constraints are yet to be inferred. Both elaboration functions
-generate new constraints by checking well-formedness and subtype
-relationships among types of their constituents. A fresh constraint
-($C_4$) generated by $\elabClass$ captures a couple of safety
-conditions: first, the allocation regions of objects referred by the
-instance variables should outlive the allocation region of the
-instance itself, and second, the allocation regions of a class type
-and its superclass type must be the same.
+Elaborating a class table ($CT$) entails elaborating every class
+definition in $CT$, while collecting constraints. The constraints thus
+collected are finally solved to obtain assignments for free region and
+predicate variables introduced during templatization and elaboration
+stages. The solutions are applied to the class table (and to the
+artifacts that make up the class table, recursively) to compute a
+class table that maps classes to their fully region-annotated
+definitions in $\FB$.
+
+% \subsection{Method and Class Elaboration}
+
+% Elaborating a method involves performingexpression elaboration over
+% its body, while collecting constraints. Method elaboration can be
+% lifted to the level of classes
+
+% Functions $\elabMeth$ and $\elabClass$ shown in
+% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+% class definitions, respectively. Both functions first build a context
+% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
+% live, a map ($\aenv$) mapping type variables to their bounds, and a
+% constraint formula ($\phicx$) capturing constraints over live region
+% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
+% capture constraints over variables in $\rhoenv$ denoting the fact that
+% such constraints are yet to be inferred. Both elaboration functions
+% generate new constraints by checking well-formedness and subtype
+% relationships among types of their constituents. A fresh constraint
+% ($C_4$) generated by $\elabClass$ captures a couple of safety
+% conditions: first, the allocation regions of objects referred by the
+% instance variables should outlive the allocation region of the
+% instance itself, and second, the allocation regions of a class type
+% and its superclass type must be the same.
 
@@
 
-Function $\elabClassTable$ (not shown) elaborates
-every definition in the class table $CT$, while accumulating
-constraints. The constraints are finally solved for free region and
-predicate variables introduced during templatization and elaboration
-stages. The solutions are applied to the class table (and to the
-artifacts that make up the class table, recursively) to compute a
-class table that maps classes to their fully region-annotated
-definitions in $\FB$.
+% Function $\elabClassTable$ (not shown) elaborates
+% every definition in the class table $CT$, while accumulating
+% constraints. The constraints are finally solved for free region and
+% predicate variables introduced during templatization and elaboration
+% stages. The solutions are applied to the class table (and to the
+% artifacts that make up the class table, recursively) to compute a
+% class table that maps classes to their fully region-annotated
+% definitions in $\FB$.
 
@@
 
-Our constraint generation algorithm generates three kinds of
-constraints:
+Three kinds of constraints are generated during elaboration:
 \begin{itemize}
@@
 
-Our constraint generation algorithm traverses the entire program,
-performing elaboration and collecting constraints. However, we do not
-require region type inference to be a whole-program analysis.
-Compositionality of type inference is reclaimed by solving the
-constraints in a compositional fashion. Concretely, this means that
-our constraint solving algorithm visits and solves every constraint
-(or, every set of mutually dependent constraints) only once. It
-composes computed solutions to solve other constraints that depend on
-the solved constraints. Importantly, the failure to solve a dependent
-constraint does not result in backtracking. We now describe our
-constraint solving algorithm by demonstrating its performance over the
-example constraints shown in the previous section. A formal account of
-the algorithm can be found in our technical report~\cite{techrep}.
+Although we collect constraints for the entire program during
+elaboration, we solve the constraints in a compositional fashion.
+Concretely, this means that our constraint solving algorithm visits
+and solves every constraint (or, every set of mutually dependent
+constraints) only once. It composes computed solutions to solve other
+constraints that depend on the solved constraints. Importantly, the
+failure to solve a dependent constraint does not result in
+backtracking. We now describe our constraint solving algorithm by
+demonstrating its performance over the example constraints shown in
+the previous section. A formal account of the algorithm can be found
+in the supplement.
 % To simplify the presentation of our algorithm, we assume that are no
@@
 constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
-shown in Fig.~\ref{fig:pair-deps}. Each node (labeled $C_i$) is a set
-of constraints that belong to a strongly connected component in the
-dependency graph ($G_c$), hence are mutually dependent. All
-dependencies, except the self-dependency on $\mathbf{c_{10}}$, are
-type-2 dependencies.
+shown below: %in Fig.~\ref{fig:pair-deps}. 
+\vspace*{-0.08in}
+\begin{figure}[H]
+\includegraphics[scale=0.6]{DepDAG.png}
+\end{figure}
+\vspace*{-0.08in}
+Each node (labeled $C_i$) is a set of constraints that belong to a
+strongly connected component in the dependency graph ($G_c$), hence
+are mutually dependent. All dependencies, except the self-dependency
+on $\mathbf{c_{10}}$, are type-2 dependencies.
 
@@
 simultaneously, whereas constraints in different sets can be solved as
-per any valid topological ordering of the graph. Accordingly, we
-obtain a topological ordering of nodes in the graph $G_{{C}}$, and
-solve the sets of constraints in that order. The solutions obtained
-after solving a constraint set are applied to the constraints in
-subsequent sets before attempting to solve them. For the DAG in
-Fig.~\ref{fig:pair-deps}, we consider the topological order $[C_1,
-C_2, C_3]$, and solve the sets of constraints in that order.
+per any valid topological ordering of the graph's transpose.
+Accordingly, we obtain a topological ordering of nodes in the graph
+$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
+in that order. The solutions obtained after solving a constraint set
+are applied to the constraints in subsequent sets before attempting to
+solve them. For the DAG in figure above, we consider the topological
+order $[C_1, C_2, C_3]$ of its transpose, and solve the sets of
+constraints in that order.
 
@@
 conjoining antecedents and consequents of all constraints in the
-set\footnote{\cite{techrep} formally justifies the soundness of this
-step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
+set\footnote{Supplement contains formal justification for this step}.
+For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
 constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
@@
 $\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
-this case). Such constraints are solved for $\varphi$ by computing a
-maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
-\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
-well-formedness condition on $\varphi$, then it is clearly a solution
-for $\varphi$, because there cannot exist a weaker solution for
-$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
-maximally weak $\phisol$ fails the well-formedness condition, then we
-establish~\cite{techrep} that it is impossible to satisfy the
-constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
-that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
-and check if it satisfies the well-formedness condition on $\varphi$.
-The former problem can be cast as an instance of graph augmentation
-problem, as described below.
-\vspace*{-0.05in}
-\paragraph{Definitions} Given a constraint formula $\phi$, we define
-its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
-vertices ($V(\phi)$) are free region variables and identifiers in
-$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
-in\footnote{We alternate between viewing a constraint formula ($\phi$)
-as a set of constraints and a conjunction of constraints in this
-description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
-\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
-$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
-conjunction of symmetric outlives constraints for the purpose of graph
-encoding. Conversely, given a digraph $G$, we define its constraint
-encoding $\Phi(G)$ in a straightforward manner. We say that a graph
-$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
-\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
-exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
-exist a path between same pair of vertices in $G_1$.
-
-A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
-\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
-graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
-connected as $G(\phicstr)$. The problem of computing such a $G$ is
-equivalent to the problem finding of finding minimum number of edges
-to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
-Efficient algorithms to solve the later problem are known to
-exist~\cite{graph}, and we rely on them for the solution.
-
-Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
-graph augmentation problems, as described above, results in the
-following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
-outlives constraints are replaced with equalities):
+this case). Note that the constraint describes an abduction problem,
+whose solution is the weakest $\phisol$ such that $\isvalid{\phictxt
+\conj \phisol}{\phicstr}$. We solve the problem by casting it as an
+instance of graph augmentation problem~\cite{siam92}. Our intuition is
+based on the observation that a constraint can be viewed as a directed
+graph with region variables as vertices and outlives relationships
+between them as edges (equality is encoded as a conjunction of
+symmetric outlives relationships).  Let $G(\phictxt)$ and
+$G(\phicstr)$ represent such graphs for $\phictxt$ and $\phicstr$,
+respectively. Now, $\isvalid{\phictxt}{\phicstr}$ if and only if every
+pair $(\rho_1,\rho_2)$ of connected vertices in $G(\phicstr)$ are also
+connected in $G(\phictxt)$. If $\rho_1$ and $\rho_2$ are not connected
+in $G(\phictxt)$, then it means that $\isvalid{\phicstr}{\rho_1
+\outlives \rho_2}$ but $\isnotvalid{\phictxt}{\rho_1 \outlives
+\rho_2}$. Since $\phisol$ needs to satisfy $\isvalid{\phictxt \conj
+\phisol}{\phicstr}$, it must be the case that $\rho_1$ and $\rho_2$
+are connected in $G(\phictxt) \cup G(\phisol)$. Since $\phisol$ must
+be weakest such formula, it can be obtained by computing the minimum
+number of edges that need to be added to $G(\phictxt)$ such that it is
+``as connected as'' $G(\phicstr)$. Using this technique, we obtain the
+following solutions to $\varphi_0$ and $\varphi_1$ for constraints in
+the running example:
 \begin{center}
 \(
-  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+  \varphi_0 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
@@
 \(
-  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
+  \varphi_1 \Leftrightarrow \rho_5 = \rho_0
 \)
@@
 \end{center}
-Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
-$F$ is a substitution function, not necessarily idempotent. Let
-$G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
-$\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
-recursive constraints  is based on some observations. Firstly, the set
-of all possible constraints over a finite set of region
-variables/identifiers is a complete meet semi-lattice, where (a).
-meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
-\conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
-\triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
-function $H$ is a monotone over the lattice since $\forall
-\phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
-\Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
-theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
-that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
-the recursive constraint shown above, we first compute the greatest
-fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
-following non-recursive constraint:
-\begin{center}
-\(
-  \isvalid{\phictxt \conj \varphi}{\phi_f}
-\)
-\end{center}
-Which can then be solved via the graph-based approach described
-before.
-
-Let $H(\varphi_2)$ denote the consequent of the recursive constraint
-$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
-\begin{center}
-\(
-  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
-  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
-  \rgn_0 = \rgn_0
-\)
-\end{center}
-Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
-\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
-results in the following solution for $\varphi_2$ that meets its
-well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
-\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
+Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and $F$ is
+a substitution function, not necessarily idempotent. Our solution to the
+constraints of the above form is based on the observation that $H(\varphi) =
+\phicstr \conj F(\varphi)$ is a monotone over the complete meet semi-lattice
+of all possible constraints over the domain of region variables in the
+program. Hence, by Tarski's fixpoint theorem~\cite{tarski}, $H$ has a
+greatest fixed point. To solve the recursive constraint, we first compute the
+greatest fixed point ($\phi_f$) for $H$, and then solve the non-recursive
+constraint $\isvalid{\phictxt \conj \varphi}{\phi_f}$ using the graph
+approach described above. Using this approach, we solve the recursive
+constraint $c_{14}$, to obtain the following solution to $\varphi_2$:
 \begin{center}
@@
 
+% If such $\phisol$ satisfies the
+% well-formedness condition on $\varphi$, then it is clearly a solution
+% for $\varphi$, because there cannot exist a weaker solution for
+% $\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
+% maximally weak $\phisol$ fails the well-formedness condition, then we
+% establish~\cite{techrep} that it is impossible to satisfy the
+% constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
+% that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
+% and check if it satisfies the well-formedness condition on $\varphi$.
+
+% \vspace*{-0.05in}
+% \paragraph{Definitions} Given a constraint formula $\phi$, we define
+% its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
+% vertices ($V(\phi)$) are free region variables and identifiers in
+% $\phi$, and whose edges ($E(\phi)$) denote outlives constraints
+% in\footnote{We alternate between viewing a constraint formula ($\phi$)
+% as a set of constraints and a conjunction of constraints in this
+% description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
+% \outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
+% $(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
+% conjunction of symmetric outlives constraints for the purpose of graph
+% encoding. Conversely, given a digraph $G$, we define its constraint
+% encoding $\Phi(G)$ in a straightforward manner. We say that a graph
+% $G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
+% \subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
+% exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
+% exist a path between same pair of vertices in $G_1$.
+
+% A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
+% \Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
+% graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
+% connected as $G(\phicstr)$. The problem of computing such a $G$ is
+% equivalent to the problem finding of finding minimum number of edges
+% to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
+% Efficient algorithms to solve the later problem are known to
+% exist~\cite{graph}, and we rely on them for the solution.
+
+% Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
+% graph augmentation problems, as described above, results in the
+% following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
+% outlives constraints are replaced with equalities):
+% \begin{center}
+% \(
+%   \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+%      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+%      \rho_2 = \rho_0 \conj \rho_3 = \rho_1
+% \)\\
+% \(
+%   \varphi_2 \Leftrightarrow \rho_5 = \rho_0
+% \)
+% \end{center}
+% Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
+% $c_{14}$ gives us a recursive constraint of the following form:
+% \begin{center}
+% \(
+%   \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
+% \)
+% \end{center}
+% Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
+% $F$ is a substitution function, not necessarily idempotent. Let
+% $G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
+% $\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
+% recursive constraints  is based on some observations. Firstly, the set
+% of all possible constraints over a finite set of region
+% variables/identifiers is a complete meet semi-lattice, where (a).
+% meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
+% \conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
+% \triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
+% function $H$ is a monotone over the lattice since $\forall
+% \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
+% \Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
+% theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
+% that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
+% the recursive constraint shown above, we first compute the greatest
+% fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
+% following non-recursive constraint:
+% \begin{center}
+% \(
+%   \isvalid{\phictxt \conj \varphi}{\phi_f}
+% \)
+% \end{center}
+% Which can then be solved via the graph-based approach described
+% before.
+
+% Let $H(\varphi_2)$ denote the consequent of the recursive constraint
+% $c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
+% \begin{center}
+% \(
+%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+%   \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
+%   \rgn_0 = \rgn_0
+% \)
+% \end{center}
+% Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
+% \conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
+% results in the following solution for $\varphi_2$ that meets its
+% well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
+% \rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
+% \begin{center}
+% \(
+%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+%   \rho_9 \conj \rho_8 = \rho_9
+% \)
+% \end{center}
+
 \subsection{Higher-Order Type Inference}
 
-We have so far ignored function types in the context of type
-inference. The region type system for $\FB$ admits function types with
-higher-rank region polymorphism, allowing region variables to be
-quantified in the types of higher-order arguments. The type system
-clearly does not have the principal typing property, as demonstrated
-by the \C{app2} method below:
-\begin{codejava}
-bool app2(A x, A y, (A $\rightarrow$ bool) f) 
-    { return (f x && f y); }
-\end{codejava}
-There are two possibilities for the region type of \C{app2}: one,
-where it allows \C{x} and \C{y} to be allocated in different regions,
-but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
-arguments from any region), and second, where it requires \C{x} and
-\C{y} to be allocated in the same region, and allows \C{f} to be
-monomorphic. None of the two types are more general than the other,
-and it is not clear which one is more preferable. The concrete
-manifestation of this dilemma is a constraint of form 
-$\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
-no additional constraints to help determine $\varphi_1$ or $\varphi_2$
-before solving this constraint. Since it is always possible to
-strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
-does not exist unique solution to such constraints. Nonetheless, it is
-possible to bias the constraint-solving algorithm to consistently
-prefer one solution over the other.
+% The region type system for $\FB$ admits function types with
+% higher-rank region polymorphism, allowing region variables to be
+% quantified in the types of higher-order arguments. This allows, for
+% example, a user-defined function to be applied on arguments from
+% different regions in different contexts. However, like most type
+% systems, our region type system loses principal typing property in
+% presence of higher-rank polymorphism, which means that there can be
+% two possible types for a function, where neither one is more general
+% than the other. Consequently, it is not possible to infer a unique
+% type for a function. Nonetheless, it is possible to engineer type
+% inference to to consistently prefer one type over the other.
+Our type inference algorithm does not generalize to higher-order case,
+due to the lack of principal typing in presence of higher-rank
+polymorphism. We therefore adopt a trivial approach of assigning the
+same region to the arguments and return values in (higher-order)
+function types. For user-defined functions (e.g., predicate in the
+\C{WHERE} clause of the \C{SELECT} dataflow query) and lambda
+arguments to the $\RgnZ$ class constructor, which are the prominent
+uses of higher-order functions in \name, the simplistic approach seems
+to suffice in practice. However, it is always possible that this
+approach makes type checker reject a program that is otherwise valid.
+In such cases, programmer can help the type checker by writing region
+annotations for only the higher-order functions.
 
-In context of \name, however, we adopt a trivial solution of
-assigning the same region to the arguments and return values in
-(higher-order) function types. For user-defined functions (e.g.,
-predicate in the \C{WHERE} clause of the \C{SELECT} dataflow query)
-and lambda arguments to the $\RgnZ$ class constructor, which are the
-prominent uses of higher-order functions in \name, the simplistic
-approach seems to suffice in practice. However, it is always possible
-that this approach makes type checker reject a program that is
-otherwise valid. In such cases, programmer can help the type checker
-by writing region annotations for only the higher-order functions.
+% \begin{codejava}
+% bool app2(A x, A y, (A $\rightarrow$ bool) f) 
+%     { return (f x && f y); }
+% \end{codejava}
+% There are two possibilities for the region type of \C{app2}: one,
+% where it allows \C{x} and \C{y} to be allocated in different regions,
+% but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
+% arguments from any region), and second, where it requires \C{x} and
+% \C{y} to be allocated in the same region, and allows \C{f} to be
+% monomorphic. None of the two types are more general than the other,
+% and it is not clear which one is more preferable. The concrete
+% manifestation of this dilemma is a constraint of form 
+% $\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
+% no additional constraints to help determine $\varphi_1$ or $\varphi_2$
+% before solving this constraint. Since it is always possible to
+% strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
+% does not exist unique solution to such constraints. Nonetheless, it is
+% possible to bias the constraint-solving algorithm to consistently
+% prefer one solution over the other.
+
+% In context of \name, however, we adopt a trivial solution of
+% assigning the same region to the arguments and return values in
+% (higher-order) function types. For user-defined functions (e.g.,
+% predicate in the \C{WHERE} clause of the \C{SELECT} dataflow query)
+% and lambda arguments to the $\RgnZ$ class constructor, which are the
+% prominent uses of higher-order functions in \name, the simplistic
+% approach seems to suffice in practice. However, it is always possible
+% that this approach makes type checker reject a program that is
+% otherwise valid. In such cases, programmer can help the type checker
+% by writing region annotations for only the higher-order functions.
 

### 2015-11-18 paper/type-system.tex
Sec 4. done.

@@
 \end{itemize}
-Like the judgments in FGJ~\ref{fgj}, all the judgments defined by the
+Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
 rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
@@
 
-To help state the type safety theorem, we extend \FB's expressions
-with $\invalidexn$, and define the syntactic class of values:
+To help state the type safety theorem, we define the syntactic class of values:
 \begin{smathpar}

### 2015-11-19 paper/implementation.tex
Edits.

\section{Implementation and Evaluation}
\label{sec:implementation}

We have implemented the prototype of \name compiler frontend,
including its region type system and type inference, in 3k+ lines of
OCaml. The input to our system is a program in $\absof{\FB^+}$, an extended
version of $\absof{\FB}$ that includes assignments, conditionals,
loops, more primitive datatypes (\eg strings), and a null value. 
% If needed, a specification file containing region type annotations
% for higher-order functions can also be provided. 
Our implementation of region type inference closely follows the
description given in Sec.~\ref{sec:type-inference}. To solve the
constraints that arise during type inference, we built a solver called
\csolve that implements constraint solving approach based on fixpoint
computation and graph augmentation described in \S~\ref{sec:csolve}. 
% Since graph augmentation is NP-hard, \csolve uses an approximation
% algorithm that repeatedly adds an edge and recomputes paths in graph
% $G_1$ until it is as connected as graph $G_2$. Though this approach
% may not necessarily return the weakest solution to the abduction
% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
% constraints are not often complex. 
If the input $\absof{\FB^+}$ program does not create any unsafe
references, our compiler annotates it with region types, which act as
a witness to program's memory safety. On contrary, if the input
program does create a reference that is potentially unsafe, then the
type inference fails during the constraint solving phase. We currently
do not implement error localization and feedback mechanisms.

To evaluate the practical utility of our region type system and type
inference, we 
% implemented some of the C\# libraries in $\absof{\FB^+}$, and 
translated some of the Naiad streaming query operator benchmarks
(Naiad vertices) used in~\cite{Broom:HotOS} to $\absof{\FB^+}$, and
used our prototype compiler to assign region types to these programs,
and thus prove their safety. During the process, we found some
instances of potential memory safety violations in the $\absof{FB^+}$
translation of benchmarks, which we verified to be present in the
original C\# implementation as well. The cause of all safety
violations is the creation of a reference from the outgoing message (a
transferable region) to the payload of the incoming message. For
example, the implementation of \C{RegionSelectVertex} contains the
following:
\begin{codejava}
  if (this.selectFn(inMsg.payload[i])) {
    outMsg.set(outputOffset, inMsg.payload[i]);
    ...
  } 
\end{codejava}
The \C{outMsg} is later transferred to a downstream actor, where the
reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
reference could have gone unnoticed during experiments
in~\cite{Broom:HotOS} because their experimental setup included only
one actor.}. We eliminated such unsafe references by creating a clone
of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
subsequently able to certify the safety of all references.

% We now briefly describe our experience of
% working with these benchmarks. Note that the performance benefits of
% using transferable regions in dataflow systems have been established
% in~\cite{Broom:HotOS}. Our current evaluation only focuses on safety
% aspects.

% \paragraph{Memory safety violation}

### 2015-11-19 paper/related.tex
Edits.

@@
 Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
-concept of a region type system to statically ensure the safety of
-region-based memory management in ML. Following their seminal work,
-static type systems for safe region-based memory management have been
-extensively studied in the context of various languages and problem
-settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
-WW01, rust, gpu14}. Our work differs from the existing proposals in a
-number of ways. Firstly, our problem setting includes lexically scoped
-stack regions and dynamic transferable regions (both
-programmer-managed) in context of an object-oriented programming
-language equipped with higher-order functions. Second, we adopt a
-two-pronged approach to memory safety that relies on a combination of
-a simple static type discipline and lightweight runtime checks. 
+concept of a region type system to statically enforce the safety of
+their region-based memory management scheme in ML. Following their
+seminal work, static type systems for safe region-based memory
+management have been extensively studied in the context of various
+languages and problem settings~\cite{cyclone02, cyclone04, yates99,
+MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
+existing proposals in a number of ways. Firstly, our problem setting
+includes lexically scoped stack regions and dynamic transferable
+regions (both programmer-managed) in context of an object-oriented
+programming language equipped with higher-order functions. Second, we
+adopt a two-pronged approach to memory safety that relies on a
+combination of a simple static type discipline and lightweight runtime
+checks. 
 % The type system reduces the problem of ensuring safety of all memory
@@
 in order to guarantee safety. Lastly, our region type system comes
-equipped with full type inference that completely eliminates the need
-to write region annotations on types to convince the type checker that
-the program is safe.
+equipped with full type inference that more or less eliminates the
+need to write region annotations on types to convince the type checker
+that the program is safe.
 
@@
 object-oriented program that already uses (stack and dynamic) regions.
-Similar to their region inference algorithm, our region type inference
-algorithm can deal with region-polymorphic recursion by computing
-fixed points for recursive constraints. While their inference
-algorithm only ever generates equality constraints, which can be
-solved via unification, our type inference algorithm also generates
-partial order outlives constraints, which are required to capture
-subtle relationships between lifetimes of transferable regions and
-stack regions. Consequently, our constraint solving algorithm is more
-sophisticated, and is capable of inferring unknown outlives
+% Similar to their region inference algorithm, our region type inference
+% algorithm can deal with region-polymorphic recursion by computing
+% fixed points for recursive constraints. 
+Our type inference is analogous to their region inference. While their
+inference algorithm only ever generates equality constraints, which
+can be solved via unification, our type inference algorithm also
+generates partial order outlives constraints, which are required to
+capture subtle relationships between lifetimes of transferable regions
+and stack regions. Consequently, our constraint solving algorithm is
+more sophisticated, and is capable of inferring unknown outlives
 constraints over region arguments of polymorphic recursive functions.
@@
 An ownership type system for safe region-based memory management in
-real-time Java has been proposed by~\cite{MIT03}. Like us, they too
-assume a source language with programmer-managed memory regions, and
-focus on proving safety of programs written in that language. Their
-source language admits various kinds of regions in order to support
-shared-memory concurrency. However, all their regions have
-lexically-scoped lifetimes. In contrast, we admit regions with
+real-time Java has been proposed by~\cite{MIT03}.  
+% Like us, they too assume a source language with programmer-managed
+% memory regions, and focus on proving safety of programs written in
+% that language.  
+Their source language admits various kinds of regions with
+lexically-scoped lifetimes (i.e., stack regions) in order to support
+shared-memory concurrency. In contrast, we admit regions with
 dynamically determined lifetimes in order to support message-passing
@@
 However, our language also admits parametric polymorphism (generics)
-and higher-order functions, whose interaction is non-trivial in
-context of a region type system. On the other hand, our type system
-eschews the notions of region kinds and ownership, leading to a more
-succinct formalization. Furthermore, we establish a type safety result
-that formalizes its guarantees with respect to a well-defined
-operational semantics. Like Cyclone, ~\cite{MIT03}'s language is
-explicitly typed. Although there is some support for local type
-inference, region types for methods and classes need to be written
-explicitly. In contrast, we support full type inference that
-eliminates any such need.
+and higher-order functions.
+%, whose interaction is non-trivial in context of a region type system. 
+% On the other hand, our type system eschews the notions of region kinds
+% and ownership, leading to a more succinct formalization. 
+Furthermore, we establish type safety and transfer safety results that
+formalize the guarantees provided by our system. While ~\cite{MIT03}'s
+language is explicitly typed, our language comes equipped with full
+type inference.
+% Although there is some support for local
+% type inference, region types for methods and classes need to be
+% written explicitly. In contrast, our region type inference that
+% eliminates any such need.
 

### 2015-11-19 paper/introduction.tex
updated intro

@@
 big-data analysis systems, garbage collection overhead becomes
-significant. Importantly, much of this overhead results from the
-collector performing avoidable or unproductive
-work~\cite{Broom:HotOS}. For example, a consumer process may trigger a
-GC as soon as it receives a large data-structure from a producer
-process, which has just finished its GC. In such case, consumer's GC
-traverses the entire data-structure performing little or no work.
-Another example is the process executing the code in
-Fig.~\ref{fig:motivating-eg}, whose GC might repeatedly traverse the
-\C{map} data-structure, although its objects cannot be collected until
-a suitable timing message arrives.
+significant~\cite{Broom:HotOS}.
+Furthermore, in a distributed dataflow system, the GC pause at one node can
+have a cascading adverse effect on the performance of other nodes.
+For instance, a GC pause at an upstream actor can block
+downstream actors that are waiting for messages. 
+Much of the GC overhead results from the collector performing avoidable or unproductive work.
+For example, in the process executing the code from
+Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the \C{map}
+data-structure, although its objects cannot be collected until a
+suitable timing message arrives.
+
+% actors.
+% performance degradation due to
+% GC overhead becomes aggravated in the distributed 
+% For example, a consumer process may trigger a
+% GC as soon as it receives a large data-structure from a producer
+% process, which has just finished its GC. In such case, consumer's GC
+% traverses the entire data-structure performing little or no work.
+
 % as the collector scans these large data-structures For instance, in
@@
 described above, is that the data-structures exchanged between them
-have can be partitioned into sets of fate-sharing objects with common
+can be partitioned into sets of fate-sharing objects with common
 lifetimes, which makes them good candidates for a region-based memory

### 2015-11-19 paper/abstract.tex
Edits.

@@
 inference algorithm that infers principal region types for first-order
-programs. This eliminates the need for programmers to write region
-annotations on types, while facilitating the reuse of existing C\#
-libraries with no modifications. Various case studies demonstrate the
-practical utility of our approach.
+programs, and useful types for higher-order programs. This eliminates
+the need for programmers to write region annotations on types, while
+facilitating the reuse of existing C\# libraries with no
+modifications. Experiments demonstrate the practical utility of our
+approach.
 

### 2015-11-19 paper/related.tex
Edits.

@@
 calculus with first-class regions with dynamic lifetimes, and imposes
-linear typing to control accesses to regions. All type annotations
-need to be written explicitly. Moreover, invariants needed to ensure
-safety under region transfers, such as the absence of references that
-escape a transferable region, cannot be enforced under~\cite{WW01}'s
-type system.
+linear typing to control accesses to regions. Our open/close lexical
+block for transferable regions traces its orgins to the \C{let!}
+expression in~\cite{WW01} and~\cite{wadler90} that safely relaxes
+linear typing restrictions, allowing variables to be temporarily
+aliased. We don't have linear typing, thus admit unrestricted
+aliasing. Non-linear typing keeps our type system simple, and makes it
+possible to infer types, thus eliminating the annotation burden.
+Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
+the invariants needed to ensure safety under region transfers, such as
+the absence of references that escape a transferable region.
 

### 2015-11-19 paper/type-inference.tex
Edits.

@@
 \subsection{Constraint Solving}
-\label{sec:fb-constraintsolving}
+\label{sec:csolve}
 
@@
 % inference to to consistently prefer one type over the other.
-Our type inference algorithm does not generalize to higher-order case,
-due to the lack of principal typing in presence of higher-rank
-polymorphism. We therefore adopt a trivial approach of assigning the
-same region to the arguments and return values in (higher-order)
-function types. For user-defined functions (e.g., predicate in the
-\C{WHERE} clause of the \C{SELECT} dataflow query) and lambda
-arguments to the $\RgnZ$ class constructor, which are the prominent
-uses of higher-order functions in \name, the simplistic approach seems
-to suffice in practice. However, it is always possible that this
-approach makes type checker reject a program that is otherwise valid.
-In such cases, programmer can help the type checker by writing region
-annotations for only the higher-order functions.
+Our type inference algorithm does not straightforwardly generalize to
+the higher-order case, due to the lack of principal typing in presence of
+higher-rank polymorphism. Fortunately, higher-order type inference, in
+its full generality, is rarely needed in the context of \name, where
+higher-order control flow is predominantly due to the second-order
+arguments to constructors (For \eg, user-defined functions in dataflow
+queries, and lambda arguments to the $\RgnZ$ class constructor). We
+therefore extend our inference algorithm with a heuristics tailor-made
+for the aforementioned second-order cases. The simplistic approach
+seems to suffice in our experience, but, it is always possible that it
+makes the type checker reject a program that is otherwise valid. In
+such cases, programmer can help the type checker by writing region
+annotations only over higher-order arguments.  Nonetheless, we note
+that it is possible to infer most general (if not principal) types for
+higher-order arguments, by extending our constraint solver with an
+approach modeled after~\cite{gulwani09}, which lets it compute maximally
+weak/strong solutions to multiple competing $\varphi$'s
+simultaneously.
 

### 2015-11-20 paper/broom.tex
Progress lemma.
Progress.
Operational Semantics Updated.
Preservation done.
Proofs proofs.
updates to section 3
Edits.
Made "fresh" fix to the opsem.
Edits to Sec.1
Some edits to Sec.3.
Sec 3 full pass.
Edits to Sec.4. Formatting changes.

@@
 
-\documentclass[10pt,pldi]{sigplanconf}
+%\documentclass[10pt,pldi]{sigplanconf}
+\documentclass[pldi]{sigplanconf-pldi16}
 
@@
 \newcommand{\mang}{\inang{\rhoalloc_m,\bar{\rho_m} \,|\, \phi_m}}
-\newcommand{\rhoset}{\rhoenv}
+\newcommand{\rhoset}{\Sigma}
+\newcommand{\rhomap}{\Sigma}
 \newcommand{\rhoenv}{\Delta}
@@
 \newcommand{\phisol}{\phi_{sol}}
-\newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
+\newcommand{\subtypcx}{\rhoset, \rhoenv,\aenv,\phicx}
 \newcommand{\A}{{\mathcal{A}}}
@@
 \newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
-\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\;#2\;\C{withroot}\;#3\;\C{in}\;#4}
+\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\;#3@#2\;\C{in}\;#4}
 \newcommand{\openalloc}[3]{\C{openalloc}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
 \newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
-\newcommand{\invalidexn}{\C{InvalidOpnException}}
+\newcommand{\invalidexn}{\bot}
 \newcommand{\toprgn}{\rgn_{\top}}
@@
 \newcommand{\frv}{{\sf frv}}
+\newcommand{\rv}{{\sf RVars}}
 \newcommand{\solve}{{\sf solve}}
+\newcommand{\valuee}{{\sf value}}
 \newcommand{\elabClass}{{\sf elabClass}}
@@
 \newcommand{\elabClassTable}{{\sf elabClassTable}}
+\newcommand{\CLOSED}{\mathtt{C}}
+\newcommand{\OPEN}{\mathtt{O}}
+\newcommand{\XFERRED}{\mathtt{X}}
+\newcommand{\transfer}{{\sf transfer}}
 \newcommand{\CSP}{(C\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi}, \bar{\rhoenv})}
@@
 \newcommand{\csid}[1]{\lbrack \mathbf{c_{#1}} \rbrack:\;}
+\newcommand{\qqquad}{\quad\quad\quad}
+\newcommand{\emptyA}[0]{(\rhomap,\rhoenv,\cdot,\phicx)}
+\newcommand{\emptyASigp}[0]{(\rhomap',\rhoenv,\cdot,\phicx)}
+\newcommand{\emptyASigpp}[0]{(dom(\rhomap''),\rhoenv,\cdot,\phicx)}
+\newcommand{\emptyADelcup}[1]{(\rhomap,\rhoenv \cup \{#1\},\cdot,\phicx)}
+\newcommand{\emptyASigpDelcup}[1]{(\rhomap',\rhoenv \cup \{#1\},
+                                    \cdot,\phicx)}
+\newcommand{\emptyASigxDelcup}[2]{(dom(\rhomap[#1]),\rhoenv \cup \{#2\},
+                                    \cdot,\phicx)}
+\newcommand{\emptyADelcupPhicap}[2]{(\rhomap,\rhoenv \cup \{#1\},
+                                    \cdot,\phicx \conj \rhoenv \outlives #2)}
+\newcommand{\emptyADelcupPhicapp}[2]{(\rhomap,\rhoenv \cup \{#1\},
+                                    \cdot,\phicx \conj #2)}
+\newcommand{\empTyADelcup}[1]{(\rhomap,\rhoenv \cup \{#1\},\cdot,\phicx)}
+\newcommand{\empTyADelcupPhisubst}[2]{(\rhomap,\rhoenv \cup \{#1\},\cdot,[#2]\phicx)}
+\newcommand{\vacantA}[1]{(\emptyset,\{#1\},\cdot,true)}
+\newcommand{\RgnZT}[1]{\RgnZ\inang{T}\inang{#1}}
+\newcommand{\BZ}[0]{B\inang{\tbar}}
+\newcommand{\BZT}[1]{B\inang{\tbar}\inang{#1}}
+\newcommand{\redstoo}[2]{\redsto{\rhoenv}{#1}{#2}}
+\newcommand{\redstocup}[3]{\redsto{\rhoenv \cup \{#1\}}{#2}{#3}}
 
@@
 basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
-keywordstyle=\color{Bittersweet},
+keywordstyle=\color{Bittersweet}\bf,
 % numbers=left,                   % where to put the line-numbers
@@
 %escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={var, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
+morekeywords={var, foreach, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
 }}
@@
 
-\authorinfo{}
-           {}
-           {}
+%\authorinfo{} {} {}
 \maketitle
@@
 
+\input{implementation}
+
 \input{related}

### 2015-11-20 paper/implementation.tex
Edits.

@@
 used our prototype compiler to assign region types to these programs,
-and thus prove their safety. During the process, we found some
+and thus prove their safety. During the process, we found multiple
 instances of potential memory safety violations in the $\absof{FB^+}$

### 2015-11-20 paper/introduction.tex
Edits.

@@
 big-data analysis systems, garbage collection overhead becomes
-significant~\cite{Broom:HotOS}.
-Furthermore, in a distributed dataflow system, the GC pause at one node can
-have a cascading adverse effect on the performance of other nodes.
-For instance, a GC pause at an upstream actor can block
-downstream actors that are waiting for messages. 
-Much of the GC overhead results from the collector performing avoidable or unproductive work.
+significant~\cite{Broom:HotOS}. This is a cause of concern because,
+firstly, in a distributed dataflow system, GC overhead at one node can
+have compounding adverse effect on the performance of other nodes.
+This can happen, for instance, when a GC is triggered at an upstream
+actor that is about to send a message, while the downstream actors are
+blocked waiting for messages. Secondly, much of the GC overhead
+results from the collector performing avoidable or unproductive work.
 For example, in the process executing the code from

### 2015-11-20 paper/related.tex
Edits.

@@
 in order to guarantee safety. Lastly, our region type system comes
-equipped with full type inference that more or less eliminates the
-need to write region annotations on types to convince the type checker
-that the program is safe.
+equipped with full type inference that eliminates the need to write
+region annotations on types to convince the type checker that the
+program is safe.
 
@@
 functions. Later proposal~\cite{cyclone04} extends Cyclone with
-dynamic regions, but has to impose unique pointer restrictions to
-guarantee safety. Objects that need to be shared within data
-structures are allowed to have multiple references, and safety for
-such objects was ensured via runtime reference counting.
+dynamic regions, imposes unique pointer restrictions to guarantee
+safety. Objects that need to be shared within data structures are
+allowed to have multiple references, and safety for such objects was
+ensured via runtime reference counting.
 
@@
 % and ownership, leading to a more succinct formalization. 
-Furthermore, we establish type safety and transfer safety results that
-formalize the guarantees provided by our system. While ~\cite{MIT03}'s
+Moreover, we establish type safety and transfer safety results that
+formalize the guarantees provided by our system. While~\cite{MIT03}'s
 language is explicitly typed, our language comes equipped with full
@@
 block for transferable regions traces its orgins to the \C{let!}
-expression in~\cite{WW01} and~\cite{wadler90} that safely relaxes
+expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes
 linear typing restrictions, allowing variables to be temporarily

### 2015-11-20 paper/type-inference.tex
Edits.

@@
 (i.e., \FB without region types, but with \C{letregion} and \C{open}
-expressions) to programs in \FB. The key aspects of this elaboration
-are described below.
+expressions, similar to the language introduced in
+\S~\ref{sec:overview}) to programs in \FB. The key aspects of this
+elaboration are described below.
 

### 2015-11-20 paper/type-system.tex
Edits.

@@
 and apply functions\footnote{We distinguish between functions and
-methods. The formed result from lambda expressions, whereas the latter
+methods. The former result from lambda expressions, whereas the latter
 come from class definitions}. Functions, like methods, exhibit
@@
 ($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
-was declared to extend.  The map $\tyvar \mapsto \fgjN$ is denoted
-$\Delta$ in FGJ formalism}, and
+was declared to extend.}, and
 \item A constraint formula ($\phicx$) that captures the outlives

### 2015-11-20 paper/introduction.tex
Merged paper to techrep.

@@
 described above, is that the data-structures exchanged between them
-can be partitioned into sets of fate-sharing objects with common
+have can be partitioned into sets of fate-sharing objects with common
 lifetimes, which makes them good candidates for a region-based memory

### 2015-11-20 paper/overview.tex
Merged paper to techrep.

@@
-\section{Overview of \name}
-\label{sec:overview}
+\newcommand{\COMMENT}[1]{}
 
-Stripped of its region-specific constructs, the core language of \name
-is a familar subset of \csharp that includes features such as
-class-based inheritance, dynamic dispatch, parametric polymorphism,
-and lambdas.  We therefore elide a discussion of the core language,
-focusing instead on the concepts specific to region-based memory
-management in \name.
+\section{An Informal Overview of \name} \label{sec:overview}
 
-\subsection{Stack Regions and Outlives Relation}
+\name enriches a simple object-oriented language (supporting
+parametric polymorphism and lambdas) with a set of region-specific
+constructs.  In this section, we present an informal overview of these
+region-specific constructs.
 
-Simplest of the region-specific constructs in \name is a \C{letregion}
-lexical block that creates a new memory region adressable via its
-static identifier. The semantics of the \C{letregion} block is similar
-to Tofte and Talpin~\cite{ttpopl94}'s \C{letregion} expression; the
-newly introduced region is live only inside the block, where its
-identifier is in scope for \C{new} expressions to make allocations
-inside the region. For example, in the following code, object \C{a} is
-allocated in the region with identifier \C{R0} introduced by the
-\C{letregion}:
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    Object a = new@R0 Object();
-  }
-\end{codejava}
-\end{center}
-The \C{letregion} blocks can be nested, leading to a stack of regions
-that are deallocated in the reverse order in which they are allocated.
-Following ~\cite{cyclonepldi02}, we therefore call regions introduced
-by \C{letregion} expressions as \emph{stack regions}. The stack
-discipline induces an \emph{outlives} relationship among regions created
-by nested \C{letregion}s, where the region introduced by the outer
-\C{letregion} is guaranteed to outlive the one introduced by the inner
-\C{letregion}. It is therefore safe to refer to an object allocated in
-outer region from the inner region, but the converse is not true. For
-example, consider the following code (assume that class \C{A} has
-field \C{x} of type \C{Object}):
-\begin{center}
-\begin{codejava}
-  letregion R0 {
-    A a0 = new@R0 A();
-    letregion R1 {
-      A a1 = new@R1 A();
-      a1.x = new@R0 Object(); // safe & legal
-      a0.x = new@R1 Object(); // unsafe & illegal
-      ...
-    }
-    ...
-  }
-\end{codejava}
-\end{center}
-The code creates two stack regions with identifiers \C{R0} and \C{R1},
-where the region \C{R0} outlives the region \C{R1} (denoted as $\C{R0}
-\outlives \C{R1}$).  Objects \C{a0} and \C{a1} are allocated in regions
-\C{R0} and \C{R1}, respectively. The first assignment statement
-assigns to \C{a1.x} an object allocated in outer region (\C{R0}). This
-assignment is safe as \C{a1.x} refers to a longer living object, hence
-is guaranteed to be a valid reference throughout the lifetime of
-\C{a1}.  In contrast, the second assignment is unsafe, as it assigns
-to \C{a0.x} an object, whose lifetime is shorter than the lifetime of
-\C{a0}, making it unsafe to dereference \C{a0.x} outside the inner
-block. 
-% Unsafe assignments can also happen indirectly via a function call..
-Preempting such unsafe assignments is the \emph{raison d'etre} of the
-region type system.
+\subsection{Using Regions in \name}
+\label{sec:alloc-ctxt}
 
-\paragraph{A Note on region identifiers} Every stack region in \name
+\paragraph{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
+creates a new stack region, with a static identifier \C{R}, whose
+scope is restricted to the statement \C{S}. The semantics of
+\C{letregion} is similar to Tofte and Talpin~\cite{tofte94}'s
+\C{letregion} expression: objects can be allocated by \C{S} in the
+newly created region while \C{R} is in scope, but the region and all
+objects allocated within it are freed at the end of \C{S}.
+
+\paragraph{Object Allocation} The ``\C{new@R T()}'' construct creates
+a new object of type \C{T} in the region \C{R}. The specification of
+the allocation region \C{R} in this construct is optional.  At
+runtime, \name maintains a stack of \emph{active} regions, and we
+refer to the region at the top of the stack as the \emph{allocation
+context}. The statement \C{new T()} allocates the newly created object
+in the current allocation context.
+%
+This is important as it enables \name applications to use existing
+region-oblivious C\# libraries. In particular, given a  C\# library
+function \C{f} (that makes no use of \name's region constructs), the
+statement ``\C{letregion R \{ f(); \}}'' invokes \C{f}, but has the
+effect that all objects allocated by this invocation are allocated in
+the new region \C{R}.
+
+\paragraph{Region Identifiers} Every live region in \name
 is associated with a static identifier that uniquely identifies the
@@
 %\begin{minipage}{1.55in}
-\begin{numcodejava}
+\begin{codejava}
 for (int i=0; i<=10; i++) {
@@
       A a1 = new@R1 A();
-      ...
-    }
-  }
-}
-\end{numcodejava}
-%\end{minipage}
-%\begin{minipage}{1.4in}
-%\begin{codejava}
-%void m(int i) {
-%  letregion R0 {
-%    A a = new@R0 A();
-%  }
-%  this.m(++i);
-%}
-%...
-%this.m(1);
-%\end{codejava}
-%\end{minipage}
+      ...  } } }
+\end{codejava}
 The identifiers \C{R0} and \C{R1} are bound to new stack regions each
@@
 
-% Interestingly, it is possible for multiple stack regions introduced by
-% evaluating a the same \C{letregion} expression to be in outlives
-% relationship. For instance, in the following example:
-% \begin{codejava}
-% void m(A a) {
-%   letregion R0 {
-%     B b = new@R0 B();
-%     b.x = a;
-%     this.m(new@R0 A());
-%   }
-% }
-% \end{codejava}
-% Stack regions identified by \C{R0} introduced in successive recursive
-% calls are in outlives relationship. Consequently, the assignment to
-% \C{b.x} should be considered safe. Region type system has to allow for
-% this possibility.
+\paragraph{Transferable Regions} \name's \emph{transferable regions}
+are an encapsulation of a data-structure that can be transferred
+between autonomous entities (\eg, between two concurrently executing
+threads or actors).  Hence, unlike stack regions, transferable regions
+are not constrained to have a lexically scoped lifetime.  (Hence, we
+also refer to them as \emph{dynamic} regions.)
+
+Furthermore, transferable regions, unlike stack regions, are first
+class values of \name: they are objects of the class \C{Region}, they
+are created using the \C{new} keyword, and can be passed as arguments,
+stored in data structures, and returned from methods.  A transferable
+region is intended to encapsulate a single data-structure, consisting
+of a collection of objects with a distinguished root object of some
+type \C{T}, which we refer to as the region's \emph{root} object.  The
+class \C{Region} is parametric over the type \C{T} of this root
+object.
+
+The \C{Region} constructor takes as a parameter a function that
+constructs the root object: it creates a new region and invokes this
+function, with the new region as the allocation context, to create the
+root object of the region. The following code illustrates the
+creation of a transferable region, whose root is an object of type
+\C{A}.
+% (\C{r}) of type \C{T}:
+\begin{codejava} 
+  Region<A> rgn = new Region<T>(() => new A())
+\end{codejava} 
+In the above code, \C{rgn} is called the \emph{handler} to the newly
+created region, and is required to read the contents of the region, or
+change its state. The class \C{Region} offers two methods: a \C{free}
+method that deallocates the region (and all the objects allocated
+within it), and \C{transfer} method that transfers the region to a
+consumer process. It is an abstraction of two possible forms of
+transfer: a transfer between two processes in a shared memory setting
+or a transfer between two processes in a distributed, message-passing,
+setting. The precise semantics of \C{transfer} are unimportant in the
+context of the region type system and we will not discuss them
+further.
+
+% \begin{figure} \begin{codejava} public class Region<T> { public
+% Region(Func<T> mkRoot); public void free(); public void transfer();
+% } \end{codejava} \caption{The type signature of the \C{Region}
+% class} \label{fig:region-class} \end{figure}
+
+% The type signature of the class \C{Region} appears in
+% Fig.~\ref{fig:region-class}.  The method \C{free} deallocates the
+% region (including all objects allocated within it).  The method
+% \C{transfer} transfers the region to a downstream actor as
+% determined by the run-time.  It is an abstraction of two possible
+% forms of  transfer: a transfer between two actors in a shared memory
+% setting or a transfer between two actors in a distributed,
+% message-passing, setting.  The precise semantics of \C{transfer} are
+% unimportant in the context of the region type system and we will not
+% discuss them further.
+
+\paragraph{Open and Closed Regions} A transferable region must be
+explicitly \emph{opened} using \name's \C{open} construct in order to
+either read or update or allocate objects in the region.
+Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
+following: (a). It opens the transferable region handled by \C{rgn}
+for allocation (i.e., makes it the current allocation context), (b).
+binds the identifier \C{R} to this open region, and (c). initializes
+the newly introduced local variable \C{v} to refer to the root object
+of the region.  The \C{@R} part of the statement is optional and may
+be omitted.  The \C{open} construct is intended to simplify the
+problem of ensuring memory safety, as will be explained soon.  We
+refer to a transferable region that has not been opened as a
+\emph{closed} region.
+
+\paragraph{Motivating Example}
+% We now illustrate how the features of \name can be used to 
+Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
+of Fig.~\ref{fig:motivating-eg} can be written in \name.  The
+\C{onReceive} method receives its input message in a
+\emph{transferred} region (\ie, a \emph{closed} region whose ownership
+is transferred to the recipient).  Line 7 creates a new region to
+store the output for time \C{t}, initializing it to contain an empty
+list.  Line 9 opens the input region to process it\footnote{We omit
+\C{@R} annotation in \C{open} when we don't need \C{R}.}.  Line 10
+creates a stack region \C{R0}.  Thus, the temporary objects created by
+the iteration in line 11, for example, will be allocated in this stack
+region that lives just long enough.  We open the desired output region
+in line 12, so that the new output objects created by the invocation
+of \C{selector} in line 13 are allocated in the output region.
+Finally, the input region is freed in line 15. The output region at
+\C{map[t]} stays as along as input messages with timestamp \C{t} keep
+arriving. When the timing message for \C{t} arrives, the \C{onNotify}
+method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
+% After the transfer, \C{SelectVertex} is no longer allowed to access
+% the \C{outRgn}, facilitating its deallocation.
+
+\input{motivating-eg-in-broom}
+
+\subsection{Memory Safety} \label{sec:memory-safety}
+
+Our goal is a type system that can ensure the memory safety of
+programs that use the region constructs described above.  The key to
+memory safety in \name is the following restriction: an object $o_1$
+in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
+a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
+similar restriction applies in the case where $o_1$ is a
+stack-allocated variable.)
+
+Enforcing this restriction is simple in the case of stack regions
+since the outlives relation between stack regions can be inferred from
+their lexical nesting. Unfortunately, inferring outlives relations
+in presence of transferable regions is not easy.  \name imposes the following
+protocol on the use of transferable regions to help simplify this
+check.
+
+A transferable region (that has not been freed or transferred) can be
+in one of two possible states, \emph{open} or \emph{closed}. A newly
+created region is in the closed state.  A region must be opened, using
+the open construct (as explained previously), in order to read or
+update or allocate an object within that region.  An open region
+cannot be freed or transferred.  In particular, an open region is
+guaranteed to be live for the entire duration of the open construct.
+This allows the type system to infer a valid outlives relation between
+the opened region and any stack region that is nested within the open
+construct.
+
+\begin{figure} 
+\includegraphics[scale=0.45]{region-fsm.png}
+\caption{The lifetime of a dynamic (transferable) region in \name}
+\label{fig:region-fsm} 
+\vspace*{-0.25in} 
+\end{figure}
+
+The protocol for transferable regions is presented as a finite state
+machine in Fig.~\ref{fig:region-fsm}.  
+% A transferable region starts
+% its lifetime in a \emph{closed} state, when it is created.
+%
+The safety of memory accesses in \name is now subject to the condition
+that every transferable region correctly follows the state transition
+discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
+region type system statically guarantees the safety of all memory
+accesses.
+% In other words, the type system reduces the problem of ensuring
+% memory safety in \name programs to the problem of enforcing the
+% state transition discipline for transferable regions.
+
+In \name, this enforcement is done at runtime by explicitly keeping
+track of the \emph{current state} for $\RgnZ$ objects, and checking
+the validity of every open, transfer, or free operation and throwing
+an exception if it is invalid.  
+% The challenge in enforcing this
+% discipline statically is that transferable regions are first-class
+% objects. Hence, the program can create multiple aliases for the same
+% region, \eg, open it via one alias and free it via another.  Typestate
+% verification in the presence of aliases is hard.  The checking can be
+% done statically by preventing the creation of aliases using, \eg,
+% linear types or unique types. However, this would be quite
+% restrictive, in terms of expressiveness.
+As explained previously, this is a reasonable trade-off in the context
+of \name, as regions are coarse-grained objects, which are manipulated
+infrequently, when compared to fine-grained objects that reside inside
+these regions. Therefore, runtime overhead of checking the region's
+state transition discipline is acceptable.
+
+\input{fb-syntax}
+
+\paragraph{Cloning} Note that in the example from
+Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
+\C{selector} (on Line 13) should not contain any references to the
+input object, since the input region, where the object resides, will
+be freed at the end of the method. If there is a need for the output
+object to point to subobjects of the input object, such subobjects
+must be cloned (to copy them from the input region to the output
+region).  Fortunately, \name's region type system
+(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
+type of \C{selector} and the type checker will ensure correctness.
+Furthermore, the type can be automatically inferred by \name's region
+type inference (\S~\ref{sec:type-inference}), which can perform the
+above reasoning on behalf of the programmer.
+
+%\begin{codejava} Region<string> rgn = new Region<List<String>> (() =>
+%new List<String>()); open rgn as  strList@R0 {
+%strList.addAtHead("World"); strList.addAtHead("Hello"); }
+%rgn.transfer(); \end{codejava}
+
+\COMMENT{ \subsection{Stack Regions and Outlives Relation}
+
+The \C{letregion} blocks can be nested, leading to a stack of regions
+that are deallocated in the reverse order in which they are allocated.
+Following ~\cite{cyclonepldi02}, we therefore call regions introduced
+by \C{letregion} expressions as \emph{stack regions}. The stack
+discipline induces an \emph{outlives} relationship among regions
+created by nested \C{letregion}s, where the region introduced by the
+outer \C{letregion} is guaranteed to outlive the one introduced by the
+inner \C{letregion}. It is therefore safe to refer to an object
+allocated in outer region from the inner region, but the converse is
+not true. For example, consider the following code (assume that class
+\C{A} has field \C{x} of type \C{Object}): \begin{center}
+\begin{codejava} letregion R0 { A a0 = new@R0 A(); letregion R1 { A a1
+= new@R1 A(); a1.x = new@R0 Object(); // safe & legal a0.x = new@R1
+Object(); // unsafe & illegal ...  } ...  } \end{codejava}
+\end{center} The code creates two stack regions with identifiers
+\C{R0} and \C{R1}, where the region \C{R0} outlives the region \C{R1}
+(denoted as $\C{R0} \outlives \C{R1}$).  Objects \C{a0} and \C{a1} are
+allocated in regions \C{R0} and \C{R1}, respectively. The first
+assignment statement assigns to \C{a1.x} an object allocated in outer
+region (\C{R0}). This assignment is safe as \C{a1.x} refers to a
+longer living object, hence is guaranteed to be a valid reference
+throughout the lifetime of \C{a1}.  In contrast, the second assignment
+is unsafe, as it assigns to \C{a0.x} an object, whose lifetime is
+shorter than the lifetime of \C{a0}, making it unsafe to dereference
+\C{a0.x} outside the inner block. 
+% Unsafe assignments can also happen indirectly via a function call..
+Preempting such unsafe assignments is the \emph{raison d'etre} of the
+region type system.
 
@@
 \label{sec:alloc-ctxt}
-
-As the preceding examples illustrate, the suffix \C{@R} in
-\C{new@R} indicates that the newly created object should be allocated in the region labeled \C{R}.
-However, \name does not require the allocation region to be specified following a \C{new}.
-This is critical, as it enables \name applications to use existing region-oblivious C\# libraries.
-In the
-absence of explicit specification, \name allocates the newly created
-object in its \emph{allocation context}, which is the region at the
-top of the stack when \C{new} is executed. For instance, the
-\C{findMin} function shown below, which makes explicit use of regions,
-calls the standard \C{List} library's \C{listIterator} method:
-\begin{center}
-\begin{codejava}
-  int findMin(List<int> l) {
-    assert(!l.isEmpty());
-    int min = l.get(0);
-    letregion R0 {
-      ListIterator<int> iter = l.listIterator();
-      while(iter.hasNext()) {
-        int n = iter.getNext();
-        if (n<min) min = n;
-      }
-    }
-    return min;
-  }
-\end{codejava}
-\end{center}
-The \C{listIterator} method neither creates new regions, nor does it
-annotate \C{new} expressions with region specifications. Consequently,
-all new objects created by the method are allocated in the allocation
-context for the call to \C{listIterator}, which in this case is
-\C{R0}.
 
@@
 
-%\subsection{Boxed and Unboxed Values}
-%
-%In \name, like in \csharp, values of some primitive types, such as
-%integers and bools, are unboxed, whereas others, such as strings are
-%arrays, are boxed. Unlike unboxed values, boxed values are objects,
-%hence passed by reference. It is therefore important to be mindful
-%of allocation regions for strings and arrays, but not so much for
-%integers and bools. For instance, in the following code, the
-%assignment to integer variable \C{x} is safe, but the assignment to
-%the string variable \C{y} is unsafe, as the allocation region (\C{R0})
-%of the string constant has lesser lifetime than \C{y}.
-%\begin{center}
-%\begin{codejava}
-%  int x;
-%  string y;
-%  letregion R0 {
-%    x = 2;
-%    y = "Hello, World!";
-%  }
-%  ...
-%\end{codejava}
-%\end{center}
-%Accordingly, \name's region type system lets the former assignment
-%pass by, but flags the later as an error.
-
 \subsection{Dynamic (Transferable) Regions}
 
-\name's stack regions have lexically scoped lifetimes, hence are not
-suitable for applications that need to transfer data between
-autonomous entities, such as \naiad dataflow operators. To support
-these applications, it is imperative to support regions whose
-lifetimes transcend the boundaries of lexical blocks, functions, or
-even programs. Regions that fit this description in \name are called
-\emph{dynamic regions}, or \emph{transferable regions}.
+In a typical dataflow computation, an upstream actor (e.g., a
+\C{SELECT} operator) constructs a transferable region, sets its root
+to the data structure containing intermediate results, and then
+transfers it to a downstream actor (e.g., a \C{COUNT} operator), which
+performs further processing. Since a transferable region escapes the
+lifetime of the sender, there must be no references from inside of the
+transferable region to objects allocated in other memory regions of
+the sender. Such references, if exist, may become invalid references
+in the context of recipient's address space, jeopardizing memory
+safety. \name relies on its region type system to prevent such unsafe
+references from being created.
 
-Transferable regions, like stack regions, are memory regions, but
-unlike stack regions, they are first class values of \name; they are
-considered objects of \C{Region} class, and, like objects of other
-classes, they are created using the \C{new} keyword, can be passed as
-arguments, stored in data structures, and returned from methods. Like
-a stack region, a transferable region facilitates arbitrary object
-allocations, but unlike a stack region, it has a well-defined
-\emph{root} object denoting the root of the data structure being
-transferred. In a typical dataflow computation, an upstream actor
-(e.g., a \C{SELECT} operator) constructs a transferable region, sets
-its root to the data structure containing intermediate results, and
-then transfers it to a downstream actor (e.g., a \C{COUNT} operator),
-which performs further processing. Since a transferable region escapes
-the lifetime of the sender, there must be no references from inside of
-the transferable region to objects allocated in other memory regions
-of the sender. Such references, if exist, may become invalid
-references in the context of recipient's address space, jeopardizing
-memory safety. \name relies on its region type system to prevent such
-unsafe references from being created.
-
-\begin{figure}
-\begin{codejava}
-  public class Region<T> {
-    public Region(Func<void,T> mkRoot);
-    public void free();
-    public void transfer();
-  }
-\end{codejava}
-\caption{The type signature of the \C{Region} class}
-\label{fig:region-class}
-\end{figure}
-The type signature of \C{Region} class is described in
-Fig.~\ref{fig:region-class}. The class is parametric over the type of
-the root object (i.e., the \emph{root type}) of the transferable
-region it creates. The \C{Region} constructor expects a \C{Func}
-object - a higher-order function argument that returns an object of
-root type.  The idea is to execute this function with the newly
-created transferable region as its allocation context. The object thus
-created inside the transferable region will serve as its root object.
-For instance, the following code creates a transferable region
-(\C{rgn}) and sets its root to the string object returned by its
-\C{Func} argument (a lambda expression):
-\begin{codejava}
-  Region<string> rgn = new Region<string>
-                        (() => "Hello");
-\end{codejava}
-Among the methods of the \C{Region} class, \C{free} deallocates
-region's memory, while \C{transfer} transfers the region to a
-downstream actor as determined by the run-time. The precise semantics
-of \C{transfer} are unimportant in the context of the region type
-system; it suffices to understand \C{transfer} as an alias of
-\C{free}.
-
-\name imposes a simple discipline on how $\RgnZ$ objects can be used
-to perform operations on transferable regions. This discipline is
-imposed in the interest of memory safety, which is discussed in the
-following subsection.
-
-\subsection{Memory Safety}
-\label{sec:memory-safety}
-
-The key to memory safety in \name is the following restriction:
-an object $o_1$ in a region $R_1$ is allowed to store a pointer to
-an object $o_2$ in a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.
-(A similar restriction applies in the case where $o_1$ is a stack-allocated variable.)
-
-Enforcing this restriction is simple in the case of stack regions since the outlives relation
-between stack regions can be inferred from their lexical nesting. Unfortunately,
-inferring outlives relations between transferable regions is not easy.
-\name imposes the following protocol on the use of transferable regions to help simplify
-this check.
-
-A transferable region (that has not been freed or transferred) can be in one of two possible
-states, \emph{open} or \emph{closed}. A newly created region is in the closed state.
-A region must be opened, using the open construct (explained in detail below), in order
-to read or update or allocate an object within that region.
-An open region cannot be freed or transferred. 
-In particular, an open region is guaranteed to be live for the entire duration of the open construct.
-This allows the type system to infer a valid outlives relation between the opened region
-and any stack region that is nested within the open construct.
-
-%Memory safety in \name is primarily due to its region type system,
-%although few runtime checks are needed.
-%% However, the type system does not ensure the integrity of a small
-%% subset of pointers, called the \emph{unsafe set}, for which runtime
-%% verification is needed to guarantee safety. This is more a design
-%% decision rather than an unfortunate inevitability, as the following
-%% discussion elaborates.
-%At a fundamental level, the region type system's approach to memory
-%safety involves approximating (a). the set of memory regions that are
-%live at any given program location\footnote{We say the type system
-%tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
-%relationships between them, and using this information to preempt
-%potentially unsafe references from being created. These include:
-%\begin{itemize}
-%\item References into dead/non-existing regions,
-%\item References from longer-living stack regions to shorter-living ones, and 
-%\item References that escape transferable regions.
-%\end{itemize}
-%Maintaining precise approximation of liveness and outlives information
-%for stack regions is easy. However, for dynamic (transferable) regions,
-%statically determining liveness is equivalent to determining if a C
-%pointer is null in presence of dynamic memory allocation, which is
-%undecidable in general. Systems that attempt to ensure full memory
-%safety in presence of dynamic memory (e.g., ~\cite{henglein},
-%~\cite{rust}) do so by conservatively approximating liveness
-%information for memory (regions). Maintaining a good conservative
-%approximation requires them to employ techniques such as linear
-%types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
-%restrictions that are often too hard to program
-%around\footnote{Internet has many tutorials on how to \emph{fight} the
-%borrow checker in Rust.}. This is particularly unacceptable in the
-%context of \name, as dataflow programs are often run offline, and the
-%concern for performance and developer productivity outweighs the
-%concern for safety. 
-%
-%\name's region type system therefore adopts the novel approach of
-%overapproximating liveness information for transferable regions, i.e.,
-%it might assume that a transferable region is live when, in fact, it
-%might have already been transferred or freed. This imprecision should
-%be a cause of concern as it can lead to cases where creation of unsafe
-%references, and their subsquent dereferencing go unnoticed by the type
-%system. Considering that stack regions can contain pointers into
-%transferable regions, the worst case cost of overapproximating
-%liveness is therefore the integrity of \emph{every} pointer on stack.
-
-%Fortunately, \name avoids this worst case scenario.  \name achieves
-%this by imposing a simple state transition discipline on the lifetime
-%of a transferable region.  The relevant finite state machine is shown
-%in Fig.~\ref{fig:region-fsm}.
-
-The protocol for transferable regions is presented as a finite state machine in Fig.~\ref{fig:region-fsm}.
-
-\begin{figure}
-\includegraphics[scale=0.45]{region-fsm.png}
-\caption{The lifetime of a dynamic (transferable) region in \name}
-\label{fig:region-fsm}
-\end{figure}
-
-A transferable region starts its lifetime in a \emph{closed} state,
-when it is created through \C{new Region} expresion. \name provides
-\C{open} lexical construct to \emph{open} a transferable region as an
-allocation context, and to access its root object:
-\begin{codejava}
-  Region<string> rgn = new Region<List<String>>
-                        (() => new List<String>());
-  open rgn as  strList@R0 {
-    strList.addAtHead("World");
-    strList.addAtHead("Hello");
-  }
-  rgn.transfer();
-\end{codejava}
-In the above code, the \C{open} lexical block performs the
-following operations: (a). It opens the transferable region handled by
-\C{rgn} for allocation (i.e., makes it the current allocation
-context), (b). assigns the identifier \C{R0} to the open region, and
-(c). sets the newly introduced local variable (\C{strList}) to refer
-to its root object (a list of strings). As Fig.~\ref{fig:region-fsm}
-indicates, it is not possible to open a region that is already
-transferred/freed. The fact that \C{rgn} is open within the block
-therefore guarantees that it is not yet transferred/freed, and that
-dereferencing \C{strList} within the block is safe. The end of
-\C{open} block marks the return of transferable region to the closed
-state.  While closed, the region is eligible to be transferred to a
-downstream actor, or to be freed.  An actor that receives the
-transferable region, receives it in the closed state. It can then
-reopen the received region to read its contents, possibly add more
-data and transfer it to another actor, or free the region.
+%% USING TEMPORARY STACK REGIONS AND ITS CORRECTNESS
 
@@
 with the data stored in a transferable region. Consider the following
-code, for instance\footnote{For brevity, we drop the region identifier binding part of the
-\C{open} expression whenever the identifier is not used.}:
-\begin{codejava}
-void onReceive(Region<List<String>> rgn)
-  open rgn as strList {
-    letregion R1 {
-      String s = "";
-      ListIterator<String> i = strList.listIterator();
-      while(i.hasNext()) {
-        s += i.getNext();
+code, for instance\footnote{For brevity, we drop the region identifier
+binding part of the \C{open} expression whenever the identifier is not
+used.}: \begin{codejava} void onReceive(Region<List<String>> rgn) open
+rgn as strList { letregion R1 { String s = ""; ListIterator<String> i
+= strList.listIterator(); while(i.hasNext()) { s += i.getNext(); }
+print s; //prints "HelloWorld" } } rgn.free(); \end{codejava} The
+stack region \C{R1} is being used in the above code to provide working
+memory to work with the objects of transferable region \C{rgn}. Since
+a transferable region cannot be transferred/freed while it is still
+open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to outlive the
+stack region \C{R1} in the above code, making it safe for the later to
+contain references to the former. \name therefore allows such
+references.
+
+As Fig.~\ref{fig:region-fsm} indicates, it is not possible to open a
+region that is already transferred/freed. The fact that \C{rgn} is
+open within the block therefore guarantees that it is not yet
+transferred/freed, and that dereferencing \C{strList} within the block
+is safe. The end of \C{open} block marks the return of transferable
+region to the closed state.  While closed, the region is eligible to
+be transferred to a downstream actor, or to be freed.  An actor that
+receives the transferable region, receives it in the closed state. It
+can then reopen the received region to read its contents, possibly add
+more data and transfer it to another actor, or free the region.
+
 }
-      print s; //prints "HelloWorld"
-    }
-  }
-  rgn.free();
-}
-\end{codejava}
-The stack region \C{R1} is being used in the above code to provide
-working memory to work with the objects of transferable region
-\C{rgn}. Since a transferable region cannot be transferred/freed while
-it is still open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to
-outlive the stack region \C{R1} in the above code, making it safe for
-the later to contain references to the former. \name therefore allows
-such references.
-
-% The design decisions reflected above, \emph{viz.}, to ascribe a root
-% object to every transferable region, to make use of a higher-order
-% argument to initialize the root object, and to require that a region
-% be explictly opened in a lexical block for allocations, were made to
-% reflect the message semantics of transferable regions, to promote
-% uniform style of programming with different kinds of regions, and to
-% precisely state and enforce the safety guarantees offered by \name's
-% region type system (\S~\ref{sec:type-system}).
-
-As evident in the above examples, the safety of memory accesses in
-\name is now subject to the condition that every transferable region
-correctly follows the state transition discipline shown
-in Fig.~\ref{fig:region-fsm}. If this is
-guaranteed, then \name's region type system statically guarantees the
-safety of all memory accesses. In other words, the type system reduces
-the problem of ensuring memory safety in \name programs to the problem
-of enforcing the state transition discipline for transferable regions.
-
-In \name, this enforcement is done at runtime by explicitly keeping
-track of the \emph{current state} for $\RgnZ$ objects, and
-checking the validity of every open, transfer, or free operation
-and throwing an exception if it is invalid.
-The challenge in enforcing this discipline statically is that transferable regions
-are first-class objects. Hence, the program can create multiple aliases for
-the same region, \eg, open it via one alias and free it via another.
-Typestate verification in the presence of aliases is hard.
-The checking can be done statically by preventing the creation of aliases using, \eg, linear types
-or unique types. However, this would be quite restrictive, in terms of expressiveness.
-
-\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
-in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
-as well as the runtime overhead of checking the region's state transition discipline is acceptable.
-
-\input{motivating-eg-in-broom}
-
-\subsection{An Example}
-
-Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
-of Fig.~\ref{fig:motivating-eg} can be written in \name.
-The \C{onReceive} method receives its input message in a \emph{transferred}
-region (\ie, a \emph{closed} region whose ownership is transferred to the
-recipient).
-Line 7 creates a new region to store the output for time \C{t},
-initializing it to contain an empty list.
-Line 9 opens the input region to process it.
-Line 10 creates a stack region \C{R0}.
-Thus, the temporary objects created by the iteration in line 11,
-for example, will be allocated in this stack region that lives just
-long enough.
-We open the desired output region in line 12, so that the new output
-objects created by the invocation of \C{selector} in line 13
-are allocated in the output region.
-Finally, the input region is freed in line 19.
-
-Note that this approach works correctly only if \C{selector}
-creates an output object that does not share any subobjects with its input
-object (which resides in the input region and will be freed at the end of
-the method).
-If desired, such subobjects must be cloned (to copy them from the input
-region to the output region).
-Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of \C{selector}
-and the type checker will ensure correctness.
-Furthermore, the type can be automatically inferred by \name's region type
-inference (\S~\ref{sec:type-inference}), which can peform the above
-non-trivial reasoning on behalf of the programmer.

### 2015-11-20 paper/type-inference.tex
Merged paper to techrep.

@@
 expressions, similar to the language introduced in
-\S~\ref{sec:overview}) to programs in \FB. The key aspects of this
-elaboration are described below.
-
-% We formalize region type inference as an elaboration function from
-% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
-% programs in \FB.  The elaboration proceeds in four steps. In the first
-% step we make use of the observation that region types are
-% \emph{refinements} of FGJ types with region annotations and
-% constraints over such region annotations, and compute polymorphic
-% region type templates for methods and classes from their FGJ types.
-% The templates contain region variables ($\rho$) to denote unknown
-% region annotations, and predicate variables ($\varphi$) to denote
-% unknown constraints over such region annotations. Free region
-% variables are generalized in types (hence, polymorphic). Second, we
-% elaborate expressions by introducing region variables to denote
-% unknown region arguments in \C{new} expressions, method calls and
-% function applications. While elaborating expressions, we make use of
-% the computed region type templates to build a system of constraints
-% that capture well-formedness requirements and subtyping relationships
-% between type templates that must hold (as per the static semantics in
-% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
-% lift expression elaboration and constraint generation to methods and
-% classes. Finally, we solve the constraints to compute solutions to
-% region ($\rho$) and predicate ($\varphi$) variables.
-%
+\S~\ref{sec:overview}) to programs in \FB.  The elaboration proceeds
+in four steps. In the first step we make use of the observation that
+region types are \emph{refinements} of FGJ types with region
+annotations and constraints over such region annotations, and compute
+polymorphic region type templates for methods and classes from their
+FGJ types. The templates contain region variables ($\rho$) to denote
+unknown region annotations, and predicate variables ($\varphi$) to
+denote unknown constraints over such region annotations. Free region
+variables are generalized in types (hence, polymorphic). Second, we
+make use of the computed region type templates to elaborate
+expressions by introducing region variables to denote unknown region
+arguments in \C{new} expressions, method calls and function
+applications. While elaborating expressions, we also build a system of
+constraints that capture well-formedness requirements and subtyping
+relationships between type templates that must hold (as per the static
+semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
+Third, we lift expression elaboration and constraint generation to
+methods and classes. Finally, we solve the constraints by making use
+of our fixpoint constraint solving algorithm \csolvestar, which
+reduces the constraint solving problem to an abduction problem in a
+Herbrand constraint system, and then relies on \csolve, our abduction
+solver for that domain. 
 % Due to the presence of region-polymorphic
@@
 template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
-region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
-templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
-region-annotated definition of \C{Pair} class from
-\S\ref{sec:fb-syntax} a region type template for a method with FGJ
-type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
-our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
-subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
-accept one region parameter - the allocation region of their
-objects.}\footnote{We abuse arrow notation to also represent types of
-methods, but unlike function types, there is no allocation region
-annotation atop the arrow.}
-$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
-\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
-\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
-region variables, and $\varphi_0$ is a fresh predicate variable
-denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
-Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
-$\rho$ is fresh. For a class, a template is computed based on the 
-type templates of its fields, constructors and methods. Additional
-region and predicate variables are added as necessary. Following template 
-of the \C{Pair} class illustrates this point:
-% For a class, a template is computed in two steps. In
-% the first step, we templatize the types of all its fields, constructor
-% arguments, and arguments and return values of all its methods, along
-% with the type of its superclass. In the second step, we generalize all
-% the free region variables occuring in the templates computed in the
-% first step as region parameters of the class. Finally, we add a fresh
-% allocation region parameter ($\rhoalloc$) to the list of parameters,
-% and introduce a new predicate variable ($\varphi$) to denote unknown
-% constraints on region parameters. For example, consider the standard
-% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
-% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
-% respectively. It can be templatized as following :
-\vspace*{-0.1in}
+region variable. Likewise, given the region-annotated definition of
+\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
+method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
+is \footnote{In our exposition, we assume that classes $\C{A}$ and
+$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
+$\ObjZ$, they accept one region parameter - the allocation region of
+their objects.}\footnote{We abuse arrow notation to also represent
+types of methods, but unlike function types, there is no allocation
+region annotation atop the arrow in a method type.}
+$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
+\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
+\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
+variables, and $\varphi_0$ is a fresh predicate variable denoting
+unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
+template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
+fresh. For a class, a template is computed in two steps. In the first
+step, we templatize the types of all its fields, constructor
+arguments, and arguments and return values of all its methods, along
+with the type of its superclass. In the second step, we generalize all
+the free region variables occuring in the templates computed in the
+first step as region parameters of the class. Finally, we add a fresh
+allocation region parameter ($\rhoalloc$) to the list of parameters,
+and introduce a new predicate variable ($\varphi$) to denote unknown
+constraints on region parameters. For example, consider the standard
+FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
+and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
+respectively. It can be templatized as following:
 \begin{center}
@@
           $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
-  $a@\rho_0$ fst; $b@\rho_1$ snd;
-  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
-    super(); this.fst=fst; this.snd=snd; 
-  }
-  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
+  $a@\rho_0$ fst;
+  $b@\rho_1$ snd;
+  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
+  $...$
 }
@@
 \end{center}
-\vspace*{-0.1in}
 % \footnote{In reality, the generalized region variables only come
@@
 % types do not contain any free region variables.}
-% The template elides $\C{Pair}$'s methods, whose region type
-% templates contain no free variables. Among the region parameters of
-% the class template, $\rho_{0-4}$ are obtained by generalizing free
-% region variables in the types of its class's fields, constructor
-% arguments, and its superclass type.  The remaining parameter
-% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
-% region argument. Fresh predicate variable $\varphi_0$ denotes
-% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
-% hold for template to be a well-formed region-annotated class
-% definition in $\FB$. 
-%
-% The type template for a recursively defined class is computed in two
-% steps. First, all recursive occurances of the class among the types of
-% its fields are ignored and the class is templatized as if it is a
-% non-recursive class. Next, all the recursive occurances are
-% templatized with respect to the class template computed in the first
-% step, such that their region annotations are \emph{exactly} same as
-% the region parameters of the class. Applying this technique on a
-% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
-% example, results in a template that uses same region variable to
-% denote allocation regions of the current node, and the node that the
-% current node refers to. 
-%
-% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
-% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
-% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
-% data stored in the linked list node.  To templatize the
-% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
-% occurance in the type of $\C{next}$ field, and templatize the type
-% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
-% template of $\C{data}$ field, we compute the class's template as
-% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
-% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
-% \end{center} Next, we trivially templatize the type of $\C{next}$
-% field in the body of the class as
-% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
-% resulting class represents a linked list with spine in the region
-% $\rhoalloc_0$ and data objects in the region $\rho_0$.
+The template elides $\C{Pair}$'s methods, whose region type templates
+contain no free variables. Among the region parameters of the class
+template, $\rho_{0-4}$ are obtained by generalizing free region
+variables in the types of its class's fields, constructor arguments,
+and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
+fresh region variable denoting the allocation region argument. Fresh
+predicate variable $\varphi_0$ denotes unknown constraints over
+$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
+well-formed region-annotated class definition in $\FB$.
+
+The type template for a recursively defined class is computed in two
+steps. First, all recursive occurances of the class among the types of
+its fields are ignored and the class is templatized as if it is a
+non-recursive class. Next, all the recursive occurances are
+templatized with respect to the class template computed in the first
+step, such that their region annotations are exactly same as the
+region parameters of the class. For example, consider a generic
+$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
+type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
+where $\C{a}$ is the type of the data stored in the linked list node.
+To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
+recursive occurance in the type of $\C{next}$ field, and templatize
+the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
+type template of $\C{data}$ field, we compute the class's template as
+following:
+\begin{center}
+$\C{ListNode}\inang{\C{a}\extends\ObjZ}
+\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
+\end{center}
+Next, we trivially templatize the type of $\C{next}$ field in the body
+of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
+The resulting class represents a linked list with spine in the region
+$\rhoalloc_0$ and data objects in the region $\rho_0$.
+
 % Observe that our strategy of trivially templatizing the recursive
@@
 % region variables are introduced.
-The type template for for a recursively defined class is computed in
-two steps. First, a template is computed by ignoring the recursive
-occurrences. Next, all the recursive occurrences are templatized based
-on the template computed in first step.  This approach can be extended
-to mutually recursive definitions in a straightforward manner. Using
-the techniques outlined above, we compute region type templates for
-all classes bound in the class table of the $\absof{\FB}$ program
-before we proceed to elaborate expressions.
+The templatization technique we described above for recursive class
+definitions can be extended to mutually recursive definitions in a
+straightforward manner, by simultaneously templatizing them. Using the
+techniques outlined above, we compute region type templates for all
+classes bound in the class table of the $\absof{\FB}$ program before
+we proceed to elaborate expressions.
 
-%\input{fb-elabexpr}
-%\input{fb-elabmeth}
+\input{fb-elabexpr}
+\input{fb-elabmeth}
 
-\subsection{Elaboration and Constraint Generation}
+\subsection{Expression Elaboration}
 
 Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
-(a). replacing core types occurring in expressions with fresh region
-type templates, and (b). explicitly instantiating region parameters
-with fresh region variables in method calls and function applications.
-This elaboration is performed with respect to the polymorphic type
-templates of classes and methods computed as per
-\S\ref{sec:fb-templatization}. Our expression elaboration
-function\footnote{Supplement contains the pseudocode for all the
-elaboration functions.} traverses expressions in a syntax-directed
-manner of a type checker, introducing fresh region type templates for
-unknown region types, while generating constraints over region and
-predicate variables. The precise nature of generated constraints is
-explained in \S\ref{sec:constraints}, but in summary, they
-capture the relationships between the type templates of various
-subexpressions and their well-formedness.
+(a). replacing core types in variable declarations and \C{new}
+expressions with fresh region type templates, and (b). explicitly
+instantiating region parameters of methods with fresh region variables
+in method calls and function applications. This elaboration is
+performed with respect to the polymorphic type templates of classes
+and methods computed as per \S\ref{sec:fb-templatization}.
 
-% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-% this elaboration for a subset of expressions in $\absof{\FB}$
-% %, whose corresponding $\FB$ expressions have been ascribed static
-% %semantics in Fig.~\ref{fig:fb-staticsem}. 
-% $\elabExpr$ is defined under the same context as the expression typing
-% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
-% and $\env$ retaining their meaning. The function traverses expressions
-% in a syntax-directed manner of a type checker, introducing fresh
-% region type templates for unknown region types, while generating
-% constraints over region and predicate variables. The precise nature of
-% generated constraints is explained in \S\ref{sec:fb-constraintsem},
-% but in summary, they capture the relationships between the type
-% templates of various subexpressions and their well-formedness. Note
-% that $\elabExpr$ returns the region type template of the
-% subexpression, which is used to generate constraints for the
-% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
-% shown) used by $\elabExpr$ implement type well-formedness and subtype
-% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.
+Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+this elaboration for a subset of expressions in $\absof{\FB}$, whose
+corresponding $\FB$ expressions have been ascribed static semantics in
+Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
+context as the expression typing judgment in
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
+retaining their meaning. The function traverses expressions in a
+syntax-directed manner of a type checker, introducing fresh region
+type templates for unknown region types, while generating constraints
+over region and predicate variables. The precise nature of generated
+constraints is explained in \S\ref{sec:fb-constraintsem}, but in
+summary, they capture the relationships between the type templates of
+various subexpressions and their well-formedness. Note that
+$\elabExpr$ returns the region type template of the subexpression,
+which is used to generate constraints for the expression. Functions
+$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
+implement type well-formedness and subtype judgments from
+Fig.~\ref{fig:fb-staticsem}, respectively.
 
-Expression elaboration can be trivially lifted to the level of a
-method, and method elaboration to the level of a class, while
-accumulating constraints. Besides such constraints, elaboration
-functions for methods and classes also generate new constraints by
-checking well-formedness and subtype relationships among the types of
-their constituents. For instance, method elaboration generates a new
-constraint that captures the requirement that, under the guarantees
-provided by the context, the region type of a method body must be a
-subtype\footnote{Subtyping is type equality in \FB. In \name,
-subtyping between region types is a straightforward extension of
-subtyping between core types.} of method's return type.  Likewise, the
-elaboration function for a class generates a fresh constraint that
-captures a couple of safety conditions: first, the allocation regions
+\subsection{Method and Class Elaboration}
+
+Functions $\elabMeth$ and $\elabClass$ shown in
+Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+class definitions, respectively. Both functions first build a context
+($\A$) containing a set ($ \rhoenv$) of region variables denoting
+regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ denoting the fact that such constraints are yet
+to be inferred.
+
+Function $\elabMeth$ elaborates a method definition of class $B$. It
+calls $\elabExpr$ with the context $\A$, its allocation context
+parameter ($\rhoallocm$), and a type environment ($\env$) that
+contains region type bindings for all the arguments of the method,
+including the implicit $\C{this}$ argument. The region type template
+returned by $\elabExpr$ for the method body is checked against its
+expected type (derived from the type template of the method)
+generating more constraints. The function then returns the elaborated
+method definition and the set of constraints.
+
+$\elabClass$ elaborates the definition of a class $B$. It relies on
+$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
+hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
+constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+the set of constraints returned by these functions, $\elabClass$ adds
+constraints generated by checking the well-formedness of the type
+templates of its superclass and fields, and also a new constraint
+capturing a couple of safety conditions: first, the allocation regions
 of objects referred by the instance variables should outlive the
@@
 
-Elaborating a class table ($CT$) entails elaborating every class
-definition in $CT$, while collecting constraints. The constraints thus
-collected are finally solved to obtain assignments for free region and
-predicate variables introduced during templatization and elaboration
-stages. The solutions are applied to the class table (and to the
-artifacts that make up the class table, recursively) to compute a
-class table that maps classes to their fully region-annotated
-definitions in $\FB$.
+Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
+every definition in the class table $CT$, while accumulating
+constraints. The constraints are finally solved by $\solve$
+(\S\ref{sec:fb-constraintsolving}), which returns substitution
+functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
+predicate variables, respectively, introduced during templatization
+and elaboration stages. The substitutions are applied to the class
+table (and to the artifacts that make up the class table, recursively)
+to compute a class table that maps classes to their fully
+region-annotated definitions in $\FB$.
 
-% \subsection{Method and Class Elaboration}
-
-% Elaborating a method involves performingexpression elaboration over
-% its body, while collecting constraints. Method elaboration can be
-% lifted to the level of classes
-
-% Functions $\elabMeth$ and $\elabClass$ shown in
-% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
-% class definitions, respectively. Both functions first build a context
-% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
-% live, a map ($\aenv$) mapping type variables to their bounds, and a
-% constraint formula ($\phicx$) capturing constraints over live region
-% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
-% capture constraints over variables in $\rhoenv$ denoting the fact that
-% such constraints are yet to be inferred. Both elaboration functions
-% generate new constraints by checking well-formedness and subtype
-% relationships among types of their constituents. A fresh constraint
-% ($C_4$) generated by $\elabClass$ captures a couple of safety
-% conditions: first, the allocation regions of objects referred by the
-% instance variables should outlive the allocation region of the
-% instance itself, and second, the allocation regions of a class type
-% and its superclass type must be the same.
-
-% Function $\elabMeth$ elaborates a method definition of class $B$. It
-% calls $\elabExpr$ with the context $\A$, its allocation context
-% parameter ($\rhoallocm$), and a type environment ($\env$) that
-% contains region type bindings for all the arguments of the method,
-% including the implicit $\C{this}$ argument. The region type template
-% returned by $\elabExpr$ for the method body is checked against its
-% expected type (derived from the type template of the method)
-% generating more constraints. The function then returns the elaborated
-% method definition and the set of constraints.
-
-% $\elabClass$ elaborates the definition of a class $B$. It relies on
-% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
-% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
-% the set of constraints returned by these functions, $\elabClass$ adds
-% constraints generated by checking the well-formedness of the type
-% templates of its superclass and fields, and also a new constraint
-% capturing a couple of safety conditions: first, the allocation regions
-% of objects referred by the instance variables should outlive the
-% allocation region of the instance itself, and second, the allocation
-% regions of a class type and its superclass type must be the same.
-
-% Function $\elabClassTable$ (not shown) elaborates
-% every definition in the class table $CT$, while accumulating
-% constraints. The constraints are finally solved for free region and
-% predicate variables introduced during templatization and elaboration
-% stages. The solutions are applied to the class table (and to the
-% artifacts that make up the class table, recursively) to compute a
-% class table that maps classes to their fully region-annotated
-% definitions in $\FB$.
-
-% Note that if the original program in $\absof{\FB}$ contains unsafe
-% references, for example, a reference from a transferable region to a
-% stack regions, then the constraints generated during the elaboration
-% are not satisfiable. In such case, $\solve$ fails to solve
-% constraints, causing the program to be rejected.
+Note that if the original program in $\absof{\FB}$ contains unsafe
+references, for example, a reference from a transferable region to a
+stack regions, then the constraints generated during the elaboration
+are not satisfiable. In such case, $\solve$ fails to solve
+constraints, causing the program to be rejected.
 
 \subsection{Constraints}
-\label{sec:constraints}
+\label{sec:fb-constraintsem}
 
-Three kinds of constraints are generated during elaboration:
+Our constraint generation algorithm generates three kinds of
+constraints:
 \begin{itemize}
@@
 deducible from $\rgn_1 \outlives \rgn_2$.
+
 % Validity constraints primarily result from checking if a region type
@@
 % region-polymorphic methods, where similar situation arises. 
-% In general, validity constraints generated by our algorithm assume one
-% of the following two forms:
-% \begin{center}
-% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
-%            {\phicstr} \qquad
-%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
-%            {F_j(\varphi_j)}
-% \)
-% \end{center}
-% Where $\varphi_i$'s denote the unknown preconditions of the class and
-% the method under elaboration. If the constraint is generated while
-% checking the well-formedness of a type or elaborating an expression,
-% then $\varphi_j$'s denote the unknown preconditions of classes and
-% methods that were used in that type or expression. Each use of a
-% (region-polymorphic) class or a method has to instantiate region
-% parameters, resulting in pending substitutions ($F_j$). Formulas
-% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
-% variables and pending substitutions. While $\phictxt$ captures
-% relationships that are \emph{known} to hold between concrete region
-% identifiers (i.e., $\rgn$'s) when the constraint was generated,
-% $\phicstr$ captures relationships that are \emph{required} to hold
-% among region varibles (i.e., $\rho$'s), or relationships between
-% region variables and identifiers. Each region variable occuring in
-% $\phicstr$ has an associated well-formedness constraint, which
-% specifies its unification domain. The unification domain of a
-% constraint is the union of unification domains of all region variables
-% occuring in the constraint.
+In general, validity constraints generated by our algorithm assume one
+of the following two forms:
+\begin{center}
+\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+           {\phicstr} \qquad
+   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+           {F_j(\varphi_j)}
+\)
+\end{center}
+Where $\varphi_i$'s denote the unknown preconditions of the class and
+the method under elaboration. If the constraint is generated while
+checking the well-formedness of a type or elaborating an expression,
+then $\varphi_j$'s denote the unknown preconditions of classes and
+methods that were used in that type or expression.  Each use of a
+(region-polymorphic) class or a method may instantiate region
+parameters differently, resulting in a different pending substitution
+($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
+of predicate variables and pending substitutions. While $\phictxt$
+captures relationships that are \emph{known} to hold between concrete
+region identifiers (i.e., $\rgn$'s) when the constraint was generated,
+$\phicstr$ captures relationships that are \emph{required} to hold
+among region varibles (i.e., $\rho$'s), or relationships between
+region variables and identifiers. Each region variable occuring in
+$\phicstr$ has an associated well-formedness constraint, which
+specifies its unification domain. The unification domain of a
+constraint is the union of unification domains of all region variables
+occuring in the constraint.
 
@@
 \end{smathpar}
-% The first constraint is generated by $\typeOk$ on the type of
-% $\C{p}$.  Second and third are generated by the the assignment
-% expressions, and the last constraint is generated from the recursive
-% call.
+
+% \paragraph{Constraints Example 1} The original source of all
+% outlives constraints is the validity constraint $C_3$ in $\elabClass$
+% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
+% class template from \S\ref{sec:fb-templatization}, the constraint is
+% as following:
+% \begin{center}
+%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
+%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
+% \end{center}
+
+% \paragraph{Constraints Example 2} Consider a contrived method
+% $\C{alt}$ that accepts an argument $\C{p}$ of type
+% $\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
+% of type $A$. It then assigns one object to $\C{p.fst}$ and other to
+% $\C{p.snd}$, and calls itself recursively without ever terminating.
+% Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
+% alternated between recursive calls. The region-annotated definition of
+% $\C{alt}$ is shown below:
+% \begin{codejava}
+% unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
+%                           A<$\rho_3$> x, A<$\rho_4$> y) {
+%   return p.fst = x; p.snd = y; 
+%          this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
+% }
+% \end{codejava}
+% Note that, to avoid clutter, we have already resolved appropriate
+% region arguments to the recursive call, instead of introducing new
+% region variables and generating equality constraints on them. Rest of
+% the validity constraints generated while elaborating $\C{alt}$ to the
+% above region-annotated definition are shown below:
+% \begin{center}
+% \(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
+% $\quad$
+% \(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
+% $\quad$
+% \(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
+% \(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
+% \end{center}
+% The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
+% Second and third are generated by the the assignment expressions, and
+% the last constraint is generated from the recursive call.
 
 \subsection{Constraint Solving}
-\label{sec:csolve}
+\label{sec:fb-constraintsolving}
 
-Although we collect constraints for the entire program during
-elaboration, we solve the constraints in a compositional fashion.
-Concretely, this means that our constraint solving algorithm visits
-and solves every constraint (or, every set of mutually dependent
-constraints) only once. It composes computed solutions to solve other
-constraints that depend on the solved constraints. Importantly, the
-failure to solve a dependent constraint does not result in
-backtracking. We now describe our constraint solving algorithm by
-demonstrating its performance over the example constraints shown in
-the previous section. A formal account of the algorithm can be found
-in the supplement.
-% To simplify the presentation of our algorithm, we assume that are no
-% mutually recursive definitions in the source program. Recursive
-% definitions are nonetheless allowed.
+Our constraint generation algorithm traverses the entire program,
+performing elaboration and collecting constraints, which are
+subsequently solved en masse. The motivation behind the whole-program
+approach to constraint generation is twofold: it simplifies
+elaboration functions and makes presentation easier, and second, it
+naturally generalizes to mutual recursion. Nonetheless, we do not
+intend our type inference to be a whole-program analysis for (a). it
+preempts opportunities for separate compilation and dynamic linking,
+and (b). it is expensive and an overkill in most practical cases. We
+therefore reclaim the compositionality of type inference by solving
+the constraints in a compositional fashion. In more practical terms
+this means that our constraint solving algorithm visits and solves
+every constraint (or, every set of mutually dependent constraints)
+only once. It composes computed solutions to solve other constraints
+that depend on the solved constraints. Importantly, the failure to
+solve a dependent constraint does not result in backtracking. We now
+describe our compositional constraint solving algorithm in detail. To
+simplify the presentation of our algorithm, we assume that are no
+mutually recursive definitions in the source program. Recursive
+definitions are nonetheless allowed.
 
 \paragraph{Terminology} In a validity constraint, a predicate variable
-occurring on the left side (resp. right side) of the turnstile is said
-to occur negatively (resp. positively). A validity constraint
-\emph{constrains} the set of predicate variables that occur negatively
-in the constraint, while it \emph{uses} the set of predicate variables
-that occur positively. A constraint is said to be \emph{recursive} if
-it constrains and uses a predicate variable.
-% A pair of validity constraints are said to be \emph{mutually
-% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
-% predicate variables that occur with opposite polarities in both the
-% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
-% predicate variables are also said to be mutually dependent. We call
-% the transitive closure of mutual dependency relation as
-% \emph{transitive dependency} relation. Like mutual dependency,
-% transitive dependency is also extended to predicate variables.
+occuring on the left side of the turnstile is said to occur
+negatively, or with \emph{negative polarity}. In contrast, a predicate
+variable occuring on the right side is said to occur positively, or
+with \emph{positive polarity}. A validity constraint
+\emph{constrains} the set of predicate variables that occur
+negatively in the constraint, while it \emph{uses} the set of
+predicate variables that occur positively. A constraint is said to be
+\emph{recursive} if it constrains and uses a predicate variable.
+% A pair of validity
+% constraints are said to be \emph{mutually dependent} if there exists a
+% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
+% with opposite polarities in both the constraints. In such case, the
+% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
+% be mutually dependent. We call the transitive closure of mutual
+% dependency relation as \emph{transitive dependency} relation. Like
+% mutual dependency, transitive dependency is also extended to predicate
+% variables.
 
@@
 \begin{itemize}
-\item \textbf{Type 1} $c_1$ constrains a predicate variable that $c_2$ uses. 
-\item \textbf{Type 2} $c_1$ constrains a (non-strict) subset of predicate variables
-that $c_2$ constrains.
+\item $c_1$ constrains a predicate variable that $c_2$ uses. 
+\item $c_1$ constrains a (non-strict) subset of predicate variables
+that $c_2$ constraint.
 \end{itemize}
-
-% The first condition intuitively corresponds to a case, where
-% expression or type, whose region elaboration is constrained by $c_2$
-% refers to a method or a class, whose unknown precondition is
-% constrained by $c_1$. The common predicate variable ($\varphi$)
-% represents the unknown precondition in this case. The dependency from
-% $c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
-% before $c_2$ is solved, thus enforcing the rule that the precondition
-% of a method must not depend on its calling context. 
+The first condition intuitively corresponds to a case, where
+expression or type, whose region elaboration is constrained by $c_2$
+refers to a method or a class, whose unknown precondition is
+constrained by $c_1$. The common predicate variable ($\varphi$)
+represents the unknown precondition in this case. The dependency from
+$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
+before $c_2$ is solved, thus enforcing the rule that the precondition
+of a method must not depend on its calling context. 
 % Note that the first condition results in self-loops over recursive
@@
 % (and self-loops over recursive constraints). 
-% The second condition captures two kinds of dependencies. First, the
-% dependency from the constraints over a method precondition to the
-% constraints over the precondition of the class containing the method.
-% This captures our preference that the constraints over a class's
-% region parameters should not depend on the idiosyncracies of its
-% methods. Any additional constraints required by any of its methods
-% must be captured in the precondition of the method itself
-% (well-formedness rules allow this possibility). The second condition
-% adds bidirectional dependencies between validity constraints that
-% constraint the same set of predicate variables. 
+The second condition captures two kinds of dependencies. First, the
+dependency from the constraints over a method precondition to the
+constraints over the precondition of the class containing the method.
+This captures our preference that the constraints over a class's
+region parameters should not depend on the idiosyncracies of its
+methods. Any additional constraints required by any of its methods
+must be captured in the precondition of the method itself
+(well-formedness rules allow this possibility). The second condition
+adds bidirectional dependencies between validity constraints that
+constraint the same set of predicate variables.
+
 Next, we convert the dependency graph over constraints into a
@@
 are applied to the constraints in subsequent sets before attempting to
-solve them. For the DAG in figure above, we consider the topological
-order $[C_1, C_2, C_3]$ of its transpose, and solve the sets of
-constraints in that order.
+solve them.  Consequently, when the turn of a constraint set ($C$)
+arrives during the constraint solving process, it satisfies certain
+properties:
+\begin{itemize}
+\item There exists only one predicate variable ($\varphi$) that is either
+constrained or used by the constraints in the set ($C$). The variable is
+called set's \emph{subject}. This property follows from (a). the fact
+that all the dependency constraints have already been solved (and
+solutions applied), and (b). the assumption that there are no mutually
+recursive definitions. 
+\item All the constraints that constrain the set's subject are present
+in the set. This follows from our definition of the dependency relation.
+\end{itemize}
 
-% Consequently, when the turn of a constraint set ($C$) arrives during
-% the constraint solving process, it satisfies certain properties:
-% \begin{itemize}
-% \item There exists only one predicate variable ($\varphi$) that is either
-% constrained or used by the constraints in the set ($C$). The variable is
-% called set's \emph{subject}. This property follows from (a). the fact
-% that all the dependency constraints have already been solved (and
-% solutions applied), and (b). the assumption that there are no mutually
-% recursive definitions. 
-% \item All the constraints that constrain the set's subject are present
-% in the set. This follows from our definition of the dependency relation.
-% \end{itemize}
+For the DAG in figure above, we consider the topological order $[C_1,
+C_2, C_3]$ of its transpose, and solve the sets of constraints in that
+order.  Solving the set ($C$) of constraints entails finding an
+assignment for $C$'s subject ($\varphi$), and also all the region
+variables ($\rhobar$) that occur free in $C$, such that the solution
+satisfies well-formedness constraints on $\varphi$ and $\rhobar$. To
+simplify presentation, we think of $C$ as being parameterized on
+$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
+\rbrack$. We now formalize the constraint satisfaction problem, and
+its solution.
 
-Solving a set ($C$) of constraints involves finding assignments for
-free region ($\rho$) and predicate ($\varphi$) variables in the set.
-Assignments to region variables can be found via unification (for the
-examples in \S~\ref{sec:constraints}, we have already performed the
-unification, so $C_{1-3}$ do not contain free region variables). Once
-the region variables have been unified, we solve $C$ to find an
-assignment for free predicate variables. The first step is to reduce
-the set of constraints into an equivalent single constraint by
-conjoining antecedents and consequents of all constraints in the
-set\footnote{Supplement contains formal justification for this step}.
-For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
-constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
-below:
+\begin{definition}
+\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
+satisfaction problem is a tuple $\CSP$, where $C\lbrack
+\varphi,\rhobar \rbrack$ is a set of validity constraints, where
+$i$'th validity constraint assumes one of the following forms:
+\begin{center}
+\(
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {\phicstr^{i}}\qquad
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {F_i(\varphi)}
+\)
+\end{center}
+$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
+union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
+unification domain ($\rhoenv$) of the CSP.  The solution to the
+constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
+$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
+$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
+constraint formula such that:
+\begin{itemize}
+\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
+$\tywf{\rhoenv_{\varphi}}{\phisol}$).
+\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
+is valid.
+% \begin{center}
+% \(
+%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
+%           {\phicstr^{i} \conj  F_i(\phisol))}
+% \)
+% \end{center}
+\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
+that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
+\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
+(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
+$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
+\end{itemize}
+\end{definition}
+
+\subsection{Solving the CSP}
+
+The first step of solving the CSP $\CSP$ is to cast it as an
+equivalent problem $\csp$ involving a single validity constraint. The
+single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
+\begin{center}
+\(
+  \isvalid{ \Phicx \conj \varphi}
+        {\Phics \conj  \bigwedge_i F_i(\varphi)}
+\)
+\end{center}
+Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
+\phicstr^{i}$. To see why both CSPs are equivalent, consider two
+distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
+$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
+\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
+generated before $c_j$ by the constraint generation algorithm. Observe
+that when our constraint generation algorithm generates a constraint,
+all relationships between concrete region identifiers referred by the
+consequent of constraint are already present in its antecedent (this
+includes the unification domains of region variables referred by the
+consequent).  Since no new relationships between existing regions are
+added when a new constraints is generated, $\phictxt^j$ either
+describes the same relationships that are already described by
+$\phictxt^i$, or describes relationships among region identifiers that
+are new, and not relevant to $c_i$. Therefore, strengthening the
+context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
+nor strengthens $c_i$ (resp.  $c_j$).
+
+For the constraint sets $C_1$, $C_2$, and $C_3$ in the running
+example, equivalent constraints are $c_{12}$, $c_{13}$, and $c_{14}$,
+respectively, shown below:
 \begin{smathpar}
@@
 \end{smathpar}
-The constraint $c_{12}$ is a non-recursive constraint of the form
-$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
-$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
-this case). Note that the constraint describes an abduction problem,
-whose solution is the weakest $\phisol$ such that $\isvalid{\phictxt
-\conj \phisol}{\phicstr}$. We solve the problem by casting it as an
-instance of graph augmentation problem~\cite{siam92}. Our intuition is
-based on the observation that a constraint can be viewed as a directed
-graph with region variables as vertices and outlives relationships
-between them as edges (equality is encoded as a conjunction of
-symmetric outlives relationships).  Let $G(\phictxt)$ and
-$G(\phicstr)$ represent such graphs for $\phictxt$ and $\phicstr$,
-respectively. Now, $\isvalid{\phictxt}{\phicstr}$ if and only if every
-pair $(\rho_1,\rho_2)$ of connected vertices in $G(\phicstr)$ are also
-connected in $G(\phictxt)$. If $\rho_1$ and $\rho_2$ are not connected
-in $G(\phictxt)$, then it means that $\isvalid{\phicstr}{\rho_1
-\outlives \rho_2}$ but $\isnotvalid{\phictxt}{\rho_1 \outlives
-\rho_2}$. Since $\phisol$ needs to satisfy $\isvalid{\phictxt \conj
-\phisol}{\phicstr}$, it must be the case that $\rho_1$ and $\rho_2$
-are connected in $G(\phictxt) \cup G(\phisol)$. Since $\phisol$ must
-be weakest such formula, it can be obtained by computing the minimum
-number of edges that need to be added to $G(\phictxt)$ such that it is
-``as connected as'' $G(\phicstr)$. Using this technique, we obtain the
-following solutions to $\varphi_0$ and $\varphi_1$ for constraints in
-the running example:
+
+\subsubsection{Non-Recursive Constraints}
+\label{sec:csolve}
+
+We first describe how we solve a non-recursive CSP $\csp$ where
+$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
+\conj \varphi}{\Phics}$.
+
+Our approach is based on the observation that $\FB$ does not admit
+null values or uninitialized variables, thus forcing every region
+variable to be unified with some concrete region. The constraint
+formula $\Phics$ captures all such unification constraints on
+$\rhobar$. Since the set of all concrete region identifiers is the
+unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
+with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
+exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
+\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
+in which case the well-formedness constraint on $\rho_i$ is not
+satisfied, and constraint solving must fail.  Therefore, there exists
+a unique assignment $\substFn$ such that $c\lbrack
+\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
+$\phi_{sol}$.
+
+To obtain $\phisol$, we make use of another observation. Let $c\lbrack
+\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
 \begin{center}
 \(
-  \varphi_0 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+  \isvalid{ \Phicx \conj \varphi}
+        {\Phics'}
+\)
+\end{center}
+Consider a maximally weak formula $\phi$ such that
+$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
+% Since $\phi$ is
+% maximally weak, there does not exist a $\phi'$ satisfying all the
+% below three conditions:
+% \begin{itemize}
+% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
+% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
+% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
+% \end{itemize}
+Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
+\subseteq \rhoenv$, we have two cases:
+\begin{itemize}
+
+\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
+a solution to $\varphi$. 
+% Furthermore, it is the weakest solution, because it is equivalent to
+% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
+% \substFn(\rho_j))$, and any other solution has to stronger than
+% $\Phics$ under the same context.
+
+\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
+contains at least one equality or outlives constraint on two region
+identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
+\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
+\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
+constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
+is not maximally weak). Let us denote such constraint on $\rgn_i$ and
+$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
+$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
+Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
+$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
+$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
+\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
+this is impossible. To see why, recall that all the constraints on
+identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
+by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
+Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
+only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
+solution $\phisol$ cannot exist.
+\end{itemize}
+The above discussion hints at an algorithm to compute a solution to
+$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
+{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
+{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
+solution to $\varphi$. We now describe a graph-based algorithm to
+compute maximally weak $\phisol$.
+
+\paragraph{Definitions} Given a constraint formula $\phi$, we define
+its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
+vertices ($V(\phi)$) are free region variables and identifiers in
+$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
+in\footnote{We alternate between viewing a constraint formula ($\phi$)
+as a set of constraints and a conjunction of constraints in this
+description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
+\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
+$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
+conjunction of symmetric outlives constraints for the purpose of graph
+encoding. Conversely, given a digraph $G$, we define its constraint
+encoding $\Phi(G)$ in a straightforward manner. We say that a graph
+$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
+\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
+exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
+exist a path between same pair of vertices in $G_1$.
+
+A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
+\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
+graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
+connected as $G(\phicstr)$. The problem of computing such a $G$ is
+equivalent to the problem finding of finding minimum number of edges
+to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
+Efficient algorithms to solve the later problem are known to
+exist~\cite{graph}.
+
+Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
+graph augmentation problems, as described above, results in the
+following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
+outlives constraints are replaced with equalities):
+\begin{center}
+\(
+  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
@@
 \(
-  \varphi_1 \Leftrightarrow \rho_5 = \rho_0
+  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
 \)
 \end{center}
+
+\subsubsection{Solving Recursive Constraints}
+
 Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
@@
 \end{center}
-Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and $F$ is
-a substitution function, not necessarily idempotent. Our solution to the
-constraints of the above form is based on the observation that $H(\varphi) =
-\phicstr \conj F(\varphi)$ is a monotone over the complete meet semi-lattice
-of all possible constraints over the domain of region variables in the
-program. Hence, by Tarski's fixpoint theorem~\cite{tarski}, $H$ has a
-greatest fixed point. To solve the recursive constraint, we first compute the
-greatest fixed point ($\phi_f$) for $H$, and then solve the non-recursive
-constraint $\isvalid{\phictxt \conj \varphi}{\phi_f}$ using the graph
-approach described above. Using this approach, we solve the recursive
-constraint $c_{14}$, to obtain the following solution to $\varphi_2$:
+Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
+$F$ is a substitution function, not necessarily idempotent. We now
+extend constraint solving to recursive CSP $\csp$, where $c\lbrack
+\varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx \conj
+\varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake of
+brevity, we define $G(\varphi) = \Phicx \conj \bigwedge_i
+F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
+\begin{center}
+\(
+  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
+\)
+\end{center}
+To solve the above recursive constraint, we start with the observation
+that the set of all constaints ($\phi$) over $\rhoenv \cup
+\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
+\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
+monotone over the lattice:
+\begin{center}
+  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
+  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
+\end{center}
+From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
+point ($\phi_f$), with following properties:
+\begin{itemize}
+\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
+$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
+\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
+G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
+{\cdot} {\phi_f' \Rightarrow \phi_f}$.
+\end{itemize}
+We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
+convert the recursive constraint to the following non-recursive
+constraint:
+\begin{center}
+\(
+  \isvalid{\Phicx \conj \varphi}{\phi_f}
+\)
+\end{center}
+The technique described in \S~\ref{sec:csolve} now suffices to solve
+the above constraint.
+
+Let $H(\varphi_2)$ denote the consequent of the recursive constraint
+$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
+\begin{center}
+\(
+  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
+  \rgn_0 = \rgn_0
+\)
+\end{center}
+Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
+\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
+results in the following solution for $\varphi_2$ that meets its
+well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
+\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
 \begin{center}
@@
 
-% If such $\phisol$ satisfies the
-% well-formedness condition on $\varphi$, then it is clearly a solution
-% for $\varphi$, because there cannot exist a weaker solution for
-% $\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
-% maximally weak $\phisol$ fails the well-formedness condition, then we
-% establish~\cite{techrep} that it is impossible to satisfy the
-% constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
-% that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
-% and check if it satisfies the well-formedness condition on $\varphi$.
-
-% \vspace*{-0.05in}
-% \paragraph{Definitions} Given a constraint formula $\phi$, we define
-% its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
-% vertices ($V(\phi)$) are free region variables and identifiers in
-% $\phi$, and whose edges ($E(\phi)$) denote outlives constraints
-% in\footnote{We alternate between viewing a constraint formula ($\phi$)
-% as a set of constraints and a conjunction of constraints in this
-% description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
-% \outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
-% $(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
-% conjunction of symmetric outlives constraints for the purpose of graph
-% encoding. Conversely, given a digraph $G$, we define its constraint
-% encoding $\Phi(G)$ in a straightforward manner. We say that a graph
-% $G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
-% \subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
-% exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
-% exist a path between same pair of vertices in $G_1$.
-
-% A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
-% \Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
-% graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
-% connected as $G(\phicstr)$. The problem of computing such a $G$ is
-% equivalent to the problem finding of finding minimum number of edges
-% to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
-% Efficient algorithms to solve the later problem are known to
-% exist~\cite{graph}, and we rely on them for the solution.
-
-% Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
-% graph augmentation problems, as described above, results in the
-% following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
-% outlives constraints are replaced with equalities):
+% Note that $F$ is a non-idempotent substitution function with following
+% properties:
+% \begin{itemize}
+% \item $F$ is a monotone: 
 % \begin{center}
-% \(
-%   \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
-%      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
-%      \rho_2 = \rho_0 \conj \rho_3 = \rho_1
-% \)\\
-% \(
-%   \varphi_2 \Leftrightarrow \rho_5 = \rho_0
-% \)
-% \end{center}
-% Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
-% $c_{14}$ gives us a recursive constraint of the following form:
-% \begin{center}
-% \(
-%   \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
-% \)
-% \end{center}
-% Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
-% $F$ is a substitution function, not necessarily idempotent. Let
-% $G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
-% $\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
-% recursive constraints  is based on some observations. Firstly, the set
-% of all possible constraints over a finite set of region
-% variables/identifiers is a complete meet semi-lattice, where (a).
-% meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
-% \conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
-% \triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
-% function $H$ is a monotone over the lattice since $\forall
-% \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
-% \Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
-% theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
-% that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
-% the recursive constraint shown above, we first compute the greatest
-% fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
-% following non-recursive constraint:
-% \begin{center}
-% \(
-%   \isvalid{\phictxt \conj \varphi}{\phi_f}
-% \)
-% \end{center}
-% Which can then be solved via the graph-based approach described
-% before.
-
-% Let $H(\varphi_2)$ denote the consequent of the recursive constraint
-% $c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
-% \begin{center}
-% \(
-%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
-%   \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
-%   \rgn_0 = \rgn_0
-% \)
-% \end{center}
-% Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
-% \conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
-% results in the following solution for $\varphi_2$ that meets its
-% well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
-% \rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
-% \begin{center}
-% \(
-%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
-%   \rho_9 \conj \rho_8 = \rho_9
-% \)
+%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
+%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
 % \end{center}
 
-\subsection{Higher-Order Type Inference}
+% \item $F$ distributes over conjunction (mind the syntactic equality):
+% \begin{center}
+%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
+%   F(\phi_2)$
+% \end{center}
+% \end{itemize}
 
-% The region type system for $\FB$ admits function types with
-% higher-rank region polymorphism, allowing region variables to be
-% quantified in the types of higher-order arguments. This allows, for
-% example, a user-defined function to be applied on arguments from
-% different regions in different contexts. However, like most type
-% systems, our region type system loses principal typing property in
-% presence of higher-rank polymorphism, which means that there can be
-% two possible types for a function, where neither one is more general
-% than the other. Consequently, it is not possible to infer a unique
-% type for a function. Nonetheless, it is possible to engineer type
-% inference to to consistently prefer one type over the other.
-Our type inference algorithm does not straightforwardly generalize to
-the higher-order case, due to the lack of principal typing in presence of
-higher-rank polymorphism. Fortunately, higher-order type inference, in
-its full generality, is rarely needed in the context of \name, where
-higher-order control flow is predominantly due to the second-order
-arguments to constructors (For \eg, user-defined functions in dataflow
-queries, and lambda arguments to the $\RgnZ$ class constructor). We
-therefore extend our inference algorithm with a heuristics tailor-made
-for the aforementioned second-order cases. The simplistic approach
-seems to suffice in our experience, but, it is always possible that it
-makes the type checker reject a program that is otherwise valid. In
-such cases, programmer can help the type checker by writing region
-annotations only over higher-order arguments.  Nonetheless, we note
-that it is possible to infer most general (if not principal) types for
-higher-order arguments, by extending our constraint solver with an
-approach modeled after~\cite{gulwani09}, which lets it compute maximally
-weak/strong solutions to multiple competing $\varphi$'s
-simultaneously.
+% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
+% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
+% $\phi_f$ be the GFP of $G$. Following properties ensue:
+% \begin{itemize}
+% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
+% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
+% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
+% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
+% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
+% \end{itemize}
 
-% \begin{codejava}
-% bool app2(A x, A y, (A $\rightarrow$ bool) f) 
-%     { return (f x && f y); }
-% \end{codejava}
-% There are two possibilities for the region type of \C{app2}: one,
-% where it allows \C{x} and \C{y} to be allocated in different regions,
-% but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
-% arguments from any region), and second, where it requires \C{x} and
-% \C{y} to be allocated in the same region, and allows \C{f} to be
-% monomorphic. None of the two types are more general than the other,
-% and it is not clear which one is more preferable. The concrete
-% manifestation of this dilemma is a constraint of form 
-% $\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
-% no additional constraints to help determine $\varphi_1$ or $\varphi_2$
-% before solving this constraint. Since it is always possible to
-% strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
-% does not exist unique solution to such constraints. Nonetheless, it is
-% possible to bias the constraint-solving algorithm to consistently
-% prefer one solution over the other.
+% \begin{theorem}
+% $\phi_f$ is a solution.
+% \end{theorem}
+% \begin{proof}
+% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
+% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
+% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
+% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
+% solution for $\varphi$.
+% \end{proof}
 
-% In context of \name, however, we adopt a trivial solution of
-% assigning the same region to the arguments and return values in
-% (higher-order) function types. For user-defined functions (e.g.,
-% predicate in the \C{WHERE} clause of the \C{SELECT} dataflow query)
-% and lambda arguments to the $\RgnZ$ class constructor, which are the
-% prominent uses of higher-order functions in \name, the simplistic
-% approach seems to suffice in practice. However, it is always possible
-% that this approach makes type checker reject a program that is
-% otherwise valid. In such cases, programmer can help the type checker
-% by writing region annotations for only the higher-order functions.
+% \begin{theorem}
+% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
+% \phisol}{\phi_f}$
+% \end{theorem}
+% \begin{proof}
+%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
+%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
+%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
+%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$
 
-% \begin{center}
-%   \C{app2} : $\inang{\rho_0,\rho_1}$(\C{A}$\inang{\rho_0}$, 
-%               \C{A}$\inang{\rho_0}$,
-%               $\inang{\rho_2}$\C{A}$\inang{\rho_2}$ $\rightarrow$
-%               \C{bool}) $\rightarrow$ \C{bool}
-% \end{center}
+% \end{proof}

### 2015-11-20 paper/type-system.tex
Merged paper to techrep.

@@
 \label{sec:type-system}
+
+\input{fb-auxdef}
 
@@
 \input{fb-staticsem}
+\input{fb-morewfrules}
 
@@
 ($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
-\tau$).
+\tau$). Note that any of the $\tau$'s in the arrow type can themselves
+be region-parametric arrow types. In this respect, our region type
+system is comparable to System F's type system, which admits
+higher-rank parametric polymorphism. 
 % at the prenex position, similar to ML type schemes. However,
@@
 % instantiate and generalize region variables. 
-Since a function closure can escape the context in which it is
-created, it is important to keep track of the region in which it is
-created in order to avoid unsafe dereferences. The $\rgn$ annotation
-above the arrow in the arrow type denotes the allocation region of the
-corresponding closure. 
-% Note that it is important to distinguish
-% between the allocation context argument ($\rhoalloc$) of a function
-% and the allocation region ($\rgn$) of its closure. In \name, the later
-% corresponds to the region where a \C{Func} object is allocated, while
-% the former corresponds to the region where it is applied
-. 
-% For instance, in the following example:
-% \begin{codejava}
-% letregion $\rgn$ {
-%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
-%   in f
-% }
-% \end{codejava}
-% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
-% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
-% in $\rgn$, and (b). when executed under an allocation context
-% $\rho^a$, the closure returns an object allocated in $\rho^a$.
+Note that a lambda expression creates a closure, which can escape the
+context in which it is created. It is therefore important to keep track of
+the region in which a closure is allocated in order to avoid unsafe
+dereferences. The $\rgn$ annotation above the arrow in the arrow
+type denotes the allocation region of the corresponding closure. Note
+that it is important to distinguish between the allocation context
+argument ($\rhoalloc$) of a function and the allocation region
+($\rgn$) of its closure. In \name, the later corresponds to the region where
+a \C{Func} object is allocated, while the former corresponds to the
+region where it is applied.  For instance, in the following example:
+\begin{codejava}
+letregion $\rgn$ {
+  let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
+  in f
+}
+\end{codejava}
+The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
+\ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
+in $\rgn$, and (b). when executed under an allocation context
+$\rho^a$, the closure returns an object allocated in $\rho^a$.
 
@@
 is considered to have a certain region type, respectively.
-Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
-rules\footnote{The full formal development can be found in the
-supplement}. The rules refer to a context ($\A$), which is a tuple of:
+Fig.~\ref{fig:fb-staticsem} contains the entire set of the rules. The
+rules refer to a context ($\A$), which is a tuple of:
 \begin{itemize}
@@
 definitions in \FB.
-% \begin{figure}
-
-% \begin{codejava}
-% class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
-%   <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
-%   Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
-%     super(); 
-%     this.thunk = thunk;
-%   }
-%   unit transfer$\inang{\rho}$(unit u) {
-%     return u;
-%   }
-% }
-% \end{codejava}
-
-% \caption{The stub used for $\RgnZ$ class}
-% \label{fig:region-stub}
-% \end{figure}
 
@@
 region annotations on types, and region arguments in expressions
-($\absof{\cdot}$ denotes the region erasure operation). 
-% For judgments in FGJ, we use a
-% class table that maps class names to their region-erased definitions.
-% We denote the region-erased class table as ($\absof{CT}$). The precise
-% semantics of region erasure ($\absof{\cdot}$) are defined
-% in~\cite{techrep}, but it can be understood as an operation that
-% erases region annotations on types. The class table is assumed to bind
-% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
-% This mostly eliminates the need to treat $\RgnZ$ objects, which double
-% up as transferable region handlers, as a special case in semantics.
-The well-formedness rule for class types
+($\absof{\cdot}$ denotes the region erasure operation).  The
+well-formedness rule for class types
 ($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
@@
 references between transferable regions, which become unsafe in
-context of the recipient's address space. Fortunately, such
-references are explicitly prohibited by the type rule of \C{\RgnZ}
+context of the recipient's address space. Fortunately, such references
+are explicitly prohibited by the type rule of \C{\RgnZ}
 objects, as described below.
@@
 
+\input{fb-opsem}
+
 The type rules distinguish between the \C{new} expressions that create
@@
 \C{letregion} expression contains no references to the temporary
-objects inside the region\footnote{The supplement contains a formal
+objects inside the region\footnote{The appendix contains a formal
 proof of this claim}.
@@
 
-The supplement defines a small-step operational semantics for \fbname
-via a five-place reduction relation:
+Fig.~\ref{fig:fb-opsem} defines a small-step operational semantics for
+\fbname via a five-place reduction relation:
 \vspace*{-0.15in}
@@
 identifier ($\rgn$) for this region. A binding is also added to
-$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
-defines a type rule for such values, allowing them to be typed.
-Type safety theorem is now stated thus:
+$\rhomap$, mapping $\rgn$ to the ``closed'' state.
+Fig.~\ref{fig:fb-staticsem} defines a type rule for such values,
+allowing them to be typed.  Type safety theorem is now stated
+thus\footnote{Theorems~\ref{thm:fb-type-safety-stmt}
+and~\ref{thm:fb-transfer-safety-stmt} are restated and proved in the
+appendix}:
 \begin{theorem}
 \emph{(\textbf{Type Safety})}
-\label{thm:fb-type-safety}
+\label{thm:fb-type-safety-stmt}
 $\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
@@
 \emph{(\textbf{Transfer Safety})}
-\label{thm:fb-type-safety}
+\label{thm:fb-transfer-safety-stmt}
 $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,

### 2015-11-20 paper/abstract.tex
small things in intro.

@@
-Recent work in high-performance systems written in managed languages
+Recent work in high-performance 
+systems written in managed languages 
 (such as Java or C\#) has shown that garbage-collection can be a
-significant performance bottleneck in such systems.  A class of these
-systems, focused on big-data, create huge data-structures with
+significant performance bottleneck.  A class of these
+systems, focused on big-data, create many and often large data-structures with
 well-defined lifetimes.  In this paper, we present a language and

### 2015-11-20 paper/introduction.tex
small things in intro.

--- 2015-11-19 paper/introduction.tex
@@
 Furthermore, in a distributed dataflow system, the GC pause at one node can
-have a cascading adverse effect on the performance of other nodes.
-For instance, a GC pause at an upstream actor can block
+have a cascading adverse effect on the performance of other nodes, particularly
+when real-time streaming performance is required.
+\dv{citation missing here, Tim Haris's work and our HotOS submission.}
+In particular, a GC pause at an upstream actor can block
 downstream actors that are waiting for messages. 
-Much of the GC overhead results from the collector performing avoidable or unproductive work.
+However, often much of the GC overhead results from the collector performing avoidable or unproductive work.
 For example, in the process executing the code from
@@
 
-An important observation, in the context of processes of kind
+An important observation, in the context of processes of the kind
 described above, is that the data-structures exchanged between them
@@
 pointers in another dynamic region (\eg, the output-message).
-However, this is not sufficient! When processing the input-message
+However, this is not sufficient! \dv{for what} When processing the input-message
 $D$, we necessarily will have to create pointers that point to objects

### 2015-11-20 paper/introduction.tex
More small comments.

@@
   \item We describe an implementation of \name frontend in OCaml,
-  along with experimental results, and case studies where the region
+  along with experimental results, \dv{Experimental results of what?} and case studies where the region
   type system was able to identify unsafe memory accesses statically.

### 2015-11-20 paper/introduction.tex
Rewrote introduction.

--- 2015-11-20 paper/introduction.tex
@@
 This example is an instance of a general pattern, where a producer
-creates a data-structure and passes it to a consumer. In a system
+creates a data structure and passes it to a consumer. In a system
 where most of the computation takes this form, and these
-data-structures are very large, as is the case with many streaming
+data structures are very large, as is the case with many streaming
 big-data analysis systems, garbage collection overhead becomes
-significant~\cite{Broom:HotOS}. This is a cause of concern because,
-firstly, in a distributed dataflow system, GC overhead at one node can
-have compounding adverse effect on the performance of other nodes.
-This can happen, for instance, when a GC is triggered at an upstream
-actor that is about to send a message, while the downstream actors are
-blocked waiting for messages. Secondly, much of the GC overhead
-results from the collector performing avoidable or unproductive work.
+significant~\cite{Broom:HotOS}.
+Furthermore, in a distributed dataflow system, the GC pause at one node can
+have a cascading adverse effect on the performance of other nodes.
+For instance, a GC pause at an upstream actor can block
+downstream actors that are waiting for messages. 
+Much of the GC overhead results from the collector performing avoidable or unproductive work.
 For example, in the process executing the code from
 Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the \C{map}
-data-structure, although its objects cannot be collected until a
+data structure, although its objects cannot be collected until a
 suitable timing message arrives.
@@
 % For example, a consumer process may trigger a
-% GC as soon as it receives a large data-structure from a producer
+% GC as soon as it receives a large data structure from a producer
 % process, which has just finished its GC. In such case, consumer's GC
-% traverses the entire data-structure performing little or no work.
+% traverses the entire data structure performing little or no work.
 
-% as the collector scans these large data-structures For instance, in
+% as the collector scans these large data structures For instance, in
 % the above example, the list at \C{map[t]} cannot be collected until
@@
 An important observation, in the context of processes of kind
-described above, is that the data-structures exchanged between them
-have can be partitioned into sets of fate-sharing objects with common
+described above, is that the data structures exchanged between them
+can be partitioned into sets of fate-sharing objects with common
 lifetimes, which makes them good candidates for a region-based memory
@@
 region is a good fit for a set of fate-sharing objects.
-% that is used to realize a data-structure with a well-defined
+% that is used to realize a data structure with a well-defined
 % lifetime. 
-In this example, the output to be constructed for each time window $t$
-can be implemented using a separate region.
+In Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
+time window $t$ (i.e., \C{map[t]}) can be a separate region that is
+allocated when the first message with timestamp $t$ arrives, and
+deallocated after \C{map[t]} is transferred in \C{onNotify}.
 
@@
 been known for a long time. Manual region-based memory management
-suffers from the usual challenges, namely the potential for dangling
-references and lack of memory safety. Automatic region-based memory
+suffers from the usual drawbacks, namely the potential for invalid
+references and the consequent lack of memory safety. Automatic region-based memory
 management systems guarantee memory safety, but impose various
-restrictions.  Tofte and Talpin's system~\cite{tofte94,tofte97} , for
-example, uses lexically scoped regions.  At runtime, the set of all
-regions (existing at a point in time) forms a stack. Thus, the
-lifetimes of all regions must be well-nested: it is not possible to
-have two regions whose lifetimes overlap, with neither one's lifetime
-contained within the other.  Unfortunately, the data-structures in the
-above example do not satisfy this restriction (as the output messages
-for multiple time windows may be simultaneously live, without any
-containment relation between their lifetimes).  We refer to regions
-with lexically scoped lifetimes as \emph{stack regions} and to regions
-that do not have such a lexically scoped region as \emph{dynamic}
-regions.
+restrictions.  MLKit, which implements the approach pioneered by Tofte
+and Talpin~\cite{tofte94,tofte97}, for example, uses lexically scoped
+regions.  At runtime, the set of all regions (existing at a point in
+time) forms a stack. Thus, the lifetimes of all regions must be
+well-nested: it is not possible to have two regions whose lifetimes
+overlap, with neither one's lifetime contained within the other.
+Unfortunately, the data structures in the above example do not satisfy
+this restriction (as the output messages for multiple time windows may
+be simultaneously live, without any containment relation between their
+lifetimes).  We refer to regions with lexically scoped lifetimes as
+\emph{stack regions} and to regions that do not have such a lexically
+scoped region as \emph{dynamic} regions.
 
@@
 management technique that supports dynamic regions as first-class
-objects. Though we focus, in this paper, on memory-safety, our
-approach offers other benefits.  When a producer transfers a
-data-structure to a consumer, in a shared-memory context , our
-approach provides an ownership discipline that ensures that the
-producer cannot subsequently modify the data-structure. When the same
-transfer happens in a distributed setting, the system guarantees that
-the data-structure is self-contained and can be safely transmitted
-across address spaces.  We refer to dynamic regions that provide these
-guarantees as \emph{transferable} regions.
+objects. Our focus, in this paper, is on dynamic regions, which can be
+safely transferred across address spaces. We refer to such dynamic
+regions as \emph{transferable} regions. As with allocation and
+deallocation, transferring a memory region is fast, and therefore,
+transferring a data structure contained in a region is more efficient
+that traversing its objects in heap, and transferring them
+independently\footnote{Empirical studies in~\cite{Broom:HotOS} support
+this claim}. In the \C{SelectVertex} example from
+Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
+output for each time window $t$ must be transferable due to the
+\C{transfer} operation on Line 16. As it is the case with
+\C{SelectVertex}, the transferred data is no longer accessed by the
+producer, so the transfer operation in our system deallocates the
+region once the transfer is complete.
 
-The key memory safety property we wish to ensure is that there are no
-dangling references: i.e., a reference to an object that has been
-freed.  A key invariant and restriction that we utilize to achieve
-this is that no object $o_1$ in one dynamic region $D_1$ can contain a
-pointer to an object $o_2$ in another dynamic region $D_2$.  In cases
-where this is too restrictive, the user has to resort to (deep)
-copying $o_2$ to $D_1$.  This is conceptually quite reasonable given
-the perspective that $D_1$ is a self-contained data-structure.
-Furthermore, this is quite analogous, from a performance perspective,
-to the copying that happens when a garbage collector promotes an
-object from a lower generation to a higher generation.  From this
-perspective, a region may be seen as a user-controlled generation.
+With respect to memory safety, the key property we wish to ensure is
+that there are no invalid references: i.e., references to objects that
+were deallocated, or simply never existed. Transfer operation, with no
+additional checks, may cause memory safety violations, both at the
+producer of the data structure, and its (possibly remote) consumer. At
+the producer, any existing references into the data structure become
+invalid post transfer. If the data structure contains references to
+objects outside its (transferable) region, then such references become
+invalid in the context of the consumer. Safety violations of this kind
+are particularly unwelcome as the program with GC did not have them to
+begin with. Note that, while the references of later kind (i.e.,
+references that escape a region), defeat the very purpose of a
+transferable region, hence need to be prohibited, references of the
+former kind (i.e., references into a region from outside) are not at
+odds with the concept of a transferable region, hence need to be
+permitted. In fact, allowing such references is crucial to
+performance, as any non-trivial program creates temporary objects, and
+it is absurd to allocate them in a transferable region; such regions
+are meant for the data being transferred. Since transferable regions
+are first class objects in our setting, controlling references to and
+from such regions while ensuring their safety without significantly
+diluting the performance advantage of regions over GC is a challenging
+exercise.
 
-This ensures that we can safely free one dynamic region (\eg, the
-input-message in our example) without affecting the validity of
-pointers in another dynamic region (\eg, the output-message).
-However, this is not sufficient! When processing the input-message
-$D$, we necessarily will have to create pointers that point to objects
-that are internal to $D$. Such pointers may be stack-allocated or even
-reside in the heap (\eg, consider the iterator object used to iterate
-over the list in the input-message).  How do we ensure the safety of
-these pointers?
+In this paper, we describe an approach that restores memory safety in
+presence of transferable regions through a combination of a static
+typing discipline and lightweight runtime checks. Our approach is
+based on the following observations. Firstly, we observe that by
+nesting a TT-style \C{letregion} lexical block, that delimits the
+lifetime of a stack region, inside an \C{open} lexical block for
+transferable regions, which ``opens'' a transferable region and
+guarantees that the region won't be transferred/freed inside the
+block, we can guarantee that the transferable region will remain live
+as long as the stack region is live. We say that the former
+``outlives'' the later\footnote{We borrow the \emph{outlives} relation
+from~\cite{MIT03}. A comparison of our approach with~\cite{MIT03} can
+be found in \S~\ref{sec:related-work}.}, and any references from the stack
+region to the transferable region are therefore safe. 
+% In general, we can allow references from a region
+% $R_0$ to another region $R_1$, provided that we establish that $R_1$ outlives
+% $R_0$. 
+Second, we note that by controlling the ``outlives'' relationships
+between various regions, we only allow safe cross-region references,
+while prohibiting unsafe ones. In the above example, an outlives
+relationship \emph{from} the stack region \emph{to} the transferable
+region means that the references in that direction are allowed, but
+not the references in the opposite direction. In contrast, if an
+\C{open} block of a transferable region $R_0$ is nested inside an
+\C{open} block of another transferable region $R_1$, we do not
+establish any outlives relationships, thus declaring our intention to
+not allow any cross-region references between $R_0$ and $R_1$.  Third,
+outlives relationships are established based on the lexical structure
+of the program, hence a static type system can enforce them
+effectively. By assigning region types to objects, which capture the
+regions such objects are allocated in, and by maintaining outlives
+relationships between various regions, we can statically decide the
+safety of a reference.
 
-We use \emph{stack regions} (regions with lexically-scoped lifetimes),
-as well as the stack, for such temporary pointers (to objects inside a
-dynamic region) and use the following protocol to ensure safety of
-such pointers.  To work with a dynamic region $D$, we must first
-\emph{open} region $D$, to indicate that $D$ should not be freed
-during this period.  When the processing of the region is complete, we
-\emph{close} the region $D$, to indicate that it is safe to free $D$.
-In general, a dynamic region may be opened and closed multiple times
-before it is finally freed.
+Of course, the utility of our approach described above is predicated on
+the assumption that we can enforce certain invariants on transferable
+regions. Firstly, a transferable region cannot be transferred/freed
+inside an open block of that region (i.e, while it is still open).
+Secondly, a transferred/freed region cannot be opened. These are
+typestate invariants on the transferable region objects, which are
+hard to enforce statically due to the presence of unrestricted
+aliasing. Techniques like linear types and unique pointers can be used
+to restrict aliasing, but the constraints they impose are often hard
+to program around, and are not often acceptable to systems
+programmers. We therefore enforce typestate invariants at runtime via
+lightweight runt-time checks. In particular, we define an acceptable
+state transition discipline for transferable regions
+(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
+transition of a transferable region (\eg, from \emph{open} state to
+\emph{freed} state) is valid or not. The check is lightweight since it
+only involves checking a single tag that captures the current state.
+We believe that this is a reasonable choice since regions are
+coarse-grained objects manipulated infrequently, when compared to the
+fine-grained objects that are present inside these regions, for which
+safety is enforced statically. An added advantage of delegating the
+enforcement of typestate invariants to runtime is that our region type
+system is simple\footnote{Supplement lists all the type rules of our
+language, which occupy less than a page.}, which made it possible to
+formulate a type inference that completely eliminates the need to
+write region type annotations. This, we believe, significantly reduces
+the impediment to adopt our approach in practical setting.
 
-Our type-system permits the creation of pointers to internal objects
-of \emph{open dynamic} regions, but uses lexical scoping to ensure
-that such pointers are no longer live when the region is closed.
-Thus, such pointers may be either stack-allocated or reside in a
-region whose lifetime is statically guaranteed to be contained within
-the interval when $D$ is open.  As a consequence, the system ensures
-the strong invariant that for a \emph{closed dynamic} region $D$ no
-pointer from outside $D$ points to an internal object of $D$.
+% transferableally checking for the second category of
+% invariants is easy and does not pose a performance concern (unlike,
+% \eg, the first category of constraints); a static typechecker that
+% rules out the second category of violations would be very restrictive
+% from an expressiveness perspective; finally, the human effort required
+% to ensure the second set of invariants is not as burdensome (as
+% regions are coarse-grained objects manipulated infrequently), \eg, as
+% it would be for the first set of invariants.
 
-The final piece required to ensure memory-safety is ensuring that the
-program correctly follows the above-mentioned protocol for dynamic
-regions: \eg, an open region should not be freed and, dually, a freed
-region should not be opened.  (A complete typestate specification of
-the protocol is presented later.)
+% when a producer transfers a
+% region containing a data structure to a consumer, in a shared-memory
+% context, we enforce an ownership discipline that ensures that the
+% producer cannot subsequently modify the data structure. When the same
+% transfer happens in a distributed setting, we guarantee that (a). the
+% data structure is contained in the region, and can be safely
+% transmitted across address spaces, and (b). once the transfer
+% concludes, the producer can safely deallocate the memory for the
+% data structure without the fear of violating memory safety.
 
-Dynamic regions are first-class objects in our system. For instance,
-in our running example, we would like to use a dictionary that stores
-dynamic regions. However, this means that the program may create
-multiple aliases (pointers) to the same region, which, in turn, means
-that statically checking if the program correctly follows the dynamic
-region protocol is hard (undecidable, in fact). Safety can be ensured
-by using, \eg, linear typing to prevent the creation of aliases, but
-this would be quite restrictive. Hence, in our system, we dynamically
-check for adherence to this protocol.
+%  Transfer operation,
+% with no additional checks and balances, may cause memory safety
+% violations, both at the producer of the data structure, and the
+% (possibly remote) consumer of the data structure. A key invariant and restriction
+% that we utilize
+% to achieve this is that no object $o_1$ in one transferable region $D_1$
+% can contain a pointer to an object $o_2$ in another transferable region
+% $D_2$.  In cases where this is too restrictive, the user has to resort
+% to (deep) copying $o_2$ to $D_1$.  This is conceptually quite
+% reasonable given the perspective that $D_1$ is a self-contained
+% data structure.  Furthermore, this is quite analogous, from a
+% performance perspective, to the copying that happens when a garbage
+% collector promotes an object from a lower generation to a higher
+% generation.  From this perspective, a region may be seen as a
+% user-controlled generation.
 
-In summary, our approach reduces memory safety to two sub-problems:
-(a) Ensuring  low-level invariants about pointers to (fine-grained)
-objects inside regions and (b) Ensuring higher-level (typestate)
-invariants about the regions themselves.  We use a type system to
-statically ensure the first set of invariants. We use a dynamic check
-to ensure that the second set of invariants are satisfied at runtime.
-%
-This is a novel aspect of our system. We believe that this is a
-reasonable choice: dynamically checking for the second category of
-invariants is easy and does not pose a performance concern (unlike,
-\eg, the first category of constraints); a static typechecker that
-rules out the second category of violations would be very restrictive
-from an expressiveness perspective; finally, the human effort required
-to ensure the second set of invariants is not as burdensome (as
-regions are coarse-grained objects manipulated infrequently), \eg, as
-it would be for the first set of invariants.
+% This ensures that we can safely free one transferable region (\eg, the
+% input-message in our example) without affecting the validity of
+% pointers in another transferable region (\eg, the output-message).
+% However, this is not sufficient! When processing the input-message
+% $D$, we necessarily will have to create pointers that point to objects
+% that are internal to $D$. Such pointers may be stack-allocated or even
+% reside in the heap (\eg, consider the iterator object used to iterate
+% over the list in the input-message).  How do we ensure the safety of
+% these pointers?
+
+% We use \emph{stack regions} (regions with lexically-scoped lifetimes),
+% as well as the stack, for such temporary pointers (to objects inside a
+% transferable region) and use the following protocol to ensure safety
+% of such pointers.  To work with a transferable region $D$, we must
+% first \emph{open} the region $D$, to indicate that $D$ should not be
+% freed during this period. When the processing of the region is
+% complete, we \emph{close} the region $D$, to indicate that it is safe
+% to free $D$.  In general, a transferable region may be opened and
+% closed multiple times before it is eventually transferred or freed.
+
+% Our type-system permits the creation of pointers to internal objects
+% of \emph{open transferable} regions, but uses lexical scoping to ensure
+% that such pointers are no longer live when the region is closed.
+% Thus, such pointers may be either stack-allocated or reside in a
+% region whose lifetime is statically guaranteed to be contained within
+% the interval when $D$ is open.  As a consequence, the system ensures
+% the strong invariant that for a \emph{closed transferable} region $D$ no
+% pointer from outside $D$ points to an internal object of $D$.
+
+% The final piece required to ensure memory-safety is ensuring that the
+% program correctly follows the above-mentioned protocol for transferable
+% regions: \eg, an open region should not be freed and, dually, a freed
+% region should not be opened.  (A complete typestate specification of
+% the protocol is presented later.)
+
+% Transferable regions are first-class objects in our system. For instance,
+% in our running example, we would like to use a dictionary that stores
+% transferable regions. However, this means that the program may create
+% multiple aliases (pointers) to the same region, which, in turn, means
+% that statically checking if the program correctly follows the transferable
+% region protocol is hard (undecidable, in fact). Safety can be ensured
+% by using, \eg, linear typing to prevent the creation of aliases, but
+% this would be quite restrictive. Hence, in our system, we transferableally
+% check for adherence to this protocol.
+
+% In summary, our approach reduces memory safety to two sub-problems:
+% (a) Ensuring  low-level invariants about pointers to (fine-grained)
+% objects inside regions and (b) Ensuring higher-level (typestate)
+% invariants about the regions themselves.  We use a type system to
+% statically ensure the first set of invariants. We use a transferable check
+% to ensure that the second set of invariants are satisfied at runtime.
+% %
+% This is a novel aspect of our system. We believe that this is a
+% reasonable choice: transferableally checking for the second category of
+% invariants is easy and does not pose a performance concern (unlike,
+% \eg, the first category of constraints); a static typechecker that
+% rules out the second category of violations would be very restrictive
+% from an expressiveness perspective; finally, the human effort required
+% to ensure the second set of invariants is not as burdensome (as
+% regions are coarse-grained objects manipulated infrequently), \eg, as
+% it would be for the first set of invariants.
 
@@
 \emph{generics} (parametric polymorphism), with constructs to create,
-manage and destroy static and dynamic memory regions. Dynamic regions
+manage and destroy static and transferable memory regions. Transferable regions
 are first-class values in \name.

### 2015-11-20 paper/introduction.tex

@@
 
-This example is an instance of a general pattern, where a producer
-creates a data structure and passes it to a consumer. In a system
-where most of the computation takes this form, and these
-data structures are very large, as is the case with many streaming
-big-data analysis systems, garbage collection overhead becomes
-significant~\cite{Broom:HotOS}.
-Furthermore, in a distributed dataflow system, the GC pause at one node can
-have a cascading adverse effect on the performance of other nodes.
-For instance, a GC pause at an upstream actor can block
-downstream actors that are waiting for messages. 
-Much of the GC overhead results from the collector performing avoidable or unproductive work.
-For example, in the process executing the code from
-Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the \C{map}
-data structure, although its objects cannot be collected until a
-suitable timing message arrives.
+This example is an instance of a general pattern, where a producer creates a
+data structure and passes it to a consumer. In a system where most of the
+computation takes this form, and these data structures are very large, as is
+the case with many streaming big-data analysis systems, garbage collection
+overhead becomes significant~\cite{Broom:HotOS}.  Furthermore, in a
+distributed dataflow system, the GC pause at one node can have a cascading
+adverse effect on the performance of other nodes, particularly when real-time
+streaming performance is required~\cite{Broom:HotOS, harris15}.  In
+particular, a GC pause at an upstream actor can block downstream actors that
+are waiting for messages.  However, often much of the GC overhead results
+from the collector performing avoidable or unproductive work.  For example,
+in the process executing the code from Fig.~\ref{fig:motivating-eg}, GC might
+repeatedly traverse the \C{map} data structure, although its objects cannot
+be collected until a suitable timing message arrives.
 
@@
 
-An important observation, in the context of processes of kind
-described above, is that the data structures exchanged between them
+An important observation, in the context of processes of the kind
+described above, is that the data-structures exchanged between them
 can be partitioned into sets of fate-sharing objects with common
@@
 % it would be for the first set of invariants.
+% =======
+% objects. Though we focus, in this paper, on memory-safety, our
+% approach offers other benefits.  When a producer transfers a
+% data-structure to a consumer, in a shared-memory context , our
+% approach provides an ownership discipline that ensures that the
+% producer cannot subsequently modify the data-structure. When the same
+% transfer happens in a distributed setting, the system guarantees that
+% the data-structure is self-contained and can be safely transmitted
+% across address spaces.  We refer to dynamic regions that provide these
+% guarantees as \emph{transferable} regions.
+
+% The key memory safety property we wish to ensure is that there are no
+% dangling references: i.e., a reference to an object that has been
+% freed.  A key invariant and restriction that we utilize to achieve
+% this is that no object $o_1$ in one dynamic region $D_1$ can contain a
+% pointer to an object $o_2$ in another dynamic region $D_2$.  In cases
+% where this is too restrictive, the user has to resort to (deep)
+% copying $o_2$ to $D_1$.  This is conceptually quite reasonable given
+% the perspective that $D_1$ is a self-contained data-structure.
+% Furthermore, this is quite analogous, from a performance perspective,
+% to the copying that happens when a garbage collector promotes an
+% object from a lower generation to a higher generation.  From this
+% perspective, a region may be seen as a user-controlled generation.
+
+% This ensures that we can safely free one dynamic region (\eg, the
+% input-message in our example) without affecting the validity of
+% pointers in another dynamic region (\eg, the output-message).
+% However, this is not sufficient! \dv{for what} When processing the input-message
+% $D$, we necessarily will have to create pointers that point to objects
+% that are internal to $D$. Such pointers may be stack-allocated or even
+% reside in the heap (\eg, consider the iterator object used to iterate
+% over the list in the input-message).  How do we ensure the safety of
+% these pointers?
+
+% We use \emph{stack regions} (regions with lexically-scoped lifetimes),
+% as well as the stack, for such temporary pointers (to objects inside a
+% dynamic region) and use the following protocol to ensure safety of
+% such pointers.  To work with a dynamic region $D$, we must first
+% \emph{open} region $D$, to indicate that $D$ should not be freed
+% during this period.  When the processing of the region is complete, we
+% \emph{close} the region $D$, to indicate that it is safe to free $D$.
+% In general, a dynamic region may be opened and closed multiple times
+% before it is finally freed.
+
+% Our type-system permits the creation of pointers to internal objects
+% of \emph{open dynamic} regions, but uses lexical scoping to ensure
+% that such pointers are no longer live when the region is closed.
+% Thus, such pointers may be either stack-allocated or reside in a
+% region whose lifetime is statically guaranteed to be contained within
+% the interval when $D$ is open.  As a consequence, the system ensures
+% the strong invariant that for a \emph{closed dynamic} region $D$ no
+% pointer from outside $D$ points to an internal object of $D$.
+
+% The final piece required to ensure memory-safety is ensuring that the
+% program correctly follows the above-mentioned protocol for dynamic
+% regions: \eg, an open region should not be freed and, dually, a freed
+% region should not be opened.  (A complete typestate specification of
+% the protocol is presented later.)
+
+% Dynamic regions are first-class objects in our system. For instance,
+% in our running example, we would like to use a dictionary that stores
+% dynamic regions. However, this means that the program may create
+% multiple aliases (pointers) to the same region, which, in turn, means
+% that statically checking if the program correctly follows the dynamic
+% region protocol is hard (undecidable, in fact). Safety can be ensured
+% by using, \eg, linear typing to prevent the creation of aliases, but
+% this would be quite restrictive. Hence, in our system, we dynamically
+% check for adherence to this protocol.
+
+% In summary, our approach reduces memory safety to two sub-problems:
+% (a) Ensuring  low-level invariants about pointers to (fine-grained)
+% objects inside regions and (b) Ensuring higher-level (typestate)
+% invariants about the regions themselves.  We use a type system to
+% statically ensure the first set of invariants. We use a dynamic check
+% to ensure that the second set of invariants are satisfied at runtime.
+% %
+% This is a novel aspect of our system. We believe that this is a
+% reasonable choice: dynamically checking for the second category of
+% invariants is easy and does not pose a performance concern (unlike,
+% \eg, the first category of constraints); a static typechecker that
+% rules out the second category of violations would be very restrictive
+% from an expressiveness perspective; finally, the human effort required
+% to ensure the second set of invariants is not as burdensome (as
+% regions are coarse-grained objects manipulated infrequently), \eg, as
+% it would be for the first set of invariants.
+% >>>>>>> af17b22a358a6165ae10628a5247cd4fa66adc07
 
@@
 
-\begin{itemize} \item We present \name, a \csharp-like typed
+\begin{itemize} 
+  \item We present \name, a \csharp-like typed
   object-oriented language that eschews garbage collection in favour of
@@
   \item We describe an implementation of \name frontend in OCaml,
-  along with experimental results, and case studies where the region
-  type system was able to identify unsafe memory accesses statically.
+  along with case studies where the region type system was able to
+  identify unsafe memory accesses statically.
 

### 2015-11-20 paper/overview2.tex
Edits.
More small comments.

@@
 
-\paragraph{Transferable Regions} \name's \emph{transferable regions}
-are an encapsulation of a data-structure that can be transferred
-between autonomous entities (\eg, between two concurrently executing
-threads or actors).  Hence, unlike stack regions, transferable regions
-are not constrained to have a lexically scoped lifetime.  (Hence, we
-also refer to them as \emph{dynamic} regions.)
+\paragraph{Transferable Regions} 
+% \name's \emph{transferable regions} are an encapsulation of a
+% data-structure that can be transferred between autonomous entities
+% (\eg, between two concurrently executing threads or actors).  Hence,
+% unlike stack regions, transferable regions are not constrained to have
+% a lexically scoped lifetime.  (Hence, we also refer to them as
+% \emph{dynamic} regions.)
 
-Furthermore, transferable regions, unlike stack regions, are first
-class values of \name: they are objects of the class \C{Region}, they
-are created using the \C{new} keyword, and can be passed as arguments,
-stored in data structures, and returned from methods.  A transferable
-region is intended to encapsulate a single data-structure, consisting
-of a collection of objects with a distinguished root object of some
-type \C{T}, which we refer to as the region's \emph{root} object.  The
-class \C{Region} is parametric over the type \C{T} of this root
-object.
+Transferable regions are first class values of \name: they are objects
+of the class \C{Region}, they are created using the \C{new} keyword,
+and can be passed as arguments, stored in data structures, and
+returned from methods.  A transferable region is intended to
+encapsulate a single data-structure, consisting of a collection of
+objects with a distinguished root object of some type \C{T}, which we
+refer to as the region's \emph{root} object.  The class \C{Region} is
+parametric over the type \C{T} of this root object.
 
@@
 within it), and \C{transfer} method that transfers the region to a
-consumer process. It is an abstraction of two possible forms of
-transfer: a transfer between two processes in a shared memory setting
-or a transfer between two processes in a distributed, message-passing,
-setting. The precise semantics of \C{transfer} are unimportant in the
-context of the region type system and we will not discuss them
-further.
+(possibly remote) consumer process. 
+% It is an abstraction of two possible forms of
+% transfer: a transfer between two processes in a shared memory setting
+% or a transfer between two processes in a distributed, message-passing,
+% setting. 
+% The precise semantics of \C{transfer} are unimportant in the
+% context of the region type system and we will not discuss them
+% further. \dv{this is a bit counter-intuitive.}
 
@@
 refer to a transferable region that has not been opened as a
-\emph{closed} region.
+\emph{closed} region. A transferable region can only be transferred or
+freed when it is in closed state. The acceptable state transition
+discipline over the lifetime of a transferable region is described
+in Fig.~\ref{fig:region-fsm}. Enforcement of this discipline is done
+at runtime.
 
@@
 store the output for time \C{t}, initializing it to contain an empty
-list.  Line 9 opens the input region to process it.  Line 10 creates a
-stack region \C{R0}.  Thus, the temporary objects created by the
-iteration in line 11, for example, will be allocated in this stack
+list.  Line 9 opens the input region to process it\footnote{We omit
+\C{@R} annotation in \C{open} when we don't need \C{R}.}.  Line 10
+creates a stack region \C{R0}.  Thus, the temporary objects created by
+the iteration in line 11, for example, will be allocated in this stack
 region that lives just long enough.  We open the desired output region
@@
 method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
-After the transfer, \C{SelectVertex} is no longer allowed to access
-the \C{outRgn}, facilitating its deallocation.
+% After the transfer, \C{SelectVertex} is no longer allowed to access
+% the \C{outRgn}, facilitating its deallocation.
 
@@
 
-\subsection{Memory Safety} \label{sec:memory-safety}
+% \subsection{Memory Safety} \label{sec:memory-safety}
 
-Our goal is a type system that can ensure the memory safety of
-programs that use the region constructs described above.  The key to
-memory safety in \name is the following restriction: an object $o_1$
-in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
-a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
-similar restriction applies in the case where $o_1$ is a
-stack-allocated variable.)
+% Our goal is a type system that can ensure the memory safety of
+% programs that use the region constructs described above.  The key to
+% memory safety in \name is the following restriction: an object $o_1$
+% in a region $R_1$ is allowed to store a pointer to an object $o_2$ in
+% a region $R_2$ only if $R_2$ is guaranteed to outlive $R_1$.  (A
+% similar restriction applies in the case where $o_1$ is a
+% stack-allocated variable.)
 
-Enforcing this restriction is simple in the case of stack regions
-since the outlives relation between stack regions can be inferred from
-their lexical nesting. Unfortunately, inferring outlives relations
-in presence of transferable regions is not easy.  \name imposes the following
-protocol on the use of transferable regions to help simplify this
-check.
+% Enforcing this restriction is simple in the case of stack regions
+% since the outlives relation between stack regions can be inferred from
+% their lexical nesting. Unfortunately, inferring outlives relations
+% in presence of transferable regions is not easy.  \name imposes the following
+% protocol on the use of transferable regions to help simplify this
+% check.
 
-A transferable region (that has not been freed or transferred) can be
-in one of two possible states, \emph{open} or \emph{closed}. A newly
-created region is in the closed state.  A region must be opened, using
-the open construct (as explained previously), in order to read or
-update or allocate an object within that region.  An open region
-cannot be freed or transferred.  In particular, an open region is
-guaranteed to be live for the entire duration of the open construct.
-This allows the type system to infer a valid outlives relation between
-the opened region and any stack region that is nested within the open
-construct.
+% A transferable region (that has not been freed or transferred) can be
+% in one of two possible states, \emph{open} or \emph{closed}. A newly
+% created region is in the closed state.  A region must be opened, using
+% the open construct (as explained previously), in order to read or
+% update or allocate an object within that region.  An open region
+% cannot be freed or transferred.  In particular, an open region is
+% guaranteed to be live for the entire duration of the open construct.
+% This allows the type system to infer a valid outlives relation between
+% the opened region and any stack region that is nested within the open
+% construct.
 
@@
 \label{fig:region-fsm} 
-\vspace*{-0.25in} 
+\vspace*{-0.1in} 
 \end{figure}
 
-The protocol for transferable regions is presented as a finite state
-machine in Fig.~\ref{fig:region-fsm}.  
-% A transferable region starts
-% its lifetime in a \emph{closed} state, when it is created.
-%
-The safety of memory accesses in \name is now subject to the condition
-that every transferable region correctly follows the state transition
-discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
-region type system statically guarantees the safety of all memory
-accesses.
-% In other words, the type system reduces the problem of ensuring
-% memory safety in \name programs to the problem of enforcing the
-% state transition discipline for transferable regions.
+% The protocol for transferable regions is presented as a finite state
+% machine in Fig.~\ref{fig:region-fsm}.  
+% % A transferable region starts
+% % its lifetime in a \emph{closed} state, when it is created.
+% %
+% The safety of memory accesses in \name is now subject to the condition
+% that every transferable region correctly follows the state transition
+% discipline of Fig.~\ref{fig:region-fsm}. Under this condition, \name's
+% region type system statically guarantees the safety of all memory
+% accesses.
+% % In other words, the type system reduces the problem of ensuring
+% % memory safety in \name programs to the problem of enforcing the
+% % state transition discipline for transferable regions.
 
-In \name, this enforcement is done at runtime by explicitly keeping
-track of the \emph{current state} for $\RgnZ$ objects, and checking
-the validity of every open, transfer, or free operation and throwing
-an exception if it is invalid.  
-% The challenge in enforcing this
-% discipline statically is that transferable regions are first-class
-% objects. Hence, the program can create multiple aliases for the same
-% region, \eg, open it via one alias and free it via another.  Typestate
-% verification in the presence of aliases is hard.  The checking can be
-% done statically by preventing the creation of aliases using, \eg,
-% linear types or unique types. However, this would be quite
-% restrictive, in terms of expressiveness.
-As explained previously, this is a reasonable trade-off in the context
-of \name, as regions are coarse-grained objects, which are manipulated
-infrequently, when compared to fine-grained objects that reside inside
-these regions. Therefore, runtime overhead of checking the region's
-state transition discipline is acceptable.
+% In \name, this enforcement is done at runtime by explicitly keeping
+% track of the \emph{current state} for $\RgnZ$ objects, and checking
+% the validity of every open, transfer, or free operation and throwing
+% an exception if it is invalid.  
+% % The challenge in enforcing this
+% % discipline statically is that transferable regions are first-class
+% % objects. Hence, the program can create multiple aliases for the same
+% % region, \eg, open it via one alias and free it via another.  Typestate
+% % verification in the presence of aliases is hard.  The checking can be
+% % done statically by preventing the creation of aliases using, \eg,
+% % linear types or unique types. However, this would be quite
+% % restrictive, in terms of expressiveness.
+% As explained previously, this is a reasonable trade-off in the context
+% of \name, as regions are coarse-grained objects, which are manipulated
+% infrequently, when compared to fine-grained objects that reside inside
+% these regions. Therefore, runtime overhead of checking the region's
+% state transition discipline is acceptable.
 

### 2015-11-20 paper/type-inference.tex
Edits.

--- 2015-11-20 paper/type-inference.tex
@@
   q.fst := y; q.snd := y; 
-  open r as $\rgn_0$ withroot p in
+  open r as $p@\rgn_0$ in
     letregion $\rgn_1$ in

### 2015-11-20 paper/implementation.tex
Edits to the implementation section.

@@
 including its region type system and type inference, in 3k+ lines of
-OCaml. The input to our system is a program in $\absof{\FB^+}$, an extended
-version of $\absof{\FB}$ that includes assignments, conditionals,
-loops, more primitive datatypes (\eg strings), and a null value. 
+OCaml. Our implementation is called \namec. The input to \namec is a
+program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
+includes assignments, conditionals, loops, more primitive datatypes
+(\eg, integers), and a null value. 
 % If needed, a specification file containing region type annotations
 % for higher-order functions can also be provided. 
-Our implementation of region type inference closely follows the
-description given in Sec.~\ref{sec:type-inference}. To solve the
-constraints that arise during type inference, we built a solver called
-\csolve that implements constraint solving approach based on fixpoint
-computation and graph augmentation described in \S~\ref{sec:csolve}. 
+Our implementation of region type inference and constraint solving
+closely follows the description given in
+Sec.~\ref{sec:type-inference}. The one difference is that our
+constraint solver uses an approximation algorithm to compute the
+minimal graph augmentation (\S~\ref{sec:csolve}), since the general
+problem is NP-hard~\cite{siam92}.
+% To solve the
+% constraints that arise during type inference, we built a solver called
+% \csolve that implements constraint solving approach based on fixpoint
+% computation and graph augmentation described in \S~\ref{sec:csolve}. 
 % Since graph augmentation is NP-hard, \csolve uses an approximation
@@
 % constraints are not often complex. 
-If the input $\absof{\FB^+}$ program does not create any unsafe
-references, our compiler annotates it with region types, which act as
-a witness to program's memory safety. On contrary, if the input
-program does create a reference that is potentially unsafe, then the
-type inference fails during the constraint solving phase. We currently
-do not implement error localization and feedback mechanisms.
+% If the input $\absof{\FB^+}$ program does not create any unsafe
+% references, \namec annotates it with region types, which act as a
+% witness to program's memory safety. If potential safety violations are
+% encountered, then the type inference fails during the constraint
+% solving phase.
+
+\begin{figure}
+\begin{codejava}
+class LinkedList<T><R5,R4 | R4$\outlives$R5> {
+  ListNode<T><R5,R4> head;
+  ..
+  List<T><R17,R4> rev<R17,R4 | R4$\outlives$R17>(unit u) {
+    List<T><R17,R4> xs = 
+        new List<T><R17,R4>(this.head.val);
+    ListNode<T><R5,R4> cur = this.head.next;
+    while (!cur == Null) {
+      xs.add<R17>(cur.val)
+      cur = cur.next;
+    }
+    return xs;
+  }
+\end{codejava}
+
+\caption{Region-annotated definition of \C{rev} computed by \namec}
+\label{fig:rev}
+\end{figure}
 
@@
 % implemented some of the C\# libraries in $\absof{\FB^+}$, and 
-translated some of the Naiad streaming query operator benchmarks
-(Naiad vertices) used in~\cite{Broom:HotOS} to $\absof{\FB^+}$, and
-used our prototype compiler to assign region types to these programs,
-and thus prove their safety. During the process, we found multiple
-instances of potential memory safety violations in the $\absof{FB^+}$
-translation of benchmarks, which we verified to be present in the
-original C\# implementation as well. The cause of all safety
-violations is the creation of a reference from the outgoing message (a
-transferable region) to the payload of the incoming message. For
-example, the implementation of \C{RegionSelectVertex} contains the
-following:
+performed two kinds of experiments. First, we implemented some of the
+microbenchmarks ($\le$100 LOC), which are the standard libraries such
+as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
+our inference engine to infer their principal region types. Since a
+library class is region-oblivious, if it is well-typed as per the core
+type system, then \namec must be able to automatically construct its
+region-type-annotated definition without fail. As expected, \namec was
+able to infer principal region types for all the library classes,
+under 10ms. Fig~\ref{fig:rev} shows the region-type-annotated
+definition computed for the list reverse method. Observe that \namec
+was able to infer that the list and its data (of type \C{T}) can be
+allocated in different regions, as long as the later outlives the
+former. This allows, for instance, a \C{preOrder} method to traverse a
+tree in a transferable region, and return a list of its nodes, where
+the list itself is allocated in the stack region.
+
+Next, we translated some of the Naiad streaming query operator
+benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
+$\absof{\FB^+}$, and used \namec to verify their safety. During the
+process, we found multiple instances of potential memory safety
+violations in the $\absof{FB^+}$ translation of benchmarks, which we
+verified to be present in the original C\# implementation as well. The
+cause of all safety violations is the creation of a reference from the
+outgoing message (a transferable region) to the payload of the
+incoming message. For example, the implementation of \C{SelectVertex}
+contains the following:
 \begin{codejava}
-  if (this.selectFn(inMsg.payload[i])) {
+  if (this.selector(inMsg.payload[i])) {
     outMsg.set(outputOffset, inMsg.payload[i]);
@@
 
+Our experience with Naiad benchmarks suggests that, although our
+approach cannot statically enforce memory safety, it is nonetheless
+useful, particularly because the static verification comes at no
+additional cost to the developer. If the developer is satisfied with
+the confidence gained by static verification, she may even choose to
+turn the runtime safety checks off. Since we did not implement code
+generation in \namec, we could not measure the runtime overhead of
+checks needed to enforce safety. However, the number of LOC peforming
+operations on \C{Region} objects relative to the total LOC never
+exceeds 8\% in Naiad benchmarks. 
+% \C{SelectVertex}, for example, contains one each of
+% \C{open} and \C{transfer} operations on one region in its 45LOC,
+% whereas, \C{JoinVertex} contains 5 \C{open}s, 2 \C{transfer}s, and 1
+% \C{free} operation.
+
 % We now briefly describe our experience of

### 2015-11-21 paper/abstract.tex
Rewrote introduction.
Final changes.

@@
-Recent work in high-performance 
-systems written in managed languages 
+Recent work in high-performance systems written in managed languages
 (such as Java or C\#) has shown that garbage-collection can be a
-significant performance bottleneck.  A class of these
-systems, focused on big-data, create many and often large data-structures with
-well-defined lifetimes.  In this paper, we present a language and
+significant performance bottleneck.  A class of these systems, focused
+on big-data, create many and often large data structures with
+well-defined lifetimes.  In this paper, we present a language and a
 memory management scheme based on user-managed memory regions (called
 \emph{transferable regions}) that allow programmers to exploit
-knowledge of data-structures' lifetimes to achieve significant
+knowledge of data structures' lifetimes to achieve significant
 performance improvements.
@@
 Manual memory management is susceptible to the usual perils of
-dangling pointers.  A key contribution of this paper is a region type
-system that ensures the memory safety of C\# programs in the presence
-of transferable regions. We compliment our type system with a type
-inference algorithm that infers principal region types for first-order
-programs, and useful types for higher-order programs. This eliminates
-the need for programmers to write region annotations on types, while
-facilitating the reuse of existing C\# libraries with no
-modifications. Experiments demonstrate the practical utility of our
-approach.
+dangling pointers. A key contribution of this paper is a
+refinement-based region type system that ensures the memory safety of
+C\# programs in the presence of transferable regions. We complement
+our type system with a type inference algorithm that infers principal
+region types for first-order programs, and practically useful types
+for higher-order programs. This eliminates the need for programmers to
+write region annotations on types, while facilitating the reuse of
+existing C\# libraries with no modifications. Experiments demonstrate
+the practical utility of our approach.
 

### 2015-11-21 paper/implementation.tex
Final changes.

@@
 constraint solver uses an approximation algorithm to compute the
-minimal graph augmentation (\S~\ref{sec:csolve}), since the general
-problem is NP-hard~\cite{siam92}.
+minimal graph augmentation (\S~\ref{sec:csolve}).
 % To solve the
@@
 \label{fig:rev}
+\vspace*{-0.15in}
 \end{figure}

### 2015-11-21 paper/introduction.tex
Final changes.

@@
 begin with. Note that, while the references of later kind (i.e.,
-references that escape a region), defeat the very purpose of a
+references that escape a region) defeat the very purpose of a
 transferable region, hence need to be prohibited, references of the
@@
 performance, as any non-trivial program creates temporary objects, and
-it is absurd to allocate them in a transferable region; such regions
-are meant for the data being transferred. Since transferable regions
-are first class objects in our setting, controlling references to and
-from such regions while ensuring their safety without significantly
-diluting the performance advantage of regions over GC is a challenging
-exercise.
+it is undesirable to allocate them in a transferable region; such
+regions are meant for the data being transferred. Since transferable
+regions are first class objects in our setting, controlling references
+to and from such regions while ensuring their safety without
+significantly diluting the performance advantage of regions over GC is
+a challenging exercise.
 
@@
 presence of transferable regions through a combination of a static
-typing discipline and lightweight runtime checks. Our approach is
-based on the following observations. Firstly, we observe that by
-nesting a TT-style \C{letregion} lexical block, that delimits the
-lifetime of a stack region, inside an \C{open} lexical block for
-transferable regions, which ``opens'' a transferable region and
-guarantees that the region won't be transferred/freed inside the
-block, we can guarantee that the transferable region will remain live
-as long as the stack region is live. We say that the former
-``outlives'' the later\footnote{We borrow the \emph{outlives} relation
-from~\cite{MIT03}. A comparison of our approach with~\cite{MIT03} can
-be found in \S~\ref{sec:related-work}.}, and any references from the stack
-region to the transferable region are therefore safe. 
+typing discipline and lightweight runtime checks. The cornerstone of
+our approach is an \C{open} lexical block for transferable regions,
+that ``opens'' a transferable region and guarantees that the region
+won't be transferred/freed while it is open. Our observation is that
+by nesting a~\cite{tofte94}-style \C{letregion} lexical block, that
+delimits the lifetime of a stack region, inside an \C{open} lexical
+block for a transferable region, we can guarantee that the
+transferable region will remain live as long as the stack region is
+live. We say that the former ``outlives'' the later\footnote{We borrow
+the \emph{outlives} relation from~\cite{MIT03}. A comparison of our
+approach with~\cite{MIT03} can be found in
+\S~\ref{sec:related-work}.}, and any references from the stack region
+to the transferable region are therefore safe. 
 % In general, we can allow references from a region
@@
 % $R_0$. 
-Second, we note that by controlling the ``outlives'' relationships
+Next, we note that by controlling the ``outlives'' relationships
 between various regions, we only allow safe cross-region references,
@@
 establish any outlives relationships, thus declaring our intention to
-not allow any cross-region references between $R_0$ and $R_1$.  Third,
-outlives relationships are established based on the lexical structure
-of the program, hence a static type system can enforce them
-effectively. By assigning region types to objects, which capture the
-regions such objects are allocated in, and by maintaining outlives
-relationships between various regions, we can statically decide the
-safety of a reference.
+not allow any cross-region references between $R_0$ and $R_1$.
+Finally, we observe that outlives relationships are established based
+on the lexical structure of the program, hence a static type system
+can enforce them effectively. By assigning region types to objects,
+which capture the regions such objects are allocated in, and by
+maintaining outlives relationships between various regions, we can
+statically decide the safety of all references in the program.
 
-Of course, the utility of our approach described above is predicated on
-the assumption that we can enforce certain invariants on transferable
-regions. Firstly, a transferable region cannot be transferred/freed
-inside an open block of that region (i.e, while it is still open).
-Secondly, a transferred/freed region cannot be opened. These are
-typestate invariants on the transferable region objects, which are
-hard to enforce statically due to the presence of unrestricted
-aliasing. Techniques like linear types and unique pointers can be used
-to restrict aliasing, but the constraints they impose are often hard
-to program around, and are not often acceptable to systems
-programmers. We therefore enforce typestate invariants at runtime via
-lightweight runt-time checks. In particular, we define an acceptable
-state transition discipline for transferable regions
-(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
-transition of a transferable region (\eg, from \emph{open} state to
-\emph{freed} state) is valid or not. The check is lightweight since it
-only involves checking a single tag that captures the current state.
-We believe that this is a reasonable choice since regions are
-coarse-grained objects manipulated infrequently, when compared to the
-fine-grained objects that are present inside these regions, for which
-safety is enforced statically. An added advantage of delegating the
-enforcement of typestate invariants to runtime is that our region type
-system is simple\footnote{Supplement lists all the type rules of our
-language, which occupy less than a page.}, which made it possible to
-formulate a type inference that completely eliminates the need to
-write region type annotations. This, we believe, significantly reduces
-the impediment to adopt our approach in practical setting.
+Indeed, the utility of our approach described above is predicated
+on the assumption that we can enforce certain invariants on
+transferable regions. Firstly, a transferable region cannot be
+transferred/freed inside an open block of that region (i.e, while it
+is still open).  Secondly, a transferred/freed region cannot be
+opened. These are typestate invariants on the transferable region
+objects, which are hard to enforce statically due to the presence of
+unrestricted aliasing. Techniques like linear types and unique
+pointers can be used to restrict aliasing, but the constraints they
+impose are often hard to program around. We therefore enforce
+typestate invariants at runtime via lightweight checks. In
+particular, we define an acceptable state transition discipline for
+transferable regions (Fig.~\ref{fig:region-fsm}), and check, at
+runtime, whether a given transition of a transferable region (\eg,
+from \emph{open} state to \emph{freed} state) is valid or not. The
+check is lightweight since it only involves checking a single tag that
+captures the current state.  We believe that this is a reasonable
+choice since regions are coarse-grained objects manipulated
+infrequently, when compared to the fine-grained objects that are
+present inside these regions, for which safety is enforced statically.
+An added advantage of delegating the enforcement of typestate
+invariants to runtime is that our region type system is
+simple\footnote{Supplement lists all the type rules of our language,
+which occupy less than a page.}, which made it possible to formulate a
+type inference that completely eliminates the need to write region
+type annotations. This, we believe, significantly reduces the
+impediment to adopt our approach in practical setting.
 

### 2015-11-21 paper/related.tex
Final changes.

@@
 
-~\cite{tofte97} proposes to extend Standard ML, a higher-order
-functional language, with lexically-scoped stack regions, and defines
-an elaboration from Standard ML to the region-annotated version of
-Standard ML. The aim of the elaboration is to introduce stack regions
-and do away with GC in a transparent fashion without jeopardizing
-memory safety. We too define an elaboration, but our focus is on
-introducing region annotations necessary to prove the safety of an
-object-oriented program that already uses (stack and dynamic) regions.
+~\cite{tofte97} elaborates SML programs to introduce stack regions and
+do away with GC in a transparent fashion. We too define an
+elaboration, but our focus is on introducing region annotations
+necessary to prove the safety of an object-oriented program that
+already uses regions.
 % Similar to their region inference algorithm, our region type inference
@@
 aliased. We don't have linear typing, thus admit unrestricted
-aliasing. Non-linear typing keeps our type system simple, and makes it
-possible to infer types, thus eliminating the annotation burden.
+aliasing. 
+% Non-linear typing keeps our type system simple, and makes it
+% possible to infer types, thus eliminating the annotation burden.
 Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce

### 2015-11-21 paper/type-inference.tex
Final changes.

@@
+\vspace*{-0.1in}
 \section{Type Inference}
@@
 
+\vspace*{-0.1in}
 \subsection{Region Type Templates}
@@
 with fresh region variables in method calls and function applications.
-This elaboration is performed with respect to the polymorphic type
-templates of classes and methods computed as per
-\S\ref{sec:fb-templatization}. Our expression elaboration
-function\footnote{Supplement contains the pseudocode for all the
-elaboration functions.} traverses expressions in a syntax-directed
-manner of a type checker, introducing fresh region type templates for
-unknown region types, while generating constraints over region and
-predicate variables. The precise nature of generated constraints is
-explained in \S\ref{sec:constraints}, but in summary, they
-capture the relationships between the type templates of various
-subexpressions and their well-formedness.
+% This elaboration is performed with respect to the polymorphic type
+% templates of classes and methods computed as per
+% \S\ref{sec:fb-templatization}. 
+Our expression elaboration function\footnote{Supplement contains the
+pseudocode for all the elaboration functions.} traverses expressions
+in a syntax-directed manner of a type checker, introducing fresh
+region type templates for unknown region types, while generating
+constraints over region and predicate variables. The precise nature of
+generated constraints is explained in \S\ref{sec:constraints}, but in
+summary, they capture the relationships between the type templates of
+various subexpressions and their well-formedness.
 
@@
 their constituents. For instance, method elaboration generates a new
-constraint that captures the requirement that, under the guarantees
-provided by the context, the region type of a method body must be a
-subtype\footnote{Subtyping is type equality in \FB. In \name,
-subtyping between region types is a straightforward extension of
-subtyping between core types.} of method's return type.  Likewise, the
-elaboration function for a class generates a fresh constraint that
-captures a couple of safety conditions: first, the allocation regions
-of objects referred by the instance variables should outlive the
-allocation region of the instance itself, and second, the allocation
-regions of a class type and its superclass type must be the same.
+constraint by typechecking the method body against its return type.
+% subtype\footnote{Subtyping is type equality in \FB. In \name,
+% subtyping between region types is a straightforward extension of
+% subtyping between core types.} of method's return type.  
+Likewise, the elaboration function for a class generates a fresh
+constraint that captures a couple of safety conditions: first, the
+allocation regions of objects referred by the instance variables
+should outlive the allocation region of the instance itself, and
+second, the allocation regions of a class type and its superclass type
+must be the same.
 
@@
 from~\cite{ltpldi08}}:
+\vspace*{-0.05in}
 \begin{smathpar}
@@
 \end{smathpar}
+\vspace*{-0.1in}
 \end{itemize}
@@
 the running example:
+\vspace*{-0.05in}
 \begin{center}
@@
 \end{center}
+\vspace*{-0.1in}
 Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
 $c_{14}$ gives us a recursive constraint of the following form:
+\vspace*{-0.05in}
 \begin{center}
@@
 \end{center}
+\vspace*{-0.05in}
 Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and $F$ is

### 2015-11-21 paper/type-system.tex
Final changes.

--- 2015-11-20 paper/type-system.tex
@@
 region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
-$R_2$ is guaranteed to outlive $R_1$. Intuitively, the invariant can
-be enforced by (a). tracking \emph{outlives} relationships between
-various regions in the program (b). tagging the C\# type of every
-object in the program with its allocation region, and (c).  ensuring
-that, when a reference is created, its target object ($o_2$) is
-allocated in a region that is known to be in outlives relationship
-with the object containing the reference. We now formally develop this
-intuition via \fbname (\FB), our explicitly typed core language (with region
-types) that incorporates the features introduced in the previous
-section. \fbname builds on the Featherweight Generic Java
-(FGJ)~\cite{fgj} formalism, and reuses notations and various
-definitions from~\cite{fgj}, such as the definition of type
-well-formedness for the core (region-free) language.
+$R_2$ is guaranteed to outlive $R_1$. 
+% Intuitively, the invariant can
+% be enforced by (a). tracking \emph{outlives} relationships between
+% various regions in the program (b). tagging the C\# type of every
+% object in the program with its allocation region, and (c).  ensuring
+% that, when a reference is created, its target object ($o_2$) is
+% allocated in a region that is known to be in outlives relationship
+% with the object containing the reference. 
+We now formally develop this intuition via \fbname (\FB), our
+explicitly typed core language (with region types) that incorporates
+the features introduced in the previous section. \fbname builds on the
+Featherweight Generic Java (FGJ)~\cite{fgj} formalism, and reuses
+notations and various definitions from~\cite{fgj}, such as the
+definition of type well-formedness for the core (region-free)
+language.
 
@@
 $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
-$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
-$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
-\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
+$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in
+\rhoenv'$, $\rgn_i \neq \toprgn$ and
+$\rgn_i \notin dom(\rhomap') \cup \rhoenv'$, if $\hastyp{\emptyA,\rgn,
+\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
 $\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
-{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
+{\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
 \end{theorem}
@@
 needs to add a binding for the region to its $\rhomap$ in order to
-preserve its type safety. Notably, this result could not have been
-established if it were possible for a transferable region to contain
-references to objects outside the region.
+preserve its type safety. 
+% Notably, this result could not have been
+% established if it were possible for a transferable region to contain
+% references to objects outside the region.

### 2016-01-31 paper/related.tex
Related work with Cyclone added.

@@
 regions, and a typing discipline that statically guarantees the safety
-of all pointer dereferences. However, types need to be written
-explicitly. While some region annotations can be inferred, type
-inference is purely intraprocedural, and cannot infer region types for
-functions. Later proposal~\cite{cyclone04} extends Cyclone with
-dynamic regions, imposes unique pointer restrictions to guarantee
-safety. Objects that need to be shared within data structures are
-allowed to have multiple references, and safety for such objects was
-ensured via runtime reference counting.
+of all pointer dereferences. Later
+proposals~\cite{cyclone04,cycloneSCP} extends Cyclone with dynamic
+regions. \name differs from Cyclone fundamentally because of its
+non-intrusiveness design principle, which requires its safety
+mechanisms to not intrude on the the programming practices of C\#.
+\name programmers, for example, shouldn’t be forced to abandon
+iterators in favor of for-loops, annotate region types, or rewrite
+C\#'s standard libraries to use in \name.  Non-intrusiveness is not a
+design consideration for Cyclone, which requires C programmers to use
+new language constructs and abandon some standard programming idioms
+in the interest of preserving safety. For instance, Cyclone
+programmers are required to write region types for functions; the type
+inference is only intraprocedural. Ensuring safety in presence of
+dynamic regions requires using either unique pointers or
+reference-counted objects.  Both approaches are intrusive. For
+example, unique pointers constrain, or in some cases forbid, the use
+of the familiar iterator pattern, which requires creation of aliases
+to objects in a collection. Some standard library functions, for
+example, those that use caching, may need to be rewritten.  Moreover,
+even with unique pointers, safety cannot be guaranteed statically;
+checks against \C{NULL} are needed at run-time to enforce safety. For
+ref-counted objects, Cyclone requires programmers to use special
+functions (\C{alias\_refptr} and \C{drop\_refptr}) to create and
+destroy aliases.  Reference count is affected only by these
+functions. An alias going out of scope, for instance, does not
+decrement the ref-count. The requirement to use additional constructs
+to manage aliases makes reference counting more-or-less as intrusive
+as unique pointers.
+
+Our work differs from Cyclone also in terms of its technical
+contributions. While Cyclone equips C with a range of region
+constructs~\cite{cycloneSCP}, the semantics of (a significant subset
+of) such constructs, and the safety guarantees of the language are not
+formalized. In contrast, the (static and dynamic) semantics of Broom
+has been rigorously defined with respect to a well-understood formal
+system (FGJ). The safety guarantees have been formalized and proved.
+The core of Broom is very simple; the rules that make up static and
+dynamic semantics occupy less than a page each. We believe that the
+rigor and simplicity of Broom makes it easy to understand the the
+underlying ideas, and apply them in various problem settings. Similar
+contrast can be made of region type inference in both the languages.
+Cyclone’s type inference was only ever described as being similar to
+Tofte and Talpin’s, and it’s effectiveness in presence of tracked
+pointers is not clear. In contrast, the complete Ocaml (pseudo) code
+of Broom's inference algorithm, was given in the supplement and the
+ideas underlying type inference have been described elaborately in the
+paper
 

### 2016-05-21 paper/broom.tex
Merged paper to techrep.
Hyperref.
small things in intro.
Edits to the implementation section.
Typos fixed. Some changes from master merged.

@@
 %\documentclass[10pt,pldi]{sigplanconf}
-\documentclass[pldi]{sigplanconf-pldi16}
+\documentclass[pldi-cameraready]{sigplanconf-pldi16}
 
@@
 \newcommand{\shape}{{\sf shape}}
+\newcommand{\ctype}{{\sf ctype}}
+\newcommand{\override}{{\sf override}}
 %\newcommand{\smallskip}{\vspace*{0.1in}}
@@
 \newcommand{\qqquad}{\quad\quad\quad}
-\newcommand{\emptyA}[0]{(\rhomap,\rhoenv,\cdot,\phicx)}
-\newcommand{\emptyASigp}[0]{(\rhomap',\rhoenv,\cdot,\phicx)}
+\newcommand{\emptyA}[0]{(dom(\rhomap),\rhoenv,\cdot,\phicx)}
+\newcommand{\emptyASigp}[0]{(dom(\rhomap'),\rhoenv,\cdot,\phicx)}
 \newcommand{\emptyASigpp}[0]{(dom(\rhomap''),\rhoenv,\cdot,\phicx)}
-\newcommand{\emptyADelcup}[1]{(\rhomap,\rhoenv \cup \{#1\},\cdot,\phicx)}
-\newcommand{\emptyASigpDelcup}[1]{(\rhomap',\rhoenv \cup \{#1\},
+\newcommand{\emptyADelcup}[1]{(dom(\rhomap),\rhoenv \cup \{#1\},\cdot,\phicx)}
+\newcommand{\emptyASigpDelcup}[1]{(dom(\rhomap'),\rhoenv \cup \{#1\},
                                     \cdot,\phicx)}
@@
                                     \cdot,\phicx)}
-\newcommand{\emptyADelcupPhicap}[2]{(\rhomap,\rhoenv \cup \{#1\},
+\newcommand{\emptyADelcupPhicap}[2]{(dom(\rhomap),\rhoenv \cup \{#1\},
                                     \cdot,\phicx \conj \rhoenv \outlives #2)}
-\newcommand{\emptyADelcupPhicapp}[2]{(\rhomap,\rhoenv \cup \{#1\},
+\newcommand{\emptyADelcupPhicapp}[2]{(dom(\rhomap),\rhoenv \cup \{#1\},
                                     \cdot,\phicx \conj #2)}
@@
 %---------------------
-\title{Safe Transferable Regions}
+\title{Safe Memory Regions for Big Data Processing}
 
-%\authorinfo{} {} {}
+\authorinfo{Gowtham Kaki}
+           {Purdue University, USA}
+           {gkaki@purdue.edu}
+
+\authorinfo{G Ramalingam}
+           {Microsoft Research, India}
+           {grama@microsoft.com}
+
+\authorinfo{Kapil Vaswani}
+           {Microsoft Research, India}
+           {kapilv@microsoft.com}
+
+\authorinfo{Dimitrios Vytiniotis}
+           {Microsoft Research, UK}
+           {dimitris@microsoft.com}
+
 \maketitle
@@
 
-\input{overview2}
+\input{overview}
 
@@
 \input{related}
+
 % \appendix
@@
 
+\appendix
+\input{fb-proofs}
+
 \end{document}

### 2016-05-21 paper/introduction.tex
Typos fixed. Some changes from master merged.

@@
 begin with. Note that, while the references of later kind (i.e.,
-references that escape a region) defeat the very purpose of a
+references that escape a region), defeat the very purpose of a
 transferable region, hence need to be prohibited, references of the
@@
 
-Indeed, the utility of our approach described above is predicated
+Of course, the utility of our approach described above is predicated
 on the assumption that we can enforce certain invariants on
@@
 impose are often hard to program around. We therefore enforce
-typestate invariants at runtime via lightweight checks. In
+typestate invariants at runtime via lightweight runt-time checks. In
 particular, we define an acceptable state transition discipline for
@@
 invariants to runtime is that our region type system is
-simple\footnote{Supplement lists all the type rules of our language,
-which occupy less than a page.}, which made it possible to formulate a
-type inference that completely eliminates the need to write region
-type annotations. This, we believe, significantly reduces the
-impediment to adopt our approach in practical setting.
+simple\footnote{\S~\ref{sec:type-system} contains all the type rules
+of our language, which occupy less than a page.}, which made it
+possible to formulate a type inference that completely eliminates the
+need to write region type annotations. This, we believe, significantly
+reduces the impediment to adopt our approach in practical setting.
 

### 2016-05-21 paper/related.tex
Typos fixed. Some changes from master merged.

@@
 Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
-concept of a region type system to statically enforce the safety of
-their region-based memory management scheme in ML. Following their
-seminal work, static type systems for safe region-based memory
-management have been extensively studied in the context of various
-languages and problem settings~\cite{cyclone02, cyclone04, yates99,
-MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
-existing proposals in a number of ways. Firstly, our problem setting
-includes lexically scoped stack regions and dynamic transferable
-regions (both programmer-managed) in context of an object-oriented
-programming language equipped with higher-order functions. Second, we
-adopt a two-pronged approach to memory safety that relies on a
-combination of a simple static type discipline and lightweight runtime
-checks. 
+concept of a region type system to statically ensure the safety of
+region-based memory management in ML. Following their seminal work,
+static type systems for safe region-based memory management have been
+extensively studied in the context of various languages and problem
+settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
+WW01, rust, gpu14}. Our work differs from the existing proposals in a
+number of ways. Firstly, our problem setting includes lexically scoped
+stack regions and dynamic transferable regions (both
+programmer-managed) in context of an object-oriented programming
+language equipped with higher-order functions. Second, we adopt a
+two-pronged approach to memory safety that relies on a combination of
+a simple static type discipline and lightweight runtime checks. 
 % The type system reduces the problem of ensuring safety of all memory
@@
 in order to guarantee safety. Lastly, our region type system comes
-equipped with full type inference that eliminates the need to write
-region annotations on types to convince the type checker that the
-program is safe.
+equipped with full type inference that completely eliminates the need
+to write region annotations on types to convince the type checker that
+the program is safe.
 
-~\cite{tofte97} elaborates SML programs to introduce stack regions and
-do away with GC in a transparent fashion. We too define an
-elaboration, but our focus is on introducing region annotations
-necessary to prove the safety of an object-oriented program that
-already uses regions.
-% Similar to their region inference algorithm, our region type inference
-% algorithm can deal with region-polymorphic recursion by computing
-% fixed points for recursive constraints. 
-Our type inference is analogous to their region inference. While their
-inference algorithm only ever generates equality constraints, which
-can be solved via unification, our type inference algorithm also
-generates partial order outlives constraints, which are required to
-capture subtle relationships between lifetimes of transferable regions
-and stack regions. Consequently, our constraint solving algorithm is
-more sophisticated, and is capable of inferring unknown outlives
+~\cite{tofte97} proposes to extend Standard ML, a higher-order
+functional language, with lexically-scoped stack regions, and defines
+an elaboration from Standard ML to the region-annotated version of
+Standard ML. The aim of the elaboration is to introduce stack regions
+and do away with GC in a transparent fashion without jeopardizing
+memory safety. We too define an elaboration, but our focus is on
+introducing region annotations necessary to prove the safety of an
+object-oriented program that already uses (stack and dynamic) regions.
+Similar to their region inference algorithm, our region type inference
+algorithm can deal with region-polymorphic recursion by computing
+fixed points for recursive constraints. While their inference
+algorithm only ever generates equality constraints, which can be
+solved via unification, our type inference algorithm also generates
+partial order outlives constraints, which are required to capture
+subtle relationships between lifetimes of transferable regions and
+stack regions. Consequently, our constraint solving algorithm is more
+sophisticated, and is capable of inferring unknown outlives
 constraints over region arguments of polymorphic recursive functions.
@@
 An ownership type system for safe region-based memory management in
-real-time Java has been proposed by~\cite{MIT03}.  
-% Like us, they too assume a source language with programmer-managed
-% memory regions, and focus on proving safety of programs written in
-% that language.  
-Their source language admits various kinds of regions with
-lexically-scoped lifetimes (i.e., stack regions) in order to support
-shared-memory concurrency. In contrast, we admit regions with
+real-time Java has been proposed by~\cite{MIT03}. Like us, they too
+assume a source language with programmer-managed memory regions, and
+focus on proving safety of programs written in that language. Their
+source language admits various kinds of regions in order to support
+shared-memory concurrency. However, all their regions have
+lexically-scoped lifetimes. In contrast, we admit regions with
 dynamically determined lifetimes in order to support message-passing
 concurrency. We borrow outlives relation from their formal
-development, and our type system bears some similarities to theirs.
+development, and our type system bears some similarities to their's.
 However, our language also admits parametric polymorphism (generics)
-and higher-order functions.
-%, whose interaction is non-trivial in context of a region type system. 
-% On the other hand, our type system eschews the notions of region kinds
-% and ownership, leading to a more succinct formalization. 
-Moreover, we establish type safety and transfer safety results that
-formalize the guarantees provided by our system. While~\cite{MIT03}'s
-language is explicitly typed, our language comes equipped with full
-type inference.
-% Although there is some support for local
-% type inference, region types for methods and classes need to be
-% written explicitly. In contrast, our region type inference that
-% eliminates any such need.
+and higher-order functions, whose interaction is non-trivial in
+context of a region type system. On the other hand, our type system
+eschews the notions of region kinds and ownership, leading to a more
+succinct formalization. Furthermore, we establish a type safety result
+that formalizes its guarantees with respect to a well-defined
+operational semantics. Like Cyclone, ~\cite{MIT03}'s language is
+explicitly typed. Although there is some support for local type
+inference, region types for methods and classes need to be written
+explicitly. In contrast, we support full type inference that
+eliminates any such need.
 

### 2016-05-21 paper/type-system.tex
Typos fixed. Some changes from master merged.

--- 2015-11-20 paper/type-system.tex
@@
 $\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
-$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
-\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
+$\rgn_i \notin dom(\rhomap') \cup \rhoenv \cup \{\toprgn\}$, if $\hastyp{\emptyA,\rgn,
+\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
 $\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
-{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
+{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
 \end{theorem}

### 2016-05-27 paper/type-inference.tex
Typos fixed. Some changes from master merged.
Changes to type inference description (beginning)

--- 2015-11-20 paper/type-inference.tex
@@
 expressions, similar to the language introduced in
-\S~\ref{sec:overview}) to programs in \FB.  The elaboration proceeds
-in four steps. In the first step we make use of the observation that
-region types are \emph{refinements} of FGJ types with region
-annotations and constraints over such region annotations, and compute
-polymorphic region type templates for methods and classes from their
-FGJ types. The templates contain region variables ($\rho$) to denote
-unknown region annotations, and predicate variables ($\varphi$) to
-denote unknown constraints over such region annotations. Free region
-variables are generalized in types (hence, polymorphic). Second, we
-make use of the computed region type templates to elaborate
-expressions by introducing region variables to denote unknown region
-arguments in \C{new} expressions, method calls and function
-applications. While elaborating expressions, we also build a system of
-constraints that capture well-formedness requirements and subtyping
-relationships between type templates that must hold (as per the static
-semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
-Third, we lift expression elaboration and constraint generation to
-methods and classes. Finally, we solve the constraints by making use
-of our fixpoint constraint solving algorithm \csolvestar, which
-reduces the constraint solving problem to an abduction problem in a
-Herbrand constraint system, and then relies on \csolve, our abduction
-solver for that domain. 
+\S~\ref{sec:overview}) to programs in \FB.
+
+The elaboration consists of the following steps:
+\begin{enumerate}
+ \item The first step consists of the following two mutually dependent steps:
+   \begin{enumerate}
+     \item for every class determine a set of region variable parameters ($\rho$)
+       it should be parametrized over, and
+     \item for every field of the class, determine its region parametrized type by
+       generating as many fresh region variables as required by its class (type).
+   \end{enumerate}
+ \item  For every method and function type, determine its corresponding region
+   type template by introducing fresh region variable parameters as required by
+   the type of each parameter and return value (in addition to the allocation region
+   parameter).
+ \item Elaborate every \C{new} expression, method call or function application by
+   introducing fresh region variable parameters (as required by the region type
+   template of the called method or function).
+ \item Introduce for each class and method, a predicate variable ($\varphi$) to denote
+   an undetermined set of constraints over the region variable parameters of that
+   class/method.
+ \item We analyze each method, to generate a set of constraints that capture
+   well-formedness requirements and subtyping relationships between type templates
+   that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).
+ \item We solve the generated set of constraints using our fixpoint constraint
+   solving algorithm \csolvestar, which reduces the constraint solving problem to
+   an abduction problem in a Herbrand constraint system, and then relies on \csolve,
+   our abduction solver for that domain. 
+ \item We finally simplify the elaborated program by coalescing region variables
+   determined to be the same in the constraint solution.
+\end{enumerate}
+
+%%    This is used to replace every FGJ type in the program
+%%    compute a polymorphic region type template
+%%     for every class and method from their FGJ types. The templates contain
+%%     region variables ($\rho$) to denote unknown region annotations, and
+%%     predicate variables ($\varphi$) to denote unknown constraints over
+%%     such region annotations. Free region variables are generalized in these
+%%     types (which are, hence, polymorphic).
+%%  
+%%    Second, we make use of the computed region type templates to elaborate
+%% expressions by introducing region variables to denote unknown region
+%% arguments in \C{new} expressions, method calls and function
+%% applications.
+%% While elaborating expressions, we also build a system of
+%% constraints that capture well-formedness requirements and subtyping
+%% relationships between type templates that must hold (as per the static
+%% semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
+%% Third, we lift expression elaboration and constraint generation to
+%% methods and classes. Finally, we solve the constraints by making use
+%% of our fixpoint constraint solving algorithm \csolvestar, which
+%% reduces the constraint solving problem to an abduction problem in a
+%% Herbrand constraint system, and then relies on \csolve, our abduction
+%% solver for that domain.
+
 % Due to the presence of region-polymorphic
@@
 The type template for a recursively defined class is computed in two
-steps. First, all recursive occurances of the class among the types of
+steps. First, all recursive occurrences of the class among the types of
 its fields are ignored and the class is templatized as if it is a
-non-recursive class. Next, all the recursive occurances are
+non-recursive class. Next, all the recursive occurrences are
 templatized with respect to the class template computed in the first
@@
 To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
-recursive occurance in the type of $\C{next}$ field, and templatize
+recursive occurrence in the type of $\C{next}$ field, and templatize
 the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
@@
 % Observe that our strategy of trivially templatizing the recursive
-% occurances of a class type in its definition ensures that no new
+% occurrences of a class type in its definition ensures that no new
 % region variables are introduced.
@@
 to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
-Efficient algorithms to solve the later problem are known to
-exist~\cite{graph}.
+Algorithms to solve the later problem are known to
+exist~\cite{siam92}.
 

### 2016-05-27 paper/type-system.tex
Added labels to op. semantics rules
Changes to type inference description (beginning)

--- 2015-11-20 paper/type-system.tex
@@
 
-Fig.~\ref{fig:fb-opsem} defines a small-step operational semantics for
-\fbname via a five-place reduction relation:
+Fig.~\ref{fig:fb-opsem-1} and Fig.~\ref{fig:fb-opsem-2} define a small-step
+operational semantics for \fbname via a five-place reduction relation:
 \vspace*{-0.15in}
@@
 $\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
-$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
-\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
-$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
-{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
+$\rgn_i \notin dom(\rhomap') \cup \rhoenv \cup \{\toprgn\}$, if
+\[
+  \hastyp{\emptyA,\rgn, \cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
+\]
+, then
+\[
+  \hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
+{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
+\]
 \end{theorem}

### 2016-05-30 paper/type-inference.tex
Added high-level outline of type-inference

@@
 
-The elaboration consists of the following steps:
+Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
+inference algorithm.
+The algorithm consists of the following steps:
 \begin{enumerate}
- \item The first step consists of the following two mutually dependent steps:
+ \item The first step elaborates the input program by introducing \emph{(formal and actual) region parameters},
+and \emph{region variables} (representing yet undetermined \emph{actual region parameters})
+as follows:
 \begin{enumerate}
-     \item for every class determine a set of region variable parameters ($\rho$)
-       it should be parametrized over, and
-     \item for every field of the class, determine its region parametrized type by
-       generating as many fresh region variables as required by its class (type).
-   \end{enumerate}
+\item For every class determine a set of \emph{region parameters} ($\rho$)
+       it should be parametrized over.
+\item For every field of the class, determine its region parametrized type by
+       generating as many fresh \emph{actual region parameters} as required by its class (type),
+       which in turn become formal region parameters of the containing class.
 \item  For every method and function type, determine its corresponding region
-   type template by introducing fresh region variable parameters as required by
+   type template by introducing fresh region parameters as required by
    the type of each parameter and return value (in addition to the allocation region
@@
  \item Elaborate every \C{new} expression, method call or function application by
-   introducing fresh region variable parameters (as required by the region type
+   introducing \emph{fresh region variables} (as required by the region type
    template of the called method or function).
- \item Introduce for each class and method, a predicate variable ($\varphi$) to denote
-   an undetermined set of constraints over the region variable parameters of that
+\item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
+   an undetermined set of constraints over the region parameters of that
    class/method.
- \item We analyze each method, to generate a set of constraints that capture
+\end{enumerate}
+ \item In the second step, we analyze the program to generate a set of constraints that capture
    well-formedness requirements and subtyping relationships between type templates
@@
    our abduction solver for that domain. 
- \item We finally simplify the elaborated program by coalescing region variables
-   determined to be the same in the constraint solution.
+ \item We finally simplify the elaborated program by substituting the solution determined
+   (for the region variables and predicate variables).
 \end{enumerate}
+
+\begin{figure}
+\begin{numcodeml}
+Infer ($p$) =
+  let $q$ = IntroduceVariables($p$) in
+  let $(r,C)$ = GenerateConstraints($q$) in
+  match (SolveConstraints($C$) with
+  | None $\longrightarrow$ None
+  | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
+\end{numcodeml}
+
+\caption{The type inference algorithm}
+\label{fig:type-inference-algo}
+\end{figure}
+
+\begin{theorem}
+\emph{(\textbf{Soundness})}
+For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
+(1) $\absof{t} = p$, and
+(2) $t$ is well-typed.
+\end{theorem}
 

### 2016-05-31 paper/type-inference.tex
Updates to type inference description

@@
 
+\input{fb-constraint-generation}
+
 \input{fb-elabexpr}
@@
 \begin{center}
-\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+\( \isvalid{\phictxt \conj \varphi_i}
            {\phicstr} \qquad
-   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+   \isvalid{\phictxt \conj \varphi_i}
            {F_j(\varphi_j)}
@@
 \end{center}
-Where $\varphi_i$'s denote the unknown preconditions of the class and
-the method under elaboration. If the constraint is generated while
-checking the well-formedness of a type or elaborating an expression,
-then $\varphi_j$'s denote the unknown preconditions of classes and
-methods that were used in that type or expression.  Each use of a
-(region-polymorphic) class or a method may instantiate region
+where $\varphi_i$ and $\varphi_j$ denote the unknown preconditions of some
+methods or constructors.
+%
+%% If the constraint is generated while checking the well-formedness of a
+%% type or elaborating an expression, then $\varphi_j$'s denote the unknown
+%% preconditions of classes and methods that were used in that type or expression.
+%
+Each use of a (region-polymorphic) class or a method may instantiate region
 parameters differently, resulting in a different pending substitution
@@
 among region varibles (i.e., $\rho$'s), or relationships between
-region variables and identifiers. Each region variable occuring in
-$\phicstr$ has an associated well-formedness constraint, which
-specifies its unification domain. The unification domain of a
+region variables and identifiers.
+%
+Each region variable occuring in $\phicstr$ has an associated well-formedness
+constraint, which specifies its unification domain. The unification domain of a
 constraint is the union of unification domains of all region variables
@@
 constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
-any of the following conditions hold:
-\begin{itemize}
-\item $c_1$ constrains a predicate variable that $c_2$ uses. 
-\item $c_1$ constrains a (non-strict) subset of predicate variables
-that $c_2$ constraint.
-\end{itemize}
-The first condition intuitively corresponds to a case, where
+$c_1$ constrains a predicate variable that $c_2$ uses.
+
+The above condition intuitively corresponds to a case, where
 expression or type, whose region elaboration is constrained by $c_2$
@@
 % (and self-loops over recursive constraints).
-The second condition captures two kinds of dependencies. First, the
-dependency from the constraints over a method precondition to the
-constraints over the precondition of the class containing the method.
-This captures our preference that the constraints over a class's
-region parameters should not depend on the idiosyncracies of its
-methods. Any additional constraints required by any of its methods
-must be captured in the precondition of the method itself
-(well-formedness rules allow this possibility). The second condition
-adds bidirectional dependencies between validity constraints that
-constraint the same set of predicate variables.
 
@@
 
-\paragraph{Example} The dependency DAG ($G_C$) over validity
-constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
-shown below: %in Fig.~\ref{fig:pair-deps}. 
-\vspace*{-0.08in}
-\begin{figure}[H]
-\includegraphics[scale=0.6]{DepDAG.png}
-\end{figure}
-\vspace*{-0.08in}
-Each node (labeled $C_i$) is a set of constraints that belong to a
-strongly connected component in the dependency graph ($G_c$), hence
-are mutually dependent. All dependencies, except the self-dependency
-on $\mathbf{c_{10}}$, are type-2 dependencies.
+%% \paragraph{Example} The dependency DAG ($G_C$) over validity
+%% constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
+%% shown below: %in Fig.~\ref{fig:pair-deps}. 
+%% \vspace*{-0.08in}
+%% \begin{figure}[H]
+%% \includegraphics[scale=0.6]{DepDAG.png}
+%% \end{figure}
+%% \vspace*{-0.08in}
+%% Each node (labeled $C_i$) is a set of constraints that belong to a
+%% strongly connected component in the dependency graph ($G_c$), hence
+%% are mutually dependent. All dependencies, except the self-dependency
+%% on $\mathbf{c_{10}}$, are type-2 dependencies.
 
@@
 
-For the DAG in figure above, we consider the topological order $[C_1,
-C_2, C_3]$ of its transpose, and solve the sets of constraints in that
-order.  Solving the set ($C$) of constraints entails finding an
+%% For the DAG in figure above, we consider the topological order $[C_1,
+%% C_2, C_3]$ of its transpose, and solve the sets of constraints in that
+%% order.
+
+Solving the set ($C$) of constraints entails finding an
 assignment for $C$'s subject ($\varphi$), and also all the region

### 2016-06-01 paper/type-inference.tex
Type inference description

@@
    solving algorithm \csolvestar, which reduces the constraint solving problem to
-   an abduction problem in a Herbrand constraint system, and then relies on \csolve,
-   our abduction solver for that domain. 
- \item We finally simplify the elaborated program by substituting the solution determined
-   (for the region variables and predicate variables).
+   an abduction problem. 
+% in a Herbrand constraint system, and then relies on \csolve,
+% our abduction solver for that domain. 
+ \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
+  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in the earlier
+phases. We apply these substitutions to the elaborated program to produce the final program.
+Note that if the original program in $\absof{\FB}$ contains unsafe
+references, for example, a reference from a transferable region to a
+stack regions, then the constraints generated during the elaboration
+are not satisfiable. In such case, $\solve$ fails to solve
+constraints, causing the program to be rejected.
 \end{enumerate}
@@
 every definition in the class table $CT$, while accumulating
-constraints. The constraints are finally solved by $\solve$
-(\S\ref{sec:fb-constraintsolving}), which returns substitution
-functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
-predicate variables, respectively, introduced during templatization
-and elaboration stages. The substitutions are applied to the class
-table (and to the artifacts that make up the class table, recursively)
-to compute a class table that maps classes to their fully
-region-annotated definitions in $\FB$.
-
-Note that if the original program in $\absof{\FB}$ contains unsafe
-references, for example, a reference from a transferable region to a
-stack regions, then the constraints generated during the elaboration
-are not satisfiable. In such case, $\solve$ fails to solve
-constraints, causing the program to be rejected.
+constraints.
 

### 2016-06-01 paper/type-inference.tex
Type inference

@@
 \S~\ref{sec:overview}) to programs in \FB.
+
+\textbf{TODO: Mention that constraints are in the theory of partial orders?}
 
@@
 template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
-fresh. For a class, a template is computed in two steps. In the first
+fresh.
+\textbf{TODO: What about bounded types?}
+For a class, a template is computed in two steps. In the first
 step, we templatize the types of all its fields, constructor
@@
 well-formed region-annotated class definition in $\FB$.
+
+\textbf{TODO: discuss base class parameters}
 
@@
 
-Our constraint generation algorithm generates three kinds of
-constraints:
+Our constraint generation algorithm generates the following kinds of constraints:
 \begin{itemize}
@@
 over region variables ($\rhoenv$) in scope, and
-\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
-$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
-extended with predicate variables and \emph{pending
-substitutions}\footnote{we borrowed this terminology
-from~\cite{ltpldi08}}:
+\item Validity constraints of form $\isvalid{\phictxt \conj \varphi_i} {\phicstr}$
+where $\varphi_i$ is a predicate variable (representing the precondition of a
+method to be determined), $\phicstr$ is a region constraint that is \emph{required}
+to hold at a particular program point (within the method), and $\phictxt$ is
+a region constraint that is \emph{known} to hold at that program point.
+Each $\phi$ is a region constraint over region variables and region parameters:
 \begin{smathpar}
 \begin{array}{lcl}
-\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
-    \ALT F(\varphi) \ALT \phi \conj \phi\\
+\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
+\end{array}
+\end{smathpar}
+%
+%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
+%% of predicate variables and pending substitutions. While $\phictxt$
+%% captures relationships that are \emph{known} to hold between concrete
+%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
+%% $\phicstr$ captures relationships that are \emph{required} to hold
+%% among region varibles (i.e., $\rho$'s), or relationships between
+%% region variables and identifiers.
+%
+\item Validity constraints of the form $\isvalid{\phictxt \conj \varphi_i} {F_j(\varphi_j)}$
+generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
+$\varphi_i$ are as above).
+Here, $F$ denotes \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}}:
+it serves to bind formal region parameters in $\varphi_j$ to the actual region parameters
+at a particular invocation site.
+\begin{smathpar}
+\begin{array}{lcl}
 F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
@@
 \end{itemize}
+
 A pending substitution ($F$) is a substitution function over region
@@
 % region-polymorphic methods, where similar situation arises.
-In general, validity constraints generated by our algorithm assume one
-of the following two forms:
-\begin{center}
-\( \isvalid{\phictxt \conj \varphi_i}
-           {\phicstr} \qquad
-   \isvalid{\phictxt \conj \varphi_i}
-           {F_j(\varphi_j)}
-\)
-\end{center}
-where $\varphi_i$ and $\varphi_j$ denote the unknown preconditions of some
-methods or constructors.
+
 %
@@
 %
-Each use of a (region-polymorphic) class or a method may instantiate region
-parameters differently, resulting in a different pending substitution
-($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-of predicate variables and pending substitutions. While $\phictxt$
-captures relationships that are \emph{known} to hold between concrete
-region identifiers (i.e., $\rgn$'s) when the constraint was generated,
-$\phicstr$ captures relationships that are \emph{required} to hold
-among region varibles (i.e., $\rho$'s), or relationships between
-region variables and identifiers.
-%
+% Each use of a (region-polymorphic) class or a method may instantiate region
+% parameters differently, resulting in a different pending substitution
+% ($F_j$).
+
 Each region variable occuring in $\phicstr$ has an associated well-formedness
@@
   q.fst := y; q.snd := y; 
-  open r as $\rgn_0$ withroot p in
+  open r as p@$\rgn_0$ in
     letregion $\rgn_1$ in
@@
 
-\subsection{Constraint Solving}
-\label{sec:fb-constraintsolving}
-
-Our constraint generation algorithm traverses the entire program,
-performing elaboration and collecting constraints, which are
-subsequently solved en masse. The motivation behind the whole-program
-approach to constraint generation is twofold: it simplifies
-elaboration functions and makes presentation easier, and second, it
-naturally generalizes to mutual recursion. Nonetheless, we do not
-intend our type inference to be a whole-program analysis for (a). it
-preempts opportunities for separate compilation and dynamic linking,
-and (b). it is expensive and an overkill in most practical cases. We
-therefore reclaim the compositionality of type inference by solving
-the constraints in a compositional fashion. In more practical terms
-this means that our constraint solving algorithm visits and solves
-every constraint (or, every set of mutually dependent constraints)
-only once. It composes computed solutions to solve other constraints
-that depend on the solved constraints. Importantly, the failure to
-solve a dependent constraint does not result in backtracking. We now
-describe our compositional constraint solving algorithm in detail. To
-simplify the presentation of our algorithm, we assume that are no
-mutually recursive definitions in the source program. Recursive
-definitions are nonetheless allowed.
-
-\paragraph{Terminology} In a validity constraint, a predicate variable
-occuring on the left side of the turnstile is said to occur
-negatively, or with \emph{negative polarity}. In contrast, a predicate
-variable occuring on the right side is said to occur positively, or
-with \emph{positive polarity}. A validity constraint
-\emph{constrains} the set of predicate variables that occur
-negatively in the constraint, while it \emph{uses} the set of
-predicate variables that occur positively. A constraint is said to be
-\emph{recursive} if it constrains and uses a predicate variable.
-% A pair of validity
-% constraints are said to be \emph{mutually dependent} if there exists a
-% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
-% with opposite polarities in both the constraints. In such case, the
-% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
-% be mutually dependent. We call the transitive closure of mutual
-% dependency relation as \emph{transitive dependency} relation. Like
-% mutual dependency, transitive dependency is also extended to predicate
-% variables.
-
-Given a set of validity constraints, we first build a dependency graph
-($G_c$) with constraints as nodes, and dependencies between them
-captured as edges. There exists an edge from a constraint $c_2$ to a
-constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
-$c_1$ constrains a predicate variable that $c_2$ uses.
-
-The above condition intuitively corresponds to a case, where
-expression or type, whose region elaboration is constrained by $c_2$
-refers to a method or a class, whose unknown precondition is
-constrained by $c_1$. The common predicate variable ($\varphi$)
-represents the unknown precondition in this case. The dependency from
-$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
-before $c_2$ is solved, thus enforcing the rule that the precondition
-of a method must not depend on its calling context. 
-% Note that the first condition results in self-loops over recursive
-% constraints in the dependency graph.  Since mutually recursive
-% definitions refer each other, the first condition also results in
-% bidirectional dependencies between mutually recursive constraints
-% (and self-loops over recursive constraints).
-
-Next, we convert the dependency graph over constraints into a
-dependency DAG ($G_C$) over sets of constraints, where each set
-represents a strongly connected component in the dependency graph.
-
-%% \paragraph{Example} The dependency DAG ($G_C$) over validity
-%% constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
-%% shown below: %in Fig.~\ref{fig:pair-deps}. 
-%% \vspace*{-0.08in}
-%% \begin{figure}[H]
-%% \includegraphics[scale=0.6]{DepDAG.png}
-%% \end{figure}
-%% \vspace*{-0.08in}
-%% Each node (labeled $C_i$) is a set of constraints that belong to a
-%% strongly connected component in the dependency graph ($G_c$), hence
-%% are mutually dependent. All dependencies, except the self-dependency
-%% on $\mathbf{c_{10}}$, are type-2 dependencies.
-
-A dependency DAG makes the dependencies between constraints explicit.
-Constraints in each set are mutually dependent, and need to be solved
-simultaneously, whereas constraints in different sets can be solved as
-per any valid topological ordering of the graph's transpose.
-Accordingly, we obtain a topological ordering of nodes in the graph
-$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
-in that order. The solutions obtained after solving a constraint set
-are applied to the constraints in subsequent sets before attempting to
-solve them.  Consequently, when the turn of a constraint set ($C$)
-arrives during the constraint solving process, it satisfies certain
-properties:
-\begin{itemize}
-\item There exists only one predicate variable ($\varphi$) that is either
-constrained or used by the constraints in the set ($C$). The variable is
-called set's \emph{subject}. This property follows from (a). the fact
-that all the dependency constraints have already been solved (and
-solutions applied), and (b). the assumption that there are no mutually
-recursive definitions. 
-\item All the constraints that constrain the set's subject are present
-in the set. This follows from our definition of the dependency relation.
-\end{itemize}
-
-%% For the DAG in figure above, we consider the topological order $[C_1,
-%% C_2, C_3]$ of its transpose, and solve the sets of constraints in that
-%% order.
-
+\paragraph{Constraint Solution}
 Solving the set ($C$) of constraints entails finding an
-assignment for $C$'s subject ($\varphi$), and also all the region
-variables ($\rhobar$) that occur free in $C$, such that the solution
-satisfies well-formedness constraints on $\varphi$ and $\rhobar$. To
-simplify presentation, we think of $C$ as being parameterized on
-$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-\rbrack$. We now formalize the constraint satisfaction problem, and
-its solution.
+assignment for each predicate variable $\varphi$) and each region
+variable ($\rho$) that occurs free in $C$, such that the solution
+satisfies well-formedness constraints on $\varphi$ and $\rhobar$.
+% To simplify presentation, we think of $C$ as being parameterized on
+% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
+% \rbrack$. We now formalize the constraint satisfaction problem, and
+% its solution.
 
@@
 % \end{center}
-\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
-that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
-\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
-(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
-$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
 \end{itemize}
 \end{definition}
+
+\paragraph{Properties of generated constraints}
+The set of generated constraints have certain properties that allow us to
+solve the constraints efficiently.
+%
+Since $\FB$ does not admit null values or uninitialized variables, every region
+variable is forced to be unified with some concrete region, by the constraints.
 
@@
 null values or uninitialized variables, thus forcing every region
-variable to be unified with some concrete region. The constraint
-formula $\Phics$ captures all such unification constraints on
-$\rhobar$. Since the set of all concrete region identifiers is the
+variable to be unified with some concrete region.
+%
+The constraint formula $\Phics$ captures all such unification constraints on $\rhobar$.
+Since the set of all concrete region identifiers is the
 unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
@@
 \end{itemize}
+
+\textbf{TODO: Is the above inequality correct?}
+
 The above discussion hints at an algorithm to compute a solution to
@@
 
+\textbf{TODO: FP iteration should use $\Phicx$}
+
 Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
@@
 \varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake of
-brevity, we define $G(\varphi) = \Phicx \conj \bigwedge_i
+brevity, we define $G(\varphi) = \Phics \conj \bigwedge_i
 F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
@@
 % \end{proof}
+
+\subsection{Modular Constraint Solving}
+\label{sec:fb-constraintsolving}
+
+Our constraint generation algorithm traverses the entire program,
+performing elaboration and collecting constraints, which are
+subsequently solved en masse. The motivation behind the whole-program
+approach to constraint generation is twofold: it simplifies
+elaboration functions and makes presentation easier, and second, it
+naturally generalizes to mutual recursion. Nonetheless, we do not
+intend our type inference to be a whole-program analysis for (a). it
+preempts opportunities for separate compilation and dynamic linking,
+and (b). it is expensive and an overkill in most practical cases. We
+therefore reclaim the compositionality of type inference by solving
+the constraints in a compositional fashion. In more practical terms
+this means that our constraint solving algorithm visits and solves
+every constraint (or, every set of mutually dependent constraints)
+only once. It composes computed solutions to solve other constraints
+that depend on the solved constraints. Importantly, the failure to
+solve a dependent constraint does not result in backtracking. We now
+describe our compositional constraint solving algorithm in detail. To
+simplify the presentation of our algorithm, we assume that are no
+mutually recursive definitions in the source program. Recursive
+definitions are nonetheless allowed.
+
+\paragraph{Terminology} In a validity constraint, a predicate variable
+occuring on the left side of the turnstile is said to occur
+negatively, or with \emph{negative polarity}. In contrast, a predicate
+variable occuring on the right side is said to occur positively, or
+with \emph{positive polarity}. A validity constraint
+\emph{constrains} the set of predicate variables that occur
+negatively in the constraint, while it \emph{uses} the set of
+predicate variables that occur positively. A constraint is said to be
+\emph{recursive} if it constrains and uses a predicate variable.
+% A pair of validity
+% constraints are said to be \emph{mutually dependent} if there exists a
+% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
+% with opposite polarities in both the constraints. In such case, the
+% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
+% be mutually dependent. We call the transitive closure of mutual
+% dependency relation as \emph{transitive dependency} relation. Like
+% mutual dependency, transitive dependency is also extended to predicate
+% variables.
+
+Given a set of validity constraints, we first build a dependency graph
+($G_c$) with constraints as nodes, and dependencies between them
+captured as edges. There exists an edge from a constraint $c_2$ to a
+constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
+$c_1$ constrains a predicate variable that $c_2$ uses.
+
+The above condition intuitively corresponds to a case, where
+expression or type, whose region elaboration is constrained by $c_2$
+refers to a method or a class, whose unknown precondition is
+constrained by $c_1$. The common predicate variable ($\varphi$)
+represents the unknown precondition in this case. The dependency from
+$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
+before $c_2$ is solved, thus enforcing the rule that the precondition
+of a method must not depend on its calling context. 
+% Note that the first condition results in self-loops over recursive
+% constraints in the dependency graph.  Since mutually recursive
+% definitions refer each other, the first condition also results in
+% bidirectional dependencies between mutually recursive constraints
+% (and self-loops over recursive constraints).
+
+Next, we convert the dependency graph over constraints into a
+dependency DAG ($G_C$) over sets of constraints, where each set
+represents a strongly connected component in the dependency graph.
+
+%% \paragraph{Example} The dependency DAG ($G_C$) over validity
+%% constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
+%% shown below: %in Fig.~\ref{fig:pair-deps}. 
+%% \vspace*{-0.08in}
+%% \begin{figure}[H]
+%% \includegraphics[scale=0.6]{DepDAG.png}
+%% \end{figure}
+%% \vspace*{-0.08in}
+%% Each node (labeled $C_i$) is a set of constraints that belong to a
+%% strongly connected component in the dependency graph ($G_c$), hence
+%% are mutually dependent. All dependencies, except the self-dependency
+%% on $\mathbf{c_{10}}$, are type-2 dependencies.
+
+A dependency DAG makes the dependencies between constraints explicit.
+Constraints in each set are mutually dependent, and need to be solved
+simultaneously, whereas constraints in different sets can be solved as
+per any valid topological ordering of the graph's transpose.
+Accordingly, we obtain a topological ordering of nodes in the graph
+$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
+in that order. The solutions obtained after solving a constraint set
+are applied to the constraints in subsequent sets before attempting to
+solve them.  Consequently, when the turn of a constraint set ($C$)
+arrives during the constraint solving process, it satisfies certain
+properties:
+\begin{itemize}
+\item There exists only one predicate variable ($\varphi$) that is either
+constrained or used by the constraints in the set ($C$). The variable is
+called set's \emph{subject}. This property follows from (a). the fact
+that all the dependency constraints have already been solved (and
+solutions applied), and (b). the assumption that there are no mutually
+recursive definitions. 
+\item All the constraints that constrain the set's subject are present
+in the set. This follows from our definition of the dependency relation.
+\end{itemize}
+
+%% For the DAG in figure above, we consider the topological order $[C_1,
+%% C_2, C_3]$ of its transpose, and solve the sets of constraints in that
+%% order.

### 2016-06-01 paper/type-inference.tex
type inference

@@
 
-\name's region type system imposes a heavy annotation burden, and the
-cost of manually annotating C\# standard libraries with region types
-is prohibitive. We now present our region type inference algorithm
-that completely eliminates the need to write region type annotations,
-except on some higher-order functions. Formally, the type inference
+\textbf{TO DO:}
+\begin{itemize}
+\item Add destructive object-field-update \emph{x.f := y} to the language.
+\item Add if-then-else to the language.
+\item Note that the claim that all region variables will be unified with some concrete
+region fails in the case of a method that recursively calls itself (and, hence, does not
+terminate). In this case, the region variables in the return type are not constrained.
+Similarly for non-terminating mutual recursion. These are uninteresting (in essence, the
+return type can be taken to be unit instead).
+\item Discuss the safety of virtual method invocation, which provides a limited form
+of existentially quantifying regions and instantiating (unpacking) them because a
+derived class may have more region parameters than a base class.
+\item Discuss the subtyping rule used to check an overriding method (in the static
+semantics).
+\item Mention that constraints are in the theory of partial orders?
+\end{itemize}
+
+\name's region type system imposes a heavy annotation burden, and
+manually annotating C\# standard libraries with region types
+can be tedious. We now present our region type inference algorithm
+that eliminates the need to write region type annotations.
+% except on some higher-order functions.
+Formally, the type inference
 algorithm is an elaboration function from programs in $\absof{\FB}$
@@
 \S~\ref{sec:overview}) to programs in \FB.
-
-\textbf{TODO: Mention that constraints are in the theory of partial orders?}
 
@@
 \label{sec:fb-templatization}
+
+The first step involves the following two mutually dependent ideas.
+
+\emph{Region Parametrization}. For every class \C{C}, we identify
+a sequence of region parameters $\pi_0, \cdots \pi_n$ that \C{C}
+should be parametric over.
+
+\emph{Instantiation of Region Parametric Classes}. For every region
+parametric class $\C{C}\langle \pi_0, \cdots, \pi_n \rangle$ so determined,
+we replace every instance of class \C{C} in the program by an instance
+$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
+are fresh identifiers.
+
+In particular, we determine the sequence of region parameters for a class \C{C}
+by starting with the sequence of region parameters required for the type of
+the base class of \C{C}, and adding fresh region parameters for the type of every field
+in the class.
+The class \C{Object} has a single region parameter $\pi_0$ (the allocation region).
 

### 2016-06-02 paper/type-inference.tex
type inference

@@
 \S~\ref{sec:overview}) to programs in \FB.
+
+\paragraph{Terminology.}
+In the sequel, we will be using \emph{region identifiers} in four different roles.
+A region identifier may serve as
+(a) A \emph{formal region parameter} of a class or method, or
+(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
+(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
+(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
+of a method invocation or object alloation,
+which will be bound to a region identifier of one of the three
+preceding kinds by the end of the type inference.
 
@@
 
-The first step involves the following two mutually dependent ideas.
+The first phase is an iterative process involving the following two
+mutually dependent steps.
 
@@
 
-\emph{Instantiation of Region Parametric Classes}. For every region
+\emph{Parameter Instantiation of Types}. For every region
 parametric class $\C{C}\langle \pi_0, \cdots, \pi_n \rangle$ so determined,
@@
 
-In particular, we determine the sequence of region parameters for a class \C{C}
-by starting with the sequence of region parameters required for the type of
-the base class of \C{C}, and adding fresh region parameters for the type of every field
-in the class.
-The class \C{Object} has a single region parameter $\pi_0$ (the allocation region).
+We identify the region parameters of classes as follows.
+The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
+The region parameters for any other non-recursive class \C{C} is determined
+only after the region parameters of any class that \C{C} depends on have been
+determined: this includes the base-class \C{B} of \C{C} and the class (type)
+of any of its fields.
+We replace the type \C{T} of every field in \C{C} by its instantiated type,
+using fresh region parameters as needed.
+The sequence of region parameters for \C{C} is defined to be
+the sequence of region parameters for the base class \C{B} concatenated
+with the list of all  fresh region parameters introduced while instantiating the types
+of the fields in the class.
 

### 2016-06-02 paper/type-inference.tex
type inference

@@
 
-\subsection{Region Type Templates}
+\subsection{Region Parametrization and Instantiation}
 \label{sec:fb-templatization}
@@
 \emph{Region Parametrization}. For every class \C{C}, we identify
-a sequence of region parameters $\pi_0, \cdots \pi_n$ that \C{C}
+a sequence of formal region parameters $\pi_0, \cdots \pi_n$ that \C{C}
 should be parametric over.
 
-\emph{Parameter Instantiation of Types}. For every region
-parametric class $\C{C}\langle \pi_0, \cdots, \pi_n \rangle$ so determined,
-we replace every instance of class \C{C} in the program by an instance
+\emph{Parameter Instantiation}.
+We then replace every instance of class \C{C} in the program by an instance
 $\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
-are fresh identifiers.
+are fresh identifiers denoting actual region parameters.
 
@@
 of any of its fields.
-We replace the type \C{T} of every field in \C{C} by its instantiated type,
+We replace every dependee type \C{T} in \C{C} by its instantiated type,
 using fresh region parameters as needed.
@@
 of the fields in the class.
+(The class inherits its allocation region from its base class. Note that if
+a class does not specify an explicit base class, it has an implicit base class
+\C{Object}.)
 
-Region type templates are FGJ types extended with fresh region
-variables ($\rho$) and predicate variables ($\varphi$) to denote
-unknown region annotations and region constraints, respectively. For
-instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
-template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
-region variable. Likewise, given the region-annotated definition of
+This transformation is illustrated below, using a non-generic \C{Pair} class:
+
+\begin{tabular}{ccc}
+\begin{minipage}{0.28\linewidth}
+\begin{codejava}
+class Pair
+  $\extends$ $\ObjZ$
+{
+  $\ObjZ$ fst;
+  $\ObjZ$ snd;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class Pair $\langle \rho_0, \rho_1, \rho_2 \rangle$
+  $\extends$ $\ObjZ \langle \rho_0 \rangle$
+{
+  $\ObjZ \langle \rho_1 \rangle$ fst;
+  $\ObjZ \langle \rho_2 \rangle$ snd;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
+
+The region parameters for a recursive class is computed in
+a similar fashion, with the following difference: any recursive
+field is ignored while instantiating region parameters for the fields of
+the class, and the region parameters of the recursive class are computed
+as before. We then do parameter instantiation for all recursive fields,
+such that their region annotations (the actial region parameters) are
+exactly the same as the (formal) region parameters of the class.
+The following example illustrates this for a non-generic \C{List} class.
+The resulting class represents a linked list with spine in the region
+$\rho_0$ and data objects in the region $\rho_1$.
+
+\begin{tabular}{ccc}
+\begin{minipage}{0.28\linewidth}
+\begin{codejava}
+class List
+  $\extends$ $\ObjZ$
+{
+  $\ObjZ$ data;
+  List next;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class List $\langle \rho_0, \rho_1 \rangle$
+  $\extends$ $\ObjZ \langle \rho_0 \rangle$
+{
+  $\ObjZ \langle \rho_1 \rangle$ data;
+  List $\langle \rho_0, \rho_1 \rangle$ next;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
+
+The above technique can be extended to mutually recursive classes in a
+straightforward manner, by simultaneously parametrizing them (and
+then instantiating them).
+
+\emph{Extensions for parametrized classes.}
+The techniques described above can be used for parametrized classes
+with the following extension. Consider a parametrized class \C{C $\langle$ T $\extends$ B $\rangle$}.
+The parametric type \C{T} is instantiated using the number of region parameters
+that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
+to be \C{Object}, and \C{T} is instantiated with one region parameter.
+
+Likewise, given the region-annotated definition of
 \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
@@
 variables, and $\varphi_0$ is a fresh predicate variable denoting
-unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
-template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
-fresh.
-\textbf{TODO: What about bounded types?}
-For a class, a template is computed in two steps. In the first
-step, we templatize the types of all its fields, constructor
-arguments, and arguments and return values of all its methods, along
-with the type of its superclass. In the second step, we generalize all
-the free region variables occuring in the templates computed in the
-first step as region parameters of the class. Finally, we add a fresh
-allocation region parameter ($\rhoalloc$) to the list of parameters,
-and introduce a new predicate variable ($\varphi$) to denote unknown
-constraints on region parameters. For example, consider the standard
-FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
-and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
-respectively. It can be templatized as following:
-\begin{center}
-\begin{codejava}
-class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
-          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
-  $a@\rho_0$ fst;
-  $b@\rho_1$ snd;
-  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
-  $...$
-}
-\end{codejava}
-\end{center}
-% \footnote{In reality, the generalized region variables only come
-% from the type templates of class's fields and superclass. Method
-% types do not contain any free region variables.}
-The template elides $\C{Pair}$'s methods, whose region type templates
-contain no free variables. Among the region parameters of the class
-template, $\rho_{0-4}$ are obtained by generalizing free region
-variables in the types of its class's fields, constructor arguments,
-and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
-fresh region variable denoting the allocation region argument. Fresh
-predicate variable $\varphi_0$ denotes unknown constraints over
+unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+
+Fresh predicate variable $\varphi_0$ denotes unknown constraints over
 $\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
 well-formed region-annotated class definition in $\FB$.
-
-\textbf{TODO: discuss base class parameters}
-
-The type template for a recursively defined class is computed in two
-steps. First, all recursive occurrences of the class among the types of
-its fields are ignored and the class is templatized as if it is a
-non-recursive class. Next, all the recursive occurrences are
-templatized with respect to the class template computed in the first
-step, such that their region annotations are exactly same as the
-region parameters of the class. For example, consider a generic
-$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
-type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
-where $\C{a}$ is the type of the data stored in the linked list node.
-To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
-recursive occurrence in the type of $\C{next}$ field, and templatize
-the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
-type template of $\C{data}$ field, we compute the class's template as
-following:
-\begin{center}
-$\C{ListNode}\inang{\C{a}\extends\ObjZ}
-\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
-\end{center}
-Next, we trivially templatize the type of $\C{next}$ field in the body
-of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
-The resulting class represents a linked list with spine in the region
-$\rhoalloc_0$ and data objects in the region $\rho_0$.
 
@@
 % region variables are introduced.
-The templatization technique we described above for recursive class
-definitions can be extended to mutually recursive definitions in a
-straightforward manner, by simultaneously templatizing them. Using the
+Using the
 techniques outlined above, we compute region type templates for all

### 2016-06-02 paper/type-inference.tex
type inference

@@
 
-\subsection{Region Parametrization and Instantiation}
+\subsection{Region Parametrization/Instantiation for Classes}
 \label{sec:fb-templatization}
@@
 
-Likewise, given the region-annotated definition of
-\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
-method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
-is \footnote{In our exposition, we assume that classes $\C{A}$ and
-$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
-$\ObjZ$, they accept one region parameter - the allocation region of
-their objects.}\footnote{We abuse arrow notation to also represent
-types of methods, but unlike function types, there is no allocation
-region annotation atop the arrow in a method type.}
-$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
-\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
-\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
-variables, and $\varphi_0$ is a fresh predicate variable denoting
-unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+\TODO{Discuss function-typed fields}
+
+%% Likewise, given the region-annotated definition of
+%% \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
+%% method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
+%% is \footnote{In our exposition, we assume that classes $\C{A}$ and
+%% $\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
+%% $\ObjZ$, they accept one region parameter - the allocation region of
+%% their objects.}\footnote{We abuse arrow notation to also represent
+%% types of methods, but unlike function types, there is no allocation
+%% region annotation atop the arrow in a method type.}
+%% $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
+%% \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
+%% \unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
+%% variables, and $\varphi_0$ is a fresh predicate variable denoting
+%% unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+
+\subsection{Region Parametrization/Instantiation for Methods}
+
+As the next step, we introduce region parameters for every method.
+We do this by instantiating the types of all parameters and the
+return value (of the method) using fresh region identifiers (as explained previously),
+and then generalizing these region identifiers as formal region parameters
+of the method.
+
+We then consider every method invocation in the program, and introduce
+fresh region variables representing the (yet unknown) actual region
+parameters for this particular invocation.
+%
+We similarly perform instantiation for every constructor invocation
+of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
+before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
+where $\rho_0, \cdots, \rho_n$ are fresh region variables.
 
@@
 well-formed region-annotated class definition in $\FB$.
-
-% Observe that our strategy of trivially templatizing the recursive
-% occurrences of a class type in its definition ensures that no new
-% region variables are introduced.
-Using the
-techniques outlined above, we compute region type templates for all
-classes bound in the class table of the $\absof{\FB}$ program before
-we proceed to elaborate expressions.
 

### 2016-06-03 paper/type-inference.tex
typeinference

@@
 \item Add if-then-else to the language.
+\item Omit constructors from classes (and use an implicit constructor).
 \item Note that the claim that all region variables will be unified with some concrete
@@
 
+\subsection{Constraints}
+\label{sec:fb-constraintsem}
+
+\input{fb-constraint-generation}
+
 Fresh predicate variable $\varphi_0$ denotes unknown constraints over
@@
 well-formed region-annotated class definition in $\FB$.
-
-\input{fb-constraint-generation}
-
-\input{fb-elabexpr}
-\input{fb-elabmeth}
-
-\subsection{Expression Elaboration}
-
-Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
-(a). replacing core types in variable declarations and \C{new}
-expressions with fresh region type templates, and (b). explicitly
-instantiating region parameters of methods with fresh region variables
-in method calls and function applications. This elaboration is
-performed with respect to the polymorphic type templates of classes
-and methods computed as per \S\ref{sec:fb-templatization}.
-
-Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-this elaboration for a subset of expressions in $\absof{\FB}$, whose
-corresponding $\FB$ expressions have been ascribed static semantics in
-Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
-context as the expression typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning. The function traverses expressions in a
-syntax-directed manner of a type checker, introducing fresh region
-type templates for unknown region types, while generating constraints
-over region and predicate variables. The precise nature of generated
-constraints is explained in \S\ref{sec:fb-constraintsem}, but in
-summary, they capture the relationships between the type templates of
-various subexpressions and their well-formedness. Note that
-$\elabExpr$ returns the region type template of the subexpression,
-which is used to generate constraints for the expression. Functions
-$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
-implement type well-formedness and subtype judgments from
-Fig.~\ref{fig:fb-staticsem}, respectively.
-
-\subsection{Method and Class Elaboration}
-
-Functions $\elabMeth$ and $\elabClass$ shown in
-Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
-class definitions, respectively. Both functions first build a context
-($\A$) containing a set ($ \rhoenv$) of region variables denoting
-regions that are currently live, a map ($\aenv$) mapping type
-variables to their bounds, and a constraint formula ($\phicx$)
-capturing constraints over live region variables. We use predicate
-variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
-
-Function $\elabMeth$ elaborates a method definition of class $B$. It
-calls $\elabExpr$ with the context $\A$, its allocation context
-parameter ($\rhoallocm$), and a type environment ($\env$) that
-contains region type bindings for all the arguments of the method,
-including the implicit $\C{this}$ argument. The region type template
-returned by $\elabExpr$ for the method body is checked against its
-expected type (derived from the type template of the method)
-generating more constraints. The function then returns the elaborated
-method definition and the set of constraints.
-
-$\elabClass$ elaborates the definition of a class $B$. It relies on
-$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
-hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
-constructor ($k$) and method definitions ($\bar{d}$), respectively. To
-the set of constraints returned by these functions, $\elabClass$ adds
-constraints generated by checking the well-formedness of the type
-templates of its superclass and fields, and also a new constraint
-capturing a couple of safety conditions: first, the allocation regions
-of objects referred by the instance variables should outlive the
-allocation region of the instance itself, and second, the allocation
-regions of a class type and its superclass type must be the same.
-
-Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
-every definition in the class table $CT$, while accumulating
-constraints.
-
-\subsection{Constraints}
-\label{sec:fb-constraintsem}
 
@@
 %% order.
+
+\subsection{Expression Elaboration}
+
+\input{fb-elabexpr}
+\input{fb-elabmeth}
+
+%Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
+%(a). replacing core types in variable declarations and \C{new}
+%expressions with fresh region type templates, and (b). explicitly
+%instantiating region parameters of methods with fresh region variables
+%in method calls and function applications. This elaboration is
+%performed with respect to the polymorphic type templates of classes
+%and methods computed as per \S\ref{sec:fb-templatization}.
+
+Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+this elaboration for a subset of expressions in $\absof{\FB}$, whose
+corresponding $\FB$ expressions have been ascribed static semantics in
+Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
+context as the expression typing judgment in
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
+retaining their meaning. The function traverses expressions in a
+syntax-directed manner of a type checker, introducing fresh region
+type templates for unknown region types, while generating constraints
+over region and predicate variables. The precise nature of generated
+constraints is explained in \S\ref{sec:fb-constraintsem}, but in
+summary, they capture the relationships between the type templates of
+various subexpressions and their well-formedness. Note that
+$\elabExpr$ returns the region type template of the subexpression,
+which is used to generate constraints for the expression. Functions
+$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
+implement type well-formedness and subtype judgments from
+Fig.~\ref{fig:fb-staticsem}, respectively.
+
+\subsection{Method and Class Elaboration}
+
+Functions $\elabMeth$ and $\elabClass$ shown in
+Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+class definitions, respectively. Both functions first build a context
+($\A$) containing a set ($ \rhoenv$) of region variables denoting
+regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ denoting the fact that such constraints are yet
+to be inferred.
+
+Function $\elabMeth$ elaborates a method definition of class $B$. It
+calls $\elabExpr$ with the context $\A$, its allocation context
+parameter ($\rhoallocm$), and a type environment ($\env$) that
+contains region type bindings for all the arguments of the method,
+including the implicit $\C{this}$ argument. The region type template
+returned by $\elabExpr$ for the method body is checked against its
+expected type (derived from the type template of the method)
+generating more constraints. The function then returns the elaborated
+method definition and the set of constraints.
+
+$\elabClass$ elaborates the definition of a class $B$. It relies on
+$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
+hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
+constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+the set of constraints returned by these functions, $\elabClass$ adds
+constraints generated by checking the well-formedness of the type
+templates of its superclass and fields, and also a new constraint
+capturing a couple of safety conditions: first, the allocation regions
+of objects referred by the instance variables should outlive the
+allocation region of the instance itself, and second, the allocation
+regions of a class type and its superclass type must be the same.
+
+Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
+every definition in the class table $CT$, while accumulating
+constraints.

### 2016-06-03 paper/type-inference.tex
typeinference

@@
 
-\paragraph{Terminology.}
+\paragraph{Overview.}
 In the sequel, we will be using \emph{region identifiers} in four different roles.
@@
 \begin{enumerate}
- \item The first step elaborates the input program by introducing \emph{(formal and actual) region parameters},
-and \emph{region variables} (representing yet undetermined \emph{actual region parameters})
-as follows:
-\begin{enumerate}
-\item For every class determine a set of \emph{region parameters} ($\rho$)
-       it should be parametrized over.
-\item For every field of the class, determine its region parametrized type by
-       generating as many fresh \emph{actual region parameters} as required by its class (type),
-       which in turn become formal region parameters of the containing class.
-\item  For every method and function type, determine its corresponding region
-   type template by introducing fresh region parameters as required by
-   the type of each parameter and return value (in addition to the allocation region
-   parameter).
- \item Elaborate every \C{new} expression, method call or function application by
-   introducing \emph{fresh region variables} (as required by the region type
-   template of the called method or function).
-\item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
-   an undetermined set of constraints over the region parameters of that
-   class/method.
-\end{enumerate}
- \item In the second step, we analyze the program to generate a set of constraints that capture
-   well-formedness requirements and subtyping relationships between type templates
+ \item The first step elaborates the input program by introducing \emph{formal region parameters}
+   (for each class and method), and \emph{region variables} (representing yet undetermined
+   \emph{actual region parameters}). We also introduce for each class and method, a
+   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
+   over the region parameters of that class/method.
+
+%%    as follows:
+%% \begin{enumerate}
+%% \item For every class determine a set of \emph{region parameters} ($\rho$)
+%%        it should be parametrized over.
+%% \item For every field of the class, determine its region parametrized type by
+%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
+%%        which in turn become formal region parameters of the containing class.
+%% \item  For every method and function type, determine its corresponding region
+%%    type template by introducing fresh region parameters as required by
+%%    the type of each parameter and return value (in addition to the allocation region
+%%    parameter).
+%%  \item Elaborate every \C{new} expression, method call or function application by
+%%    introducing \emph{fresh region variables} (as required by the region type
+%%    template of the called method or function).
+%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
+%%    an undetermined set of constraints over the region parameters of that
+%%    class/method.
+%% \end{enumerate}
+
+ \item In the second step, we analyze the program to generate a set of constraints
+   (over the region identifiers and the predicate variables)
    that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).
+
  \item We solve the generated set of constraints using our fixpoint constraint
    solving algorithm \csolvestar, which reduces the constraint solving problem to
-   an abduction problem. 
+   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
+   references, for example, a reference from a transferable region to a
+   stack region, then the constraints generated during the elaboration
+   are not satisfiable. In such a case, \csolvestar{} fails to solve
+   the constraints.
 % in a Herbrand constraint system, and then relies on \csolve,
 % our abduction solver for that domain.
+
  \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
-  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in the earlier
-phases. We apply these substitutions to the elaborated program to produce the final program.
-Note that if the original program in $\absof{\FB}$ contains unsafe
-references, for example, a reference from a transferable region to a
-stack regions, then the constraints generated during the elaboration
-are not satisfiable. In such case, $\solve$ fails to solve
-constraints, causing the program to be rejected.
+  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
+  step 1. We apply these substitutions to the elaborated program to produce the final program.
+
 \end{enumerate}

### 2016-06-03 paper/type-inference.tex
type inference

@@
-\section{Type Inference}
-\label{sec:type-inference}
-
-\textbf{TO DO:}
+\section{TO DO Items}
+
 \begin{itemize}
@@
 
+\section{Type Inference}
+\label{sec:type-inference}
+
 \name's region type system imposes a heavy annotation burden, and
@@
 \begin{enumerate}
- \item The first step elaborates the input program by introducing \emph{formal region parameters}
+ \item \emph{Region Parametrization}.
+   The first step elaborates the input program by introducing \emph{formal region parameters}
    (for each class and method), and \emph{region variables} (representing yet undetermined
@@
 
- \item In the second step, we analyze the program to generate a set of constraints
+ \item \emph{Constraint Generation}.
+   In the second step, we analyze the program to generate a set of constraints
    (over the region identifiers and the predicate variables)
-   that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).
-
- \item We solve the generated set of constraints using our fixpoint constraint
+   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).
+
+ \item \emph{Constraint Solving}.
+   We solve the generated set of constraints using our fixpoint constraint
    solving algorithm \csolvestar, which reduces the constraint solving problem to
@@
 Infer ($p$) =
-  let $q$ = IntroduceVariables($p$) in
+  let $q$ = IntroduceRegionParameters($p$) in
   let $(r,C)$ = GenerateConstraints($q$) in
@@
 
-The first phase is an iterative process involving the following two
-mutually dependent steps.
+The first phase is an iterative process involving the following three steps,
+the first two of which are mutually dependent on each other.
 
@@
 
+\emph{Predicate Variable Introduction}. For every class \C{C}, we introduce
+a fresh predicate variable $\varphi$, which represents the yet undetermined
+outlives constraints between the formal region parameters of class \C{C}.
+
 We identify the region parameters of classes as follows.
+
+\emph{Non-Recursive Classes}.
 The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
@@
 \begin{codejava}
-class Pair $\langle \rho_0, \rho_1, \rho_2 \rangle$
+class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
   $\extends$ $\ObjZ \langle \rho_0 \rangle$
@@
 
+\emph{Recursive Classes}.
 The region parameters for a recursive class is computed in
@@
 \begin{codejava}
-class List $\langle \rho_0, \rho_1 \rangle$
+class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
   $\extends$ $\ObjZ \langle \rho_0 \rangle$
@@
 
-\emph{Extensions for parametrized classes.}
+\emph{Type-Parametric Classes}.
 The techniques described above can be used for parametrized classes
@@
 
-\TODO{Discuss function-typed fields}
+\emph{Function Types}.
+Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
+instantiates function types as follows: the type of every parameter as well as the return value is
+instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
+are generalized as formal region parameters of the function type.
+For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
+$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_0,\rho_1} \rightarrow \C{Object}\inang{\rho_2}$.
 
@@
 and then generalizing these region identifiers as formal region parameters
-of the method.
+of the method. As for the classes, we also introduce a fresh predicate variable
+$\varphi$ for every method.
 
@@
 
-Fresh predicate variable $\varphi_0$ denotes unknown constraints over
-$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
-well-formed region-annotated class definition in $\FB$.
+%Fresh predicate variable $\varphi_0$ denotes unknown constraints over
+%$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
+%well-formed region-annotated class definition in $\FB$.
 

### 2016-06-06 paper/type-inference.tex
type inference

@@
 \paragraph{Overview.}
-In the sequel, we will be using \emph{region identifiers} in four different roles.
-A region identifier may serve as
-(a) A \emph{formal region parameter} of a class or method, or
-(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
-(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
-(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
-of a method invocation or object alloation,
-which will be bound to a region identifier of one of the three
-preceding kinds by the end of the type inference.
 
@@
 
-%Fresh predicate variable $\varphi_0$ denotes unknown constraints over
-%$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
-%well-formed region-annotated class definition in $\FB$.
-
-Our constraint generation algorithm generates the following kinds of constraints:
+The constraint generation algorithm mimics the static type checker, but accumulates
+constraints that must hold for the type checking to succeed.
+
+\paragraph{Syntax of Constraints.}
+The constraints are expressed in terms of region identifiers ($\pi$ and $\rho$)
+and predicate variables ($\varphi$).
+
+Recall that \emph{region identifiers} are used in four different roles.
+A region identifier may serve as
+(a) A \emph{formal region parameter} of a class or method, or
+(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
+(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
+(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
+of a method invocation or object alloation,
+which will be bound to a region identifier of one of the three
+preceding kinds by the end of the type inference.
+
+Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
+%% over a set of fixed formal region parameters,
+where the set of region-constraints $\phi$ is defined by:
+\begin{smathpar}
+\begin{array}{lcl}
+\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
+\end{array}
+\end{smathpar}
+
+Our constraints also make uses of \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}} $F$.
+A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
+used in a particular context:
+\begin{smathpar}
+\begin{array}{lcl}
+F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
+\end{array}
+\end{smathpar}
+For instance, in the validity constraint $\isvalid{\rgn_1 \outlives
+\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
+is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
+$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
+$\varphi$ if and only the formula obtained by substituting $\rgn_1$
+and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
+deducible from $\rgn_1 \outlives \rgn_2$.
+
+The constraints generated are of the following kinds:
 \begin{itemize}
+
 \item Well-formedness constraints of form $\rho \in \rhoenv$,
 restricting the domain of unification for a region variable ($\rho$)
-to the set ($\rhoenv$) of regions in scope,
-\item Well-formedness constraints of form
-$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
-variable ($\varphi$) to the set of all possible constraint formulas
-over region variables ($\rhoenv$) in scope, and
+to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
+
+\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
+variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
+regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
+
+\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
+  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
+  the region-constraint of its base class \C{B}).
+
+\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
+  (e.g., to indicate the region-constraint of a class \C{C} must imply the
+  region-constraint of each of its fields).
+
 \item Validity constraints of form $\isvalid{\phictxt \conj \varphi_i} {\phicstr}$
@@
 %
+
 \item Validity constraints of the form $\isvalid{\phictxt \conj \varphi_i} {F_j(\varphi_j)}$
@@
 $\varphi_i$ are as above).
-Here, $F$ denotes \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}}:
-it serves to bind formal region parameters in $\varphi_j$ to the actual region parameters
-at a particular invocation site.
-\begin{smathpar}
-\begin{array}{lcl}
-F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
-\end{array}
-\end{smathpar}
+
 \end{itemize}
 
-A pending substitution ($F$) is a substitution function over region
-variables/identifiers. They represent the substitutions that need to
-be carried out when a predicate variable ($\varphi$) is replaced by a
-concrete formula in a validity constraint. For instance, in the
-validity constraint $\isvalid{\rgn_1 \outlives
-\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
-is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
-$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only the formula obtained by substituting $\rgn_1$
-and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
-deducible from $\rgn_1 \outlives \rgn_2$.
+% A pending substitution ($F$) is a substitution function over region
+% variables/identifiers. They represent the substitutions that need to
+% be carried out when a predicate variable ($\varphi$) is replaced by a
+% concrete formula in a validity constraint.
 
@@
 occuring in the constraint.
+
+\emph{Class Constraints}. 
+Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
+fairly straightforward. It is the conjunction of
+\begin{itemize}
+\item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
+\item $\varphi_{\C{B}}$
+\item $[\rho_a/\rho_f]\varphi_{\C{F}}$
+\end{itemize}
 

### 2016-06-07 paper/type-inference.tex
type inference

@@
 
-\subsection{Region Parametrization/Instantiation for Classes}
+\subsection{Region Parametrization for Classes}
 \label{sec:fb-templatization}
 
-The first phase is an iterative process involving the following three steps,
+Region parametrization is an iterative process involving the following three steps,
 the first two of which are mutually dependent on each other.
 
-\emph{Region Parametrization}. For every class \C{C}, we identify
-a sequence of formal region parameters $\pi_0, \cdots \pi_n$ that \C{C}
-should be parametric over.
-
-\emph{Parameter Instantiation}.
+\emph{Introduction of Formal Region Parameters}.
+For every class \C{C}, we identify a sequence of formal region parameters
+$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.
+
+\emph{Introduction of Actual Region Parameters}.
 We then replace every instance of class \C{C} in the program by an instance
@@
   $\ObjZ \langle \rho_1 \rangle$ data;
-  List $\langle \rho_0, \rho_1 \rangle$ next;
+  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
 }
@@
 \emph{Type-Parametric Classes}.
-The techniques described above can be used for parametrized classes
-with the following extension. Consider a parametrized class \C{C $\langle$ T $\extends$ B $\rangle$}.
+Type parameters of classes are handled as follows.
+Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
 The parametric type \C{T} is instantiated using the number of region parameters
@@
 
-\subsection{Region Parametrization/Instantiation for Methods}
+\subsection{Region Parametrization for Methods}
 
@@
 \end{smathpar}
+We will use the term \emph{validity constraint} to denote an entailment constraint
+of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
 
@@
 \end{smathpar}
-For instance, in the validity constraint $\isvalid{\rgn_1 \outlives
+E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
 \rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
@@
 
-\item Validity constraints of form $\isvalid{\phictxt \conj \varphi_i} {\phicstr}$
+\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
 where $\varphi_i$ is a predicate variable (representing the precondition of a
@@
 a region constraint that is \emph{known} to hold at that program point.
-Each $\phi$ is a region constraint over region variables and region parameters:
-\begin{smathpar}
-\begin{array}{lcl}
-\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
-\end{array}
-\end{smathpar}
+
 %
@@
 
-\item Validity constraints of the form $\isvalid{\phictxt \conj \varphi_i} {F_j(\varphi_j)}$
+\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
 generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
@@
 
-% A pending substitution ($F$) is a substitution function over region
-% variables/identifiers. They represent the substitutions that need to
-% be carried out when a predicate variable ($\varphi$) is replaced by a
-% concrete formula in a validity constraint.
-
-% Validity constraints primarily result from checking if a region type
-% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
-% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
-% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
-% formal region parameters with actual region arguments satisfy its
-% precondition, where actuals are subsituted for formals. Since $B$'s
-% precondition could be unknown, the requirement results in a in a
-% validity constraint with predicate variables and pending
-% substitutions. Other sources of validity constraints are calls to
-% region-polymorphic methods, where similar situation arises.
-
-%
-%% If the constraint is generated while checking the well-formedness of a
-%% type or elaborating an expression, then $\varphi_j$'s denote the unknown
-%% preconditions of classes and methods that were used in that type or expression.
-%
-% Each use of a (region-polymorphic) class or a method may instantiate region
-% parameters differently, resulting in a different pending substitution
-% ($F_j$).
-
-Each region variable occuring in $\phicstr$ has an associated well-formedness
-constraint, which specifies its unification domain. The unification domain of a
-constraint is the union of unification domains of all region variables
-occuring in the constraint.
-
-\emph{Class Constraints}. 
-Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
-fairly straightforward. It is the conjunction of
+\paragraph{Constraint Solution}
+Solving the set ($C$) of constraints entails finding an
+assignment for each predicate variable $(\varphi$) and each region
+variable ($\rho$) that occurs free in $C$, that satisfies all
+the validity constraints as well as the well-formedness constraints on
+$\varphi$ and $\rhobar$.
+% To simplify presentation, we think of $C$ as being parameterized on
+% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
+% \rbrack$. We now formalize the constraint satisfaction problem, and
+% its solution.
+
+\begin{definition}
+\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
+A constraint satisfaction problem is a tuple
+$(\bar{\pi}, \bar{\rho}, \bar{\varphi}, \regionDeltaMap, \predDeltaMap, C)$,
+where $\bar{\pi}$ is a set of region constants,
+$\bar{\rho}$ is a set of region variables,
+$\bar{\varphi}$ is a set of predicate variables,
+$regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
+specifies a unification domain for each region variable,
+$\predDeltaMap : \bar{\varphi} \rightarrow 2^{\bar{\pi}}$ is a function that
+specifies the domain for each predicate variable,
+and $C$ is a set of validity constraints, where the $i$'th validity constraint
+assumes one of the following forms:
+\begin{center}
+\(
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {\phicstr^{i}}\qquad
+    \isvalid{\phictxt^{i} \conj \varphi}
+            {F_i(\varphi)}
+\)
+\end{center}
+The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
+where $\regionSubstFn$ is a map from $\rhobar$ to $\bar{\pi}$
+and $\predSubstFn$ is a map from $\bar{\varphi}$ to a region-constraint formula such that
 \begin{itemize}
-\item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
-\item $\varphi_{\C{B}}$
-\item $[\rho_a/\rho_f]\varphi_{\C{F}}$
+  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \bar{\rho}$,
+
+  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
+    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \bar{\varphi}$).
+
+  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
+    \regionSubstFn$.
+  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
 \end{itemize}
+\end{definition}
+
+\textbf{TODO: The following not yet integrated into above.}
 
@@
 
-\paragraph{Constraint Solution}
-Solving the set ($C$) of constraints entails finding an
-assignment for each predicate variable $\varphi$) and each region
-variable ($\rho$) that occurs free in $C$, such that the solution
-satisfies well-formedness constraints on $\varphi$ and $\rhobar$.
-% To simplify presentation, we think of $C$ as being parameterized on
-% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-% \rbrack$. We now formalize the constraint satisfaction problem, and
-% its solution.
-
-\begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
-satisfaction problem is a tuple $\CSP$, where $C\lbrack
-\varphi,\rhobar \rbrack$ is a set of validity constraints, where
-$i$'th validity constraint assumes one of the following forms:
-\begin{center}
-\(
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {\phicstr^{i}}\qquad
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {F_i(\varphi)}
-\)
-\end{center}
-$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
-union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
-unification domain ($\rhoenv$) of the CSP.  The solution to the
-constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
-$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
-$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
-constraint formula such that:
-\begin{itemize}
-\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
-$\tywf{\rhoenv_{\varphi}}{\phisol}$).
-\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
-is valid.
-% \begin{center}
-% \(
-%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
-%           {\phicstr^{i} \conj  F_i(\phisol))}
-% \)
-% \end{center}
-\end{itemize}
-\end{definition}
+\subsection{Solving the CSP}
+
+Solving the constraints is an iterative process, consisting of the following
+steps, and this process fails at any point if it determines that no solution
+exists for the set of constraints:
+
+\begin{enumerate}
+
+\item
+    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
+$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
+If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
+solver fails.
+
+\item
+When a region variable $\rho$ is unified with a region constant $\pi$,
+if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
+and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
+If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
+
+\item
+Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$,
+where $\pi_i$ and $\pi_j$ are both region constants.
+If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
+$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
+Otherwise, the constraint solver fails, since no valid solution is possible.
+
+\item
+  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
+  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
+  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$.
+
+\end{enumerate}
+
+%% \emph{Class Constraints}. 
+%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
+%% fairly straightforward. It is the conjunction of
+%% \begin{itemize}
+%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
+%% \item $\varphi_{\C{B}}$
+%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
+%% \end{itemize}
 
@@
 variable is forced to be unified with some concrete region, by the constraints.
-
-\subsection{Solving the CSP}
-
-The first step of solving the CSP $\CSP$ is to cast it as an
-equivalent problem $\csp$ involving a single validity constraint. The
-single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
-\begin{center}
-\(
-  \isvalid{ \Phicx \conj \varphi}
-        {\Phics \conj  \bigwedge_i F_i(\varphi)}
-\)
-\end{center}
-Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
-\phicstr^{i}$. To see why both CSPs are equivalent, consider two
-distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
-$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
-\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
-generated before $c_j$ by the constraint generation algorithm. Observe
-that when our constraint generation algorithm generates a constraint,
-all relationships between concrete region identifiers referred by the
-consequent of constraint are already present in its antecedent (this
-includes the unification domains of region variables referred by the
-consequent).  Since no new relationships between existing regions are
-added when a new constraints is generated, $\phictxt^j$ either
-describes the same relationships that are already described by
-$\phictxt^i$, or describes relationships among region identifiers that
-are new, and not relevant to $c_i$. Therefore, strengthening the
-context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
-nor strengthens $c_i$ (resp.  $c_j$).
-
-For the constraint sets $C_1$, $C_2$, and $C_3$ in the running
-example, equivalent constraints are $c_{12}$, $c_{13}$, and $c_{14}$,
-respectively, shown below:
-\begin{smathpar}
-\begin{array}{l}
-  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
-     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
-     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
-  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
-  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
-    {
-        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
-        \rho_7 = \rho_9 \conj 
-    }\\
-  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
-        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
-\end{array}
-\end{smathpar}
 

### 2016-06-08 paper/type-inference.tex
type inference

@@
 
+\paragraph{Constraint Generation.}
+
 \paragraph{Constraint Solution}
@@
 $\bar{\varphi}$ is a set of predicate variables,
-$regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
+$\regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
 specifies a unification domain for each region variable,
@@
 
+\paragraph{Properties of generated constraints}
+The set of generated constraints have two interesting properties that allow us to
+solve the constraints efficiently.
+%
+The first property is that every region variable is forced to be unified with some
+region constant by the constraints, except in one special case.
+This follows because $\FB$ does not admit uninitialized variables.
+(The only case where this does not hold is in the case of a recursive function
+that calls itself in a non-terminating fashion: in this case, the function never
+returns a value, and so the return-value can be typed as anything. In the sequel,
+we assume that the return value of such a function is typed to be \C{unit}.)
+% , and the type system uniquely determines the region parameters of every value
+%
+The second property is that the \emph{context region constraint} $\phictxt$
+occurring on the antecedent of any validity constraint is a conjunction of
+outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
+is a static region identifier and $\pi$ is a region constant (either a formal
+region parameter, or static region identifier, or open transferable region identifier).
+Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
+it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+
+\paragraph{Constraint Solver.}
 Solving the constraints is an iterative process, consisting of the following
@@
 \item
-Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$,
+\label{item:context}
+Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
 where $\pi_i$ and $\pi_j$ are both region constants.
@@
 $\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-Otherwise, the constraint solver fails, since no valid solution is possible.
+Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
 
@@
   we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
-
 \end{enumerate}
+
+Point~\ref{item:context} makes use of the special structure of the generated
+constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
+form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
+region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+
+This iterative process is a standard fixed point computation process.
+It is possible to encode this computation using a set of Datalog rules, and compute the
+solution using any Datalog engine.
+Alternatively, note that the above computation, in the absence of context constraints
+$\phictxt$ on the left-hand-side of the entailment, is a special instance of context-free
+reachability in graphs. Algorithms for context-free reachability can be adapted to
+incorporate the above treatment of context constraints $\phictxt$.
 
@@
 %% \end{itemize}
-
-\paragraph{Properties of generated constraints}
-The set of generated constraints have certain properties that allow us to
-solve the constraints efficiently.
-%
-Since $\FB$ does not admit null values or uninitialized variables, every region
-variable is forced to be unified with some concrete region, by the constraints.
 

### 2016-06-10 paper/type-inference.tex
type inference

@@
-\section{TO DO Items}
-
-\begin{itemize}
-\item Add destructive object-field-update \emph{x.f := y} to the language.
-\item Add if-then-else to the language.
-\item Omit constructors from classes (and use an implicit constructor).
-\item Note that the claim that all region variables will be unified with some concrete
-region fails in the case of a method that recursively calls itself (and, hence, does not
-terminate). In this case, the region variables in the return type are not constrained.
-Similarly for non-terminating mutual recursion. These are uninteresting (in essence, the
-return type can be taken to be unit instead).
-\item Discuss the safety of virtual method invocation, which provides a limited form
-of existentially quantifying regions and instantiating (unpacking) them because a
-derived class may have more region parameters than a base class.
-\item Discuss the subtyping rule used to check an overriding method (in the static
-semantics).
-\item Mention that constraints are in the theory of partial orders?
-\end{itemize}
-
 \section{Type Inference}
@@
 
-\textbf{TODO: The following not yet integrated into above.}
+\subsection{Solving the CSP}
+
+\paragraph{Properties of generated constraints}
+The set of generated constraints have two interesting properties that allow us to
+solve the constraints efficiently.
+%
+The first property is that every region variable is forced to be unified with some
+region constant by the constraints, except in one special case.
+This follows because $\FB$ does not admit uninitialized variables.
+(The only case where this does not hold is in the case of a recursive function
+that calls itself in a non-terminating fashion: in this case, the function never
+returns a value, and so the return-value can be typed as anything. In the sequel,
+we assume that the return value of such a function is typed to be \C{unit}.)
+% , and the type system uniquely determines the region parameters of every value
+%
+The second property is that the \emph{context region constraint} $\phictxt$
+occurring on the antecedent of any validity constraint is a conjunction of
+outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
+is a static region identifier and $\pi$ is a region constant (either a formal
+region parameter, or static region identifier, or open transferable region identifier).
+Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
+it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+
+\paragraph{Constraint Solver.}
+Solving the constraints is an iterative process, consisting of the following
+steps, and this process fails at any point if it determines that no solution
+exists for the set of constraints:
+
+\begin{enumerate}
+
+\item
+    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
+$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
+If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
+solver fails.
+
+\item
+When a region variable $\rho$ is unified with a region constant $\pi$,
+if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
+and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
+If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
+
+\item
+\label{item:context}
+Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
+where $\pi_i$ and $\pi_j$ are both region constants.
+If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
+$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
+Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
+
+\item
+  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
+  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
+  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
+\end{enumerate}
+
+Point~\ref{item:context} makes use of the special structure of the generated
+constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
+form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
+region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+
+This iterative process is a standard fixed point computation process.
+It is possible to encode this computation using a set of Datalog rules,
+and compute the solution using any Datalog engine.
+Alternatively, algorithms for context-free reachability can be adapted
+to solve the constraints, as explained below.
+
+Given a set $S$ of outlives constraints, we define the directed
+graph $G(S)=(V(S),E(S))$ as follows.
+Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
+which we will also refer to as $\rho$.
+Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
+an edge from $\rho_1$ to $\rho_2$.
+Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
+for the pair of constraints $\rho_1 \outlives \rho_2$ and
+$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
+by a pair of edges.
+%
+It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
+there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
+%
+Thus, simple graph reachability (transitive closure) algorithms can be
+used to identify all the logical consequences of a set of outlives
+constraints.
+
+Given a set of context-free validity constraints, we can generalize the
+above approach and reduce the problem to one of context-free reachability
+in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
+%
+Algorithms for context-free reachability can also be adapted to incorporate
+the above treatment of context constraints $\phictxt$.
+%
+Further details omitted.
+
+%% \emph{Class Constraints}. 
+%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
+%% fairly straightforward. It is the conjunction of
+%% \begin{itemize}
+%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
+%% \item $\varphi_{\C{B}}$
+%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
+%% \end{itemize}
+
+\section{TO DO Items}
+
+\begin{itemize}
+\item Add destructive object-field-update \emph{x.f := y} to the language.
+\item Add if-then-else to the language.
+\item Omit constructors from classes (and use an implicit constructor).
+\item Discuss the safety of virtual method invocation, which provides a limited form
+of existentially quantifying regions and instantiating (unpacking) them because a
+derived class may have more region parameters than a base class.
+\item Discuss the subtyping rule used to check an overriding method (in the static
+semantics).
+\item Mention that constraints are in the theory of partial orders?
+\item Update discussion of examples
+\end{itemize}
+
+\textbf{TODO: Miscellaneous points to be integrated.}
+
+To see why, recall that all the constraints on
+identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
+by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
+Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
+only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
+solution $\phisol$ cannot exist.
+
+\textbf{TODO: Is the above inequality correct?}
+
+\subsection{Example}
 
@@
 % the last constraint is generated from the recursive call.
-
-\subsection{Solving the CSP}
-
-\paragraph{Properties of generated constraints}
-The set of generated constraints have two interesting properties that allow us to
-solve the constraints efficiently.
-%
-The first property is that every region variable is forced to be unified with some
-region constant by the constraints, except in one special case.
-This follows because $\FB$ does not admit uninitialized variables.
-(The only case where this does not hold is in the case of a recursive function
-that calls itself in a non-terminating fashion: in this case, the function never
-returns a value, and so the return-value can be typed as anything. In the sequel,
-we assume that the return value of such a function is typed to be \C{unit}.)
-% , and the type system uniquely determines the region parameters of every value
-%
-The second property is that the \emph{context region constraint} $\phictxt$
-occurring on the antecedent of any validity constraint is a conjunction of
-outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-is a static region identifier and $\pi$ is a region constant (either a formal
-region parameter, or static region identifier, or open transferable region identifier).
-Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-
-\paragraph{Constraint Solver.}
-Solving the constraints is an iterative process, consisting of the following
-steps, and this process fails at any point if it determines that no solution
-exists for the set of constraints:
-
-\begin{enumerate}
-
-\item
-    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
-$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
-If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
-solver fails.
-
-\item
-When a region variable $\rho$ is unified with a region constant $\pi$,
-if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
-and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
-If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
-
-\item
-\label{item:context}
-Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
-where $\pi_i$ and $\pi_j$ are both region constants.
-If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
-$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
-
-\item
-  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
-  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
-  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
-\end{enumerate}
-
-Point~\ref{item:context} makes use of the special structure of the generated
-constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
-form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
-region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-
-This iterative process is a standard fixed point computation process.
-It is possible to encode this computation using a set of Datalog rules, and compute the
-solution using any Datalog engine.
-Alternatively, note that the above computation, in the absence of context constraints
-$\phictxt$ on the left-hand-side of the entailment, is a special instance of context-free
-reachability in graphs. Algorithms for context-free reachability can be adapted to
-incorporate the above treatment of context constraints $\phictxt$.
-
-%% \emph{Class Constraints}. 
-%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
-%% fairly straightforward. It is the conjunction of
-%% \begin{itemize}
-%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
-%% \item $\varphi_{\C{B}}$
-%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
-%% \end{itemize}
-
-\subsubsection{Non-Recursive Constraints}
-\label{sec:csolve}
-
-We first describe how we solve a non-recursive CSP $\csp$ where
-$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
-\conj \varphi}{\Phics}$.
-
-Our approach is based on the observation that $\FB$ does not admit
-null values or uninitialized variables, thus forcing every region
-variable to be unified with some concrete region.
-%
-The constraint formula $\Phics$ captures all such unification constraints on $\rhobar$.
-Since the set of all concrete region identifiers is the
-unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
-with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
-exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
-\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
-in which case the well-formedness constraint on $\rho_i$ is not
-satisfied, and constraint solving must fail.  Therefore, there exists
-a unique assignment $\substFn$ such that $c\lbrack
-\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
-$\phi_{sol}$.
-
-To obtain $\phisol$, we make use of another observation. Let $c\lbrack
-\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
-\begin{center}
-\(
-  \isvalid{ \Phicx \conj \varphi}
-        {\Phics'}
-\)
-\end{center}
-Consider a maximally weak formula $\phi$ such that
-$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
-% Since $\phi$ is
-% maximally weak, there does not exist a $\phi'$ satisfying all the
-% below three conditions:
-% \begin{itemize}
-% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
-% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
-% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
-% \end{itemize}
-Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
-\subseteq \rhoenv$, we have two cases:
-\begin{itemize}
-
-\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
-a solution to $\varphi$. 
-% Furthermore, it is the weakest solution, because it is equivalent to
-% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
-% \substFn(\rho_j))$, and any other solution has to stronger than
-% $\Phics$ under the same context.
-
-\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
-contains at least one equality or outlives constraint on two region
-identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
-\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
-\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
-constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
-is not maximally weak). Let us denote such constraint on $\rgn_i$ and
-$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
-$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
-Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
-$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
-$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
-\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
-this is impossible. To see why, recall that all the constraints on
-identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
-by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
-Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
-only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
-solution $\phisol$ cannot exist.
-\end{itemize}
-
-\textbf{TODO: Is the above inequality correct?}
-
-The above discussion hints at an algorithm to compute a solution to
-$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
-{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
-{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
-solution to $\varphi$. We now describe a graph-based algorithm to
-compute maximally weak $\phisol$.
-
-\paragraph{Definitions} Given a constraint formula $\phi$, we define
-its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
-vertices ($V(\phi)$) are free region variables and identifiers in
-$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
-in\footnote{We alternate between viewing a constraint formula ($\phi$)
-as a set of constraints and a conjunction of constraints in this
-description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
-\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
-$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
-conjunction of symmetric outlives constraints for the purpose of graph
-encoding. Conversely, given a digraph $G$, we define its constraint
-encoding $\Phi(G)$ in a straightforward manner. We say that a graph
-$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
-\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
-exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
-exist a path between same pair of vertices in $G_1$.
-
-A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
-\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
-graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
-connected as $G(\phicstr)$. The problem of computing such a $G$ is
-equivalent to the problem finding of finding minimum number of edges
-to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
-Algorithms to solve the later problem are known to
-exist~\cite{siam92}.
-
-Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
-graph augmentation problems, as described above, results in the
-following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
-outlives constraints are replaced with equalities):
-\begin{center}
-\(
-  \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
-     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
-     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
-\)\\
-\(
-  \varphi_2 \Leftrightarrow \rho_5 = \rho_0
-\)
-\end{center}
-
-\subsubsection{Solving Recursive Constraints}
-
-\textbf{TODO: FP iteration should use $\Phicx$}
-
-Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
-$c_{14}$ gives us a recursive constraint of the following form:
-\begin{center}
-\(
-  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
-\)
-\end{center}
-Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
-$F$ is a substitution function, not necessarily idempotent. We now
-extend constraint solving to recursive CSP $\csp$, where $c\lbrack
-\varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx \conj
-\varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake of
-brevity, we define $G(\varphi) = \Phics \conj \bigwedge_i
-F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
-\begin{center}
-\(
-  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
-\)
-\end{center}
-To solve the above recursive constraint, we start with the observation
-that the set of all constaints ($\phi$) over $\rhoenv \cup
-\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
-\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
-monotone over the lattice:
-\begin{center}
-  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
-  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
-\end{center}
-From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
-point ($\phi_f$), with following properties:
-\begin{itemize}
-\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
-$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
-\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
-G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
-{\cdot} {\phi_f' \Rightarrow \phi_f}$.
-\end{itemize}
-We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
-convert the recursive constraint to the following non-recursive
-constraint:
-\begin{center}
-\(
-  \isvalid{\Phicx \conj \varphi}{\phi_f}
-\)
-\end{center}
-The technique described in \S~\ref{sec:csolve} now suffices to solve
-the above constraint.
-
-Let $H(\varphi_2)$ denote the consequent of the recursive constraint
-$c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
-\begin{center}
-\(
-  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
-  \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
-  \rgn_0 = \rgn_0
-\)
-\end{center}
-Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
-\conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
-results in the following solution for $\varphi_2$ that meets its
-well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
-\rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
-\begin{center}
-\(
-  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
-  \rho_9 \conj \rho_8 = \rho_9
-\)
-\end{center}
-
-% Note that $F$ is a non-idempotent substitution function with following
-% properties:
-% \begin{itemize}
-% \item $F$ is a monotone: 
-% \begin{center}
-%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
-%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
-% \end{center}
-
-% \item $F$ distributes over conjunction (mind the syntactic equality):
-% \begin{center}
-%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
-%   F(\phi_2)$
-% \end{center}
-% \end{itemize}
-
-% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
-% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
-% $\phi_f$ be the GFP of $G$. Following properties ensue:
-% \begin{itemize}
-% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
-% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
-% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
-% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
-% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
-% \end{itemize}
-
-% \begin{theorem}
-% $\phi_f$ is a solution.
-% \end{theorem}
-% \begin{proof}
-% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
-% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
-% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
-% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
-% solution for $\varphi$.
-% \end{proof}
-
-% \begin{theorem}
-% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
-% \phisol}{\phi_f}$
-% \end{theorem}
-% \begin{proof}
-%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
-%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
-%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
-%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$
-
-% \end{proof}
 

### 2016-06-10 paper/type-inference.tex
type inference

@@
 
-This iterative process is a standard fixed point computation process.
-It is possible to encode this computation using a set of Datalog rules,
+\paragraph{Algorithmic Aspects.}
+The iterative process described above is a standard fixed point computation.
+We can encode this computation using a set of Datalog rules,
 and compute the solution using any Datalog engine.
@@
 Further details omitted.
+
+\paragraph{Modularity Aspects.}
+The type inference algorithm, as presented, traverses the entire program to
+generate the set of constraints, which are solved en masse, using an iterative
+fixed point computation. However, the type inference can be realized in a
+modular and compositional fashion, subject only to the restrictions imposed
+by recursion.
 

### 2016-06-13 paper/type-inference.tex
type inference

@@
 
+In the elaboration phase, we can process a class \C{C} only after any class
+\C{B} that \C{C} depends on has been processed: class \C{C} depends on
+class \C{B} if \C{B} is either \C{C}'s base class or the type of any field
+of \C{C} depends on \C{B}. In effect, this means that any collection of
+mutually recursive classes must be processed together. Non-recursive
+dependences can be handled in a compositional fashion: if class \C{C}
+depends on \C{B} non-recursively, then the elaboration can be done for
+\C{B} first, and then \C{C} can be processed.
+
+The same idea applies to the constraint-solving phase as well.
+Given a set of constraints, we say that a predicate variable $\varphi_1$
+\emph{directly-depends} on another predicate variable $\varphi_2$ if the set of
+constraints includes a constraint $\isvalid{\varphi_1 \conj \phictxt}{F(\varphi_2)}$.
+We say that $\varphi_1$ \emph{depends} on $\varphi_2$ if $\varphi_1$ transitively
+depends on $\varphi_2$.
+The constraint solver needs to process any collection of mutually dependent
+predicate variables together.
+In effect, this requires the type inference to process any collection of
+mutually recursive methods together.
+However, methods that are not mutually recursive can be processed separately.
+
 %% \emph{Class Constraints}. 
@@
 %% \end{itemize}
+
+\subsection{Elaboration and Constraint Generation}
+
+\input{fb-elabexpr}
+\input{fb-elabmeth}
+
+%Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
+%(a). replacing core types in variable declarations and \C{new}
+%expressions with fresh region type templates, and (b). explicitly
+%instantiating region parameters of methods with fresh region variables
+%in method calls and function applications. This elaboration is
+%performed with respect to the polymorphic type templates of classes
+%and methods computed as per \S\ref{sec:fb-templatization}.
+
+We now present an integrated algorithm that processes all the methods
+in a given program, elaborating them (by introducing region variables
+to stand for actual region parameters) and simultaneously generating
+the set of constraints that must be solved.
+
+Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+this elaboration for a subset of expressions in $\absof{\FB}$, whose
+corresponding $\FB$ expressions have been ascribed static semantics in
+Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
+context as the expression typing judgment in
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
+retaining their meaning. The function traverses expressions in a
+syntax-directed manner of a type checker, introducing fresh region
+type templates for unknown region types, while generating constraints
+over region and predicate variables.
+Note that $\elabExpr$ returns the type of the subexpression,
+which is used to generate constraints for the expression. Functions
+$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
+implement type well-formedness and subtype judgments from
+Fig.~\ref{fig:fb-staticsem}, respectively.
+
+Functions $\elabMeth$ and $\elabClass$ shown in
+Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+class definitions, respectively. Both functions first build a context
+($\A$) containing a set ($ \rhoenv$) of region variables denoting
+regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ denoting the fact that such constraints are yet
+to be inferred.
+
+Function $\elabMeth$ elaborates a method definition of class $B$. It
+calls $\elabExpr$ with the context $\A$, its allocation context
+parameter ($\rhoallocm$), and a type environment ($\env$) that
+contains region type bindings for all the arguments of the method,
+including the implicit $\C{this}$ argument. The type 
+returned by $\elabExpr$ for the method body is checked against its
+expected type (derived from the type of the method)
+generating more constraints. The function then returns the elaborated
+method definition and the set of constraints.
+
+$\elabClass$ elaborates the definition of a class $B$. It relies on
+$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
+hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
+constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+the set of constraints returned by these functions, $\elabClass$ adds
+constraints generated by checking the well-formedness of the type
+templates of its superclass and fields, and also a new constraint
+capturing a couple of safety conditions: first, the allocation regions
+of objects referred by the instance variables should outlive the
+allocation region of the instance itself, and second, the allocation
+regions of a class type and its superclass type must be the same.
+
+Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
+every definition in the class table $CT$, while accumulating
+constraints.
 
@@
 % the last constraint is generated from the recursive call.
-
-\subsection{Modular Constraint Solving}
-\label{sec:fb-constraintsolving}
-
-Our constraint generation algorithm traverses the entire program,
-performing elaboration and collecting constraints, which are
-subsequently solved en masse. The motivation behind the whole-program
-approach to constraint generation is twofold: it simplifies
-elaboration functions and makes presentation easier, and second, it
-naturally generalizes to mutual recursion. Nonetheless, we do not
-intend our type inference to be a whole-program analysis for (a). it
-preempts opportunities for separate compilation and dynamic linking,
-and (b). it is expensive and an overkill in most practical cases. We
-therefore reclaim the compositionality of type inference by solving
-the constraints in a compositional fashion. In more practical terms
-this means that our constraint solving algorithm visits and solves
-every constraint (or, every set of mutually dependent constraints)
-only once. It composes computed solutions to solve other constraints
-that depend on the solved constraints. Importantly, the failure to
-solve a dependent constraint does not result in backtracking. We now
-describe our compositional constraint solving algorithm in detail. To
-simplify the presentation of our algorithm, we assume that are no
-mutually recursive definitions in the source program. Recursive
-definitions are nonetheless allowed.
-
-\paragraph{Terminology} In a validity constraint, a predicate variable
-occuring on the left side of the turnstile is said to occur
-negatively, or with \emph{negative polarity}. In contrast, a predicate
-variable occuring on the right side is said to occur positively, or
-with \emph{positive polarity}. A validity constraint
-\emph{constrains} the set of predicate variables that occur
-negatively in the constraint, while it \emph{uses} the set of
-predicate variables that occur positively. A constraint is said to be
-\emph{recursive} if it constrains and uses a predicate variable.
-% A pair of validity
-% constraints are said to be \emph{mutually dependent} if there exists a
-% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
-% with opposite polarities in both the constraints. In such case, the
-% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
-% be mutually dependent. We call the transitive closure of mutual
-% dependency relation as \emph{transitive dependency} relation. Like
-% mutual dependency, transitive dependency is also extended to predicate
-% variables.
-
-Given a set of validity constraints, we first build a dependency graph
-($G_c$) with constraints as nodes, and dependencies between them
-captured as edges. There exists an edge from a constraint $c_2$ to a
-constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
-$c_1$ constrains a predicate variable that $c_2$ uses.
-
-The above condition intuitively corresponds to a case, where
-expression or type, whose region elaboration is constrained by $c_2$
-refers to a method or a class, whose unknown precondition is
-constrained by $c_1$. The common predicate variable ($\varphi$)
-represents the unknown precondition in this case. The dependency from
-$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
-before $c_2$ is solved, thus enforcing the rule that the precondition
-of a method must not depend on its calling context. 
-% Note that the first condition results in self-loops over recursive
-% constraints in the dependency graph.  Since mutually recursive
-% definitions refer each other, the first condition also results in
-% bidirectional dependencies between mutually recursive constraints
-% (and self-loops over recursive constraints).
-
-Next, we convert the dependency graph over constraints into a
-dependency DAG ($G_C$) over sets of constraints, where each set
-represents a strongly connected component in the dependency graph.
-
-%% \paragraph{Example} The dependency DAG ($G_C$) over validity
-%% constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
-%% shown below: %in Fig.~\ref{fig:pair-deps}. 
-%% \vspace*{-0.08in}
-%% \begin{figure}[H]
-%% \includegraphics[scale=0.6]{DepDAG.png}
-%% \end{figure}
-%% \vspace*{-0.08in}
-%% Each node (labeled $C_i$) is a set of constraints that belong to a
-%% strongly connected component in the dependency graph ($G_c$), hence
-%% are mutually dependent. All dependencies, except the self-dependency
-%% on $\mathbf{c_{10}}$, are type-2 dependencies.
-
-A dependency DAG makes the dependencies between constraints explicit.
-Constraints in each set are mutually dependent, and need to be solved
-simultaneously, whereas constraints in different sets can be solved as
-per any valid topological ordering of the graph's transpose.
-Accordingly, we obtain a topological ordering of nodes in the graph
-$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
-in that order. The solutions obtained after solving a constraint set
-are applied to the constraints in subsequent sets before attempting to
-solve them.  Consequently, when the turn of a constraint set ($C$)
-arrives during the constraint solving process, it satisfies certain
-properties:
-\begin{itemize}
-\item There exists only one predicate variable ($\varphi$) that is either
-constrained or used by the constraints in the set ($C$). The variable is
-called set's \emph{subject}. This property follows from (a). the fact
-that all the dependency constraints have already been solved (and
-solutions applied), and (b). the assumption that there are no mutually
-recursive definitions. 
-\item All the constraints that constrain the set's subject are present
-in the set. This follows from our definition of the dependency relation.
-\end{itemize}
-
-%% For the DAG in figure above, we consider the topological order $[C_1,
-%% C_2, C_3]$ of its transpose, and solve the sets of constraints in that
-%% order.
-
-\subsection{Expression Elaboration}
-
-\input{fb-elabexpr}
-\input{fb-elabmeth}
-
-%Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
-%(a). replacing core types in variable declarations and \C{new}
-%expressions with fresh region type templates, and (b). explicitly
-%instantiating region parameters of methods with fresh region variables
-%in method calls and function applications. This elaboration is
-%performed with respect to the polymorphic type templates of classes
-%and methods computed as per \S\ref{sec:fb-templatization}.
-
-Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-this elaboration for a subset of expressions in $\absof{\FB}$, whose
-corresponding $\FB$ expressions have been ascribed static semantics in
-Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
-context as the expression typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning. The function traverses expressions in a
-syntax-directed manner of a type checker, introducing fresh region
-type templates for unknown region types, while generating constraints
-over region and predicate variables. The precise nature of generated
-constraints is explained in \S\ref{sec:fb-constraintsem}, but in
-summary, they capture the relationships between the type templates of
-various subexpressions and their well-formedness. Note that
-$\elabExpr$ returns the region type template of the subexpression,
-which is used to generate constraints for the expression. Functions
-$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
-implement type well-formedness and subtype judgments from
-Fig.~\ref{fig:fb-staticsem}, respectively.
-
-\subsection{Method and Class Elaboration}
-
-Functions $\elabMeth$ and $\elabClass$ shown in
-Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
-class definitions, respectively. Both functions first build a context
-($\A$) containing a set ($ \rhoenv$) of region variables denoting
-regions that are currently live, a map ($\aenv$) mapping type
-variables to their bounds, and a constraint formula ($\phicx$)
-capturing constraints over live region variables. We use predicate
-variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
-
-Function $\elabMeth$ elaborates a method definition of class $B$. It
-calls $\elabExpr$ with the context $\A$, its allocation context
-parameter ($\rhoallocm$), and a type environment ($\env$) that
-contains region type bindings for all the arguments of the method,
-including the implicit $\C{this}$ argument. The region type template
-returned by $\elabExpr$ for the method body is checked against its
-expected type (derived from the type template of the method)
-generating more constraints. The function then returns the elaborated
-method definition and the set of constraints.
-
-$\elabClass$ elaborates the definition of a class $B$. It relies on
-$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
-hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
-constructor ($k$) and method definitions ($\bar{d}$), respectively. To
-the set of constraints returned by these functions, $\elabClass$ adds
-constraints generated by checking the well-formedness of the type
-templates of its superclass and fields, and also a new constraint
-capturing a couple of safety conditions: first, the allocation regions
-of objects referred by the instance variables should outlive the
-allocation region of the instance itself, and second, the allocation
-regions of a class type and its superclass type must be the same.
-
-Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
-every definition in the class table $CT$, while accumulating
-constraints.

### 2016-06-15 paper/type-inference.tex
type inference

@@
 
-\input{fb-constraint-generation}
-
 The constraint generation algorithm mimics the static type checker, but accumulates
@@
 
+\input{fb-constraint-generation}
+
 \paragraph{Constraint Generation.}
+The constraint generation algorithm is a direct adaption of the type checker:
+each type checking rule is modified to produce a set of constraints that must hold
+for the type checker to succeed.
+%
+Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
 
@@
 
-\subsection{Elaboration and Constraint Generation}
+\subsection{Elaboration and Constraint Generation: Details}
 

### 2016-06-21 paper/type-inference.tex
type inference

@@
 as before. We then do parameter instantiation for all recursive fields,
-such that their region annotations (the actial region parameters) are
+such that their region annotations (the actual region parameters) are
 exactly the same as the (formal) region parameters of the class.
@@
 
-Our constraints also make uses of \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}} $F$.
+Our constraints also make uses of \emph{pending substitutions} $F$:
+% \footnote{we borrowed this terminology from~\cite{ltpldi08}}
 A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
@@
 $\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only the formula obtained by substituting $\rgn_1$
+$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
 and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
@@
 
+\paragraph{Soundness and Completeness.}
+
+\textbf{TO DO:}
+The entire algorithm is sound.
+The second phase is complete as well: if there is any possible instantiation (of the
+region variables and predicate variables) of the elaborated program produced by the first
+phase that type checks, then the second phase will identify it.
+Incompleteness exists only in a few choices made during the first phase (elaboration phase).
+
 \subsection{Elaboration and Constraint Generation: Details}

### 2016-06-21 paper/abstract.tex
updated abstract

@@
 significant performance bottleneck.  A class of these systems, focused
-on big-data, create many and often large data structures with
+on big-data, create many large data structures with
 well-defined lifetimes.  In this paper, we present a language and a
@@
 C\# programs in the presence of transferable regions. We complement
-our type system with a type inference algorithm that infers principal
-region types for first-order programs, and practically useful types
-for higher-order programs. This eliminates the need for programmers to
+our type system with a type inference algorithm that infers 
+region types for higher-order programs. This eliminates the need for programmers to
 write region annotations on types, while facilitating the reuse of

### 2016-06-22 paper/type-inference.tex
type inference

@@
 A constraint satisfaction problem is a tuple
-$(\bar{\pi}, \bar{\rho}, \bar{\varphi}, \regionDeltaMap, \predDeltaMap, C)$,
-where $\bar{\pi}$ is a set of region constants,
-$\bar{\rho}$ is a set of region variables,
-$\bar{\varphi}$ is a set of predicate variables,
-$\regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
+$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
+where $\regionConstants$ is a set of region constants,
+$\regionVars$ is a set of region variables,
+$\predVars$ is a set of predicate variables,
+$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
 specifies a unification domain for each region variable,
-$\predDeltaMap : \bar{\varphi} \rightarrow 2^{\bar{\pi}}$ is a function that
+$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
 specifies the domain for each predicate variable,
-and $C$ is a set of validity constraints, where the $i$'th validity constraint
-assumes one of the following forms:
+and $\constraintSet$ is a set of validity constraints in one of the following forms:
 \begin{center}
 \(
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {\phicstr^{i}}\qquad
-    \isvalid{\phictxt^{i} \conj \varphi}
-            {F_i(\varphi)}
+    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
+    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
 \)
@@
 The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
-where $\regionSubstFn$ is a map from $\rhobar$ to $\bar{\pi}$
-and $\predSubstFn$ is a map from $\bar{\varphi}$ to a region-constraint formula such that
+where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
+and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
 \begin{itemize}
-  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \bar{\rho}$,
+  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
 
   \item $\predSubstFn$ is well-formed under $\predDeltaMap$
-    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \bar{\varphi}$).
+    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
 

### 2016-06-22 paper/type-inference.tex
type inference

@@
   step 1. We apply these substitutions to the elaborated program to produce the final program.
-
 \end{enumerate}
@@
 
-\begin{theorem}
-\emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
-(1) $\absof{t} = p$, and
-(2) $t$ is well-typed.
-\end{theorem}
-
-%%    This is used to replace every FGJ type in the program
-%%    compute a polymorphic region type template
-%%     for every class and method from their FGJ types. The templates contain
-%%     region variables ($\rho$) to denote unknown region annotations, and
-%%     predicate variables ($\varphi$) to denote unknown constraints over
-%%     such region annotations. Free region variables are generalized in these
-%%     types (which are, hence, polymorphic).
-%%  
-%%    Second, we make use of the computed region type templates to elaborate
-%% expressions by introducing region variables to denote unknown region
-%% arguments in \C{new} expressions, method calls and function
-%% applications.
-%% While elaborating expressions, we also build a system of
-%% constraints that capture well-formedness requirements and subtyping
-%% relationships between type templates that must hold (as per the static
-%% semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
-%% Third, we lift expression elaboration and constraint generation to
-%% methods and classes. Finally, we solve the constraints by making use
-%% of our fixpoint constraint solving algorithm \csolvestar, which
-%% reduces the constraint solving problem to an abduction problem in a
-%% Herbrand constraint system, and then relies on \csolve, our abduction
-%% solver for that domain.
-
-% Due to the presence of region-polymorphic
-% recursion in \FB, constraints generated by the algorithm can be
-% circular. More precisely, constraints generated can assume the form
-% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
-% non-idempotent substitution function for region variables in
-% $\varphi$. In the constraint solving phase, the algorithm then relies
-% on a fixpoint constraint solving algorithm called \csolvestar to solve
-% the constraints and determine assignments for unknown region and
-% predicate variables.
+We later on establish the soundness of the type inference algorithm, and also establish
+partial completeness results.
 
@@
 For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_0,\rho_1} \rightarrow \C{Object}\inang{\rho_2}$.
+$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.
 
@@
 
-Recall that \emph{region identifiers} are used in four different roles.
-A region identifier may serve as
-(a) A \emph{formal region parameter} of a class or method, or
+\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
+A \emph{region variable} is introduced to represent an unknown actual region parameter
+of a method invocation or object alloation, which will be bound to a region constant by the end of the
+type inference.
+A region constant may be 
+(a) a \emph{formal region parameter} of a class or method, or
 (b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
-(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct, or
-(d) a \emph{region variable}, introduced to represent an unknown actual region parameter
-of a method invocation or object alloation,
-which will be bound to a region identifier of one of the three
-preceding kinds by the end of the type inference.
+(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
 
@@
 %
-The first property is that every region variable is forced to be unified with some
-region constant by the constraints, except in one special case.
+The first property is that every region variable is guaranteed to be unified with some
+region constant by the constraints\footnote{
+There is one special case where this does not hold: a recursive function
+that calls itself in a non-terminating fashion. Such a function never
+returns a value, and so the return-value can be typed as anything.
+We assume that the return value of such a function is typed to be \C{unit},
+which will not introduce any region variables at the call-site of such a function.}
 This follows because $\FB$ does not admit uninitialized variables.
-(The only case where this does not hold is in the case of a recursive function
-that calls itself in a non-terminating fashion: in this case, the function never
-returns a value, and so the return-value can be typed as anything. In the sequel,
-we assume that the return value of such a function is typed to be \C{unit}.)
 % , and the type system uniquely determines the region parameters of every value
@@
 
-\textbf{TO DO:}
-The entire algorithm is sound.
-The second phase is complete as well: if there is any possible instantiation (of the
-region variables and predicate variables) of the elaborated program produced by the first
-phase that type checks, then the second phase will identify it.
-Incompleteness exists only in a few choices made during the first phase (elaboration phase).
+The type inference algorithm is sound:
+
+\begin{theorem}
+\emph{(\textbf{Soundness})}
+For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
+(1) $\absof{t} = p$, and
+(2) $t$ is well-typed.
+\end{theorem}
+
+Furthermore, the second phase of the type inference algorithm is complete as well.
+If there is any possible instantiation (of the region variables and predicate variables) of
+the elaborated program produced by the first phase that type checks, then the second phase
+will identify it.
+
+Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
+explained below.
+
+\begin{enumerate}
+\item Our technique for determining the set of region parameters for a recursive class
+incorporates a specific heuristic, namely that the recursive occurrence of the class has
+the same parameters, in the same order, as the class itself. This technique will fail, for
+example. if the program uses a recursive list type whose elements alternatively come
+from two different regions.
+% , $R_1$ and $R_2$.
+% both of which are region parameters of the class.
+
+\item Our technique for region parametrization also uses a heuristic in the case of
+higher order programs. Consider a higher-order method that has a function $f$ as a
+parameter. We have a choice in terms of where the region-parameters introduced for
+$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
+of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
+source of incompleteness in the type-inference.
+
+The treatment of class fields with a function type also uses the same (incomplete)
+heuristic.
+
+\end{enumerate}
+
+Region annotations provided by the user can help the type-inference overcome these
+sources of incompleteness.
 

### 2016-06-23 paper/type-inference.tex
incompleteness
removed constructor

@@
 
-\subsection{Constraints}
+\subsection{The Constraint Satisfaction Problem (CSP)}
 \label{sec:fb-constraintsem}
@@
 
-\subsection{Solving the CSP}
-
 \paragraph{Properties of generated constraints}
-The set of generated constraints have two interesting properties that allow us to
-solve the constraints efficiently.
+In the sequel, we will restrict our attention to a special class of CSPs that satisfy
+two interesting properties that allow us to solve the constraints efficiently.
+The constraints generated our type inference algorithm satisfy these two properties.
 %
@@
 it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+
+\subsection{Solving the CSP}
 
@@
 example. if the program uses a recursive list type whose elements alternatively come
-from two different regions.
+from two different regions. Such a program would require the following elaboration,
+which is beyond the scope of our approach:
 % , $R_1$ and $R_2$.
 % both of which are region parameters of the class.
+\begin{tabular}{ccc}
+\begin{minipage}{0.3\linewidth}
+\begin{codejava}
+class List
+  $\extends$ $\ObjZ$
+{
+  $\ObjZ$ data;
+  List next;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
+  $\extends$ $\ObjZ \langle \rho_0 \rangle$
+{
+  $\ObjZ \langle \rho_1 \rangle$ data;
+  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
 

### 2016-06-29 paper/abstract.tex
Formatting changes. Figure added in intro.

--- 2015-11-21 paper/abstract.tex
@@
 significant performance bottleneck.  A class of these systems, focused
-on big-data, create many and often large data structures with
-well-defined lifetimes.  In this paper, we present a language and a
-memory management scheme based on user-managed memory regions (called
-\emph{transferable regions}) that allow programmers to exploit
+on big-data, create numerous large data structures with well-defined
+lifetimes.  In this paper, we present a language and a memory
+management scheme based on user-managed memory regions (called
+\emph{transferable regions}) that allow programmers to exploit the
 knowledge of data structures' lifetimes to achieve significant
@@
 the practical utility of our approach.
-
-%As a testament to efficacy of our approach, we demonstrate performance
-%improvements over a range of realistic dataflow programs that were
-%rewritten to use region-based manual memory management, but are
-%nevertheless judged to be safe by our type system.

### 2016-06-29 paper/introduction.tex
Formatting changes. Figure added in intro.

--- 2015-11-21 paper/introduction.tex
@@
 
-Consider the example, from~\cite{Broom:HotOS}, shown in
-Fig.~\ref{fig:motivating-eg}.  This code represents the logic for a
-streaming query operator.  The operator receives a stream of input
-messages, each associated with a time (window) $t$, processed by
-method \texttt{onReceive}.  Each input message contains a list of
-inputs, each of which is processed by applying a user-defined function
-to create a corresponding output.  The operator may receive multiple
+Consider the example shown in Fig.~\ref{fig:motivating-eg}.  This code
+represents the logic for a \C{SELECT} query operator for streaming
+data.  The operator receives a stream of input messages, each
+associated with a time (window) $t$, processed by method
+\texttt{onReceive}.  Each input message contains a list of inputs,
+each of which is processed by applying a user-defined function to
+create a corresponding output.  The operator may receive multiple
 messages with the same timestamp (and messages with different
@@
 
-This example is an instance of a general pattern, where a producer creates a
-data structure and passes it to a consumer. In a system where most of the
-computation takes this form, and these data structures are very large, as is
-the case with many streaming big-data analysis systems, garbage collection
-overhead becomes significant~\cite{Broom:HotOS}.  Furthermore, in a
-distributed dataflow system, the GC pause at one node can have a cascading
-adverse effect on the performance of other nodes, particularly when real-time
-streaming performance is required~\cite{Broom:HotOS, harris15}.  In
-particular, a GC pause at an upstream actor can block downstream actors that
-are waiting for messages.  However, often much of the GC overhead results
-from the collector performing avoidable or unproductive work.  For example,
-in the process executing the code from Fig.~\ref{fig:motivating-eg}, GC might
-repeatedly traverse the \C{map} data structure, although its objects cannot
-be collected until a suitable timing message arrives.
-
-% actors.
-% performance degradation due to
-% GC overhead becomes aggravated in the distributed 
-% For example, a consumer process may trigger a
-% GC as soon as it receives a large data structure from a producer
-% process, which has just finished its GC. In such case, consumer's GC
-% traverses the entire data structure performing little or no work.
-
-% as the collector scans these large data structures For instance, in
-% the above example, the list at \C{map[t]} cannot be collected until
-% the the timing message arrives, and the list is transferred.
-% \C{transfer} operation is complete.  using alternative memory
-% management techniques.
+This example is an instance of a general pattern, where a producer
+creates a data structure and passes it to a consumer. In a system
+where most of the computation takes this form, and these data
+structures are very large, as is the case with many streaming big-data
+analysis systems, garbage collection overhead becomes
+significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
+dataflow system, the GC pause at one node can have a cascading adverse
+effect on the performance of other nodes, particularly when real-time
+streaming performance is required.  In particular, a GC pause at an
+upstream actor can block downstream actors that are waiting for
+messages.  However, often much of the GC overhead results from the
+collector performing avoidable or unproductive work.  For example, in
+the process executing the code from Fig.~\ref{fig:motivating-eg}, GC
+might repeatedly traverse the \C{map} data structure, although its
+objects cannot be collected until a suitable timing message arrives.
 
@@
 while they are deallocated en masse when the region is freed.  Thus, a
-region is a good fit for a set of fate-sharing objects.
-% that is used to realize a data structure with a well-defined
-% lifetime. 
-In Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
+region is a good fit for a set of fate-sharing objects.  In
+Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
 time window $t$ (i.e., \C{map[t]}) can be a separate region that is
@@
 suffers from the usual drawbacks, namely the potential for invalid
-references and the consequent lack of memory safety. Automatic region-based memory
-management systems guarantee memory safety, but impose various
-restrictions.  MLKit, which implements the approach pioneered by Tofte
-and Talpin~\cite{tofte94,tofte97}, for example, uses lexically scoped
-regions.  At runtime, the set of all regions (existing at a point in
-time) forms a stack. Thus, the lifetimes of all regions must be
-well-nested: it is not possible to have two regions whose lifetimes
-overlap, with neither one's lifetime contained within the other.
-Unfortunately, the data structures in the above example do not satisfy
-this restriction (as the output messages for multiple time windows may
-be simultaneously live, without any containment relation between their
-lifetimes).  We refer to regions with lexically scoped lifetimes as
-\emph{stack regions} and to regions that do not have such a lexically
-scoped region as \emph{dynamic} regions.
+references and the consequent lack of memory safety. Automatic
+region-based memory management systems guarantee memory safety, but
+impose various restrictions.  MLKit, which implements the approach
+pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, for example,
+uses lexically scoped regions.  At runtime, the set of all regions
+(existing at a point in time) forms a stack. Thus, the lifetimes of
+all regions must be well-nested: it is not possible to have two
+regions whose lifetimes overlap, with neither one's lifetime contained
+within the other.  Unfortunately, the data structures in the above
+example do not satisfy this restriction (as the output messages for
+multiple time windows may be simultaneously live, without any
+containment relation between their lifetimes).  We refer to regions
+with lexically scoped lifetimes as \emph{stack regions} and to regions
+that do not have such a lexically scoped region as \emph{dynamic}
+regions.
 
@@
 that traversing its objects in heap, and transferring them
-independently\footnote{Empirical studies in~\cite{Broom:HotOS} support
-this claim}. In the \C{SelectVertex} example from
+independently. In the \C{SelectVertex} example from
 Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
@@
 region once the transfer is complete.
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  Before transfer.
+  \label{fig:after-transfer}
+} [
+  0.35\columnwidth
+] {
+  \includegraphics[scale=0.36]{producer-heap}
+}
+\subcaptionbox {
+  After transfer.
+  \label{fig:before-transfer}
+} {
+  \includegraphics[scale=0.36]{producer-consumer-heap}
+}
+
+\caption{References in and out of a transferable region ($R$) become
+invalid after transfer.}
+\label{fig:transfer-violations}
+\end{figure}
 
@@
 objects outside its (transferable) region, then such references become
-invalid in the context of the consumer. Safety violations of this kind
-are particularly unwelcome as the program with GC did not have them to
-begin with. Note that, while the references of later kind (i.e.,
-references that escape a region) defeat the very purpose of a
-transferable region, hence need to be prohibited, references of the
-former kind (i.e., references into a region from outside) are not at
-odds with the concept of a transferable region, hence need to be
-permitted. In fact, allowing such references is crucial to
-performance, as any non-trivial program creates temporary objects, and
-it is undesirable to allocate them in a transferable region; such
-regions are meant for the data being transferred. Since transferable
-regions are first class objects in our setting, controlling references
-to and from such regions while ensuring their safety without
-significantly diluting the performance advantage of regions over GC is
-a challenging exercise.
+invalid in the context of the consumer. Both scenarios are depicted in
+Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
+particularly unwelcome in the context of languages like C\# and Java
+that are otherwise memory-safe. While references that escape a
+transferable region can be disallowed in the interest of safety,
+references into a transferable region need to be allowed while the
+region is not yet transferred. In fact, allowing such references is
+crucial to performance, as any non-trivial program creates temporary
+objects, and it is undesirable to allocate them in a transferable
+region; such regions are meant for the data being transferred. Since
+transferable regions are first class objects in our setting,
+controlling references to and from such regions while ensuring their
+safety without significantly diluting the performance advantage of
+regions over GC is a challenging exercise.
 
@@
 won't be transferred/freed while it is open. Our observation is that
-by nesting a~\cite{tofte94}-style \C{letregion} lexical block, that
-delimits the lifetime of a stack region, inside an \C{open} lexical
-block for a transferable region, we can guarantee that the
-transferable region will remain live as long as the stack region is
-live. We say that the former ``outlives'' the later\footnote{We borrow
-the \emph{outlives} relation from~\cite{MIT03}. A comparison of our
-approach with~\cite{MIT03} can be found in
-\S~\ref{sec:related-work}.}, and any references from the stack region
-to the transferable region are therefore safe. 
-% In general, we can allow references from a region
-% $R_0$ to another region $R_1$, provided that we establish that $R_1$ outlives
-% $R_0$. 
-Next, we note that by controlling the ``outlives'' relationships
-between various regions, we only allow safe cross-region references,
-while prohibiting unsafe ones. In the above example, an outlives
-relationship \emph{from} the stack region \emph{to} the transferable
-region means that the references in that direction are allowed, but
-not the references in the opposite direction. In contrast, if an
-\C{open} block of a transferable region $R_0$ is nested inside an
-\C{open} block of another transferable region $R_1$, we do not
-establish any outlives relationships, thus declaring our intention to
-not allow any cross-region references between $R_0$ and $R_1$.
-Finally, we observe that outlives relationships are established based
-on the lexical structure of the program, hence a static type system
-can enforce them effectively. By assigning region types to objects,
-which capture the regions such objects are allocated in, and by
-maintaining outlives relationships between various regions, we can
-statically decide the safety of all references in the program.
+by nesting an MLKit-style \C{letregion} lexical block, that delimits
+the lifetime of a stack region, inside an \C{open} lexical block for a
+transferable region, we can guarantee that the transferable region
+will remain live as long as the stack region is live. We say that the
+former \emph{outlives} the later., and any references from the stack
+region to the transferable region are therefore safe.  Next, we note
+that by controlling the outlives relationships between various
+regions, we only allow safe cross-region references, while prohibiting
+unsafe ones. In the above example, an outlives relationship
+\emph{from} the stack region \emph{to} the transferable region means
+that the references in that direction are allowed, but not the
+references in the opposite direction. In contrast, if an \C{open}
+block of a transferable region $R_0$ is nested inside an \C{open}
+block of another transferable region $R_1$, we do not establish any
+outlives relationships, thus declaring our intention to not allow any
+cross-region references between $R_0$ and $R_1$.  Finally, we observe
+that outlives relationships are established based on the lexical
+structure of the program, hence a static type system can enforce them
+effectively. By assigning region types to objects, which capture the
+regions such objects are allocated in, and by maintaining outlives
+relationships between various regions, we can statically decide the
+safety of all references in the program.
 
-Indeed, the utility of our approach described above is predicated
-on the assumption that we can enforce certain invariants on
-transferable regions. Firstly, a transferable region cannot be
-transferred/freed inside an open block of that region (i.e, while it
-is still open).  Secondly, a transferred/freed region cannot be
-opened. These are typestate invariants on the transferable region
-objects, which are hard to enforce statically due to the presence of
-unrestricted aliasing. Techniques like linear types and unique
-pointers can be used to restrict aliasing, but the constraints they
-impose are often hard to program around. We therefore enforce
-typestate invariants at runtime via lightweight checks. In
-particular, we define an acceptable state transition discipline for
-transferable regions (Fig.~\ref{fig:region-fsm}), and check, at
-runtime, whether a given transition of a transferable region (\eg,
-from \emph{open} state to \emph{freed} state) is valid or not. The
-check is lightweight since it only involves checking a single tag that
-captures the current state.  We believe that this is a reasonable
-choice since regions are coarse-grained objects manipulated
-infrequently, when compared to the fine-grained objects that are
-present inside these regions, for which safety is enforced statically.
-An added advantage of delegating the enforcement of typestate
-invariants to runtime is that our region type system is
-simple\footnote{Supplement lists all the type rules of our language,
-which occupy less than a page.}, which made it possible to formulate a
-type inference that completely eliminates the need to write region
-type annotations. This, we believe, significantly reduces the
-impediment to adopt our approach in practical setting.
+Indeed, the utility of our approach described above is predicated on
+the assumption that we can enforce certain invariants on transferable
+regions. Firstly, a transferable region cannot be transferred/freed
+inside an open block of that region (i.e, while it is still open).
+Secondly, a transferred/freed region cannot be opened. These are
+typestate invariants on the transferable region objects, which are
+hard to enforce statically due to the presence of unrestricted
+aliasing. Techniques like linear types and unique pointers can be used
+to restrict aliasing, but the constraints they impose are often hard
+to program around. We therefore enforce typestate invariants at
+runtime via lightweight checks. In particular, we define an acceptable
+state transition discipline for transferable regions
+(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
+transition of a transferable region (\eg, from \emph{open} state to
+\emph{freed} state) is valid or not. The check is lightweight since it
+only involves checking a single tag that captures the current state.
+We believe that this is a reasonable choice since regions are
+coarse-grained objects manipulated infrequently, when compared to the
+fine-grained objects that are present inside these regions, for which
+safety is enforced statically.  An added advantage of delegating the
+enforcement of typestate invariants to runtime is that our region type
+system is simple, which made it possible to formulate a type inference
+that completely eliminates the need to write region type annotations.
+This, we believe, significantly reduces the impediment to adopt our
+approach in practical setting.
 

### 2016-06-29 paper/related.tex
Formatting changes. Figure added in intro.

--- 2016-01-31 paper/related.tex
@@
 MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
-existing proposals in a number of ways. Firstly, our problem setting
+existing proposals in the following ways. Firstly, our problem setting
 includes lexically scoped stack regions and dynamic transferable
@@
 % guarantee safety in presence of
-In particular, our approach requires neither restrictive static
-mechanisms (e.g., linear types and unique pointers) nor expensive
-runtime mechanisms (e.g., garbage collection and reference counting)
-in order to guarantee safety. Lastly, our region type system comes
-equipped with full type inference that eliminates the need to write
-region annotations on types to convince the type checker that the
-program is safe.
+In particular, our approach circumvents the need for restrictive
+static mechanisms (e.g., linear types and unique pointers) and
+expensive runtime mechanisms (e.g., garbage collection and reference
+counting) in order to guarantee safety. Lastly, our region type system
+comes equipped with full type inference that eliminates the need to
+write region annotations on types to convince the type checker that
+the program is safe.
 
@@
 iterators in favor of for-loops, annotate region types, or rewrite
-C\#'s standard libraries to use in \name.  Non-intrusiveness is not a
-design consideration for Cyclone, which requires C programmers to use
-new language constructs and abandon some standard programming idioms
-in the interest of preserving safety. For instance, Cyclone
-programmers are required to write region types for functions; the type
-inference is only intraprocedural. Ensuring safety in presence of
-dynamic regions requires using either unique pointers or
+C\#'s standard libraries to use in \name. Cyclone requires C
+programmers to use new language constructs and abandon some standard
+programming idioms in the interest of preserving safety. For instance,
+Cyclone programmers are required to write region types for functions;
+the type inference is only intraprocedural. Ensuring safety in
+presence of dynamic regions requires using either unique pointers or
 reference-counted objects.  Both approaches are intrusive. For
@@
 functions (\C{alias\_refptr} and \C{drop\_refptr}) to create and
-destroy aliases.  Reference count is affected only by these
-functions. An alias going out of scope, for instance, does not
-decrement the ref-count. The requirement to use additional constructs
+destroy aliases.  Reference count is affected only by these functions.
+An alias going out of scope, for instance, does not decrement the
+ref-count. The requirement to use additional constructs
 to manage aliases makes reference counting more-or-less as intrusive

### 2016-06-30 paper/introduction.tex
abstract and intro

@@
 
-Consider the example shown in Fig.~\ref{fig:motivating-eg}.  This code
-represents the logic for a \C{SELECT} query operator for streaming
-data.  The operator receives a stream of input messages, each
+The motivating example in Fig.~\ref{fig:motivating-eg} represents
+a \C{SELECT} query operator for streaming data. 
+The operator receives a stream of input messages, each
 associated with a time (window) $t$, processed by method
@@
 
-This example is an instance of a general pattern, where a producer
-creates a data structure and passes it to a consumer. In a system
-where most of the computation takes this form, and these data
-structures are very large, as is the case with many streaming big-data
+This example (of  a producer creating a data structure and transferring it to a consumer)
+is an instance of a general idiom of a set of concurrent actors interacting via messages
+(representing data-structures).
+Our focus in this paper is a region-based approach to realizing such transferable
+data, motivated by the following concerns.
+
+\emph{Performance}.
+When the transferred data structures are large, as is the case with many streaming big-data
 analysis systems, garbage collection overhead becomes
@@
 dataflow system, the GC pause at one node can have a cascading adverse
-effect on the performance of other nodes, particularly when real-time
-streaming performance is required.  In particular, a GC pause at an
-upstream actor can block downstream actors that are waiting for
+effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
+a GC pause at an upstream actor can block downstream actors that are waiting for
 messages.  However, often much of the GC overhead results from the
-collector performing avoidable or unproductive work.  For example, in
-the process executing the code from Fig.~\ref{fig:motivating-eg}, GC
-might repeatedly traverse the \C{map} data structure, although its
-objects cannot be collected until a suitable timing message arrives.
+collector performing avoidable or unproductive work.  For the example
+in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
+\C{map} data structure, although its objects cannot be collected until a
+suitable timing message arrives.
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  Before transfer.
+  \label{fig:after-transfer}
+} [
+  0.35\columnwidth
+] {
+  \includegraphics[scale=0.36]{producer-heap}
+}
+\subcaptionbox {
+  After transfer.
+  \label{fig:before-transfer}
+} {
+  \includegraphics[scale=0.36]{producer-consumer-heap}
+}
+
+\caption{References in and out of a transferable region ($R$) become
+invalid after transfer.}
+\label{fig:transfer-violations}
+\end{figure}
+
+\emph{Safety}.
+These systems are designed to allow the producer and consumer to
+execute in the same address space or in different address spaces
+(as determined by the compiler and runtime).
+We wish to ensure memory safety in the presence of such transferred data by ensuring
+that there are no invalid references. A transfer operation, with no
+additional checks, may cause memory safety violations, both at the
+producer of the data structure, and its (possibly remote) consumer. At
+the producer, any existing references into the transferred data structure become
+invalid post transfer. If the data structure contains references to
+objects outside (the transferred data), then such references become
+invalid in the context of the consumer. Both scenarios are depicted in
+Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
+particularly unwelcome in the context of languages like C\# and Java
+that are otherwise memory-safe. While references from within the
+transferable data to outside can be disallowed in the interest of safety,
+references into transferable data needs to be allowed before the data
+is transferred. Allowing such references is crucial to performance,
+as any non-trivial program creates temporary references to the internal
+objects of a data-structure.
+
+% , and it is undesirable to allocate them in a transferable
+% region; such regions are meant for the data being transferred.
+
+Since
+transferable regions are first class objects in our setting,
+controlling references to and from such regions while ensuring their
+safety without significantly diluting the performance advantage of
+regions over GC is a challenging exercise.
 
@@
 management discipline. A region is a block of memory that is allocated
-and freed in one shot, consuming constant time. A region may contain
+and freed in one shot, often in constant time. A region may contain
 one or more contiguous range of memory locations, and individual
@@
 region once the transfer is complete.
-
-\begin{figure}
-\centering
-\subcaptionbox {
-  Before transfer.
-  \label{fig:after-transfer}
-} [
-  0.35\columnwidth
-] {
-  \includegraphics[scale=0.36]{producer-heap}
-}
-\subcaptionbox {
-  After transfer.
-  \label{fig:before-transfer}
-} {
-  \includegraphics[scale=0.36]{producer-consumer-heap}
-}
-
-\caption{References in and out of a transferable region ($R$) become
-invalid after transfer.}
-\label{fig:transfer-violations}
-\end{figure}
-
-With respect to memory safety, the key property we wish to ensure is
-that there are no invalid references: i.e., references to objects that
-were deallocated, or simply never existed. Transfer operation, with no
-additional checks, may cause memory safety violations, both at the
-producer of the data structure, and its (possibly remote) consumer. At
-the producer, any existing references into the data structure become
-invalid post transfer. If the data structure contains references to
-objects outside its (transferable) region, then such references become
-invalid in the context of the consumer. Both scenarios are depicted in
-Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
-particularly unwelcome in the context of languages like C\# and Java
-that are otherwise memory-safe. While references that escape a
-transferable region can be disallowed in the interest of safety,
-references into a transferable region need to be allowed while the
-region is not yet transferred. In fact, allowing such references is
-crucial to performance, as any non-trivial program creates temporary
-objects, and it is undesirable to allocate them in a transferable
-region; such regions are meant for the data being transferred. Since
-transferable regions are first class objects in our setting,
-controlling references to and from such regions while ensuring their
-safety without significantly diluting the performance advantage of
-regions over GC is a challenging exercise.
 

### 2016-07-01 paper/introduction.tex
introduction

@@
 
-\emph{Performance}.
+\paragraph{Performance.}
 When the transferred data structures are large, as is the case with many streaming big-data
@@
 
-\emph{Safety}.
+\paragraph{Safety.}
 These systems are designed to allow the producer and consumer to
@@
 
-% , and it is undesirable to allocate them in a transferable
-% region; such regions are meant for the data being transferred.
-
-Since
-transferable regions are first class objects in our setting,
-controlling references to and from such regions while ensuring their
-safety without significantly diluting the performance advantage of
-regions over GC is a challenging exercise.
-
-An important observation, in the context of processes of the kind
-described above, is that the data-structures exchanged between them
-can be partitioned into sets of fate-sharing objects with common
-lifetimes, which makes them good candidates for a region-based memory
-management discipline. A region is a block of memory that is allocated
+\paragraph{Safe Transferable Regions.}
+In this paper, we present a language-based approach to safely and
+efficiently realize such transferable data that uses regions as
+the implementation technique.
+%
+A region is a block of memory that is allocated
 and freed in one shot, often in constant time. A region may contain
@@
 while they are deallocated en masse when the region is freed.  Thus, a
-region is a good fit for a set of fate-sharing objects.  In
+region is a good fit for a transferable data-structure.  In
 Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
@@
 
-Region-based memory management, both manual as well as automatic, has
-been known for a long time. Manual region-based memory management
-suffers from the usual drawbacks, namely the potential for invalid
-references and the consequent lack of memory safety. Automatic
-region-based memory management systems guarantee memory safety, but
-impose various restrictions.  MLKit, which implements the approach
-pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, for example,
-uses lexically scoped regions.  At runtime, the set of all regions
+The use of regions alleviates the performance concern related to
+garbage collection described earlier.
+However, manual region memory management introduces memory safety
+issues such as dangling pointers.
+We present a single type system that simultaneously addresses
+this safety concern, as well as those described above related to
+the use of \emph{transferable data}.
+
+% Region-based memory management, both manual as well as automatic, has
+% been known for a long time. Manual region-based memory management
+% suffers from the usual drawbacks, namely the potential for invalid
+% references and the consequent lack of memory safety. Automatic
+% region-based memory management systems guarantee memory safety, but
+% impose various restrictions.  MLKit, which implements the approach
+Safe region-based memory management using types was
+pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
+use only lexically scoped regions.  At runtime, the set of all regions
 (existing at a point in time) forms a stack. Thus, the lifetimes of
@@
 regions.
+%
+Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
+that can be safely transferred across address spaces. We refer to such dynamic
+regions as \emph{transferable} regions.
 
-The goal of this work is a \emph{memory-safe} region-based memory
-management technique that supports dynamic regions as first-class
-objects. Our focus, in this paper, is on dynamic regions, which can be
-safely transferred across address spaces. We refer to such dynamic
-regions as \emph{transferable} regions. As with allocation and
-deallocation, transferring a memory region is fast, and therefore,
-transferring a data structure contained in a region is more efficient
-that traversing its objects in heap, and transferring them
-independently. In the \C{SelectVertex} example from
-Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
-output for each time window $t$ must be transferable due to the
-\C{transfer} operation on Line 16. As it is the case with
-\C{SelectVertex}, the transferred data is no longer accessed by the
-producer, so the transfer operation in our system deallocates the
-region once the transfer is complete.
-
-In this paper, we describe an approach that restores memory safety in
-presence of transferable regions through a combination of a static
-typing discipline and lightweight runtime checks. The cornerstone of
-our approach is an \C{open} lexical block for transferable regions,
+Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
+that combine linear types and regions to suport dynamic regions.
+Unlike the prior work, we ensure memory safety in the presence of
+transferable regions through a combination of a static
+typing discipline and lightweight runtime checks (in place of linear types).
+The cornerstone of this approach is an \C{open} lexical block for transferable regions,
 that ``opens'' a transferable region and guarantees that the region
-won't be transferred/freed while it is open. Our observation is that
-by nesting an MLKit-style \C{letregion} lexical block, that delimits
+won't be transferred/freed while it is open. By nesting an
+MLKit-style \C{letregion} lexical block, that delimits
 the lifetime of a stack region, inside an \C{open} lexical block for a
@@
 
-Indeed, the utility of our approach described above is predicated on
-the assumption that we can enforce certain invariants on transferable
-regions. Firstly, a transferable region cannot be transferred/freed
+Ensuring memory safety using this approach requires ensuring
+that the use of transferable regions satisfies certain temporal properties.
+Firstly, a transferable region cannot be transferred/freed
 inside an open block of that region (i.e, while it is still open).
@@
 typestate invariants on the transferable region objects, which are
-hard to enforce statically due to the presence of unrestricted
+hard to enforce statically in the presence of unrestricted
 aliasing. Techniques like linear types and unique pointers can be used
@@
   %\S~\ref{sec:conclusion} concludes.
+
+% , and it is undesirable to allocate them in a transferable
+% region; such regions are meant for the data being transferred.
+
+\textbf{TO DO:}
+
+As with allocation and deallocation, transferring a memory region is fast, and therefore,
+transferring a data structure contained in a region is more efficient
+that traversing its objects in heap, and transferring them
+independently.
+In the \C{SelectVertex} example from
+Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
+output for each time window $t$ must be transferable due to the
+\C{transfer} operation on Line 16. As it is the case with
+\C{SelectVertex}, the transferred data is no longer accessed by the
+producer, so the transfer operation in our system deallocates the
+region once the transfer is complete.
+
+Since transferable regions are first class objects in our setting,
+controlling references to and from such regions while ensuring their
+safety without significantly diluting the performance advantage of
+regions over GC is a challenging exercise.
+
+An important observation, in the context of processes of the kind
+described above, is that the data-structures exchanged between them
+can be partitioned into sets of fate-sharing objects with common
+lifetimes, which makes them good candidates for a region-based memory
+management discipline.

### 2016-07-01 paper/introduction.tex
introduction

@@
 
-The motivating example in Fig.~\ref{fig:motivating-eg} represents
-a \C{SELECT} query operator for streaming data. 
+In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
+to \emph{transferring data} between a set of concurrent actors, motivated by the following concerns.
+
+Consider the example (of  a producer creating a data structure and transferring it to a consumer)
+in Fig.~\ref{fig:motivating-eg}. This represents a \C{SELECT} query operator for streaming data. 
 The operator receives a stream of input messages, each
@@
 \input{motivating-eg}
-
-This example (of  a producer creating a data structure and transferring it to a consumer)
-is an instance of a general idiom of a set of concurrent actors interacting via messages
-(representing data-structures).
-Our focus in this paper is a region-based approach to realizing such transferable
-data, motivated by the following concerns.
 
@@
 regions.
-%
+
 Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
@@
 regions as \emph{transferable} regions.
-
+%
 Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
@@
 Unlike the prior work, we ensure memory safety in the presence of
-transferable regions through a combination of a static
-typing discipline and lightweight runtime checks (in place of linear types).
+transferable regions through a \emph{combination of a static
+typing discipline and lightweight runtime checks in place of linear types}.
+
 The cornerstone of this approach is an \C{open} lexical block for transferable regions,
@@
 that the use of transferable regions satisfies certain temporal properties.
-Firstly, a transferable region cannot be transferred/freed
+Firstly, a transferable region should not be transferred/freed
 inside an open block of that region (i.e, while it is still open).
-Secondly, a transferred/freed region cannot be opened. These are
+Secondly, a transferred/freed region should not be opened. These are
 typestate invariants on the transferable region objects, which are
@@
 fine-grained objects that are present inside these regions, for which
-safety is enforced statically.  An added advantage of delegating the
-enforcement of typestate invariants to runtime is that our region type
-system is simple, which made it possible to formulate a type inference
-that completely eliminates the need to write region type annotations.
+safety is enforced statically.
+
+One of the novel contributions of this paper is a type inference algorithm
+that eliminates the need for users to write region type annotations.
 This, we believe, significantly reduces the impediment to adopt our

### 2016-07-01 paper/introduction.tex
introduction

@@
 The use of regions alleviates the performance concern related to
-garbage collection described earlier.
+garbage collection described earlier. (It also enables more efficient
+serialization/deserialization of data-structures across address spaces,
+which is also a significant performance bottleneck in such systems.)
 However, manual region memory management introduces memory safety
@@
 
+Formally, our type system introduces \emph{region parameters} (for both
+classes and methods) and uses constrained parametric polymorphism
+over these parameters, where the constraints capture outlives constraints
+between the region parameters.
+
 Ensuring memory safety using this approach requires ensuring
@@
 
-One of the novel contributions of this paper is a type inference algorithm
+One of the novel contributions of this paper is a sound type inference algorithm
 that eliminates the need for users to write region type annotations.
@@
 approach in practical setting.
+Our inference algorithm proceeds in two stages: in the first stage,
+we introduce region parameters (for classes and methods), and in
+the second stage we infer the outlives constraints between these
+parameters (using an interprocedural fixed point computation).
+% While the first
 
@@
 
-\textbf{TO DO:}
-
-As with allocation and deallocation, transferring a memory region is fast, and therefore,
-transferring a data structure contained in a region is more efficient
-that traversing its objects in heap, and transferring them
-independently.
-In the \C{SelectVertex} example from
-Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
-output for each time window $t$ must be transferable due to the
-\C{transfer} operation on Line 16. As it is the case with
-\C{SelectVertex}, the transferred data is no longer accessed by the
-producer, so the transfer operation in our system deallocates the
-region once the transfer is complete.
-
-Since transferable regions are first class objects in our setting,
-controlling references to and from such regions while ensuring their
-safety without significantly diluting the performance advantage of
-regions over GC is a challenging exercise.
-
-An important observation, in the context of processes of the kind
-described above, is that the data-structures exchanged between them
-can be partitioned into sets of fate-sharing objects with common
-lifetimes, which makes them good candidates for a region-based memory
-management discipline.
+%As with allocation and deallocation, transferring a memory region is fast, and therefore,
+%transferring a data structure contained in a region is more efficient
+%that traversing its objects in heap, and transferring them
+%independently.
+%In the \C{SelectVertex} example from
+%Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
+%output for each time window $t$ must be transferable due to the
+%\C{transfer} operation on Line 16. As it is the case with
+%\C{SelectVertex}, the transferred data is no longer accessed by the
+%producer, so the transfer operation in our system deallocates the
+%region once the transfer is complete.
+%
+%Since transferable regions are first class objects in our setting,
+%controlling references to and from such regions while ensuring their
+%safety without significantly diluting the performance advantage of
+%regions over GC is a challenging exercise.
+%
+%An important observation, in the context of processes of the kind
+%described above, is that the data-structures exchanged between them
+%can be partitioned into sets of fate-sharing objects with common
+%lifetimes, which makes them good candidates for a region-based memory
+%management discipline.

### 2016-07-01 paper/broom.tex
Added high-level outline of type-inference
type inference
Formatting changes. Figure added in intro.
integrated new type inference section

--- 2015-11-13 paper/broom.tex
@@
 
-\documentclass[10pt,pldi]{sigplanconf}
+\documentclass[preprint,numbers]{sigplanconf}
 
@@
 \usepackage{breakurl}
-\usepackage[breaklinks]{hyperref}
+\usepackage[breaklinks,draft=false,backref]{hyperref}
+\hypersetup{
+     colorlinks   = true,
+     citecolor    = blue,
+     linkcolor    = blue,
+     urlcolor    = blue
+}
 \usepackage[multiple]{footmisc}
 
-%Formatting Commands
-%-------------------
-\newcommand{\C}[1]{\code{#1}}
-\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
-\newcommand{\code}[1]{\,{\tt #1}\,}
-\newcommand{\conj}{\wedge}
-\newcommand{\disj}{\vee}
-\newcommand{\seqn}[1]{\overline {#1}}
-\newcommand{\tuplee}[1]{\langle #1 \rangle}
-\newcommand{\ALT}{~\mid~}
-\newcommand{\lamb}[2]{\lambda #1.\,#2}
-\newcommand{\spc}[0]{\quad}
-\newcommand{\U}[0]{\cup}
-\newcommand{\X}[0]{\times}
-\newcommand{\xn}[0]{\cap}
-\newcommand{\redsto}[0]{\longrightarrow}
-\newcommand{\rredsto}[0]{\rightsquigarrow_r }
-\newcommand{\evalsto}[0]{\longrightarrow }
-\newcommand{\env}[0]{\Gamma}
-\newcommand{\finmaparrow}[0]{\overset { Fin }{ \longrightarrow } }
-\newcommand{\envwf}[1]{\vdash\,#1}
-\newcommand{\tywf}[2]{#1\,\vdash\,#2 \; \texttt{ok}}
-\newcommand{\fgjtywf}[2]{#1\,\Vdash\,#2 \; \texttt{ok}}
-\newcommand{\isvalid}[2]{#1\,\vdash\,#2}
-\newcommand{\isnotvalid}[2]{#1\,\nvdash\,#2}
-\newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
-\newcommand{\hasfgjtyp}[3]{#1\,\Vdash^{}\,#2\,:\,#3}
-\newcommand{\hassort}[3]{#1\,\vdash\,#2\,::\,#3}
-\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
-\newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
-\newcommand{\eqtyp}[3]{#1\,\vdash\,#2 \equiv #3}
-\newcommand{\generalizes}[3]{#1\,\vdash\,#2 \succ #3}
-\newcommand{\msfolsig}[0]{{\Lambda \equiv}}
-%\newcommand{\msentails}[2]{#1\,\models_L\,#2}
-\newcommand{\msentails}[2]{#1\,\models\,#2}
-\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
-\newcommand{\defeq}[0]{ \triangleq }
-\newcommand{\tymap}[0]{\mathcal{F}}
-\newcommand{\etaeq}[0]{\eta_{eq}}
-\newcommand{\semof}[1]{\thesemof{#1}}
-\newcommand{\mssemof}[1]{\thesemof{#1}}
-\newcommand{\rwsto}[0]{=}
-\newcommand{\rrwsto}[0]{\hookrightarrow}
-\newcommand{\rwbindfwd}[0]{\gamma_{\Rightarrow}}
-\newcommand{\rwbindbwd}[0]{\gamma_{\Leftarrow}}
-\newcommand{\eqhlt}[1]{\underline{#1}}
-\newcommand{\tfr}[0]{F_R}
-\newcommand{\ground}[1]{\underline{#1}}
-\newcommand{\semeq}[0]{\equiv}
-\newcommand{\elabsto}[0]{\looparrowright }
-\newcommand{\arr}[0]{$\rightarrow$}
-\newcommand{\eef}[0]{$\Leftrightarrow$}
-\newcommand{\imp}[0]{$\Rightarrow$}
-\newcommand{\RNull}[0]{\emptyset}
-\newcommand{\cty}[0]{ty}
-\newcommand{\hsp}[0]{\hspace*{-0.1in}}
-\newcommand{\hspa}[0]{\hspace*{-0.08in}}
-% removable:
-\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
-%\newcommand{\absof}[1]{\left\| #1 \right\| }
-\newcommand{\absof}[1]{\thesemof{#1}}
-\newcommand{\broom}{{\sc Broom}\xspace}
-\newcommand{\name}{{\sc Broom}\xspace}
-\newcommand{\fbname}{{\sc Featherweight Broom}\xspace}
-\newcommand{\FB}{{\sc FB}\xspace}
-\newcommand{\naiad}{{\sc Naiad}\xspace}
-\newcommand{\csolve}{{\sc CSolve}\xspace}
-\newcommand{\GK}[1]{{\textcolor{red}{{GK: #1}}}}
-\newcommand{\csharp}{C\#~}
-\newcommand{\outlives}{\succeq}
-\newcommand{\underlives}{\preceq}
-\newcommand{\coloneqq}{::=}
-\newcommand{\R}[1]{\textrm{#1}}
-\newcommand{\N}[1]{{\normalfont #1}}
-\newcommand{\M}[1]{\mathtt{#1}}
-\newcommand{\ObjZ}{\C{Object}}
-\newcommand{\RgnZ}{\C{Region}}
-\newcommand{\thisZ}{\C{this}}
-\newcommand{\superZ}{\C{super}}
-\newcommand{\unitZ}{\C{unit}}
-\newcommand{\FuncZ}{\C{Func}}
-\newcommand{\unitval}{\C{()}}
-\newcommand{\inang}[1]{\langle #1 \rangle}
-\newcommand{\rhoalloc}{\rho^a}
-\newcommand{\rhobar}{\bar{\rho}}
-\newcommand{\rhoallocm}{\rho^a_m}
-\newcommand{\rhobarm}{\bar{\rho_m}}
-\newcommand{\ralloc}{\pi^a}
-\newcommand{\rgn}{\pi}
-\newcommand{\rbar}{\bar{\pi}}
-\newcommand{\tbar}{\bar{T}}
-\newcommand{\taubar}{\bar{\tau}}
-\newcommand{\xbar}{\bar{x}}
-\newcommand{\vbar}{\bar{v}}
-\newcommand{\ebar}{\bar{e}}
-\newcommand{\extends}{\triangleleft}
-\newcommand{\rextends}{\blacktriangleleft}
-\renewcommand{\bar}[1]{\overline{#1}}
-\newcommand{\angR}{\inang{\rhoalloc,\rhobar \,|\, \phi}}
-\newcommand{\angT}{\inang{\tbar}}
-\newcommand{\tyvar}{a}
-\newcommand{\tyvarb}{b}
-\newcommand{\angAlpha}{\inang{\bar{\tyvar} \extends \bar{\fgjN}}}
-\newcommand{\fgjN}{K}
-\newcommand{\fbN}{N}
-\newcommand{\headerOf}[1]{\C{class}\; #1\angAlpha\angR \extends \fbN}
-\newcommand{\substFn}{\mathcal{S}}
-\newcommand{\mang}{\inang{\rhoalloc_m,\bar{\rho_m} \,|\, \phi_m}}
-\newcommand{\rhoset}{\rhoenv}
-\newcommand{\rhoenv}{\Delta}
-\newcommand{\aenv}{\Theta}
-\newcommand{\phicx}{\Phi}
-\newcommand{\phictxt}{\phi_{cx}}
-\newcommand{\phicstr}{\phi_{cs}}
-\newcommand{\phisol}{\phi_{sol}}
-\newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
-\newcommand{\A}{{\mathcal{A}}}
-\newcommand{\exptycx}[2]{\A,#1,#2}
-\newcommand{\void}{\C{void}}
-\newcommand{\functy}[3]{\FuncZ\inang{#1}\inang{#2,#3}}
-\newcommand{\stmtsemcx}{\A,\env}
-\newcommand{\stmtsemcxp}{\A',\env'}
-\newcommand{\stmtsemcxpp}{\A'',\env''}
-\newcommand{\sredsto}{\Rightarrow}
-\newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
-\newcommand{\okin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
-\newcommand{\lambdaexp}[3]{\lambda\inang{#1}(#2). #3}
-\newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
-\newcommand{\letexp}[3]{\C{let}\;#1\,=\,#2\;\C{in}\;#3}
-\newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
-\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\;#2\;\C{withroot}\;#3\;\C{in}\;#4}
-\newcommand{\openalloc}[3]{\C{openalloc}\;#1\;\C{withroot}\;#2\;\C{in}\;#3}
-\newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
-\newcommand{\invalidexn}{\C{InvalidOpnException}}
-\newcommand{\toprgn}{\rgn_{\top}}
-\newcommand{\csolvestar}{{\sc CSolve*}\xspace}
-\newcommand{\elabExpr}{{\sf elabExpr}}
-\newcommand{\fields}{{\sf fields}}
-\newcommand{\mtype}{{\sf mtype}}
-\newcommand{\bound}{{\sf bound}}
-\newcommand{\fresh}{{\sf fresh}}
-\newcommand{\typeOk}{{\sf typeOk}}
-\newcommand{\subtypeOk}{{\sf subtypeOk}}
-\newcommand{\templateTy}{{\sf templateTy}}
-\newcommand{\allocRgn}{{\sf allocRgn}}
-\newcommand{\shape}{{\sf shape}}
-%\newcommand{\smallskip}{\vspace*{0.1in}}
-\newcommand{\elabMeth}{{\sf elabMeth}}
-\newcommand{\frv}{{\sf frv}}
-\newcommand{\solve}{{\sf solve}}
-\newcommand{\elabClass}{{\sf elabClass}}
-\newcommand{\elabCons}{{\sf elabCons}}
-\newcommand{\elabClassTable}{{\sf elabClassTable}}
-\newcommand{\CSP}{(C\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi}, \bar{\rhoenv})}
-\newcommand{\csp}{(c\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi},\bar{\rhoenv})}
-\newcommand{\Phicx}{\Phi_{cx}}
-\newcommand{\Phics}{\Phi_{cs}}
-\newcommand{\tynwf}[2]{#1\,\nvdash\,#2 \; \texttt{ok}}
-\newcommand{\csid}[1]{\lbrack \mathbf{c_{#1}} \rbrack:\;}
+\input{macros}
+\begin{document}
 
-%Rules
-%-----
-\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
-\newcommand{\ilrulelabel}[1]{{\sc #1}}
-\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
-                           {\begin{array}{c}#2\end{array}}}
-
-% Math mode
-%-----------
-\newenvironment{nop}{}{}
-\newenvironment{smathpar}{
-\begin{nop}\small\begin{mathpar}}{
-\end{mathpar}\end{nop}\ignorespacesafterend}
-
-% Theorems Environment
-%----------------------
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{proposition}[theorem]{Proposition}
-\newtheorem{corollary}[theorem]{Corollary}
-
-\newenvironment{proof}[1][Proof]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-\newenvironment{definition}[1][Definition]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-\newenvironment{example}[1][Example]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-\newenvironment{remark}[1][Remark]{\begin{trivlist}
-\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-
-\newcommand{\qed}{\nobreak \ifvmode \relax \else
-      \ifdim\lastskip<1.5em \hskip-\lastskip
-      \hskip1.5em plus0em minus0.5em \fi \nobreak
-      \vrule height0.75em width0.5em depth0.25em\fi}
-
-\usepackage{listings}
-\usepackage{subcaption}
-\usepackage{stmaryrd}
-\usepackage[skip=5pt]{caption}
-
-\newcommand{\lstjava}{\lstset{ %
-language=Java, % choose the language of the code
-basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
-keywordstyle=\color{Bittersweet},
-% numbers=left,                   % where to put the line-numbers
-numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
-stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-numbersep=5pt,                  % how far the line-numbers are from the code
-showspaces=false,               % show spaces adding particular underscores
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-% frame=single,                   % adds a frame around the code
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to bottom
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-commentstyle=\itshape\color{MidnightBlue},
-%escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={var, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
-}}
-\lstnewenvironment{codejava}
-    { % \centering
-			\lstjava
-      \lstset{mathescape=true}%
-      \csname lst@setfirstlabel\endcsname}
-    { %\centering
-      \csname lst@savefirstlabel\endcsname}
-
-\lstnewenvironment{numcodejava}
-    { % \centering
-			\lstjava
-      \lstset{numbers=left,firstnumber=1}%
-      \csname lst@setfirstlabel\endcsname}
-    { %\centering
-      \csname lst@savefirstlabel\endcsname}
-
-\newcommand{\lstml}{\lstset{ %
-language=[Objective]Caml, % choose the language of the code
-basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
-keywordstyle=\color{Bittersweet},
-% numbers=left,                   % where to put the line-numbers
-numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
-stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
-numbersep=5pt,                  % how far the line-numbers are from the code
-showspaces=false,               % show spaces adding particular underscores
-showstringspaces=false,         % underline spaces within strings
-showtabs=false,                 % show tabs within strings adding particular underscores
-% frame=single,                   % adds a frame around the code
-tabsize=2,                      % sets default tabsize to 2 spaces
-captionpos=b,                   % sets the caption-position to bottom
-breaklines=true,                % sets automatic line breaking
-breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-commentstyle=\itshape\color{MidnightBlue},
-escapeinside={\%*}{*)},         % if you want to add a comment within your code
-morekeywords={foreach, do, done, not, : , /\\}
-}}
-\lstnewenvironment{codeml}
-    { % \centering
-			\lstml
-      \lstset{mathescape=true}%
-      \csname lst@setfirstlabel\endcsname}
-    { %\centering
-      \csname lst@savefirstlabel\endcsname}
-
-\begin{document}
+\newcommand{\dv}[1]{{\bf DV:}\begin{color}{red}#1\end{color}}
 
@@
 
-\authorinfo{}
-           {}
-           {}
+\authorinfo{} {} {}
 \maketitle
@@
 \bibliographystyle{abbrvnat}
+\small
 \bibliography{broom}

### 2016-07-01 paper/introduction.tex
integrated new type inference section

@@
 In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
-to \emph{transferring data} between a set of concurrent actors, motivated by the following concerns.
+to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
 

### 2016-07-01 paper/type-inference.tex
integrated new type inference section

@@
 
-\section{TO DO Items}
-
-\begin{itemize}
-\item Add destructive object-field-update \emph{x.f := y} to the language.
-\item Add if-then-else to the language.
-\item Omit constructors from classes (and use an implicit constructor).
-\item Discuss the safety of virtual method invocation, which provides a limited form
-of existentially quantifying regions and instantiating (unpacking) them because a
-derived class may have more region parameters than a base class.
-\item Discuss the subtyping rule used to check an overriding method (in the static
-semantics).
-\item Mention that constraints are in the theory of partial orders?
-\item Update discussion of examples
-\end{itemize}
-
-\textbf{TODO: Miscellaneous points to be integrated.}
-
-To see why, recall that all the constraints on
-identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
-by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
-Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
-only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
-solution $\phisol$ cannot exist.
-
-\textbf{TODO: Is the above inequality correct?}
+\textbf{TOD: Update example discussion}
 

### 2016-07-01 paper/introduction.tex
fixed typos minor corrections

@@
 \paragraph{Safe Transferable Regions.}
-In this paper, we present a language-based approach to safely and
-efficiently realize such transferable data that uses regions as
-the implementation technique.
+In this paper, we present a language-based approach to cleanly encapsulated
+transferable data and a safe and efficient implementation of such data based
+on regions. 
 %

### 2016-07-01 paper/type-inference.tex
fixed typos minor corrections

@@
 contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
-\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
+\C{Pair} object \C{q}, and an object \C{y}. It assigns
 \C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
 recursively with the same region, a new \C{Pair} object allocated in a
-local region, and an \C{A} object referred by the \C{snd} field of the
+local region, and an object referred by the \C{snd} field of the
 pair inside the region. \C{alt} never terminates.  Elaboration phase
@@
 definition\footnote{In reality, elaboration uses new region variables
-as parameters to the constructor and method calls, and then generates
+as parameters to the method calls, and then generates
 constraints that unify them with actuals. In our examples, to avoid
@@
 \begin{codejava}
-unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
-                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
+unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
+                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
   q.fst := y; q.snd := y; 
@@
     letregion $\rgn_1$ in
-      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
+      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
                       (p.fst,p.fst) in

### 2016-07-01 paper/type-system.tex
fixed typos minor corrections

--- 2015-11-21 paper/type-system.tex
@@
     let fst = new Object<$\rgn_1$>() in
-    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
+    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
                   (fst,snd);
@@
 In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
-$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
+$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
 and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
@@
 via a five-place reduction relation:
-\vspace*{-0.15in}
+% \vspace*{-0.15in}
 \begin{smathpar}
@@
 defines a type rule for such values, allowing them to be typed.
-Type safety theorem is now stated thus:
+The following type safety theorem shows that a well-typed program will never
+attempt to dereference an ``invalid'' reference (a reference to an object
+in a region that has been transferred or freed):
 \begin{theorem}

### 2016-07-01 paper/elaboration.tex
added appendix and moved some material there

--- 2016-07-01 paper/type-inference.tex
@@
-\section{Type Inference}
-\label{sec:type-inference}
-
-\name's region type system imposes a heavy annotation burden, and
-manually annotating C\# standard libraries with region types
-can be tedious. We now present our region type inference algorithm
-that eliminates the need to write region type annotations.
-% except on some higher-order functions.
-Formally, the type inference
-algorithm is an elaboration function from programs in $\absof{\FB}$
-(i.e., \FB without region types, but with \C{letregion} and \C{open}
-expressions, similar to the language introduced in
-\S~\ref{sec:overview}) to programs in \FB.
-
-\paragraph{Overview.}
-
-Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
-inference algorithm.
-The algorithm consists of the following steps:
-\begin{enumerate}
- \item \emph{Region Parametrization}.
-   The first step elaborates the input program by introducing \emph{formal region parameters}
-   (for each class and method), and \emph{region variables} (representing yet undetermined
-   \emph{actual region parameters}). We also introduce for each class and method, a
-   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
-   over the region parameters of that class/method.
-
-%%    as follows:
-%% \begin{enumerate}
-%% \item For every class determine a set of \emph{region parameters} ($\rho$)
-%%        it should be parametrized over.
-%% \item For every field of the class, determine its region parametrized type by
-%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
-%%        which in turn become formal region parameters of the containing class.
-%% \item  For every method and function type, determine its corresponding region
-%%    type template by introducing fresh region parameters as required by
-%%    the type of each parameter and return value (in addition to the allocation region
-%%    parameter).
-%%  \item Elaborate every \C{new} expression, method call or function application by
-%%    introducing \emph{fresh region variables} (as required by the region type
-%%    template of the called method or function).
-%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
-%%    an undetermined set of constraints over the region parameters of that
-%%    class/method.
-%% \end{enumerate}
-
- \item \emph{Constraint Generation}.
-   In the second step, we analyze the program to generate a set of constraints
-   (over the region identifiers and the predicate variables)
-   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).
-
- \item \emph{Constraint Solving}.
-   We solve the generated set of constraints using our fixpoint constraint
-   solving algorithm \csolvestar, which reduces the constraint solving problem to
-   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
-   references, for example, a reference from a transferable region to a
-   stack region, then the constraints generated during the elaboration
-   are not satisfiable. In such a case, \csolvestar{} fails to solve
-   the constraints.
-% in a Herbrand constraint system, and then relies on \csolve,
-% our abduction solver for that domain.
-
- \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
-  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
-  step 1. We apply these substitutions to the elaborated program to produce the final program.
-\end{enumerate}
-
-\begin{figure}
-\begin{numcodeml}
-Infer ($p$) =
-  let $q$ = IntroduceRegionParameters($p$) in
-  let $(r,C)$ = GenerateConstraints($q$) in
-  match (SolveConstraints($C$) with
-  | None $\longrightarrow$ None
-  | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
-\end{numcodeml}
-
-\caption{The type inference algorithm}
-\label{fig:type-inference-algo}
-\end{figure}
-
-We later on establish the soundness of the type inference algorithm, and also establish
-partial completeness results.
-
-\subsection{Region Parametrization for Classes}
-\label{sec:fb-templatization}
-
-Region parametrization is an iterative process involving the following three steps,
-the first two of which are mutually dependent on each other.
-
-\emph{Introduction of Formal Region Parameters}.
-For every class \C{C}, we identify a sequence of formal region parameters
-$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.
-
-\emph{Introduction of Actual Region Parameters}.
-We then replace every instance of class \C{C} in the program by an instance
-$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
-are fresh identifiers denoting actual region parameters.
-
-\emph{Predicate Variable Introduction}. For every class \C{C}, we introduce
-a fresh predicate variable $\varphi$, which represents the yet undetermined
-outlives constraints between the formal region parameters of class \C{C}.
-
-We identify the region parameters of classes as follows.
-
-\emph{Non-Recursive Classes}.
-The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
-The region parameters for any other non-recursive class \C{C} is determined
-only after the region parameters of any class that \C{C} depends on have been
-determined: this includes the base-class \C{B} of \C{C} and the class (type)
-of any of its fields.
-We replace every dependee type \C{T} in \C{C} by its instantiated type,
-using fresh region parameters as needed.
-The sequence of region parameters for \C{C} is defined to be
-the sequence of region parameters for the base class \C{B} concatenated
-with the list of all  fresh region parameters introduced while instantiating the types
-of the fields in the class.
-(The class inherits its allocation region from its base class. Note that if
-a class does not specify an explicit base class, it has an implicit base class
-\C{Object}.)
-
-This transformation is illustrated below, using a non-generic \C{Pair} class:
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
-\begin{codejava}
-class Pair
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ fst;
-  $\ObjZ$ snd;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ fst;
-  $\ObjZ \langle \rho_2 \rangle$ snd;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-\emph{Recursive Classes}.
-The region parameters for a recursive class is computed in
-a similar fashion, with the following difference: any recursive
-field is ignored while instantiating region parameters for the fields of
-the class, and the region parameters of the recursive class are computed
-as before. We then do parameter instantiation for all recursive fields,
-such that their region annotations (the actual region parameters) are
-exactly the same as the (formal) region parameters of the class.
-The following example illustrates this for a non-generic \C{List} class.
-The resulting class represents a linked list with spine in the region
-$\rho_0$ and data objects in the region $\rho_1$.
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
-\begin{codejava}
-class List
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-The above technique can be extended to mutually recursive classes in a
-straightforward manner, by simultaneously parametrizing them (and
-then instantiating them).
-
-\emph{Type-Parametric Classes}.
-Type parameters of classes are handled as follows.
-Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
-The parametric type \C{T} is instantiated using the number of region parameters
-that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
-to be \C{Object}, and \C{T} is instantiated with one region parameter.
-
-\emph{Function Types}.
-Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
-instantiates function types as follows: the type of every parameter as well as the return value is
-instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
-are generalized as formal region parameters of the function type.
-For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.
-
-%% Likewise, given the region-annotated definition of
-%% \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
-%% method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
-%% is \footnote{In our exposition, we assume that classes $\C{A}$ and
-%% $\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
-%% $\ObjZ$, they accept one region parameter - the allocation region of
-%% their objects.}\footnote{We abuse arrow notation to also represent
-%% types of methods, but unlike function types, there is no allocation
-%% region annotation atop the arrow in a method type.}
-%% $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
-%% \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
-%% \unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
-%% variables, and $\varphi_0$ is a fresh predicate variable denoting
-%% unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
-
-\subsection{Region Parametrization for Methods}
-
-As the next step, we introduce region parameters for every method.
-We do this by instantiating the types of all parameters and the
-return value (of the method) using fresh region identifiers (as explained previously),
-and then generalizing these region identifiers as formal region parameters
-of the method. As for the classes, we also introduce a fresh predicate variable
-$\varphi$ for every method.
-
-We then consider every method invocation in the program, and introduce
-fresh region variables representing the (yet unknown) actual region
-parameters for this particular invocation.
-%
-We similarly perform instantiation for every constructor invocation
-of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
-before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
-where $\rho_0, \cdots, \rho_n$ are fresh region variables.
-
-\subsection{The Constraint Satisfaction Problem (CSP)}
-\label{sec:fb-constraintsem}
-
-The constraint generation algorithm mimics the static type checker, but accumulates
-constraints that must hold for the type checking to succeed.
-
-\paragraph{Syntax of Constraints.}
-The constraints are expressed in terms of region identifiers ($\pi$ and $\rho$)
-and predicate variables ($\varphi$).
-
-\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
-A \emph{region variable} is introduced to represent an unknown actual region parameter
-of a method invocation or object alloation, which will be bound to a region constant by the end of the
-type inference.
-A region constant may be 
-(a) a \emph{formal region parameter} of a class or method, or
-(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
-(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
-
-Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
-%% over a set of fixed formal region parameters,
-where the set of region-constraints $\phi$ is defined by:
-\begin{smathpar}
-\begin{array}{lcl}
-\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
-\end{array}
-\end{smathpar}
-We will use the term \emph{validity constraint} to denote an entailment constraint
-of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
-
-Our constraints also make uses of \emph{pending substitutions} $F$:
-% \footnote{we borrowed this terminology from~\cite{ltpldi08}}
-A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
-used in a particular context:
-\begin{smathpar}
-\begin{array}{lcl}
-F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
-\end{array}
-\end{smathpar}
-E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
-\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
-is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
-$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
-and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
-deducible from $\rgn_1 \outlives \rgn_2$.
-
-The constraints generated are of the following kinds:
-\begin{itemize}
-
-\item Well-formedness constraints of form $\rho \in \rhoenv$,
-restricting the domain of unification for a region variable ($\rho$)
-to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
-
-\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
-variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
-regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
-  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
-  the region-constraint of its base class \C{B}).
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
-  (e.g., to indicate the region-constraint of a class \C{C} must imply the
-  region-constraint of each of its fields).
-
-\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
-where $\varphi_i$ is a predicate variable (representing the precondition of a
-method to be determined), $\phicstr$ is a region constraint that is \emph{required}
-to hold at a particular program point (within the method), and $\phictxt$ is
-a region constraint that is \emph{known} to hold at that program point.
-
-%
-%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-%% of predicate variables and pending substitutions. While $\phictxt$
-%% captures relationships that are \emph{known} to hold between concrete
-%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
-%% $\phicstr$ captures relationships that are \emph{required} to hold
-%% among region varibles (i.e., $\rho$'s), or relationships between
-%% region variables and identifiers.
-%
-
-\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
-generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
-$\varphi_i$ are as above).
-
-\end{itemize}
-
-\input{fb-constraint-generation}
-
-\paragraph{Constraint Generation.}
-The constraint generation algorithm is a direct adaption of the type checker:
-each type checking rule is modified to produce a set of constraints that must hold
-for the type checker to succeed.
-%
-Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
-
-\paragraph{Constraint Solution}
-Solving the set ($C$) of constraints entails finding an
-assignment for each predicate variable $(\varphi$) and each region
-variable ($\rho$) that occurs free in $C$, that satisfies all
-the validity constraints as well as the well-formedness constraints on
-$\varphi$ and $\rhobar$.
-% To simplify presentation, we think of $C$ as being parameterized on
-% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-% \rbrack$. We now formalize the constraint satisfaction problem, and
-% its solution.
-
-\begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
-A constraint satisfaction problem is a tuple
-$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-where $\regionConstants$ is a set of region constants,
-$\regionVars$ is a set of region variables,
-$\predVars$ is a set of predicate variables,
-$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies a unification domain for each region variable,
-$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies the domain for each predicate variable,
-and $\constraintSet$ is a set of validity constraints in one of the following forms:
-\begin{center}
-\(
-    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
-    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
-\)
-\end{center}
-The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
-where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
-\begin{itemize}
-  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
-
-  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
-    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
-
-  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
-    \regionSubstFn$.
-  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
-\end{itemize}
-\end{definition}
-
-\paragraph{Properties of generated constraints}
-In the sequel, we will restrict our attention to a special class of CSPs that satisfy
-two interesting properties that allow us to solve the constraints efficiently.
-The constraints generated our type inference algorithm satisfy these two properties.
-%
-The first property is that every region variable is guaranteed to be unified with some
-region constant by the constraints\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
-%
-The second property is that the \emph{context region constraint} $\phictxt$
-occurring on the antecedent of any validity constraint is a conjunction of
-outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-is a static region identifier and $\pi$ is a region constant (either a formal
-region parameter, or static region identifier, or open transferable region identifier).
-Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-
-\subsection{Solving the CSP}
-
-\paragraph{Constraint Solver.}
-Solving the constraints is an iterative process, consisting of the following
-steps, and this process fails at any point if it determines that no solution
-exists for the set of constraints:
-
-\begin{enumerate}
-
-\item
-    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
-$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
-If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
-solver fails.
-
-\item
-When a region variable $\rho$ is unified with a region constant $\pi$,
-if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
-and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
-If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
-
-\item
-\label{item:context}
-Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
-where $\pi_i$ and $\pi_j$ are both region constants.
-If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
-$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
-
-\item
-  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
-  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
-  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
-\end{enumerate}
-
-Point~\ref{item:context} makes use of the special structure of the generated
-constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
-form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
-region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-
-\paragraph{Algorithmic Aspects.}
-The iterative process described above is a standard fixed point computation.
-We can encode this computation using a set of Datalog rules,
-and compute the solution using any Datalog engine.
-Alternatively, algorithms for context-free reachability can be adapted
-to solve the constraints, as explained below.
-
-Given a set $S$ of outlives constraints, we define the directed
-graph $G(S)=(V(S),E(S))$ as follows.
-Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
-which we will also refer to as $\rho$.
-Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
-an edge from $\rho_1$ to $\rho_2$.
-Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
-for the pair of constraints $\rho_1 \outlives \rho_2$ and
-$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
-by a pair of edges.
-%
-It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
-there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
-%
-Thus, simple graph reachability (transitive closure) algorithms can be
-used to identify all the logical consequences of a set of outlives
-constraints.
-
-Given a set of context-free validity constraints, we can generalize the
-above approach and reduce the problem to one of context-free reachability
-in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
-%
-Algorithms for context-free reachability can also be adapted to incorporate
-the above treatment of context constraints $\phictxt$.
-%
-Further details omitted.
-
-\paragraph{Modularity Aspects.}
-The type inference algorithm, as presented, traverses the entire program to
-generate the set of constraints, which are solved en masse, using an iterative
-fixed point computation. However, the type inference can be realized in a
-modular and compositional fashion, subject only to the restrictions imposed
-by recursion.
-
-In the elaboration phase, we can process a class \C{C} only after any class
-\C{B} that \C{C} depends on has been processed: class \C{C} depends on
-class \C{B} if \C{B} is either \C{C}'s base class or the type of any field
-of \C{C} depends on \C{B}. In effect, this means that any collection of
-mutually recursive classes must be processed together. Non-recursive
-dependences can be handled in a compositional fashion: if class \C{C}
-depends on \C{B} non-recursively, then the elaboration can be done for
-\C{B} first, and then \C{C} can be processed.
-
-The same idea applies to the constraint-solving phase as well.
-Given a set of constraints, we say that a predicate variable $\varphi_1$
-\emph{directly-depends} on another predicate variable $\varphi_2$ if the set of
-constraints includes a constraint $\isvalid{\varphi_1 \conj \phictxt}{F(\varphi_2)}$.
-We say that $\varphi_1$ \emph{depends} on $\varphi_2$ if $\varphi_1$ transitively
-depends on $\varphi_2$.
-The constraint solver needs to process any collection of mutually dependent
-predicate variables together.
-In effect, this requires the type inference to process any collection of
-mutually recursive methods together.
-However, methods that are not mutually recursive can be processed separately.
-
-%% \emph{Class Constraints}. 
-%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
-%% fairly straightforward. It is the conjunction of
-%% \begin{itemize}
-%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
-%% \item $\varphi_{\C{B}}$
-%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
-%% \end{itemize}
-
-\paragraph{Soundness and Completeness.}
-
-The type inference algorithm is sound:
-
-\begin{theorem}
-\emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
-(1) $\absof{t} = p$, and
-(2) $t$ is well-typed.
-\end{theorem}
-
-Furthermore, the second phase of the type inference algorithm is complete as well.
-If there is any possible instantiation (of the region variables and predicate variables) of
-the elaborated program produced by the first phase that type checks, then the second phase
-will identify it.
-
-Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
-explained below.
-
-\begin{enumerate}
-\item Our technique for determining the set of region parameters for a recursive class
-incorporates a specific heuristic, namely that the recursive occurrence of the class has
-the same parameters, in the same order, as the class itself. This technique will fail, for
-example. if the program uses a recursive list type whose elements alternatively come
-from two different regions. Such a program would require the following elaboration,
-which is beyond the scope of our approach:
-% , $R_1$ and $R_2$.
-% both of which are region parameters of the class.
-\begin{tabular}{ccc}
-\begin{minipage}{0.3\linewidth}
-\begin{codejava}
-class List
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-\item Our technique for region parametrization also uses a heuristic in the case of
-higher order programs. Consider a higher-order method that has a function $f$ as a
-parameter. We have a choice in terms of where the region-parameters introduced for
-$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
-of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
-source of incompleteness in the type-inference.
-
-The treatment of class fields with a function type also uses the same (incomplete)
-heuristic.
-
-\end{enumerate}
-
-Region annotations provided by the user can help the type-inference overcome these
-sources of incompleteness.
-
-\subsection{Elaboration and Constraint Generation: Details}
+\section{Elaboration and Constraint Generation}
+\label{sec:elaboration}
 
@@
 constraints.
-
-\textbf{TOD: Update example discussion}
-
-\subsection{Example}
-
-\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
-template from \S\ref{sec:fb-templatization}. Following constraints are
-generated during its elaboration (Constrains are identified with
-$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
-\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
-\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
-\{\rho_5\}$, have been elided): 
-\begin{smathpar}
-\begin{array}{l}
-  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
-  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
-     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
-  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
-  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
-  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
-  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
-\end{array}
-\end{smathpar}
-
-\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
-contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
-\C{Pair} object \C{q}, and an object \C{y}. It assigns
-\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
-recursively with the same region, a new \C{Pair} object allocated in a
-local region, and an object referred by the \C{snd} field of the
-pair inside the region. \C{alt} never terminates.  Elaboration phase
-elaborates the method to the following region-annotated
-definition\footnote{In reality, elaboration uses new region variables
-as parameters to the method calls, and then generates
-constraints that unify them with actuals. In our examples, to avoid
-clutter due to trivial constraints,we coalesced both steps and show
-the actuals instead.}(The original definition of \C{alt} can be
-obtained by erasing all the region annotations from the elaborated
-version):
-% \begin{codejava}
-% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
-%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
-%   p.fst = x; p.snd = y; 
-%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
-% }
-% \end{codejava}
-\begin{codejava}
-unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
-                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
-  q.fst := y; q.snd := y; 
-  open r as p@$\rgn_0$ in
-    letregion $\rgn_1$ in
-      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
-                      (p.fst,p.fst) in
-        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
-}
-\end{codejava}
-% Note that, to avoid clutter, we have already resolved appropriate
-% region arguments to the recursive call, instead of introducing new
-% region variables and generating equality constraints on them. 
-Constraints generated during the elaboration are shown below
-(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
-\begin{smathpar}
-\begin{array}{l}
-\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
-\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
-    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
-\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
-\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
-\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
-\rgn_1} 
-    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
-\end{array}
-\end{smathpar}
-
-% \paragraph{Constraints Example 1} The original source of all
-% outlives constraints is the validity constraint $C_3$ in $\elabClass$
-% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
-% class template from \S\ref{sec:fb-templatization}, the constraint is
-% as following:
-% \begin{center}
-%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
-%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
-% \end{center}
-
-% \paragraph{Constraints Example 2} Consider a contrived method
-% $\C{alt}$ that accepts an argument $\C{p}$ of type
-% $\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
-% of type $A$. It then assigns one object to $\C{p.fst}$ and other to
-% $\C{p.snd}$, and calls itself recursively without ever terminating.
-% Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
-% alternated between recursive calls. The region-annotated definition of
-% $\C{alt}$ is shown below:
-% \begin{codejava}
-% unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
-%                           A<$\rho_3$> x, A<$\rho_4$> y) {
-%   return p.fst = x; p.snd = y; 
-%          this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
-% }
-% \end{codejava}
-% Note that, to avoid clutter, we have already resolved appropriate
-% region arguments to the recursive call, instead of introducing new
-% region variables and generating equality constraints on them. Rest of
-% the validity constraints generated while elaborating $\C{alt}$ to the
-% above region-annotated definition are shown below:
-% \begin{center}
-% \(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
-% \(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
-% \end{center}
-% The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
-% Second and third are generated by the the assignment expressions, and
-% the last constraint is generated from the recursive call.

### 2016-07-01 paper/type-inference.tex
added appendix and moved some material there

@@
 Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
+%
+The appendix presents a more detailed algorithm that performs the elaboration and
+constraint generation in a single integrated pass.
 
@@
 The constraints generated our type inference algorithm satisfy these two properties.
-%
+
 The first property is that every region variable is guaranteed to be unified with some
@@
 % , and the type system uniquely determines the region parameters of every value
-%
+
 The second property is that the \emph{context region constraint} $\phictxt$
@@
 it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
+simplifying the constraint solver.
 
@@
 
-%% \emph{Class Constraints}. 
-%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
-%% fairly straightforward. It is the conjunction of
-%% \begin{itemize}
-%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
-%% \item $\varphi_{\C{B}}$
-%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
-%% \end{itemize}
-
 \paragraph{Soundness and Completeness.}
@@
 
-\subsection{Elaboration and Constraint Generation: Details}
-
-\input{fb-elabexpr}
-\input{fb-elabmeth}
-
-%Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
-%(a). replacing core types in variable declarations and \C{new}
-%expressions with fresh region type templates, and (b). explicitly
-%instantiating region parameters of methods with fresh region variables
-%in method calls and function applications. This elaboration is
-%performed with respect to the polymorphic type templates of classes
-%and methods computed as per \S\ref{sec:fb-templatization}.
-
-We now present an integrated algorithm that processes all the methods
-in a given program, elaborating them (by introducing region variables
-to stand for actual region parameters) and simultaneously generating
-the set of constraints that must be solved.
-
-Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
-this elaboration for a subset of expressions in $\absof{\FB}$, whose
-corresponding $\FB$ expressions have been ascribed static semantics in
-Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
-context as the expression typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning. The function traverses expressions in a
-syntax-directed manner of a type checker, introducing fresh region
-type templates for unknown region types, while generating constraints
-over region and predicate variables.
-Note that $\elabExpr$ returns the type of the subexpression,
-which is used to generate constraints for the expression. Functions
-$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
-implement type well-formedness and subtype judgments from
-Fig.~\ref{fig:fb-staticsem}, respectively.
-
-Functions $\elabMeth$ and $\elabClass$ shown in
-Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
-class definitions, respectively. Both functions first build a context
-($\A$) containing a set ($ \rhoenv$) of region variables denoting
-regions that are currently live, a map ($\aenv$) mapping type
-variables to their bounds, and a constraint formula ($\phicx$)
-capturing constraints over live region variables. We use predicate
-variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
-
-Function $\elabMeth$ elaborates a method definition of class $B$. It
-calls $\elabExpr$ with the context $\A$, its allocation context
-parameter ($\rhoallocm$), and a type environment ($\env$) that
-contains region type bindings for all the arguments of the method,
-including the implicit $\C{this}$ argument. The type 
-returned by $\elabExpr$ for the method body is checked against its
-expected type (derived from the type of the method)
-generating more constraints. The function then returns the elaborated
-method definition and the set of constraints.
-
-$\elabClass$ elaborates the definition of a class $B$. It relies on
-$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
-hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
-constructor ($k$) and method definitions ($\bar{d}$), respectively. To
-the set of constraints returned by these functions, $\elabClass$ adds
-constraints generated by checking the well-formedness of the type
-templates of its superclass and fields, and also a new constraint
-capturing a couple of safety conditions: first, the allocation regions
-of objects referred by the instance variables should outlive the
-allocation region of the instance itself, and second, the allocation
-regions of a class type and its superclass type must be the same.
-
-Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
-every definition in the class table $CT$, while accumulating
-constraints.
-
-\textbf{TOD: Update example discussion}
+\textbf{TODO: Update example discussion}
 

### 2016-07-02 paper/type-inference.tex
minor fix

@@
 \emph{Type-Parametric Classes}.
-Type parameters of classes are handled as follows.
-Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
-The parametric type \C{T} is instantiated using the number of region parameters
-that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
-to be \C{Object}, and \C{T} is instantiated with one region parameter.
+% Type parameters of classes are handled as follows.
+% Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
+The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
+% number of region parameters
+% that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
+% to be \C{Object}, and \C{T} is instantiated with one region parameter.
 

### 2016-07-02 paper/type-inference.tex
soundness, completeness, examples

@@
 
-The type inference algorithm is sound:
+We now state the soundness theorem:
 
@@
 
-Furthermore, the second phase of the type inference algorithm is complete as well.
-If there is any possible instantiation (of the region variables and predicate variables) of
-the elaborated program produced by the first phase that type checks, then the second phase
-will identify it.
+We do not present a formal proof of the theorem. The proof essentially requires
+establishing the following lemma inductively: if
+$
+ \exprok {\exptycx{\ralloc}{\env}}{e}{\tau}{C}
+$
+and $(\regionSubstFn,\predSubstFn)$ satisfies $C$, then
+$
+ \hastyp {\exptycx{\ralloc}{\env}}{e[\regionSubstFn,\predSubstFn]}{\tau}
+$,
+where $e[\regionSubstFn,\predSubstFn]$ indicates $e$ with the substitutions
+in the solution $(\regionSubstFn,\predSubstFn)$,
+as well as analogous lemmas for type well-formedness.
 
-Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
-explained below.
+However, in this context, completeness is more interesting than soundness.
+(Note that an unsound inference algorithm can be made sound by simply type-checking
+the resulting program.)
+We believe that the converse of the above lemma can also be established:
+any substitution for the region variables and the predicate variables that makes the
+expression well-typed will also satisfy the generated constraints.
+Thus, if there is any possible instantiation (of the region variables and predicate variables) of
+the elaborated program produced by the first phase that type checks, then the generated
+constraints are satisfiable, and the second phase will identify it.
+
+However, the entire type inference algorithm is not complete, because of some heuristic
+choices made during the first phase (elaboration phase), as explained below.
 
@@
 
-\textbf{TODO: Update example discussion}
-
-\subsection{Example}
-
-\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
-template from \S\ref{sec:fb-templatization}. Following constraints are
-generated during its elaboration (Constrains are identified with
-$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
-\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
-\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
-\{\rho_5\}$, have been elided): 
-\begin{smathpar}
-\begin{array}{l}
-  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
-  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
-     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
-  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
-  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
-  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
-  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
-\end{array}
-\end{smathpar}
-
-\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
-contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
-\C{Pair} object \C{q}, and an object \C{y}. It assigns
-\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
-recursively with the same region, a new \C{Pair} object allocated in a
-local region, and an object referred by the \C{snd} field of the
-pair inside the region. \C{alt} never terminates.  Elaboration phase
-elaborates the method to the following region-annotated
-definition\footnote{In reality, elaboration uses new region variables
-as parameters to the method calls, and then generates
-constraints that unify them with actuals. In our examples, to avoid
-clutter due to trivial constraints,we coalesced both steps and show
-the actuals instead.}(The original definition of \C{alt} can be
-obtained by erasing all the region annotations from the elaborated
-version):
-% \begin{codejava}
-% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
-%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
-%   p.fst = x; p.snd = y; 
-%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
-% }
-% \end{codejava}
-\begin{codejava}
-unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
-                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
-  q.fst := y; q.snd := y; 
-  open r as p@$\rgn_0$ in
-    letregion $\rgn_1$ in
-      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
-                      (p.fst,p.fst) in
-        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
-}
-\end{codejava}
-% Note that, to avoid clutter, we have already resolved appropriate
-% region arguments to the recursive call, instead of introducing new
-% region variables and generating equality constraints on them. 
-Constraints generated during the elaboration are shown below
-(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
-\begin{smathpar}
-\begin{array}{l}
-\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
-\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
-    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
-\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
-\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
-\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
-\rgn_1} 
-    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
-\end{array}
-\end{smathpar}
-
+%\textbf{TODO: Update example discussion}
+%
+%\subsection{Example}
+%
+%\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
+%template from \S\ref{sec:fb-templatization}. Following constraints are
+%generated during its elaboration (Constrains are identified with
+%$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
+%\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
+%\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
+%\{\rho_5\}$, have been elided): 
+%\begin{smathpar}
+%\begin{array}{l}
+%  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
+%  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
+%     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
+%  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
+%  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
+%  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
+%  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
+%\end{array}
+%\end{smathpar}
+%
+%\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
+%contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
+%\C{Pair} object \C{q}, and an object \C{y}. It assigns
+%\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
+%recursively with the same region, a new \C{Pair} object allocated in a
+%local region, and an object referred by the \C{snd} field of the
+%pair inside the region. \C{alt} never terminates.  Elaboration phase
+%elaborates the method to the following region-annotated
+%definition\footnote{In reality, elaboration uses new region variables
+%as parameters to the method calls, and then generates
+%constraints that unify them with actuals. In our examples, to avoid
+%clutter due to trivial constraints,we coalesced both steps and show
+%the actuals instead.}(The original definition of \C{alt} can be
+%obtained by erasing all the region annotations from the elaborated
+%version):
+%% \begin{codejava}
+%% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
+%%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
+%%   p.fst = x; p.snd = y; 
+%%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
+%% }
+%% \end{codejava}
+%\begin{codejava}
+%unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
+%                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
+%  q.fst := y; q.snd := y; 
+%  open r as p@$\rgn_0$ in
+%    letregion $\rgn_1$ in
+%      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
+%                      (p.fst,p.fst) in
+%        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
+%}
+%\end{codejava}
+%% Note that, to avoid clutter, we have already resolved appropriate
+%% region arguments to the recursive call, instead of introducing new
+%% region variables and generating equality constraints on them. 
+%Constraints generated during the elaboration are shown below
+%(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
+%\begin{smathpar}
+%\begin{array}{l}
+%\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
+%\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
+%    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
+%\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
+%\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
+%\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
+%\rgn_1} 
+%    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
+%\end{array}
+%\end{smathpar}
+%
 % \paragraph{Constraints Example 1} The original source of all

### 2016-07-02 paper/type-system.tex
added op semantics
added appendix and moved some material there
removed constructor

@@
 $R_2$ is guaranteed to outlive $R_1$. 
-% Intuitively, the invariant can
-% be enforced by (a). tracking \emph{outlives} relationships between
-% various regions in the program (b). tagging the C\# type of every
-% object in the program with its allocation region, and (c).  ensuring
-% that, when a reference is created, its target object ($o_2$) is
-% allocated in a region that is known to be in outlives relationship
-% with the object containing the reference. 
 We now formally develop this intuition via \fbname (\FB), our
@@
   b@$\rho_2$ snd;
-  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
-    super(); this.fst = fst; this.snd = snd;
-  }
   a@$\rho_1$ getFst() { 
@@
 \end{codejava}
-\vspace*{-0.15in}
+% \vspace*{-0.15in}
+
+%  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
+%    super(); this.fst = fst; this.snd = snd;
+%  }
 
@@
 
-The supplement defines a small-step operational semantics for \fbname
+Fig.~\ref{fig:fb-opsem} in the appendix defines a small-step operational semantics for \fbname
 via a five-place reduction relation:

### 2016-07-03 paper/abstract.tex
abstract and intro
Wibbles plus new reference

--- 2015-11-21 paper/abstract.tex
@@
 dangling pointers. A key contribution of this paper is a
-refinement-based region type system that ensures the memory safety of
-C\# programs in the presence of transferable regions. We complement
+refinement-based region type system that ensures memory safety
+in the presence of transferable regions. We complement
 our type system with a type inference algorithm that infers principal
@@
 write region annotations on types, while facilitating the reuse of
-existing C\# libraries with no modifications. Experiments demonstrate
-the practical utility of our approach.
+existing libraries with few modifications. Experiments demonstrate
+the expressiveness and practical utility of our approach.
 

### 2016-07-03 paper/introduction.tex
Wibbles plus new reference

--- 2016-07-01 paper/introduction.tex
@@
 
-In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
-to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
-
-Consider the example (of  a producer creating a data structure and transferring it to a consumer)
-in Fig.~\ref{fig:motivating-eg}. This represents a \C{SELECT} query operator for streaming data. 
-The operator receives a stream of input messages, each
-associated with a time (window) $t$, processed by method
-\texttt{onReceive}.  Each input message contains a list of inputs,
-each of which is processed by applying a user-defined function to
-create a corresponding output.  The operator may receive multiple
+Consider the example, from~\cite{Broom:HotOS}, shown in
+Fig.~\ref{fig:motivating-eg}.  This code represents the logic for a
+streaming query operator.  The operator receives a stream of input
+messages, each associated with a time (window) $t$, processed by
+method \texttt{onReceive}.  Each input message contains a list of
+inputs, each of which is processed by applying a user-defined function
+to create a corresponding output.  The operator may receive multiple
 messages with the same timestamp (and messages with different
@@
 
-\paragraph{Performance.}
-When the transferred data structures are large, as is the case with many streaming big-data
-analysis systems, garbage collection overhead becomes
-significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
-dataflow system, the GC pause at one node can have a cascading adverse
-effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
-a GC pause at an upstream actor can block downstream actors that are waiting for
-messages.  However, often much of the GC overhead results from the
-collector performing avoidable or unproductive work.  For the example
-in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
-\C{map} data structure, although its objects cannot be collected until a
-suitable timing message arrives.
+This example is an instance of a general pattern, where a producer creates a
+data structure and passes it to a consumer. In a system where most of the
+computation takes this form, and these data structures are very large, as is
+the case with many streaming big-data analysis systems, garbage collection
+overhead becomes significant~\cite{Broom:HotOS,Nguyen+:facade}.  Furthermore, in a
+distributed dataflow system, the GC pause at one node can have a cascading
+adverse effect on the performance of other nodes, particularly when real-time
+streaming performance is required~\cite{Broom:HotOS, harris15}.  In
+particular, a GC pause at an upstream actor can block downstream actors that
+are waiting for messages.  However, often much of the GC overhead results
+from the collector performing avoidable or unproductive work.  For example,
+in the process executing the code from Fig.~\ref{fig:motivating-eg}, GC might
+repeatedly traverse the \C{map} data structure, although its objects cannot
+be collected until a suitable timing message arrives.
 
-\begin{figure}
-\centering
-\subcaptionbox {
-  Before transfer.
-  \label{fig:after-transfer}
-} [
-  0.35\columnwidth
-] {
-  \includegraphics[scale=0.36]{producer-heap}
-}
-\subcaptionbox {
-  After transfer.
-  \label{fig:before-transfer}
-} {
-  \includegraphics[scale=0.36]{producer-consumer-heap}
-}
+% actors.
+% performance degradation due to
+% GC overhead becomes aggravated in the distributed 
+% For example, a consumer process may trigger a
+% GC as soon as it receives a large data structure from a producer
+% process, which has just finished its GC. In such case, consumer's GC
+% traverses the entire data structure performing little or no work.
 
-\caption{References in and out of a transferable region ($R$) become
-invalid after transfer.}
-\label{fig:transfer-violations}
-\end{figure}
+% as the collector scans these large data structures For instance, in
+% the above example, the list at \C{map[t]} cannot be collected until
+% the the timing message arrives, and the list is transferred.
+% \C{transfer} operation is complete.  using alternative memory
+% management techniques.
 
-\paragraph{Safety.}
-These systems are designed to allow the producer and consumer to
-execute in the same address space or in different address spaces
-(as determined by the compiler and runtime).
-We wish to ensure memory safety in the presence of such transferred data by ensuring
-that there are no invalid references. A transfer operation, with no
-additional checks, may cause memory safety violations, both at the
-producer of the data structure, and its (possibly remote) consumer. At
-the producer, any existing references into the transferred data structure become
-invalid post transfer. If the data structure contains references to
-objects outside (the transferred data), then such references become
-invalid in the context of the consumer. Both scenarios are depicted in
-Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
-particularly unwelcome in the context of languages like C\# and Java
-that are otherwise memory-safe. While references from within the
-transferable data to outside can be disallowed in the interest of safety,
-references into transferable data needs to be allowed before the data
-is transferred. Allowing such references is crucial to performance,
-as any non-trivial program creates temporary references to the internal
-objects of a data-structure.
-
-\paragraph{Safe Transferable Regions.}
-In this paper, we present a language-based approach to safely and
-efficiently realize such transferable data that uses regions as
-the implementation technique.
-%
-A region is a block of memory that is allocated
-and freed in one shot, often in constant time. A region may contain
+An important observation, in the context of processes of the kind
+described above, is that the data-structures exchanged between them
+can be partitioned into sets of fate-sharing objects with common
+lifetimes, which makes them good candidates for a region-based memory
+management discipline. A region is a block of memory that is allocated
+and freed in one shot, consuming constant time. A region may contain
 one or more contiguous range of memory locations, and individual
@@
 while they are deallocated en masse when the region is freed.  Thus, a
-region is a good fit for a transferable data-structure.  In
-Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
+region is a good fit for a set of fate-sharing objects.
+% that is used to realize a data structure with a well-defined
+% lifetime. 
+In Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
 time window $t$ (i.e., \C{map[t]}) can be a separate region that is
@@
 
-The use of regions alleviates the performance concern related to
-garbage collection described earlier. (It also enables more efficient
-serialization/deserialization of data-structures across address spaces,
-which is also a significant performance bottleneck in such systems.)
-However, manual region memory management introduces memory safety
-issues such as dangling pointers.
-We present a single type system that simultaneously addresses
-this safety concern, as well as those described above related to
-the use of \emph{transferable data}.
+Region-based memory management, both manual as well as automatic, has
+been known for a long time. Manual region-based memory management
+suffers from the usual drawbacks, namely the potential for invalid
+references and the consequent lack of memory safety. Automatic region-based memory
+management systems guarantee memory safety, but impose various
+restrictions.  MLKit, which implements the approach pioneered by Tofte
+and Talpin~\cite{tofte94,tofte97}, for example, uses lexically scoped
+regions.  At runtime, the set of all regions (existing at a point in
+time) forms a stack. Thus, the lifetimes of all regions must be
+well-nested: it is not possible to have two regions whose lifetimes
+overlap, with neither one's lifetime contained within the other.
+Unfortunately, the data structures in the above example do not satisfy
+this restriction (as the output messages for multiple time windows may
+be simultaneously live, without any containment relation between their
+lifetimes).  We refer to regions with lexically scoped lifetimes as
+\emph{stack regions} and to regions that do not have such a lexically
+scoped region as \emph{dynamic} regions.
 
-% Region-based memory management, both manual as well as automatic, has
-% been known for a long time. Manual region-based memory management
-% suffers from the usual drawbacks, namely the potential for invalid
-% references and the consequent lack of memory safety. Automatic
-% region-based memory management systems guarantee memory safety, but
-% impose various restrictions.  MLKit, which implements the approach
-Safe region-based memory management using types was
-pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
-use only lexically scoped regions.  At runtime, the set of all regions
-(existing at a point in time) forms a stack. Thus, the lifetimes of
-all regions must be well-nested: it is not possible to have two
-regions whose lifetimes overlap, with neither one's lifetime contained
-within the other.  Unfortunately, the data structures in the above
-example do not satisfy this restriction (as the output messages for
-multiple time windows may be simultaneously live, without any
-containment relation between their lifetimes).  We refer to regions
-with lexically scoped lifetimes as \emph{stack regions} and to regions
-that do not have such a lexically scoped region as \emph{dynamic}
-regions.
+The goal of this work is a \emph{memory-safe} region-based memory
+management technique that supports dynamic regions as first-class
+objects. Our focus, in this paper, is on dynamic regions, which can be
+safely transferred across address spaces. We refer to such dynamic
+regions as \emph{transferable} regions. As with allocation and
+deallocation, transferring a memory region is fast, and therefore,
+transferring a data structure contained in a region is more efficient
+that traversing its objects in heap, and transferring them
+independently\footnote{Empirical studies in~\cite{Broom:HotOS} support
+this claim}. In the \C{SelectVertex} example from
+Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
+output for each time window $t$ must be transferable due to the
+\C{transfer} operation on Line 16. As it is the case with
+\C{SelectVertex}, the transferred data is no longer accessed by the
+producer, so the transfer operation in our system deallocates the
+region once the transfer is complete.
 
-Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
-that can be safely transferred across address spaces. We refer to such dynamic
-regions as \emph{transferable} regions.
-%
-Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
-that combine linear types and regions to suport dynamic regions.
-Unlike the prior work, we ensure memory safety in the presence of
-transferable regions through a \emph{combination of a static
-typing discipline and lightweight runtime checks in place of linear types}.
+With respect to memory safety, the key property we wish to ensure is
+that there are no invalid references: i.e., references to objects that
+were deallocated, or simply never existed. Transfer operation, with no
+additional checks, may cause memory safety violations, both at the
+producer of the data structure, and its (possibly remote) consumer. At
+the producer, any existing references into the data structure become
+invalid post transfer. If the data structure contains references to
+objects outside its (transferable) region, then such references become
+invalid in the context of the consumer. Safety violations of this kind
+are particularly unwelcome as the program with GC did not have them to
+begin with. Note that, while the references of later kind (i.e.,
+references that escape a region) defeat the very purpose of a
+transferable region, hence need to be prohibited, references of the
+former kind (i.e., references into a region from outside) are not at
+odds with the concept of a transferable region, hence need to be
+permitted. In fact, allowing such references is crucial to
+performance, as any non-trivial program creates temporary objects, and
+it is undesirable to allocate them in a transferable region; such
+regions are meant for the data being transferred. Since transferable
+regions are first class objects in our setting, controlling references
+to and from such regions while ensuring their safety without
+significantly diluting the performance advantage of regions over GC is
+a challenging exercise.
 
-The cornerstone of this approach is an \C{open} lexical block for transferable regions,
+In this paper, we describe an approach that restores memory safety in
+presence of transferable regions through a combination of a static
+typing discipline and lightweight runtime checks. The cornerstone of
+our approach is an \C{open} lexical block for transferable regions,
 that ``opens'' a transferable region and guarantees that the region
-won't be transferred/freed while it is open. By nesting an
-MLKit-style \C{letregion} lexical block, that delimits
-the lifetime of a stack region, inside an \C{open} lexical block for a
-transferable region, we can guarantee that the transferable region
-will remain live as long as the stack region is live. We say that the
-former \emph{outlives} the later., and any references from the stack
-region to the transferable region are therefore safe.  Next, we note
-that by controlling the outlives relationships between various
-regions, we only allow safe cross-region references, while prohibiting
-unsafe ones. In the above example, an outlives relationship
-\emph{from} the stack region \emph{to} the transferable region means
-that the references in that direction are allowed, but not the
-references in the opposite direction. In contrast, if an \C{open}
-block of a transferable region $R_0$ is nested inside an \C{open}
-block of another transferable region $R_1$, we do not establish any
-outlives relationships, thus declaring our intention to not allow any
-cross-region references between $R_0$ and $R_1$.  Finally, we observe
-that outlives relationships are established based on the lexical
-structure of the program, hence a static type system can enforce them
-effectively. By assigning region types to objects, which capture the
-regions such objects are allocated in, and by maintaining outlives
-relationships between various regions, we can statically decide the
-safety of all references in the program.
+won't be transferred/freed while it is open. Our observation is that
+by nesting a~\cite{tofte94}-style \C{letregion} lexical block, that
+delimits the lifetime of a stack region, inside an \C{open} lexical
+block for a transferable region, we can guarantee that the
+transferable region will remain live as long as the stack region is
+live. We say that the former ``outlives'' the later\footnote{We borrow
+the \emph{outlives} relation from~\cite{MIT03}. A comparison of our
+approach with~\cite{MIT03} can be found in
+\S~\ref{sec:related-work}.}, and any references from the stack region
+to the transferable region are therefore safe. 
+% In general, we can allow references from a region
+% $R_0$ to another region $R_1$, provided that we establish that $R_1$ outlives
+% $R_0$. 
+Next, we note that by controlling the ``outlives'' relationships
+between various regions, we only allow safe cross-region references,
+while prohibiting unsafe ones. In the above example, an outlives
+relationship \emph{from} the stack region \emph{to} the transferable
+region means that the references in that direction are allowed, but
+not the references in the opposite direction. In contrast, if an
+\C{open} block of a transferable region $R_0$ is nested inside an
+\C{open} block of another transferable region $R_1$, we do not
+establish any outlives relationships, thus declaring our intention to
+not allow any cross-region references between $R_0$ and $R_1$.
+Finally, we observe that outlives relationships are established based
+on the lexical structure of the program, hence a static type system
+can enforce them effectively. By assigning region types to objects,
+which capture the regions such objects are allocated in, and by
+maintaining outlives relationships between various regions, we can
+statically decide the safety of all references in the program.
 
-Formally, our type system introduces \emph{region parameters} (for both
-classes and methods) and uses constrained parametric polymorphism
-over these parameters, where the constraints capture outlives constraints
-between the region parameters.
-
-Ensuring memory safety using this approach requires ensuring
-that the use of transferable regions satisfies certain temporal properties.
-Firstly, a transferable region should not be transferred/freed
-inside an open block of that region (i.e, while it is still open).
-Secondly, a transferred/freed region should not be opened. These are
-typestate invariants on the transferable region objects, which are
-hard to enforce statically in the presence of unrestricted
-aliasing. Techniques like linear types and unique pointers can be used
-to restrict aliasing, but the constraints they impose are often hard
-to program around. We therefore enforce typestate invariants at
-runtime via lightweight checks. In particular, we define an acceptable
-state transition discipline for transferable regions
-(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
-transition of a transferable region (\eg, from \emph{open} state to
-\emph{freed} state) is valid or not. The check is lightweight since it
-only involves checking a single tag that captures the current state.
-We believe that this is a reasonable choice since regions are
-coarse-grained objects manipulated infrequently, when compared to the
-fine-grained objects that are present inside these regions, for which
-safety is enforced statically.
-
-One of the novel contributions of this paper is a sound type inference algorithm
-that eliminates the need for users to write region type annotations.
-This, we believe, significantly reduces the impediment to adopt our
-approach in practical setting.
-Our inference algorithm proceeds in two stages: in the first stage,
-we introduce region parameters (for classes and methods), and in
-the second stage we infer the outlives constraints between these
-parameters (using an interprocedural fixed point computation).
-% While the first
+Indeed, the utility of our approach described above is predicated
+on the assumption that we can enforce certain invariants on
+transferable regions. Firstly, a transferable region cannot be
+transferred/freed inside an open block of that region (i.e, while it
+is still open).  Secondly, a transferred/freed region cannot be
+opened. These are typestate invariants on the transferable region
+objects, which are hard to enforce statically due to the presence of
+unrestricted aliasing. Techniques like linear types and unique
+pointers can be used to restrict aliasing, but the constraints they
+impose are often hard to program around. We therefore enforce
+typestate invariants at runtime via lightweight checks. In
+particular, we define an acceptable state transition discipline for
+transferable regions (Fig.~\ref{fig:region-fsm}), and check, at
+runtime, whether a given transition of a transferable region (\eg,
+from \emph{open} state to \emph{freed} state) is valid or not. The
+check is lightweight since it only involves checking a single tag that
+captures the current state.  We believe that this is a reasonable
+choice since regions are coarse-grained objects manipulated
+infrequently, when compared to the fine-grained objects that are
+present inside these regions, for which safety is enforced statically.
+An added advantage of delegating the enforcement of typestate
+invariants to runtime is that our region type system is
+simple\footnote{Supplement lists all the type rules of our language,
+which occupy less than a page.}, which made it possible to formulate a
+type inference that completely eliminates the need to write region
+type annotations. This, we believe, significantly reduces the
+impediment to adopt our approach in practical setting.
 
@@
   %\S~\ref{sec:conclusion} concludes.
-
-% , and it is undesirable to allocate them in a transferable
-% region; such regions are meant for the data being transferred.
-
-%As with allocation and deallocation, transferring a memory region is fast, and therefore,
-%transferring a data structure contained in a region is more efficient
-%that traversing its objects in heap, and transferring them
-%independently.
-%In the \C{SelectVertex} example from
-%Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
-%output for each time window $t$ must be transferable due to the
-%\C{transfer} operation on Line 16. As it is the case with
-%\C{SelectVertex}, the transferred data is no longer accessed by the
-%producer, so the transfer operation in our system deallocates the
-%region once the transfer is complete.
-%
-%Since transferable regions are first class objects in our setting,
-%controlling references to and from such regions while ensuring their
-%safety without significantly diluting the performance advantage of
-%regions over GC is a challenging exercise.
-%
-%An important observation, in the context of processes of the kind
-%described above, is that the data-structures exchanged between them
-%can be partitioned into sets of fate-sharing objects with common
-%lifetimes, which makes them good candidates for a region-based memory
-%management discipline.

### 2016-07-03 paper/introduction.tex
Wording.

@@
 invariants to runtime is that our region type system is
-simple\footnote{Supplement lists all the type rules of our language,
+simple\footnote{The accompanying material lists all the type rules of our language,
 which occupy less than a page.}, which made it possible to formulate a

### 2016-07-03 paper/type-system.tex
Comments from DV

--- 2016-07-01 paper/type-system.tex
@@
 
+\dv{Why do we have the type parameters of $a$ and $b$ extending \C{Object} versus
+  the actual class extending \C{Object<$\rho^a$>}? Also, instead of explicitly introducing
+  all those $\rho^a$ variables, wouldn't it be nicer if we overloaded \C{this} to be a valid region,
+  meaning {\bf the} region of this object? I think notationally that might be more lightweight?}
+
 A class in \FB is parametric over zero or more type variables (as in
@@
     let fst = new Object<$\rgn_1$>() in
-    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
+    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                   (fst,snd);
@@
 %\vspace*{-0.1in}
+
+\dv{Note that we slightly changed notation here compared to the introduction, which
+  had introduded n\C{new C@r} for new object allocations. I think we should make the
+  connection in the beginning of this section.}
+
 In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
-$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
+$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
 and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
@@
 via a five-place reduction relation:
-% \vspace*{-0.15in}
+\vspace*{-0.15in}
 \begin{smathpar}
@@
 defines a type rule for such values, allowing them to be typed.
-The following type safety theorem shows that a well-typed program will never
-attempt to dereference an ``invalid'' reference (a reference to an object
-in a region that has been transferred or freed):
+Type safety theorem is now stated thus:
 \begin{theorem}

### 2016-07-03 paper/type-system.tex
More comments on syntax.
Commentary updates.

@@
 
-\dv{Why do we have the type parameters of $a$ and $b$ extending \C{Object} versus
-  the actual class extending \C{Object<$\rho^a$>}? Also, instead of explicitly introducing
+\dv{Explain why do we have the type parameters of $a$ and $b$ extending \C{Object} versus
+  the actual class extending \C{Object<$\rho^a$>}? What is the point of having it in \C{Object<$\rho^a$>?}.
+  I think it does not make sense to include it there, exactly as we do not include regions in the type parameters,
+  it's all about the type inheritance hierarchy. A second point: instead of explicitly introducing
   all those $\rho^a$ variables, wouldn't it be nicer if we overloaded \C{this} to be a valid region,
@@
 as such regions, and objects allocated in them, don't find their way
-into the result of evaluating $e$}.
+into the result of evaluating $e$}. \dv{Which one is that rule? If it's the upper
+  left rule, is there a typo? What is it's type? What is the context?}
 
@@
 via a five-place reduction relation:
-\vspace*{-0.15in}
+%\vspace*{-0.15in}
 \begin{smathpar}

### 2016-07-03 paper/type-inference.tex
Suggestion for the type inference section.

--- 2016-07-01 paper/type-inference.tex
@@
 
-\name's region type system imposes a heavy annotation burden, and
-manually annotating C\# standard libraries with region types
-can be tedious. We now present our region type inference algorithm
-that eliminates the need to write region type annotations.
-% except on some higher-order functions.
-Formally, the type inference
+\name's region type system imposes a heavy annotation burden, and the
+cost of manually annotating C\# standard libraries with region types
+is prohibitive. We now present our region type inference algorithm
+that completely eliminates the need to write region type annotations,
+except on some higher-order functions. Formally, the type inference
 algorithm is an elaboration function from programs in $\absof{\FB}$
@@
 expressions, similar to the language introduced in
-\S~\ref{sec:overview}) to programs in \FB.
+\S~\ref{sec:overview}) to programs in \FB. The key aspects of this
+elaboration are described below.
 
-\paragraph{Overview.}
+% We formalize region type inference as an elaboration function from
+% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
+% programs in \FB.  The elaboration proceeds in four steps. In the first
+% step we make use of the observation that region types are
+% \emph{refinements} of FGJ types with region annotations and
+% constraints over such region annotations, and compute polymorphic
+% region type templates for methods and classes from their FGJ types.
+% The templates contain region variables ($\rho$) to denote unknown
+% region annotations, and predicate variables ($\varphi$) to denote
+% unknown constraints over such region annotations. Free region
+% variables are generalized in types (hence, polymorphic). Second, we
+% elaborate expressions by introducing region variables to denote
+% unknown region arguments in \C{new} expressions, method calls and
+% function applications. While elaborating expressions, we make use of
+% the computed region type templates to build a system of constraints
+% that capture well-formedness requirements and subtyping relationships
+% between type templates that must hold (as per the static semantics in
+% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
+% lift expression elaboration and constraint generation to methods and
+% classes. Finally, we solve the constraints to compute solutions to
+% region ($\rho$) and predicate ($\varphi$) variables.
+%
+% Due to the presence of region-polymorphic
+% recursion in \FB, constraints generated by the algorithm can be
+% circular. More precisely, constraints generated can assume the form
+% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
+% non-idempotent substitution function for region variables in
+% $\varphi$. In the constraint solving phase, the algorithm then relies
+% on a fixpoint constraint solving algorithm called \csolvestar to solve
+% the constraints and determine assignments for unknown region and
+% predicate variables.
 
-Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
-inference algorithm.
-The algorithm consists of the following steps:
-\begin{enumerate}
- \item \emph{Region Parametrization}.
-   The first step elaborates the input program by introducing \emph{formal region parameters}
-   (for each class and method), and \emph{region variables} (representing yet undetermined
-   \emph{actual region parameters}). We also introduce for each class and method, a
-   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
-   over the region parameters of that class/method.
-
-%%    as follows:
-%% \begin{enumerate}
-%% \item For every class determine a set of \emph{region parameters} ($\rho$)
-%%        it should be parametrized over.
-%% \item For every field of the class, determine its region parametrized type by
-%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
-%%        which in turn become formal region parameters of the containing class.
-%% \item  For every method and function type, determine its corresponding region
-%%    type template by introducing fresh region parameters as required by
-%%    the type of each parameter and return value (in addition to the allocation region
-%%    parameter).
-%%  \item Elaborate every \C{new} expression, method call or function application by
-%%    introducing \emph{fresh region variables} (as required by the region type
-%%    template of the called method or function).
-%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
-%%    an undetermined set of constraints over the region parameters of that
-%%    class/method.
-%% \end{enumerate}
-
- \item \emph{Constraint Generation}.
-   In the second step, we analyze the program to generate a set of constraints
-   (over the region identifiers and the predicate variables)
-   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).
-
- \item \emph{Constraint Solving}.
-   We solve the generated set of constraints using our fixpoint constraint
-   solving algorithm \csolvestar, which reduces the constraint solving problem to
-   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
-   references, for example, a reference from a transferable region to a
-   stack region, then the constraints generated during the elaboration
-   are not satisfiable. In such a case, \csolvestar{} fails to solve
-   the constraints.
-% in a Herbrand constraint system, and then relies on \csolve,
-% our abduction solver for that domain.
-
- \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
-  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
-  step 1. We apply these substitutions to the elaborated program to produce the final program.
-\end{enumerate}
-
-\begin{figure}
-\begin{numcodeml}
-Infer ($p$) =
-  let $q$ = IntroduceRegionParameters($p$) in
-  let $(r,C)$ = GenerateConstraints($q$) in
-  match (SolveConstraints($C$) with
-  | None $\longrightarrow$ None
-  | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
-\end{numcodeml}
-
-\caption{The type inference algorithm}
-\label{fig:type-inference-algo}
-\end{figure}
-
-We later on establish the soundness of the type inference algorithm, and also establish
-partial completeness results.
-
-\subsection{Region Parametrization for Classes}
+\vspace*{-0.1in}
+\subsection{Region Type Templates}
 \label{sec:fb-templatization}
 
-Region parametrization is an iterative process involving the following three steps,
-the first two of which are mutually dependent on each other.
-
-\emph{Introduction of Formal Region Parameters}.
-For every class \C{C}, we identify a sequence of formal region parameters
-$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.
-
-\emph{Introduction of Actual Region Parameters}.
-We then replace every instance of class \C{C} in the program by an instance
-$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
-are fresh identifiers denoting actual region parameters.
-
-\emph{Predicate Variable Introduction}. For every class \C{C}, we introduce
-a fresh predicate variable $\varphi$, which represents the yet undetermined
-outlives constraints between the formal region parameters of class \C{C}.
-
-We identify the region parameters of classes as follows.
-
-\emph{Non-Recursive Classes}.
-The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
-The region parameters for any other non-recursive class \C{C} is determined
-only after the region parameters of any class that \C{C} depends on have been
-determined: this includes the base-class \C{B} of \C{C} and the class (type)
-of any of its fields.
-We replace every dependee type \C{T} in \C{C} by its instantiated type,
-using fresh region parameters as needed.
-The sequence of region parameters for \C{C} is defined to be
-the sequence of region parameters for the base class \C{B} concatenated
-with the list of all  fresh region parameters introduced while instantiating the types
-of the fields in the class.
-(The class inherits its allocation region from its base class. Note that if
-a class does not specify an explicit base class, it has an implicit base class
-\C{Object}.)
-
-This transformation is illustrated below, using a non-generic \C{Pair} class:
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
+Region type templates are FGJ types extended with fresh region
+variables ($\rho$) and predicate variables ($\varphi$) to denote
+unknown region annotations and region constraints, respectively. For
+instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
+template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
+region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
+templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
+region-annotated definition of \C{Pair} class from
+\S\ref{sec:fb-syntax} a region type template for a method with FGJ
+type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
+our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
+subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
+accept one region parameter - the allocation region of their
+objects.}\footnote{We abuse arrow notation to also represent types of
+methods, but unlike function types, there is no allocation region
+annotation atop the arrow.}
+$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
+\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
+\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
+region variables, and $\varphi_0$ is a fresh predicate variable
+denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
+$\rho$ is fresh. For a class, a template is computed based on the 
+type templates of its fields, constructors and methods. Additional
+region and predicate variables are added as necessary. Following template 
+of the \C{Pair} class illustrates this point:
+% For a class, a template is computed in two steps. In
+% the first step, we templatize the types of all its fields, constructor
+% arguments, and arguments and return values of all its methods, along
+% with the type of its superclass. In the second step, we generalize all
+% the free region variables occuring in the templates computed in the
+% first step as region parameters of the class. Finally, we add a fresh
+% allocation region parameter ($\rhoalloc$) to the list of parameters,
+% and introduce a new predicate variable ($\varphi$) to denote unknown
+% constraints on region parameters. For example, consider the standard
+% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
+% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
+% respectively. It can be templatized as following :
+\vspace*{-0.1in}
+\begin{center}
 \begin{codejava}
-class Pair
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ fst;
-  $\ObjZ$ snd;
+class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
+          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
+  $a@\rho_0$ fst; $b@\rho_1$ snd;
+  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
+    super(); this.fst=fst; this.snd=snd; 
+  }
+  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
 }
 \end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ fst;
-  $\ObjZ \langle \rho_2 \rangle$ snd;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
+\end{center}
+\vspace*{-0.1in}
+% \footnote{In reality, the generalized region variables only come
+% from the type templates of class's fields and superclass. Method
+% types do not contain any free region variables.}
+% The template elides $\C{Pair}$'s methods, whose region type
+% templates contain no free variables. Among the region parameters of
+% the class template, $\rho_{0-4}$ are obtained by generalizing free
+% region variables in the types of its class's fields, constructor
+% arguments, and its superclass type.  The remaining parameter
+% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
+% region argument. Fresh predicate variable $\varphi_0$ denotes
+% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
+% hold for template to be a well-formed region-annotated class
+% definition in $\FB$. 
+%
+% The type template for a recursively defined class is computed in two
+% steps. First, all recursive occurances of the class among the types of
+% its fields are ignored and the class is templatized as if it is a
+% non-recursive class. Next, all the recursive occurances are
+% templatized with respect to the class template computed in the first
+% step, such that their region annotations are \emph{exactly} same as
+% the region parameters of the class. Applying this technique on a
+% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
+% example, results in a template that uses same region variable to
+% denote allocation regions of the current node, and the node that the
+% current node refers to. 
+%
+% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
+% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
+% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
+% data stored in the linked list node.  To templatize the
+% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
+% occurance in the type of $\C{next}$ field, and templatize the type
+% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
+% template of $\C{data}$ field, we compute the class's template as
+% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
+% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
+% \end{center} Next, we trivially templatize the type of $\C{next}$
+% field in the body of the class as
+% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
+% resulting class represents a linked list with spine in the region
+% $\rhoalloc_0$ and data objects in the region $\rho_0$.
+% Observe that our strategy of trivially templatizing the recursive
+% occurances of a class type in its definition ensures that no new
+% region variables are introduced.
+The type template for for a recursively defined class is computed in
+two steps. First, a template is computed by ignoring the recursive
+occurrences. Next, all the recursive occurrences are templatized based
+on the template computed in first step.  This approach can be extended
+to mutually recursive definitions in a straightforward manner. Using
+the techniques outlined above, we compute region type templates for
+all classes bound in the class table of the $\absof{\FB}$ program
+before we proceed to elaborate expressions.
 
-\emph{Recursive Classes}.
-The region parameters for a recursive class is computed in
-a similar fashion, with the following difference: any recursive
-field is ignored while instantiating region parameters for the fields of
-the class, and the region parameters of the recursive class are computed
-as before. We then do parameter instantiation for all recursive fields,
-such that their region annotations (the actual region parameters) are
-exactly the same as the (formal) region parameters of the class.
-The following example illustrates this for a non-generic \C{List} class.
-The resulting class represents a linked list with spine in the region
-$\rho_0$ and data objects in the region $\rho_1$.
+%\input{fb-elabexpr}
+%\input{fb-elabmeth}
 
-\begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
-\begin{codejava}
-class List
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
+\subsection{Elaboration and Constraint Generation}
 
-The above technique can be extended to mutually recursive classes in a
-straightforward manner, by simultaneously parametrizing them (and
-then instantiating them).
+Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
+(a). replacing core types occurring in expressions with fresh region
+type templates, and (b). explicitly instantiating region parameters
+with fresh region variables in method calls and function applications.
+% This elaboration is performed with respect to the polymorphic type
+% templates of classes and methods computed as per
+% \S\ref{sec:fb-templatization}. 
+Our expression elaboration function\footnote{Supplement contains the
+pseudocode for all the elaboration functions.} traverses expressions
+in a syntax-directed manner of a type checker, introducing fresh
+region type templates for unknown region types, while generating
+constraints over region and predicate variables. The precise nature of
+generated constraints is explained in \S\ref{sec:constraints}, but in
+summary, they capture the relationships between the type templates of
+various subexpressions and their well-formedness.
 
-\emph{Type-Parametric Classes}.
-Type parameters of classes are handled as follows.
-Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
-The parametric type \C{T} is instantiated using the number of region parameters
-that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
-to be \C{Object}, and \C{T} is instantiated with one region parameter.
+% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
+% this elaboration for a subset of expressions in $\absof{\FB}$
+% %, whose corresponding $\FB$ expressions have been ascribed static
+% %semantics in Fig.~\ref{fig:fb-staticsem}. 
+% $\elabExpr$ is defined under the same context as the expression typing
+% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
+% and $\env$ retaining their meaning. The function traverses expressions
+% in a syntax-directed manner of a type checker, introducing fresh
+% region type templates for unknown region types, while generating
+% constraints over region and predicate variables. The precise nature of
+% generated constraints is explained in \S\ref{sec:fb-constraintsem},
+% but in summary, they capture the relationships between the type
+% templates of various subexpressions and their well-formedness. Note
+% that $\elabExpr$ returns the region type template of the
+% subexpression, which is used to generate constraints for the
+% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
+% shown) used by $\elabExpr$ implement type well-formedness and subtype
+% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.
 
-\emph{Function Types}.
-Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
-instantiates function types as follows: the type of every parameter as well as the return value is
-instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
-are generalized as formal region parameters of the function type.
-For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.
+Expression elaboration can be trivially lifted to the level of a
+method, and method elaboration to the level of a class, while
+accumulating constraints. Besides such constraints, elaboration
+functions for methods and classes also generate new constraints by
+checking well-formedness and subtype relationships among the types of
+their constituents. For instance, method elaboration generates a new
+constraint by typechecking the method body against its return type.
+% subtype\footnote{Subtyping is type equality in \FB. In \name,
+% subtyping between region types is a straightforward extension of
+% subtyping between core types.} of method's return type.  
+Likewise, the elaboration function for a class generates a fresh
+constraint that captures a couple of safety conditions: first, the
+allocation regions of objects referred by the instance variables
+should outlive the allocation region of the instance itself, and
+second, the allocation regions of a class type and its superclass type
+must be the same.
 
-%% Likewise, given the region-annotated definition of
-%% \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
-%% method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
-%% is \footnote{In our exposition, we assume that classes $\C{A}$ and
-%% $\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
-%% $\ObjZ$, they accept one region parameter - the allocation region of
-%% their objects.}\footnote{We abuse arrow notation to also represent
-%% types of methods, but unlike function types, there is no allocation
-%% region annotation atop the arrow in a method type.}
-%% $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
-%% \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
-%% \unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
-%% variables, and $\varphi_0$ is a fresh predicate variable denoting
-%% unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
+Elaborating a class table ($CT$) entails elaborating every class
+definition in $CT$, while collecting constraints. The constraints thus
+collected are finally solved to obtain assignments for free region and
+predicate variables introduced during templatization and elaboration
+stages. The solutions are applied to the class table (and to the
+artifacts that make up the class table, recursively) to compute a
+class table that maps classes to their fully region-annotated
+definitions in $\FB$.
 
-\subsection{Region Parametrization for Methods}
+% \subsection{Method and Class Elaboration}
 
-As the next step, we introduce region parameters for every method.
-We do this by instantiating the types of all parameters and the
-return value (of the method) using fresh region identifiers (as explained previously),
-and then generalizing these region identifiers as formal region parameters
-of the method. As for the classes, we also introduce a fresh predicate variable
-$\varphi$ for every method.
+% Elaborating a method involves performingexpression elaboration over
+% its body, while collecting constraints. Method elaboration can be
+% lifted to the level of classes
 
-We then consider every method invocation in the program, and introduce
-fresh region variables representing the (yet unknown) actual region
-parameters for this particular invocation.
-%
-We similarly perform instantiation for every constructor invocation
-of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
-before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
-where $\rho_0, \cdots, \rho_n$ are fresh region variables.
+% Functions $\elabMeth$ and $\elabClass$ shown in
+% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
+% class definitions, respectively. Both functions first build a context
+% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
+% live, a map ($\aenv$) mapping type variables to their bounds, and a
+% constraint formula ($\phicx$) capturing constraints over live region
+% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
+% capture constraints over variables in $\rhoenv$ denoting the fact that
+% such constraints are yet to be inferred. Both elaboration functions
+% generate new constraints by checking well-formedness and subtype
+% relationships among types of their constituents. A fresh constraint
+% ($C_4$) generated by $\elabClass$ captures a couple of safety
+% conditions: first, the allocation regions of objects referred by the
+% instance variables should outlive the allocation region of the
+% instance itself, and second, the allocation regions of a class type
+% and its superclass type must be the same.
 
-\subsection{The Constraint Satisfaction Problem (CSP)}
-\label{sec:fb-constraintsem}
+% Function $\elabMeth$ elaborates a method definition of class $B$. It
+% calls $\elabExpr$ with the context $\A$, its allocation context
+% parameter ($\rhoallocm$), and a type environment ($\env$) that
+% contains region type bindings for all the arguments of the method,
+% including the implicit $\C{this}$ argument. The region type template
+% returned by $\elabExpr$ for the method body is checked against its
+% expected type (derived from the type template of the method)
+% generating more constraints. The function then returns the elaborated
+% method definition and the set of constraints.
 
-The constraint generation algorithm mimics the static type checker, but accumulates
-constraints that must hold for the type checking to succeed.
+% $\elabClass$ elaborates the definition of a class $B$. It relies on
+% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
+% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
+% the set of constraints returned by these functions, $\elabClass$ adds
+% constraints generated by checking the well-formedness of the type
+% templates of its superclass and fields, and also a new constraint
+% capturing a couple of safety conditions: first, the allocation regions
+% of objects referred by the instance variables should outlive the
+% allocation region of the instance itself, and second, the allocation
+% regions of a class type and its superclass type must be the same.
 
-\paragraph{Syntax of Constraints.}
-The constraints are expressed in terms of region identifiers ($\pi$ and $\rho$)
-and predicate variables ($\varphi$).
+% Function $\elabClassTable$ (not shown) elaborates
+% every definition in the class table $CT$, while accumulating
+% constraints. The constraints are finally solved for free region and
+% predicate variables introduced during templatization and elaboration
+% stages. The solutions are applied to the class table (and to the
+% artifacts that make up the class table, recursively) to compute a
+% class table that maps classes to their fully region-annotated
+% definitions in $\FB$.
 
-\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
-A \emph{region variable} is introduced to represent an unknown actual region parameter
-of a method invocation or object alloation, which will be bound to a region constant by the end of the
-type inference.
-A region constant may be 
-(a) a \emph{formal region parameter} of a class or method, or
-(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
-(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
+% Note that if the original program in $\absof{\FB}$ contains unsafe
+% references, for example, a reference from a transferable region to a
+% stack regions, then the constraints generated during the elaboration
+% are not satisfiable. In such case, $\solve$ fails to solve
+% constraints, causing the program to be rejected.
 
-Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
-%% over a set of fixed formal region parameters,
-where the set of region-constraints $\phi$ is defined by:
+\subsection{Constraints}
+\label{sec:constraints}
+
+Three kinds of constraints are generated during elaboration:
+\begin{itemize}
+\item Well-formedness constraints of form $\rho \in \rhoenv$,
+restricting the domain of unification for a region variable ($\rho$)
+to the set ($\rhoenv$) of regions in scope,
+\item Well-formedness constraints of form
+$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
+variable ($\varphi$) to the set of all possible constraint formulas
+over region variables ($\rhoenv$) in scope, and
+\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
+$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
+extended with predicate variables and \emph{pending
+substitutions}\footnote{we borrowed this terminology
+from~\cite{ltpldi08}}:
+\vspace*{-0.05in}
 \begin{smathpar}
 \begin{array}{lcl}
-\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
-\end{array}
-\end{smathpar}
-We will use the term \emph{validity constraint} to denote an entailment constraint
-of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
-
-Our constraints also make uses of \emph{pending substitutions} $F$:
-% \footnote{we borrowed this terminology from~\cite{ltpldi08}}
-A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
-used in a particular context:
-\begin{smathpar}
-\begin{array}{lcl}
+\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
+    \ALT F(\varphi) \ALT \phi \conj \phi\\
 F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
@@
 \end{smathpar}
-E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
-\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
+\vspace*{-0.1in}
+\end{itemize}
+A pending substitution ($F$) is a substitution function over region
+variables/identifiers. They represent the substitutions that need to
+be carried out when a predicate variable ($\varphi$) is replaced by a
+concrete formula in a validity constraint. For instance, in the
+validity constraint $\isvalid{\rgn_1 \outlives
+\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
 is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
 $\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
+$\varphi$ if and only the formula obtained by substituting $\rgn_1$
 and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
 deducible from $\rgn_1 \outlives \rgn_2$.
-
-The constraints generated are of the following kinds:
-\begin{itemize}
-
-\item Well-formedness constraints of form $\rho \in \rhoenv$,
-restricting the domain of unification for a region variable ($\rho$)
-to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
-
-\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
-variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
-regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
-  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
-  the region-constraint of its base class \C{B}).
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
-  (e.g., to indicate the region-constraint of a class \C{C} must imply the
-  region-constraint of each of its fields).
-
-\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
-where $\varphi_i$ is a predicate variable (representing the precondition of a
-method to be determined), $\phicstr$ is a region constraint that is \emph{required}
-to hold at a particular program point (within the method), and $\phictxt$ is
-a region constraint that is \emph{known} to hold at that program point.
-
-%
-%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-%% of predicate variables and pending substitutions. While $\phictxt$
-%% captures relationships that are \emph{known} to hold between concrete
-%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
-%% $\phicstr$ captures relationships that are \emph{required} to hold
-%% among region varibles (i.e., $\rho$'s), or relationships between
-%% region variables and identifiers.
-%
-
-\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
-generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
-$\varphi_i$ are as above).
-
-\end{itemize}
-
-\input{fb-constraint-generation}
-
-\paragraph{Constraint Generation.}
-The constraint generation algorithm is a direct adaption of the type checker:
-each type checking rule is modified to produce a set of constraints that must hold
-for the type checker to succeed.
-%
-Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
-%
-The appendix presents a more detailed algorithm that performs the elaboration and
-constraint generation in a single integrated pass.
-
-\paragraph{Constraint Solution}
-Solving the set ($C$) of constraints entails finding an
-assignment for each predicate variable $(\varphi$) and each region
-variable ($\rho$) that occurs free in $C$, that satisfies all
-the validity constraints as well as the well-formedness constraints on
-$\varphi$ and $\rhobar$.
-% To simplify presentation, we think of $C$ as being parameterized on
-% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-% \rbrack$. We now formalize the constraint satisfaction problem, and
-% its solution.
-
-\begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
-A constraint satisfaction problem is a tuple
-$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-where $\regionConstants$ is a set of region constants,
-$\regionVars$ is a set of region variables,
-$\predVars$ is a set of predicate variables,
-$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies a unification domain for each region variable,
-$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies the domain for each predicate variable,
-and $\constraintSet$ is a set of validity constraints in one of the following forms:
-\begin{center}
-\(
-    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
-    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
-\)
-\end{center}
-The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
-where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
-\begin{itemize}
-  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
-
-  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
-    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
-
-  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
-    \regionSubstFn$.
-  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
-\end{itemize}
-\end{definition}
-
-\paragraph{Properties of generated constraints}
-In the sequel, we will restrict our attention to a special class of CSPs that satisfy
-two interesting properties that allow us to solve the constraints efficiently.
-The constraints generated our type inference algorithm satisfy these two properties.
-
-The first property is that every region variable is guaranteed to be unified with some
-region constant by the constraints\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
-
-The second property is that the \emph{context region constraint} $\phictxt$
-occurring on the antecedent of any validity constraint is a conjunction of
-outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-is a static region identifier and $\pi$ is a region constant (either a formal
-region parameter, or static region identifier, or open transferable region identifier).
-Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
-simplifying the constraint solver.
-
-\subsection{Solving the CSP}
-
-\paragraph{Constraint Solver.}
-Solving the constraints is an iterative process, consisting of the following
-steps, and this process fails at any point if it determines that no solution
-exists for the set of constraints:
-
-\begin{enumerate}
-
-\item
-    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
-$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
-If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
-solver fails.
-
-\item
-When a region variable $\rho$ is unified with a region constant $\pi$,
-if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
-and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
-If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
-
-\item
-\label{item:context}
-Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
-where $\pi_i$ and $\pi_j$ are both region constants.
-If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
-$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
-
-\item
-  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
-  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
-  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
-\end{enumerate}
-
-Point~\ref{item:context} makes use of the special structure of the generated
-constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
-form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
-region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-
-\paragraph{Algorithmic Aspects.}
-The iterative process described above is a standard fixed point computation.
-We can encode this computation using a set of Datalog rules,
-and compute the solution using any Datalog engine.
-Alternatively, algorithms for context-free reachability can be adapted
-to solve the constraints, as explained below.
-
-Given a set $S$ of outlives constraints, we define the directed
-graph $G(S)=(V(S),E(S))$ as follows.
-Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
-which we will also refer to as $\rho$.
-Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
-an edge from $\rho_1$ to $\rho_2$.
-Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
-for the pair of constraints $\rho_1 \outlives \rho_2$ and
-$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
-by a pair of edges.
-%
-It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
-there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
-%
-Thus, simple graph reachability (transitive closure) algorithms can be
-used to identify all the logical consequences of a set of outlives
-constraints.
-
-Given a set of context-free validity constraints, we can generalize the
-above approach and reduce the problem to one of context-free reachability
-in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
-%
-Algorithms for context-free reachability can also be adapted to incorporate
-the above treatment of context constraints $\phictxt$.
-%
-Further details omitted.
-
-\paragraph{Modularity Aspects.}
-The type inference algorithm, as presented, traverses the entire program to
-generate the set of constraints, which are solved en masse, using an iterative
-fixed point computation. However, the type inference can be realized in a
-modular and compositional fashion, subject only to the restrictions imposed
-by recursion.
-
-In the elaboration phase, we can process a class \C{C} only after any class
-\C{B} that \C{C} depends on has been processed: class \C{C} depends on
-class \C{B} if \C{B} is either \C{C}'s base class or the type of any field
-of \C{C} depends on \C{B}. In effect, this means that any collection of
-mutually recursive classes must be processed together. Non-recursive
-dependences can be handled in a compositional fashion: if class \C{C}
-depends on \C{B} non-recursively, then the elaboration can be done for
-\C{B} first, and then \C{C} can be processed.
-
-The same idea applies to the constraint-solving phase as well.
-Given a set of constraints, we say that a predicate variable $\varphi_1$
-\emph{directly-depends} on another predicate variable $\varphi_2$ if the set of
-constraints includes a constraint $\isvalid{\varphi_1 \conj \phictxt}{F(\varphi_2)}$.
-We say that $\varphi_1$ \emph{depends} on $\varphi_2$ if $\varphi_1$ transitively
-depends on $\varphi_2$.
-The constraint solver needs to process any collection of mutually dependent
-predicate variables together.
-In effect, this requires the type inference to process any collection of
-mutually recursive methods together.
-However, methods that are not mutually recursive can be processed separately.
-
-\paragraph{Soundness and Completeness.}
-
-The type inference algorithm is sound:
-
-\begin{theorem}
-\emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
-(1) $\absof{t} = p$, and
-(2) $t$ is well-typed.
-\end{theorem}
-
-Furthermore, the second phase of the type inference algorithm is complete as well.
-If there is any possible instantiation (of the region variables and predicate variables) of
-the elaborated program produced by the first phase that type checks, then the second phase
-will identify it.
-
-Incompleteness exists only in a few choices made during the first phase (elaboration phase), as
-explained below.
-
-\begin{enumerate}
-\item Our technique for determining the set of region parameters for a recursive class
-incorporates a specific heuristic, namely that the recursive occurrence of the class has
-the same parameters, in the same order, as the class itself. This technique will fail, for
-example. if the program uses a recursive list type whose elements alternatively come
-from two different regions. Such a program would require the following elaboration,
-which is beyond the scope of our approach:
-% , $R_1$ and $R_2$.
-% both of which are region parameters of the class.
-\begin{tabular}{ccc}
-\begin{minipage}{0.3\linewidth}
-\begin{codejava}
-class List
-  $\extends$ $\ObjZ$
-{
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.5\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-\item Our technique for region parametrization also uses a heuristic in the case of
-higher order programs. Consider a higher-order method that has a function $f$ as a
-parameter. We have a choice in terms of where the region-parameters introduced for
-$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
-of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
-source of incompleteness in the type-inference.
-
-The treatment of class fields with a function type also uses the same (incomplete)
-heuristic.
-
-\end{enumerate}
-
-Region annotations provided by the user can help the type-inference overcome these
-sources of incompleteness.
-
-\textbf{TODO: Update example discussion}
-
-\subsection{Example}
+% Validity constraints primarily result from checking if a region type
+% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
+% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
+% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
+% formal region parameters with actual region arguments satisfy its
+% precondition, where actuals are subsituted for formals. Since $B$'s
+% precondition could be unknown, the requirement results in a in a
+% validity constraint with predicate variables and pending
+% substitutions. Other sources of validity constraints are calls to
+% region-polymorphic methods, where similar situation arises. 
+% In general, validity constraints generated by our algorithm assume one
+% of the following two forms:
+% \begin{center}
+% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+%            {\phicstr} \qquad
+%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
+%            {F_j(\varphi_j)}
+% \)
+% \end{center}
+% Where $\varphi_i$'s denote the unknown preconditions of the class and
+% the method under elaboration. If the constraint is generated while
+% checking the well-formedness of a type or elaborating an expression,
+% then $\varphi_j$'s denote the unknown preconditions of classes and
+% methods that were used in that type or expression. Each use of a
+% (region-polymorphic) class or a method has to instantiate region
+% parameters, resulting in pending substitutions ($F_j$). Formulas
+% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
+% variables and pending substitutions. While $\phictxt$ captures
+% relationships that are \emph{known} to hold between concrete region
+% identifiers (i.e., $\rgn$'s) when the constraint was generated,
+% $\phicstr$ captures relationships that are \emph{required} to hold
+% among region varibles (i.e., $\rho$'s), or relationships between
+% region variables and identifiers. Each region variable occuring in
+% $\phicstr$ has an associated well-formedness constraint, which
+% specifies its unification domain. The unification domain of a
+% constraint is the union of unification domains of all region variables
+% occuring in the constraint.
 
@@
 contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
-\C{Pair} object \C{q}, and an object \C{y}. It assigns
+\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
 \C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
 recursively with the same region, a new \C{Pair} object allocated in a
-local region, and an object referred by the \C{snd} field of the
+local region, and an \C{A} object referred by the \C{snd} field of the
 pair inside the region. \C{alt} never terminates.  Elaboration phase
@@
 definition\footnote{In reality, elaboration uses new region variables
-as parameters to the method calls, and then generates
+as parameters to the constructor and method calls, and then generates
 constraints that unify them with actuals. In our examples, to avoid
@@
 \begin{codejava}
-unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
-                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
+unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
+                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
   q.fst := y; q.snd := y; 
-  open r as p@$\rgn_0$ in
+  open r as $p@\rgn_0$ in
     letregion $\rgn_1$ in
-      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
+      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
                       (p.fst,p.fst) in
@@
 \end{smathpar}
+% The first constraint is generated by $\typeOk$ on the type of
+% $\C{p}$.  Second and third are generated by the the assignment
+% expressions, and the last constraint is generated from the recursive
+% call.
 
-% \paragraph{Constraints Example 1} The original source of all
-% outlives constraints is the validity constraint $C_3$ in $\elabClass$
-% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
-% class template from \S\ref{sec:fb-templatization}, the constraint is
-% as following:
+\subsection{Constraint Solving}
+\label{sec:csolve}
+
+Although we collect constraints for the entire program during
+elaboration, we solve the constraints in a compositional fashion.
+Concretely, this means that our constraint solving algorithm visits
+and solves every constraint (or, every set of mutually dependent
+constraints) only once. It composes computed solutions to solve other
+constraints that depend on the solved constraints. Importantly, the
+failure to solve a dependent constraint does not result in
+backtracking. We now describe our constraint solving algorithm by
+demonstrating its performance over the example constraints shown in
+the previous section. A formal account of the algorithm can be found
+in the supplement.
+% To simplify the presentation of our algorithm, we assume that are no
+% mutually recursive definitions in the source program. Recursive
+% definitions are nonetheless allowed.
+
+\paragraph{Terminology} In a validity constraint, a predicate variable
+occurring on the left side (resp. right side) of the turnstile is said
+to occur negatively (resp. positively). A validity constraint
+\emph{constrains} the set of predicate variables that occur negatively
+in the constraint, while it \emph{uses} the set of predicate variables
+that occur positively. A constraint is said to be \emph{recursive} if
+it constrains and uses a predicate variable.
+% A pair of validity constraints are said to be \emph{mutually
+% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
+% predicate variables that occur with opposite polarities in both the
+% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
+% predicate variables are also said to be mutually dependent. We call
+% the transitive closure of mutual dependency relation as
+% \emph{transitive dependency} relation. Like mutual dependency,
+% transitive dependency is also extended to predicate variables.
+
+Given a set of validity constraints, we first build a dependency graph
+($G_c$) with constraints as nodes, and dependencies between them
+captured as edges. There exists an edge from a constraint $c_2$ to a
+constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
+any of the following conditions hold:
+\begin{itemize}
+\item \textbf{Type 1} $c_1$ constrains a predicate variable that $c_2$ uses. 
+\item \textbf{Type 2} $c_1$ constrains a (non-strict) subset of predicate variables
+that $c_2$ constrains.
+\end{itemize}
+
+% The first condition intuitively corresponds to a case, where
+% expression or type, whose region elaboration is constrained by $c_2$
+% refers to a method or a class, whose unknown precondition is
+% constrained by $c_1$. The common predicate variable ($\varphi$)
+% represents the unknown precondition in this case. The dependency from
+% $c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
+% before $c_2$ is solved, thus enforcing the rule that the precondition
+% of a method must not depend on its calling context. 
+% Note that the first condition results in self-loops over recursive
+% constraints in the dependency graph.  Since mutually recursive
+% definitions refer each other, the first condition also results in
+% bidirectional dependencies between mutually recursive constraints
+% (and self-loops over recursive constraints). 
+% The second condition captures two kinds of dependencies. First, the
+% dependency from the constraints over a method precondition to the
+% constraints over the precondition of the class containing the method.
+% This captures our preference that the constraints over a class's
+% region parameters should not depend on the idiosyncracies of its
+% methods. Any additional constraints required by any of its methods
+% must be captured in the precondition of the method itself
+% (well-formedness rules allow this possibility). The second condition
+% adds bidirectional dependencies between validity constraints that
+% constraint the same set of predicate variables. 
+Next, we convert the dependency graph over constraints into a
+dependency DAG ($G_C$) over sets of constraints, where each set
+represents a strongly connected component in the dependency graph.
+
+\paragraph{Example} The dependency DAG ($G_C$) over validity
+constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
+shown below: %in Fig.~\ref{fig:pair-deps}. 
+\vspace*{-0.08in}
+\begin{figure}[H]
+\includegraphics[scale=0.6]{DepDAG.png}
+\end{figure}
+\vspace*{-0.08in}
+Each node (labeled $C_i$) is a set of constraints that belong to a
+strongly connected component in the dependency graph ($G_c$), hence
+are mutually dependent. All dependencies, except the self-dependency
+on $\mathbf{c_{10}}$, are type-2 dependencies.
+
+A dependency DAG makes the dependencies between constraints explicit.
+Constraints in each set are mutually dependent, and need to be solved
+simultaneously, whereas constraints in different sets can be solved as
+per any valid topological ordering of the graph's transpose.
+Accordingly, we obtain a topological ordering of nodes in the graph
+$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
+in that order. The solutions obtained after solving a constraint set
+are applied to the constraints in subsequent sets before attempting to
+solve them. For the DAG in figure above, we consider the topological
+order $[C_1, C_2, C_3]$ of its transpose, and solve the sets of
+constraints in that order.
+
+% Consequently, when the turn of a constraint set ($C$) arrives during
+% the constraint solving process, it satisfies certain properties:
+% \begin{itemize}
+% \item There exists only one predicate variable ($\varphi$) that is either
+% constrained or used by the constraints in the set ($C$). The variable is
+% called set's \emph{subject}. This property follows from (a). the fact
+% that all the dependency constraints have already been solved (and
+% solutions applied), and (b). the assumption that there are no mutually
+% recursive definitions. 
+% \item All the constraints that constrain the set's subject are present
+% in the set. This follows from our definition of the dependency relation.
+% \end{itemize}
+
+Solving a set ($C$) of constraints involves finding assignments for
+free region ($\rho$) and predicate ($\varphi$) variables in the set.
+Assignments to region variables can be found via unification (for the
+examples in \S~\ref{sec:constraints}, we have already performed the
+unification, so $C_{1-3}$ do not contain free region variables). Once
+the region variables have been unified, we solve $C$ to find an
+assignment for free predicate variables. The first step is to reduce
+the set of constraints into an equivalent single constraint by
+conjoining antecedents and consequents of all constraints in the
+set\footnote{Supplement contains formal justification for this step}.
+For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
+constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
+below:
+\begin{smathpar}
+\begin{array}{l}
+  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
+     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
+  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
+  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
+    {
+        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
+        \rho_7 = \rho_9 \conj 
+    }\\
+  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
+        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
+\end{array}
+\end{smathpar}
+The constraint $c_{12}$ is a non-recursive constraint of the form
+$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
+$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
+this case). Note that the constraint describes an abduction problem,
+whose solution is the weakest $\phisol$ such that $\isvalid{\phictxt
+\conj \phisol}{\phicstr}$. We solve the problem by casting it as an
+instance of graph augmentation problem~\cite{siam92}. Our intuition is
+based on the observation that a constraint can be viewed as a directed
+graph with region variables as vertices and outlives relationships
+between them as edges (equality is encoded as a conjunction of
+symmetric outlives relationships).  Let $G(\phictxt)$ and
+$G(\phicstr)$ represent such graphs for $\phictxt$ and $\phicstr$,
+respectively. Now, $\isvalid{\phictxt}{\phicstr}$ if and only if every
+pair $(\rho_1,\rho_2)$ of connected vertices in $G(\phicstr)$ are also
+connected in $G(\phictxt)$. If $\rho_1$ and $\rho_2$ are not connected
+in $G(\phictxt)$, then it means that $\isvalid{\phicstr}{\rho_1
+\outlives \rho_2}$ but $\isnotvalid{\phictxt}{\rho_1 \outlives
+\rho_2}$. Since $\phisol$ needs to satisfy $\isvalid{\phictxt \conj
+\phisol}{\phicstr}$, it must be the case that $\rho_1$ and $\rho_2$
+are connected in $G(\phictxt) \cup G(\phisol)$. Since $\phisol$ must
+be weakest such formula, it can be obtained by computing the minimum
+number of edges that need to be added to $G(\phictxt)$ such that it is
+``as connected as'' $G(\phicstr)$. Using this technique, we obtain the
+following solutions to $\varphi_0$ and $\varphi_1$ for constraints in
+the running example:
+\vspace*{-0.05in}
+\begin{center}
+\(
+  \varphi_0 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+     \rho_2 = \rho_0 \conj \rho_3 = \rho_1
+\)\\
+\(
+  \varphi_1 \Leftrightarrow \rho_5 = \rho_0
+\)
+\end{center}
+\vspace*{-0.1in}
+Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
+$c_{14}$ gives us a recursive constraint of the following form:
+\vspace*{-0.05in}
+\begin{center}
+\(
+  \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
+\)
+\end{center}
+\vspace*{-0.05in}
+Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and $F$ is
+a substitution function, not necessarily idempotent. Our solution to the
+constraints of the above form is based on the observation that $H(\varphi) =
+\phicstr \conj F(\varphi)$ is a monotone over the complete meet semi-lattice
+of all possible constraints over the domain of region variables in the
+program. Hence, by Tarski's fixpoint theorem~\cite{tarski}, $H$ has a
+greatest fixed point. To solve the recursive constraint, we first compute the
+greatest fixed point ($\phi_f$) for $H$, and then solve the non-recursive
+constraint $\isvalid{\phictxt \conj \varphi}{\phi_f}$ using the graph
+approach described above. Using this approach, we solve the recursive
+constraint $c_{14}$, to obtain the following solution to $\varphi_2$:
+\begin{center}
+\(
+  \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+  \rho_9 \conj \rho_8 = \rho_9
+\)
+\end{center}
+
+% If such $\phisol$ satisfies the
+% well-formedness condition on $\varphi$, then it is clearly a solution
+% for $\varphi$, because there cannot exist a weaker solution for
+% $\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
+% maximally weak $\phisol$ fails the well-formedness condition, then we
+% establish~\cite{techrep} that it is impossible to satisfy the
+% constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
+% that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
+% and check if it satisfies the well-formedness condition on $\varphi$.
+
+% \vspace*{-0.05in}
+% \paragraph{Definitions} Given a constraint formula $\phi$, we define
+% its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
+% vertices ($V(\phi)$) are free region variables and identifiers in
+% $\phi$, and whose edges ($E(\phi)$) denote outlives constraints
+% in\footnote{We alternate between viewing a constraint formula ($\phi$)
+% as a set of constraints and a conjunction of constraints in this
+% description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
+% \outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
+% $(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
+% conjunction of symmetric outlives constraints for the purpose of graph
+% encoding. Conversely, given a digraph $G$, we define its constraint
+% encoding $\Phi(G)$ in a straightforward manner. We say that a graph
+% $G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
+% \subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
+% exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
+% exist a path between same pair of vertices in $G_1$.
+
+% A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
+% \Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
+% graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
+% connected as $G(\phicstr)$. The problem of computing such a $G$ is
+% equivalent to the problem finding of finding minimum number of edges
+% to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
+% Efficient algorithms to solve the later problem are known to
+% exist~\cite{graph}, and we rely on them for the solution.
+
+% Solving the constraints $c_{12}$ and $c_{13}$ by reducing them to
+% graph augmentation problems, as described above, results in the
+% following solutions for $\varphi_0$ and $\varphi_1$ (symmetric
+% outlives constraints are replaced with equalities):
 % \begin{center}
-%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
-%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
+% \(
+%   \varphi_1 \Leftrightarrow \rho_0 \outlives \rhoalloc_0 \conj \rho_1
+%      \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
+%      \rho_2 = \rho_0 \conj \rho_3 = \rho_1
+% \)\\
+% \(
+%   \varphi_2 \Leftrightarrow \rho_5 = \rho_0
+% \)
+% \end{center}
+% Substituting the above solutions for $\varphi_0$ and $\varphi_1$ in
+% $c_{14}$ gives us a recursive constraint of the following form:
+% \begin{center}
+% \(
+%   \isvalid{\phictxt \conj \varphi}{\phicstr \conj F(\varphi)}
+% \)
+% \end{center}
+% Where $\phictxt$ and $\phicstr$ are concrete constraint formulas, and
+% $F$ is a substitution function, not necessarily idempotent. Let
+% $G(\varphi) = \phicstr \conj F(\varphi)$, so that the constraint is
+% $\isvalid{\phictxt \conj \varphi}{H(\varphi)}$. Our solution to such
+% recursive constraints  is based on some observations. Firstly, the set
+% of all possible constraints over a finite set of region
+% variables/identifiers is a complete meet semi-lattice, where (a).
+% meet operation is defined as $\phi_1 \sqcap \phi_2 \triangleq \phi_1
+% \conj \phi_2$, and (b). lattice order is defined as $\phi_1 \le \phi_2
+% \triangleq \isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Secondly,
+% function $H$ is a monotone over the lattice since $\forall
+% \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow \phi_2)
+% \Rightarrow (H(\phi_1) \Rightarrow H(\phi_2))}$. From Knaster-Tarski's
+% theorem, we know that $H$ has a greatest fixed point $\phi_f$ such
+% that $\isvalid{\cdot}{\phi_f \Leftrightarrow H(\phi_f) }$. To solve
+% the recursive constraint shown above, we first compute the greatest
+% fixed point ($\phi_f$) of $H$, and then reduce the constraint to the
+% following non-recursive constraint:
+% \begin{center}
+% \(
+%   \isvalid{\phictxt \conj \varphi}{\phi_f}
+% \)
+% \end{center}
+% Which can then be solved via the graph-based approach described
+% before.
+
+% Let $H(\varphi_2)$ denote the consequent of the recursive constraint
+% $c_{14}$. Its greatest fixed point ($\phi_f$) is shown below:
+% \begin{center}
+% \(
+%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+%   \rho_9 \conj \rho_8 = \rho_9 \conj \rgn_0 \outlives \rgn_1 \conj
+%   \rgn_0 = \rgn_0
+% \)
+% \end{center}
+% Computing a maximally weak $\phisol$ such that $\isvalid{\varphi_1
+% \conj \rgn_0 \outlives \rgn_1}{\phisol \Leftrightarrow \phi_f}$
+% results in the following solution for $\varphi_2$ that meets its
+% well-formedness requirement ($\tywf{\{\rhoalloc_0,\rho_{}0-4,
+% \rhoalloc_2, \rho_{6-9}\}}{\varphi_2} $):
+% \begin{center}
+% \(
+%   \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj \rho_7 =
+%   \rho_9 \conj \rho_8 = \rho_9
+% \)
 % \end{center}
 
-% \paragraph{Constraints Example 2} Consider a contrived method
-% $\C{alt}$ that accepts an argument $\C{p}$ of type
-% $\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
-% of type $A$. It then assigns one object to $\C{p.fst}$ and other to
-% $\C{p.snd}$, and calls itself recursively without ever terminating.
-% Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
-% alternated between recursive calls. The region-annotated definition of
-% $\C{alt}$ is shown below:
+\subsection{Higher-Order Type Inference}
+
+% The region type system for $\FB$ admits function types with
+% higher-rank region polymorphism, allowing region variables to be
+% quantified in the types of higher-order arguments. This allows, for
+% example, a user-defined function to be applied on arguments from
+% different regions in different contexts. However, like most type
+% systems, our region type system loses principal typing property in
+% presence of higher-rank polymorphism, which means that there can be
+% two possible types for a function, where neither one is more general
+% than the other. Consequently, it is not possible to infer a unique
+% type for a function. Nonetheless, it is possible to engineer type
+% inference to to consistently prefer one type over the other.
+Our type inference algorithm does not straightforwardly generalize to
+the higher-order case, due to the lack of principal typing in presence of
+higher-rank polymorphism. Fortunately, higher-order type inference, in
+its full generality, is rarely needed in the context of \name, where
+higher-order control flow is predominantly due to the second-order
+arguments to constructors (For \eg, user-defined functions in dataflow
+queries, and lambda arguments to the $\RgnZ$ class constructor). We
+therefore extend our inference algorithm with a heuristics tailor-made
+for the aforementioned second-order cases. The simplistic approach
+seems to suffice in our experience, but, it is always possible that it
+makes the type checker reject a program that is otherwise valid. In
+such cases, programmer can help the type checker by writing region
+annotations only over higher-order arguments.  Nonetheless, we note
+that it is possible to infer most general (if not principal) types for
+higher-order arguments, by extending our constraint solver with an
+approach modeled after~\cite{gulwani09}, which lets it compute maximally
+weak/strong solutions to multiple competing $\varphi$'s
+simultaneously.
+
+\dv{I believe it is actually worth explaining the problem, as well as what our heuristic is, at a level
+where the readers might understand that this is challenging and get convinced that our heuristic
+is a good one. After all, although we undermine higher-orderness, in data processing systems its
+all about those second order functions.}
+
 % \begin{codejava}
-% unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
-%                           A<$\rho_3$> x, A<$\rho_4$> y) {
-%   return p.fst = x; p.snd = y; 
-%          this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
-% }
+% bool app2(A x, A y, (A $\rightarrow$ bool) f) 
+%     { return (f x && f y); }
 % \end{codejava}
-% Note that, to avoid clutter, we have already resolved appropriate
-% region arguments to the recursive call, instead of introducing new
-% region variables and generating equality constraints on them. Rest of
-% the validity constraints generated while elaborating $\C{alt}$ to the
-% above region-annotated definition are shown below:
+% There are two possibilities for the region type of \C{app2}: one,
+% where it allows \C{x} and \C{y} to be allocated in different regions,
+% but requires \C{f} to be region-polymorphic (i.e, \C{f} should accept
+% arguments from any region), and second, where it requires \C{x} and
+% \C{y} to be allocated in the same region, and allows \C{f} to be
+% monomorphic. None of the two types are more general than the other,
+% and it is not clear which one is more preferable. The concrete
+% manifestation of this dilemma is a constraint of form 
+% $\isvalid{\phictxt \conj \varphi_1}{\phicstr \conj \varphi_2}$, with
+% no additional constraints to help determine $\varphi_1$ or $\varphi_2$
+% before solving this constraint. Since it is always possible to
+% strengthen $\varphi_2$ and then find a suitable $\varphi_1$, there
+% does not exist unique solution to such constraints. Nonetheless, it is
+% possible to bias the constraint-solving algorithm to consistently
+% prefer one solution over the other.
+
+% In context of \name, however, we adopt a trivial solution of
+% assigning the same region to the arguments and return values in
+% (higher-order) function types. For user-defined functions (e.g.,
+% predicate in the \C{WHERE} clause of the \C{SELECT} dataflow query)
+% and lambda arguments to the $\RgnZ$ class constructor, which are the
+% prominent uses of higher-order functions in \name, the simplistic
+% approach seems to suffice in practice. However, it is always possible
+% that this approach makes type checker reject a program that is
+% otherwise valid. In such cases, programmer can help the type checker
+% by writing region annotations for only the higher-order functions.
+
 % \begin{center}
-% \(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
-% \(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
+%   \C{app2} : $\inang{\rho_0,\rho_1}$(\C{A}$\inang{\rho_0}$, 
+%               \C{A}$\inang{\rho_0}$,
+%               $\inang{\rho_2}$\C{A}$\inang{\rho_2}$ $\rightarrow$
+%               \C{bool}) $\rightarrow$ \C{bool}
 % \end{center}
-% The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
-% Second and third are generated by the the assignment expressions, and
-% the last constraint is generated from the recursive call.

### 2016-07-03 paper/implementation.tex
Comments on un-safety detection.

@@
 
+\dv{Hmmmm what is \C{inMsg.payload[i]} in this case? Is it an object? The example
+  demonstrates that we should really have passed in a region that stored other
+  transferrable regions inside, i.e. we should proably make the \C{inMsg.payload[i]} a
+  region itself. This way we don't have to pay the cost of clone. I think that might be
+  an interesting additional discussion in this paragraph. Otherwise I am left as a reader
+  with this bittersweat feeling that ``ok you fixed it but you introduced a copy''. But if
+  the payload was also on it own transferrable region we would be fine right? We probably
+  need a \C{GiveUp} primitive of a region too as we will transfer only the root transferrable
+  region (\C{inMsg}) but we want to declare that we've given up ownership on all the nested
+  ones too.}
+
 Our experience with Naiad benchmarks suggests that, although our
-approach cannot statically enforce memory safety, it is nonetheless
-useful, particularly because the static verification comes at no
+approach cannot statically enforce memory safety, \dv{Valid point,
+  also worth re-visiting in the light of the above?}
+it is nonetheless useful, particularly because the static verification comes at no
 additional cost to the developer. If the developer is satisfied with

### 2016-07-04 paper/type-inference.tex
theorem for type inference?
constraint generation

--- 2016-07-02 paper/type-inference.tex
@@
 %
-Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
+Fig.~\ref{fig:constraint-gen} illustrates this for a couple of rules.
+The algorithm proceeds top-down, analyzes an expression $e$ in a context
+$\exptycx{\ralloc}{\env}$, and returns a type $\tau$ and a set of constraints $C$
+(expressed in the rules as $\exprok {\exptycx{\ralloc}{\env}} {e} {\tau} {C}$ ),
+indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.
 %

### 2016-07-04 paper/type-inference.tex
added let-region constraint generation rule
Removed elaboration from appendix

@@
 for the type checker to succeed.
-%
-Fig.~\ref{fig:constraint-gen} illustrates this for a couple of rules.
+
+Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
+These rules use the same context as the expression typing judgment in
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
+retaining their meaning.
 The algorithm proceeds top-down, analyzes an expression $e$ in a context
@@
 indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.
-%
-The appendix presents a more detailed algorithm that performs the elaboration and
-constraint generation in a single integrated pass.
+
+The rules for generating constraints from a method and
+class definition first build a context ($\A$) containing a set ($ \rhoenv$) of region variables denoting
+regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ denoting the fact that such constraints are yet
+to be inferred.
 

### 2016-07-05 paper/implementation.tex
More comments from DV.
minor changes

--- 2015-11-21 paper/implementation.tex
@@
 closely follows the description given in
-Sec.~\ref{sec:type-inference}. The one difference is that our
-constraint solver uses an approximation algorithm to compute the
-minimal graph augmentation (\S~\ref{sec:csolve}).
+Sec.~\ref{sec:type-inference}.
+
 % To solve the
@@
 
-Our experience with Naiad benchmarks suggests that, although our
-approach cannot statically enforce memory safety, it is nonetheless
-useful, particularly because the static verification comes at no
-additional cost to the developer. If the developer is satisfied with
+Our experience with Naiad benchmarks shows the utility of our
+type inference/checking tool,
+%suggests that, although our
+%approach cannot statically enforce memory safety, it is nonetheless
+%useful, 
+particularly because it comes at no additional cost to the developer.
+If the developer is satisfied with
 the confidence gained by static verification, she may even choose to

### 2016-07-05 paper/related.tex
minor changes

@@
 In particular, our approach circumvents the need for restrictive
-static mechanisms (e.g., linear types and unique pointers) and
+static mechanisms (e.g., linear types and unique pointers) or
 expensive runtime mechanisms (e.g., garbage collection and reference
@@
 Tofte and Talpin’s, and it’s effectiveness in presence of tracked
-pointers is not clear. In contrast, the complete Ocaml (pseudo) code
-of Broom's inference algorithm, was given in the supplement and the
-ideas underlying type inference have been described elaborately in the
-paper
+pointers is not clear.
+%In contrast, the complete Ocaml (pseudo) code
+%of Broom's inference algorithm, was given in the supplement and the
+%ideas underlying type inference have been described elaborately in the
+%paper
 

### 2016-07-05 paper/type-inference.tex
minor changes

@@
 The constraint generation algorithm is a direct adaption of the type checker:
-each type checking rule is modified to produce a set of constraints that must hold
+each type checking judgment is modified to produce a set of constraints that must hold
 for the type checker to succeed.
@@
 Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
-These rules use the same context as the expression typing judgment in
+These rules use the same context as the corresponding typing judgment in
 Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
@@
 to be inferred.
+
+One notable aspect here is the following. Consider a method with predicate
+variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
+We use $\varphi_m$ as the constraint formula in the initial context used for
+the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
+It is possible to show that there is no loss of generality in doing so.
+Specifically, it can be shown for any well-typed program, replacing the constraint
+$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
+where $\varphi_c$ is the constraint of the class containing the method,
+produces a well-typed program.
+%
+This simplification produces a restricted form of constraints that is easier to solve.
 
@@
 \item Our technique for region parametrization also uses a heuristic in the case of
-higher order programs. Consider a higher-order method that has a function $f$ as a
-parameter. We have a choice in terms of where the region-parameters introduced for
-$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
-of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
-source of incompleteness in the type-inference.
+higher order programs. The following examples illustrates that principal types may
+not exist for higher order functions.
+\begin{codejava}
+unit apply ( T $\rightarrow$ unit f, T x, T y) {
+   f(x); f(y);
+}
+\end{codejava}
+This method may be typed assuming either that \C{f} is polymorphic over the region
+that its parameter is allocated in (permitting \C{x} and \C{y} to be allocated in any
+regions), or by assuming that \C{x} and \C{y} are allocated in the same region $\rho_1$
+that \C{f} expects its parameters to be allocated in. Neither type subsumes the other.
+Our algorithm heuristically chooses the first option, as it appears to be the more likely
+and useful candidate.
 
-The treatment of class fields with a function type also uses the same (incomplete)
-heuristic.
+%as either\\
+%$
+%\inang{ \rho_0 \rho_1 \rho_2}
+%    (\inang{\rho_0 \rho_3} (T \inang{\rho_3} \xrightarrow{} unit, 
+%      T \inang{\rho_1}, 
+%      T \inang{\rho_2})
+%   \xrightarrow{} 
+%      unit
+%$\\
+%or as\\
+%$
+%\inang{ \rho_0 \rho_1}
+%    (\inang{\rho_0} (T \inang{\rho_1} \xrightarrow{} unit, 
+%      T \inang{\rho_1}, 
+%      T \inang{\rho_1})
+%   \xrightarrow{} 
+%      unit
+%$\\
+
+%Consider a higher-order method that has a function $f$ as a
+%parameter. We have a choice in terms of where the region-parameters introduced for
+%$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
+%of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
+%source of incompleteness in the type-inference.
+
+% The treatment of class fields with a function type also uses the same (incomplete) heuristic.
 

### 2016-07-05 paper/type-system.tex

--- 2016-07-02 paper/type-system.tex
@@
 %
-The following definition of \C{Pair} class in \FB illustrates some
-of the key elements of the formal language\footnote{The symbol
-$\extends$ should be read \emph{extends}, and the symbol $\outlives$
-stands for \emph{outlives}}:
+The following definition of \C{Pair} class in \FB illustrates some of
+the key elements of the formal language (the symbol $\extends$ should
+be read \emph{extends}, and the symbol $\outlives$ stands for
+\emph{outlives}):
 \begin{codejava}[mathescape=true]
@@
 % \vspace*{-0.15in}
-
 %  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
@@
 %  }
+Note that we elide showing constructors since they are uninteresting
+from the type system's standpoint; their behavior in \FB is same as in
+FGJ.
 
-A class in \FB is parametric over zero or more type variables (as in
-FGJ) as well as one or more region variables $\rho$.
+A class in \FB is parametric over zero or more type variables (as
+in FGJ) as well as one or more region variables $\rho$.
 % In the above example, the class \C{Pair} is parameterized over
@@
 % context. Every class definition in \FB is necessarily polymorphic with
-% respect to the allocation region of its objects, i.e., the allocation
+% respect to the allocation region of its objects, i.e., the
+% allocation
 % context of its constructor. We adopt a convention that requires the
@@
 % method in \FB. 
-Note that if a method is not intended to be polymorphic with respect
-to its allocation context (for example, if its allocation context
-needs to be same as the allocation region of its \emph{this}
-argument), then the required monomorphism can be captured as an
-equality constraint in $\phi$.
+If a method is not intended to be polymorphic with respect to its
+allocation context (for example, if its allocation context needs to be
+same as the allocation region of its \emph{this} argument), then the
+required monomorphism can be captured as an equality constraint in
+$\phi$.
 
@@
 \item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
-variables to their bounds\footnote{A bound of a type variable
-($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
-was declared to extend.}, and
+variables to their \emph{bounds}, i.e., classes they are declared to
+extend, and
 \item A constraint formula ($\phicx$) that captures the outlives
@@
 ($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
-judgments on core types \footnote{We use a double-piped turnstile
-($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
-($\vdash$) for those in \FB.}. The class table ($\absof{CT}$) for such
-judgments is derived from \FB's class table ($CT$) by erasing all
-region annotations on types, and region arguments in expressions
+judgments on core types. We use a double-piped turnstile ($\Vdash$)
+for judgments in FGJ~\cite{fgj}, and a simple turnstile ($\vdash$) for
+those in \FB. The class table ($\absof{CT}$) for FGJ judgments is
+derived from \FB's class table ($CT$) by erasing all region
+annotations on types, and region arguments in expressions
 ($\absof{\cdot}$ denotes the region erasure operation). 
@@
 parameters. The later is enforced by checking the validity of $\phi$,
-with actual region arguments substituted\footnote{The notation
-$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
+with actual region arguments substituted for formal
 region parameters, under the conditions ($\A.\phicx$) guaranteed by
@@
 objects of  the \RgnZ class, and \C{new} expressions that create
-objects of other classes. The rule for the later relies on an
+objects of other classes. The rule for the latter relies on an
 auxiliary definition called $\fields$ that returns the sequence of
@@
 preventing cross-region references originating from transferable
-regions\footnote{The body of the function ($e$) might, however, create
+regions. The body ($e$) of the function might, however, create
 new (transferable) regions while execution, but that is fine as long
 as such regions, and objects allocated in them, don't find their way
-into the result of evaluating $e$}.
+into the result of evaluating $e$.
 

### 2016-07-06 paper/type-inference.tex
Updated section 4.2 function type parametrization

@@
 \emph{Function Types}.
-Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
-instantiates function types as follows: the type of every parameter as well as the return value is
-instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
-are generalized as formal region parameters of the function type.
-For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.
+Since \FB{} is higher-order, fields of function type are allowed.We explain how the parameter instantiation step
+instantiates function types below, after discussing parametrization for methods.
 
-%% Likewise, given the region-annotated definition of
-%% \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
-%% method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
-%% is \footnote{In our exposition, we assume that classes $\C{A}$ and
-%% $\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
-%% $\ObjZ$, they accept one region parameter - the allocation region of
-%% their objects.}\footnote{We abuse arrow notation to also represent
+%% \footnote{We abuse arrow notation to also represent
 %% types of methods, but unlike function types, there is no allocation
@@
 
-\subsection{Region Parametrization for Methods}
+\subsection{Region Parametrization for Methods and Function Types}
 
@@
 and then generalizing these region identifiers as formal region parameters
-of the method. As for the classes, we also introduce a fresh predicate variable
-$\varphi$ for every method.
+of the method. In addition, a fresh region identifier is introduced to represent the
+allocation region.
+As for the classes, we also introduce a fresh predicate variable $\varphi$ for every method.
+Thus, the method
+
+\C{Object m (List x) \{...\} } \\
+is instantiated as
+
+\C{Object$\inang{\rho_3}$ m$\inang{\rho_0,\rho_1,\rho_2,\rho_3}$ (List$\inang{\rho_1,\rho_2}$ x) \{...\} }
 
@@
 where $\rho_0, \cdots, \rho_n$ are fresh region variables.
+
+Function types (of fields and parameters) are instantiated just like methods above, except that
+an additional fresh region identifier is introduced to represent the region where the closure is
+allocated.
+For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
+$\inang{\rho_0, \rho_1, \rho_2, \rho_3, \rho_4 \,|\, \varphi } \C{List} \inang{\rho_1,\rho_2} \xrightarrow{\rho_4} \C{Object}\inang{\rho_3}$.
+Note that the newly introduced region identifiers are generalized as formal region parameters of the function type.
+This is a (heuristic) choice made in the case of higher order functions.
+Consider a higher order function \emph{f} with a function typed parameter $g$.
+The fresh region identifiers introduced while instantiating the type of $g$ could be
+alternatively generalized as formal region parameters of $f$, but we choose to
+generalize them as formal region parameters of $g$.
 

### 2016-07-06 paper/introduction.tex
updated intro (rust, scala, trill)

--- 2016-07-01 paper/introduction.tex
@@
 suitable timing message arrives.
+There has been increasing interest in off-heap memory management for better performance
+in the context of several  systems and languages (e.g., Spark~\cite{SPARK:2015}, Scala, Rust).
+Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in
+the absence of language support.
 
@@
 \paragraph{Safety.}
-These systems are designed to allow the producer and consumer to
+Several of these systems are designed to allow the producer and consumer to
 execute in the same address space or in different address spaces
@@
 
+\paragraph{Language.}
 The cornerstone of this approach is an \C{open} lexical block for transferable regions,
 that ``opens'' a transferable region and guarantees that the region
-won't be transferred/freed while it is open. By nesting an
-MLKit-style \C{letregion} lexical block, that delimits
+won't be transferred/freed while it is open. By nesting a
+Tofte-Talpin style \C{letregion} lexical block, that delimits
 the lifetime of a stack region, inside an \C{open} lexical block for a
@@
 former \emph{outlives} the later., and any references from the stack
-region to the transferable region are therefore safe.  Next, we note
-that by controlling the outlives relationships between various
+region to the transferable region are therefore safe.
+This is particular useful, since such stack regions serve as temporary
+working storage while working with the (opened) transferable regions.
+
+By controlling the outlives relationships between various
 regions, we only allow safe cross-region references, while prohibiting
@@
 
+\paragraph{Type System.}
 Formally, our type system introduces \emph{region parameters} (for both
@@
 
+\paragraph{Lightweight Runtime Checks.}
 Ensuring memory safety using this approach requires ensuring
@@
 
+\paragraph{Type Inference.}
 One of the novel contributions of this paper is a sound type inference algorithm

### 2016-07-06 paper/type-inference.tex
updated intro (rust, scala, trill)

@@
 generalize them as formal region parameters of $g$.
+We will discuss this aspect again later.
 
@@
 two interesting properties that allow us to solve the constraints efficiently.
-The constraints generated our type inference algorithm satisfy these two properties.
+The constraints generated by our type inference algorithm satisfy these two properties.
 
@@
 
-Region annotations provided by the user can help the type-inference overcome these
-sources of incompleteness.
+If users provide partial region annotations, especially in situations (such as above) where elaboration
+makes a heuristic choice, the elaboration procedure can use the user-provided choices instead.
+This can help the type-inference overcome these sources of incompleteness.
 

### 2016-07-06 paper/related.tex
updated to related work section

@@
 MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
-existing proposals in the following ways. Firstly, our problem setting
-includes lexically scoped stack regions and dynamic transferable
-regions (both programmer-managed) in context of an object-oriented
-programming language equipped with higher-order functions. Second, we
-adopt a two-pronged approach to memory safety that relies on a
+existing proposals in one or more of the following respects.
+
+\begin{enumerate}
+\item 
+   Our design choice focuses on ensuring memory-safety while giving programmers
+   control over  region management and allocation of objects in regions.
+   In constrast, some systems automate all aspects of memory management.
+   This is a convenience-performance tradeoff.
+
+\item We support both lexically scoped (stack) regions and dynamic transferable
+regions (both programmer-managed).
+
+\item We adopt a two-pronged approach to memory safety that relies on a
 combination of a simple static type discipline and lightweight runtime
-checks. 
-% The type system reduces the problem of ensuring safety of all memory
-% accesses in a program into a problem of enforcing a state-transition
-% discipline on memory accesses over a small subset of objects in the
-% program. The later can be done efficiently at runtime. This lets us
-% guarantee safety in presence of
-In particular, our approach circumvents the need for restrictive
+checks. In particular, our approach circumvents the need for restrictive
 static mechanisms (e.g., linear types and unique pointers) or
 expensive runtime mechanisms (e.g., garbage collection and reference
-counting) in order to guarantee safety. Lastly, our region type system
-comes equipped with full type inference that eliminates the need to
-write region annotations on types to convince the type checker that
-the program is safe.
+counting) in order to guarantee safety.
 
-~\cite{tofte97} elaborates SML programs to introduce stack regions and
-do away with GC in a transparent fashion. We too define an
-elaboration, but our focus is on introducing region annotations
-necessary to prove the safety of an object-oriented program that
-already uses regions.
+\item We present a full (interprocedural) type inference algorithm
+that eliminates the need to write region annotations on types.
+
+\item Our underlying language is an object-oriented programming language,
+equipped with higher-order functions and parametrized (generic) types.
+These language features necessitate some non-trivial choices in the design
+of the region-parametricity aspect of the language, which also have an
+impact on aspects such type inference.
+
+\end{enumerate}
+
+Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed lexically
+scoped (stack) regions (as a replacement for GC).
 % Similar to their region inference algorithm, our region type inference
@@
 % fixed points for recursive constraints. 
-Our type inference is analogous to their region inference. While their
-inference algorithm only ever generates equality constraints, which
+Our type inference is analogous to their region inference in some respects,
+while differing in others. While their inference algorithm only generates equality constraints, which
 can be solved via unification, our type inference algorithm also
 generates partial order outlives constraints, which are required to
-capture subtle relationships between lifetimes of transferable regions
+capture relationships between lifetimes of transferable regions
 and stack regions. Consequently, our constraint solving algorithm is
@@
 system (FGJ). The safety guarantees have been formalized and proved.
-The core of Broom is very simple; the rules that make up static and
-dynamic semantics occupy less than a page each. We believe that the
-rigor and simplicity of Broom makes it easy to understand the the
-underlying ideas, and apply them in various problem settings. Similar
-contrast can be made of region type inference in both the languages.
+%The core of Broom is very simple; the rules that make up static and
+%dynamic semantics occupy less than a page each. We believe that the
+%rigor and simplicity of Broom makes it easy to understand the the
+%underlying ideas, and apply them in various problem settings.
+Similar contrast can be made of region type inference in both the languages.
 Cyclone’s type inference was only ever described as being similar to
@@
 pointers is not clear.
+In contrast, a detailed type inference algorithm is one of our core contributions.
 %In contrast, the complete Ocaml (pseudo) code

### 2016-07-06 paper/type-inference.tex
explained generation of constraints with one variable

@@
 produces a well-typed program.
-%
-This simplification produces a restricted form of constraints that is easier to solve.
+
+We use the above simplification to avoid generating validity constraints of the form
+$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
+with more than one predicate variable on the left hand side, which introduce complications).
+%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.
 

### 2016-07-06 paper/related.tex
moved discussion of walker and watkins

@@
 constraints over region arguments of polymorphic recursive functions.
+
+Walker and Watkins~\cite{WW01} extend lambda calculus with first-class regions
+with dynamic lifetimes, and imposes linear typing to control accesses to regions.
+Our open/close lexical block for transferable regions traces its orgins to the \C{let!}
+expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes
+linear typing restrictions, allowing variables to be temporarily
+aliased.
+We don't use linear typing (for references to regions), thus admit unrestricted aliasing,
+but use lightweight runtime checks for safety.
+% Non-linear typing keeps our type system simple, and makes it
+% possible to infer types, thus eliminating the annotation burden.
+Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
+the invariants needed to ensure safety under region transfers, such as
+the absence of references that escape a transferable region.
 
@@
 not first-class values of the language, and reference counting is
-nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
-calculus with first-class regions with dynamic lifetimes, and imposes
-linear typing to control accesses to regions. Our open/close lexical
-block for transferable regions traces its orgins to the \C{let!}
-expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes
-linear typing restrictions, allowing variables to be temporarily
-aliased. We don't have linear typing, thus admit unrestricted
-aliasing. 
-% Non-linear typing keeps our type system simple, and makes it
-% possible to infer types, thus eliminating the annotation burden.
-Moreover, ~\cite{WW01}'s linear type system is insufficient to enforce
-the invariants needed to ensure safety under region transfers, such as
-the absence of references that escape a transferable region.
+nonetheless needed to ensure memory safety.
 

### 2016-07-07 paper/type-inference.tex
formatting and multiply defined label

@@
 \begin{figure}
-\begin{numcodeml}
+\begin{codeml}
 Infer ($p$) =
@@
   | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
-\end{numcodeml}
+\end{codeml}
 

### 2016-07-07 paper/type-system.tex
Fixed bug in type safety theorem stmt.
minor edits (comparing languages of section 2 and 3)
formatting and multiply defined label

@@
 language. 
+(The language used in Section~\ref{sec:overview} is essentially
+a version of FGJ without region types and with some syntactic sugar.)
 
@@
 $e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
-and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
+and $\hastyp{\emptyASigp,\rgn, \cdot}{e'}{\tau}$.
 \end{theorem}
@@
 \emph{(\textbf{Transfer Safety})}
-\label{thm:fb-type-safety}
+\label{thm:fb-transfer-safety}
 $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,

### 2016-07-07 paper/introduction.tex
Added scala link. Explained T@R advantages.

@@
 There has been increasing interest in off-heap memory management for better performance
-in the context of several  systems and languages (e.g., Spark~\cite{SPARK:2015}, Scala, Rust).
+in the context of several  systems and languages (e.g.,
+Spark~\cite{SPARK:2015}, Scala~\cite{ScalaOffHeap}, Rust).
 Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in

### 2016-07-07 paper/type-system.tex
Added scala link. Explained T@R advantages.

@@
 region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
-$R_2$ is guaranteed to outlive $R_1$. 
-We now formally develop this intuition via \fbname (\FB), our
-explicitly typed core language (with region types) that incorporates
-the features introduced in the previous section. \fbname builds on the
-Featherweight Generic Java (FGJ)~\cite{fgj} formalism, and reuses
-notations and various definitions from~\cite{fgj}, such as the
-definition of type well-formedness for the core (region-free)
-language. 
-(The language used in Section~\ref{sec:overview} is essentially
-a version of FGJ without region types and with some syntactic sugar.)
+$R_2$ is guaranteed to outlive $R_1$.  We now formally develop this
+intuition via \fbname (\FB), our explicitly typed core language (with
+region types) that incorporates the features introduced in the
+previous section. \fbname builds on the Featherweight Generic Java
+(FGJ)~\cite{fgj} formalism, and reuses notations and various
+definitions from~\cite{fgj}, such as the definition of type
+well-formedness for the core (region-free) language.  (The language
+used in Section~\ref{sec:overview} is essentially a version of FGJ
+without region types and with some syntactic sugar.)
 
@@
 type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
-we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
+we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, we treat
 $B\inang{\tbar}@\rgn$ as being equivalent to
-$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
+$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
 $a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
 instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
-\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
-\FB, hence it has no region annotations.
+\C{Pair} object contained in  $\rgn$. Separate parameterization over
+types and regions is needed in \FB to support region-polymorphic
+higher-order functions (explained below) as class fields. Class fields
+are not allowed to be type polymorphic, in keeping with C\#, which
+constitutes the core of \name. For class fields that are functions,
+regions are parameterized at the (class) field level whereas types are
+parameterized at the class level.  This choice also simplifies type
+inference in the presence of higher-order functions. The type $\unitZ$
+is unboxed in \FB, hence it has no region annotations.
+
 % The region type of $\ObjZ$ class in \FB has no fields, hence its
@@
 \emph{(\textbf{Transfer Safety})}
-\label{thm:fb-transfer-safety}
+\label{thm:fb-type-safety}
 $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,

### 2016-07-07 paper/overview.tex
Typos fixed.

--- 2015-11-12 paper/overview.tex
@@
 Stripped of its region-specific constructs, the core language of \name
-is a familar subset of \csharp that includes features such as
+is a familiar subset of \csharp that includes features such as
 class-based inheritance, dynamic dispatch, parametric polymorphism,
@@
 Simplest of the region-specific constructs in \name is a \C{letregion}
-lexical block that creates a new memory region adressable via its
+lexical block that creates a new memory region addressable via its
 static identifier. The semantics of the \C{letregion} block is similar
@@
 The identifiers \C{R0} and \C{R1} are bound to new stack regions each
-time the loop is evaulated. Nonetheless, the propositions (a). $\C{R0}
+time the loop is evaluated. Nonetheless, the propositions (a). $\C{R0}
 \outlives \C{R1}$, and (b). $\C{a1}:\C{A@R1}$ (read as \emph{\C{a1}
@@
 A transferable region starts its lifetime in a \emph{closed} state,
-when it is created through \C{new Region} expresion. \name provides
+when it is created through \C{new Region} expression. \name provides
 \C{open} lexical construct to \emph{open} a transferable region as an
@@
 
-\name chooses a reasonable tradeoff. Regions are coarse-grained objects, manipulated relatively infrequently,
+\name chooses a reasonable trade-off. Regions are coarse-grained objects, manipulated relatively infrequently,
 in comparison to manipulations of the fine-grained objects that reside inside regions. Hence, the programmer burden
@@
 Furthermore, the type can be automatically inferred by \name's region type
-inference (\S~\ref{sec:type-inference}), which can peform the above
+inference (\S~\ref{sec:type-inference}), which can perform the above
 non-trivial reasoning on behalf of the programmer.

### 2016-07-07 paper/related.tex
Typos fixed.

@@
    control over  region management and allocation of objects in regions.
-   In constrast, some systems automate all aspects of memory management.
-   This is a convenience-performance tradeoff.
+   In contrast, some systems automate all aspects of memory management.
+   This is a convenience-performance trade-off.
 
@@
 \item Our underlying language is an object-oriented programming language,
-equipped with higher-order functions and parametrized (generic) types.
+equipped with higher-order functions and parameterized (generic) types.
 These language features necessitate some non-trivial choices in the design
@@
 with dynamic lifetimes, and imposes linear typing to control accesses to regions.
-Our open/close lexical block for transferable regions traces its orgins to the \C{let!}
+Our open/close lexical block for transferable regions traces its or gins to the \C{let!}
 expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes

### 2016-07-07 paper/type-inference.tex
Typos fixed.

@@
 \begin{enumerate}
- \item \emph{Region Parametrization}.
+ \item \emph{Region Parameterization}.
    The first step elaborates the input program by introducing \emph{formal region parameters}
@@
 
-\subsection{Region Parametrization for Classes}
+\subsection{Region Parameterization for Classes}
 \label{sec:fb-templatization}
 
-Region parametrization is an iterative process involving the following three steps,
+Region parameterization is an iterative process involving the following three steps,
 the first two of which are mutually dependent on each other.
@@
 The above technique can be extended to mutually recursive classes in a
-straightforward manner, by simultaneously parametrizing them (and
+straightforward manner, by simultaneously parameterizing them (and
 then instantiating them).
@@
 Since \FB{} is higher-order, fields of function type are allowed.We explain how the parameter instantiation step
-instantiates function types below, after discussing parametrization for methods.
+instantiates function types below, after discussing parameterization for methods.
 
@@
 
-\subsection{Region Parametrization for Methods and Function Types}
+\subsection{Region Parameterization for Methods and Function Types}
 
@@
 A \emph{region variable} is introduced to represent an unknown actual region parameter
-of a method invocation or object alloation, which will be bound to a region constant by the end of the
+of a method invocation or object allocation, which will be bound to a region constant by the end of the
 type inference.
@@
 
-\item Our technique for region parametrization also uses a heuristic in the case of
+\item Our technique for region parameterization also uses a heuristic in the case of
 higher order programs. The following examples illustrates that principal types may

### 2016-07-07 paper/related.tex
Fixed a typo that aspell introduced.

@@
 with dynamic lifetimes, and imposes linear typing to control accesses to regions.
-Our open/close lexical block for transferable regions traces its or gins to the \C{let!}
+Our open/close lexical block for transferable regions traces its origins to the \C{let!}
 expression in~\cite{WW01} and~\cite{wadler90}, which safely relaxes

### 2016-09-30 paper/typeinference-proofs.tex
added type inference proof file

\begin{lemma}
The antecedent of any generated validity constraint is of the form
$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
$\phictxt$ is a conjunction of zero or more outlives-constraints.
Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
a conjunct in $\phictxt$.
\end{lemma}

\begin{proof}
  By induction over the constraint-generation rules.
  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
  The only rule that modifies $\phicx$ is the rule for \C{letregion}
  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
  as conjuncts to $\phicx$.
\end{proof}

\begin{theorem}
Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
variables in an execution of the type inference algorithm.
\begin{enumerate}
\item $\absof{q}$ = $p$
\item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
\item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
\item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
$C$ has a solution.
\item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
\item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
\end{enumerate}
\end{theorem}

### 2016-10-03 paper/typeinference-proofs.tex
switched to ESOP formatting

@@
+Invariants and other points:
+\begin{itemize}
+  \item Every predicate variable has a single domain-constraint.
+  \item Progress: Every region-variable will be unified with some constant.
+  \item Notation: should we use a different notation to distinguish a validity constraint
+    from actual entailment?
+\end{itemize}
+
 \begin{lemma}
@@
 
+\begin{lemma}
+  Consider any constraint satisfying the above invariant with a rhs of the
+  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
+  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+\end{lemma}
+
 \begin{theorem}

### 2016-10-03 paper/related.tex
fixed spaces

@@
 management have been extensively studied in the context of various
-languages and problem settings~\cite{cyclone02, cyclone04, yates99,
-MIT03, DPJ09, HMN01, WW01, rust, gpu14}. Our work differs from the
+languages and problem settings~\cite{cyclone02,cyclone04,yates99,MIT03,DPJ09,HMN01,WW01,rust,gpu14}.
+Our work differs from the
 existing proposals in one or more of the following respects.

### 2016-10-04 paper/typeinference-proofs.tex
minor fixes

@@
-Invariants and other points:
+Invariants, other points, and to-dos:
 \begin{itemize}
+  \item use of a set to represent conjunctions
+  \item equality is shorthand for two inequalities
+  \item unify notation for substitution
+  \item substitution required in correctness theorem as well
+  \item use of alpha-renaming, and its use to discharge not-in contraints;
+also, incorporate this into first stage rather than second (in Fig. 7).
+  \item Modify CSP definition to be same as preceding; we can define the $\Delta$s 
+    as auxiliary notation
+  \item Note that we can also clarify that the well-formedness constraints are
+    syntactic sugar from basic set constraints
   \item Every predicate variable has a single domain-constraint.
-  \item Progress: Every region-variable will be unified with some constant.
+  \item Prove: Progress: Every region-variable will be unified with some constant.
   \item Notation: should we use a different notation to distinguish a validity constraint
     from actual entailment?
+  \item Proof by relating proof-trees with constraint-generation trees; each can
+    be constructed from the other; inductive lemma is as earlier
 \end{itemize}

### 2016-10-05 paper/type-inference.tex
updated constraint generation
switched to ESOP formatting
do

@@
+\newcommand{\soln}{\eta}
+
 \section{Type Inference}
@@
 
-%%    as follows:
-%% \begin{enumerate}
-%% \item For every class determine a set of \emph{region parameters} ($\rho$)
-%%        it should be parametrized over.
-%% \item For every field of the class, determine its region parametrized type by
-%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
-%%        which in turn become formal region parameters of the containing class.
-%% \item  For every method and function type, determine its corresponding region
-%%    type template by introducing fresh region parameters as required by
-%%    the type of each parameter and return value (in addition to the allocation region
-%%    parameter).
-%%  \item Elaborate every \C{new} expression, method call or function application by
-%%    introducing \emph{fresh region variables} (as required by the region type
-%%    template of the called method or function).
-%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
-%%    an undetermined set of constraints over the region parameters of that
-%%    class/method.
-%% \end{enumerate}
-
  \item \emph{Constraint Generation}.
@@
    We solve the generated set of constraints using our fixpoint constraint
-   solving algorithm \csolvestar, which reduces the constraint solving problem to
+   solving algorithm, which reduces the constraint solving problem to
    an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
@@
    stack region, then the constraints generated during the elaboration
-   are not satisfiable. In such a case, \csolvestar{} fails to solve
+   are not satisfiable. In such a case, the solver fails to solve
    the constraints.
-% in a Herbrand constraint system, and then relies on \csolve,
-% our abduction solver for that domain.
 
- \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
-  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
-  step 1. We apply these substitutions to the elaborated program to produce the final program.
+ \item If the solver succeeds, it returns a solution $\soln$ consisting of a
+   pair of substitution functions $\substFn_\rho$ and $\substFn_\varphi$ for
+   free region and predicate variables, respectively, introduced in step 1.
+   We apply these substitutions to the elaborated program to produce the final program.
 \end{enumerate}
@@
   let $q$ = IntroduceRegionParameters($p$) in
-  let $(r,C)$ = GenerateConstraints($q$) in
+  let $C$ = GenerateConstraints($q$) in
   match (SolveConstraints($C$) with
   | None $\longrightarrow$ None
-  | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
+  | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
 \end{codeml}
@@
 We later on establish the soundness of the type inference algorithm, and also establish
-partial completeness results.
+that the constraint-generation and constraint-solving steps (the second and third steps
+above) are complete.
 
@@
 \begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
+\begin{minipage}{0.35\linewidth}
 \begin{codejava}
-class Pair
-  $\extends$ $\ObjZ$
-{
+class Pair $\extends$ $\ObjZ$ {
   $\ObjZ$ fst;
@@
 &
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.6\linewidth}
 \begin{codejava}
-class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
+class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
   $\ObjZ \langle \rho_1 \rangle$ fst;
@@
 the class, and the region parameters of the recursive class are computed
+
 as before. We then do parameter instantiation for all recursive fields,
@@
 \begin{tabular}{ccc}
-\begin{minipage}{0.28\linewidth}
+\begin{minipage}{0.35\linewidth}
 \begin{codejava}
-class List
-  $\extends$ $\ObjZ$
-{
+class List $\extends$ $\ObjZ$ {
   $\ObjZ$ data;
@@
 &
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.65\linewidth}
 \begin{codejava}
-class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
+class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
   $\ObjZ \langle \rho_1 \rangle$ data;
@@
 The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
+(This can be extended to use the bound specified for \C{T}, if any. We do not
+discuss this due to space restrictions.)
 % number of region parameters
@@
 allocation region.
-As for the classes, we also introduce a fresh predicate variable $\varphi$ for every method.
+We also introduce a fresh predicate variable $\varphi$ for every method,
+just as we did for each class.
 Thus, the method
@@
 
-\subsection{The Constraint Satisfaction Problem (CSP)}
+\subsection{Constraint Generation}
 \label{sec:fb-constraintsem}
@@
 \paragraph{Syntax of Constraints.}
-The constraints are expressed in terms of region identifiers ($\pi$ and $\rho$)
-and predicate variables ($\varphi$).
+The constraints are expressed using
+a set $\regionConstants$ of region constants,
+a set $\regionVars$ of region variables,
+and
+a set $\predVars$ of predicate variables.
+% 
+% ($\pi$ and $\rho$) and predicate variables ($\varphi$).
+% $(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
 
-\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
-A \emph{region variable} is introduced to represent an unknown actual region parameter
+% \emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
+Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
 of a method invocation or object allocation, which will be bound to a region constant by the end of the
@@
 \begin{array}{lcl}
-\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
+\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
 \end{array}
 \end{smathpar}
+We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
+We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
+as shorthand for $\conj_{\phi \in S} \phi$.
+
 We will use the term \emph{validity constraint} to denote an entailment constraint
@@
 
+\paragraph{Constraint Solution.}
+The solution $\soln$ to a set of constraints $C$ is a pair $(\regionSubstFn,\predSubstFn)$,
+where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
+and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
+every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.
+
 \input{fb-constraint-generation}
@@
 These rules use the same context as the corresponding typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
-retaining their meaning.
+Fig.~\ref{fig:fb-staticsem} with symbols $\A$ and $\env$ retaining their meaning.
 The algorithm proceeds top-down, analyzes an expression $e$ in a context
-$\exptycx{\ralloc}{\env}$, and returns a type $\tau$ and a set of constraints $C$
-(expressed in the rules as $\exprok {\exptycx{\ralloc}{\env}} {e} {\tau} {C}$ ),
+$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
+(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
 indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.
@@
 simplifying the constraint solver.
+
+\begin{theorem}
+  \label{thm:constraint-generation-sc}
+Let $C$ be the set of constraints generated from a program $q$.
+Let $\soln$ be any substitution for the region and predicate variables in $q$.
+$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
+\end{theorem}
+\begin{proof}
+  We inductively establish a correspondence between derivation trees under
+  the static semantics and the constraint-generation trees.
+
+% We do not present a formal proof of the theorem.
+The proof essentially requires establishing the following lemma inductively: if
+$\exprok {\stdcontext}{e}{\tau}{C}$
+and $\soln$ satisfies $C$, then
+$\hastyp {\stdcontext}{e[\soln]}{\tau}$,
+where $e[\soln]$ indicates $e$ with the substitutions
+in the solution $\soln$, as well as analogous lemmas for type well-formedness.
+\end{proof}
 
@@
 
-\paragraph{Soundness and Completeness.}
+\paragraph{Soundness and Completeness of the Constraint Solver.}
+
+\begin{theorem}
+  \label{thm:constraint-solver-sc}
+Let $C$ be the set of constraints generated from some $p \in \absof{\FB}$.
+If the constraint solver returns Some($\soln$), then $\soln$ is a valid solution
+for $C$.
+Conversely, if $C$ has some valid solution, then the constraint solver will return
+a value of the form Some($\soln$).
+\end{theorem}
+
+\subsection{Discussion}
 
@@
 \end{theorem}
-
-We do not present a formal proof of the theorem. The proof essentially requires
-establishing the following lemma inductively: if
-$
- \exprok {\exptycx{\ralloc}{\env}}{e}{\tau}{C}
-$
-and $(\regionSubstFn,\predSubstFn)$ satisfies $C$, then
-$
- \hastyp {\exptycx{\ralloc}{\env}}{e[\regionSubstFn,\predSubstFn]}{\tau}
-$,
-where $e[\regionSubstFn,\predSubstFn]$ indicates $e$ with the substitutions
-in the solution $(\regionSubstFn,\predSubstFn)$,
-as well as analogous lemmas for type well-formedness.
 
@@
 % the last constraint is generated from the recursive call.
+
+\subsection{Proofs}
+
+\input{typeinference-proofs}

### 2016-10-06 paper/type-inference.tex
type inference

@@
 \paragraph{Constraint Solution.}
-The solution $\soln$ to a set of constraints $C$ is a pair $(\regionSubstFn,\predSubstFn)$,
+We define an \emph{assignment} $\soln$ to be a pair of functions $(\regionSubstFn,\predSubstFn)$,
 where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
+and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula.
+Such an assignment is said to be a solution to a set of constraints $C$ if
 every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.
+
+%\begin{definition}
+%\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
+%A constraint satisfaction problem is a tuple
+%$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
+%where $\regionConstants$ is a set of region constants,
+%$\regionVars$ is a set of region variables,
+%$\predVars$ is a set of predicate variables,
+%$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
+%specifies a unification domain for each region variable,
+%$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
+%specifies the domain for each predicate variable,
+%and $\constraintSet$ is a set of validity constraints in one of the following forms:
+%\begin{center}
+%\(
+%    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
+%    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
+%\)
+%\end{center}
+%The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
+%where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
+%and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
+%\begin{itemize}
+%  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
+%
+%  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
+%    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
+%
+%  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
+%    \regionSubstFn$.
+%  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
+%\end{itemize}
+%\end{definition}
 
@@
 
+\TODO{Change the static semantics to be consistent with this, and
+make this a comment there.}
 One notable aspect here is the following. Consider a method with predicate
@@
 
-\paragraph{Constraint Solution}
-Solving the set ($C$) of constraints entails finding an
-assignment for each predicate variable $(\varphi$) and each region
-variable ($\rho$) that occurs free in $C$, that satisfies all
-the validity constraints as well as the well-formedness constraints on
-$\varphi$ and $\rhobar$.
-% To simplify presentation, we think of $C$ as being parameterized on
-% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
-% \rbrack$. We now formalize the constraint satisfaction problem, and
-% its solution.
+\begin{theorem}
+  \label{thm:constraint-generation-sc}
+Let $C$ be the set of constraints generated from a program $q$.
+Let $\soln$ be any substitution for the region and predicate variables in $q$.
+$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
+\end{theorem}
+\begin{proof}
+  We inductively establish a correspondence between derivation trees under
+  the static semantics and the constraint-generation trees.
 
-\begin{definition}
-\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
-A constraint satisfaction problem is a tuple
-$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-where $\regionConstants$ is a set of region constants,
-$\regionVars$ is a set of region variables,
-$\predVars$ is a set of predicate variables,
-$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies a unification domain for each region variable,
-$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
-specifies the domain for each predicate variable,
-and $\constraintSet$ is a set of validity constraints in one of the following forms:
-\begin{center}
-\(
-    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
-    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
-\)
-\end{center}
-The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
-where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
-\begin{itemize}
-  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
-
-  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
-    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
-
-  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
-    \regionSubstFn$.
-  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
-\end{itemize}
-\end{definition}
+% We do not present a formal proof of the theorem.
+The proof essentially requires establishing the following lemma inductively: if
+$\exprok {\stdcontext}{e}{\tau}{C}$
+and $\soln$ satisfies $C$, then
+$\hastyp {\stdcontext}{e[\soln]}{\tau}$,
+where $e[\soln]$ indicates $e$ with the substitutions
+in the solution $\soln$, as well as analogous lemmas for type well-formedness.
+\end{proof}
 
@@
 simplifying the constraint solver.
-
-\begin{theorem}
-  \label{thm:constraint-generation-sc}
-Let $C$ be the set of constraints generated from a program $q$.
-Let $\soln$ be any substitution for the region and predicate variables in $q$.
-$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
-\end{theorem}
-\begin{proof}
-  We inductively establish a correspondence between derivation trees under
-  the static semantics and the constraint-generation trees.
-
-% We do not present a formal proof of the theorem.
-The proof essentially requires establishing the following lemma inductively: if
-$\exprok {\stdcontext}{e}{\tau}{C}$
-and $\soln$ satisfies $C$, then
-$\hastyp {\stdcontext}{e[\soln]}{\tau}$,
-where $e[\soln]$ indicates $e$ with the substitutions
-in the solution $\soln$, as well as analogous lemmas for type well-formedness.
-\end{proof}
 

### 2016-10-06 paper/type-inference.tex
type inference

@@
 \newcommand{\soln}{\eta}
+\newcommand{\saturate}[1]{{#1}^*}
+\newcommand{\groundc}[1]{\saturate{#1}_g}
 
@@
 These rules use the same context as the corresponding typing judgment in
-Fig.~\ref{fig:fb-staticsem} with symbols $\A$ and $\env$ retaining their meaning.
+Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
+use of region variables and predicate variables.
+Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
 The algorithm proceeds top-down, analyzes an expression $e$ in a context
@@
 The rules for generating constraints from a method and
-class definition first build a context ($\A$) containing a set ($ \rhoenv$) of region variables denoting
+class definition first build a context ($\A$) containing a set ($ \rhoenv$) denoting
 regions that are currently live, a map ($\aenv$) mapping type
@@
 \begin{proof}
-  We inductively establish a correspondence between derivation trees under
-  the static semantics and the constraint-generation trees.
+The correspondence between the static semantics rules and the constraint generation
+rules induces a correspondence between derivation trees produced by the
+two systems.
+% We do not present a formal proof of the theorem.
+We can establish the following lemma inductively:
+given a derivation tree $\zeta_1$ for
+$\exprok {\stdcontext}{e}{\tau}{C}$
+and any substitution $\soln$,
+we can construct a corresponding \emph{candidate} derivation
+tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
+We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
+(We make use of analogous lemmas for type well-formedness rules as well.)
+\end{proof}
 
-% We do not present a formal proof of the theorem.
-The proof essentially requires establishing the following lemma inductively: if
-$\exprok {\stdcontext}{e}{\tau}{C}$
-and $\soln$ satisfies $C$, then
-$\hastyp {\stdcontext}{e[\soln]}{\tau}$,
-where $e[\soln]$ indicates $e$ with the substitutions
-in the solution $\soln$, as well as analogous lemmas for type well-formedness.
-\end{proof}
+\subsection{Solving the CSP}
 
@@
 
-\subsection{Solving the CSP}
+\begin{lemma}
+The antecedent of any generated validity constraint is of the form
+$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
+$\phictxt$ is a conjunction of zero or more outlives-constraints.
+Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
+Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
+Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
+every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
+a conjunct in $\phictxt$.
+\end{lemma}
+
+\begin{proof}
+  By induction over the constraint-generation rules.
+  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
+  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
+  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
+  The only rule that modifies $\phicx$ is the rule for \C{letregion}
+  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
+  as conjuncts to $\phicx$.
+\end{proof}
+
+\begin{lemma}
+  Consider any constraint satisfying the above invariant with a rhs of the
+  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
+  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+\end{lemma}
 
 \paragraph{Constraint Solver.}
+Our algorithm for solving a set of constraints $C$ consists of the following steps:
+\begin{enumerate}
+\item We use an iterative fixed point (saturation) algorithm to compute
+$\saturate{C}$, the set of all constraints logically implied by $C$.
+\item Let $\groundc{C}$ denote the subset of all ground constraints
+(\ie, constraints without any region variable or predicate variable)
+in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
+partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
+as below. Otherwise, the constraint solver fails, as no valid solution exists.
+\begin{itemize}
+\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
+to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
+\item For any region variable $\rho$,  we are guaranteed that there exists a unique
+ $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
+with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
+\end{itemize}
+\end{enumerate}
+
 Solving the constraints is an iterative process, consisting of the following
@@
 
+\paragraph{Soundness and Completeness of the Constraint Solver.}
+
+\begin{theorem}
+  \label{thm:constraint-solver-sc}
+Let $C$ be the set of constraints generated from some $p \in \absof{\FB}$.
+If the constraint solver returns Some($\soln$), then $\soln$ is a valid solution
+for $C$.
+Conversely, if $C$ has some valid solution, then the constraint solver will return
+a value of the form Some($\soln$).
+\end{theorem}
+
 \paragraph{Modularity Aspects.}
@@
 However, methods that are not mutually recursive can be processed separately.
-
-\paragraph{Soundness and Completeness of the Constraint Solver.}
-
-\begin{theorem}
-  \label{thm:constraint-solver-sc}
-Let $C$ be the set of constraints generated from some $p \in \absof{\FB}$.
-If the constraint solver returns Some($\soln$), then $\soln$ is a valid solution
-for $C$.
-Conversely, if $C$ has some valid solution, then the constraint solver will return
-a value of the form Some($\soln$).
-\end{theorem}
 

### 2016-10-06 paper/typeinference-proofs.tex
do
type inference

@@
 \begin{itemize}
+  \item discuss computation of least-solution to constraints?
   \item use of a set to represent conjunctions
@@
 
-\begin{lemma}
-The antecedent of any generated validity constraint is of the form
-$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
-$\phictxt$ is a conjunction of zero or more outlives-constraints.
-Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
-Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
-Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
-every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
-a conjunct in $\phictxt$.
-\end{lemma}
-
-\begin{proof}
-  By induction over the constraint-generation rules.
-  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
-  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
-  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
-  The only rule that modifies $\phicx$ is the rule for \C{letregion}
-  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
-  as conjuncts to $\phicx$.
-\end{proof}
-
-\begin{lemma}
-  Consider any constraint satisfying the above invariant with a rhs of the
-  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
-  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
-\end{lemma}
-
 \begin{theorem}

### 2016-10-06 paper/type-inference.tex
constraint solver

@@
 to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$,  we are guaranteed that there exists a unique
- $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
-with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
+\item For any region variable $\rho$,  the set
+ $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
+is guarantted to contain a single element $\pi$.
+We define $\regionSubstFn(\rho)$ to be $\pi$.
+%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
+% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
+%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
 \end{itemize}
@@
 
-Solving the constraints is an iterative process, consisting of the following
-steps, and this process fails at any point if it determines that no solution
-exists for the set of constraints:
+We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
+point is reached:
 
@@
 
+\item (Initialization) $\isvalid{\ell}{r} \in C \Rightarrow \isvalid{\ell}{r} \in \saturate{C}$
+
+\item (Transitivity)
+$\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
+$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$,
+$\Rightarrow$
+$\isvalid{\ell}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+
+\item (Substitution)
+$\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
+$\isvalid{\varphi}{\phi} \in \saturate{C}$
+$\Rightarrow$ $\isvalid{\ell}{F(\phi)} \in \saturate{C}$
+
+\item (Abduction Decomposition)
+\label{item:context}
+\begin{enumerate}
 \item
-    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
-$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
-If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
-solver fails.
+$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
+$\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$
+$\Rightarrow$
+$\isvalid{\varphi}{\pi_i \outlives \pi_j} \in \saturate{C} $
 
 \item 
-When a region variable $\rho$ is unified with a region constant $\pi$,
-if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
-and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
-If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
+$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
+$\{ \pi_i, \pi_j \} \subseteq \regionConstants \setminus \predDeltaMap(\varphi)$
+$\Rightarrow$
+$\isvalid{\phictxt}{\pi_i \outlives \pi_j} \in \saturate{C} $
+\end{enumerate}
 
-\item
-\label{item:context}
-Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
-where $\pi_i$ and $\pi_j$ are both region constants.
-If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
-$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
-
-\item
-  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
-  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
-  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
+%\item
+%\label{item:context}
+%Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
+%where $\pi_i$ and $\pi_j$ are both region constants.
+%If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
+%$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
+%Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+%If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
+%
+%\item
+%    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
+%$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
+%If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
+%solver fails.
+%
+%\item
+%When a region variable $\rho$ is unified with a region constant $\pi$,
+%if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
+%and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
+%If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
+%\item
+%  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
+%  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
+%  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
 \end{enumerate}

### 2016-10-06 paper/type-inference.tex
constraint solver

@@
 \newcommand{\groundc}[1]{\saturate{#1}_g}
+\newcommand{\satC}{\saturate{C}}
 
@@
 
-\subsection{Solving the CSP}
-
-\paragraph{Properties of generated constraints}
-In the sequel, we will restrict our attention to a special class of CSPs that satisfy
-two interesting properties that allow us to solve the constraints efficiently.
-The constraints generated by our type inference algorithm satisfy these two properties.
-
-The first property is that every region variable is guaranteed to be unified with some
-region constant by the constraints\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
-
-The second property is that the \emph{context region constraint} $\phictxt$
-occurring on the antecedent of any validity constraint is a conjunction of
-outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-is a static region identifier and $\pi$ is a region constant (either a formal
-region parameter, or static region identifier, or open transferable region identifier).
-Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
-simplifying the constraint solver.
-
-\begin{lemma}
-The antecedent of any generated validity constraint is of the form
-$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
-$\phictxt$ is a conjunction of zero or more outlives-constraints.
-Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
-Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
-Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
-every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
-a conjunct in $\phictxt$.
-\end{lemma}
-
-\begin{proof}
-  By induction over the constraint-generation rules.
-  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
-  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
-  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
-  The only rule that modifies $\phicx$ is the rule for \C{letregion}
-  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
-  as conjuncts to $\phicx$.
-\end{proof}
-
-\begin{lemma}
-  Consider any constraint satisfying the above invariant with a rhs of the
-  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
-  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
-\end{lemma}
+\subsection{The Constraint Solver}
 
@@
 $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
-$\{ \pi_i, \pi_j \} \subseteq \regionConstants \setminus \predDeltaMap(\varphi)$
+$\{ \pi_i, \pi_j \} \not\subseteq \predDeltaMap(\varphi)$,
+$\{ \pi_i, \pi_j \} \subseteq \regionConstants$
 $\Rightarrow$
@@
 
+\paragraph{Properties of generated constraints}
+Our algorithm makes use of some special properties of $C$ guaranteed by
+our constraint-generation algorithm, which we now describe.
+
+For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
+of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
+$\predDeltaMap{\varphi}$.
+
+We refer to any validity constraint whose antecedent contains a predicate variable
+(\ie, is of the form $\varphi$ or $\varphi \conj \phictxt$) as an \emph{abduction
+constraint}. We refer to an abduction constraint in $C$ as \emph{decomposable}
+if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
+$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
+$\pi_1 \outlives \pi_2$ satisfying the following conditions:
+(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
+(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
+every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
+a conjunct in $\phictxt$.
+
+\begin{lemma}
+Let $C$ be the set of constraints generated by our algorithm.
+Every abduction constraint in $C$ is decomposable.
+\end{lemma}
+
+\begin{proof}
+  By induction over the constraint-generation rules.
+  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
+  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
+  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
+  The only rule that modifies $\phicx$ is the rule for \C{letregion}
+  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
+  as conjuncts to $\phicx$.
+\end{proof}
+
+%The second property is that the \emph{context region constraint} $\phictxt$
+%occurring on the antecedent of any validity constraint is a conjunction of
+%outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
+%is a static region identifier and $\pi$ is a region constant (either a formal
+%region parameter, or static region identifier, or open transferable region identifier).
+%Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
+%it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+%This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+%iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
+%simplifying the constraint solver.
+
+\begin{lemma}
+  Consider any constraint satisfying the above invariant with a rhs of the
+  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
+  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+\end{lemma}
+
 \paragraph{Algorithmic Aspects.}
@@
 \paragraph{Soundness and Completeness of the Constraint Solver.}
+
+Correctness follows by establishing:
+\begin{theorem}
+(a) For any assignment $\soln$, $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+(b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
+(c) For any region variable $\rho$, $\satC$ contains some constraint of the
+form $\isvalid{\ell}{\rho = \pi}$.
+\end{theorem}
+
+The last item in the above theorem follows because every region variable is guaranteed to be unified with some
+region constant by the generated constraints\footnote{
+There is one special case where this does not hold: a recursive function
+that calls itself in a non-terminating fashion. Such a function never
+returns a value, and so the return-value can be typed as anything.
+We assume that the return value of such a function is typed to be \C{unit},
+which will not introduce any region variables at the call-site of such a function.}
+This follows because $\FB$ does not admit uninitialized variables.
+% , and the type system uniquely determines the region parameters of every value
 

### 2016-10-06 paper/type-inference.tex
constraint solver

@@
 \begin{lemma}
-  Consider any constraint satisfying the above invariant with a rhs of the
-  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
+  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+where both $\pi_i$ and $\pi_j$ are region constants.
 (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
   (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
 \end{lemma}
+
+Note that the above lemma shows how the conjunction in the antecedent of
+any decomposable constraint can be eliminated.
 

### 2016-10-07 paper/type-inference.tex
type inference updates

@@
 
-Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
-inference algorithm.
+% Fig.~\ref{fig:type-inference-algo}
+We now present a high-level outline of the type inference algorithm.
 The algorithm consists of the following steps:
@@
  \item If the solver succeeds, it returns a solution $\soln$ consisting of a
-   pair of substitution functions $\substFn_\rho$ and $\substFn_\varphi$ for
+   pair of substitution functions $\regionSubstFn$ and $\predSubstFn$ for
    free region and predicate variables, respectively, introduced in step 1.
@@
 
-\begin{figure}
-\begin{codeml}
-Infer ($p$) =
-  let $q$ = IntroduceRegionParameters($p$) in
-  let $C$ = GenerateConstraints($q$) in
-  match (SolveConstraints($C$) with
-  | None $\longrightarrow$ None
-  | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
-\end{codeml}
-
-\caption{The type inference algorithm}
-\label{fig:type-inference-algo}
-\end{figure}
+% \begin{figure}
+% \begin{codeml}
+% Infer ($p$) =
+%   let $q$ = IntroduceRegionParameters($p$) in
+%   let $C$ = GenerateConstraints($q$) in
+%   match (SolveConstraints($C$) with
+%   | None $\longrightarrow$ None
+%   | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
+% \end{codeml}
+% 
+% \caption{The type inference algorithm}
+% \label{fig:type-inference-algo}
+% \end{figure}
 
@@
 the class, and the region parameters of the recursive class are computed
-
 as before. We then do parameter instantiation for all recursive fields,
@@
 \emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if Infer($p$) returns Some($t$), then
+For any $p \in \absof{\FB}$, if type inference succeeds, it returns a $t$ such that
+% Infer($p$) returns Some($t$), then
 (1) $\absof{t} = p$, and
@@
 
+% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
+
 However, in this context, completeness is more interesting than soundness.
@@
 the resulting program.)
-We believe that the converse of the above lemma can also be established:
-any substitution for the region variables and the predicate variables that makes the
-expression well-typed will also satisfy the generated constraints.
-Thus, if there is any possible instantiation (of the region variables and predicate variables) of
-the elaborated program produced by the first phase that type checks, then the generated
-constraints are satisfiable, and the second phase will identify it.
-
-However, the entire type inference algorithm is not complete, because of some heuristic
-choices made during the first phase (elaboration phase), as explained below.
+Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
+show that the second and third steps of the type-inference algorithm are complete.
+%% We believe that the converse of the above lemma can also be established:
+%% any substitution for the region variables and the predicate variables that makes the
+%% expression well-typed will also satisfy the generated constraints.
+%% Thus, if there is any possible instantiation (of the region variables and predicate variables) of
+%% the elaborated program produced by the first phase that type checks, then the generated
+%% constraints are satisfiable, and the second phase will identify it.
+%% 
+The only sources of incompleteness in the type inference algorithm are some heuristic
+choices made in the first step (the elaboration phase), as explained below.
 
@@
 
-\subsection{Proofs}
+\subsection{Miscellaneous points and to-dos}
 
-\input{typeinference-proofs}
+% \input{typeinference-proofs}
+
+\begin{itemize}
+  \item for function-calls: need basic typing from core language to introduce region parameters
+  \item discuss computation of least-solution to constraints?
+  \item use of a set to represent conjunctions
+  \item equality is shorthand for two inequalities
+  \item use of alpha-renaming, and its use to discharge not-in contraints;
+also, incorporate this into first stage rather than second (in Fig. 7).
+  \item Note that we can also clarify that the well-formedness constraints are
+    syntactic sugar from basic set constraints
+  \item Prove: Progress: Every region-variable will be unified with some constant.
+  \item Notation: should we use a different notation to distinguish a validity constraint
+    from actual entailment?
+\end{itemize}

### 2016-10-07 paper/typeinference-proofs.tex
type inference updates

@@
-Invariants, other points, and to-dos:
-\begin{itemize}
-  \item discuss computation of least-solution to constraints?
-  \item use of a set to represent conjunctions
-  \item equality is shorthand for two inequalities
-  \item unify notation for substitution
-  \item substitution required in correctness theorem as well
-  \item use of alpha-renaming, and its use to discharge not-in contraints;
-also, incorporate this into first stage rather than second (in Fig. 7).
-  \item Modify CSP definition to be same as preceding; we can define the $\Delta$s 
-    as auxiliary notation
-  \item Note that we can also clarify that the well-formedness constraints are
-    syntactic sugar from basic set constraints
-  \item Every predicate variable has a single domain-constraint.
-  \item Prove: Progress: Every region-variable will be unified with some constant.
-  \item Notation: should we use a different notation to distinguish a validity constraint
-    from actual entailment?
-  \item Proof by relating proof-trees with constraint-generation trees; each can
-    be constructed from the other; inductive lemma is as earlier
-\end{itemize}
-
-\begin{theorem}
-Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
-variables in an execution of the type inference algorithm.
-\begin{enumerate}
-\item $\absof{q}$ = $p$
-\item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
-\item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
-\item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
-$C$ has a solution.
-\item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
-\item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
-\end{enumerate}
-\end{theorem}
+% \begin{theorem}
+% Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
+% variables in an execution of the type inference algorithm.
+% \begin{enumerate}
+% \item $\absof{q}$ = $p$
+% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
+% \item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
+% \item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
+% $C$ has a solution.
+% \item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
+% \item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
+% \end{enumerate}
+% \end{theorem}

### 2016-10-07 paper/type-inference.tex
type inference updates

@@
 \newcommand{\groundc}[1]{\saturate{#1}_g}
+\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
 \newcommand{\satC}{\saturate{C}}
@@
 and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula.
-Such an assignment is said to be a solution to a set of constraints $C$ if
+Such an assignment is said to \emph{satisfy} a set of constraints $C$ if
 every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.
+We say that $C$ is \emph{satisfiable} if it has a satisfying assignment.
 
@@
 \label{thm:constraint-generation-sc}
-Let $C$ be the set of constraints generated from a program $q$.
-Let $\soln$ be any substitution for the region and predicate variables in $q$.
-$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
+Let $C = \consOf{q}$.
+Any assignment $\soln$ (for the region and predicate variables in $q$)
+satisfies $C$ iff $q[\soln]$ is well-typed.
 \end{theorem}
@@
 
+% \paragraph{Properties of generated constraints}
+Our constraint solver makes use of some special properties of the set of constraints $C$
+generated by our algorithm, which we now describe.
+
+For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
+of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
+$\predDeltaMap(\varphi)$.
+
+The formalism below is motivated by the following intuition.
+Suppose some construct in a method's body requires an outlives-constraint
+$\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
+of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
+as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
+$\pi_1 \outlives \pi_2$ must be established by the context within the method
+(\eg, by the \code{letregion} constructs in the method).
+We now generalize this idea to the setting of our abstract constraint-satisfication
+problem.
+
+We refer to any validity constraint whose antecedent contains a predicate variable
+(\ie, is of the form $\varphi$ or $\varphi \conj \phictxt$) as an \emph{abduction
+constraint}. We refer to an abduction constraint in $C$ as \emph{decomposable}
+if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
+$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
+$\pi_1 \outlives \pi_2$ satisfying the following conditions:
+(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
+(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
+every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
+a conjunct in $\phictxt$.
+
+\begin{lemma}
+  \label{lemma:gc-is-decomposable}
+  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
+\end{lemma}
+
+\begin{proof}
+  By induction over the constraint-generation rules.
+  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
+  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
+  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
+  The only rule that modifies $\phicx$ is the rule for \C{letregion}
+  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
+  as conjuncts to $\phicx$.
+\end{proof}
+
+%The second property is that the \emph{context region constraint} $\phictxt$
+%occurring on the antecedent of any validity constraint is a conjunction of
+%outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
+%is a static region identifier and $\pi$ is a region constant (either a formal
+%region parameter, or static region identifier, or open transferable region identifier).
+%Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
+%it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
+%This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+%iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
+%simplifying the constraint solver.
+
+\begin{lemma}
+  \label{lemma:decomposition}
+  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+where both $\pi_i$ and $\pi_j$ are region constants.
+(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
+\end{lemma}
+
+\begin{proof}
+  Straightforward.
+\end{proof}
+
+Note that the above lemma shows how the conjunction in the antecedent of
+any decomposable constraint can be eliminated.
+
 \paragraph{Constraint Solver.}
@@
 to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$,  the set
- $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
-is guarantted to contain a single element $\pi$.
-We define $\regionSubstFn(\rho)$ to be $\pi$.
+\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
+set $\{ \pi \in \regionDeltaMap(\rho) \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
+if this set is non-empty.
+Otherwise, the solver fails.
 %\item For any region variable $\rho$,  we are guaranteed that there exists a unique
@@
 
+\paragraph{Step 1.}
 We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
@@
 \item (Transitivity)
+  \label{item:transitivity}
 $\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
@@
 \item (Substitution)
+  \label{item:subst}
 $\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
@@
 
-Point~\ref{item:context} makes use of the special structure of the generated
-constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
-form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
-region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
+Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+% the special structure of the generated
+% constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
+% form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
+% region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
+% iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
 
-\paragraph{Properties of generated constraints}
-Our algorithm makes use of some special properties of $C$ guaranteed by
-our constraint-generation algorithm, which we now describe.
-
-For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
-of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
-$\predDeltaMap{\varphi}$.
-
-We refer to any validity constraint whose antecedent contains a predicate variable
-(\ie, is of the form $\varphi$ or $\varphi \conj \phictxt$) as an \emph{abduction
-constraint}. We refer to an abduction constraint in $C$ as \emph{decomposable}
-if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
-$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
-$\pi_1 \outlives \pi_2$ satisfying the following conditions:
-(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
-(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
-every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
-a conjunct in $\phictxt$.
-
-\begin{lemma}
-Let $C$ be the set of constraints generated by our algorithm.
-Every abduction constraint in $C$ is decomposable.
-\end{lemma}
-
-\begin{proof}
-  By induction over the constraint-generation rules.
-  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
-  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
-  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
-  The only rule that modifies $\phicx$ is the rule for \C{letregion}
-  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
-  as conjuncts to $\phicx$.
-\end{proof}
-
-%The second property is that the \emph{context region constraint} $\phictxt$
-%occurring on the antecedent of any validity constraint is a conjunction of
-%outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-%is a static region identifier and $\pi$ is a region constant (either a formal
-%region parameter, or static region identifier, or open transferable region identifier).
-%Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-%it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-%This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-%iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
-%simplifying the constraint solver.
-
-\begin{lemma}
-  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
-where both $\pi_i$ and $\pi_j$ are region constants.
-(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
-\end{lemma}
-
-Note that the above lemma shows how the conjunction in the antecedent of
-any decomposable constraint can be eliminated.
+\paragraph{Step 2.}
+Checking the validity of a ground constraint is straightforward.
+A ground constraint is of the form $\isvalid{\conj_{i \in I} \ell_i}{r}$, where each
+$\ell_i$ and $r$ is an outlives-constraint.
+Let $D$ denote $\{ \isvalid{}{\ell_i} \;|\; i \in I \}$.
+The given constraint is valid iff $r$ belongs to $\saturate{D}$.
 
 \paragraph{Algorithmic Aspects.}
-The iterative process described above is a standard fixed point computation.
-We can encode this computation using a set of Datalog rules,
-and compute the solution using any Datalog engine.
-Alternatively, algorithms for context-free reachability can be adapted
-to solve the constraints, as explained below.
-
-Given a set $S$ of outlives constraints, we define the directed
+Note that checking the validity of a ground constraint can be realized using
+a simple graph reachability algorithm.
+%
+Given a set $S$ of outlives constraints, define the directed
 graph $G(S)=(V(S),E(S))$ as follows.
@@
 an edge from $\rho_1$ to $\rho_2$.
-Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
+(Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
 for the pair of constraints $\rho_1 \outlives \rho_2$ and
 $\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
-by a pair of edges.
+by a pair of edges.)
 %
-It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
+It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$ iff
 there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
 %
-Thus, simple graph reachability (transitive closure) algorithms can be
-used to identify all the logical consequences of a set of outlives
-constraints.
+Thus, a simple graph reachability algorithm can be used to check the validity of
+ground constraints.
+% all the logical consequences of a set of outlives constraints.
 
-Given a set of context-free validity constraints, we can generalize the
-above approach and reduce the problem to one of context-free reachability
-in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
-%
-Algorithms for context-free reachability can also be adapted to incorporate
-the above treatment of context constraints $\phictxt$.
-%
-Further details omitted.
+This idea generalizes. Consider a set $C$ of constraints that contains no
+constraint with an antecedent $\varphi \conj \phi$ where $\phi$ is a non-empty conjunction. 
+Then, the algorithm for computing the closure $\satC$ does not need to use
+rule~\ref{item:context} (for Abduction Decomposition). The closure in this
+case can be computed using context-free reachability in graphs~\cite{Reps:Reachability}
+(as usual for context-sensitive interprocedural analysis).
+
+Algorithms for context-free reachability can be adapted to incorporate
+the Abduction Decomposition step.
+% Further details omitted.
+Alternatively, the iterative process described above is a standard fixed point computation.
+We can encode this computation using a set of Datalog rules,
+and compute the solution using any Datalog engine.
 
@@
 \begin{theorem}
-(a) For any assignment $\soln$, $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+(a) Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
 (b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
@@
 \label{thm:constraint-solver-sc}
-Let $C$ be the set of constraints generated from some $p \in \absof{\FB}$.
-If the constraint solver returns Some($\soln$), then $\soln$ is a valid solution
-for $C$.
-Conversely, if $C$ has some valid solution, then the constraint solver will return
-a value of the form Some($\soln$).
+Let $C = \consOf{q}$.
+(Soundness) If the constraint solver succeeds, it will return an assignment $\soln$ 
+that satisfies $C$.
+(Completeness)
+Conversely, if $C$ is satisfiable, then the constraint solver will succeed.
 \end{theorem}
@@
 
-\subsection{Discussion}
+\subsection{Soundness and Completeness: Discussion}
 
-We now state the soundness theorem:
+We now state the soundness theorem, which follows from
+Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
+as well as the following observation about the elaboration phase.
+For any program $p \in \absof{\FB}$, the elaboration phase produces a $q$
+such that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
 
@@
 \begin{enumerate}
-\item Our technique for determining the set of region parameters for a recursive class
-incorporates a specific heuristic, namely that the recursive occurrence of the class has
-the same parameters, in the same order, as the class itself. This technique will fail, for
-example. if the program uses a recursive list type whose elements alternatively come
-from two different regions. Such a program would require the following elaboration,
-which is beyond the scope of our approach:
-% , $R_1$ and $R_2$.
-% both of which are region parameters of the class.
+\item We determine the set of region parameters for a recursive class
+using the heuristic that a recursive occurrence of the class has
+the same parameters, in the same order, as the class itself.
+This heuristic fails, for example, if the program uses a recursive list type whose
+elements alternatively come from two different regions. Such a program would require
+the following elaboration, which is beyond the scope of our approach:
+
 \begin{tabular}{ccc}
-\begin{minipage}{0.3\linewidth}
+\begin{minipage}{0.35\linewidth}
 \begin{codejava}
-class List
-  $\extends$ $\ObjZ$
+class List $\extends$ $\ObjZ$
 {
@@
 &
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.65\linewidth}
 \begin{codejava}
-class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
-  $\extends$ $\ObjZ \langle \rho_0 \rangle$
+class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
 {
@@
 \begin{codejava}
-unit apply ( T $\rightarrow$ unit f, T x, T y) {
-   f(x); f(y);
-}
+    unit apply ( T $\rightarrow$ unit f, T x, T y) { f(x); f(y); }
 \end{codejava}
@@
 \begin{itemize}
+  \item explain: non-determinism in defining region-variable binding is immaterial.
+  \item explain lemma 2's use case (a) and (b)
+  \item notation: decomposable is dependent on C
+  \item cleanup: syntax of constraints (region-constraint, validity constraint, etc.)
+    and later usage (abduction constraint)
+  \item adjust spacing in RULEs
   \item for function-calls: need basic typing from core language to introduce region parameters
@@
     from actual entailment?
+  \item update abstract/intro to stress aspects of constraint solver
+  \item Change static semantics to use method precondition, without class precondition
 \end{itemize}

### 2016-10-07 paper/type-inference.tex
type inference updates

@@
 \newcommand{\saturate}[1]{{#1}^*}
-\newcommand{\groundc}[1]{\saturate{#1}_g}
+\newcommand{\myground}[1]{\saturate{#1}_g}
 \newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
 \newcommand{\satC}{\saturate{C}}
+\newcommand{\groundC}{\myground{C}}
 
@@
 $\saturate{C}$, the set of all constraints logically implied by $C$.
-\item Let $\groundc{C}$ denote the subset of all ground constraints
+\item Let $\myground{C}$ denote the subset of all ground constraints
 (\ie, constraints without any region variable or predicate variable)
-in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
-partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
+in $\saturate{C}$. We check if $\myground{C}$ is valid in the theory of
+partial-order. If $\myground{C}$ is valid, the solution $\soln$ is determined
 as below. Otherwise, the constraint solver fails, as no valid solution exists.
@@
 
-Correctness follows by establishing:
+% Correctness follows by establishing:
 \begin{theorem}
-(a) Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
-(b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
-(c) For any region variable $\rho$, $\satC$ contains some constraint of the
-form $\isvalid{\ell}{\rho = \pi}$.
+  \label{thm:closure}
+Let $C = \consOf{q}$.
+Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+%% (b) $\satC$ is satisfiable iff $\myground{C}$ is valid.
+%% (c) For any region variable $\rho$, $\satC$ contains some constraint of the
+%% form $\isvalid{\ell}{\rho = \pi}$.
 \end{theorem}
 
-The last item in the above theorem follows because every region variable is guaranteed to be unified with some
-region constant by the generated constraints\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
+\begin{proof}
+  For the forward implication, we use induction to show that every constraint
+  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
+  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
+  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+  The reverse implication is trivial.
+\end{proof}
 
@@
 \end{theorem}
+
+\begin{proof}
+
+Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
+
+\TODO{The following is an informal sketch, requiring some formalization.}
+
+The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
+Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
+It is easy to show that we take care of these.
+
+The more difficult case are validity constraints, involving either region variables
+($\rho$) or predicate variables ($\varphi$) or both.
+
+The constraint-generation rules guarantee that every region variable will
+be unified with some region constant\footnote{
+There is one special case where this does not hold: a recursive function
+that calls itself in a non-terminating fashion. Such a function never
+returns a value, and so the return-value can be typed as anything.
+We assume that the return value of such a function is typed to be \C{unit},
+which will not introduce any region variables at the call-site of such a function.}
+This follows because $\FB$ does not admit uninitialized variables.
+% , and the type system uniquely determines the region parameters of every value
+
+We use this property to show that it is okay to ignore validity constraints
+in $\satC$ containing region variables. For every
+region variable $\rho$ unified with a constant $\pi$, and for every validity
+constraint containing $\rho$, we will have a corresponding validity constraint
+containing $\pi$ instead.
+Note: We likely need to add another rule to the saturation rules for computing
+$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
+We need this to propagate constraints from an outer scope into an inner scope.
+
+As for validity constraints containing a predicate variable $\varphi$ on the consequent,
+we argue that the use of the Substitution rule in computing $\satC$ makes it okay
+to ignore these constraints.
+
+As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
+we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
+us convert these into simpler constraints that do not have $\varphi$ on the
+antecedent or into a constraint that is directly incorporated into definition
+of $\predSubstFn$.
+\end{proof}
 

### 2016-10-10 paper/type-inference.tex
type inference proofs

@@
 \S~\ref{sec:overview}) to programs in \FB.
+For simplicity, we assume that each \C{letregion} and \C{open} construct
+in the input program introduces a distinct region identifier $\pi$.
 
@@
 
-%\item
-%\label{item:context}
-%Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
-%where $\pi_i$ and $\pi_j$ are both region constants.
-%If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
-%$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
-%Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
-%If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
-%
-%\item
-%    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
-%$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
-%If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
-%solver fails.
-%
-%\item
-%When a region variable $\rho$ is unified with a region constant $\pi$,
-%if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
-%and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
-%If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
-%\item
-%  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
-%  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
-%  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
 \end{enumerate}
@@
 Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
-% the special structure of the generated
-% constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
-% form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
-% region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-% iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
 
@@
 
-\begin{theorem}
-\label{thm:constraint-solver-sc}
+Consider a set of constraints $D$. We say that a region variable $\rho$
+\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
+constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
+We say that a region variable $\rho$ is \emph{bound} to a region constant
+$\pi$ in a context $\ell$ if
+$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
+We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
+$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.
+
+\begin{lemma}
+  \label{lemma:completely-bound}
 Let $C = \consOf{q}$.
-(Soundness) If the constraint solver succeeds, it will return an assignment $\soln$ 
-that satisfies $C$.
-(Completeness)
-Conversely, if $C$ is satisfiable, then the constraint solver will succeed.
-\end{theorem}
+Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
+\end{lemma}
 
 \begin{proof}
-
-Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
-
-\TODO{The following is an informal sketch, requiring some formalization.}
-
-The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
-Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
-It is easy to show that we take care of these.
-
-The more difficult case are validity constraints, involving either region variables
-($\rho$) or predicate variables ($\varphi$) or both.
-
 The constraint-generation rules guarantee that every region variable will
@@
 
-We use this property to show that it is okay to ignore validity constraints
-in $\satC$ containing region variables. For every
-region variable $\rho$ unified with a constant $\pi$, and for every validity
-constraint containing $\rho$, we will have a corresponding validity constraint
-containing $\pi$ instead.
 Note: We likely need to add another rule to the saturation rules for computing
@@
 
-As for validity constraints containing a predicate variable $\varphi$ on the consequent,
-we argue that the use of the Substitution rule in computing $\satC$ makes it okay
-to ignore these constraints.
-
-As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
-we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
-us convert these into simpler constraints that do not have $\varphi$ on the
-antecedent or into a constraint that is directly incorporated into definition
-of $\predSubstFn$.
+\TODO{Need to establish that there are no independent variables with same name.}
+\end{proof}
+
+\begin{lemma}
+Let $C = \consOf{q}$.
+Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
+and $\pi_2$ in $\satC$.
+Let $\soln$ satisfy $\satC$.
+For any context $\varphi \conj \phi$ that contains $\rho$, we must have
+$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
+$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
+\end{lemma}
+
+\begin{proof}
+  Assume that a region variable $\rho$ is bound to two different
+  region constants $\pi_1$ and $\pi_2$ in a context $\varphi \conj \phi$.
+  By transitivity, $\satC$ is guaranteed to have the constraints
+  $\isvalid{\varphi \conj \phi}{\pi_1 \outlives \pi_2}$
+  and $\isvalid{\varphi \conj \phi}{\pi_2 \outlives \pi_1}$.
+  By abduction decomposition, one of the following conditions must hold:
+  (a) $\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
+  both $\isvalid{\varphi}{\pi_1 \outlives \pi_2}$ and
+  $\isvalid{\varphi}{\pi_2 \outlives \pi_1}$ are in $\satC$, or
+  (b) $\{ \pi_1, \pi_2 \} \not\subseteq \predDeltaMap(\varphi)$ and
+  both $\isvalid{\phi}{\pi_1 \outlives \pi_2}$ and
+  $\isvalid{\phi}{\pi_2 \outlives \pi_1}$ are in $\satC$.
+  In case (b), $\satC$ will not be satisfiable, so we can ignore this case.
+  In case (a), $\varphi$ must be bound to a conjunction that includes both
+  $\pi_1 \outlives \pi_2$ and $\pi_2 \outlives \pi_1$ in any satisfying assignment
+  for $\satC$.
+\end{proof}
+
+\begin{lemma}
+Let $C = \consOf{q}$.
+Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
+that contain a region variable.
+$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
+\end{lemma}
+
+\begin{proof}
+  The forward implication is trivial.
+  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
+
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-soundness}
+Let $C = \consOf{q}$.
+If the constraint solver returns an assignment $\soln$,
+then $\soln$ satisfies $\satC$.
+\end{theorem}
+
+\begin{proof}
+Assume that the constraint solver returns an assignment $\soln$.
+
+By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
+is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.
+
+The definition of $\soln$ ensures that it satisfies the well-formedness constraints
+for any region variable $\rho$.
+
+Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
+where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
+ensures that $\soln$ satisfies this constraint.
+Furthermore, $\soln$ must also satisfy the well-formedness constraint
+$\tywf{predDeltaMap(\varphi)}{\varphi}$.
+(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
+be in $\satC$, which means $\groundC$ must be invalid.)
+
+Consider any constraint of the form $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ in $\satC$,
+It follows from abduction decomposition that either
+$\isvalid{\varphi}{\pi_i \outlives \pi_j}$
+or
+$\isvalid{\phi}{\pi_i \outlives \pi_j}$ must be in $\satC$, which we know is satisfied by $\soln$.
+Hence, $\soln$ must satisfy
+$\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ as well.
+
+This shows that $\soln$ satisfies all outlives-constraints in $\satC$ that have no variables in the consequent.
+
+Consider any constraint of the form $\isvalid{\ell}{F_j(\varphi_j)}$.
+Substituting the value $\predSubstFn(\varphi_j)$ reduces this constraint to one of
+the form $\isvalid{\ell}{\conj_i \pi_i \outlives \pi_i'}$ where the substitution rule
+guarantees $\isvalid{\ell}{\pi_i \outlives \pi_i'}$ is already in $\satC$ and, hence,
+satisfied by $\soln$.
+Hence, $\soln$ satisfies $\isvalid{\ell}{F_j(\varphi_j)}$ as well.
+
+Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
+of a region variable, say $\rho$.
+It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
+constant $\pi$.
+Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
+It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
+and that $\soln$ satisfies this constraint.
+Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
+Hence, $\soln$ satisfies this constraint as well.
+\TODO{Flesh this out.}
+
+The same idea works for constraints with occurrences of two region variables in the consequent.
+
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-completeness}
+Let $C = \consOf{q}$.
+If $\satC$ is satisfiable, then the constraint solver will return some assignment.
+\end{theorem}
+
+\begin{proof}
+If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
+We just need to show that the definition $\regionSubstFn(\rho)$ for any
+region variable $\rho$ will succeed.
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-sc}
+Let $C = \consOf{q}$.
+(Soundness) If the constraint solver succeeds, it will return an assignment $\soln$ 
+that satisfies $C$.
+(Completeness)
+Conversely, if $C$ is satisfiable, then the constraint solver will succeed.
+\end{theorem}
+
+\begin{proof}
+
+Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
+
+% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
+% 
+% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
+% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
+% It is easy to show that we take care of these.
+% 
+% The more difficult case are validity constraints, involving either region variables
+% ($\rho$) or predicate variables ($\varphi$) or both.
+% 
+% We use this property to show that it is okay to ignore validity constraints
+% in $\satC$ containing region variables. For every
+% region variable $\rho$ unified with a constant $\pi$, and for every validity
+% constraint containing $\rho$, we will have a corresponding validity constraint
+% containing $\pi$ instead.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
+% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
+% to ignore these constraints.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
+% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
+% us convert these into simpler constraints that do not have $\varphi$ on the
+% antecedent or into a constraint that is directly incorporated into definition
+% of $\predSubstFn$.
 \end{proof}
@@
   \item use of alpha-renaming, and its use to discharge not-in contraints;
-also, incorporate this into first stage rather than second (in Fig. 7).
   \item Note that we can also clarify that the well-formedness constraints are

### 2016-10-10 paper/type-inference.tex
type inference proof

@@
 Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
-and $\pi_2$ in $\satC$.
+and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
 Let $\soln$ satisfy $\satC$.
@@
 If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
-We just need to show that the definition $\regionSubstFn(\rho)$ for any
-region variable $\rho$ will succeed.
+If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
+$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
+Consider any such $\rho$.
+Let $\soln$ be a satisfying assignment for $\satC$.
+Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' in \regionDeltaMap(\rho)$.
+From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
+Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
+Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
+If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
+$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
+If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
+in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
+We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
+some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
+It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
+The result follows.
 \end{proof}

### 2016-10-11 paper/type-inference.tex
type inference

@@
 \newcommand{\groundC}{\myground{C}}
+\newcommand{\rhoC}{C_\rho}
+\newcommand{\solveCon}[1]{\textsc{Solve}(#1)}
 
@@
 \paragraph{Constraint Solver.}
-Our algorithm for solving a set of constraints $C$ consists of the following steps:
-\begin{enumerate}
-\item We use an iterative fixed point (saturation) algorithm to compute
-$\saturate{C}$, the set of all constraints logically implied by $C$.
-\item Let $\myground{C}$ denote the subset of all ground constraints
-(\ie, constraints without any region variable or predicate variable)
-in $\saturate{C}$. We check if $\myground{C}$ is valid in the theory of
-partial-order. If $\myground{C}$ is valid, the solution $\soln$ is determined
-as below. Otherwise, the constraint solver fails, as no valid solution exists.
-\begin{itemize}
-\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
-to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
-set $\{ \pi \in \regionDeltaMap(\rho) \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
-if this set is non-empty.
-Otherwise, the solver fails.
-%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
-% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
-%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
-\end{itemize}
-\end{enumerate}
-
-\paragraph{Step 1.}
+The first step in our algorithm for solving a set of constraints $C$ uses an
+iterative fixed point algorithm to compute a $\satC \supseteq C$ of
+constraints by adding  a set of constraints that are logically implied by $C$.
+% \paragraph{Step 1.}
 We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
@@
 
-Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
-
-\paragraph{Step 2.}
+% Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+
+\begin{theorem}
+  \label{thm:closure}
+Let $C = \consOf{q}$.
+An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+\end{theorem}
+
+\begin{proof}
+  For the forward implication, we use induction to show that every constraint
+  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
+  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
+  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+  The reverse implication is trivial.
+\end{proof}
+
+One key point of the first step is that it has the effect of unifying region variables with the
+region constants that they should be bound to in the solution, as explained below.
+
+Consider a set of constraints $D$. We say that a region variable $\rho$
+\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
+constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
+We say that a region variable $\rho$ is \emph{bound} to a region constant
+$\pi$ in a context $\ell$ if
+$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
+We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
+$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.
+
+\begin{lemma}
+  \label{lemma:completely-bound}
+Let $C = \consOf{q}$.
+Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
+\end{lemma}
+
+\begin{proof}
+The constraint-generation rules guarantee that every region variable will
+be unified with some region constant\footnote{
+There is one special case where this does not hold: a recursive function
+that calls itself in a non-terminating fashion. Such a function never
+returns a value, and so the return-value can be typed as anything.
+We assume that the return value of such a function is typed to be \C{unit},
+which will not introduce any region variables at the call-site of such a function.}
+This follows because $\FB$ does not admit uninitialized variables.
+% , and the type system uniquely determines the region parameters of every value
+
+Note: We likely need to add another rule to the saturation rules for computing
+$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
+We need this to propagate constraints from an outer scope into an inner scope.
+\end{proof}
+
+The set $\satC$ makes it easy to identify a solution $\soln$ (if one exists), as below.
+\begin{itemize}
+\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
+to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
+\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
+set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$.
+\end{itemize}
+
+% \paragraph{Step 2.}
+
+% \begin{enumerate}
+% \item We use an iterative fixed point (saturation) algorithm to compute
+% $\saturate{C}$, the set of all constraints logically implied by $C$.
+% \item
+
+The set $\satC$ also makes it easy to check if $C$ is satisfiable.
+Let $\myground{C}$ denote the subset of all ground constraints
+(\ie, constraints without any region variable or predicate variable)
+in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
+constraints for region variables in $\satC$.
+Define $\solveCon(C)$ as below.
+% It is possible to show that $\soln$ satisfies $\satC4
+% iff  $\myground{C}$ is valid in the theory of partial-order.
+
+% \end{enumerate}
+
+\[
+\solveCon{C} = 
+\begin{cases}
+\textsc{Some} (\soln) & \textrm{if $\groundC$ is valid and $\soln$ satisfies $\rhoC$} \\
+\textsc{None} & \textrm{otherwise}
+\end{cases}
+\]
+
+\begin{theorem}
+\label{thm:constraint-solver-sc}
+Let $C = \consOf{q}$.
+(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.
+(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
+\end{theorem}
+
+\begin{proof}
+
+Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
+
+% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
+% 
+% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
+% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
+% It is easy to show that we take care of these.
+% 
+% The more difficult case are validity constraints, involving either region variables
+% ($\rho$) or predicate variables ($\varphi$) or both.
+% 
+% We use this property to show that it is okay to ignore validity constraints
+% in $\satC$ containing region variables. For every
+% region variable $\rho$ unified with a constant $\pi$, and for every validity
+% constraint containing $\rho$, we will have a corresponding validity constraint
+% containing $\pi$ instead.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
+% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
+% to ignore these constraints.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
+% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
+% us convert these into simpler constraints that do not have $\varphi$ on the
+% antecedent or into a constraint that is directly incorporated into definition
+% of $\predSubstFn$.
+\end{proof}
+
 Checking the validity of a ground constraint is straightforward.
@@
 % Correctness follows by establishing:
-\begin{theorem}
-  \label{thm:closure}
-Let $C = \consOf{q}$.
-Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
-%% (b) $\satC$ is satisfiable iff $\myground{C}$ is valid.
-%% (c) For any region variable $\rho$, $\satC$ contains some constraint of the
-%% form $\isvalid{\ell}{\rho = \pi}$.
-\end{theorem}
-
-\begin{proof}
-  For the forward implication, we use induction to show that every constraint
-  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
-  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
-  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
-  The reverse implication is trivial.
-\end{proof}
-
-Consider a set of constraints $D$. We say that a region variable $\rho$
-\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
-constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
-We say that a region variable $\rho$ is \emph{bound} to a region constant
-$\pi$ in a context $\ell$ if
-$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
-We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
-$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.
-
-\begin{lemma}
-  \label{lemma:completely-bound}
-Let $C = \consOf{q}$.
-Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
-\end{lemma}
-
-\begin{proof}
-The constraint-generation rules guarantee that every region variable will
-be unified with some region constant\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
-
-Note: We likely need to add another rule to the saturation rules for computing
-$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
-We need this to propagate constraints from an outer scope into an inner scope.
-
-\TODO{Need to establish that there are no independent variables with same name.}
-\end{proof}
 
@@
 The result follows.
-\end{proof}
-
-\begin{theorem}
-\label{thm:constraint-solver-sc}
-Let $C = \consOf{q}$.
-(Soundness) If the constraint solver succeeds, it will return an assignment $\soln$ 
-that satisfies $C$.
-(Completeness)
-Conversely, if $C$ is satisfiable, then the constraint solver will succeed.
-\end{theorem}
-
-\begin{proof}
-
-Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
-
-% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
-% 
-% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
-% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
-% It is easy to show that we take care of these.
-% 
-% The more difficult case are validity constraints, involving either region variables
-% ($\rho$) or predicate variables ($\varphi$) or both.
-% 
-% We use this property to show that it is okay to ignore validity constraints
-% in $\satC$ containing region variables. For every
-% region variable $\rho$ unified with a constant $\pi$, and for every validity
-% constraint containing $\rho$, we will have a corresponding validity constraint
-% containing $\pi$ instead.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
-% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
-% to ignore these constraints.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
-% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
-% us convert these into simpler constraints that do not have $\varphi$ on the
-% antecedent or into a constraint that is directly incorporated into definition
-% of $\predSubstFn$.
 \end{proof}
@@
   \item use of alpha-renaming, and its use to discharge not-in contraints;
+\item also remind reader that variable names are unique
   \item Note that we can also clarify that the well-formedness constraints are

### 2016-10-12 paper/type-inference.tex
type inference proofs

@@
 \end{theorem}
-\begin{proof}
-The correspondence between the static semantics rules and the constraint generation
-rules induces a correspondence between derivation trees produced by the
-two systems.
-% We do not present a formal proof of the theorem.
-We can establish the following lemma inductively:
-given a derivation tree $\zeta_1$ for
-$\exprok {\stdcontext}{e}{\tau}{C}$
-and any substitution $\soln$,
-we can construct a corresponding \emph{candidate} derivation
-tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
-We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
-(We make use of analogous lemmas for type well-formedness rules as well.)
-\end{proof}
 
@@
 
-% \paragraph{Properties of generated constraints}
-Our constraint solver makes use of some special properties of the set of constraints $C$
-generated by our algorithm, which we now describe.
+We refer to any validity constraint whose antecedent contains a predicate variable
+(\ie, is of the form $\varphi \conj \phictxt$) as an \emph{abduction constraint}.
+Here,  $\phictxt$ is either \emph{true} (in which case, we call the constraint a trivial
+abduction constraint) or a conjunction of one or more outlives-constraints (in
+which case, we call the constraint a non-trivial abduction constraint).
+
+Non-trivial abduction constraints are a key challenge in solving the constraints.
+We now describe some special properties of the set of constraints $C$
+generated by our algorithm, which allow us to handle abduction constraints efficiently.
 
@@
 
-We refer to any validity constraint whose antecedent contains a predicate variable
-(\ie, is of the form $\varphi$ or $\varphi \conj \phictxt$) as an \emph{abduction
-constraint}. We refer to an abduction constraint in $C$ as \emph{decomposable}
+ We refer to an abduction constraint in $C$ as \emph{decomposable}
 if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
@@
 \end{lemma}
-
-\begin{proof}
-  By induction over the constraint-generation rules.
-  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
-  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
-  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
-  The only rule that modifies $\phicx$ is the rule for \C{letregion}
-  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
-  as conjuncts to $\phicx$.
-\end{proof}
 
@@
 
-\begin{proof}
-  Straightforward.
-\end{proof}
-
 Note that the above lemma shows how the conjunction in the antecedent of
@@
   \label{item:transitivity}
+\begin{enumerate}
+\item
 $\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
-$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$,
+$\isvalid{\ell \conj \phi}{\rho_2 \outlives \rho_3} \in \saturate{C}$
 $\Rightarrow$
-$\isvalid{\ell}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+\item
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
+$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$
+$\Rightarrow$
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+\end{enumerate}
 
@@
 
-\begin{proof}
-  For the forward implication, we use induction to show that every constraint
-  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
-  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
-  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
-  The reverse implication is trivial.
-\end{proof}
-
 One key point of the first step is that it has the effect of unifying region variables with the
@@
 
-\begin{proof}
-The constraint-generation rules guarantee that every region variable will
-be unified with some region constant\footnote{
-There is one special case where this does not hold: a recursive function
-that calls itself in a non-terminating fashion. Such a function never
-returns a value, and so the return-value can be typed as anything.
-We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
-% , and the type system uniquely determines the region parameters of every value
-
-Note: We likely need to add another rule to the saturation rules for computing
-$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
-We need this to propagate constraints from an outer scope into an inner scope.
-\end{proof}
-
 The set $\satC$ makes it easy to identify a solution $\soln$ (if one exists), as below.
@@
 \end{theorem}
-
-\begin{proof}
-
-Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
-
-% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
-% 
-% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
-% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
-% It is easy to show that we take care of these.
-% 
-% The more difficult case are validity constraints, involving either region variables
-% ($\rho$) or predicate variables ($\varphi$) or both.
-% 
-% We use this property to show that it is okay to ignore validity constraints
-% in $\satC$ containing region variables. For every
-% region variable $\rho$ unified with a constant $\pi$, and for every validity
-% constraint containing $\rho$, we will have a corresponding validity constraint
-% containing $\pi$ instead.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
-% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
-% to ignore these constraints.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
-% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
-% us convert these into simpler constraints that do not have $\varphi$ on the
-% antecedent or into a constraint that is directly incorporated into definition
-% of $\predSubstFn$.
-\end{proof}
 
@@
 and compute the solution using any Datalog engine.
-
-\paragraph{Soundness and Completeness of the Constraint Solver.}
-
-% Correctness follows by establishing:
-
-\begin{lemma}
-Let $C = \consOf{q}$.
-Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
-and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
-Let $\soln$ satisfy $\satC$.
-For any context $\varphi \conj \phi$ that contains $\rho$, we must have
-$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
-$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
-\end{lemma}
-
-\begin{proof}
-  Assume that a region variable $\rho$ is bound to two different
-  region constants $\pi_1$ and $\pi_2$ in a context $\varphi \conj \phi$.
-  By transitivity, $\satC$ is guaranteed to have the constraints
-  $\isvalid{\varphi \conj \phi}{\pi_1 \outlives \pi_2}$
-  and $\isvalid{\varphi \conj \phi}{\pi_2 \outlives \pi_1}$.
-  By abduction decomposition, one of the following conditions must hold:
-  (a) $\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
-  both $\isvalid{\varphi}{\pi_1 \outlives \pi_2}$ and
-  $\isvalid{\varphi}{\pi_2 \outlives \pi_1}$ are in $\satC$, or
-  (b) $\{ \pi_1, \pi_2 \} \not\subseteq \predDeltaMap(\varphi)$ and
-  both $\isvalid{\phi}{\pi_1 \outlives \pi_2}$ and
-  $\isvalid{\phi}{\pi_2 \outlives \pi_1}$ are in $\satC$.
-  In case (b), $\satC$ will not be satisfiable, so we can ignore this case.
-  In case (a), $\varphi$ must be bound to a conjunction that includes both
-  $\pi_1 \outlives \pi_2$ and $\pi_2 \outlives \pi_1$ in any satisfying assignment
-  for $\satC$.
-\end{proof}
-
-\begin{lemma}
-Let $C = \consOf{q}$.
-Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
-that contain a region variable.
-$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
-\end{lemma}
-
-\begin{proof}
-  The forward implication is trivial.
-  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
-
-\end{proof}
-
-\begin{theorem}
-\label{thm:constraint-solver-soundness}
-Let $C = \consOf{q}$.
-If the constraint solver returns an assignment $\soln$,
-then $\soln$ satisfies $\satC$.
-\end{theorem}
-
-\begin{proof}
-Assume that the constraint solver returns an assignment $\soln$.
-
-By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
-is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.
-
-The definition of $\soln$ ensures that it satisfies the well-formedness constraints
-for any region variable $\rho$.
-
-Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
-where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
-ensures that $\soln$ satisfies this constraint.
-Furthermore, $\soln$ must also satisfy the well-formedness constraint
-$\tywf{predDeltaMap(\varphi)}{\varphi}$.
-(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
-be in $\satC$, which means $\groundC$ must be invalid.)
-
-Consider any constraint of the form $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ in $\satC$,
-It follows from abduction decomposition that either
-$\isvalid{\varphi}{\pi_i \outlives \pi_j}$
-or
-$\isvalid{\phi}{\pi_i \outlives \pi_j}$ must be in $\satC$, which we know is satisfied by $\soln$.
-Hence, $\soln$ must satisfy
-$\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ as well.
-
-This shows that $\soln$ satisfies all outlives-constraints in $\satC$ that have no variables in the consequent.
-
-Consider any constraint of the form $\isvalid{\ell}{F_j(\varphi_j)}$.
-Substituting the value $\predSubstFn(\varphi_j)$ reduces this constraint to one of
-the form $\isvalid{\ell}{\conj_i \pi_i \outlives \pi_i'}$ where the substitution rule
-guarantees $\isvalid{\ell}{\pi_i \outlives \pi_i'}$ is already in $\satC$ and, hence,
-satisfied by $\soln$.
-Hence, $\soln$ satisfies $\isvalid{\ell}{F_j(\varphi_j)}$ as well.
-
-Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
-of a region variable, say $\rho$.
-It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
-constant $\pi$.
-Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
-It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
-and that $\soln$ satisfies this constraint.
-Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
-Hence, $\soln$ satisfies this constraint as well.
-\TODO{Flesh this out.}
-
-The same idea works for constraints with occurrences of two region variables in the consequent.
-
-\end{proof}
-
-\begin{theorem}
-\label{thm:constraint-solver-completeness}
-Let $C = \consOf{q}$.
-If $\satC$ is satisfiable, then the constraint solver will return some assignment.
-\end{theorem}
-
-\begin{proof}
-If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
-If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
-$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
-Consider any such $\rho$.
-Let $\soln$ be a satisfying assignment for $\satC$.
-Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' in \regionDeltaMap(\rho)$.
-From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
-Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
-Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
-If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
-$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
-If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
-in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
-We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
-some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
-It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
-The result follows.
-\end{proof}
 

### 2016-10-12 paper/typeinference-proofs.tex
type inference proofs

@@
-% \begin{theorem}
-% Let $p$, $q$, $r$ and $C$ denote the values of the corresponding
-% variables in an execution of the type inference algorithm.
-% \begin{enumerate}
-% \item $\absof{q}$ = $p$
-% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
-% \item If $\sigma$ is any solution to $C$, then $q[\sigma]$ is well-typed.
-% \item If there is any substitution $\sigma$ such that $q[\sigma]$ is well-typed, then
-% $C$ has a solution.
-% \item If SolveConstraints($C$) returns Some($s$), then $s$ is a solution to $C$.
-% \item If $C$ has a  solution, then SolveConstraints($C$) will return some solution.
-% \end{enumerate}
-% \end{theorem}
+\section{Type Inference: Proofs}
+
+\begin{theorem}
+\label{thm:constraint-generation-sc}
+Let $C = \consOf{q}$.
+Any assignment $\soln$ (for the region and predicate variables in $q$)
+satisfies $C$ iff $q[\soln]$ is well-typed.
+\end{theorem}
+\begin{proof}
+The correspondence between the static semantics rules and the constraint generation
+rules induces a correspondence between derivation trees produced by the
+two systems.
+% We do not present a formal proof of the theorem.
+We can establish the following lemma inductively:
+given a derivation tree $\zeta_1$ for
+$\exprok {\stdcontext}{e}{\tau}{C}$
+and any substitution $\soln$,
+we can construct a corresponding \emph{candidate} derivation
+tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
+We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
+(We make use of analogous lemmas for type well-formedness rules as well.)
+\end{proof}
+
+\begin{lemma}
+  \label{lemma:gc-is-decomposable}
+  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
+\end{lemma}
+
+\begin{proof}
+  By induction over the constraint-generation rules.
+  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
+  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
+  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
+  The only rule that modifies $\phicx$ is the rule for \C{letregion}
+  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
+  as conjuncts to $\phicx$.
+\end{proof}
+
+\begin{lemma}
+  \label{lemma:decomposition}
+  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+where both $\pi_i$ and $\pi_j$ are region constants.
+(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+iff
+$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
+\end{lemma}
+
+\begin{proof}
+  Straightforward.
+\end{proof}
+
+\begin{theorem}
+  \label{thm:closure}
+Let $C = \consOf{q}$.
+An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+\end{theorem}
+
+\begin{proof}
+  For the forward implication, we use induction to show that every constraint
+  added to $\satC$ preserves satisfiability (with respect to an assignment $\soln$).
+  The Transitivity and Substitution steps are straightforward. The Abduction Decomposition
+  step is justified by Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+  The reverse implication is trivial.
+\end{proof}
+
+\begin{lemma}
+  \label{lemma:completely-bound}
+Let $C = \consOf{q}$.
+Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
+\end{lemma}
+
+\begin{proof}
+The constraint-generation rules guarantee that every region variable will
+be unified with some region constant\footnote{
+There is one special case where this does not hold: a recursive function
+that calls itself in a non-terminating fashion. Such a function never
+returns a value, and so the return-value can be typed as anything.
+We assume that the return value of such a function is typed to be \C{unit},
+which will not introduce any region variables at the call-site of such a function.}
+This follows because $\FB$ does not admit uninitialized variables.
+% , and the type system uniquely determines the region parameters of every value
+\end{proof}
+
+\paragraph{Soundness and Completeness of the Constraint Solver.}
+\begin{lemma}
+Let $C = \consOf{q}$.
+Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
+and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
+Let $\soln$ satisfy $\satC$.
+For any context $\varphi \conj \phi$ that contains $\rho$, we must have
+$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
+$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
+\end{lemma}
+
+\begin{proof}
+  Assume that a region variable $\rho$ is bound to two different
+  region constants $\pi_1$ and $\pi_2$ in a context $\varphi \conj \phi$.
+  By transitivity, $\satC$ is guaranteed to have the constraints
+  $\isvalid{\varphi \conj \phi}{\pi_1 \outlives \pi_2}$
+  and $\isvalid{\varphi \conj \phi}{\pi_2 \outlives \pi_1}$.
+  By abduction decomposition, one of the following conditions must hold:
+  (a) $\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
+  both $\isvalid{\varphi}{\pi_1 \outlives \pi_2}$ and
+  $\isvalid{\varphi}{\pi_2 \outlives \pi_1}$ are in $\satC$, or
+  (b) $\{ \pi_1, \pi_2 \} \not\subseteq \predDeltaMap(\varphi)$ and
+  both $\isvalid{\phi}{\pi_1 \outlives \pi_2}$ and
+  $\isvalid{\phi}{\pi_2 \outlives \pi_1}$ are in $\satC$.
+  In case (b), $\satC$ will not be satisfiable, so we can ignore this case.
+  In case (a), $\varphi$ must be bound to a conjunction that includes both
+  $\pi_1 \outlives \pi_2$ and $\pi_2 \outlives \pi_1$ in any satisfying assignment
+  for $\satC$.
+\end{proof}
+
+\begin{lemma}
+Let $C = \consOf{q}$.
+Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
+that contain a region variable.
+$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
+\end{lemma}
+
+\begin{proof}
+  The forward implication is trivial.
+  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
+
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-soundness}
+Let $C = \consOf{q}$.
+If the constraint solver returns an assignment $\soln$,
+then $\soln$ satisfies $\satC$.
+\end{theorem}
+
+\begin{proof}
+Assume that the constraint solver returns an assignment $\soln$.
+
+By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
+is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.
+
+The definition of $\soln$ ensures that it satisfies the well-formedness constraints
+for any region variable $\rho$.
+
+Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
+where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
+ensures that $\soln$ satisfies this constraint.
+Furthermore, $\soln$ must also satisfy the well-formedness constraint
+$\tywf{predDeltaMap(\varphi)}{\varphi}$.
+(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
+be in $\satC$, which means $\groundC$ must be invalid.)
+
+Consider any constraint of the form $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ in $\satC$,
+It follows from abduction decomposition that either
+$\isvalid{\varphi}{\pi_i \outlives \pi_j}$
+or
+$\isvalid{\phi}{\pi_i \outlives \pi_j}$ must be in $\satC$, which we know is satisfied by $\soln$.
+Hence, $\soln$ must satisfy
+$\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ as well.
+
+This shows that $\soln$ satisfies all outlives-constraints in $\satC$ that have no variables in the consequent.
+
+Consider any constraint of the form $\isvalid{\ell}{F_j(\varphi_j)}$.
+Substituting the value $\predSubstFn(\varphi_j)$ reduces this constraint to one of
+the form $\isvalid{\ell}{\conj_i \pi_i \outlives \pi_i'}$ where the substitution rule
+guarantees $\isvalid{\ell}{\pi_i \outlives \pi_i'}$ is already in $\satC$ and, hence,
+satisfied by $\soln$.
+Hence, $\soln$ satisfies $\isvalid{\ell}{F_j(\varphi_j)}$ as well.
+
+Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
+of a region variable, say $\rho$.
+It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
+constant $\pi$.
+Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
+It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
+and that $\soln$ satisfies this constraint.
+Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
+Hence, $\soln$ satisfies this constraint as well.
+\TODO{Flesh this out.}
+
+The same idea works for constraints with occurrences of two region variables in the consequent.
+
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-completeness}
+Let $C = \consOf{q}$.
+If $\satC$ is satisfiable, then the constraint solver will return some assignment.
+\end{theorem}
+
+\begin{proof}
+If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
+If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
+$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
+Consider any such $\rho$.
+Let $\soln$ be a satisfying assignment for $\satC$.
+Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' in \regionDeltaMap(\rho)$.
+From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
+Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
+Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
+If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
+$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
+If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
+in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
+We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
+some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
+It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
+The result follows.
+\end{proof}
+
+\begin{theorem}
+\label{thm:constraint-solver-sc}
+Let $C = \consOf{q}$.
+(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.
+(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
+\end{theorem}
+
+\begin{proof}
+
+Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
+
+% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
+% 
+% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
+% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
+% It is easy to show that we take care of these.
+% 
+% The more difficult case are validity constraints, involving either region variables
+% ($\rho$) or predicate variables ($\varphi$) or both.
+% 
+% We use this property to show that it is okay to ignore validity constraints
+% in $\satC$ containing region variables. For every
+% region variable $\rho$ unified with a constant $\pi$, and for every validity
+% constraint containing $\rho$, we will have a corresponding validity constraint
+% containing $\pi$ instead.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
+% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
+% to ignore these constraints.
+% 
+% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
+% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
+% us convert these into simpler constraints that do not have $\varphi$ on the
+% antecedent or into a constraint that is directly incorporated into definition
+% of $\predSubstFn$.
+\end{proof}

### 2016-10-13 paper/abstract.tex
minor updates

--- 2016-06-29 paper/abstract.tex
@@
-Recent work in high-performance systems written in managed languages
-(such as Java or C\#) has shown that garbage-collection can be a
-significant performance bottleneck.  A class of these systems, focused
-on big-data, create numerous large data structures with well-defined
-lifetimes.  In this paper, we present a language and a memory
+% Recent work in high-performance systems written in managed languages
+% (such as Java or C\#) has shown that garbage-collection can be a
+% significant performance bottleneck.
+% A class of these systems, focused on big-data, create numerous large
+% data structures with well-defined lifetimes.
+%
+There is increasing interest in alternative memory management techniques
+that strike a balance between the convenience of garbage collection and
+the better performance that manual memory management enables, without
+compromising safety.
+%
+In this paper, we present a language and a memory
 management scheme based on user-managed memory regions (called
 \emph{transferable regions}) that allow programmers to exploit the
-knowledge of data structures' lifetimes to achieve significant
-performance improvements.
+knowledge of data structures' lifetimes to improve performance.
 
@@
 refinement-based region type system that ensures the memory safety of
-C\# programs in the presence of transferable regions. We complement
-our type system with a type inference algorithm that infers principal
-region types for first-order programs, and practically useful types
-for higher-order programs. This eliminates the need for programmers to
-write region annotations on types, while facilitating the reuse of
-existing C\# libraries with no modifications. Experiments demonstrate
-the practical utility of our approach.
+C\# programs in the presence of transferable regions.
+We complement our type system with a type inference algorithm.
+This eliminates the need for programmers to write region annotations on types.
+We establish the soundness of the type inference algorithm and
+show that the key steps of constraint-generation and constraint-solving
+in the type inference algorithm are complete.
+
+% while facilitating the reuse of existing C\# libraries with no modifications.
+% Experiments demonstrate the practical utility of our approach.

### 2016-10-13 paper/type-inference.tex
minor updates

@@
 % the last constraint is generated from the recursive call.
-
-\subsection{Miscellaneous points and to-dos}
-
-% \input{typeinference-proofs}
-
-\begin{itemize}
-  \item explain: non-determinism in defining region-variable binding is immaterial.
-  \item explain lemma 2's use case (a) and (b)
-  \item notation: decomposable is dependent on C
-  \item cleanup: syntax of constraints (region-constraint, validity constraint, etc.)
-    and later usage (abduction constraint)
-  \item adjust spacing in RULEs
-  \item for function-calls: need basic typing from core language to introduce region parameters
-  \item discuss computation of least-solution to constraints?
-  \item use of a set to represent conjunctions
-  \item equality is shorthand for two inequalities
-  \item use of alpha-renaming, and its use to discharge not-in contraints;
-\item also remind reader that variable names are unique
-  \item Note that we can also clarify that the well-formedness constraints are
-    syntactic sugar from basic set constraints
-  \item Prove: Progress: Every region-variable will be unified with some constant.
-  \item Notation: should we use a different notation to distinguish a validity constraint
-    from actual entailment?
-  \item update abstract/intro to stress aspects of constraint solver
-  \item Change static semantics to use method precondition, without class precondition
-\end{itemize}

### 2016-10-13 paper/type-inference.tex
minor

@@
 Let $C = \consOf{q}$.
-Any assignment $\soln$ (for the region and predicate variables in $q$)
+An assignment $\soln$ (for the region and predicate variables in $q$)
 satisfies $C$ iff $q[\soln]$ is well-typed.
@@
 $\predDeltaMap(\varphi)$.
-
-The formalism below is motivated by the following intuition.
-Suppose some construct in a method's body requires an outlives-constraint
-$\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
-of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
-as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
-$\pi_1 \outlives \pi_2$ must be established by the context within the method
-(\eg, by the \code{letregion} constructs in the method).
-We now generalize this idea to the setting of our abstract constraint-satisfication
-problem.
-
+%
+%% The formalism below is motivated by the following intuition.
+%% Suppose some construct in a method's body requires an outlives-constraint
+%% $\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
+%% of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
+%% as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
+%% $\pi_1 \outlives \pi_2$ must be established by the context within the method
+%% (\eg, by the \code{letregion} constructs in the method).
+%% We now generalize this idea to the setting of our abstract constraint-satisfication
+%% problem.
+%
 We refer to an abduction constraint in $C$ as \emph{decomposable}
@@
 a conjunct in $\phictxt$.
+%
+(The above concepts parametrically depend on the set $C$, though we omit $C$
+from the notation for simplicity.)
 
@@
 
-Note that the above lemma shows how the conjunction in the antecedent of
-any decomposable constraint can be eliminated.
+The above lemma shows how we can reduce a non-trivial abduction constraint to
+either a trivial abduction constraint or a non-abduction constraint.
 
 \paragraph{Constraint Solver.}
-The first step in our algorithm for solving a set of constraints $C$ uses an
-iterative fixed point algorithm to compute a $\satC \supseteq C$ of
-constraints by adding  a set of constraints that are logically implied by $C$.
-% \paragraph{Step 1.}
-We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
-point is reached:
+The first step in our algorithm for solving a set of constraints $C$
+computes a set $\satC \supseteq C$ of constraints by iteratively applying
+the following rules until a fixed point is reached:
 
@@
 % Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
+We can show that every new constraint added by the above rules is implied by existing
+constraints:
 
@@
 
-One key point of the first step is that it has the effect of unifying region variables with the
-region constants that they should be bound to in the solution, as explained below.
-
+A key goal of this step is to identify the value every region variable must
+have in any solution of the set of constraints, as explained below.
+%
+% unifying region variables with the
+% region constants that they should be bound to in the solution, as explained below.
+%
 Consider a set of constraints $D$. We say that a region variable $\rho$
@@
 constraints for region variables in $\satC$.
-Define $\solveCon(C)$ as below.
+Define $\solveCon{C}$ as below.
 % It is possible to show that $\soln$ satisfies $\satC4
@@
 an edge from $\rho_1$ to $\rho_2$.
-(Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
-for the pair of constraints $\rho_1 \outlives \rho_2$ and
-$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
-by a pair of edges.)
+%
+%% (Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
+%% for the pair of constraints $\rho_1 \outlives \rho_2$ and
+%% $\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
+%% by a pair of edges.)
 %
@@
 
-This idea generalizes. Consider a set $C$ of constraints that contains no
-constraint with an antecedent $\varphi \conj \phi$ where $\phi$ is a non-empty conjunction. 
-Then, the algorithm for computing the closure $\satC$ does not need to use
-rule~\ref{item:context} (for Abduction Decomposition). The closure in this
-case can be computed using context-free reachability in graphs~\cite{Reps:Reachability}
+This idea generalizes.
+% Consider a set $C$ of constraints that contains no non-trivial abduction constraint.
+% Then, the algorithm for computing the closure $\satC$ does not need to use
+% rule~\ref{item:context} (for Abduction Decomposition).
+Extending the simple graph reachability algorithm to incorporate the Substitution rule
+(in computing $\satC$) turns the problem into a context-free reachability problem in graphs~\cite{Reps:Reachability}
 (as usual for context-sensitive interprocedural analysis).
 
-Algorithms for context-free reachability can be adapted to incorporate
-the Abduction Decomposition step.
+Algorithms for context-free reachability can be adapted to incorporate the Abduction Decomposition step.
+Alternatively, the iterative process described above is a standard fixed point computation
+and can be encoded using a set of Datalog rules, allowing us to compute the closure
+using any Datalog engine.
 % Further details omitted.
-Alternatively, the iterative process described above is a standard fixed point computation.
-We can encode this computation using a set of Datalog rules,
-and compute the solution using any Datalog engine.
 

### 2016-10-13 paper/typeinference-proofs.tex
minor

@@
 
-\paragraph{Soundness and Completeness of the Constraint Solver.}
 \begin{lemma}
+  \label{lemma:two-bindings}
 Let $C = \consOf{q}$.
@@
 
-\begin{lemma}
-Let $C = \consOf{q}$.
-Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
-that contain a region variable.
-$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
-\end{lemma}
-
-\begin{proof}
-  The forward implication is trivial.
-  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
-
-\end{proof}
+% \begin{lemma}
+% Let $C = \consOf{q}$.
+% Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
+% that contain a region variable.
+% $\satC$ is satisfiable iff $\satC_1$ is satisfiable.
+% \end{lemma}
+% 
+% \begin{proof}
+%   The forward implication is trivial.
+%   The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.
+% \end{proof}
 
@@
 Let $C = \consOf{q}$.
-If the constraint solver returns an assignment $\soln$,
+If $\solveCon{C} = \textsc{Some}(\soln)$,
 then $\soln$ satisfies $\satC$.
@@
 \begin{proof}
-Assume that the constraint solver returns an assignment $\soln$.
+Assume $\solveCon{C} = \textsc{Some}(\soln)$.
 
-By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
+By definition (of $\solveCon{C}$), $\groundC$ (the set of all variable-free constraints in $\satC$)
 is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.
-
-The definition of $\soln$ ensures that it satisfies the well-formedness constraints
+%
+The definition also ensures that $\soln$ satisfies the well-formedness constraints
 for any region variable $\rho$.
@@
 Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
-where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
+where $\pi_i$ and $\pi_j$ are distinct region constants. The definition of $\predSubstFn(\varphi)$
 ensures that $\soln$ satisfies this constraint.
 Furthermore, $\soln$ must also satisfy the well-formedness constraint
-$\tywf{predDeltaMap(\varphi)}{\varphi}$.
-(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
-be in $\satC$, which means $\groundC$ must be invalid.)
+$\tywf{\predDeltaMap(\varphi)}{\varphi}$.
+(Otherwise, the application of Abduction Decomposition to this constraint will add
+$\isvalid{}{\pi_i \outlives \pi_j}$ to $\satC$. This constraint is invalid, which
+contradicts the fact that $\groundC$ is valid.)
 
@@
 of a region variable, say $\rho$.
-It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
+It follows from Lemma~\ref{lemma:completely-bound} that $\rho$ is bound to some region
 constant $\pi$.
 Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
-It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
-and that $\soln$ satisfies this constraint.
-Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
-Hence, $\soln$ satisfies this constraint as well.
-\TODO{Flesh this out.}
+It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$.
+Since this constraint has no variables on the consequent, we have already shown that $\soln$
+must satisfy this constraint ($\isvalid{\ell}{r'}$).
+Let $\pi' = \regionSubstFn(\rho)$.
+We can show (as in the proof of Lemma~\ref{lemma:two-bindings}) that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$.
+Hence, $\soln$ satisfies $\isvalid{\ell}{r}$ as well.
 
@@
 Let $\soln$ be a satisfying assignment for $\satC$.
-Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' in \regionDeltaMap(\rho)$.
+Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
 From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
@@
 \label{thm:constraint-solver-sc}
-Let $C = \consOf{q}$.
-(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.
+Let $C = \consOf{q}$.\\
+(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.\\
 (Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.

### 2016-10-14 paper/type-inference.tex
proofs

@@
 \newcommand{\soln}{\eta}
+\newcommand{\solnR}{\soln_R}
+\newcommand{\solnP}{\soln_P}
+\newcommand{\thesoln}{\hat{\eta}}
+\newcommand{\thesolnR}{\thesoln_R}
+\newcommand{\thesolnP}{\thesoln_P}
+
 \newcommand{\saturate}[1]{{#1}^*}
@@
 
-The set $\satC$ makes it easy to identify a solution $\soln$ (if one exists), as below.
+The set $\satC$ makes it easy to identify a solution $\thesoln$ (if one exists), as below.
 \begin{itemize}
-\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
+\item For any predicate variable $\varphi$, $\thesolnP(\varphi)$ is defined
 to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
+\item For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
 set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$.
@@
 \begin{cases}
-\textsc{Some} (\soln) & \textrm{if $\groundC$ is valid and $\soln$ satisfies $\rhoC$} \\
+\textsc{Some} (\thesoln) & \textrm{if $\groundC$ is valid and $\thesoln$ satisfies $\rhoC$} \\
 \textsc{None} & \textrm{otherwise}

### 2016-10-14 paper/typeinference-proofs.tex
proofs

@@
   \label{lemma:two-bindings}
-Let $C = \consOf{q}$.
+Let $C = \consOf{q}$. Assume that $\groundC$ is valid.
 Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
 and $\pi_2$ in a context $\varphi \conj \phi$. % in $\satC$.
-Let $\soln$ satisfy $\satC$.
-For any context $\varphi \conj \phi$ that contains $\rho$, we must have
-$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
-$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
+% Let $\soln$ satisfy $\satC$.
+% For any context $\varphi \conj \phi$ that contains $\rho$, we must have
+Then, 
+(a) $\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$,
+(b) $\isvalid{\thesolnP(\varphi)}{\pi_1 = \pi_2}$.
+(c) For any $\soln$ that satisfies $\satC$, we must have $\isvalid{\solnP(\varphi)}{\pi_1 = \pi_2}$.
 \end{lemma}
@@
 Let $\pi' = \regionSubstFn(\rho)$.
-We can show (as in the proof of Lemma~\ref{lemma:two-bindings}) that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$.
+We can show (using Lemma~\ref{lemma:two-bindings}) that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$.
 Hence, $\soln$ satisfies $\isvalid{\ell}{r}$ as well.
@@
 Let $C = \consOf{q}$.
-If $\satC$ is satisfiable, then the constraint solver will return some assignment.
+If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
+% If $\satC$ is satisfiable, then the constraint solver will return some assignment.
 \end{theorem}

### 2016-10-14 paper/abstract.tex
proofs and space compression

@@
 %
-In this paper, we present a language and a memory
-management scheme based on user-managed memory regions (called
-\emph{transferable regions}) that allow programmers to exploit the
-knowledge of data structures' lifetimes to improve performance.
+In this paper, we present a language and a memory management scheme based
+on user-managed memory regions (called \emph{transferable regions})
+that allow programmers to exploit knowledge of data structures' lifetimes
+to improve performance.
 
@@
 dangling pointers. A key contribution of this paper is a
-refinement-based region type system that ensures the memory safety of
-C\# programs in the presence of transferable regions.
-We complement our type system with a type inference algorithm.
-This eliminates the need for programmers to write region annotations on types.
-We establish the soundness of the type inference algorithm and
-show that the key steps of constraint-generation and constraint-solving
+refinement-based region type system, for a C\#-like typed object-oriented
+language with higher-order functions and generics, that ensures the
+memory safety of programs in the presence of transferable regions.
+We complement our type system with a type inference algorithm,
+which eliminates the need for programmers to write region annotations on types.
+The cornerstone of our inference algorithm is a novel constraint solver that
+performs abduction in a partial-order constraint domain to infer weakest solutions
+to recursive constraints.
+We establish soundness and partial completeness results for the type inference algorithm:
+in particular, the key steps of constraint-generation and constraint-solving
 in the type inference algorithm are complete.

### 2016-10-14 paper/typeinference-proofs.tex
proofs and space compression

@@
 If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
-% If $\satC$ is satisfiable, then the constraint solver will return some assignment.
 \end{theorem}
@@
 \begin{proof}
-If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
-If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
-$\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
-Consider any such $\rho$.
-Let $\soln$ be a satisfying assignment for $\satC$.
-Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
-From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
-Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
-Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
-If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
-$\pi \in S_\rho$ and $S_\rho$ cannot be empty.
-If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
-in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
-We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
-some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
-It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
-The result follows.
+We prove the theorem by contradiction.
+Assume that $\satC$ is satisfiable.
+We will show that the precondition for the first case in the definition of $\textsc{Solve}$
+holds and, hence, $\solveCon{C}$ cannot be $\textsc{None}$.
+
+Since $\satC$ is satisfiable, $\groundC$ is trivially valid.
+We just need to show that $\thesolnR$ satisfies $\rhoC$.
+Recall that $\rhoC$ is the subset of $C$ of well-formedness constraints
+on region variables.
+So, we just need to show for any constraint $\rho \in \rhoenv$ in $C$ that
+$\thesolnR(\rho) \in \rhoenv$.
+Since we have assumed that $\satC$ is satisfiable, there exists some $\soln$ that satisfies $\satC$
+and, in particular, the constraint $\rho \in \rhoenv$.
+Let $\pi$ denote $\soln(\rho)$. Thus, we have $\pi \in \rhoenv$.
+
+Let $\hat{\pi}$ denote $\thesolnR(\rho)$. If $\pi$ and $\hat{\pi}$ are the same region
+constant, then $\hat{\pi}\ \in \rhoenv$ and the proof is complete.
+%
+Otherwise, it follows from the definition of $\thesoln$, that we have some constraint
+$\isvalid{\varphi \conj \phi}{\rho = \hat{\pi}}$ in $\satC$.
+Since $\soln$ satisfies this constraint, we have
+$\isvalid{\solnP(\varphi) \conj \phi}{\pi = \hat{\pi}}$.
+We can show that $\{ \pi, \hat{\pi} \} \subseteq \predDeltaMap(\varphi)$
+(as in the proof of Lemma~\ref{lemma:two-bindings}).
+We can show, by induction over the constraint-generation rules, that if $\Delta$
+includes some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
+Hence, it follows that $\hat{\pi} \in \Delta$.
+% for some predicate variable $\varphi$.
+
+% If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
+% $\{ \pi \in \regionDeltaMap(\rho) \; | \; \isvalid{\ell}{\rho = \pi} \in \satC \}$ is empty for some region variable $\rho$.
+% Consider any such $\rho$.
+% Let $\soln$ be a satisfying assignment for $\satC$.
+% Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
+% From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.
+% Thus, $\isvalid{\ell}{\rho = \pi} \in \satC$.
+% Since $\soln$ satisfies this constraint, we have $\isvalid{\ell[\soln]}{\pi' = \pi}$.
+% If $\pi$ and $\pi'$ are the same region identifier, we have a contradiction, since
+% $\pi \in S_\rho$ and $S_\rho$ cannot be empty.
+% If $\pi$ and $\pi'$ are distinct region identifiers, then it follows that the precondition of case (a) must hold
+% in Lemma~\ref{lemma:decomposition}, and we must have $\{ \pi, \pi' \} \subseteq \predDeltaMap(\varphi)$.
+% We can show, by induction over the constraint-generation rules, that if $\regionDeltaMap(\rho)$ includes
+% some element of $\predDeltaMap(\varphi)$, then it includes all elements of $\predDeltaMap(\varphi)$.
+% It follows that $\pi \in \regionDeltaMap(\rho)$. Hence, the set $S_\rho$ is non-empty (as it contains $\pi$.
+% The result follows.
 \end{proof}
@@
 
-% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
-% 
-% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
-% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
-% It is easy to show that we take care of these.
-% 
-% The more difficult case are validity constraints, involving either region variables
-% ($\rho$) or predicate variables ($\varphi$) or both.
-% 
-% We use this property to show that it is okay to ignore validity constraints
-% in $\satC$ containing region variables. For every
-% region variable $\rho$ unified with a constant $\pi$, and for every validity
-% constraint containing $\rho$, we will have a corresponding validity constraint
-% containing $\pi$ instead.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
-% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
-% to ignore these constraints.
-% 
-% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
-% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
-% us convert these into simpler constraints that do not have $\varphi$ on the
-% antecedent or into a constraint that is directly incorporated into definition
-% of $\predSubstFn$.
 \end{proof}

### 2016-10-14 paper/abstract.tex
various

@@
 memory safety of programs in the presence of transferable regions.
-We complement our type system with a type inference algorithm,
-which eliminates the need for programmers to write region annotations on types.
-The cornerstone of our inference algorithm is a novel constraint solver that
-performs abduction in a partial-order constraint domain to infer weakest solutions
-to recursive constraints.
-We establish soundness and partial completeness results for the type inference algorithm:
-in particular, the key steps of constraint-generation and constraint-solving
-in the type inference algorithm are complete.
+We complement our type system with a type inference algorithm, which eliminates the need for programmers to write region annotations on types.
+% The cornerstone of our inference algorithm is a novel constraint solver that
+% performs abduction in a partial-order constraint domain to infer weakest solutions
+% to recursive constraints.
+We establish soundness and partial completeness results for the type inference algorithm: in particular, the key steps of constraint-generation and constraint-solving in the type inference algorithm are complete.
+We describe a prototype implementation of our inference algorithm and the constraint solver, and our experience of using it to enforce memory safety in dataflow programs.
 

### 2016-10-14 paper/type-inference.tex
proofs and space compression
various

@@
 \newcommand{\groundC}{\myground{C}}
-\newcommand{\rhoC}{C_\rho}
+\newcommand{\rhoC}{\text{WF}_R}
 \newcommand{\solveCon}[1]{\textsc{Solve}(#1)}
@@
 
+\input{constraint-syntax}
+
 % \emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
@@
 Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
-%% over a set of fixed formal region parameters,
-where the set of region-constraints $\phi$ is defined by:
-\begin{smathpar}
-\begin{array}{lcl}
-\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
-\end{array}
-\end{smathpar}
-We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
-We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
-as shorthand for $\conj_{\phi \in S} \phi$.
-
+over a set of fixed formal region parameters.
+%% where the set of region-constraints $\phi$ is defined by:
+%% \begin{smathpar}
+%% \begin{array}{lcl}
+%% \phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
+%% \end{array}
+%% \end{smathpar}
+%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
+%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
+%% as shorthand for $\conj_{\phi \in S} \phi$.
+%% 
 We will use the term \emph{validity constraint} to denote an entailment constraint
@@
 Our constraints also make uses of \emph{pending substitutions} $F$:
-% \footnote{we borrowed this terminology from~\cite{ltpldi08}}
 A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
 used in a particular context:
-\begin{smathpar}
-\begin{array}{lcl}
-F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
-\end{array}
-\end{smathpar}
+%% \begin{smathpar}
+%% \begin{array}{lcl}
+%% F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
+%% \end{array}
+%% \end{smathpar}
 E.g., in the validity constraint $\isvalid{\rgn_1 \outlives

### 2016-10-15 paper/type-system.tex
Static semantics for locations-based opsem.

--- 2016-07-07 paper/type-system.tex
@@
 region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
-$R_2$ is guaranteed to outlive $R_1$. 
-We now formally develop this intuition via \fbname (\FB), our
-explicitly typed core language (with region types) that incorporates
-the features introduced in the previous section. \fbname builds on the
-Featherweight Generic Java (FGJ)~\cite{fgj} formalism, and reuses
-notations and various definitions from~\cite{fgj}, such as the
-definition of type well-formedness for the core (region-free)
-language. 
-(The language used in Section~\ref{sec:overview} is essentially
-a version of FGJ without region types and with some syntactic sugar.)
+$R_2$ is guaranteed to outlive $R_1$. Intuitively, the invariant can
+be enforced by (a). tracking \emph{outlives} relationships between
+various regions in the program (b). tagging the C\# type of every
+object in the program with its allocation region, and (c).  ensuring
+that, when a reference is created, its target object ($o_2$) is
+allocated in a region that is known to be in outlives relationship
+with the object containing the reference. We now formally develop this
+intuition via \fbname (\FB), our explicitly typed core language (with region
+types) that incorporates the features introduced in the previous
+section. \fbname builds on the Featherweight Generic Java
+(FGJ)~\cite{fgj} formalism, and reuses notations and various
+definitions from~\cite{fgj}, such as the definition of type
+well-formedness for the core (region-free) language.
 
@@
 %
-The following definition of \C{Pair} class in \FB illustrates some of
-the key elements of the formal language (the symbol $\extends$ should
-be read \emph{extends}, and the symbol $\outlives$ stands for
-\emph{outlives}):
+The following definition of \C{Pair} class in \FB illustrates some
+of the key elements of the formal language\footnote{The symbol
+$\extends$ should be read \emph{extends}, and the symbol $\outlives$
+stands for \emph{outlives}}:
 \begin{codejava}[mathescape=true]
@@
   b@$\rho_2$ snd;
+  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
+    super(); this.fst = fst; this.snd = snd;
+  }
   a@$\rho_1$ getFst() { 
@@
 \end{codejava}
-% \vspace*{-0.15in}
-%  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
-%    super(); this.fst = fst; this.snd = snd;
-%  }
-Note that we elide showing constructors since they are uninteresting
-from the type system's standpoint; their behavior in \FB is same as in
-FGJ.
+\vspace*{-0.15in}
 
-A class in \FB is parametric over zero or more type variables (as
-in FGJ) as well as one or more region variables $\rho$.
+A class in \FB is parametric over zero or more type variables (as in
+FGJ) as well as one or more region variables $\rho$.
 % In the above example, the class \C{Pair} is parameterized over
@@
 In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
-$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
+$\rgn_0$, and $\rho_2$ with $\rgn_1$ is allowed because (a) $\rgn_0$
 and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
@@
 % context. Every class definition in \FB is necessarily polymorphic with
-% respect to the allocation region of its objects, i.e., the
-% allocation
+% respect to the allocation region of its objects, i.e., the allocation
 % context of its constructor. We adopt a convention that requires the
@@
 % method in \FB. 
-If a method is not intended to be polymorphic with respect to its
-allocation context (for example, if its allocation context needs to be
-same as the allocation region of its \emph{this} argument), then the
-required monomorphism can be captured as an equality constraint in
-$\phi$.
+Note that if a method is not intended to be polymorphic with respect
+to its allocation context (for example, if its allocation context
+needs to be same as the allocation region of its \emph{this}
+argument), then the required monomorphism can be captured as an
+equality constraint in $\phi$.
 
@@
 \item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
-variables to their \emph{bounds}, i.e., classes they are declared to
-extend, and
+variables to their bounds\footnote{A bound of a type variable
+($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
+was declared to extend.}, and
 \item A constraint formula ($\phicx$) that captures the outlives
@@
 ($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
-judgments on core types. We use a double-piped turnstile ($\Vdash$)
-for judgments in FGJ~\cite{fgj}, and a simple turnstile ($\vdash$) for
-those in \FB. The class table ($\absof{CT}$) for FGJ judgments is
-derived from \FB's class table ($CT$) by erasing all region
-annotations on types, and region arguments in expressions
-($\absof{\cdot}$ denotes the region erasure operation). 
-% For judgments in FGJ, we use a
-% class table that maps class names to their region-erased definitions.
-% We denote the region-erased class table as ($\absof{CT}$). The precise
-% semantics of region erasure ($\absof{\cdot}$) are defined
-% in~\cite{techrep}, but it can be understood as an operation that
-% erases region annotations on types. The class table is assumed to bind
-% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
-% This mostly eliminates the need to treat $\RgnZ$ objects, which double
-% up as transferable region handlers, as a special case in semantics.
-The well-formedness rule for class types
+judgments on core types \footnote{We use a double-piped turnstile
+($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
+($\vdash$) for those in \FB.}. The class table ($\absof{CT}$) for such
+judgments is derived from \FB's class table ($CT$) by erasing all
+region annotations on types, and region arguments in expressions
+($\absof{\cdot}$ denotes the region erasure operation).  The
+well-formedness rule for class types
 ($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
@@
 parameters. The later is enforced by checking the validity of $\phi$,
-with actual region arguments substituted for formal
+with actual region arguments substituted\footnote{The notation
+$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
 region parameters, under the conditions ($\A.\phicx$) guaranteed by
@@
 
+%\input{fb-opsem}
+
 The type rules distinguish between the \C{new} expressions that create
 objects of  the \RgnZ class, and \C{new} expressions that create
-objects of other classes. The rule for the latter relies on an
+objects of other classes. The rule for the later relies on an
 auxiliary definition called $\fields$ that returns the sequence of
@@
 preventing cross-region references originating from transferable
-regions. The body ($e$) of the function might, however, create
+regions\footnote{The body of the function ($e$) might, however, create
 new (transferable) regions while execution, but that is fine as long
 as such regions, and objects allocated in them, don't find their way
-into the result of evaluating $e$.
+into the result of evaluating $e$}.
 
@@
 
-Fig.~\ref{fig:fb-opsem} in the appendix defines a small-step operational semantics for \fbname
-via a five-place reduction relation:
-% \vspace*{-0.15in}
+Fig.~\ref{fig:fb-opsem-1} and Fig.~\ref{fig:fb-opsem-2} define a small-step
+operational semantics for \fbname via a five-place reduction relation:
+\vspace*{-0.15in}
 \begin{smathpar}
@@
 identifier ($\rgn$) for this region. A binding is also added to
-$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
-defines a type rule for such values, allowing them to be typed.
-The following type safety theorem shows that a well-typed program will never
-attempt to dereference an ``invalid'' reference (a reference to an object
-in a region that has been transferred or freed):
+$\rhomap$, mapping $\rgn$ to the ``closed'' state.
+Fig.~\ref{fig:fb-staticsem} defines a type rule for such values,
+allowing them to be typed.  Type safety theorem is now stated
+thus\footnote{Theorems~\ref{thm:fb-type-safety-stmt}
+and~\ref{thm:fb-transfer-safety-stmt} are restated and proved in the
+appendix}:
 \begin{theorem}
 \emph{(\textbf{Type Safety})}
-\label{thm:fb-type-safety}
+\label{thm:fb-type-safety-stmt}
 $\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
@@
 $e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
-and $\hastyp{\emptyASigp,\rgn, \cdot}{e'}{\tau}$.
+and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
 \end{theorem}
@@
 \emph{(\textbf{Transfer Safety})}
-\label{thm:fb-transfer-safety}
+\label{thm:fb-transfer-safety-stmt}
 $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
-$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in
-\rhoenv'$, $\rgn_i \neq \toprgn$ and
-$\rgn_i \notin dom(\rhomap') \cup \rhoenv'$, if $\hastyp{\emptyA,\rgn,
-\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
-$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
-{\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
+$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
+$\rgn_i \notin dom(\rhomap') \cup \rhoenv \cup \{\toprgn\}$, if
+\[
+  \hastyp{\emptyA,\rgn, \cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
+\]
+, then
+\[
+  \hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
+{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
+\]
 \end{theorem}
@@
 needs to add a binding for the region to its $\rhomap$ in order to
-preserve its type safety. 
-% Notably, this result could not have been
-% established if it were possible for a transferable region to contain
-% references to objects outside the region.
+preserve its type safety. Notably, this result could not have been
+established if it were possible for a transferable region to contain
+references to objects outside the region.

### 2016-10-18 paper/implementation.tex
Typos fixed.
proofs and space compression
minor changes

@@
 \label{sec:implementation}
+
+As mentioned earlier, our work is a continuation of the work reported in~\cite{Broom:HotOS},
+which provides users with a C\#-based implementation of \broom (the language described in
+Section~\ref{sec:overview}). This system has no type system and provides users with no safety
+guarantees, but does shed light on the performance gains as well as the viability of the
+approach (in terms of expressiveness). In particular, it reports speedups up to 34\% for
+typical big-data analytics jobs.
+We now describe our implementation and experience with the type system and type inference
+algorithm presented in the current paper.
 
@@
 class LinkedList<T><R5,R4 | R4$\outlives$R5> {
-  ListNode<T><R5,R4> head;
-  ..
+  ListNode<T><R5,R4> head; ...
   List<T><R17,R4> rev<R17,R4 | R4$\outlives$R17>(unit u) {
-    List<T><R17,R4> xs = 
-        new List<T><R17,R4>(this.head.val);
+    List<T><R17,R4> xs = new List<T><R17,R4>(this.head.val);
     ListNode<T><R5,R4> cur = this.head.next;
@@
       xs.add<R17>(cur.val)
-      cur = cur.next;
-    }
+      cur = cur.next; }
     return xs;
@@
 generation in \namec, we could not measure the runtime overhead of
-checks needed to enforce safety. However, the number of LOC peforming
+checks needed to enforce safety. However, the number of LOC performing
 operations on \C{Region} objects relative to the total LOC never

### 2016-10-18 paper/introduction.tex
Typos fixed.
minor changes

@@
 
-In this paper, we present a \emph{safe} and \emph{efficient} region-based approach
+This paper presents a \emph{safe} and \emph{efficient} region-based approach
 to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
@@
 Consider the example (of  a producer creating a data structure and transferring it to a consumer)
-in Fig.~\ref{fig:motivating-eg}. This represents a \C{SELECT} query operator for streaming data. 
-The operator receives a stream of input messages, each
-associated with a time (window) $t$, processed by method
-\texttt{onReceive}.  Each input message contains a list of inputs,
-each of which is processed by applying a user-defined function to
-create a corresponding output.  The operator may receive multiple
-messages with the same timestamp (and messages with different
-timestamps may be delivered out of order). A timing-message (an
-invocation of method \texttt{OnNotify}) indicates that no more input
-messages with a timestamp $t$ will be subsequently delivered.  At this
-point, the operator completes the processing for time window $t$ and
-sends a corresponding output message to its successor.
+in Fig.~\ref{fig:motivating-eg}.
+This represents a \C{SELECT} query operator that receives a stream of input messages,
+each associated with a time window $t$, processed by method \texttt{onReceive}.
+Each input message contains a list of inputs, each processed by applying a user-defined function to
+create a corresponding output.  Multiple messages with the same timestamp
+are permitted and messages with different timestamps may arrive out of order.
+An invocation of method
+\texttt{OnNotify} indicates that no more input messages with a timestamp
+$t$ will be subsequently delivered.  At this point, the operator completes
+the processing for time window $t$ and sends a corresponding output message
+to its successor.
 
@@
 \paragraph{Performance.}
-When the transferred data structures are large, as is the case with many streaming big-data
-analysis systems, garbage collection overhead becomes
-significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
+When the transferred data structures are large, as is the case with many
+streaming big-data analysis systems, garbage collection overhead becomes
+significant~\cite{Broom:HotOS}.  Further, in a distributed
 dataflow system, the GC pause at one node can have a cascading adverse
@@
 a GC pause at an upstream actor can block downstream actors that are waiting for
-messages.  However, often much of the GC overhead results from the
+messages.  However, much of the GC overhead results from the
 collector performing avoidable or unproductive work.  For the example
 in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
-\C{map} data structure, although its objects cannot be collected until a
+\C{map}, although its objects cannot be collected until a
 suitable timing message arrives.
@@
 (as determined by the compiler and runtime).
-We wish to ensure memory safety in the presence of such transferred data by ensuring
-that there are no invalid references. A transfer operation, with no
+We wish to ensure memory safety in the presence of such transferred data.
+% by ensuring that there are no invalid references.
+A transfer operation, with no
 additional checks, may cause memory safety violations, both at the
@@
 invalid in the context of the consumer. Both scenarios are depicted in
-Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
-particularly unwelcome in the context of languages like C\# and Java
-that are otherwise memory-safe. While references from within the
-transferable data to outside can be disallowed in the interest of safety,
+Fig.~\ref{fig:transfer-violations}.
+% Safety violations of this kind are
+% particularly unwelcome in the context of languages like C\# and Java
+% that are otherwise memory-safe.
+While references from within the transferable data to outside can be disallowed
+in the interest of safety,
 references into transferable data needs to be allowed before the data
-is transferred. Allowing such references is crucial to performance,
+is transferred. Allowing such references is crucial, % to performance,
 as any non-trivial program creates temporary references to the internal
@@
 Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
-that combine linear types and regions to suport dynamic regions.
+that combine linear types and regions to support dynamic regions.
 Unlike the prior work, we ensure memory safety in the presence of
@@
 
-\subsection*{Contributions}
-
+\paragraph{Contributions.}
 The paper makes the following contributions:
@@
 \begin{itemize} 
-  \item We present \name, a \csharp-like typed
+  \item We present \broom, a \csharp-like typed
   object-oriented language that eschews garbage collection in favour of
-  programmer-managed memory regions . \name extends its core language,
+  programmer-managed memory regions. \broom{} extends its core language,
   which includes \emph{lambdas} (higher-order functions) and
   \emph{generics} (parametric polymorphism), with constructs to create,
-  manage and destroy static and transferable memory regions. Transferable regions
-  are first-class values in \name.
+  manage and free static and transferable memory regions. Transferable regions
+  are first-class values in \broom.
   % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
@@
 
-  \item \name is equipped with a region type system that statically
+\item \broom{} is equipped with a region type system that statically
   guarantees safety of all memory accesses in a well-typed program,
@@
   later invariants are enforced via simple runtime checks.
-% The correctness of region usage is checked efficiently at run-time.
-% The overhead of checking these conditions is less than \GK{x\%} in
-% our experiments with \naiad workloads.
 
-  \item We define an operational semantics for \name, and a type
+  \item We define an operational semantics for \broom, and a type
   safety result that clearly defines and proves safety guarantees
@@
 
-  \item We describe a region type inference algorithm for \name that
-  (a). completely eliminates the need to annotate \name programs with
+  \item We describe a region type inference algorithm for \broom that
+  (a). completely eliminates the need to annotate \broom programs with
   region types, and (b). enables seamless interoperability between
-  region-aware \name programs and legacy standard library code that is
+  region-aware \broom programs and legacy standard library code that is
   region-oblivious. The cornerstone of our inference algorithm is a
@@
 
-  \item We describe an implementation of \name frontend in OCaml,
+\item We establish partial completeness results for the type inference
+  algorithm, as well as soundness.
+
+  \item We describe an implementation of \broom frontend in OCaml,
   along with case studies where the region type system was able to
@@
   %The rest of the paper is organized as follows. The next section
-  %presents an informal overview of \name, and motivates the need for a
+  %presents an informal overview of \broom, and motivates the need for a
   %region type system.  \S~\ref{sec:type-system} formalizes the type

### 2016-10-18 paper/related.tex
minor changes

@@
 Similar contrast can be made of region type inference in both the languages.
-Cyclone’s type inference was only ever described as being similar to
-Tofte and Talpin’s, and it’s effectiveness in presence of tracked
+Cyclone's type inference was only ever described as being similar to
+Tofte and Talpin's, and its effectiveness in presence of tracked
 pointers is not clear.
@@
 
+Our region type system can also be thought of as a specialized
+ownership type system~\cite{OwnershipSurvey}, where each region is the
+owner of all objects allocated in the region.
+%
 An ownership type system for safe region-based memory management in
@@
 % that language.  
-Their source language admits various kinds of regions with
-lexically-scoped lifetimes (i.e., stack regions) in order to support
-shared-memory concurrency. In contrast, we admit regions with
-dynamically determined lifetimes in order to support message-passing
-concurrency. We borrow outlives relation from their formal
+Their language permits only lexically-scoped (stack) regions
+(in the context of shared-memory concurrency). In contrast, we permit
+regions with dynamically determined lifetimes (in the context of message-passing concurrency).
+We borrow outlives relation from their formal
 development, and our type system bears some similarities to theirs.
@@
 and higher-order functions.
-%, whose interaction is non-trivial in context of a region type system. 
-% On the other hand, our type system eschews the notions of region kinds
-% and ownership, leading to a more succinct formalization. 
 Moreover, we establish type safety and transfer safety results that
@@
 % eliminates any such need.
+% Our type inference algorithm is also novel compared to existing type
+% inference approaches for ownership types, which are based on, e.g.,
+% context-insensitive pointer analysis or boolean satisfiability.
+% E.g., papers 55 and 79.
 

### 2016-10-18 paper/overview2.tex
Final changes.
minor changes

@@
 \begin{codejava} 
-  Region<A> rgn = new Region<T>(() => new A())
+  Region<A> rgn = new Region<A>(() => new A())
 \end{codejava} 
@@
 store the output for time \C{t}, initializing it to contain an empty
-list.  Line 9 opens the input region to process it\footnote{We omit
-\C{@R} annotation in \C{open} when we don't need \C{R}.}.  Line 10
-creates a stack region \C{R0}.  Thus, the temporary objects created by
-the iteration in line 11, for example, will be allocated in this stack
+list.  Line 8 opens the input region to process it.
+% \footnote{We omit \C{@R} annotation in \C{open} when we don't need \C{R}.}
+Line 9 creates a stack region \C{R0}.  Thus, the temporary objects created by
+the iteration in line 10, for example, will be allocated in this stack
 region that lives just long enough.  We open the desired output region
-in line 12, so that the new output objects created by the invocation
-of \C{selector} in line 13 are allocated in the output region.
-Finally, the input region is freed in line 15. The output region at
+in line 11, so that the new output objects created by the invocation
+of \C{selector} in line 12 are allocated in the output region.
+Finally, the input region is freed in line 14. The output region at
 \C{map[t]} stays as along as input messages with timestamp \C{t} keep

### 2016-10-19 paper/type-inference.tex
type inference proofs

@@
  \item If the solver succeeds, it returns a solution $\soln$ consisting of a
-   pair of substitution functions $\regionSubstFn$ and $\predSubstFn$ for
+   pair of substitution functions $\solnR$ and $\solnP$ for
    free region and predicate variables, respectively, introduced in step 1.
@@
 \paragraph{Constraint Solution.}
-We define an \emph{assignment} $\soln$ to be a pair of functions $(\regionSubstFn,\predSubstFn)$,
-where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula.
+We define an \emph{assignment} $\soln$ to be a pair of functions $(\solnR,\solnP)$,
+where $\solnR$ is a map from $\regionVars$ to $\regionConstants$
+and $\solnP$ is a map from $\predVars$ to a region-constraint formula.
 Such an assignment is said to \emph{satisfy} a set of constraints $C$ if
-every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.
+every sequent in $C$ is valid after the substitutions $\solnP$ and $ \solnR$.
 We say that $C$ is \emph{satisfiable} if it has a satisfying assignment.
@@
 %\end{center}
-%The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
-%where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
-%and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
+%The solution to the constraint satisfaction problem is a pair $(\solnR,\solnP)$,
+%where $\solnR$ is a map from $\regionVars$ to $\regionConstants$
+%and $\solnP$ is a map from $\predVars$ to a region-constraint formula such that
 %\begin{itemize}
-%  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
+%  \item $\solnR(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
 %
-%  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
-%    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).
+%  \item $\solnP$ is well-formed under $\predDeltaMap$
+%    (i.e., $\tywf{\predDeltaMap(\varphi)}{\solnP(\varphi)}$, for every $\varphi \in \predVars$).
 %
-%  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
-%    \regionSubstFn$.
-%  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
+%  \item Every sequent in $C$ is valid after substitutions $\solnP$ and $
+%    \solnR$.
+%  %% $C\lbrack (\solnR,\solnP) \rbrack$ is valid.
 %\end{itemize}

### 2016-10-19 paper/typeinference-proofs.tex
type inference proofs

@@
+\newcommand{\deltaPC}{\Delta_P^C}
+
 \section{Type Inference: Proofs}
@@
 
+We recall the definition of $\predDeltaMap(\varphi)$ first. We work with sets of constraints $C$ that
+contain exactly only constraint of the form  $\tywf{\Delta}{\varphi}$ for any predicate variable $\varphi$
+and we refer to this $\Delta$ as $\predDeltaMap(\varphi)$. In the sequel, we use the notation
+$\deltaPC(\varphi)$ to clarify the dependence on $C$, and will omit the superscript if no confusion
+is likely.
+
 \begin{lemma}
@@
 where both $\pi_i$ and $\pi_j$ are region constants.
-(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
+(a) If $\{ \pi_i, \pi_j \} \subseteq \deltaPC(\varphi)$,
 then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
@@
 $\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \deltaPC(\varphi)$,
 $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
@@
 \begin{proof}
-  Straightforward.
+  Let $\soln$ be any assignment that satisfies $C$.
+  Let $\phi'$ denote $\solnP(\varphi)$.
+  The lemma follows once we show that
+  $\isvalid{\phi' \conj \phi}{\pi_i \outlives \pi_j}$ iff
+  $\isvalid{\phi'}{\pi_i \outlives \pi_j}$ or
+  $\isvalid{\phi}{\pi_i \outlives \pi_j}$.
+  The reverse implication is trivial, and we establish the forward implication below.
+
+  Assume that $\isvalid{\phi' \conj \phi}{\pi_i \outlives \pi_j}$.
+  Recall that we can represent $\phi' \conj \phi$ as a graph, with each vertex
+  representing a region identifier, and each outlives-constraint $\pi \outlives \pi'$
+  represented as an edge from $\pi$ to $\pi'$. Further,
+  $\isvalid{\phi' \conj \phi}{\pi_i \outlives \pi_j}$ holds iff there exists a path
+  from $\pi_i$ to $\pi_j$ in this graph.
+
+  Let $\Delta$ denote $\deltaPC(\varphi)$.
+  Clearly, $\phi'$ can include an outlives-constraint $\pi \outlives \pi'$ only if
+  $\{ \pi, \pi' \} \subseteq \Delta$.
+  Thus, $\phi'$ represents a set of edges between vertices in $\Delta$.
+  On the other hand, $\phi$ represents a set of edges whose targets are outside $\Delta$
+  (from the definition of a decomposable constraint).
+  Furthermore, it follows from the definition of a decomposable constraint that
+  if $\phi$ includes an outlives-constraint $\pi \outlives \pi'$ for some $\pi \in \Delta$,
+  then $\phi$ includes the outlives-constraint $\pi'' \outlives \pi'$ for every $\pi'' \in \Delta$.
+
+  The result follows.
 \end{proof}
@@
 \begin{proof}
-The constraint-generation rules guarantee that every region variable will
-be unified with some region constant\footnote{
+Intuitively, the lemma says that every region variable will be unified with
+some region constant (by the generated constraints).
 There is one special case where this does not hold: a recursive function
@@
 We assume that the return value of such a function is typed to be \C{unit},
-which will not introduce any region variables at the call-site of such a function.}
-This follows because $\FB$ does not admit uninitialized variables.
+which will not introduce any region variables at the call-site of such a function.
+
+This property follows from several aspects of $\FB$, e.g., it does not admit uninitialized variables.
+Note that the region \C{R} where a new object is allocated (by the \C{new} construct) is indicated
+by the user (as \C{new@R}). If the user does not specify this, it defaults to the allocation
+context region. Thus, this aspect is handled by the frontend, and does not require or use
+the constraint solver.
+
+% This property follows from the constraint-generation rules.
 % , and the type system uniquely determines the region parameters of every value
@@
 Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
-where $\pi_i$ and $\pi_j$ are distinct region constants. The definition of $\predSubstFn(\varphi)$
+where $\pi_i$ and $\pi_j$ are distinct region constants. The definition of $\solnP(\varphi)$
 ensures that $\soln$ satisfies this constraint.
@@
 Consider any constraint of the form $\isvalid{\ell}{F_j(\varphi_j)}$.
-Substituting the value $\predSubstFn(\varphi_j)$ reduces this constraint to one of
+Substituting the value $\solnP(\varphi_j)$ reduces this constraint to one of
 the form $\isvalid{\ell}{\conj_i \pi_i \outlives \pi_i'}$ where the substitution rule
@@
 must satisfy this constraint ($\isvalid{\ell}{r'}$).
-Let $\pi' = \regionSubstFn(\rho)$.
+Let $\pi' = \solnR(\rho)$.
 We can show (using Lemma~\ref{lemma:two-bindings}) that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$.
@@
 % Let $\soln$ be a satisfying assignment for $\satC$.
-% Let $\pi' = \predSubstFn(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
+% Let $\pi' = \solnP(\rho)$. Then, we must have $\pi' \in \regionDeltaMap(\rho)$.
 % From Lemma~\ref{lemma:completely-bound}, $\rho$ must be bound to some $\pi$ in some context $\ell$.

### 2016-10-19 paper/misc.tex
moved discussion to appendix

--- 2016-10-19 paper/type-inference.tex
@@
-\newcommand{\soln}{\eta}
-\newcommand{\solnR}{\soln_R}
-\newcommand{\solnP}{\soln_P}
-\newcommand{\thesoln}{\hat{\eta}}
-\newcommand{\thesolnR}{\thesoln_R}
-\newcommand{\thesolnP}{\thesoln_P}
+\section{Other Aspects}
 
-\newcommand{\saturate}[1]{{#1}^*}
-\newcommand{\myground}[1]{\saturate{#1}_g}
-\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
-\newcommand{\satC}{\saturate{C}}
-\newcommand{\groundC}{\myground{C}}
-\newcommand{\rhoC}{\text{WF}_R}
-\newcommand{\solveCon}[1]{\textsc{Solve}(#1)}
-
-\section{Type Inference}
-\label{sec:type-inference}
-
-\name's region type system imposes a heavy annotation burden, and
-manually annotating C\# standard libraries with region types
-can be tedious. We now present our region type inference algorithm
-that eliminates the need to write region type annotations.
-% except on some higher-order functions.
-Formally, the type inference
-algorithm is an elaboration function from programs in $\absof{\FB}$
-(i.e., \FB without region types, but with \C{letregion} and \C{open}
-expressions, similar to the language introduced in
-\S~\ref{sec:overview}) to programs in \FB.
-For simplicity, we assume that each \C{letregion} and \C{open} construct
-in the input program introduces a distinct region identifier $\pi$.
-
-\paragraph{Overview.}
-
-% Fig.~\ref{fig:type-inference-algo}
-We now present a high-level outline of the type inference algorithm.
-The algorithm consists of the following steps:
-\begin{enumerate}
- \item \emph{Region Parameterization}.
-   The first step elaborates the input program by introducing \emph{formal region parameters}
-   (for each class and method), and \emph{region variables} (representing yet undetermined
-   \emph{actual region parameters}). We also introduce for each class and method, a
-   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
-   over the region parameters of that class/method.
-
- \item \emph{Constraint Generation}.
-   In the second step, we analyze the program to generate a set of constraints
-   (over the region identifiers and the predicate variables)
-   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).
-
- \item \emph{Constraint Solving}.
-   We solve the generated set of constraints using our fixpoint constraint
-   solving algorithm, which reduces the constraint solving problem to
-   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
-   references, for example, a reference from a transferable region to a
-   stack region, then the constraints generated during the elaboration
-   are not satisfiable. In such a case, the solver fails to solve
-   the constraints.
-
- \item If the solver succeeds, it returns a solution $\soln$ consisting of a
-   pair of substitution functions $\solnR$ and $\solnP$ for
-   free region and predicate variables, respectively, introduced in step 1.
-   We apply these substitutions to the elaborated program to produce the final program.
-\end{enumerate}
-
-% \begin{figure}
-% \begin{codeml}
-% Infer ($p$) =
-%   let $q$ = IntroduceRegionParameters($p$) in
-%   let $C$ = GenerateConstraints($q$) in
-%   match (SolveConstraints($C$) with
-%   | None $\longrightarrow$ None
-%   | Some ($\soln$) $\longrightarrow$ Some ($q[\soln]$)
-% \end{codeml}
-% 
-% \caption{The type inference algorithm}
-% \label{fig:type-inference-algo}
-% \end{figure}
-
-We later on establish the soundness of the type inference algorithm, and also establish
-that the constraint-generation and constraint-solving steps (the second and third steps
-above) are complete.
-
-\subsection{Region Parameterization for Classes}
-\label{sec:fb-templatization}
-
-Region parameterization is an iterative process involving the following three steps,
-the first two of which are mutually dependent on each other.
-
-\emph{Introduction of Formal Region Parameters}.
-For every class \C{C}, we identify a sequence of formal region parameters
-$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.
-
-\emph{Introduction of Actual Region Parameters}.
-We then replace every instance of class \C{C} in the program by an instance
-$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
-are fresh identifiers denoting actual region parameters.
-
-\emph{Predicate Variable Introduction}. For every class \C{C}, we introduce
-a fresh predicate variable $\varphi$, which represents the yet undetermined
-outlives constraints between the formal region parameters of class \C{C}.
-
-We identify the region parameters of classes as follows.
-
-\emph{Non-Recursive Classes}.
-The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
-The region parameters for any other non-recursive class \C{C} is determined
-only after the region parameters of any class that \C{C} depends on have been
-determined: this includes the base-class \C{B} of \C{C} and the class (type)
-of any of its fields.
-We replace every dependee type \C{T} in \C{C} by its instantiated type,
-using fresh region parameters as needed.
-The sequence of region parameters for \C{C} is defined to be
-the sequence of region parameters for the base class \C{B} concatenated
-with the list of all  fresh region parameters introduced while instantiating the types
-of the fields in the class.
-(The class inherits its allocation region from its base class. Note that if
-a class does not specify an explicit base class, it has an implicit base class
-\C{Object}.)
-
-This transformation is illustrated below, using a non-generic \C{Pair} class:
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.35\linewidth}
-\begin{codejava}
-class Pair $\extends$ $\ObjZ$ {
-  $\ObjZ$ fst;
-  $\ObjZ$ snd;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.6\linewidth}
-\begin{codejava}
-class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
-  $\ObjZ \langle \rho_1 \rangle$ fst;
-  $\ObjZ \langle \rho_2 \rangle$ snd;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-\emph{Recursive Classes}.
-The region parameters for a recursive class is computed in
-a similar fashion, with the following difference: any recursive
-field is ignored while instantiating region parameters for the fields of
-the class, and the region parameters of the recursive class are computed
-as before. We then do parameter instantiation for all recursive fields,
-such that their region annotations (the actual region parameters) are
-exactly the same as the (formal) region parameters of the class.
-The following example illustrates this for a non-generic \C{List} class.
-The resulting class represents a linked list with spine in the region
-$\rho_0$ and data objects in the region $\rho_1$.
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.35\linewidth}
-\begin{codejava}
-class List $\extends$ $\ObjZ$ {
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.65\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-The above technique can be extended to mutually recursive classes in a
-straightforward manner, by simultaneously parameterizing them (and
-then instantiating them).
-
-\emph{Type-Parametric Classes}.
-% Type parameters of classes are handled as follows.
-% Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
-The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
-(This can be extended to use the bound specified for \C{T}, if any. We do not
-discuss this due to space restrictions.)
-% number of region parameters
-% that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
-% to be \C{Object}, and \C{T} is instantiated with one region parameter.
-
-\emph{Function Types}.
-Since \FB{} is higher-order, fields of function type are allowed.We explain how the parameter instantiation step
-instantiates function types below, after discussing parameterization for methods.
-
-%% \footnote{We abuse arrow notation to also represent
-%% types of methods, but unlike function types, there is no allocation
-%% region annotation atop the arrow in a method type.}
-%% $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
-%% \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
-%% \unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
-%% variables, and $\varphi_0$ is a fresh predicate variable denoting
-%% unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
-
-\subsection{Region Parameterization for Methods and Function Types}
-
-As the next step, we introduce region parameters for every method.
-We do this by instantiating the types of all parameters and the
-return value (of the method) using fresh region identifiers (as explained previously),
-and then generalizing these region identifiers as formal region parameters
-of the method. In addition, a fresh region identifier is introduced to represent the
-allocation region.
-We also introduce a fresh predicate variable $\varphi$ for every method,
-just as we did for each class.
-Thus, the method
-
-\C{Object m (List x) \{...\} } \\
-is instantiated as
-
-\C{Object$\inang{\rho_3}$ m$\inang{\rho_0,\rho_1,\rho_2,\rho_3}$ (List$\inang{\rho_1,\rho_2}$ x) \{...\} }
-
-We then consider every method invocation in the program, and introduce
-fresh region variables representing the (yet unknown) actual region
-parameters for this particular invocation.
-%
-We similarly perform instantiation for every constructor invocation
-of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
-before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
-where $\rho_0, \cdots, \rho_n$ are fresh region variables.
-
-Function types (of fields and parameters) are instantiated just like methods above, except that
-an additional fresh region identifier is introduced to represent the region where the closure is
-allocated.
-For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2, \rho_3, \rho_4 \,|\, \varphi } \C{List} \inang{\rho_1,\rho_2} \xrightarrow{\rho_4} \C{Object}\inang{\rho_3}$.
-Note that the newly introduced region identifiers are generalized as formal region parameters of the function type.
-This is a (heuristic) choice made in the case of higher order functions.
-Consider a higher order function \emph{f} with a function typed parameter $g$.
-The fresh region identifiers introduced while instantiating the type of $g$ could be
-alternatively generalized as formal region parameters of $f$, but we choose to
-generalize them as formal region parameters of $g$.
-We will discuss this aspect again later.
-
-\subsection{Constraint Generation}
-\label{sec:fb-constraintsem}
-
-The constraint generation algorithm mimics the static type checker, but accumulates
-constraints that must hold for the type checking to succeed.
-
-\paragraph{Syntax of Constraints.}
-The constraints are expressed using
-a set $\regionConstants$ of region constants,
-a set $\regionVars$ of region variables,
-and
-a set $\predVars$ of predicate variables.
-% 
-% ($\pi$ and $\rho$) and predicate variables ($\varphi$).
-% $(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-
-\input{constraint-syntax}
-
-% \emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
-Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
-of a method invocation or object allocation, which will be bound to a region constant by the end of the
-type inference.
-A region constant may be 
-(a) a \emph{formal region parameter} of a class or method, or
-(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
-(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
-
-Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
-over a set of fixed formal region parameters.
-%% where the set of region-constraints $\phi$ is defined by:
-%% \begin{smathpar}
-%% \begin{array}{lcl}
-%% \phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
-%% \end{array}
-%% \end{smathpar}
-%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
-%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
-%% as shorthand for $\conj_{\phi \in S} \phi$.
-%% 
-We will use the term \emph{validity constraint} to denote an entailment constraint
-of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
-
-Our constraints also make uses of \emph{pending substitutions} $F$:
-A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
-used in a particular context:
-%% \begin{smathpar}
-%% \begin{array}{lcl}
-%% F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
-%% \end{array}
-%% \end{smathpar}
-E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
-\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
-is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
-$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
-and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
-deducible from $\rgn_1 \outlives \rgn_2$.
-
-The constraints generated are of the following kinds:
-\begin{itemize}
-
-\item Well-formedness constraints of form $\rho \in \rhoenv$,
-restricting the domain of unification for a region variable ($\rho$)
-to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
-
-\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
-variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
-regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
-  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
-  the region-constraint of its base class \C{B}).
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
-  (e.g., to indicate the region-constraint of a class \C{C} must imply the
-  region-constraint of each of its fields).
-
-\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
-where $\varphi_i$ is a predicate variable (representing the precondition of a
-method to be determined), $\phicstr$ is a region constraint that is \emph{required}
-to hold at a particular program point (within the method), and $\phictxt$ is
-a region constraint that is \emph{known} to hold at that program point.
-
-%
-%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-%% of predicate variables and pending substitutions. While $\phictxt$
-%% captures relationships that are \emph{known} to hold between concrete
-%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
-%% $\phicstr$ captures relationships that are \emph{required} to hold
-%% among region varibles (i.e., $\rho$'s), or relationships between
-%% region variables and identifiers.
-%
-
-\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
-generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
-$\varphi_i$ are as above).
-
-\end{itemize}
-
-\paragraph{Constraint Solution.}
-We define an \emph{assignment} $\soln$ to be a pair of functions $(\solnR,\solnP)$,
-where $\solnR$ is a map from $\regionVars$ to $\regionConstants$
-and $\solnP$ is a map from $\predVars$ to a region-constraint formula.
-Such an assignment is said to \emph{satisfy} a set of constraints $C$ if
-every sequent in $C$ is valid after the substitutions $\solnP$ and $ \solnR$.
-We say that $C$ is \emph{satisfiable} if it has a satisfying assignment.
-
-%\begin{definition}
-%\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
-%A constraint satisfaction problem is a tuple
-%$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-%where $\regionConstants$ is a set of region constants,
-%$\regionVars$ is a set of region variables,
-%$\predVars$ is a set of predicate variables,
-%$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
-%specifies a unification domain for each region variable,
-%$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
-%specifies the domain for each predicate variable,
-%and $\constraintSet$ is a set of validity constraints in one of the following forms:
-%\begin{center}
-%\(
-%    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
-%    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
-%\)
-%\end{center}
-%The solution to the constraint satisfaction problem is a pair $(\solnR,\solnP)$,
-%where $\solnR$ is a map from $\regionVars$ to $\regionConstants$
-%and $\solnP$ is a map from $\predVars$ to a region-constraint formula such that
-%\begin{itemize}
-%  \item $\solnR(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
-%
-%  \item $\solnP$ is well-formed under $\predDeltaMap$
-%    (i.e., $\tywf{\predDeltaMap(\varphi)}{\solnP(\varphi)}$, for every $\varphi \in \predVars$).
-%
-%  \item Every sequent in $C$ is valid after substitutions $\solnP$ and $
-%    \solnR$.
-%  %% $C\lbrack (\solnR,\solnP) \rbrack$ is valid.
-%\end{itemize}
-%\end{definition}
-
-\input{fb-constraint-generation}
-
-\paragraph{Constraint Generation.}
-The constraint generation algorithm is a direct adaption of the type checker:
-each type checking judgment is modified to produce a set of constraints that must hold
-for the type checker to succeed.
-
-Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
-These rules use the same context as the corresponding typing judgment in
-Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
-use of region variables and predicate variables.
-Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
-The algorithm proceeds top-down, analyzes an expression $e$ in a context
-$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
-(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
-indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.
-
-The rules for generating constraints from a method and
-class definition first build a context ($\A$) containing a set ($ \rhoenv$) denoting
-regions that are currently live, a map ($\aenv$) mapping type
-variables to their bounds, and a constraint formula ($\phicx$)
-capturing constraints over live region variables. We use predicate
-variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
-
-\TODO{Change the static semantics to be consistent with this, and
-make this a comment there.}
-One notable aspect here is the following. Consider a method with predicate
-variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
-We use $\varphi_m$ as the constraint formula in the initial context used for
-the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
-It is possible to show that there is no loss of generality in doing so.
-Specifically, it can be shown for any well-typed program, replacing the constraint
-$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
-where $\varphi_c$ is the constraint of the class containing the method,
-produces a well-typed program.
-
-We use the above simplification to avoid generating validity constraints of the form
-$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
-with more than one predicate variable on the left hand side, which introduce complications).
-%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.
-
-\begin{theorem}
-\label{thm:constraint-generation-sc}
-Let $C = \consOf{q}$.
-An assignment $\soln$ (for the region and predicate variables in $q$)
-satisfies $C$ iff $q[\soln]$ is well-typed.
-\end{theorem}
-
-\subsection{The Constraint Solver}
-
-We refer to any validity constraint whose antecedent contains a predicate variable
-(\ie, is of the form $\varphi \conj \phictxt$) as an \emph{abduction constraint}.
-Here,  $\phictxt$ is either \emph{true} (in which case, we call the constraint a trivial
-abduction constraint) or a conjunction of one or more outlives-constraints (in
-which case, we call the constraint a non-trivial abduction constraint).
-
-Non-trivial abduction constraints are a key challenge in solving the constraints.
-We now describe some special properties of the set of constraints $C$
-generated by our algorithm, which allow us to handle abduction constraints efficiently.
-
-For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
-of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
-$\predDeltaMap(\varphi)$.
-%
-%% The formalism below is motivated by the following intuition.
-%% Suppose some construct in a method's body requires an outlives-constraint
-%% $\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
-%% of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
-%% as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
-%% $\pi_1 \outlives \pi_2$ must be established by the context within the method
-%% (\eg, by the \code{letregion} constructs in the method).
-%% We now generalize this idea to the setting of our abstract constraint-satisfication
-%% problem.
-%
-We refer to an abduction constraint in $C$ as \emph{decomposable}
-if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
-$\phictxt$ is a conjunction of zero or more outlives-constraints of the form
-$\pi_1 \outlives \pi_2$ satisfying the following conditions:
-(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
-(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
-every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
-a conjunct in $\phictxt$.
-%
-(The above concepts parametrically depend on the set $C$, though we omit $C$
-from the notation for simplicity.)
-
-\begin{lemma}
-  \label{lemma:gc-is-decomposable}
-  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
-\end{lemma}
-
-%The second property is that the \emph{context region constraint} $\phictxt$
-%occurring on the antecedent of any validity constraint is a conjunction of
-%outlives-constraints of the form $\pi \outlives \pi_s$ where $\pi_s$
-%is a static region identifier and $\pi$ is a region constant (either a formal
-%region parameter, or static region identifier, or open transferable region identifier).
-%Furthermore, if $\phictxt$ includes any conjunct $\pi \outlives \pi_s$, then
-%it includes every conjunct $\pi_f \outlives \pi_s$ for every $\pi_f \in \predDeltaMap(\varphi)$.
-%This property allows us to show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
-%iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$,
-%simplifying the constraint solver.
-
-\begin{lemma}
-  \label{lemma:decomposition}
-  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
-where both $\pi_i$ and $\pi_j$ are region constants.
-(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
-\end{lemma}
-
-The above lemma shows how we can reduce a non-trivial abduction constraint to
-either a trivial abduction constraint or a non-abduction constraint.
-
-\paragraph{Constraint Solver.}
-The first step in our algorithm for solving a set of constraints $C$
-computes a set $\satC \supseteq C$ of constraints by iteratively applying
-the following rules until a fixed point is reached:
-
-\begin{enumerate}
-
-\item (Initialization) $\isvalid{\ell}{r} \in C \Rightarrow \isvalid{\ell}{r} \in \saturate{C}$
-
-\item (Transitivity)
-  \label{item:transitivity}
-\begin{enumerate}
-\item
-$\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
-$\isvalid{\ell \conj \phi}{\rho_2 \outlives \rho_3} \in \saturate{C}$
-$\Rightarrow$
-$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
-\item
-$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
-$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$
-$\Rightarrow$
-$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
-\end{enumerate}
-
-\item (Substitution)
-  \label{item:subst}
-$\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
-$\isvalid{\varphi}{\phi} \in \saturate{C}$
-$\Rightarrow$ $\isvalid{\ell}{F(\phi)} \in \saturate{C}$
-
-\item (Abduction Decomposition)
-\label{item:context}
-\begin{enumerate}
-\item
-$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
-$\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$
-$\Rightarrow$
-$\isvalid{\varphi}{\pi_i \outlives \pi_j} \in \saturate{C} $
-
-\item 
-$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
-$\{ \pi_i, \pi_j \} \not\subseteq \predDeltaMap(\varphi)$,
-$\{ \pi_i, \pi_j \} \subseteq \regionConstants$
-$\Rightarrow$
-$\isvalid{\phictxt}{\pi_i \outlives \pi_j} \in \saturate{C} $
-\end{enumerate}
-
-\end{enumerate}
-
-% Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
-We can show that every new constraint added by the above rules is implied by existing
-constraints:
-
-\begin{theorem}
-  \label{thm:closure}
-Let $C = \consOf{q}$.
-An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
-\end{theorem}
-
-A key goal of this step is to identify the value every region variable must
-have in any solution of the set of constraints, as explained below.
-%
-% unifying region variables with the
-% region constants that they should be bound to in the solution, as explained below.
-%
-Consider a set of constraints $D$. We say that a region variable $\rho$
-\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
-constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
-We say that a region variable $\rho$ is \emph{bound} to a region constant
-$\pi$ in a context $\ell$ if
-$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
-We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
-$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.
-
-\begin{lemma}
-  \label{lemma:completely-bound}
-Let $C = \consOf{q}$.
-Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
-\end{lemma}
-
-The set $\satC$ makes it easy to identify a solution $\thesoln$ (if one exists), as below.
-\begin{itemize}
-\item For any predicate variable $\varphi$, $\thesolnP(\varphi)$ is defined
-to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
-set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$.
-\end{itemize}
-
-% \paragraph{Step 2.}
-
-% \begin{enumerate}
-% \item We use an iterative fixed point (saturation) algorithm to compute
-% $\saturate{C}$, the set of all constraints logically implied by $C$.
-% \item
-
-The set $\satC$ also makes it easy to check if $C$ is satisfiable.
-Let $\myground{C}$ denote the subset of all ground constraints
-(\ie, constraints without any region variable or predicate variable)
-in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
-constraints for region variables in $\satC$.
-Define $\solveCon{C}$ as below.
-% It is possible to show that $\soln$ satisfies $\satC4
-% iff  $\myground{C}$ is valid in the theory of partial-order.
-
-% \end{enumerate}
-
-\[
-\solveCon{C} = 
-\begin{cases}
-\textsc{Some} (\thesoln) & \textrm{if $\groundC$ is valid and $\thesoln$ satisfies $\rhoC$} \\
-\textsc{None} & \textrm{otherwise}
-\end{cases}
-\]
-
-\begin{theorem}
-\label{thm:constraint-solver-sc}
-Let $C = \consOf{q}$.
-(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.
-(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
-\end{theorem}
-
-Checking the validity of a ground constraint is straightforward.
-A ground constraint is of the form $\isvalid{\conj_{i \in I} \ell_i}{r}$, where each
-$\ell_i$ and $r$ is an outlives-constraint.
-Let $D$ denote $\{ \isvalid{}{\ell_i} \;|\; i \in I \}$.
-The given constraint is valid iff $r$ belongs to $\saturate{D}$.
-
-\paragraph{Algorithmic Aspects.}
-Note that checking the validity of a ground constraint can be realized using
-a simple graph reachability algorithm.
-%
-Given a set $S$ of outlives constraints, define the directed
-graph $G(S)=(V(S),E(S))$ as follows.
-Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
-which we will also refer to as $\rho$.
-Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
-an edge from $\rho_1$ to $\rho_2$.
-%
-%% (Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
-%% for the pair of constraints $\rho_1 \outlives \rho_2$ and
-%% $\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
-%% by a pair of edges.)
-%
-It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$ iff
-there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
-%
-Thus, a simple graph reachability algorithm can be used to check the validity of
-ground constraints.
-% all the logical consequences of a set of outlives constraints.
-
-This idea generalizes.
-% Consider a set $C$ of constraints that contains no non-trivial abduction constraint.
-% Then, the algorithm for computing the closure $\satC$ does not need to use
-% rule~\ref{item:context} (for Abduction Decomposition).
-Extending the simple graph reachability algorithm to incorporate the Substitution rule
-(in computing $\satC$) turns the problem into a context-free reachability problem in graphs~\cite{Reps:Reachability}
-(as usual for context-sensitive interprocedural analysis).
-
-Algorithms for context-free reachability can be adapted to incorporate the Abduction Decomposition step.
-Alternatively, the iterative process described above is a standard fixed point computation
-and can be encoded using a set of Datalog rules, allowing us to compute the closure
-using any Datalog engine.
-% Further details omitted.
-
-\paragraph{Modularity Aspects.}
+\paragraph{Modularity Aspects of Type Inference.}
 The type inference algorithm, as presented, traverses the entire program to
@@
 However, methods that are not mutually recursive can be processed separately.
-
-\subsection{Soundness and Completeness: Discussion}
-
-We now state the soundness theorem, which follows from
-Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
-as well as the following observation about the elaboration phase.
-For any program $p \in \absof{\FB}$, the elaboration phase produces a $q$
-such that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
-
-\begin{theorem}
-\emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if type inference succeeds, it returns a $t$ such that
-% Infer($p$) returns Some($t$), then
-(1) $\absof{t} = p$, and
-(2) $t$ is well-typed.
-\end{theorem}
-
-% \item $\absof{q[\sigma]}$ = $p$ for any substitution $\sigma$.
-
-However, in this context, completeness is more interesting than soundness.
-(Note that an unsound inference algorithm can be made sound by simply type-checking
-the resulting program.)
-Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
-show that the second and third steps of the type-inference algorithm are complete.
-%% We believe that the converse of the above lemma can also be established:
-%% any substitution for the region variables and the predicate variables that makes the
-%% expression well-typed will also satisfy the generated constraints.
-%% Thus, if there is any possible instantiation (of the region variables and predicate variables) of
-%% the elaborated program produced by the first phase that type checks, then the generated
-%% constraints are satisfiable, and the second phase will identify it.
-%% 
-The only sources of incompleteness in the type inference algorithm are some heuristic
-choices made in the first step (the elaboration phase), as explained below.
-
-\begin{enumerate}
-\item We determine the set of region parameters for a recursive class
-using the heuristic that a recursive occurrence of the class has
-the same parameters, in the same order, as the class itself.
-This heuristic fails, for example, if the program uses a recursive list type whose
-elements alternatively come from two different regions. Such a program would require
-the following elaboration, which is beyond the scope of our approach:
-
-\begin{tabular}{ccc}
-\begin{minipage}{0.35\linewidth}
-\begin{codejava}
-class List $\extends$ $\ObjZ$
-{
-  $\ObjZ$ data;
-  List next;
-}
-\end{codejava}
-\end{minipage}
-&
-$\Rightarrow$
-&
-\begin{minipage}{0.65\linewidth}
-\begin{codejava}
-class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
-{
-  $\ObjZ \langle \rho_1 \rangle$ data;
-  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
-}
-\end{codejava}
-\end{minipage}
-\end{tabular}
-
-\item Our technique for region parameterization also uses a heuristic in the case of
-higher order programs. The following examples illustrates that principal types may
-not exist for higher order functions.
-\begin{codejava}
-    unit apply ( T $\rightarrow$ unit f, T x, T y) { f(x); f(y); }
-\end{codejava}
-This method may be typed assuming either that \C{f} is polymorphic over the region
-that its parameter is allocated in (permitting \C{x} and \C{y} to be allocated in any
-regions), or by assuming that \C{x} and \C{y} are allocated in the same region $\rho_1$
-that \C{f} expects its parameters to be allocated in. Neither type subsumes the other.
-Our algorithm heuristically chooses the first option, as it appears to be the more likely
-and useful candidate.
-
-%as either\\
-%$
-%\inang{ \rho_0 \rho_1 \rho_2}
-%    (\inang{\rho_0 \rho_3} (T \inang{\rho_3} \xrightarrow{} unit, 
-%      T \inang{\rho_1}, 
-%      T \inang{\rho_2})
-%   \xrightarrow{} 
-%      unit
-%$\\
-%or as\\
-%$
-%\inang{ \rho_0 \rho_1}
-%    (\inang{\rho_0} (T \inang{\rho_1} \xrightarrow{} unit, 
-%      T \inang{\rho_1}, 
-%      T \inang{\rho_1})
-%   \xrightarrow{} 
-%      unit
-%$\\
-
-%Consider a higher-order method that has a function $f$ as a
-%parameter. We have a choice in terms of where the region-parameters introduced for
-%$f$ are bound (i.e., quantified). We always bind these region-parameters within the type
-%of $f$ (essentially, requiring $f$ to be region-polymorphic). This may also be a potential
-%source of incompleteness in the type-inference.
-
-% The treatment of class fields with a function type also uses the same (incomplete) heuristic.
-
-\end{enumerate}
-
-If users provide partial region annotations, especially in situations (such as above) where elaboration
-makes a heuristic choice, the elaboration procedure can use the user-provided choices instead.
-This can help the type-inference overcome these sources of incompleteness.
-
-%\textbf{TODO: Update example discussion}
-%
-%\subsection{Example}
-%
-%\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
-%template from \S\ref{sec:fb-templatization}. Following constraints are
-%generated during its elaboration (Constrains are identified with
-%$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
-%\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
-%\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
-%\{\rho_5\}$, have been elided): 
-%\begin{smathpar}
-%\begin{array}{l}
-%  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
-%  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
-%     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
-%  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
-%  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
-%  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
-%  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
-%\end{array}
-%\end{smathpar}
-%
-%\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
-%contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
-%\C{Pair} object \C{q}, and an object \C{y}. It assigns
-%\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
-%recursively with the same region, a new \C{Pair} object allocated in a
-%local region, and an object referred by the \C{snd} field of the
-%pair inside the region. \C{alt} never terminates.  Elaboration phase
-%elaborates the method to the following region-annotated
-%definition\footnote{In reality, elaboration uses new region variables
-%as parameters to the method calls, and then generates
-%constraints that unify them with actuals. In our examples, to avoid
-%clutter due to trivial constraints,we coalesced both steps and show
-%the actuals instead.}(The original definition of \C{alt} can be
-%obtained by erasing all the region annotations from the elaborated
-%version):
-%% \begin{codejava}
-%% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
-%%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
-%%   p.fst = x; p.snd = y; 
-%%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
-%% }
-%% \end{codejava}
-%\begin{codejava}
-%unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair><$\toprgn$> r, 
-%                Pair<$\rho_{6-8}$> q, Object<$\rho_{9}$> y) {
-%  q.fst := y; q.snd := y; 
-%  open r as p@$\rgn_0$ in
-%    letregion $\rgn_1$ in
-%      let x = new Pair<$\rgn_1$,$\rgn_0$,$\rgn_0$>
-%                      (p.fst,p.fst) in
-%        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
-%}
-%\end{codejava}
-%% Note that, to avoid clutter, we have already resolved appropriate
-%% region arguments to the recursive call, instead of introducing new
-%% region variables and generating equality constraints on them. 
-%Constraints generated during the elaboration are shown below
-%(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
-%\begin{smathpar}
-%\begin{array}{l}
-%\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
-%\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
-%    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
-%\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
-%\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
-%\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
-%\rgn_1} 
-%    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
-%\end{array}
-%\end{smathpar}
-%
-% \paragraph{Constraints Example 1} The original source of all
-% outlives constraints is the validity constraint $C_3$ in $\elabClass$
-% function (Fig.~\ref{fig:fb-elabmeth}). In context of the $\C{Pair}$
-% class template from \S\ref{sec:fb-templatization}, the constraint is
-% as following:
-% \begin{center}
-%   \(\isvalid{\varphi_0}{\rho_0 \succeq \rhoalloc_0 \conj \rho_0
-%   \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0}\)
-% \end{center}
-
-% \paragraph{Constraints Example 2} Consider a contrived method
-% $\C{alt}$ that accepts an argument $\C{p}$ of type
-% $\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
-% of type $A$. It then assigns one object to $\C{p.fst}$ and other to
-% $\C{p.snd}$, and calls itself recursively without ever terminating.
-% Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
-% alternated between recursive calls. The region-annotated definition of
-% $\C{alt}$ is shown below:
-% \begin{codejava}
-% unit alt<$\rhoalloc$,$\rho_{0-4}$ | $\varphi_0$>(Pair<A,A><$\rho_0$,$\rho_1$,$\rho_2$> p,
-%                           A<$\rho_3$> x, A<$\rho_4$> y) {
-%   return p.fst = x; p.snd = y; 
-%          this.alt<$\rhoalloc$,$\rho_1$,$\rho_2$,$\rho_4$,$\rho_3$>(A,y,x)
-% }
-% \end{codejava}
-% Note that, to avoid clutter, we have already resolved appropriate
-% region arguments to the recursive call, instead of introducing new
-% region variables and generating equality constraints on them. Rest of
-% the validity constraints generated while elaborating $\C{alt}$ to the
-% above region-annotated definition are shown below:
-% \begin{center}
-% \(\isvalid{\varphi_0}{\rho_1 \outlives \rho_0 \conj \rho_2 \outlives \rho_0}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_1 = \rho_3}\)
-% $\quad$
-% \(\isvalid{\varphi_0}{\rho_2 = \rho_4}\)\\
-% \(\isvalid{\varphi_0}{[\rho_3/\rho_4][\rho_4/\rho_3]\varphi_0}\)
-% \end{center}
-% The first constraint is generated by $\typeOk$ on the type of $\C{p}$.
-% Second and third are generated by the the assignment expressions, and
-% the last constraint is generated from the recursive call.

### 2016-10-19 paper/type-inference.tex
moved discussion to appendix

@@
 
-\paragraph{Modularity Aspects.}
-The type inference algorithm, as presented, traverses the entire program to
-generate the set of constraints, which are solved en masse, using an iterative
-fixed point computation. However, the type inference can be realized in a
-modular and compositional fashion, subject only to the restrictions imposed
-by recursion.
-
-In the elaboration phase, we can process a class \C{C} only after any class
-\C{B} that \C{C} depends on has been processed: class \C{C} depends on
-class \C{B} if \C{B} is either \C{C}'s base class or the type of any field
-of \C{C} depends on \C{B}. In effect, this means that any collection of
-mutually recursive classes must be processed together. Non-recursive
-dependences can be handled in a compositional fashion: if class \C{C}
-depends on \C{B} non-recursively, then the elaboration can be done for
-\C{B} first, and then \C{C} can be processed.
-
-The same idea applies to the constraint-solving phase as well.
-Given a set of constraints, we say that a predicate variable $\varphi_1$
-\emph{directly-depends} on another predicate variable $\varphi_2$ if the set of
-constraints includes a constraint $\isvalid{\varphi_1 \conj \phictxt}{F(\varphi_2)}$.
-We say that $\varphi_1$ \emph{depends} on $\varphi_2$ if $\varphi_1$ transitively
-depends on $\varphi_2$.
-The constraint solver needs to process any collection of mutually dependent
-predicate variables together.
-In effect, this requires the type inference to process any collection of
-mutually recursive methods together.
-However, methods that are not mutually recursive can be processed separately.
-
 \subsection{Soundness and Completeness: Discussion}
 
-We now state the soundness theorem, which follows from
 Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
-as well as the following observation about the elaboration phase.
+show that the second and third steps of the type-inference algorithm are both
+sound and complete.
+It is easy to verify the soundness of the first step (the elaboration phase):
 For any program $p \in \absof{\FB}$, the elaboration phase produces a $q$
 such that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
+%
+The soundness of the type inference algorithm follows.
+The only sources of incompleteness in the type inference algorithm are some heuristic
+choices made in the first step, as explained below.
 
-\begin{theorem}
-\emph{(\textbf{Soundness})}
-For any $p \in \absof{\FB}$, if type inference succeeds, it returns a $t$ such that
-% Infer($p$) returns Some($t$), then
-(1) $\absof{t} = p$, and
-(2) $t$ is well-typed.
-\end{theorem}
+%% \begin{theorem}
+%% \emph{(\textbf{Soundness})}
+%% For any $p \in \absof{\FB}$, if type inference succeeds, it returns a $t$ such that
+%% % Infer($p$) returns Some($t$), then
+%% (1) $\absof{t} = p$, and
+%% (2) $t$ is well-typed.
+%% \end{theorem}
 
@@
 
-However, in this context, completeness is more interesting than soundness.
-(Note that an unsound inference algorithm can be made sound by simply type-checking
-the resulting program.)
-Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
-show that the second and third steps of the type-inference algorithm are complete.
-%% We believe that the converse of the above lemma can also be established:
-%% any substitution for the region variables and the predicate variables that makes the
-%% expression well-typed will also satisfy the generated constraints.
-%% Thus, if there is any possible instantiation (of the region variables and predicate variables) of
-%% the elaborated program produced by the first phase that type checks, then the generated
-%% constraints are satisfiable, and the second phase will identify it.
-%% 
-The only sources of incompleteness in the type inference algorithm are some heuristic
-choices made in the first step (the elaboration phase), as explained below.
+%% However, in this context, completeness is more interesting than soundness.
+%% (Note that an unsound inference algorithm can be made sound by simply type-checking
+%% the resulting program.)
+%% Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
+%% show that the second and third steps of the type-inference algorithm are complete.
 

### 2016-10-19 paper/implementation.tex
implementation

@@
 \label{sec:implementation}
+
+% We now briefly describe our experience of
+% working with these benchmarks. Note that the performance benefits of
+% using transferable regions in dataflow systems have been established
+% in~\cite{Broom:HotOS}. Our current evaluation only focuses on safety
+% aspects.
 
@@
 
-We have implemented the prototype of \name compiler frontend,
+We have implemented \namec, a prototype of \name compiler frontend,
 including its region type system and type inference, in 3k+ lines of
-OCaml. Our implementation is called \namec. The input to \namec is a
-program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
+OCaml.
+% Our implementation is called \namec.
+The input to \namec is a program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
 includes assignments, conditionals, loops, more primitive datatypes
@@
 
-% To solve the
-% constraints that arise during type inference, we built a solver called
-% \csolve that implements constraint solving approach based on fixpoint
-% computation and graph augmentation described in \S~\ref{sec:csolve}. 
-% Since graph augmentation is NP-hard, \csolve uses an approximation
-% algorithm that repeatedly adds an edge and recomputes paths in graph
-% $G_1$ until it is as connected as graph $G_2$. Though this approach
-% may not necessarily return the weakest solution to the abduction
-% problem (\S~\ref{sec:csolve}), it seems to do so in practice, since
-% constraints are not often complex. 
-% If the input $\absof{\FB^+}$ program does not create any unsafe
-% references, \namec annotates it with region types, which act as a
-% witness to program's memory safety. If potential safety violations are
-% encountered, then the type inference fails during the constraint
-% solving phase.
+We performed two kinds of experiments to evaluate our region type system and type
+inference.
+First, we implemented some 
+microbenchmarks ($\le$100 LOC) consisting of standard classes such
+as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
+our inference engine to infer their region types.
+These classes are region-oblivious. Hence, as long as they are well-typed
+as per the core type system, \namec must be able to automatically construct its
+region-type-annotated definition without fail. \namec was
+able to infer the expected region types for all these classes under 10ms.
+Fig~\ref{fig:rev} shows the region-type-annotated
+definition computed for the list reverse method. Observe that \namec
+was able to infer that the list and its data (of type \C{T}) can be
+allocated in different regions, as long as the later outlives the
+former. This allows, for instance, a \C{preOrder} method to traverse a
+tree in a transferable region, and return a list of its nodes, where
+the list itself is allocated in the stack region.
 
@@
 
-To evaluate the practical utility of our region type system and type
-inference, we 
-% implemented some of the C\# libraries in $\absof{\FB^+}$, and 
-performed two kinds of experiments. First, we implemented some of the
-microbenchmarks ($\le$100 LOC), which are the standard libraries such
-as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
-our inference engine to infer their principal region types. Since a
-library class is region-oblivious, if it is well-typed as per the core
-type system, then \namec must be able to automatically construct its
-region-type-annotated definition without fail. As expected, \namec was
-able to infer principal region types for all the library classes,
-under 10ms. Fig~\ref{fig:rev} shows the region-type-annotated
-definition computed for the list reverse method. Observe that \namec
-was able to infer that the list and its data (of type \C{T}) can be
-allocated in different regions, as long as the later outlives the
-former. This allows, for instance, a \C{preOrder} method to traverse a
-tree in a transferable region, and return a list of its nodes, where
-the list itself is allocated in the stack region.
-
 Next, we translated some of the Naiad streaming query operator
 benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
-$\absof{\FB^+}$, and used \namec to verify their safety. During the
-process, we found multiple instances of potential memory safety
+$\absof{\FB^+}$, and used \namec to verify their safety.
+The number of LOC performing operations on \C{Region} objects relative
+to the total LOC is 8\% or under in the Naiad benchmarks. 
+During the process, we found multiple instances of potential memory safety
 violations in the $\absof{FB^+}$ translation of benchmarks, which we
@@
 type inference/checking tool,
-%suggests that, although our
-%approach cannot statically enforce memory safety, it is nonetheless
-%useful, 
 particularly because it comes at no additional cost to the developer.
-If the developer is satisfied with
-the confidence gained by static verification, she may even choose to
-turn the runtime safety checks off. Since we did not implement code
-generation in \namec, we could not measure the runtime overhead of
-checks needed to enforce safety. However, the number of LOC performing
-operations on \C{Region} objects relative to the total LOC never
-exceeds 8\% in Naiad benchmarks. 
+%% If the developer is satisfied with
+%% the confidence gained by static verification, she may even choose to
+%% turn the runtime safety checks off. Since we did not implement code
+%% generation in \namec, we could not measure the runtime overhead of
+%% checks needed to enforce safety.
+
 % \C{SelectVertex}, for example, contains one each of
@@
 % \C{free} operation.
-
-% We now briefly describe our experience of
-% working with these benchmarks. Note that the performance benefits of
-% using transferable regions in dataflow systems have been established
-% in~\cite{Broom:HotOS}. Our current evaluation only focuses on safety
-% aspects.
-
-% \paragraph{Memory safety violation}

### 2016-10-19 paper/related.tex
related work

@@
 
-Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
-concept of a region type system to statically enforce the safety of
-their region-based memory management scheme in ML. Following their
-seminal work, static type systems for safe region-based memory
+% Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
+% concept of a region type system to statically enforce the safety of
+% their region-based memory management scheme in ML.
+Following Tofte and Talpin's seminal work in~\cite{tofte93,tofte94,tofte97},
+static type systems for safe region-based memory
 management have been extensively studied in the context of various
@@
 
-\item We adopt a two-pronged approach to memory safety that relies on a
-combination of a simple static type discipline and lightweight runtime
-checks. In particular, our approach circumvents the need for restrictive
+\item We exploit a combination of a simple static type discipline and lightweight
+  runtime checks to ensure memory safety.
+  % adopt a two-pronged approach to memory safety that relies on a
+In particular, our approach circumvents the need for restrictive
 static mechanisms (e.g., linear types and unique pointers) or
@@
 scoped (stack) regions (as a replacement for GC).
-% Similar to their region inference algorithm, our region type inference
-% algorithm can deal with region-polymorphic recursion by computing
-% fixed points for recursive constraints. 
-Our type inference is analogous to their region inference in some respects,
-while differing in others. While their inference algorithm only generates equality constraints, which
-can be solved via unification, our type inference algorithm also
-generates partial order outlives constraints, which are required to
-capture relationships between lifetimes of transferable regions
-and stack regions. Consequently, our constraint solving algorithm is
+Our type inference is analogous to their's in some respects,
+while differing in others.
+Their inference algorithm only generates equality constraints, solvable via unification.
+Our type inference algorithm generates partial order outlives constraints.
+% which are required to capture relationships between lifetimes of transferable regions and stack regions.
+Consequently, our constraint solving algorithm is
 more sophisticated, and is capable of inferring unknown outlives
@@
 Walker and Watkins~\cite{WW01} extend lambda calculus with first-class regions
-with dynamic lifetimes, and imposes linear typing to control accesses to regions.
+with dynamic lifetimes, and impose linear typing to control accesses to regions.
 Our open/close lexical block for transferable regions traces its origins to the \C{let!}
@@
 proposals~\cite{cyclone04,cycloneSCP} extends Cyclone with dynamic
-regions. \name differs from Cyclone fundamentally because of its
-non-intrusiveness design principle, which requires its safety
-mechanisms to not intrude on the the programming practices of C\#.
-\name programmers, for example, shouldn’t be forced to abandon
+regions. \name differs from Cyclone in its non-intrusiveness design principle,
+which requires its safety
+mechanisms to not intrude on the programming practices of C\#.
+\name programmers, for example, shouldn't be forced to abandon
 iterators in favor of for-loops, annotate region types, or rewrite
@@
 regions with dynamically determined lifetimes (in the context of message-passing concurrency).
-We borrow outlives relation from their formal
-development, and our type system bears some similarities to theirs.
-However, our language also admits parametric polymorphism (generics)
-and higher-order functions.
-Moreover, we establish type safety and transfer safety results that
+% We borrow outlives relation from their formal
+% development, and our type system bears some similarities to theirs.
+Our language also admits generics and higher-order functions.
+We also establish type safety and transfer safety results that
 formalize the guarantees provided by our system. While~\cite{MIT03}'s
@@
 type inference.
+
 % Although there is some support for local
@@
 
-~\cite{HMN01} proposes a flow-sensitive region-based memory management
-for first-order programs that proposes to overcome some of the
-drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
-to regions with dynamic lifetimes. However, dynamic regions are still
-not first-class values of the language, and reference counting is
-nonetheless needed to ensure memory safety.
-
-The idea of using region-based memory management to facilitate the
-safe transfer of rich data structures between computational nodes has
-been previously explored by~\cite{gpu14} in the context of Scheme
-language. However, their setting only includes lexically-scoped
+~\cite{HMN01} proposes a flow-sensitive approach for first-order programs
+% to overcome some of the drawbacks of~\cite{tofte97}
+to generalize~\cite{tofte97}'s approach to dynamic regions.
+However, dynamic regions are still not first-class values of the language,
+and reference counting is needed to ensure memory safety.
+%
+\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
+in the context of Scheme. However, their setting only includes lexically-scoped
 regions for which Tofte and Talpin-style analysis~\cite{tofte97}
-suffices. In contrast, our language provides first-class support for
-transferable regions with dynamic lifetimes. We require this
-generality in order to support streaming query operators, such as the
-one shown in Fig.~\ref{fig:motivating-eg}.
+suffices. In contrast, we provide first-class support for
+transferable regions with dynamic lifetimes.
+% We require this generality in order to support streaming query operators, such as the
+% one shown in Fig.~\ref{fig:motivating-eg}.

### 2016-10-19 paper/type-inference.tex
minor formatting

@@
 
-\begin{enumerate}
-\item We determine the set of region parameters for a recursive class
+% \begin{enumerate}
+% \item
+  (1) We determine the set of region parameters for a recursive class
 using the heuristic that a recursive occurrence of the class has
@@
 
-\item Our technique for region parameterization also uses a heuristic in the case of
+% \item
+  (2) Our technique for region parameterization also uses a heuristic in the case of
 higher order programs. The following examples illustrates that principal types may
@@
 
-\end{enumerate}
+% \end{enumerate}
 

### 2016-10-19 paper/type-inference.tex
constraint syntax

@@
 \paragraph{Syntax of Constraints.}
-The constraints are expressed using
-a set $\regionConstants$ of region constants,
-a set $\regionVars$ of region variables,
-and
-a set $\predVars$ of predicate variables.
-% 
-% ($\pi$ and $\rho$) and predicate variables ($\varphi$).
-% $(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
+(See~Fig.~\ref{fig:constraint-syntax}.)
+The constraints are expressed using a set $\regionConstants$ of region constants,
+a set $\regionVars$ of region variables, and a set $\predVars$ of predicate variables.
 
@@
 
-% \emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
 Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
@@
 
-Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
+A predicate variable $\varphi$ is introduced to represent, \eg, the unknown precondition of a method.
+The constraint-solver will end up binding it to a \emph{region-constraint} $\phi$
 over a set of fixed formal region parameters.
+
 %% where the set of region-constraints $\phi$ is defined by:
@@
 %% 
-We will use the term \emph{validity constraint} to denote an entailment constraint
-of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
+% We will use the term \emph{validity constraint} to denote an entailment constraint
+% of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
 
@@
 
-The constraints generated are of the following kinds:
-\begin{itemize}
+The constraints generated are primarily of the form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
+or $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$.  Here, $\varphi_i$ is a predicate variable
+(representing the precondition of a method to be determined), $\phicstr$ is a region-constraint
+that is \emph{required} to hold at a particular program point (within the method), and $\phictxt$ is
+a region-constraint that is \emph{known} to hold at that program point.
+Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
+are generated by an invocation of a method with precondition $\varphi_j$.
 
-\item Well-formedness constraints of form $\rho \in \rhoenv$,
-restricting the domain of unification for a region variable ($\rho$)
+We use \emph{well-formedness constraints} of the form $\rho \in \rhoenv$
+to restrict the domain of unification for a region variable ($\rho$)
 to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
+and  well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$ to
+restrict the domain of a predicate variable ($\varphi$) to the set of all possible
+constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.
 
-\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
-variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
-regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
-  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
-  the region-constraint of its base class \C{B}).
-
-\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
-  (e.g., to indicate the region-constraint of a class \C{C} must imply the
-  region-constraint of each of its fields).
-
-\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
-where $\varphi_i$ is a predicate variable (representing the precondition of a
-method to be determined), $\phicstr$ is a region constraint that is \emph{required}
-to hold at a particular program point (within the method), and $\phictxt$ is
-a region constraint that is \emph{known} to hold at that program point.
-
+%The constraints generated are of the following kinds:
+%\begin{itemize}
 %
-%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
-%% of predicate variables and pending substitutions. While $\phictxt$
-%% captures relationships that are \emph{known} to hold between concrete
-%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
-%% $\phicstr$ captures relationships that are \emph{required} to hold
-%% among region varibles (i.e., $\rho$'s), or relationships between
-%% region variables and identifiers.
+%\item Well-formedness constraints of form $\rho \in \rhoenv$,
+%restricting the domain of unification for a region variable ($\rho$)
+%to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
 %
-
-\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
-generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
-$\varphi_i$ are as above).
-
-\end{itemize}
+%\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
+%variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
+%regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
+%
+%\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
+%  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
+%  the region-constraint of its base class \C{B}).
+%
+%\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
+%  (e.g., to indicate the region-constraint of a class \C{C} must imply the
+%  region-constraint of each of its fields).
+%
+%\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
+%where $\varphi_i$ is a predicate variable (representing the precondition of a
+%method to be determined), $\phicstr$ is a region constraint that is \emph{required}
+%to hold at a particular program point (within the method), and $\phictxt$ is
+%a region constraint that is \emph{known} to hold at that program point.
+%
+%\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
+%generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
+%$\varphi_i$ are as above).
+%
+%\end{itemize}
 
@@
 
-Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
+Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language constructs.
 These rules use the same context as the corresponding typing judgment in
@@
 
-\TODO{Change the static semantics to be consistent with this, and
-make this a comment there.}
-One notable aspect here is the following. Consider a method with predicate
-variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
-We use $\varphi_m$ as the constraint formula in the initial context used for
-the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
-It is possible to show that there is no loss of generality in doing so.
-Specifically, it can be shown for any well-typed program, replacing the constraint
-$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
-where $\varphi_c$ is the constraint of the class containing the method,
-produces a well-typed program.
-
-We use the above simplification to avoid generating validity constraints of the form
-$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
-with more than one predicate variable on the left hand side, which introduce complications).
-%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.
+%\TODO{Change the static semantics to be consistent with this, and
+%make this a comment there.}
+%One notable aspect here is the following. Consider a method with predicate
+%variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
+%We use $\varphi_m$ as the constraint formula in the initial context used for
+%the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
+%It is possible to show that there is no loss of generality in doing so.
+%Specifically, it can be shown for any well-typed program, replacing the constraint
+%$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
+%where $\varphi_c$ is the constraint of the class containing the method,
+%produces a well-typed program.
+%
+%We use the above simplification to avoid generating validity constraints of the form
+%$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
+%with more than one predicate variable on the left hand side, which introduce complications).
+%%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.
 

### 2016-10-20 paper/introduction.tex
expanded type inference discussion in intro

@@
 \paragraph{Type Inference.}
-One of the novel contributions of this paper is a sound type inference algorithm
+One of the key contributions of this paper is a sound type inference algorithm
 that eliminates the need for users to write region type annotations.
-This, we believe, significantly reduces the impediment to adopt our
-approach in practical setting.
-Our inference algorithm proceeds in two stages: in the first stage,
-we introduce region parameters (for classes and methods), and in
-the second stage we infer the outlives constraints between these
-parameters (using an interprocedural fixed point computation).
+The users write programs in the underlying language that provides
+primitives for the users to manipulate regions (create, open, free,
+transfer) and to allocate objects in regions, but has no region type
+annotations.
+% This, we believe, significantly reduces the impediment to adopt our
+% approach in practical setting.
+
+Our inference algorithm proceeds in three stages: in the first stage,
+we elaborate the program by introducing region parameters (for classes and methods);
+in the second stage we generate a set of constraints that must be satisfied for
+the program to type check; in the third stage, we solve the set of constraints,
+inferring the preconditions of all classes and methods (in terms of the
+expected outlives-constraints between the region parameters).
+We show that the second and third stages are sound as well as
+complete (while the first stage is sound).
+% (using an interprocedural fixed point computation).
 % While the first

### 2016-10-20 paper/overview2.tex
minor changes

@@
 This is important as it enables \name applications to use existing
-region-oblivious C\# libraries. In particular, given a  C\# library
-function \C{f} (that makes no use of \name's region constructs), the
-statement ``\C{letregion R \{ f(); \}}'' invokes \C{f}, but has the
-effect that all objects allocated by this invocation are allocated in
-the new region \C{R}.
+region-oblivious C\# libraries, as explained soon.
+
+%In particular, given a  C\# library
+%function \C{f} (that makes no use of \name's region constructs), the
+%statement ``\C{letregion R \{ f(); \}}'' invokes \C{f}, but has the
+%effect that all objects allocated by this invocation are allocated in
+%the new region \C{R}.
 
@@
 at runtime.
+
+Opening a region also makes it the current allocation context.
+Thus, given a  C\# library function \C{f} (that makes no use of \name's region constructs),
+opening a region \C{R} and invoking \C{f} has the effect that all objects created by this
+invocation are allocated in the region \C{R}.
 

### 2016-10-20 paper/type-inference.tex
minor fix to proofs

@@
 
-\begin{lemma}
-  \label{lemma:completely-bound}
-Let $C = \consOf{q}$.
-Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
-\end{lemma}
+%\begin{lemma}
+%  \label{lemma:completely-bound}
+%Let $C = \consOf{q}$.
+%Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
+%\end{lemma}
 
@@
 \item For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
-set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$.
+set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$,
+if this set is non-empty. If this set is empty for any $\rho$, $\thesoln$ is undefined.
 \end{itemize}
@@
 \begin{cases}
-\textsc{Some} (\thesoln) & \textrm{if $\groundC$ is valid and $\thesoln$ satisfies $\rhoC$} \\
+\textsc{Some} (\thesoln) & \textrm{if $\groundC$ is valid and $\thesoln$ is defined and satisfies $\rhoC$} \\
 \textsc{None} & \textrm{otherwise}

### 2016-10-20 paper/type-inference.tex
updates to proofs

@@
 We similarly perform instantiation for every constructor invocation
-of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
-before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
-where $\rho_0, \cdots, \rho_n$ are fresh region variables.
+of the form \C{new@$\pi_0$ T($\ldots$)}, by instantiating the type \C{T} as
+before, turning it into \C{new T$\langle \pi_0, \rho_1, \cdots, \rho_n \rangle$($\ldots$)},
+where $\rho_1, \cdots, \rho_n$ are fresh region variables. Note that the programmer
+typically specifies the region $\pi_0$ where the object is to be allocated. If the programmer
+does not specify this, it is allocated in the allocation-context region by default. Region
+variables are introduced only for the other region parameters.
 

### 2016-10-20 paper/typeinference-proofs.tex
minor fix to proofs
updates to proofs

@@
 
-\begin{theorem}
-\label{thm:constraint-generation-sc}
-Let $C = \consOf{q}$.
-An assignment $\soln$ (for the region and predicate variables in $q$)
-satisfies $C$ iff $q[\soln]$ is well-typed.
-\end{theorem}
-\begin{proof}
+%\begin{theorem}
+%\label{thm:constraint-generation-sc}
+%Let $C = \consOf{q}$.
+%An assignment $\soln$ (for the region and predicate variables in $q$)
+%satisfies $C$ iff $q[\soln]$ is well-typed.
+%\end{theorem}
+
+\begin{proof}[\textbf{Theorem~\ref{thm:constraint-generation-sc}}]
 The correspondence between the static semantics rules and the constraint generation
@@
 (We make use of analogous lemmas for type well-formedness rules as well.)
+\qed
 \end{proof}
 
-\begin{lemma}
-  \label{lemma:gc-is-decomposable}
-  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
-\end{lemma}
+%\begin{lemma}
+%  \label{lemma:gc-is-decomposable}
+%  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
+%\end{lemma}
 
-\begin{proof}
+\begin{proof}[\textbf{Lemma~\ref{lemma:gc-is-decomposable}}]
   By induction over the constraint-generation rules.
@@
   as conjuncts to $\phicx$.
+\qed
 \end{proof}
@@
 
-\begin{lemma}
-  \label{lemma:decomposition}
-Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
-where both $\pi_i$ and $\pi_j$ are region constants.
-(a) If $\{ \pi_i, \pi_j \} \subseteq \deltaPC(\varphi)$,
-then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \deltaPC(\varphi)$,
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
-iff
-$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
-\end{lemma}
+%\begin{lemma}
+%  \label{lemma:decomposition}
+%Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+%where both $\pi_i$ and $\pi_j$ are region constants.
+%(a) If $\{ \pi_i, \pi_j \} \subseteq \deltaPC(\varphi)$,
+%then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+%iff
+%$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
+%  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \deltaPC(\varphi)$,
+%$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+%iff
+%$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
+%\end{lemma}
 
-\begin{proof}
+\begin{proof}[\textbf{Lemma~\ref{lemma:decomposition}}]
   Let $\soln$ be any assignment that satisfies $C$.
@@
   The result follows.
+\qed
 \end{proof}
 
-\begin{theorem}
-  \label{thm:closure}
-Let $C = \consOf{q}$.
-An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
-\end{theorem}
+%\begin{theorem}
+%  \label{thm:closure}
+%Let $C = \consOf{q}$.
+%An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+%\end{theorem}
 
-\begin{proof}
+\begin{proof}[\textbf{Theorem~\ref{thm:closure}}]
   For the forward implication, we use induction to show that every constraint
@@
   The reverse implication is trivial.
+\qed
 \end{proof}
@@
   \label{lemma:completely-bound}
-Let $C = \consOf{q}$.
-Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
+Let $C = \consOf{q}$. If $\satC$ is satisfiable, then
+every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
 \end{lemma}
@@
 some region constant (by the generated constraints).
+%
+Region variables are introduced at call-sites and represent the formal region parameters for the call.
+Some correspond to region parameters of types of input parameters, and these will be bound by the
+actual input parameter expressions. Other region variables correspond to region parameters of the
+return type. These are determined by the actual return-expression in the called function/method.
+It can be shown that these will be bound to one of ``input region parameters'', provided the called
+function/method is typable (\ie, if $\satC$ is satisfiable).
+
 There is one special case where this does not hold: a recursive function
@@
 
-This property follows from several aspects of $\FB$, e.g., it does not admit uninitialized variables.
+% This property follows from several aspects of $\FB$, e.g., it does not admit uninitialized variables.
 Note that the region \C{R} where a new object is allocated (by the \C{new} construct) is indicated
@@
 the constraint solver.
-
+\qed
 % This property follows from the constraint-generation rules.
@@
   for $\satC$.
+\qed
 \end{proof}
@@
 The same idea works for constraints with occurrences of two region variables in the consequent.
-
+\qed
 \end{proof}
@@
 Hence, it follows that $\hat{\pi} \in \Delta$.
+\qed
 % for some predicate variable $\varphi$.
-
+%
 % If $\groundC$ is satisfiable, the solver will fail to return an assignment only if the set $S_\rho$ = 
@@
 
-\begin{theorem}
-\label{thm:constraint-solver-sc}
-Let $C = \consOf{q}$.\\
-(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.\\
-(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
-\end{theorem}
+%\begin{theorem}
+%\label{thm:constraint-solver-sc}
+%Let $C = \consOf{q}$.\\
+%(Soundness) If $\solveCon{C} = \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.\\
+%(Completeness) If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
+%\end{theorem}
 
-\begin{proof}
+\begin{proof}[\textbf{Theorem~\ref{thm:constraint-solver-sc}}]
 
 Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.
-
+\qed
 \end{proof}

### 2016-10-21 paper/typeinference-proofs.tex
unification proof

@@
 \begin{proof}
-Intuitively, the lemma says that every region variable will be unified with
+The lemma formalizes the intuition that every region variable will be ``unified'' with
 some region constant (by the generated constraints).
@@
 the constraint solver.
+
+Formally, we can prove via induction that if $\exprok{\stdcontext}{e}{\tau}{C}$ then
+every region variable in the result type $\tau$ will be bound to one of the ``live''
+region constants (\ie, region constants in $\A$).
 \qed
-% This property follows from the constraint-generation rules.
-% , and the type system uniquely determines the region parameters of every value
+
 \end{proof}

### 2016-10-21 paper/type-inference.tex
minor fixes

@@
+\newcommand{\deltaPC}{\Delta_P^C}
+
 \newcommand{\soln}{\eta}
@@
 of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
-$\predDeltaMap(\varphi)$.
+$\deltaPC(\varphi)$.
 %
@@
 %
-We refer to an abduction constraint in $C$ as \emph{decomposable}
+We say that an abduction constraint is \emph{$C$-decomposable}
 if its antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
@@
 $\pi_1 \outlives \pi_2$ satisfying the following conditions:
-(1) $\pi_2 \notin \predDeltaMap(\varphi)$.
-(2) if $\pi_1 \in \predDeltaMap(\varphi)$, then for
-every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
-a conjunct in $\phictxt$.
+(1) $\pi_2 \notin \deltaPC(\varphi)$.
+(2) if $\pi_1 \in \deltaPC(\varphi)$, then for
+every $\pi_f \in \deltaPC(\varphi)$, $\pi_f \outlives \pi_2$ is a conjunct in $\phictxt$.
 %
-(The above concepts parametrically depend on the set $C$, though we omit $C$
-from the notation for simplicity.)
+% (The above concepts parametrically depend on the set $C$, though we omit $C$
+% from the notation for simplicity.)
+We will omit the reference to $C$ in the above notation if no confusion is likely.
 
@@
   \label{lemma:gc-is-decomposable}
-  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
+  Let $C = \consOf{q}$. Every abduction constraint in $C$ is $C$-decomposable.
 \end{lemma}
@@
   \label{lemma:decomposition}
-  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+Consider any $C$-decomposable constraint $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
 where both $\pi_i$ and $\pi_j$ are region constants.
-(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
-then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+Let $\soln$ satisfy $C$.\\
+(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$:
+$\soln$ satisfies $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
 iff
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
-  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
-$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
+$\soln$ satisfies $\isvalid{\varphi}{\pi_i \outlives \pi_j}$.\\
+(b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$:
+$\soln$ satisfies $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
 iff
-$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
+$\soln$ satisfies $\isvalid{\phi}{\pi_i \outlives \pi_j}$
 \end{lemma}

### 2016-10-21 paper/introduction.tex
minor fixes

@@
 % (using an interprocedural fixed point computation).
-% While the first
+
+\paragraph{Evaluation.}
+Our work was inspired by~\cite{Broom:HotOS}, which presents evidence that
+realistic programs can be written using transferable regions
+and that this can yield significant performance improvements.
+The system described in~\cite{Broom:HotOS}, however, provides no memory
+safety guarantees, which is the subject of our contribution.
+We report on an implementation of our type inference algorithm that
+was able to identify some unsafe memory accesses and was able to successfully
+verify programs correct (once the errors were fixed).
 

### 2016-10-21 paper/type-inference.tex
minor fixes

@@
 Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language constructs.
+(The appendix contains the remaining constraint generation rules.)
 These rules use the same context as the corresponding typing judgment in
@@
 indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.
+
+Note that some of the conditions in the antecedents are checked at constraint-generation
+time, while some are encoded as constraints (to be resolved by the constraint-solver).
+E.g., we assume that the region constants introduced by
+\C{open}
+and
+\C{letregion}
+constructs are unique (and the elaboration phase ensures this by alpha-renaming
+them as needed). Thus, the precondition $\rho \not\in \Delta$ in rules \textsc{LetRegion}
+and \textsc{Open} merely validates this at constraint-generation time.
 

### 2016-10-21 paper/implementation.tex
minor fixes

@@
 As mentioned earlier, our work is a continuation of the work reported in~\cite{Broom:HotOS},
-which provides users with a C\#-based implementation of \broom (the language described in
-Section~\ref{sec:overview}). This system has no type system and provides users with no safety
-guarantees, but does shed light on the performance gains as well as the viability of the
-approach (in terms of expressiveness). In particular, it reports speedups up to 34\% for
-typical big-data analytics jobs.
+which provides users with a C\#-based implementation of transferable regions (the features
+described in Section~\ref{sec:overview}). This system has no type system and provides users
+with no safety guarantees. \cite{Broom:HotOS} presents evidence that realistic programs can
+be implemented using transferable regions and that this can yield significant performance gains.
+In particular, it reports speedups up to 34\% for typical big-data analytics jobs.
 We now describe our implementation and experience with the type system and type inference

### 2016-10-21 paper/type-inference.tex
minor fixes

@@
 
-\name's region type system imposes a heavy annotation burden, and
+\name's region type system imposes an annotation burden, and
 manually annotating C\# standard libraries with region types
@@
    \emph{actual region parameters}). We also introduce for each class and method, a
-   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
+   \emph{predicate variable} ($\varphi$) to denote an undetermined set of outlives-constraints
    over the region parameters of that class/method.
@@
 a fresh predicate variable $\varphi$, which represents the yet undetermined
-outlives constraints between the formal region parameters of class \C{C}.
+outlives-constraints between the formal region parameters of class \C{C}.
 
@@
 \begin{tabular}{ccc}
-\begin{minipage}{0.35\linewidth}
+\begin{minipage}{0.33\linewidth}
 \begin{codejava}
@@
 &
-$\Rightarrow$
+$\Rightarrow \; \;$
 &
@@
 
-Function types (of fields and parameters) are instantiated just like methods above, except that
-an additional fresh region identifier is introduced to represent the region where the closure is
-allocated.
+Function types (of fields and parameters) are instantiated just like methods above,
+(and the region where the closure is allocated is determined just as for other objects).
 For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
-$\inang{\rho_0, \rho_1, \rho_2, \rho_3, \rho_4 \,|\, \varphi } \C{List} \inang{\rho_1,\rho_2} \xrightarrow{\rho_4} \C{Object}\inang{\rho_3}$.
+$\inang{\rho_0, \rho_1, \rho_2, \rho_3 \,|\, \varphi } \C{List} \inang{\rho_1,\rho_2} \xrightarrow{\pi} \C{Object}\inang{\rho_3}$.
 Note that the newly introduced region identifiers are generalized as formal region parameters of the function type.
@@
 over a set of fixed formal region parameters.
-
-%% where the set of region-constraints $\phi$ is defined by:
-%% \begin{smathpar}
-%% \begin{array}{lcl}
-%% \phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
-%% \end{array}
-%% \end{smathpar}
+%
 %% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
@@
 % of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.
-
+%
 Our constraints also make uses of \emph{pending substitutions} $F$:
@@
 used in a particular context:
-%% \begin{smathpar}
-%% \begin{array}{lcl}
-%% F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
-%% \end{array}
-%% \end{smathpar}
 E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
 \rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
-is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
-$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
-$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
-and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
-deducible from $\rgn_1 \outlives \rgn_2$.
+is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$.
+% Any concrete formula (call it
+% $\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
+% $\varphi$ if and only if the formula obtained by substituting $\rgn_1$
+% and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
+% deducible from $\rgn_1 \outlives \rgn_2$.
 
-The constraints generated are primarily of the form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
+The constraints are primarily of the form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
 or $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$.  Here, $\varphi_i$ is a predicate variable
@@
 a region-constraint that is \emph{known} to hold at that program point.
-Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
+Constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
 are generated by an invocation of a method with precondition $\varphi_j$.
@@
 
-%\begin{definition}
-%\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
-%A constraint satisfaction problem is a tuple
-%$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
-%where $\regionConstants$ is a set of region constants,
-%$\regionVars$ is a set of region variables,
-%$\predVars$ is a set of predicate variables,
-%$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
-%specifies a unification domain for each region variable,
-%$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
-%specifies the domain for each predicate variable,
-%and $\constraintSet$ is a set of validity constraints in one of the following forms:
-%\begin{center}
-%\(
-%    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
-%    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
-%\)
-%\end{center}
-%The solution to the constraint satisfaction problem is a pair $(\solnR,\solnP)$,
-%where $\solnR$ is a map from $\regionVars$ to $\regionConstants$
-%and $\solnP$ is a map from $\predVars$ to a region-constraint formula such that
-%\begin{itemize}
-%  \item $\solnR(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,
-%
-%  \item $\solnP$ is well-formed under $\predDeltaMap$
-%    (i.e., $\tywf{\predDeltaMap(\varphi)}{\solnP(\varphi)}$, for every $\varphi \in \predVars$).
-%
-%  \item Every sequent in $C$ is valid after substitutions $\solnP$ and $
-%    \solnR$.
-%  %% $C\lbrack (\solnR,\solnP) \rbrack$ is valid.
-%\end{itemize}
-%\end{definition}
-
 \input{fb-constraint-generation}
@@
 Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
+Thus, the component $\phicx$ of $\A$ will now be a symbolic expression
+of the form $\varphi \conj \phi$, where $\varphi$ is the predicate variable
+representing the undetermined precondition of the method being analyzed.
 The algorithm proceeds top-down, analyzes an expression $e$ in a context
@@
 variables ($\varphi$ and $\varphi_m$) to capture constraints over
-variables in $\rhoenv$ denoting the fact that such constraints are yet
-to be inferred.
+variables in $\rhoenv$ that are yet to be inferred.
 
@@
 
+Let $\consOf{q}$ denote the set of constraints generated from an elaborated program $q$.
+The following theorem states that constraint-generation is sound and complete:
 \begin{theorem}
@@
 The above lemma shows how we can reduce a non-trivial abduction constraint to
-either a trivial abduction constraint or a non-abduction constraint.
+either a trivial abduction constraint or a non-abduction constraint,
+provided that the consequent is an outlives-constraint without region variables.
 
@@
 The set $\satC$ makes it easy to identify a solution $\thesoln$ (if one exists), as below.
-\begin{itemize}
-\item For any predicate variable $\varphi$, $\thesolnP(\varphi)$ is defined
+% \begin{itemize}
+% \item
+For any predicate variable $\varphi$, $\thesolnP(\varphi)$ is defined
 to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
-\item For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
+% \item
+For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
 set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$,
 if this set is non-empty. If this set is empty for any $\rho$, $\thesoln$ is undefined.
-\end{itemize}
-
-% \paragraph{Step 2.}
-
-% \begin{enumerate}
-% \item We use an iterative fixed point (saturation) algorithm to compute
-% $\saturate{C}$, the set of all constraints logically implied by $C$.
-% \item
+% \end{itemize}
 
@@
 Define $\solveCon{C}$ as below.
-% It is possible to show that $\soln$ satisfies $\satC4
-% iff  $\myground{C}$ is valid in the theory of partial-order.
-
-% \end{enumerate}
 
@@
 \begin{tabular}{ccc}
-\begin{minipage}{0.35\linewidth}
+\begin{minipage}{0.32\linewidth}
 \begin{codejava}
@@
 &
-$\Rightarrow$
+$\Rightarrow  \; \; \;$
 &

### 2016-10-22 paper/type-system.tex
minor changes
fixed fig ref
added missing param Delta
Sec.3 edited.

--- 2016-07-07 paper/type-system.tex
@@
 % well-formedness for the core (region-free) language, which are
-% elaborated further in the supplement~\cite{techrep}.
+% elaborated further in the appendix~\cite{techrep}.
 
@@
   b@$\rho_2$ snd;
-  a@$\rho_1$ getFst() { 
-    return this.fst; 
-  }
+  a@$\rho_1$ getFst() { return this.fst; }
 }
@@
 own allocation region ($\rhoalloc$), provided that the former outlive
-the later (i.e., $\rhobar \outlives \rho$). In such case, the
+the latter (i.e., $\rhobar \outlives \rhoalloc$). In such case, the
 definition of object's class needs to be parametric over allocation
@@
 To construct objects of the \C{Pair} class, its type and region
-parameters need to be instantiated with core types ($T$) and concrete
-region identifiers\footnote{Region variables ($\rho$) and region
-identifiers ($\rgn$) belong to the same syntactic class. Region
-identifiers are to region variables, as types ($T$) are to type
-variables ($a$)} ($\rgn$), respectively. For example:
+parameters need to be instantiated with core types ($T$) and region
+annotations\footnote{Region annotations ($\rgn$) include region
+variables ($\rho$) and region identifiers ($\pi$). Region identifiers
+are to region variables, as types ($T$) are to type variables ($a$)}
+($\rgn$), respectively. For example:
 \begin{codejava}
-letregion $\rgn_0$ in
-  let snd = new Object<$\rgn_0$>() in
-  letregion $\rgn_1$ in
-    let fst = new Object<$\rgn_1$>() in
-    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
-                  (fst,snd);
+letregion $\pi_0$ in
+  let snd = new Object<$\pi_0$>() in
+  letregion $\pi_1$ in
+    let fst = new Object<$\pi_1$>() in
+    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
 \end{codejava}
@@
 In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
-$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
-and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
-\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
+$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
+and $\pi_1$ are live during the instantiation, and (b). $\pi_0
+\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
 reflexive). Observe that the region type of \C{p} conveys
-the fact that (a). it is allocated in region $\rgn_1$, and (b). it
-holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
+the fact that (a). it is allocated in region $\pi_1$, and (b). it
+holds references to objects allocated in region $\pi_0$ and $\pi_1$.
 In contrast, if we choose to allocate the \C{snd} object also in
-$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
-type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
-we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, we treat
-$B\inang{\tbar}@\rgn$ as being equivalent to
+$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
+type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
+we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
+$B\inang{\tbar}@\pi$ as being equivalent to
 $B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
-$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
+$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
 instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
-\C{Pair} object contained in  $\rgn$. Separate parameterization over
-types and regions is needed in \FB to support region-polymorphic
-higher-order functions (explained below) as class fields. Class fields
-are not allowed to be type polymorphic, in keeping with C\#, which
-constitutes the core of \name. For class fields that are functions,
-regions are parameterized at the (class) field level whereas types are
-parameterized at the class level.  This choice also simplifies type
-inference in the presence of higher-order functions. The type $\unitZ$
-is unboxed in \FB, hence it has no region annotations.
+\C{Pair} object contained in  $\pi$. Class fields are not allowed to
+be type polymorphic, in keeping with C\#, which constitutes the core
+of \name. Separate parameterization over types and regions is needed
+in \FB to support region-polymorphic higher-order functions as class
+fields. For such higher-order class fields, regions are parameterized
+at the (class) field level whereas types are parameterized at the
+class level. This allows, for example, a generic class, whose type
+parameters are $a$ and $b$, to contain a region-polymorphic function of type
+$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
+\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
+\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
+region-polymorphic higher-order fields are used frequently by the
+dataflow operators, which apply them in the context of various regions
+(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
+parameterization seperate from region parameterization allows \FB to
+type such examples. This choice also simplifies type inference in the
+presence of higher-order functions.
 
@@
 \FB extends FGJ's expression language with a lambda expression and an
-application expression ($e\inang{\ralloc\rbar}(\bar{e})$) to define
-and apply functions\footnote{We distinguish between functions and
-methods. The former result from lambda expressions, whereas the latter
-come from class definitions}. Functions, like methods, exhibit
+application expression ($e\inang{\rbar}(\bar{e})$) to define
+and apply functions (lambdas). Functions, like methods, exhibit
 constrained region polymorphism, as evident in their arrow region type
-($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
+($\inang{\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
 \tau$).
@@
 % the former corresponds to the region where it is applied
-. 
 % For instance, in the following example:
@@
 Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
-rules\footnote{The full formal development can be found in the
-supplement}. The rules refer to a context ($\A$), which is a tuple of:
+rules\footnote{Full formal development can be found in the
+appendix}. The rules refer to a context ($\A$), which is a tuple of:
 \begin{itemize}
-\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
-that are estimated to be live,
+\item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
+live,
 \item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
 variables to their \emph{bounds}, i.e., classes they are declared to
-extend, and
+extend (this artifact is inherited from FGJ), and
 \item A constraint formula ($\phicx$) that captures the outlives
@@
 \end{itemize}
-$\A$ also contains $\rhomap$, 
-% a map ($\rhomap \in \rgn \mapsto
-% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
-% transferable regions to their current state in the FSM from
-% Fig.~\ref{fig:region-fsm}.
-which is primarily an artifact to facilitate the type safety proof,
-and can be ignored while type checking user-written programs
-in \FB. The context for the expression typing judgment also includes:
-\begin{itemize}
-\item A type environment ($\env \in x \mapsto \tau$) that contains the
-type bindings for variables in scope, and 
-\item The static identifier ($\rgn^a$) of the allocation context for
-the expression is being typechecked.  
-\end{itemize}
-Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
-rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
-class table ($CT \in cn \mapsto D$) that maps class names to their
+In addition, the context for the expression typing judgment also
+includes a type environment ($\env \in x \mapsto \tau$) that contains
+the type bindings for variables in scope. Like the judgments in
+FGJ~\cite{fgj}, all the judgments defined by the rules in
+Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a class
+table ($CT \in cn \mapsto D$) that maps class names to their
 definitions in \FB.
@@
 The well-formedness rule for class types
-($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
-the safety property that prevents objects from containing unsafe
-references. It does so by insisting that regions $\ralloc\rbar$
-satisfy the constraints ($\phi$) imposed by the class on its region
-parameters. The later is enforced by checking the validity of $\phi$,
-with actual region arguments substituted for formal
-region parameters, under the conditions ($\A.\phicx$) guaranteed by
-the context. The semantics of this sequent is straightforward, and
-follows directly from the properties of outlives and equality
-relations. For any well-formed core type $T$, $T@\rgn$ is a
-well-formed region type if $\rgn$ is a valid region. The type
-$\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
+($B\inang{\tbar}\inang{\rbar}$) is responsible for enforcing the
+safety property that prevents objects from containing unsafe
+references. It does so by insisting that regions $\rbar$ satisfy the
+constraints ($\phi$) imposed by the class on its region parameters.
+The later is enforced by checking the validity of $\phi$, with actual
+region arguments substituted for formal region parameters, under the
+conditions ($\phicx$) guaranteed by the context. The semantics of this
+sequent is straightforward, and follows directly from the properties
+of outlives and equality relations. For any well-formed core type $T$,
+$T@\rgn$ is a well-formed region type if $\rgn$ is a valid region. The
+type $\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
 $\toprgn$ is a special immortal region that outlives every other live
@@
 references between transferable regions, which become unsafe in
-context of the recipient's address space. Fortunately, such
-references are explicitly prohibited by the type rule of \C{\RgnZ}
-objects, as described below.
+context of the recipient's address space. Fortunately, such references
+are explicitly prohibited by the type rule of \C{\RgnZ} objects, as
+described below.
 
@@
 objects of other classes. The rule for the latter relies on an
-auxiliary definition called $\fields$ that returns the sequence of
-type bindings for fields (instance variables) of a given class type.
-Like in FGJ, the names and types of a constructor's arguments in \FB
-are same as the names and types of its class's fields, and the type
-rule relies on this fact to typecheck constructor applications. Note
-that this rule does not apply to \C{new} expressions involving \RgnZ
-class, as we do not define $\fields$ for \RgnZ.
+auxiliary definition called $\fields$ (undefined for $\RgnZ$ class)
+that returns the sequence of type bindings for fields (instance
+variables) of a given class type.  Like in FGJ, the names and types of
+a constructor's arguments in \FB are same as the names and types of
+its class's fields, and the type rule relies on this fact to typecheck
+constructor applications.
 
@@
 constructor to be called with a nullary function that returns a value
-in its allocation context. It enforces this by typechecking the body
-($e$) of the function under an empty context containing nothing but
-the allocation context of the function. This step ensures that the
-value returned by the function stores no references to objects
-allocated elsewhere, including the top region ($\toprgn$), thus
-preventing cross-region references originating from transferable
-regions. The body ($e$) of the function might, however, create
-new (transferable) regions while execution, but that is fine as long
-as such regions, and objects allocated in them, don't find their way
-into the result of evaluating $e$.
+in its allocation context. 
+% It enforces this by typechecking the body ($e$) of the function
+% under an empty context containing nothing but the allocation context
+% of the function. 
+This ensures that the value returned by the function stores no
+references to objects allocated elsewhere, including the top region
+($\toprgn$), thus preventing cross-region references originating from
+transferable regions. The body of the function might however create
+new regions while execution, but this is not a problem as long as such
+regions, and objects allocated in them, don't find their way into the
+result of its evaluation.
 
@@
 context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
-order to prevent the new region from incorrectly assuming any existing
+order to prevent the new region from incorrectly assuming existing
 outlives relationships on an eponymous region. Provided this is
@@
 \C{letregion} expression contains no references to the temporary
-objects inside the region\footnote{The supplement contains a formal
-proof of this claim}.
+objects inside the region.
 
@@
 parameters hold. The constraints ($\phi$) are required to be
-well-formed under $\rhoenv'$, which means that $\phi$ must only refer
-to the region variables in the set $\rhoenv'$. Note that the closure
-is always allocated in the current allocation context ($\rgn^a$). This
-prevents the closure from escaping the context in which it is created,
-thus trivially ensuring the safety of any dereferences inside the
-closure.
+well-formed under $\rhoenv$ extended with the function's region
+parameters ($\rhobar$).The premise $\forall
+(\rgn'\in\Delta).~(\isvalid{\phicx}{\rgn\outlives\rgn'}) ~\Rightarrow~
+\rgn = \rgn'$ ensures that the closure is always allocated in the
+last-opened region (also called the \emph{current allocation context}).
+This prevents the closure from escaping the context in which it is
+created , thus trivially ensuring the safety of any dereferences
+inside the closure.
 
@@
 
-Fig.~\ref{fig:fb-opsem} in the appendix defines a small-step operational semantics for \fbname
-via a five-place reduction relation:
+Fig.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} in the appendix
+defines a small-step operational semantics for \fbname via a
+five-place reduction relation:
 % \vspace*{-0.15in}
 \begin{smathpar}
-  \redstoo{(e,\rhomap)}{(e',\rhomap')}
+  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
 \end{smathpar}
-The reduction judgment should be read as following: given a set
-($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
-from unique identifiers of transferable regions to their current
-states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
-$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
-``stuck'' if $e$ attempts to access an object whose allocation region
-is not present in $\rhoenv$, or if $e$ tries to \C{open} a
-transferable region, whose identifier is not mapped to a state by
-$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
-on a $\RgnZ$ object that is not sanctioned by the transition
-discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
-value ($\invalidexn$).
+Operational semantics replaces abstract region annotations ($\pi$ and
+$\rho$) with \emph{locations} ($\loc$), where each \emph{location}
+abstracts all the memory locations associated with a region (i.e.,
+each memory region is associated with a single location). The
+typestate of regions is tracked by a finite map ($\mem$) from
+locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
+relation should be read as following: given a set ($\rhoenv$) of
+regions that are currently live, and an initial typestate map
+($\mem$), the expression $e$ reduces to $e'$, while updating $\mem$ to
+$\mem'$. The semantics gets ``stuck'' if $e$ attempts to access an
+object whose allocation region is not present in $\rhoenv$, or if $e$
+tries to \C{open} a transferable region that is not mapped to an
+appropriate typestate by $\mem$.  On the other hand, if $e$ attempts
+to commit an operation on a $\RgnZ$ object that is not sanctioned by
+the transition discipline in Fig.~\ref{fig:region-fsm}, then it raises
+an exception value ($\invalidexn$).
 
-To help state the type safety theorem, we define the syntactic class of values:
+Replacing $\pi$'s and $\rho$'s with $\loc$'s at runtime results in a
+new class of expressions that are not captured in
+Fig.~\ref{fig:fb-syntax}. Appendix contains full syntax that includes
+such expressions. To help state the type safety theorem, we define the
+syntactic class of (runtime) values:
 \begin{smathpar}
 \begin{array}{lclcl}
-v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
-\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
-  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
+v & \in & \mathtt{values} & \coloneqq & \C{new}\;
+    B\inang{\tbar}\inang{\overline{\loc}}(\bar{v}) \ALT
+\lambdaexp{\loc}{\rhobar\,|\,\phi}{\taubar \; \xbar}{e} \ALT
+\C{new}\; \RgnZT{\toploc\loc}(v)
+%  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
 \end{array}
 \end{smathpar}
-Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
-$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
-\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
-$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
-identifier ($\rgn$) for this region. A binding is also added to
-$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
-defines a type rule for such values, allowing them to be typed.
-The following type safety theorem shows that a well-typed program will never
-attempt to dereference an ``invalid'' reference (a reference to an object
-in a region that has been transferred or freed):
+The first two forms are obtained by using locations for region
+annotations in $\C{new}$ and lambda expressions. The last form is the
+value that $\C{new}\;\RgnZ$ expressions gets evaluated to; the
+location $\toploc$ stands for the region $\toprgn$, and $\loc$ is the
+location of the newly allocated transferable region.  The following
+type safety theorem shows that a well-typed program will never attempt
+to dereference an ``invalid'' reference (a reference to an object in a
+region that has been transferred or freed):
 \begin{theorem}

### 2016-10-22 paper/type-system.tex
Type safety stmt edited. Transfer safety removed.

@@
 \label{thm:fb-type-safety}
-$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
-\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
+$\forall e, \tau, \Delta, \mem$, such that $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
 \cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
-exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
-$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
-and $\hastyp{\emptyASigp,\rgn, \cdot}{e'}{\tau}$.
+exception ($\redstoo{\Delta}{(e,\mem)}{\invalidexn}$), or there exists an
+$e'$ and a $\mem'$ such that $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$
+and $\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
 \end{theorem}
 
-Furthermore, we prove the following theorem about \FB, which, in
-conjunction with the type safety theorem, implies the safety of region
-transfers across address spaces:
-\begin{theorem}
-\emph{(\textbf{Transfer Safety})}
-\label{thm:fb-transfer-safety}
-$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
-$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in
-\rhoenv'$, $\rgn_i \neq \toprgn$ and
-$\rgn_i \notin dom(\rhomap') \cup \rhoenv'$, if $\hastyp{\emptyA,\rgn,
-\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
-$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
-{\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
-\end{theorem}
-The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
-is well-typed under one context, then it is also well-typed under
-every other context, whose $\rhomap$ maps $\rgn_i$ to closed
-($\CLOSED$) state.  Thus, a recipient of a transferable region only
-needs to add a binding for the region to its $\rhomap$ in order to
-preserve its type safety. 
+% Furthermore, we prove the following theorem about \FB, which, in
+% conjunction with the type safety theorem, implies the safety of region
+% transfers across address spaces:
+% \begin{theorem}
+% \emph{(\textbf{Transfer Safety})}
+% \label{thm:fb-transfer-safety}
+% $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\mem$, $\mem'$, $\phicx$,
+% $\phicx'$, such that $\loc \not\in dom(\mem')$, if $\hastyp{\emptyA, \cdot}{\C{new}\;
+% \RgnZT{\toploc\loc}(v)}{\RgnZT{\toprgn}}$, then
+% $\hastyp{\rhoenv',\cdot,\phicx')}
+% {\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
+% \end{theorem}
+% The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
+% is well-typed under one context, then it is also well-typed under
+% every other context, whose $\mem$ maps $\rgn_i$ to closed
+% ($\CLOSED$) state.  Thus, a recipient of a transferable region only
+% needs to add a binding for the region to its $\mem$ in order to
+% preserve its type safety. 
 % Notably, this result could not have been

### 2016-10-22 paper/type-system.tex
Fixed lambda type rule and constr. gen rule.

@@
 well-formed under $\rhoenv$ extended with the function's region
-parameters ($\rhobar$).The premise $\forall
-(\rgn'\in\Delta).~(\isvalid{\phicx}{\rgn\outlives\rgn'}) ~\Rightarrow~
-\rgn = \rgn'$ ensures that the closure is always allocated in the
-last-opened region (also called the \emph{current allocation context}).
-This prevents the closure from escaping the context in which it is
-created , thus trivially ensuring the safety of any dereferences
-inside the closure.
+parameters ($\rhobar$). The premise $\forall
+(\rgn'\in\frv(e)\setminus\{\rhobar\}). ~\isvalid{\phicx}
+{\rgn'\outlives\rgn}$ ensures the safety of dereferences inside the
+closure by requring that all free region variables ($\frv$) trapped by
+the closure outlive the allocation region of the closure.
 

### 2016-10-22 paper/type-system.tex
Type safety theorem stmt edited.

@@
 \label{thm:fb-type-safety}
-$\forall e, \tau, \Delta, \mem$, such that $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
+$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
+and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
 \cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
 exception ($\redstoo{\Delta}{(e,\mem)}{\invalidexn}$), or there exists an
-$e'$ and a $\mem'$ such that $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$
-and $\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
+$e'$ and a $\mem'$ such that $\consistent(\Delta,\mem')$ and
+$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$ and
+$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
 \end{theorem}
+The relation $\consistent$ relates $\Delta$ and $\mem$ only if both
+make consistent assumptions about liveness of regions. Since $\Delta$
+is consistent with both $\mem$ and $\mem'$, the theorem also captures
+the key property of operational semantics that no live region is ever
+freed or transferred.
 

### 2016-10-22 paper/type-inference.tex
Removed claims of completeness.

@@
 Let $\consOf{q}$ denote the set of constraints generated from an elaborated program $q$.
-The following theorem states that constraint-generation is sound and complete:
+The following theorem states that constraint-generation is sound:
 \begin{theorem}
@@
 An assignment $\soln$ (for the region and predicate variables in $q$)
-satisfies $C$ iff $q[\soln]$ is well-typed.
+satisfies $C$ only if $q[\soln]$ is well-typed.
 \end{theorem}
@@
 
-Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
-show that the second and third steps of the type-inference algorithm are both
-sound and complete.
-It is easy to verify the soundness of the first step (the elaboration phase):
-For any program $p \in \absof{\FB}$, the elaboration phase produces a $q$
-such that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
+Theorem~\ref{thm:constraint-generation-sc}
+and~\ref{thm:constraint-solver-sc} show that the second and third
+steps of the type-inference algorithm are sound. It is easy to verify
+the soundness of the first step (the elaboration phase): For any
+program $p \in \absof{\FB}$, the elaboration phase produces a $q$ such
+that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
 %
 The soundness of the type inference algorithm follows.
-The only sources of incompleteness in the type inference algorithm are some heuristic
-choices made in the first step, as explained below.
+Theorem~\ref{thm:constraint-solver-sc} also establishes the
+completeness of constraint solving.
+
+The main source of incompleteness in the type inference algorithm is
+the set of heuristic choices made during region parameterization step,
+as explained below.
 
@@
 makes a heuristic choice, the elaboration procedure can use the user-provided choices instead.
-This can help the type-inference overcome these sources of incompleteness.
+This can help the type-inference overcome these limitations.
 

### 2016-10-22 paper/typesystem-proofs.tex
Type safety proofs added.

\section{Type System: Proofs}

We first define the $\consistent$ relation between $\Delta$ and
$\mem$:

\begin{definition}[\consistent($\Delta$,$\mem$)]
A set $\Delta \in 2^{\rgn}$ of region annotations is said to be
consistent with a map $\mem \in \loc \rightarrow s$ from region
locations to typestates if and only if forall $\loc\in\Delta$,
$\mem(\loc) = \LIVE$.
\end{definition}

Consistency between $\Delta$ and $\mem$ is preserved by the reduction
relation:

\begin{lemma}[\textbf{consistency preservation}]
\label{lem:consistency}
$\forall(e,\Delta,\mem,\phicx,\tau)$. if $\consistent(\Delta,\mem)$
and $\tywf{\Delta}{\phicx}$ and
$\hastyp{(\Delta,\cdot,\phicx),\cdot}{e}{\tau}$ and
$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then
$\consistent(\Delta,\mem')$.
\end{lemma}
\begin{proof}
Proof is by induction on $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$.
For the rules where $\mem'=\mem$, proof is trivial. For the rules
where $\mem'$ is a result of executing a subexpression under
$\Delta$, proof follows from the inductive hypothesis. Remaining rules
are discussed below:
% \begin{smathpar}
% \begin{array}{cl}
%   \forall(\Delta,\mem,\phicx,\tau).~\consistent(\Delta,\mem)
%   ~\conj ~\tywf{\Delta}{\phicx}
%   ~\conj ~\hastyp{(\Delta,\cdot,\phicx),\cdot}{e_0}{\tau} & \\
%   \hspace*{1in}
%   ~\redstoo{\Delta}{(e_0,\mem)}{(e_0',\mem')}
%   ~\Rightarrow~ \consistent(\Delta,\mem') & IH\\
% \end{array}
% \end{smathpar}
\begin{itemize}
  \item Rule $\rulelabel{LetRegionBegin}$: All live regions in $\mem$
  are also live in $\mem'$. Proof follows
  \item Rule $\rulelabel{LetRegion}$: By inductive hypothesis, $\mem'$
  is consistent with $\Delta \cup \{\loc\}$. Hence, it is consistent
  with $\Delta$.
  \item Rule $\rulelabel{LetRegionEnd}$: Here, $\loc$ is set to
  $\XFERRED$ in $\mem'$. However, since $e$ is the \C{letd}
  expression, and the type rule for the \C{letd} expression gives us
  $\loc \not\in \Delta$. Hence $\Delta$ is consisten with $\mem'$.
  \item Rule $\rulelabel{NewRegion}$: $\mem'$ extends $\mem$ with a
  new binding. Consistency is trivially preserved.
  \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\mem(\loc)=\USED$
  and $\consistent(\Delta,\mem)$, $\loc \not\in \Delta$. The proof
  follows from inductive hypothesis.
  \item Rule $\rulelabel{Open}$: As $\mem'=\mem[\loc\mapsto\LIVE]$,
  hence $\Delta$ remains consistent with $\mem'$. Here, we also take
  note of the fact that if the result expression (\C{opened}) is
  tagged with a typestate of $\USED$, then $\loc\notin \Delta$.
  \item Rule $\rulelabel{Opened}$: Inductive hypothesis guarantees
  that $\Delta\cup\{\loc\}$ is consistent with $\mem'$. Hence,
  $\Delta$ is consistent with $\mem'$. We also take note of the fact
  that the typestate tagged with the $\C{opened}$ expression remains
  invariant during reduction.
  \item Rule $\rulelabel{OpenEnd}$: $\mem'=\mem[\loc\mapsto s]$, where
  $s$ is the typestate tagged with the \C{opened} expression. As clear
  from the $\rulelabel{Open}$ rule, $s$ can be either $\USED$ or
  $\LIVE$, and if $s$ is $\USED$ then $\loc\notin\Delta$. Hence, in
  either case $\Delta$ remains consistent with $\mem$.
  \item Rules $\rulelabel{Transfer}$: changes binding for a non-live
  location $\loc$. Hence, consistency is preserved.
\end{itemize} 
\qed
\end{proof}

\begin{lemma}[value substitution preserves typing]
\label{lem:substitution}
$\forall(e,z,\tau_1,\tau_2,\Delta,\phicx)$, if $\hastyp{(\Delta,\cdot,
\phicx),\cdot[x\mapsto\tau_1]}{e}{\tau_2}$ and $\hastyp{(\Delta,\cdot,
\phicx),\cdot}{v}{\tau_1}$, then $\hastyp{(\Delta,\cdot,
\phicx),\cdot} {[v/x]e}{\tau_2}$.
\end{lemma}
\begin{proof}
The proof is by induction on typing derivation and follows on the
lines of similar proof for FGJ.
\qed
\end{proof}

% \begin{lemma}[weakening]
% $\forall(v,\tau,\Delta,\Delta_0,\phicx,\phicx_0)$, 
% if 
% $\hastyp{(\Delta \cup \Delta_0,\cdot, \phicx \conj
% \phicx_0),\cdot}{v}{\tau}$
% and
% $\tywf{(\Delta,\cdot,\phicx)}{\tau}$
% then 
% $\hastyp{(\Delta,\cdot, \phicx),\cdot} {v}{\tau}$.
% \end{lemma}
% \begin{proof}
% The proof is by induction on $\hastyp{(\Delta \cup \Delta_0,\cdot, \phicx \conj
% \phicx_0),\cdot}{v}{\tau}$. Since $v$ is a value, we have few cases:

% \begin{itemize}
%   \item Case ($v = \C{new}\; \fbN_0(\vbar)$ and $\tau = \fbN_0$): By
%   inverting 
% \end{itemize}

% \end{proof}

\begin{lemma}[progress]
\label{lem:progress}
$\forall e, \tau, \mem, \rhomap, \phicx$, if $\consistent(\Delta,\mem)$ and 
$\tywf{\Delta}{\phicx}$ and
$\hastyp{\emptyA,\cdot}{e}{\tau}$, then one of the following holds:\\
  \begin{smathpar}
  \begin{array}{rl}
    (i) & \exists (e',\rhomap').\;\redstoo{\Delta}{(e,\rhomap)}{(e',\rhomap')}\\
    (ii) & \valuee(e)\\
    (iii) & \redstoo{\Delta}{(e,\rhomap)}{\invalidexn}\\
  \end{array}
  \end{smathpar}
\end{lemma}
\begin{proof}
Proof is by induction on the typing derivation of $e:\tau$. Most cases
follow from the inductive hypothesis (IH), which claims that if a
subexpression has a typing derivation, then it can make progress. We
will consider cases where all subexpressions are values, but the
expression itself is not a value.
\begin{itemize}
  \item $B\inang{\tbar}{\locbar}(\vbar).f_i$ case: This expression has
  a type only if $\tywf{\emptyA}{B\inang{\tbar}\inang{\locbar}}$, which is
  possible only if $\locbar \in \Delta$. Hence $e$ can make progress.

  \item $\C{let} \; x = v \; \C{in} \; e$ case: Always takes step via
  \rulelabel{LetExp}.

  \item $\C{new}\; \RgnZT{\toprgn}(v)$ case: takes a step by
  \rulelabel{NewRegion} rule to $\C{new}\; \RgnZT{\toploc\loc}
  (v\inang{\loc}())$.

  \item Method call case: Since $\mtype$ is defined, $\mbody$ is also
  defined, and the execution takes a step by $\rulelabel{MethodInv}$.

  \item Lambda application case: takes step by \rulelabel{FnApply}

  \item \C{letregion} case: takes a step by \rulelabel{LetRegionBegin}
  rule.

  \item \C{open} case: takes a step by \rulee{Open} rule, or throws an
  exception by \rulee{OpenTransferred} rule, depending on the typestate
  of the region location.

  \item \C{letd} case: The subexpression is typed under $\Delta \cup
  \{\loc\}$, hence the subexpression takes a step under $\Delta \cup
  \{\loc\}$. This allows \C{letd} expression take a step via
  \rulee{LetRegion} rule. If subexpression is a value, then \C{letd}
  takes a step via \rulee{LetRegionEnd} rule.

  \item \C{opened} case: similar to \C{letd} case.
\end{itemize}
\qed
\end{proof}

\begin{lemma}[preservation]
\label{lem:preservation}
$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
\cdot}{e}{\tau}$, and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then 
$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
\end{lemma}
\begin{proof}
  Proof is by induction on the reduction step. Most cases follow from
  the inductive hypothesis, which asserts that if a subexpression
  takes a step, it preserves its type under the same $\Delta$ and
  $\phicx$. We consider interesting cases below:
  \begin{itemize}
    \item \rulee{FieldAccess} case: $\C{new}\; A\inang{\tbar}
    \inang{\loc\locbar}(\vbar).f_i$ takes a step to $v_i$. The field
    access expression has a type of $i^{th}$ field returned by
    $\fields$ definition, and so does $v_i$, if ${new}\;
    A\inang{\tbar} \inang{\loc\locbar}(\vbar)$ has to be typable.
    Hence the type is preserved.

    \item \rulee{MethodInv} case: From the method invocation type rule
    and method well-formedness condition, method body ($e'$) has a type
    $\tau^2$ under $(\{\loc,\locbar,\rhobar\},\cdot,\phi),\cdot[\xbar
    \mapsto \overline{\tau^1}]$. Also,
    $\tywf{\loc,\locbar,\rhobar}{\phi}$. Since $\overline{\loc'} \neq \rhobar$
    (locations are never equal to region variables), we have:
    \begin{center}
    $\hastyp{(\{\loc,\locbar,\overline{\loc'}\},\cdot,[\overline{\loc'}/\rhobar]\phi),\cdot[\xbar
    \mapsto \overline{\tau^1}]}{e'}{\tau^2}$ and
    $\tywf{\{\loc,\locbar,\overline{\loc'}\}}{[\overline{\loc'}/\rhobar]\phi}$
    \end{center}
    Since $\{\{\loc,\locbar,\overline{\loc'}\} \subseteq \Delta$, and
    $\phicx \vdash [\overline{\loc'}/\rhobar]\phi$, we can strengthen
    the context and derive:
    \begin{center}
      $\hastyp{(\Delta,\cdot,\phicx),\cdot[\xbar \mapsto \overline{\tau^1}]}{e'}{\tau^2}$
    \end{center}
    Applying the substitution lemma (Lemma~\ref{lem:substitution}) and
    inductive hypothesis gives the proof.

    \item \rulee{FnApply} case: proceeds on the similar lines as
    \rulee{MethodInv} case, except that no strengthening is needed;
    function body is typed under a context that includes $\Delta$.

    \item \rulee{LetRegionBegin} case: Since $\loc \notin dom(\mem)$,
    it follows that $\loc \notin \Delta$. Rest of the premises
    required to apply the \C{letd} type rule on result expression are
    obtained from the \C{letregion} type rule of the initial
    expression ($e$), by substituing $\loc$ for $\pi$.

    \item \rulee{LetRegionEnd} case: $e$ is a \C{letd} expression
    which reduces to a value $v$. Invering the typing derivation of
    $e$ yeilds the premise that $v$ is well-typed under the context
    $(\Delta \cup \{\loc\},\cdot,\phicx \conj \Delta \outlives \loc)$.
    However, we have to prove that $v$ is well-typed under the smaller
    context $(\Delta,\cdot,\phicx)$. We carry out this proof by
    induction on the structure of value $v$:
    \begin{itemize}
      \item If $v$ is an object value (e.g.,
      $B\inang{\tbar}\inang{\locbar}(\vbar)$), IH yeilds the
      well-typedness of $\vbar$ under $(\Delta,\cdot,\phicx)$, and the
      premise 
      $\tywf{(\Delta,\cdot,\phicx)} {B\inang{\tbar}\inang{\locbar}}$,
      obtained by inverting the type judgment for \C{letd}, yeilds the
      well-typedness of whole value.

      \item A region handler value ($\RgnZT{\toploc\loc'}(v)$) is
      well-typed under any context.

      \item If $v$ is a function closure, then it is well-tuped under
      $(\Delta,\cdot,\phicx)$ only if it doesn't trap any references
      to $\loc$, the \C{letd} location. Since closure's type needs to
      be well-formed under $(\Delta,\cdot,\phicx)$, its allocation
      region belongs to $\Delta$, hence outlives $\loc$. Since the
      closure typing rule requires all free region variables of the
      closure body outlive the allocation region of closure, it
      follows that all free region variables strictly outlive $\loc$,
      hence they belong to $\Delta$. Hence the function closure is
      well-typed outside \C{letd}.
    \end{itemize}

  \item \rulee{OpenEnd} case: The proof is similar to the \C{letd}
  case.
  \qed
  \end{itemize}
\end{proof}

\begin{proof}[\textbf{Theorem~\ref{thm:fb-type-safety}}]
Follows from Lemmas~\ref{lem:consistency},~\ref{lem:progress},
and~\ref{lem:preservation}.
\qed
\end{proof}

### 2016-11-29 paper/abstract.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
 dangling pointers. A key contribution of this paper is a
-refinement-based region type system, for a C\#-like typed object-oriented
-language with higher-order functions and generics, that ensures the
-memory safety of programs in the presence of transferable regions.
-We complement our type system with a type inference algorithm, which eliminates the need for programmers to write region annotations on types.
+refinement-based region type system, for a C\#-like typed
+object-oriented language with higher-order functions and generics,
+that ensures the memory safety of programs in the presence of
+transferable regions.  We complement our type system with a type
+inference algorithm, which eliminates the need for programmers to
+write region annotations on types.
 % The cornerstone of our inference algorithm is a novel constraint solver that
@@
 % to recursive constraints.
-We establish soundness and partial completeness results for the type inference algorithm: in particular, the key steps of constraint-generation and constraint-solving in the type inference algorithm are complete.
-We describe a prototype implementation of our inference algorithm and the constraint solver, and our experience of using it to enforce memory safety in dataflow programs.
+The inference algorithm has been proven sound, and the constraint
+solving algorithm used by the inference has been proven complete.  We
+describe a prototype implementation of the inference algorithm and the
+solver, and our experience of using it to enforce memory safety in
+dataflow programs.
 

### 2016-11-29 paper/introduction.tex
Removed claims of completeness.
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
 Our inference algorithm proceeds in three stages: in the first stage,
-we elaborate the program by introducing region parameters (for classes and methods);
-in the second stage we generate a set of constraints that must be satisfied for
-the program to type check; in the third stage, we solve the set of constraints,
-inferring the preconditions of all classes and methods (in terms of the
-expected outlives-constraints between the region parameters).
-We show that the second and third stages are sound as well as
-complete (while the first stage is sound).
+we elaborate the program by introducing region parameters (for classes
+and methods); in the second stage we generate a set of constraints
+that must be satisfied for the program to type check; in the third
+stage, we solve the set of constraints, inferring the preconditions of
+all classes and methods (in terms of the expected outlives-constraints
+between the region parameters).  We show that the algorithm is sound,
+and the last stage of constraint solving is complete.
 % (using an interprocedural fixed point computation).
@@
 
-\item We establish partial completeness results for the type inference
-  algorithm, as well as soundness.
+\item We establish soundness of type inference algorithm, and
+completeness of constraint solving.
 

### 2016-11-29 paper/type-inference.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
 
-We later on establish the soundness of the type inference algorithm, and also establish
-that the constraint-generation and constraint-solving steps (the second and third steps
-above) are complete.
+We later on establish the soundness of the type inference algorithm,
+and the completeness of constraint-solving step.
 
@@
 (\ie, constraints without any region variable or predicate variable)
-in $\saturate{C}$. Let $\rhoC$ denote the subset of all wellformedness
+in $\saturate{C}$. Let $\rhoC$ denote the subset of all well-formedness
 constraints for region variables in $\satC$.

### 2016-11-29 paper/type-system.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
 class Pair<a $\extends$ Object, b $\extends$ Object>
-          <$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a \conj \rho_2 \outlives \rho^a$> $\extends$ Object<$\rho^a$> {
+          <$\rho_0, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
+          \rho_2 \outlives \rho_0$> $\extends$ Object<$\rho_0$> {
   a@$\rho_1$ fst; 
@@
 % $\rho^a, \rho_1, \rho_2$.
-We refer to the first region parameter, usually denoted $\rho^a$, as
-the \emph{allocation region} of the class: it serves to identify the
-region where an instance of the class is allocated.
+We refer to the first region parameter ($\rho_0$ in the above example)
+as the \emph{allocation region} of the class: it serves to identify
+the region where an instance of the class is allocated\footnote{In
+general, $\rhobar$ denotes the sequence $\rho_1\rho_2...$ of region
+parameters of a class or a method. In some cases, we also represent
+region parameters as $\rho\rhobar$ to clearly distinguish between
+the allocation region parameter ($\rho$) and the rest.}.
 %
-An object in \FB can contain fields referring to objects allocated in regions ($\rhobar$) other than its
-own allocation region ($\rhoalloc$), provided that the former outlive
-the latter (i.e., $\rhobar \outlives \rhoalloc$). In such case, the
-definition of object's class needs to be parametric over allocation
-regions of its fields (i.e., their classes). Furthermore, the
-constraint that such regions must outlive the allocation region of the
-class needs to be made explicit in the definition, as the \C{Pair}
-class does in the above definition. We say that the \C{Pair} class
-exhibits \emph{constrained region polymorphism}.
+An object in \FB can contain fields referring to objects allocated in
+regions ($\rhobar$) other than its own allocation region ($\rho$),
+provided that the former outlive the latter (i.e., $\rhobar \outlives
+\rho$). In such case, the definition of object's class needs to be
+parametric over allocation regions of its fields (i.e., their
+classes). Furthermore, the constraint that such regions must outlive
+the allocation region of the class needs to be made explicit in the
+definition, as the \C{Pair} class does in the above definition. We say
+that the \C{Pair} class exhibits \emph{constrained region
+polymorphism}.
 %
@@
 \begin{codejava}
+
 letregion $\pi_0$ in
@@
 %\vspace*{-0.1in}
-In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
+In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
 $\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
@@
 (\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
-parameterization seperate from region parameterization allows \FB to
+parameterization separate from region parameterization allows \FB to
 type such examples. This choice also simplifies type inference in the
@@
 In addition, the context for the expression typing judgment also
-includes a type environment ($\env \in x \mapsto \tau$) that contains
-the type bindings for variables in scope. Like the judgments in
-FGJ~\cite{fgj}, all the judgments defined by the rules in
-Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a class
-table ($CT \in cn \mapsto D$) that maps class names to their
-definitions in \FB.
+includes (a). a type environment ($\env \in x \mapsto \tau$) that
+contains the type bindings for variables in scope, and (b). The region
+($\rgn$) that serves as the allocation context for the expression
+being type checked. Like the judgments in FGJ~\cite{fgj}, all the
+judgments defined by the rules in Fig.~\ref{fig:fb-staticsem} are
+implicitly parameterized on a class table ($CT \in cn \mapsto D$) that
+maps class names to their definitions in \FB.
 
@@
 identifier introduced by the expression be unique under the current
-context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
-order to prevent the new region from incorrectly assuming existing
-outlives relationships on an eponymous region. Provided this is
-satisfied, the expression ($e$) under \C{letregion} is then
-typechecked assuming that the new region is live ($\rgn \in \rhoenv$)
-and that it is outlived by all existing live regions ($\rhoenv
-\outlives \rgn$). The result of a \C{letregion} expression must have a
-type that is well-formed under a context not containing the new
-region. This ensures that the value obtained by evaluating a
-\C{letregion} expression contains no references to the temporary
-objects inside the region.
+context (i.e., $\pi \notin \Delta$). This condition is needed in order
+to prevent the new region from incorrectly assuming existing outlives
+relationships on an eponymous region. Provided this is satisfied, the
+expression ($e$) under \C{letregion} is then typechecked assuming that
+the new region is live ($\Delta \cup \{\pi\}$) and that it is outlived
+by all existing live regions ($\Delta \outlives \pi$). The result of a
+\C{letregion} expression must have a type that is well-formed under a
+context not containing the new region. This ensures that the value
+obtained by evaluating a \C{letregion} expression contains no
+references to the temporary objects inside the region.
 
@@
 
-The type rule for lambda expression typechecks
-% requires that its region
-% parameters ($\rhoalloc\rhobar$) be unique under the current context,
-% and that the constraints ($\phi$) over region parameters be
-% well-formed. 
-the lambda-bound expression ($e$) under an extended type environment
-containing bindings for function's arguments, assuming that region
-parameters are live, and that declared constraints over region
-parameters hold. The constraints ($\phi$) are required to be
-well-formed under $\rhoenv$ extended with the function's region
-parameters ($\rhobar$). The premise $\forall
-(\rgn'\in\frv(e)\setminus\{\rhobar\}). ~\isvalid{\phicx}
-{\rgn'\outlives\rgn}$ ensures the safety of dereferences inside the
-closure by requring that all free region variables ($\frv$) trapped by
-the closure outlive the allocation region of the closure.
+The type rule for lambda expression typechecks the lambda-bound
+expression ($e$) under an extended type environment containing
+bindings for function's arguments, assuming that region parameters are
+live, and that declared constraints over region parameters hold. The
+constraints ($\phi$) are required to be well-formed under $\Delta$
+extended with the function's region parameters ($\rho\rhobar$). Note
+that the closure is always allocated in the current allocation context
+($\rgn$). This prevents the closure from escaping the context in which
+it is created, thus trivially ensuring the safety of any dereferences
+inside the closure.
 
@@
 defines a small-step operational semantics for \fbname via a
-five-place reduction relation:
+four-place reduction relation:
 % \vspace*{-0.15in}
@@
 locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
-relation should be read as following: given a set ($\rhoenv$) of
-regions that are currently live, and an initial typestate map
-($\mem$), the expression $e$ reduces to $e'$, while updating $\mem$ to
-$\mem'$. The semantics gets ``stuck'' if $e$ attempts to access an
-object whose allocation region is not present in $\rhoenv$, or if $e$
-tries to \C{open} a transferable region that is not mapped to an
-appropriate typestate by $\mem$.  On the other hand, if $e$ attempts
-to commit an operation on a $\RgnZ$ object that is not sanctioned by
-the transition discipline in Fig.~\ref{fig:region-fsm}, then it raises
-an exception value ($\invalidexn$).
+relation relates an expression ($e$) and a typestate map ($\mem$) to a
+reduced expression ($e'$) and an updated typestate map ($\mem'$). The
+semantics gets ``stuck'' if $e$ attempts to access an object whose
+allocation region is not present in $\rhoenv$, or if $e$ tries to
+\C{open} a transferable region that is not mapped to an appropriate
+typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
+operation on a $\RgnZ$ object that is not sanctioned by the transition
+discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
+value ($\invalidexn$).
 

### 2016-11-29 paper/typeinference-proofs.tex
minor fixes
Removed claims of completeness.
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
-\newcommand{\deltaPC}{\Delta_P^C}
+\section{Type Inference: Proofs}
 
-\section{Type Inference: Proofs}
+Figs.~\ref{fig:constraint-gen-1} and~\ref{fig:constraint-gen-2} contain the
+extra constraint-generation rules not covered in Fig.~\ref{fig:constraint-gen-0}.
+We now present a sketch of the proofs of the soundness of type
+inference algorithm, and completeness of the constraint solver.
 
@@
 tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
-We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
+We can show that $\zeta_2$ is a valid derivation tree if $\soln$ satisfies $C$.
 (We make use of analogous lemmas for type well-formedness rules as well.)

### 2016-11-29 paper/typesystem-proofs.tex
Re-added allocation context to type judgment. Restores completeness of type inference.

@@
 \label{lem:consistency}
-$\forall(e,\Delta,\mem,\phicx,\tau)$. if $\consistent(\Delta,\mem)$
-and $\tywf{\Delta}{\phicx}$ and
-$\hastyp{(\Delta,\cdot,\phicx),\cdot}{e}{\tau}$ and
-$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then
+$\forall(e,\Delta,\mem,\phicx,\rgn,\tau)$. if $\consistent(\Delta,\mem)$
+and $\tywf{\Delta}{\phicx}$ %and $\rgn \in \Delta$ 
+and $\hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{e}{\tau}$ 
+and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then
 $\consistent(\Delta,\mem')$.
@@
   \item Rule $\rulelabel{LetRegion}$: By inductive hypothesis, $\mem'$
-  is consistent with $\Delta \cup \{\loc\}$. Hence, it is consistent
-  with $\Delta$.
+  is consistent with $\Delta$. Hence, it is consistent
+% is consistent with $\Delta \cup \{\loc\}$. Hence, it is consistent
+% with $\Delta$.
   \item Rule $\rulelabel{LetRegionEnd}$: Here, $\loc$ is set to
@@
   expression, and the type rule for the \C{letd} expression gives us
-  $\loc \not\in \Delta$. Hence $\Delta$ is consisten with $\mem'$.
+  $\loc \not\in \Delta$. Hence $\Delta$ is consistent with $\mem'$.
   \item Rule $\rulelabel{NewRegion}$: $\mem'$ extends $\mem$ with a
   new binding. Consistency is trivially preserved.
-  \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\mem(\loc)=\USED$
-  and $\consistent(\Delta,\mem)$, $\loc \not\in \Delta$. The proof
-  follows from inductive hypothesis.
+% \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\mem(\loc)=\USED$
+% and $\consistent(\Delta,\mem)$, $\loc \not\in \Delta$. The proof
+% follows from inductive hypothesis.
+  \item Rule \#2 of $\rulelabel{NewRegion}$ : Since $\consistent(\Delta,
+  \mem)$, we have $\consistent(\Delta \cup \{\loc\}, \mem[\loc\mapsto 
+  \LIVE])$. By inductive hypothesis, $\consistent(\Delta \cup \{\loc\}, 
+  \mem')$. Hence, $\consistent(\Delta, \mem')$.
   \item Rule $\rulelabel{Open}$: As $\mem'=\mem[\loc\mapsto\LIVE]$,
@@
   \item Rule $\rulelabel{Opened}$: Inductive hypothesis guarantees
-  that $\Delta\cup\{\loc\}$ is consistent with $\mem'$. Hence,
-  $\Delta$ is consistent with $\mem'$. We also take note of the fact
-  that the typestate tagged with the $\C{opened}$ expression remains
-  invariant during reduction.
+  that $\Delta$ is consistent with $\mem'$.
+% that $\Delta\cup\{\loc\}$ is consistent with $\mem'$. Hence,
+% $\Delta$ is consistent with $\mem'$. 
+  We also take note of the fact that the typestate tagged with the
+  $\C{opened}$ expression remains invariant during reduction.
   \item Rule $\rulelabel{OpenEnd}$: $\mem'=\mem[\loc\mapsto s]$, where
@@
   from the $\rulelabel{Open}$ rule, $s$ can be either $\USED$ or
-  $\LIVE$, and if $s$ is $\USED$ then $\loc\notin\Delta$. Hence, in
-  either case $\Delta$ remains consistent with $\mem$.
+  $\LIVE$. The type rule for \C{opened} tells us that if $s$ is
+  $\USED$ then $\loc\notin\Delta$. Hence, in either case $\Delta$
+  remains consistent with $\mem$.
   \item Rules $\rulelabel{Transfer}$: changes binding for a non-live
@@
 
-% \begin{lemma}[weakening]
-% $\forall(v,\tau,\Delta,\Delta_0,\phicx,\phicx_0)$, 
-% if 
-% $\hastyp{(\Delta \cup \Delta_0,\cdot, \phicx \conj
-% \phicx_0),\cdot}{v}{\tau}$
-% and
-% $\tywf{(\Delta,\cdot,\phicx)}{\tau}$
-% then 
-% $\hastyp{(\Delta,\cdot, \phicx),\cdot} {v}{\tau}$.
-% \end{lemma}
-% \begin{proof}
-% The proof is by induction on $\hastyp{(\Delta \cup \Delta_0,\cdot, \phicx \conj
-% \phicx_0),\cdot}{v}{\tau}$. Since $v$ is a value, we have few cases:
-
-% \begin{itemize}
-%   \item Case ($v = \C{new}\; \fbN_0(\vbar)$ and $\tau = \fbN_0$): By
-%   inverting 
-% \end{itemize}
-
-% \end{proof}
-
 \begin{lemma}[progress]
 \label{lem:progress}
-$\forall e, \tau, \mem, \rhomap, \phicx$, if $\consistent(\Delta,\mem)$ and 
+$\forall e, \tau, \mem, \rhomap, \phicx, \rgn$, if $\consistent(\Delta,\mem)$ and 
 $\tywf{\Delta}{\phicx}$ and
-$\hastyp{\emptyA,\cdot}{e}{\tau}$, then one of the following holds:\\
+$\hastyp{\emptyA,\cdot, \rgn}{e}{\tau}$, then one of the following holds:\\
   \begin{smathpar}
@@
 \begin{itemize}
-  \item $B\inang{\tbar}{\locbar}(\vbar).f_i$ case: This expression has
+  \item $B\inang{\tbar}\inang{\locbar}(\vbar).f_i$ case: This expression has
   a type only if $\tywf{\emptyA}{B\inang{\tbar}\inang{\locbar}}$, which is
-  possible only if $\locbar \in \Delta$. Hence $e$ can make progress.
+  possible only if $\locbar \in \Delta$. Since $\Delta$ is consistent
+  with $\mem$, we have $\mem(\locbar)=\LIVE$. Hence $e$ can make progress.
 
@@
 
-  \item Method call case: Since $\mtype$ is defined, $\mbody$ is also
-  defined, and the execution takes a step by $\rulelabel{MethodInv}$.
+  \item $\C{new}\; \RgnZT{\toploc\loc}(e)$ case: takes a step by
+  \rulelabel{NewRegion} rule \#2 to $\C{new}\; \RgnZT{\toploc\loc}(e')$.
 
-  \item Lambda application case: takes step by \rulelabel{FnApply}
+  \item Method call case
+  ($v_0.m\inang{\loc'\overline{\loc'}}(\overline{v'})$): From the type
+  rule, since $\bound$ is defined for the type of $v_0$, we infer that
+  the type is of form $A\inang{\tbar}\inang{\loc\locbar}$ and $v_0 =
+  A\inang{\tbar}\inang{\loc\locbar}(\vbar)$.  Since $\mtype$ is
+  defined, $\mbody$ is also defined. Well-formedness requirement on
+  $A\inang{\tbar}\inang{\loc\locbar}$ gives $\loc \in \Delta$. Since
+  $\consistent(\Delta,\mem)$, we have $\mem(\loc)=\LIVE$. Hence the
+  execution takes a step by $\rulelabel{MethodInv}$.
+
+  \item Lambda application case ($e = (\lambdaexp{\loc'}{\rho\rhobar \,|\, 
+  \phi}{\xbar:\bar{\tau^1}}{e_0})\inang{\rgn\rbar}(\vbar)$): The type
+  rule gives $\loc',\rgn,\rbar \in \Delta$. Since
+  $\consistent(\Delta,\mem)$ and $dom(\mem)$ only contains locations,
+  we have $\rgn=\loc$ and $\rbar=\locbar$, for some locations $\loc,
+  \locbar$. Since $\loc' \in \Delta$, $\consistent(\Delta,\mem)$ also
+  gives $\mem(\loc')=\LIVE$. Hence, $e$ takes step by
+  \rulelabel{FnApply}.
 
@@
 
-  \item \C{letd} case: The subexpression is typed under $\Delta \cup
-  \{\loc\}$, hence the subexpression takes a step under $\Delta \cup
-  \{\loc\}$. This allows \C{letd} expression take a step via
-  \rulee{LetRegion} rule. If subexpression is a value, then \C{letd}
-  takes a step via \rulee{LetRegionEnd} rule.
+  \item \C{letd} case: takes a step via \rulee{LetRegion} rule.  If
+  subexpression is a value, then \C{letd} takes a step via
+  \rulee{LetRegionEnd} rule.
 
-  \item \C{opened} case: similar to \C{letd} case.
+  \item \C{opened} case: takes a step via \rulee{Opened} rule.  If
+  subexpression is a value, then \C{letd} takes a step via
+  \rulee{OpenEnd} rule.
+
+  \item \C{transfer} case: Either takes a step via \rulee{Transfer}
+  rule, or throws an exception via \rulee{TransferOpened}, depending
+  on the typestate of the region.
+
+  \item Subtype case: The premise is that
+  $\hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{e}{\tau'}$ and 
+  $\subtyp{(\Delta,\cdot,\phicx)}{\tau'}{\tau}$. Inductive hypothesis
+  from $\hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{e}{\tau'}$ directly
+  gives the proof.
+
 \end{itemize}
@@
 \label{lem:preservation}
-$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
+$\forall e, \tau, \Delta, \mem, \rgn$, such that $\consistent(\Delta,\mem)$
 and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
-\cdot}{e}{\tau}$, and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then 
-$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
+\cdot,\rgn}{e}{\tau}$, and $\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$, then 
+$\hastyp{\emptyASigp,\cdot,\rgn}{e'}{\tau}$.
 \end{lemma}
@@
     access expression has a type of $i^{th}$ field returned by
-    $\fields$ definition, and so does $v_i$, if ${new}\;
-    A\inang{\tbar} \inang{\loc\locbar}(\vbar)$ has to be typable.
+    $\fields$ definition, and so does $v_i$, if $\C{new}\;
+    A\inang{\tbar} \inang{\loc\locbar}(\vbar)$ has to be typeable.
     Hence the type is preserved.
 
-    \item \rulee{MethodInv} case: From the method invocation type rule
-    and method well-formedness condition, method body ($e'$) has a type
-    $\tau^2$ under $(\{\loc,\locbar,\rhobar\},\cdot,\phi),\cdot[\xbar
-    \mapsto \overline{\tau^1}]$. Also,
-    $\tywf{\loc,\locbar,\rhobar}{\phi}$. Since $\overline{\loc'} \neq \rhobar$
-    (locations are never equal to region variables), we have:
+    \item \rulee{MethodInv} case ($e = (\C{new}\; A\inang{\tbar}
+    \inang{\loc\locbar}(\vbar)).m\inang{\loc'\overline{\loc'}}
+    (\overline{v'})$): Let $\mtype(A\inang{\tbar}\inang{\loc\locbar}) 
+    \,=\, \inang{\rho\rhobar \,|\, \phi}\overline{\tau^1}\rightarrow 
+    \tau^2$. Hence $\tau = \tau^2$. Method invocation type rule tells
+    us $r = \loc'$. Further, from the method well-formedness
+    condition, method body ($e_0$) has a type $\tau^2$ under
+    $(\{\loc,\locbar,\rho,\rhobar\},\cdot,\phi),\cdot[\xbar \mapsto
+    \overline{\tau^1}], \rho$. Also,
+    $\tywf{\loc,\locbar,\rho,\rhobar}{\phi}$. Since $\loc' \neq \rho$
+    and $\overline{\loc'} \neq \rhobar$ (locations are never equal to
+    region variables), we have:
     \begin{center}
-    $\hastyp{(\{\loc,\locbar,\overline{\loc'}\},\cdot,[\overline{\loc'}/\rhobar]\phi),\cdot[\xbar
-    \mapsto \overline{\tau^1}]}{e'}{\tau^2}$ and
-    $\tywf{\{\loc,\locbar,\overline{\loc'}\}}{[\overline{\loc'}/\rhobar]\phi}$
+    $\hastyp{(\{\loc,\locbar,\loc',\overline{\loc'}\},\cdot,
+      [\loc'\overline{\loc'}/\rho\rhobar]\phi),\cdot[\xbar
+      \mapsto \overline{\tau^1}],\loc'}{e_0}{\tau^2}$ and
+    $\tywf{\{\loc,\locbar,\loc',\overline{\loc'}\}}
+          {[\loc'\overline{\loc'}/\rho\rhobar]\phi}$
     \end{center}
-    Since $\{\{\loc,\locbar,\overline{\loc'}\} \subseteq \Delta$, and
-    $\phicx \vdash [\overline{\loc'}/\rhobar]\phi$, we can strengthen
+    Since $\{\{\loc,\locbar,\loc',\overline{\loc'}\} \subseteq \Delta$, and
+    $\phicx \vdash [\loc'\overline{\loc'}/\rho\rhobar]\phi$, we can strengthen
     the context and derive:
     \begin{center}
-      $\hastyp{(\Delta,\cdot,\phicx),\cdot[\xbar \mapsto \overline{\tau^1}]}{e'}{\tau^2}$
+      $\hastyp{(\Delta,\cdot,\phicx),\cdot[\xbar \mapsto
+      \overline{\tau^1}], \loc'}{e_0}{\tau^2}$
     \end{center}
-    Applying the substitution lemma (Lemma~\ref{lem:substitution}) and
-    inductive hypothesis gives the proof.
+    Applying the substitution lemma (Lemma~\ref{lem:substitution})
+    gives:
+    \begin{center}
+      $\hastyp{(\Delta,\cdot,\phicx),\cdot,
+        \loc'}{[\overline{v'}/\xbar]e_0}{\tau^2}$
+    \end{center}
+    Since $e' \,=\, [\overline{v'}/\xbar]e_0$ and $\tau = \tau_2$, we
+    have the required:
+    \begin{center}
+      $\hastyp{(\Delta,\cdot,\phicx),\cdot, \loc'}{e'}{\tau}$
+    \end{center}
+    It is important to note that if $\rho$, the allocation context
+    parameter of $m$, was not instantiated with the current allocation
+    context $\loc'$, the proof would have failed.
+    \item \rulee{FnApply} case: proceeds on the similar lines as
+    \rulee{MethodInv} case (an exception is that the typing context
+    for the method body already includes $\Delta$).
 
-    \item \rulee{FnApply} case: proceeds on the similar lines as
-    \rulee{MethodInv} case, except that no strengthening is needed;
-    function body is typed under a context that includes $\Delta$.
+    \item \rulee{LetRegionBegin} case ($e = \letregion{\pi}{e'}$):
+    Steps to $\letd{\loc}{[\loc/\pi]e'}$.
+    $\loc$ is the new location introduced. Since $\loc \notin
+    dom(\mem)$, it follows that $\loc \notin \Delta$. Rest of the
+    premises required to apply the \C{letd} type rule on result
+    expression are obtained from the \C{letregion} type rule of the
+    initial expression ($e$), by substituing $\loc$ for $\pi$.
 
-    \item \rulee{LetRegionBegin} case: Since $\loc \notin dom(\mem)$,
-    it follows that $\loc \notin \Delta$. Rest of the premises
-    required to apply the \C{letd} type rule on result expression are
-    obtained from the \C{letregion} type rule of the initial
-    expression ($e$), by substituing $\loc$ for $\pi$.
-
-    \item \rulee{LetRegionEnd} case: $e$ is a \C{letd} expression
-    which reduces to a value $v$. Invering the typing derivation of
-    $e$ yeilds the premise that $v$ is well-typed under the context
-    $(\Delta \cup \{\loc\},\cdot,\phicx \conj \Delta \outlives \loc)$.
-    However, we have to prove that $v$ is well-typed under the smaller
-    context $(\Delta,\cdot,\phicx)$. We carry out this proof by
+    \item \rulee{LetRegionEnd} case ($e = \letd{\loc}{v}$): Steps to
+    $v$. Inverting the typing derivation of $e$ yeilds the premise
+    that $v$ is well-typed under the context $(\Delta \cup
+    \{\loc\},\cdot,\phicx \conj \Delta \outlives \loc), \cdot, \loc$,
+    and that the type is well-formed under the smaller context
+    $(\Delta,\cdot,\phicx)$. Let the type be $\tau$. We now have to
+    prove that $v$ is well-typed ($\tau$) under the typing context
+    $(\Delta,\cdot,\phicx), \cdot, \loc'$. We carry out this proof by
     induction on the structure of value $v$:
     \begin{itemize}
-      \item If $v$ is an object value (e.g.,
-      $B\inang{\tbar}\inang{\locbar}(\vbar)$), IH yeilds the
-      well-typedness of $\vbar$ under $(\Delta,\cdot,\phicx)$, and the
-      premise 
-      $\tywf{(\Delta,\cdot,\phicx)} {B\inang{\tbar}\inang{\locbar}}$,
-      obtained by inverting the type judgment for \C{letd}, yeilds the
-      well-typedness of whole value.
+      \item When $v$ is an object value (i.e., $\C{new}\;
+      B\inang{\tbar}\inang{\locbar}(\vbar)$): We know that the type of
+      $\vbar$ is determined by
+      $\fields(B\inang{\tbar}\inang{\locbar})$. Since
+      $B\inang{\tbar}\inang{\locbar}$ is well-formed under
+      $(\Delta,\cdot,\phicx)$ (premise), and $\loc \notin \Delta$, we
+      know that types of $\vbar$ are well-formed under
+      $(\Delta,\cdot,\phicx)$ (see class well-formedness rule). Hence,
+      by IH, they retain their type under the typing context
+      $(\Delta,\cdot,\phicx), \cdot, \loc'$. Using the type rule of
+      \C{new}, we derive that $\C{new}\;
+      B\inang{\tbar}\inang{\locbar}(\vbar)$ has the type $\tau$ under
+      $(\Delta,\cdot,\phicx), \cdot, \loc'$.
 
-      \item A region handler value ($\RgnZT{\toploc\loc'}(v)$) is
-      well-typed under any context.
+      \item When $v$ is a region handler (i.e.,
+      $\RgnZT{\toploc\loc''}(v')$): $v'$ has type $T@\loc''$ under
+      $(\Delta \cup \{\loc,\loc''\},\cdot,\phicx \conj \Delta
+      \outlives \loc), \cdot, \loc$, and the type is well-formed under
+      $(\Delta \cup \{\loc,\loc''\},\cdot,\phicx \conj \Delta
+      \outlives \loc)$. But, $T@\loc''$ is a type that is well-formed
+      under the minimal context of $(\{\loc''\},\cdot,true)$. Since
+      the context can be strengthened without affecting
+      well-formedness, the type is also well-formed under $(\Delta
+      \cup \{\loc''\},\cdot,\phicx)$. By IH, $v'$ retains it's type
+      ($T@\loc''$) under $(\Delta \cup \{\loc''\},\cdot,\phicx)$. By
+      applying IH, we derive $\hastyp{(\Delta \cup
+      \{\loc''\},\cdot,\phicx), \cdot,
+      \loc'}{\RgnZT{\toploc\loc''}(v')}{\RgnZT{\toprgn}}$.
 
-      \item If $v$ is a function closure, then it is well-tuped under
-      $(\Delta,\cdot,\phicx)$ only if it doesn't trap any references
-      to $\loc$, the \C{letd} location. Since closure's type needs to
-      be well-formed under $(\Delta,\cdot,\phicx)$, its allocation
-      region belongs to $\Delta$, hence outlives $\loc$. Since the
-      closure typing rule requires all free region variables of the
-      closure body outlive the allocation region of closure, it
-      follows that all free region variables strictly outlive $\loc$,
-      hence they belong to $\Delta$. Hence the function closure is
-      well-typed outside \C{letd}.
+      \item When $v$ is a function closure (i.e.,
+      $\lambdaexp{\loc''}{\rho\rhobar\,|\,\phi}{\xbar:\taubar}{e_0}$).
+      By inversion on the typing deriviation of $v$, we know that
+      $\loc'' = \loc$ since a function closure is always allocated in
+      the current allocation context (i.e., the youngest region).
+      Hence it's type ($\tau$) is $\inang{\rho\rhobar \,|\,
+      \phi}\taubar\xrightarrow{\loc}\tau'$, for some $\tau'$. Since
+      $\loc \notin \Delta$, $\tau$ can not be well-formed under
+      $(\Delta, \cdot, \phicx)$ - a contradiction. In other words, a
+      function closure can never escape a region's lexical scope.
     \end{itemize}
 
-  \item \rulee{OpenEnd} case: The proof is similar to the \C{letd}
-  case.
+  \item \rulee{Open} case ($e \,=\, \open{v_a}{y}{\pi}{e_b}$): We know
+  that $v_a \,=\, \RgnZT{\toploc\loc}(v)$, for some location $\loc$
+  and value $v$. By inversion on the typing derivation:
+  \begin{smathpar}
+  \begin{array}{l}
+    \pi \notin \Delta \\
+    \hastyp{(\Delta \cup \{\pi\},\cdot,\phicx), [y\mapsto T@\pi],
+    \pi}{e_b}{\tau}\\
+    \tywf{(\Delta, \cdot, \phicx)}{\tau}\\
+    \hastyp{(\Delta,\cdot,\phicx),\cdot,\rgn}{v_a} {\RgnZT{\toprgn}}\\
+  \end{array}
+  \end{smathpar}
+  Since $\pi \notin \Delta$ and $\tywf{\Delta}{\phicx}$ and
+  $\tywf{(\Delta, \cdot, \phicx)}{\tau}$ (i.e., $\pi$ does not occur
+  free in $\phicx$ and $\tau$):
+  \begin{smathpar}
+  \begin{array}{l}
+    \hastyp{(\Delta \cup \{\loc\},\cdot,\phicx), [y\mapsto T@\loc],
+    \loc}{[\loc/\pi]e_b}{\tau}\\
+  \end{array}
+  \end{smathpar}
+  Since $\hastyp{(\Delta \cup \{\loc\},\cdot,\phicx), \cdot, \loc}
+  {v}{T@\loc}$, by substitution lemma:
+  \begin{smathpar}
+  \begin{array}{l}
+    \hastyp{(\Delta \cup \{\loc\},\cdot,\phicx), \cdot, \loc}
+      {[v/y][\loc/\pi]e_b}{\tau}\\
+  \end{array}
+  \end{smathpar}
+  We know that the reduced expression ($e'$) is $\opened{l}{s} 
+  {[v/y][\loc/\pi]e_b}$, where $s = \mem(\loc)$. Moreover, if 
+  $\loc \in \Delta$, then since $\consistent(\Delta,\mem)$, we have
+  $\mem(\loc)=\LIVE$. Therefore, by applying the type rule for
+  \C{opened} expression, we derive:
+  \begin{smathpar}
+  \begin{array}{l}
+    \hastyp{(\Delta,\cdot,\phicx), \cdot, \rgn} 
+      {\opened{\loc}{s}{[v/y][\loc/\pi]e_b}}{\tau}\\
+  \end{array}
+  \end{smathpar}
+  Hence, $e'$ has the same type as $e$.
+
+  \item \rulee{OpenEnd} case: The proof is very similar to the
+  \C{letd} case.
   \qed

### 2017-01-02 paper/broom.tex
added appendix and moved some material there
Wibbles plus new reference
Removed rhoAlloc and Sigma from type judgment. Appendix-related edits.
Progress lemma updated.
Preservation updated.
updated constraint generation
switched to ESOP formatting
New locations-based opsem.
type inference proofs
Fixed OPEN bug in opsem.
Static semantics for locations-based opsem.
SLIVE & TLIVE merged into LIVE
Restored Delta in staticsem.
Sec.3 edited.
ECOOP initial version

@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
-%               sigplanconf-guide.pdf
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
+\documentclass[a4paper,UKenglish]{lipics-v2016}
+%This is a template for producing LIPIcs articles. 
+%See lipics-manual.pdf for further information.
+%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+% for section-numbered lemmas etc., use "numberwithinsect"
 
-\documentclass[preprint,numbers]{sigplanconf}
+\usepackage{microtype}%if unwanted, comment out or use option "draft"
 
-% The following \documentclass options may be useful:
+%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
 
-% preprint      Remove this option only once the paper is in final form.
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
+\bibliographystyle{plainurl}% the recommended bibstyle
+
+%%%%%%%%%%%%%%%% OUR PRELIM STUFF BEGINS %%%%%%%%%%%%%%%%%%%%
 
@@
 \usepackage[scaled]{helvet} % see www.ctan.org/get/macros/latex/required/psnfss/psnfss2e.pdf
-\usepackage{listings}          % format code
+% \usepackage{listings}          % format code
 \usepackage{enumitem}      % adjust spacing in enums
@@
 \usepackage{graphicx}
-\usepackage[usenames,dvipsnames]{color}
+% \usepackage[usenames,dvipsnames]{color}
 \usepackage{listings}
@@
 \usepackage{breakurl}
-\usepackage[breaklinks,draft=false,backref]{hyperref}
+% \usepackage[breaklinks,draft=false,backref]{hyperref}
 \hypersetup{
@@
 \input{macros}
+
+%%%%%%%%%%%%%%%% OUR PRELIM STUFF ENDS %%%%%%%%%%%%%%%%%%%%
+
+% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title{Safe Transferable Regions}
+
+%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
+\author[1]{Gowthan Kaki}
+\author[2]{G. Ramalingam}
+\author[3]{Kapil Vaswani}
+\author[4]{Dimitrios Vytiniotis}
+\affil[1]{Purdue University}
+\affil[2]{Microsoft Research}
+\affil[3]{Microsoft Research}
+\affil[4]{Microsoft Research}
+\authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
+
+\Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+% \subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html .
+E.g., cite as "F.1.1 Models of Computation".
+\keywords{Memory management, regions, type system, type inference}% mandatory: Please provide 1-5 keywords
+% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{John Q. Open and Joan R. Acces}
+\EventNoEds{2}
+\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+\EventShortTitle{ECOOP 2017}
+\EventAcronym{ECOOP}
+\EventYear{2017}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
 
-\newcommand{\dv}[1]{{\bf DV:}\begin{color}{red}#1\end{color}}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-
-%\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-% \titlebanner{banner above paper title}        % These are ignored unless
-% \preprintfooter{short description of paper}   % 'preprint' option specified.
-
-%The Paper
-%---------------------
-\title{Safe Transferable Regions}
-
-\authorinfo{} {} {}
 \maketitle
@@
 \end{abstract}
-
-% general terms are not compulsory anymore, 
-% you may leave them out
-%% \terms
-%% term1, term2
 
@@
 
+\renewcommand{\rgn}{\pi}
 \input{type-inference}
+\renewcommand{\rgn}{r}
 
@@
 \input{related}
-% \appendix
-% \section{Appendix Title}
 
-% This is the text of the appendix, if you need one.
+%%
+%% Appendix
+%%
 
-% We recommend abbrvnat bibliography style.
+\appendix
 
-\bibliographystyle{abbrvnat}
-\small
+\input{appendix}
+
+%%
+%% Bibliography
+%%
+
 \bibliography{broom}

### 2017-01-02 paper/introduction.tex
fixed minor formatting issues

@@
 \newcommand{\ie}{\emph{i.e.}}
+\newcommand{\mypara}[1]{\textbf{#1}}
 
@@
 
-\paragraph{Performance.}
+\mypara{Performance.}
 When the transferred data structures are large, as is the case with many
@@
 
-\paragraph{Safety.}
+\mypara{Safety.}
 Several of these systems are designed to allow the producer and consumer to
@@
 
-\paragraph{Safe Transferable Regions.}
+\mypara{Safe Transferable Regions.}
 In this paper, we present a language-based approach to cleanly encapsulated
@@
 
-\paragraph{Language.}
+\mypara{Language.}
 The cornerstone of this approach is an \C{open} lexical block for transferable regions,
@@
 
-\paragraph{Type System.}
+\mypara{Type System.}
 Formally, our type system introduces \emph{region parameters} (for both
@@
 
-\paragraph{Lightweight Runtime Checks.}
+\mypara{Lightweight Runtime Checks.}
 Ensuring memory safety using this approach requires ensuring
@@
 
-\paragraph{Type Inference.}
+\mypara{Type Inference.}
 One of the key contributions of this paper is a sound type inference algorithm
@@
 
-\paragraph{Evaluation.}
+\mypara{Evaluation.}
 Our work was inspired by~\cite{Broom:HotOS}, which presents evidence that
@@
 
-\paragraph{Contributions.}
+\mypara{Contributions.}
 The paper makes the following contributions:

### 2017-01-02 paper/overview2.tex
fixed minor formatting issues

@@
 
-\paragraph{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
+\mypara{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
 creates a new stack region, with a static identifier \C{R}, whose
@@
 
-\paragraph{Object Allocation} The ``\C{new@R T()}'' construct creates
+\mypara{Object Allocation} The ``\C{new@R T()}'' construct creates
 a new object of type \C{T} in the region \C{R}. The specification of
@@
 
-\paragraph{Transferable Regions} 
+\mypara{Transferable Regions} 
 % \name's \emph{transferable regions} are an encapsulation of a
@@
 % \emph{dynamic} regions.)
-
+%
 Transferable regions are first class values of \name: they are objects
@@
 
-\paragraph{Open and Closed Regions} A transferable region must be
+\mypara{Open and Closed Regions} A transferable region must be
 explicitly \emph{opened} using \name's \C{open} construct in order to
@@
 
-\paragraph{Motivating Example}
+\mypara{Motivating Example}
 % We now illustrate how the features of \name can be used to 
@@
 
-\paragraph{Cloning} Note that in the example from
+\mypara{Cloning} Note that in the example from
 Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the

### 2017-01-02 paper/type-inference.tex
fixed minor formatting issues

@@
 
-\paragraph{Overview.}
-
+\mypara{Overview.}
+%
 % Fig.~\ref{fig:type-inference-algo}
@@
 
-\paragraph{Syntax of Constraints.}
+\mypara{Syntax of Constraints.}
 (See~Fig.~\ref{fig:constraint-syntax}.)
@@
 
-\paragraph{Constraint Solution.}
+\mypara{Constraint Solution.}
 We define an \emph{assignment} $\soln$ to be a pair of functions $(\solnR,\solnP)$,
@@
 
-\paragraph{Constraint Generation.}
+\mypara{Constraint Generation.}
 The constraint generation algorithm is a direct adaption of the type checker:
@@
 
-\paragraph{Constraint Solver.}
+\mypara{Constraint Solver.}
 The first step in our algorithm for solving a set of constraints $C$
@@
 
-\paragraph{Algorithmic Aspects.}
+\mypara{Algorithmic Aspects.}
 Note that checking the validity of a ground constraint can be realized using
@@
 %
-%\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
+%\mypara{Constraints Example 1} Consider the $\C{Pair}$ class
 %template from \S\ref{sec:fb-templatization}. Following constraints are
@@
 %
-%\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
+%\mypara{Constraints Example 2} Let us add to the \C{Pair} class a
 %contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
@@
 %
-% \paragraph{Constraints Example 1} The original source of all
+% \mypara{Constraints Example 1} The original source of all
 % outlives constraints is the validity constraint $C_3$ in $\elabClass$
@@
 
-% \paragraph{Constraints Example 2} Consider a contrived method
+% \mypara{Constraints Example 2} Consider a contrived method
 % $\C{alt}$ that accepts an argument $\C{p}$ of type

### 2017-01-05 paper/related.tex
added alloc context to constraint generation rules

@@
 % that language.  
-Their language permits only lexically-scoped (stack) regions
-(in the context of shared-memory concurrency). In contrast, we permit
-regions with dynamically determined lifetimes (in the context of message-passing concurrency).
+Their language permits only lexically-scoped (stack) regions.
+% (in the context of shared-memory concurrency).
+In contrast, we permit regions with dynamically determined lifetimes.
+% (in the context of message-passing concurrency).
 % We borrow outlives relation from their formal

### 2017-01-06 paper/abstract.tex
New abstract.

--- 2016-10-14 paper/abstract.tex
@@
-% Recent work in high-performance systems written in managed languages
-% (such as Java or C\#) has shown that garbage-collection can be a
-% significant performance bottleneck.
-% A class of these systems, focused on big-data, create numerous large
-% data structures with well-defined lifetimes.
-%
-There is increasing interest in alternative memory management techniques
-that strike a balance between the convenience of garbage collection and
-the better performance that manual memory management enables, without
-compromising safety.
-%
-In this paper, we present a language and a memory management scheme based
-on user-managed memory regions (called \emph{transferable regions})
-that allow programmers to exploit knowledge of data structures' lifetimes
-to improve performance.
+There is an increasing interest in alternative memory management
+schemes that seek to the combine the convenience of garbage collection
+and the performance of manual memory management in a single language
+framework.  Unfortunately, ensuring safety in presence of manual
+memory management remains as a challenge. In this paper, we present a
+C\#-like object-oriented language called \name that uses a combination
+of region type system and lightweight runtime checks to enforce safety
+in presence of user-managed memory regions called \emph{transferable
+regions}. Unsafe transferable regions have been previously used in
+dataflow computations to contain the latency due to unbounded GC
+pauses. Our approach shows that it is possible to restore safety
+without compromising on the benefits of transferable regions. We prove
+the type safety of \name in a formal framework that includes its
+C\#-inspired features, such as higher-order functions and generics. We
+complement our type system with a type inference algorithm, which
+eliminates the need for programmers to write region annotations on
+types. The inference algorithm has been proven sound and relatively
+complete. We describe a prototype implementation of the inference
+algorithm, and our experience of using it to enforce memory safety in
+dataflow programs.
+% %
+% In this paper, we present a language and a memory management scheme based
+% on user-managed memory regions (called \emph{transferable regions})
+% that allow programmers to exploit knowledge of data structures' lifetimes
+% to improve performance.
 
-Manual memory management is susceptible to the usual perils of
-dangling pointers. A key contribution of this paper is a
-refinement-based region type system, for a C\#-like typed object-oriented
-language with higher-order functions and generics, that ensures the
-memory safety of programs in the presence of transferable regions.
-We complement our type system with a type inference algorithm,
-which eliminates the need for programmers to write region annotations on types.
-The cornerstone of our inference algorithm is a novel constraint solver that
-performs abduction in a partial-order constraint domain to infer weakest solutions
-to recursive constraints.
-We establish soundness and partial completeness results for the type inference algorithm:
-in particular, the key steps of constraint-generation and constraint-solving
-in the type inference algorithm are complete.
-
-% while facilitating the reuse of existing C\# libraries with no modifications.
-% Experiments demonstrate the practical utility of our approach.
+% Manual memory management is susceptible to the usual perils of
+% dangling pointers. A key contribution of this paper is a
+% refinement-based region type system, for a C\#-like typed
+% object-oriented language with higher-order functions and generics,
+% that ensures the memory safety of programs in the presence of
+% transferable regions.

### 2017-01-06 paper/abstract.tex
Edit.

@@
 framework.  Unfortunately, ensuring safety in presence of manual
-memory management remains as a challenge. In this paper, we present a
-C\#-like object-oriented language called \name that uses a combination
-of region type system and lightweight runtime checks to enforce safety
-in presence of user-managed memory regions called \emph{transferable
-regions}. Unsafe transferable regions have been previously used in
-dataflow computations to contain the latency due to unbounded GC
-pauses. Our approach shows that it is possible to restore safety
-without compromising on the benefits of transferable regions. We prove
-the type safety of \name in a formal framework that includes its
-C\#-inspired features, such as higher-order functions and generics. We
-complement our type system with a type inference algorithm, which
-eliminates the need for programmers to write region annotations on
-types. The inference algorithm has been proven sound and relatively
-complete. We describe a prototype implementation of the inference
-algorithm, and our experience of using it to enforce memory safety in
-dataflow programs.
+memory management remains as great a challenge as ever. In this paper,
+we present a C\#-like object-oriented language called \name that uses
+a combination of region type system and lightweight runtime checks to
+enforce safety in presence of user-managed memory regions called
+\emph{transferable regions}. Unsafe transferable regions have been
+previously used in dataflow computations to contain the latency due to
+unbounded GC pauses. Our approach shows that it is possible to restore
+safety without compromising on the benefits of transferable regions.
+We prove the type safety of \name in a formal framework that includes
+its C\#-inspired features, such as higher-order functions and
+generics. We complement our type system with a type inference
+algorithm, which eliminates the need for programmers to write region
+annotations on types. The inference algorithm has been proven sound
+and relatively complete. We describe a prototype implementation of the
+inference algorithm, and our experience of using it to enforce memory
+safety in dataflow programs.
 % %

### 2017-01-09 paper/misc.tex
type inference minor updates

@@
-\section{Other Aspects}
+% \subsection{Other Aspects}
 
-\paragraph{Modularity Aspects of Type Inference.}
+\subsection{Modularity Aspects of Type Inference}
+
 The type inference algorithm, as presented, traverses the entire program to

### 2017-01-09 paper/type-inference.tex
type inference minor updates

@@
 We later on establish the soundness of the type inference algorithm,
-and the completeness of constraint-solving step.
+and the completeness of the constraint-generation and constraint-solving steps
+(steps 2 and 3 above).
 
@@
 &
-\begin{minipage}{0.6\linewidth}
+\begin{minipage}{0.5\linewidth}
 \begin{codejava}
@@
 &
-\begin{minipage}{0.65\linewidth}
+\begin{minipage}{0.5\linewidth}
 \begin{codejava}
@@
 The type parameter \C{T} of a class \C{C} is instantiated as $\C{T}@\rho$ using a single region parameter $\rho$. 
-(This can be extended to use the bound specified for \C{T}, if any. We do not
-discuss this due to space restrictions.)
+(This can be extended to use the bound specified for \C{T}, if any.)
 % number of region parameters
@@
 
-Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
-of a method invocation or object allocation, which will be bound to a region constant by the end of the
-type inference.
-A region constant may be 
+Recall that a \emph{region constant} may be either
 (a) a \emph{formal region parameter} of a class or method, or
@@
 (c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.
+A \emph{region variable} is introduced to represent an unknown \emph{actual} region parameter
+of a method invocation or object allocation, and the constraint-solver, if successful, 
+will bind each region variable to a region constant.
+% by the end of the type inference.
 
+A \emph{region-constraint} $\phi$ consists of a conjunction of \emph{outlives-constraints}
+of the form $\rho_1 \outlives \rho_2$.
 A predicate variable $\varphi$ is introduced to represent, \eg, the unknown precondition of a method.
@@
 restrict the domain of a predicate variable ($\varphi$) to the set of all possible
-constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.
+region-constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.
 
@@
 Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
-Thus, the component $\phicx$ of $\A$ will now be a symbolic expression
+Thus, the component $\phicx$ of $\A$ will now be a \emph{symbolic expression}
 of the form $\varphi \conj \phi$, where $\varphi$ is the predicate variable
@@
 
-Note that some of the conditions in the antecedents are checked at constraint-generation
-time, while some are encoded as constraints (to be resolved by the constraint-solver).
-E.g., we assume that the region constants introduced by
-\C{open}
-and
-\C{letregion}
-constructs are unique (and the elaboration phase ensures this by alpha-renaming
+When we adapt a type-checking judgement rule to produce a corresponding
+constraint-generation rule, we treat the antecedent conditions of the type-checking
+rule in one of two ways. Many of these antecedent conditions are converted into
+constraints which are accumulated in the set $C$. However, some of the antecedent
+conditions are expected to be trivially satisfied and are expressed as
+antecedent conditions in the constraint-generation rule as well.
+% Note that some of the conditions in the antecedents are checked at constraint-generation
+% time, while some are encoded as constraints (to be resolved by the constraint-solver).
+E.g., our frontend ensures that the region constants introduced by \C{open}
+and \C{letregion} constructs are unique (and the elaboration phase ensures this by alpha-renaming
 them as needed). Thus, the precondition $\rho \not\in \Delta$ in rules \textsc{LetRegion}
-and \textsc{Open} merely validates this at constraint-generation time.
+and \textsc{Open} is expected to hold true at constraint-generation time
+(and does not produce any constraints).
 
@@
 Let $\consOf{q}$ denote the set of constraints generated from an elaborated program $q$.
-The following theorem states that constraint-generation is sound:
+The following theorem states that constraint-generation is sound and complete:
 \begin{theorem}
@@
 An assignment $\soln$ (for the region and predicate variables in $q$)
-satisfies $C$ only if $q[\soln]$ is well-typed.
+satisfies $C$ iff $q[\soln]$ is well-typed.
 \end{theorem}

### 2017-01-10 paper/esop_rebuttal.txt
Edits.

We thank the reviewers for providing thoughtful comments and constructive
feedback.

Review 1:

(a) Reviewer 1's primary concern (that "the type system is not designed for
first-class polymorphism") is misplaced.  Our type system DOES admit first
class polymorphism (\tau in Fig. 5); higher-order arguments and higher-order
class fields can have region-polymorphic types (discussed on p.10). Polymorphic
recursion is also admitted via "this" (Method well-formedness in Fig. 12). Type
inference includes an iterative fixed-point constraint solver (Sec. 4.4) that
computes instantiations for recursive calls.

First-class polymorphism does complicate inference. Please see discussion point
(2) in section 4.5 (page 21).  Our heuristic choice in this case favors
first-class polymorphism as the more likely scenario.  As discussed in section
4.5, it is straightforward to support user (region type) annotations in such
scenarios, allowing the type inference to be effective with few annotations. We
believe that this is a very practical solution.

(b) Regarding the concern about novelty: The idea of (dynamic) regions is not
new, but we believe that there are sufficiently new ideas in the paper: E.g.,
the constraint-solver discussed in sections 4.3 and 4.4 is novel, especially as
it works with "outlives" constraints, and we are not aware of any work that
shows how we can solve such constraints.

(c) Broom region management has been used for real-world computations in the
Naiad dataflow systems ([5] has more details).  Our implementation of the type
inference engine works on all the Naiad programs described in [5] (rewritten
into FB+ (Sec. 5) mainly to convert interface-based inheritance to class-based
inheritance).

Review 2:

(d) The reviewer asks "which operations can fail at runtime." Runtime checks
are used only to enforce the state transition discipline of Fig.4 on Region
handlers. Hence, only operations on region handlers require runtime checks, as
against operations on the objects stored in the region pointed by the handler.
The contribution of the type system is to ensure that it is indeed enough to
instrument only the region handler dereferences, even though the integrity of
all dereferences is under jeopardy. Operational semantics (Fig. 15) clarifies
which operations require runtime checks.

We will explain this in more detail in the paper.

(e) The reviewer suggests that we should be able to rule out some problematic
scenarios (e.g. freeing a region that is known to currently be open)
statically.

Consider the scenario of an "open region being freed". SOME instances of this
may be easy to identify, in which case producing a static warning is useful and
appropriate. However, soundly identifying ALL instances of this, without
producing too many false positives, is hard because of possible aliases that
can be created to regions. Forbidding the creation of aliases to regions (or
forbidding placing region handles in data-structures like dictionaries) would
be restrictive. Hence, our particular design choice.

(f) The reviewer asks why subtyping is not emphasized in the paper, and notes
that while Cyclone supports subtyping based on region lifetimes, Broom does
not.

Subtyping in FB is a trivial extension of the underlying FGJ calculus and
orthogonal to outlives-constraints.  Hence, it is not emphasized. Please note
that a subtype relation based on outlives constraints is unsound. That is, if
region R1 outlives region R2, it is UNSOUND to treat T@R1 as a subtype of T@R2
because T could be a mutable object that can point to other objects. Where it
may be safe to destructively update a T@R2 object's field to point to an object
in R2, it won't be safe to update a T@R1 object's field to point to that object
in R2. Broom therefore doesn’t lift outlives constraints to subtype relation,
choosing instead to record them explicitly in class headers. The suggested list
example is typable this way (Sections 4.1, 4.2 & 4.3 in fact discuss this
example).

(g) Regarding the question about region-unaware legacy code: "open" blocks are
needed only if the code tries to allocate objects explicitly in a transferable
region. Legacy code is not region aware, hence this is not needed. A legacy
method's region constraints are captured explicitly in its region type, and as
long as the caller satisfies these constraints, the method doesn't care whether
a region is transferable or static. In other words, if a region needs to be
open for a piece of legacy code to be safe, then the type system enforces this
obligation on the caller, not the callee.

(h) Reviewer seems to suggest that, short of inferring all region annotations
(including open blocks and letregions), region type inference is not very
useful. We disagree. As mentioned in the previous point, type inference is
crucial to infer region constraints on the legacy code, so that we can be sure
that its composition with the region-annotated code is indeed safe. An example
to demonstrate this point is provided at the end of this rebuttal.

Review 3:
(i) We did not bother to distinguish "transfer" and "free" much in the paper
because, from the type-system perspective, the distinction is not much. But we
will clarify this better in the paper.

------------------------------------------------------------------------------
Example to accompany point (h)

For illustrative purpose, consider a hypothetical function foo defined as
following:

void foo(A x, A y, A z, B b, unsigned int i) {
  if (i>=4) {
    return;
  }
  x.b = b;
  foo(y,z,x,b,i+1);
}

Assume that a programmer uses foo in the following context:

let rgn = new Region<B>();
open rgn as b0@R0 {
  B b1 = new B();
  letregion R1 {
    A x = new A(b0);
    A y = new A(b0);
    A z = new A(b0);
    …
    foo(x,y,z,b1,1);
  }
}

Is this use safe? It is to determine answers to this kind of questions that we
need inference. “foo” could be standard library function, or generally any
function whose implementation may not be familiar to the programmer. Reasoning
about outlives properties in the context of nested function calls is
non-obvious (esp. in the presence of polymorphic recursion), and requires tool
support. Type inference serves precisely this purpose.

### 2017-01-10 paper/type-inference.tex
minor fixes

@@
 $\deltaPC(\varphi)$.
+(Our constraint-generation algorithm guarantees the preceding property.
+Even otherwise, a set of constraints $\tywf{\Delta_i}{\varphi}$ can be
+replaced by the single equivalent constraint $\tywf{(\cap_i \Delta_i)}{\varphi}$.)
 %
@@
 $\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.
+We say that a region variable $\rho$ is \emph{somewhere-bound} to a region constant $\pi$ (in $D$)
+if $\rho$ is bound to $\pi$ in some context $\ell$ in which it occurs.
 
@@
 For any region variable $\rho$, we define $\thesolnR(\rho)$ to be any element of the
-set $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$,
+set
+% $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$,
+$\{ \pi \in \regionConstants \:|\; \rho \text{ is somewhere-bound to } \pi \text{ in } \saturate{C} \}$,
 if this set is non-empty. If this set is empty for any $\rho$, $\thesoln$ is undefined.
@@
 The soundness of the type inference algorithm follows.
-Theorem~\ref{thm:constraint-solver-sc} also establishes the
-completeness of constraint solving.
 
-The main source of incompleteness in the type inference algorithm is
-the set of heuristic choices made during region parameterization step,
+Theorem~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc} also
+establish the completeness of the constraint-generation and constraint-solving steps.
+%
+The only source of incompleteness in the type inference algorithm is
+the set of heuristic choices made during the first step,
 as explained below.

### 2017-01-11 paper/related.tex
ownwership inference related work

@@
 type inference.
+However, several inference algorithms have been proposed in the context
+of other ownership type systems.
+Our type inference algorithm is also novel compared to these existing
+ownership inference algorithms, which are based on, e.g.,
+pointer analysis~\cite{HuangEtAl:ECOOP12} or boolean satisfiability~\cite{DietlEtAl:ECOOP11}.
+(See Section 5.2 of~\cite{OwnershipSurvey} for a more comprehensive discussion
+of ownership inference algorithms.)
+Some distinguishing characteristics of our algorithm is that it is customized
+to our problem, does not use any pointer analysis algorithm (which
+can be a source of imprecision) or SAT solvers (which can be a source of inefficiency),
+and comes with relative completeness guarantees.
 
@@
 % eliminates any such need.
-% Our type inference algorithm is also novel compared to existing type
-% inference approaches for ownership types, which are based on, e.g.,
-% context-insensitive pointer analysis or boolean satisfiability.
-% E.g., papers 55 and 79.
 

### 2017-01-13 paper/introduction.tex
type inference minor updates
Edits to the intro.

@@
 
-This paper presents a \emph{safe} and \emph{efficient} region-based approach
-to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
+% This paper presents a \emph{safe} and \emph{efficient} region-based approach
+% to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
 
-Consider the example (of  a producer creating a data structure and transferring it to a consumer)
-in Fig.~\ref{fig:motivating-eg}.
-This represents a \C{SELECT} query operator that receives a stream of input messages,
-each associated with a time window $t$, processed by method \texttt{onReceive}.
-Each input message contains a list of inputs, each processed by applying a user-defined function to
-create a corresponding output.  Multiple messages with the same timestamp
-are permitted and messages with different timestamps may arrive out of order.
-An invocation of method
-\texttt{OnNotify} indicates that no more input messages with a timestamp
-$t$ will be subsequently delivered.  At this point, the operator completes
-the processing for time window $t$ and sends a corresponding output message
-to its successor.
+Computations performed by a network of concurrent communicating actors
+often involve data transfer between producers and consumers. Consider
+for example the \C{SELECT} query operator shown in
+Fig.~\ref{fig:motivating-eg}, which functions as an actor in a
+dataflow computation involving a network of other such query
+operators. \C{SELECT} receives a stream of input messages, each
+associated with a time window $t$, processed by method
+\texttt{onReceive}.  Each input message contains a list of inputs,
+each processed by applying a user-defined function to create a
+corresponding output.  Multiple messages with the same timestamp are
+permitted and messages with different timestamps may arrive out of
+order.  An invocation of method \texttt{OnNotify} indicates that no
+more input messages with a timestamp $t$ will be subsequently
+delivered.  At this point, the operator completes the processing for
+time window $t$ and sends a corresponding output message to its
+successor.
 
@@
 \mypara{Safe Transferable Regions.}
-In this paper, we present a language-based approach to cleanly encapsulated
+In this paper, we present a language-based approach to cleanly encapsulate
 transferable data and a safe and efficient implementation of such data based
@@
 between the region parameters.
+The type system may be seen as a form of ownership type system~\cite{OwnershipSurvey},
+with a region being the owner of all objects allocated in that region.
 
@@
 \mypara{Type Inference.}
-One of the key contributions of this paper is a sound type inference algorithm
-that eliminates the need for users to write region type annotations.
-The users write programs in the underlying language that provides
-primitives for the users to manipulate regions (create, open, free,
-transfer) and to allocate objects in regions, but has no region type
-annotations.
+One of the key contributions of this paper is a type inference
+algorithm that eliminates the need for users to write region type
+annotations.  The users write programs in the underlying language that
+provides primitives for the users to manipulate regions (create, open,
+free, transfer) and to allocate objects in regions, but has no region
+type annotations.
 % This, we believe, significantly reduces the impediment to adopt our
@@
 between the region parameters).  We show that the algorithm is sound,
-and the last stage of constraint solving is complete.
+and the stages of constraint generation and solving are complete
+(i.e., the algorithm is complete relative to the first stage of
+elaboration).
 % (using an interprocedural fixed point computation).
@@
 \mypara{Evaluation.}
-Our work was inspired by~\cite{Broom:HotOS}, which presents evidence that
-realistic programs can be written using transferable regions
-and that this can yield significant performance improvements.
-The system described in~\cite{Broom:HotOS}, however, provides no memory
-safety guarantees, which is the subject of our contribution.
-We report on an implementation of our type inference algorithm that
-was able to identify some unsafe memory accesses and was able to successfully
-verify programs correct (once the errors were fixed).
+Our work was inspired by~\cite{Broom:HotOS}, which presents evidence
+that realistic programs can be written using transferable regions and
+that this can yield significant performance improvements.
+While~\cite{Broom:HotOS} addresses the engineering challenges in
+extending a managed runtime with transferable regions, it adopts an ad
+hoc approach insofar as language design is concerned, exposing the
+region functionality through an unsafe API, and in the process losing
+the safety guarantees. Our contribution is an alternative approach
+that is grounded in sound theory and restores the language safety
+guarantees. The utility of our approach is demonstrated by a prototype
+implementation of our type inference algorithm that was able to
+identify unsafe memory accesses among the benchmarks extracted
+from~\cite{Broom:HotOS}.
 
@@
 \begin{itemize} 
-  \item We present \broom, a \csharp-like typed
-  object-oriented language that eschews garbage collection in favour of
-  programmer-managed memory regions. \broom{} extends its core language,
-  which includes \emph{lambdas} (higher-order functions) and
-  \emph{generics} (parametric polymorphism), with constructs to create,
-  manage and free static and transferable memory regions. Transferable regions
-  are first-class values in \broom.
+  \item We present \broom, a \csharp-like typed object-oriented
+  language that supports programmer-managed memory regions. \broom{}
+  extends its core language, which includes \emph{lambdas}
+  (higher-order functions) and \emph{generics} (parametric
+  polymorphism), with constructs to create, manage and free static and
+  transferable memory regions. Transferable regions are first-class
+  values in \broom.
   % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
@@
   region types, and (b). enables seamless interoperability between
-  region-aware \broom programs and legacy standard library code that is
-  region-oblivious. The cornerstone of our inference algorithm is a
-  novel constraint solver that performs abduction in a partial-order
+  region-aware \broom programs and legacy standard library code that
+  is region-oblivious. Our inference algorithm is based on a novel
+  constraint solver that performs abduction in a partial-order
   constraint domain to infer weakest solutions to recursive
@@
 
-\item We establish soundness of type inference algorithm, and
-completeness of constraint solving.
+\item We establish the soundness and relative completeness of the type
+inference algorithm.
 

### 2017-01-13 paper/type-system.tex
Sec.3 made more peppy.

@@
 region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
-$R_2$ is guaranteed to outlive $R_1$.  We now formally develop this
-intuition via \fbname (\FB), our explicitly typed core language (with
-region types) that incorporates the features introduced in the
+$R_2$ is guaranteed to outlive $R_1$.  While the invariant is easily
+stated, enforcing it in the presence of first-class dynamic regions,
+parametric polymorphism (generics), and higher-order functions
+requires new reasoning principles that we formally develop in this
+section. We introduce \fbname (\FB), an explicitly typed core language
+(with region types) that incorporates the features introduced in the
 previous section. \fbname builds on the Featherweight Generic Java
@@
 without region types and with some syntactic sugar.)
-
-% In this section, we formally develop \fbname, a language that
-% incorporates the features introduced in the previous section, as well
-% as its type system, building on the Featherweight Generic Java
-% (FGJ)~\cite{fgj} formalism.  Our development reuses notations and
-% various definitions from~\cite{fgj}, such as the definition of type
-% well-formedness for the core (region-free) language, which are
-% elaborated further in the appendix~\cite{techrep}.
 
@@
 
-A class in \FB is parametric over zero or more type variables (as
-in FGJ) as well as one or more region variables $\rho$.
+A class in \FB is parametric over zero or more type variables (as in
+FGJ) as well as one or more region variables $\rho$.
 % In the above example, the class \C{Pair} is parameterized over
@@
 parameters of a class or a method. In some cases, we also represent
-region parameters as $\rho\rhobar$ to clearly distinguish between
-the allocation region parameter ($\rho$) and the rest.}.
+region parameters as $\rho\rhobar$ to clearly distinguish between the
+allocation region parameter ($\rho$) and the rest.}.
 %
@@
 be type polymorphic, in keeping with C\#, which constitutes the core
-of \name. Separate parameterization over types and regions is needed
-in \FB to support region-polymorphic higher-order functions as class
-fields. For such higher-order class fields, regions are parameterized
-at the (class) field level whereas types are parameterized at the
-class level. This allows, for example, a generic class, whose type
-parameters are $a$ and $b$, to contain a region-polymorphic function of type
+of \name.
+
+Note that classes in \FB are independently parameterized over types
+and regions. While this design decision has a downside that it allows
+type variables to only denote objects contained in a single region, it
+yeilds benefits that far outweigh the costs. In particular, it lets us
+support region-polymorphic higher-order functions as class fields.
+This allows, for example, a generic class, whose type parameters are
+$a$ and $b$, to contain a region-polymorphic function of type
 $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
@@
 (\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
-parameterization separate from region parameterization allows \FB to
-type such examples. This choice also simplifies type inference in the
-presence of higher-order functions.
+and region parameterizations separate also simplifies the type system
+to the point that inference becomes practical
+(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
+type systems suggests, type inference is pivotal to the practical
+utility of sophisticated type systems.
 
@@
 
-\input{fb-staticsem}
-
 Like classes, methods can also exhibit constrained region
 polymorphism.  A method definition in \FB is necessarily polymorphic
-over its allocation context (\S~\ref{sec:alloc-ctxt}), and
-optionally polymorphic with respect to the regions containing its
-arguments.  Region parameters, like those on classes, are qualified
-with constraints ($\phi$).
+over its allocation context (\S~\ref{sec:alloc-ctxt}), and optionally
+polymorphic with respect to the regions containing its arguments
+(i.e., a method has at least one region parameter). Region
+parameters, like those on classes, are qualified with constraints
+($\phi$).
 % Region parameters on the methods, like those on classes, are
@@
 % instantiate and generalize region variables. 
-Since a function closure can escape the context in which it is
-created, it is important to keep track of the region in which it is
-created in order to avoid unsafe dereferences. The $\rgn$ annotation
-above the arrow in the arrow type denotes the allocation region of the
-corresponding closure. 
+A function, like a method, is necessarily polymorphic w.r.t its
+allocation context.  Since a function closure can escape the context
+in which it is created, it is important to keep track of the region in
+which it is created in order to avoid unsafe dereferences. The $\rgn$
+annotation above the arrow in the arrow type denotes the allocation
+region of the corresponding closure. 
 % Note that it is important to distinguish
@@
 of outlives and equality relations. For any well-formed core type $T$,
-$T@\rgn$ is a well-formed region type if $\rgn$ is a valid region. The
-type $\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
+$T@\rgn$ is a well-formed region type if $\rgn$ is a valid region.
+The type $\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
 $\toprgn$ is a special immortal region that outlives every other live
@@
 described below.
+
+\input{fb-staticsem}
+
+% Transferable regions are first-class objects in \name, hence in \FB.
+% The type of a transferable region handler is of the form
+% $\RgnZT{\rgn}$, where, like any other object, $\rgn$ denotes the
+% region in which the handler object is allocated. However, region
+% handlers are often required to escape the scope and address
 
@@
 a constructor's arguments in \FB are same as the names and types of
-its class's fields, and the type rule relies on this fact to typecheck
-constructor applications.
+its class's fields.
 
@@
 to prevent the new region from incorrectly assuming existing outlives
-relationships on an eponymous region. Provided this is satisfied, the
-expression ($e$) under \C{letregion} is then typechecked assuming that
-the new region is live ($\Delta \cup \{\pi\}$) and that it is outlived
-by all existing live regions ($\Delta \outlives \pi$). The result of a
-\C{letregion} expression must have a type that is well-formed under a
-context not containing the new region. This ensures that the value
-obtained by evaluating a \C{letregion} expression contains no
-references to the temporary objects inside the region.
+relationships on an eponymous region. The expression ($e$) under
+\C{letregion} is typechecked with the new region as its allocation
+context, under the assumption that the region is live ($\Delta \cup
+\{\pi\}$) and that it is outlived by all existing live regions
+($\Delta \outlives \pi$). The result of a \C{letregion} expression
+must have a type that is well-formed under a context not containing
+the new region. This ensures that the value obtained by evaluating a
+\C{letregion} expression contains no references to the temporary
+objects inside the region.
 
@@
 inside the transferable region from storing references to those
-outside.  Environment ($\env$) is extended with binding for the type
-of root object while typechecking $e$.
+outside. The newly opened region becomes the allocation context for
+$e$, which is type checked under an environment ($\env$) extended with
+the type binding for the root object.
 
@@
 constraints ($\phi$) are required to be well-formed under $\Delta$
-extended with the function's region parameters ($\rho\rhobar$). Note
-that the closure is always allocated in the current allocation context
-($\rgn$). This prevents the closure from escaping the context in which
-it is created, thus trivially ensuring the safety of any dereferences
-inside the closure.
+extended with the function's region parameters ($\rho\rhobar$).
+Unlike a typical object, a function closure might capture references
+that may become unsafe if the closure escapes its allocation context.
+\FB prevents this scenario by requiring the function closure to always
+be allocated in the current allocation context ($\rgn$).
 
+The type rule for method application uses auxiliary definitions
+$\mtype$ and $\bound$ which originate from the FGJ calculus - the
+former gives the type of a class method and the latter defines the
+bound of a type variable. Method application involves instantiation of
+method's region parameters, and the type rule requires the first
+region parameter to be instantiated with the current allocation
+context ($r$). The type rule for function application does similarly.
+This requirement ensures the safety of closures as demonstrated by the
+following example.
+
+Consider a method $m$ with two region parameters - $\rho_0$ and
+$\rho_1$ (i.e., $m\inang{\rho_0,\rho_1}(\dots)$). Suppose the method
+immediately returns a function closure that contains a reference to
+(an object in) $\rho_1$. Since function closures are always allocated
+in the current allocation context, the closure is allocated in
+$\rho_0$, the allocation context for the method body\footnote{
+Fig.~\ref{fig:fb-staticsem-1} of the Appendix contains the method
+well-formedness rule, which is on the similar lines as the lambda type
+rule shown in Fig.~\ref{fig:fb-staticsem}}. Now, consider the
+following use of the method:
+\begin{codejava}
+letregion R0 in
+  let f = letregion R1 in m<R0,R1>()
+  in f()
+\end{codejava}
+Observe that the first region parameter of \C{m} is instantiated with
+\C{R0}, while the allocation context for the method call is \C{R1}.
+The function \C{f} returned by \C{m} stores a reference to \C{R1},
+which is unsafe when \C{f} is finally called. The type system
+fortunately disallows this scenario by enforcing certain restrictions
+during method and function applications as described above.
 % The type rule for the \C{new} expression is a straightforward

### 2017-01-13 paper/broom.tex
added alloc context to constraint generation rules
type inference minor updates
Formatting edits.
fixed typos in constraint generation rules
Edits to the intro.
Sec.3 edits complete.

@@
 %% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
-\author[1]{Gowthan Kaki}
-\author[2]{G. Ramalingam}
-\author[3]{Kapil Vaswani}
-\author[4]{Dimitrios Vytiniotis}
-\affil[1]{Purdue University}
-\affil[2]{Microsoft Research}
-\affil[3]{Microsoft Research}
-\affil[4]{Microsoft Research}
-\authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
+\author{}
+% \author[1]{Gowthan Kaki}
+% \author[2]{G. Ramalingam}
+% \author[2]{Kapil Vaswani}
+% \author[2]{Dimitrios Vytiniotis}
+% \affil[1]{Purdue University}
+% \affil[2]{Microsoft Research}
+% \authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
 
-\Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+% \Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 E.g., cite as "F.1.1 Models of Computation".
-\keywords{Memory management, regions, type system, type inference}% mandatory: Please provide 1-5 keywords
+\keywords{Memory Safety, Formal Methods, Type System, Type Inference,
+Regions}% mandatory: Please provide 1-5 keywords
 % Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \SeriesVolume{42}
-\ArticleNo{23}
+\ArticleNo{1}
 % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \input{type-inference}
+\input{misc}
 \renewcommand{\rgn}{r}
@@
 \input{related}
+
+\clearpage
 
@@
 \input{appendix}
+\input{typesystem-proofs}
+\input{fb-constraint-generation-extra}
+\renewcommand{\rgn}{\myownthrowexception}
+\input{typeinference-proofs}
+\input{misc}
 

### 2017-01-13 paper/type-system.tex
Sec.3 edits complete.

@@
 and regions. While this design decision has a downside that it allows
-type variables to only denote objects contained in a single region, it
-yeilds benefits that far outweigh the costs. In particular, it lets us
-support region-polymorphic higher-order functions as class fields.
-This allows, for example, a generic class, whose type parameters are
-$a$ and $b$, to contain a region-polymorphic function of type
+type variables to only denote the types of objects contained in a
+single region, it yeilds benefits that far outweigh the costs. In
+particular, it lets us support region-polymorphic higher-order
+functions as class fields.  This allows, for example, a generic class,
+whose type parameters are $a$ and $b$, to contain a region-polymorphic
+function of type
 $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
@@
 
+\input{fb-staticsem}
+
 \noindent The well-formedness judgment on region types
@@
 described below.
-
-\input{fb-staticsem}
 
@@
 constructor to be called with a nullary function that returns a value
-in its allocation context. 
-% It enforces this by typechecking the body ($e$) of the function
-% under an empty context containing nothing but the allocation context
-% of the function. 
-This ensures that the value returned by the function stores no
-references to objects allocated elsewhere, including the top region
-($\toprgn$), thus preventing cross-region references originating from
-transferable regions. The body of the function might however create
-new regions while execution, but this is not a problem as long as such
-regions, and objects allocated in them, don't find their way into the
-result of its evaluation.
+in its allocation context.  This ensures that the value returned by
+the function stores no references to objects allocated elsewhere,
+including the top region ($\toprgn$), thus preventing cross-region
+references originating from transferable regions. The body of the
+function might however create new regions while execution, but this is
+not a problem as long as such regions, and objects allocated in them,
+don't find their way into the result of its evaluation.
+
+% \begin{figure}[!t]
+%   \input{fb-morewfrules}
+%   \caption{\fbname: method and class well-formedness rules}
+%   \label{fig:fb-morewfrules}
+% \end{figure}
 
@@
 in the current allocation context, the closure is allocated in
-$\rho_0$, the allocation context for the method body\footnote{
-Fig.~\ref{fig:fb-staticsem-1} of the Appendix contains the method
-well-formedness rule, which is on the similar lines as the lambda type
-rule shown in Fig.~\ref{fig:fb-staticsem}}. Now, consider the
-following use of the method:
+$\rho_0$, the allocation context for the method body (the rule for
+methods discussed below). Now, consider the following use of the
+method:
 \begin{codejava}
@@
 
+The method well-formedness rule makes use of an auxiliary definition
+$\override$ that judges whether the current method is a valid
+overriding of any eponymous method from the super class. The type
+environment is extended with a binding that binds the $\thisZ$ keyword
+to the type of the current class. Note that the the type of the
+current method as accessed via $\thisZ$ (i.e., $\thisZ.m$) is
+region-parametric, thus admitting region-polymorphic recursion (i.e.,
+recursive call to $m$ can have different region arguments than $m$).
+
 \subsection{Operational Semantics and Type Safety}
@@
 
-Fig.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} in the appendix
-defines a small-step operational semantics for \fbname via a
-four-place reduction relation:
+\input{fb-syntax-extra}
+
+The operational semantics of \FB describes a non-trivial runtime
+component that introduces memory regions and performs run-time
+verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
+additional langauge constructs of \FB that manifest only at run-time.
+A location ($\loc$) abstracts all the memory locations associated with
+a region (i.e., each memory region is associated with a single
+location). A transferable region can be in one of three possible
+states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
+($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
+manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
+while allocating a (static) region, and \C{open} reduces to \C{opened}
+while opening a (transferable) region. An \C{opened}
+expression is tagged with the typestate ($s$) of the transferable
+region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
+transferable region can be \C{open}'d from either an open state or a
+closed state). Special value $\bot$ denotes an exception.
+
+Operational semantics of \FB defines a four-place small-step reduction
+relation of the form shown below:
 % \vspace*{-0.15in}
@@
 \end{smathpar}
-Operational semantics replaces abstract region annotations ($\pi$ and
-$\rho$) with \emph{locations} ($\loc$), where each \emph{location}
-abstracts all the memory locations associated with a region (i.e.,
-each memory region is associated with a single location). The
-typestate of regions is tracked by a finite map ($\mem$) from
+The typestate of regions is tracked by a finite map ($\mem$) from
 locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
@@
 discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
-value ($\invalidexn$).
+value ($\invalidexn$). An illustrative subset of operational semantics
+that formalize the intuitions desribed above is shown in
+Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix
+contains rest of the rules.
 
-Replacing $\pi$'s and $\rho$'s with $\loc$'s at runtime results in a
-new class of expressions that are not captured in
-Fig.~\ref{fig:fb-syntax}. Appendix contains full syntax that includes
-such expressions. To help state the type safety theorem, we define the
-syntactic class of (runtime) values:
+\input{fb-opsem}
+
+To help state the type safety theorem, we define the syntactic class
+of (runtime) values:
 \begin{smathpar}
@@
 
-% Furthermore, we prove the following theorem about \FB, which, in
-% conjunction with the type safety theorem, implies the safety of region
-% transfers across address spaces:
-% \begin{theorem}
-% \emph{(\textbf{Transfer Safety})}
-% \label{thm:fb-transfer-safety}
-% $\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\mem$, $\mem'$, $\phicx$,
-% $\phicx'$, such that $\loc \not\in dom(\mem')$, if $\hastyp{\emptyA, \cdot}{\C{new}\;
-% \RgnZT{\toploc\loc}(v)}{\RgnZT{\toprgn}}$, then
-% $\hastyp{\rhoenv',\cdot,\phicx')}
-% {\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
-% \end{theorem}
-% The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
-% is well-typed under one context, then it is also well-typed under
-% every other context, whose $\mem$ maps $\rgn_i$ to closed
-% ($\CLOSED$) state.  Thus, a recipient of a transferable region only
-% needs to add a binding for the region to its $\mem$ in order to
-% preserve its type safety. 
-% Notably, this result could not have been
-% established if it were possible for a transferable region to contain
-% references to objects outside the region.
+{\bf A note on region handlers}. \FB stores the region handlers in a
+distinguished top region, which abstracts the GC heap. Type system
+prevents transferable regions from storing references into the GC
+heap. A static region however can store references to region handles,
+which need to be taken into account while GC'ing the heap.  Traversing
+the region memory to identify references into the GC heap
+unfortunately defeats the purpose of regions. A solution is to disable
+collecting region handles as long as any static region is open. Since
+static regions are intended to store temporary objects while
+populating a dynamic region, their lifetimes are short enough to let
+region handles to be GC’d.

### 2017-01-14 paper/related.tex
Edits to appendix. CR04 added as related work.

@@
 scoped (stack) regions (as a replacement for GC).
-Our type inference is analogous to their's in some respects,
+Our type inference is analogous to theirs in some respects,
 while differing in others.
@@
 
-% Although there is some support for local
-% type inference, region types for methods and classes need to be
-% written explicitly. In contrast, our region type inference that
-% eliminates any such need.
-
-~\cite{HMN01} proposes a flow-sensitive approach for first-order programs
-% to overcome some of the drawbacks of~\cite{tofte97}
-to generalize~\cite{tofte97}'s approach to dynamic regions.
-However, dynamic regions are still not first-class values of the language,
-and reference counting is needed to ensure memory safety.
+~\cite{HMN01} proposes a flow-sensitive approach for first-order
+programs to generalize~\cite{tofte97}'s approach to dynamic
+regions.~\cite{CR04} describes a flow-insensitive and
+context-sensitive analysis that transforms Java programs to use
+(dynamic) regions.  However, neither~\cite{HMN01} nor~\cite{CR04}
+supports dynamic regions as first-class objects; they cannot be stored
+in data structures or passed to methods. Furthermore,
+while~\cite{HMN01} requires reference counting to ensure memory
+safety,~\cite{CR04} comes with no formal safety guarantees.
 %

### 2017-01-14 paper/type-system.tex
Edits to appendix. CR04 added as related work.

@@
 type variables to only denote the types of objects contained in a
-single region, it yeilds benefits that far outweigh the costs. In
+single region, it yields benefits that far outweigh the costs. In
 particular, it lets us support region-polymorphic higher-order
@@
 objects of other classes. The rule for the latter relies on an
-auxiliary definition called $\fields$ (undefined for $\RgnZ$ class)
-that returns the sequence of type bindings for fields (instance
-variables) of a given class type.  Like in FGJ, the names and types of
-a constructor's arguments in \FB are same as the names and types of
-its class's fields.
+auxiliary definition\footnote{All auxiliary definitions we use in this
+exposition originate from the FGJ calculus.} called $\fields$
+(undefined for $\RgnZ$ class) that returns the sequence of type
+bindings for fields (instance variables) of a given class type.  Like
+in FGJ, the names and types of a constructor's arguments in \FB are
+same as the names and types of its class's fields. The type rule for
+the field access expression ($e.f_i$) also uses $\fields$ and another
+definition called $\bound$, which returns the bound of a type variable
+($\bound$ is an identity function for concrete types).
 
@@
 
-The type rule for method application uses auxiliary definitions
-$\mtype$ and $\bound$ which originate from the FGJ calculus - the
-former gives the type of a class method and the latter defines the
-bound of a type variable. Method application involves instantiation of
-method's region parameters, and the type rule requires the first
-region parameter to be instantiated with the current allocation
-context ($r$). The type rule for function application does similarly.
-This requirement ensures the safety of closures as demonstrated by the
-following example.
+The type rule for method application uses an auxiliary definition
+$\mtype$ that gives the type of a class method. Method application
+involves instantiation of method's region parameters, and the type
+rule requires the first region parameter to be instantiated with the
+current allocation context ($r$). The type rule for function
+application does similarly.  This requirement ensures the safety of
+closures as demonstrated by the following example.
 
@@
 to the type of the current class. Note that the the type of the
-current method as accessed via $\thisZ$ (i.e., $\thisZ.m$) is
+current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
 region-parametric, thus admitting region-polymorphic recursion (i.e.,
-recursive call to $m$ can have different region arguments than $m$).
+recursive call to $\C{m}$ can have different region arguments than
+$\C{m}$).
 
@@
 verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
-additional langauge constructs of \FB that manifest only at run-time.
+additional language constructs of \FB that manifest only at run-time.
 A location ($\loc$) abstracts all the memory locations associated with
@@
 value ($\invalidexn$). An illustrative subset of operational semantics
-that formalize the intuitions desribed above is shown in
+that formalize the intuitions described above is shown in
 Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix

### 2017-01-14 paper/typesystem-proofs.tex
Edits to appendix. CR04 added as related work.

@@
 
-Consistency between $\Delta$ and $\mem$ is preserved by the reduction
-relation:
+The rationale behind the $\consistent$ relation is to establish the
+correspondence between the concrete memory ($\mem$) of the operational
+semantics and its abstraction ($\Delta$) used by the static semantics.
+We now prove that consistency between $\Delta$ and $\mem$ is preserved
+by the reduction relation:
 
@@
 \end{itemize} 
-\qed
+%\qed
 \end{proof}
@@
 lines of similar proof for FGJ.
-\qed
+%\qed
 \end{proof}
@@
 \end{itemize}
-\qed
+%\qed
 \end{proof}
@@
   \C{letd} case.
-  \qed
+  %\qed
   \end{itemize}
@@
 and~\ref{lem:preservation}.
-\qed
+%\qed
 \end{proof}

### 2017-03-22 paper/ecoop_rebuttal.txt
ECOOP rebuttal committed.

We thank the reviewers for their extensive review of our work, and
thoughtful suggestions for improvement.

1. Why are closures treated specially? Why don't classes suffice?
It is true that C# treats lambdas as objects of class `Func`. A flip
side of such punning is that C# classes cannot have type-polymorphic
lambdas as class fields (since all fields have monomorphic types).
Adopting C#'s approach in the region type system would lead to a
similar drawback: we can no longer support region-polymorphic lambdas
as class fields, but this feature is used frequently in practice. An
example is presented in Fig.3 (and explained in 3rd paragraph on p.8).

2. We are summarily in agreement with reviewer 24A's comparison with
HM. Separation of type and region parameters is an example of the
language aspect designed with inference in mind. "Action at distance
errors" do occur in our case, and we conjecture that the recent
advances[oopsla15,pldi15] could help us migitage the problem.

3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
system. The type system enforces two invariants on every well-typed
object: (a). its type signature explicitly captures the regions to
which its fields refer to, and (b). all such regions must outlive the
allocation region of the object. The discussion on constrained region
polymorphism (p.8) covers this informally, but formal enforcement is
done by the class well-formedness rule, which was relegated to the
appendix (Fig. 13). In the context of reviewer’s example, if
`b.boxref` is a reference to an object in region `R2`, and if `b`
itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
instantiate `R1` with `x` (rather, it’s region idenfitier), and `R2`
with `y`, but note that there does not exist any outlives relationship
between R1 and R2 (since both are transferable regions). Thus, this
instantiation is disallowed, effectively preempting the possibility of
an unsafe reference.

4. Reviewer 24B's lambda example is (incorrectly) accepted by the type
system. The reason for this is that the closure type rule of Fig. 6 is
missing the premise `Δ ≥ r` (Δ outlives r). We had this premise in the
previous draft of the paper, but got rid of it during the days leading
to the submission, perhaps reasoning (incorrectly) that it is a
tautology. We regret this decision. Restoring this premise would
prevent the example from type checking since `x` doesn't outlive `y`.

5. It is true that the FB^+ language used in the evaluation is
improverished in comparison with Java/C#, but is is enough to capture
the aspects of Naiad operators relevant to the region type system.
There are a total of 6 Naiad streaming query operators, out of which
we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
are aggregation operators, which do not contain any interesting use
of regions. Instances of the safety violation described in the paper
were found in all 4, and the violation can manifest in an actual
execution in in each case. We accept the reviewer's suggestion to
include these details in the paper.

### 2017-03-22 paper/ecoop_rebuttal.txt
Rama's changes

@@
 
-2. We are summarily in agreement with reviewer 24A's comparison with
-HM. Separation of type and region parameters is an example of the
-language aspect designed with inference in mind. "Action at distance
-errors" do occur in our case, and we conjecture that the recent
-advances[oopsla15,pldi15] could help us migitage the problem.
+2. We agree with reviewer 24A's comparison with HM. Separation of type
+and region parameters is an example of the language aspect designed with
+inference in mind. "Action at distance errors" do occur in our case,
+and we conjecture that the recent advances[oopsla15,pldi15] could help
+us migitage the problem.
 
@@
 system. The type system enforces two invariants on every well-typed
-object: (a). its type signature explicitly captures the regions to
-which its fields refer to, and (b). all such regions must outlive the
+object: (a) Its type signature explicitly captures the regions to
+which its fields refer to, and (b) All such regions must outlive the
 allocation region of the object. The discussion on constrained region
 polymorphism (p.8) covers this informally, but formal enforcement is
-done by the class well-formedness rule, which was relegated to the
+done by the class well-formedness rule, which was (unfortunately) relegated to the
 appendix (Fig. 13). In the context of reviewer’s example, if
@@
 
-4. Reviewer 24B's lambda example is (incorrectly) accepted by the type
-system. The reason for this is that the closure type rule of Fig. 6 is
-missing the premise `Δ ≥ r` (Δ outlives r). We had this premise in the
-previous draft of the paper, but got rid of it during the days leading
-to the submission, perhaps reasoning (incorrectly) that it is a
-tautology. We regret this decision. Restoring this premise would
-prevent the example from type checking since `x` doesn't outlive `y`.
+4. (Lambda allocation regions) Reviewer 24B's intuition and points in
+this regard are correct. The closure type rule of Fig. 6 is
+unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
+Adding this premise will ensure that the reviewer's example is correctly handled
+(since `x` doesn't outlive `y`). And it is also correct that with this
+premise, the allocation of lambda does not have to be restricted to the
+allocation region.
+
+We had an equivalent premise in the previous draft of the paper, but got rid
+of it during the days leading to the submission, with a faulty simplification
+considering only allocation regions. We apologize for the confusion caused by this mistake.
 

### 2017-03-22 paper/ecoop_rebuttal.txt
Minor fix
ECOOP rebuttal edited.

@@
 inference in mind. "Action at distance errors" do occur in our case,
-and we conjecture that the recent advances[oopsla15,pldi15] could help
+and we conjecture that the recent advances[oopsla14,pldi15] could help
 us migitage the problem.
@@
 unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
-Adding this premise will ensure that the reviewer's example is correctly handled
+With this added premise, the reviewer's example is correctly handled
 (since `x` doesn't outlive `y`). And it is also correct that with this
@@
 
-We had an equivalent premise in the previous draft of the paper, but got rid
-of it during the days leading to the submission, with a faulty simplification
-considering only allocation regions. We apologize for the confusion caused by this mistake.
+We had an equivalent premise in the previous draft of the paper, but
+got rid of it during the days leading to the submission, with a faulty
+simplification considering only allocation regions. We apologize for
+the confusion caused by this mistake. Thanks to the reviewer for
+catching this.
 
@@
 include these details in the paper.
+
+6. In the definition of C-decomposable, why only one of the regions is
+required to be in Δ? 
+The asymmetry in this definition reflects properties of the
+constraint-generation, namely that the context Φ is modified only by
+adding conjuncts of the form `Δ ≥ π` where `π` is not a
+region-parameter (see the rule LetRegion), as discussed in the proof
+of Lemma 3 in Appendix C.
+
+7. In the explanation of the type safety result, the statement "the
+theorem also captures the key property of operational semantics that
+no live region is ever freed or transferred" is informal for "if
+`consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
+(Lemma 8 of the appendix). Thus, operational semantics in fact never
+frees or transfers a live region. We however understand that the
+informal statement can be misconstrued as stronger guarantees provided
+by the type statement, and accept the reviewer's suggestion to include
+a clarification.
+
+### References ###
+
+[oopsla14]: Z Pavlinovic, Finding minimum type error sources, OOPSLA'14
+[pldi15]: D Zhang et al, Diagnosing type errors with class, PLDI'15

### 2017-03-22 paper/ecoop_rebuttal.txt
Edits to rebuttal.

@@
 and we conjecture that the recent advances[oopsla14,pldi15] could help
-us migitage the problem.
+us mitigate the problem.
 
@@
 is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
-instantiate `R1` with `x` (rather, it’s region idenfitier), and `R2`
+instantiate `R1` with `x` (rather, it’s region identifier), and `R2`
 with `y`, but note that there does not exist any outlives relationship
@@
 5. It is true that the FB^+ language used in the evaluation is
-improverished in comparison with Java/C#, but is is enough to capture
+impoverished in comparison with Java/C#, but is is enough to capture
 the aspects of Naiad operators relevant to the region type system.
@@
 informal statement can be misconstrued as stronger guarantees provided
-by the type statement, and accept the reviewer's suggestion to include
+by the type system. We accept the reviewer's suggestion to include
 a clarification.

### 2017-03-22 paper/ecoop_rebuttal.txt
Edit.

@@
 side of such punning is that C# classes cannot have type-polymorphic
-lambdas as class fields (since all fields have monomorphic types).
-Adopting C#'s approach in the region type system would lead to a
-similar drawback: we can no longer support region-polymorphic lambdas
-as class fields, but this feature is used frequently in practice. An
-example is presented in Fig.3 (and explained in 3rd paragraph on p.8).
+lambdas as class fields (since all fields have monomorphic types,
+i.e., there cannot be any quantifiers).  Adopting C#'s approach in the
+region type system would lead to a similar drawback: we can no longer
+support region-polymorphic lambdas as class fields, but this feature
+is used frequently in practice. An example is presented in Fig.3 (and
+explained in 3rd paragraph on p.8).
 

### 2017-03-22 paper/ecoop_rebuttal.txt
Edits.

@@
 
+8. Why do we need constraint based lifetime parameters for classes?
+What are the distinctive examples which this design enables which a
+simpler ownership system cannot provide? 
+Transferable regions are self-contained, containing no objects with
+pointers to objects in other regions. However, stack regions are more
+flexible, and are permitted to contain objects with pointers to
+objects in other regions. Stack regions are very useful as "working
+storage" (paragraph 3 on p.3). For example, consider a dataflow
+computation that indexes the intermediary results obtained from an
+upstream actor in a transferable region (`R1`), before updating the
+results, and transferring them to a downstream actor. The index,
+perhaps a B-tree, is an auxiliary structure that need not be
+transferred, hence can be stored in a stack region (`R2`). Outlives
+constraints are needed to admit safe references from `R2` into `R1`.
+
 ### References ###

### 2017-03-22 paper/ecoop_rebuttal.txt
Rebuttal edited.

@@
 
-1. Why are closures treated specially? Why don't classes suffice?
+1. **Why are closures treated specially? Why don't classes suffice?**
 It is true that C# treats lambdas as objects of class `Func`. A flip
@@
 region type system would lead to a similar drawback: we can no longer
-support region-polymorphic lambdas as class fields, but this feature
-is used frequently in practice. An example is presented in Fig.3 (and
-explained in 3rd paragraph on p.8).
+support region-polymorphic lambdas as class fields, but this
+functionality is required in practice to support user-defined
+functions (UDFs) in LINQ queries. An example is presented in Fig.3
+(and explained in 3rd paragraph on p.8).
 
-2. We agree with reviewer 24A's comparison with HM. Separation of type
-and region parameters is an example of the language aspect designed with
-inference in mind. "Action at distance errors" do occur in our case,
-and we conjecture that the recent advances[oopsla14,pldi15] could help
-us mitigate the problem.
+2. We agree with reviewer 24A's **comparison with HM**. Separation of type
+and region parameters is an example of the language aspect designed
+with inference in mind. "Action at distance errors" do occur in our
+case, and we conjecture that the recent advances
+[[oopsla14][oopsla14],[pldi15][pldi15]] could help us mitigate the
+problem.
 
-3. Reviewer 24B's IntBox example is (rightfully) rejected by the type
-system. The type system enforces two invariants on every well-typed
-object: (a) Its type signature explicitly captures the regions to
-which its fields refer to, and (b) All such regions must outlive the
-allocation region of the object. The discussion on constrained region
-polymorphism (p.8) covers this informally, but formal enforcement is
-done by the class well-formedness rule, which was (unfortunately) relegated to the
-appendix (Fig. 13). In the context of reviewer’s example, if
-`b.boxref` is a reference to an object in region `R2`, and if `b`
-itself (an `IntBoxBox`) is allocated in `R1`, then the type signature
-is `IntBoxBox<R1,R2 | R2≥R1>`. The example (implicitly) tries to
-instantiate `R1` with `x` (rather, it’s region identifier), and `R2`
-with `y`, but note that there does not exist any outlives relationship
-between R1 and R2 (since both are transferable regions). Thus, this
-instantiation is disallowed, effectively preempting the possibility of
-an unsafe reference.
+3. Reviewer 24B's **`IntBox` example** is (rightfully) rejected by the
+type system. The type system enforces two invariants on every
+well-typed object: (a) Its type signature explicitly captures the
+regions to which its fields refer to, and (b) All such regions must
+outlive the allocation region of the object. The discussion on
+constrained region polymorphism (p.8) covers this informally, but
+formal enforcement is done by the class well-formedness rule, which
+was (unfortunately) relegated to the appendix (Fig. 13). In the
+context of reviewer’s example, if `b.boxref` is a reference to an
+object in region `R2`, and if `b` itself (an `IntBoxBox`) is allocated
+in `R1`, then the type signature is `IntBoxBox<R1,R2 | R2≥R1>`. The
+example (implicitly) tries to instantiate `R1` with (the region
+identifier of) `x`, and `R2` with `y`, but note that there does not
+exist any outlives relationship between R1 and R2 (since both are
+transferable regions). Thus, this instantiation is disallowed,
+effectively preempting the possibility of an unsafe reference.
 
-4. (Lambda allocation regions) Reviewer 24B's intuition and points in
-this regard are correct. The closure type rule of Fig. 6 is
-unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ outlives r).
-With this added premise, the reviewer's example is correctly handled
-(since `x` doesn't outlive `y`). And it is also correct that with this
-premise, the allocation of lambda does not have to be restricted to the
+4. **Lambda allocation regions**: Reviewer 24B's intuition and points
+in this regard are correct. The closure type rule of Fig. 6 is
+unfortunately missing a premise, namely `Δ ≥ r` (every region in Δ
+outlives r). The premise requires a closure to be well-typed under
+only the regions that outlive its allocation region. With this added
+condition, the reviewer's example is correctly handled (since `x`
+doesn't outlive `y`). And it is also correct that with this premise,
+the allocation of lambda does not have to be restricted to the
 allocation region.
-
 We had an equivalent premise in the previous draft of the paper, but
@@
 simplification considering only allocation regions. We apologize for
-the confusion caused by this mistake. Thanks to the reviewer for
-catching this.
+the confusion caused by this mistake. We are grateful to the reviewer
+for pointing it out.
 
-5. It is true that the FB^+ language used in the evaluation is
+5. It is true that the FB^+ **language used in the evaluation** is
 impoverished in comparison with Java/C#, but is is enough to capture
@@
 There are a total of 6 Naiad streaming query operators, out of which
-we analyzed 4 (Select, SelectMany, Join and WindowJoin). The other two
-are aggregation operators, which do not contain any interesting use
-of regions. Instances of the safety violation described in the paper
-were found in all 4, and the violation can manifest in an actual
-execution in in each case. We accept the reviewer's suggestion to
-include these details in the paper.
+we analyzed 4 (Select, Join, GroupBy, and Aggregate). The other two
+are not interesting from the region type system perspective. Instances
+of the safety violation described in the paper were found in all 4,
+and the violation can manifest in an actual execution in in each case.
+We accept reviewer 24B's suggestion to include these details in the
+paper. To answer reviewer 24D's question of whether the need to clone
+objects negates the benefits of regions, we believe the answer is no.
+As explained in paragraph 2 on p.1 (citing [7]), the savings obtained
+by using regions are primarily because of (a). avoiding multiple older
+generation collections over long-lived objects, and (b). preventing
+cascading effect of long GC pauses on the latency at downstream actors.
+Both these factors are unaffected by cloning, which happens only once
+per objects (as against multiple GC walks over a large data structure
+that aggregates objects).
 
-6. In the definition of C-decomposable, why only one of the regions is
-required to be in Δ? 
+6. **In the definition of C-decomposable, why only one of the regions is
+required to be in Δ?**
 The asymmetry in this definition reflects properties of the
-constraint-generation, namely that the context Φ is modified only by
+constraint-generation, namely that the context `Φ` is modified only by
 adding conjuncts of the form `Δ ≥ π` where `π` is not a
@@
 
-7. In the explanation of the type safety result, the statement "the
-theorem also captures the key property of operational semantics that
-no live region is ever freed or transferred" is informal for "if
+7. In the **explanation of the type safety result**, the statement
+"the theorem also captures the key property of operational semantics
+that no live region is ever freed or transferred" is informal for "if
 `consistent(Δ,Σ)` and `(e,Σ)-->(e',Σ')` then `consistent(Δ,Σ')`"
-(Lemma 8 of the appendix). Thus, operational semantics in fact never
-frees or transfers a live region. We however understand that the
-informal statement can be misconstrued as stronger guarantees provided
-by the type system. We accept the reviewer's suggestion to include
-a clarification.
+(included in the type safety statement, and proved as Lemma 8 in the
+appendix). Thus, operational semantics in fact never frees or
+transfers a live region. We however understand that the informal
+statement can be misconstrued as stronger guarantees provided by the
+type system. We accept the reviewer's suggestion to include a
+clarification.
 
-8. Why do we need constraint based lifetime parameters for classes?
+8. **Why do we need constraint based lifetime parameters for classes?
 What are the distinctive examples which this design enables which a
-simpler ownership system cannot provide? 
-Transferable regions are self-contained, containing no objects with
-pointers to objects in other regions. However, stack regions are more
-flexible, and are permitted to contain objects with pointers to
-objects in other regions. Stack regions are very useful as "working
-storage" (paragraph 3 on p.3). For example, consider a dataflow
-computation that indexes the intermediary results obtained from an
-upstream actor in a transferable region (`R1`), before updating the
-results, and transferring them to a downstream actor. The index,
-perhaps a B-tree, is an auxiliary structure that need not be
-transferred, hence can be stored in a stack region (`R2`). Outlives
-constraints are needed to admit safe references from `R2` into `R1`.
+simpler ownership system cannot provide?**  Transferable regions are
+self-contained, containing no objects with pointers to objects in
+other regions. However, stack regions are more flexible, and are
+permitted to contain objects with pointers to objects in other
+regions. Stack regions are very useful as "working storage" (paragraph
+3 on p.3). For example, consider a dataflow computation that indexes
+the intermediary results obtained from an upstream actor in a
+transferable region (`R1`), before updating the results, and
+transferring them to a downstream actor. The index, perhaps a B-tree,
+is an auxiliary structure that need not be transferred, hence can be
+stored in a stack region (`R2`). Outlives constraints are needed to
+admit safe references from `R2` into `R1`.
 
-### References ###
-
-[oopsla14]: Z Pavlinovic, Finding minimum type error sources, OOPSLA'14
-[pldi15]: D Zhang et al, Diagnosing type errors with class, PLDI'15
+[oopsla14]: http://dl.acm.org/citation.cfm?id=2660230
+[pldi15]: http://dl.acm.org/citation.cfm?id=2738009

### 2018-03-20 paper/ecoop18_rebuttal.txt
Ecoop18-related changes and the rebuttal.

We thank the reviewers for their reviews and suggestions for improvement.

Review #50A
------------

1. The other two Naiad benchmarks that were left out of evaluation
were so left out because they are very similar to the existing
benchmarks insofar as region type system is concerned. In particular,
the Naiad vertex `RegionWindowJoinVertex` is similar to
`RegionJoinVertex`, and `RegionSelectManyVertex` is similar to
`RegionSelectVertex`. They were nonetheless included in the evaluation
of [7] because they allow the runtime overhead of regions to be
measured at scale (allowing [7] to make a case for regions from the
performance standpoint, as opposed to the safety standpoint). We shall
include this note in the final version of the paper.

Review #50B
-------------

1. Dealing with run-time errors: Run-time exceptions are raised either
when the program attempts to transfer/free an open region, or open a
transferred/freed region (The operational semantics (Figs. 15 and 16)
precisely identify these two cases). The former is often a mere
programming oversight, which can be rectified by moving the
`transfer`/`free` out of the lexical scope of `open`. The latter could
however denote a genuine bug in the program, where a dynamic region is
assumed to be always available at some program point, when in fact
it could have been freed/transferred along a program path leading to
that point. The solution in this case is to fix the program logic by
acknowledging that the region may no longer be available at that
program point.

Review #50C
-----------

1. How the type system differentiates between `free` and `transfer`:
The type system doesn't. Dynamic regions in Broom are primarily
intended to transfer data among actors (hence the name "Transferable
Regions"). It is therefore expected that a dynamic region is
transferred at least once during its lifetime (before it is freed),
thereby making it unsafe to allow outgoing references from the region
to other regions.  While nothing prevents a transferable region to be
used merely as dynamic memory, that is not an intended usecase for
which Broom is optimized.

2. Serializing: Broom's region runtime [7] serializes data if the
receiver of a transferable region is remote.

3. Our model does admit aliases (through the `let` expression). The
operational semantics defines a a *store* (Σ) that maps locations to
their typestates, and implicitly models a heap at the granularity of a
region (each location $l$ denotes a region, and every object whose
first region parameter is $l$ (e.g., `new B<l,..>)(...)`) is allocated
in the region at $l$). Mutable object fields constitute an ostensible
omission from our model (also the FGJ model), but this isn't a major
shortcoming considering that constructor application (during the new
object creation) includes assignments to object fields. We would
include a detailed note to this extent in the final version.

4. As far as the syntax is concerned, `transfer` and `free` are like
any other methods, hence their application is modeled by the generic
method application expression ($e.m<...>(...)$). Since the operational
semantics models a single actor (and not a system of actors), the
behavior of `free` is same as that of `transfer`: both result in the
region becoming unavailable to the actor calling `transfer`/`free`.
The rule for `transfer` captures this semantics.

5. The run-time checks are precisely captured by the premises of
the opsem rules that deal with exceptions (namely, OpenTransferred and
TransferOpened). The checks basically check whether the action being
performed (e.g., `transfer`) is allowed given the current typestate of
the region (e.g., "live"), raising an exception otherwise.

6. `inRgn.free()` in Fig. 3 is needed to prevent a memory leak.
Reference going out of scope doesn't automatically free a region,
because the region may have been aliased. Naiad code does contain
`free` statements.

### 2018-03-20 paper/implementation.tex
Ecoop18-related changes and the rebuttal.

@@
 was able to infer that the list and its data (of type \C{T}) can be
-allocated in different regions, as long as the later outlives the
+allocated in different regions, as long as the latter outlives the
 former. This allows, for instance, a \C{preOrder} method to traverse a
@@
 
-Next, we translated some of the Naiad streaming query operator
+Next, we translated 4 out of the 6 Naiad streaming query operator
 benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
-$\absof{\FB^+}$, and used \namec to verify their safety.
-The number of LOC performing operations on \C{Region} objects relative
-to the total LOC is 8\% or under in the Naiad benchmarks. 
-During the process, we found multiple instances of potential memory safety
-violations in the $\absof{FB^+}$ translation of benchmarks, which we
-verified to be present in the original C\# implementation as well. The
-cause of all safety violations is the creation of a reference from the
-outgoing message (a transferable region) to the payload of the
-incoming message. For example, the implementation of \C{SelectVertex}
-contains the following:
+$\absof{\FB^+}$, and used \namec to verify their safety (The remaining
+two Naiad vertices are not interesting from the type system
+perspective).  The number of LOC performing operations on \C{Region}
+objects relative to the total LOC is 8\% or under in the Naiad
+benchmarks.  During the process, we found multiple instances of
+potential memory safety violations in the $\absof{FB^+}$ translation
+of all the 4 Naiad vertices, which we verified to be present in the
+original C\# implementation as well. The cause of all safety
+violations is the creation of a reference from the outgoing message (a
+transferable region) to the payload of the incoming message. For
+example, the implementation of \C{SelectVertex} contains the
+following:
 \begin{codejava}

### 2018-03-20 paper/introduction.tex
Sec.3 edits complete.
Ecoop18-related changes and the rebuttal.

@@
 
-\input{motivating-eg}
-
 \mypara{Performance.}
@@
 the absence of language support.
+
+\input{motivating-eg}
 
@@
 will remain live as long as the stack region is live. We say that the
-former \emph{outlives} the later., and any references from the stack
-region to the transferable region are therefore safe.
-This is particular useful, since such stack regions serve as temporary
-working storage while working with the (opened) transferable regions.
+former \emph{outlives} the latter, and any references from the stack
+region to the transferable region are therefore safe. This is
+particular useful, since such stack regions serve as temporary working
+storage while working with the (opened) transferable regions.
 
@@
   provided that certain typestate invariants on regions hold.  The
-  later invariants are enforced via simple runtime checks.
+  latter invariants are enforced via simple runtime checks.
 

### 2018-03-20 paper/related.tex
Ecoop18-related changes and the rebuttal.

@@
 An ownership type system for safe region-based memory management in
-real-time Java has been proposed by~\cite{MIT03}.  
+real-time Java has been proposed by Boyapati \emph{et al.}~\cite{MIT03}.  
 % Like us, they too assume a source language with programmer-managed
@@
 % development, and our type system bears some similarities to theirs.
-Our language also admits generics and higher-order functions.
-We also establish type safety and transfer safety results that
-formalize the guarantees provided by our system. While~\cite{MIT03}'s
+Our language also admits generics and higher-order functions.  We also
+establish type safety and transfer safety results that formalize the
+guarantees provided by our system. While Boyapati \emph{et al.}'s
 language is explicitly typed, our language comes equipped with full
-type inference.
-However, several inference algorithms have been proposed in the context
-of other ownership type systems.
-Our type inference algorithm is also novel compared to these existing
-ownership inference algorithms, which are based on, e.g.,
-pointer analysis~\cite{HuangEtAl:ECOOP12} or boolean satisfiability~\cite{DietlEtAl:ECOOP11}.
-(See Section 5.2 of~\cite{OwnershipSurvey} for a more comprehensive discussion
-of ownership inference algorithms.)
-Some distinguishing characteristics of our algorithm is that it is customized
-to our problem, does not use any pointer analysis algorithm (which
-can be a source of imprecision) or SAT solvers (which can be a source of inefficiency),
-and comes with relative completeness guarantees.
+type inference.  However, several inference algorithms have been
+proposed in the context of other ownership type systems.  Our type
+inference algorithm is also novel compared to these existing ownership
+inference algorithms, which are based on, e.g., pointer
+analysis~\cite{HuangEtAl:ECOOP12} or boolean
+satisfiability~\cite{DietlEtAl:ECOOP11}.  (See Section 5.2 of Clark
+\emph{et al.}'s survey of ownership type
+systems~\cite{OwnershipSurvey} for a more comprehensive discussion of
+ownership inference algorithms.) Some distinguishing characteristics
+of our algorithm is that it is customized to our problem, does not use
+any pointer analysis algorithm (which can be a source of imprecision)
+or SAT solvers (which can be a source of inefficiency), and comes with
+relative completeness guarantees.
 
-~\cite{HMN01} proposes a flow-sensitive approach for first-order
-programs to generalize~\cite{tofte97}'s approach to dynamic
-regions.~\cite{CR04} describes a flow-insensitive and
-context-sensitive analysis that transforms Java programs to use
-(dynamic) regions.  However, neither~\cite{HMN01} nor~\cite{CR04}
-supports dynamic regions as first-class objects; they cannot be stored
-in data structures or passed to methods. Furthermore,
-while~\cite{HMN01} requires reference counting to ensure memory
-safety,~\cite{CR04} comes with no formal safety guarantees.
+Henglein \emph{et al.}~\cite{HMN01} propose a flow-sensitive approach
+for first-order programs to generalize Tofte and Talpin's approach to
+dynamic regions. Cherem and Rugina~\cite{CR04} describe a
+flow-insensitive and context-sensitive analysis that transforms Java
+programs to use (dynamic) regions.  However, neither of themr supports
+dynamic regions as first-class objects; they cannot be stored in data
+structures or passed to methods. Furthermore, while Henglein \emph{et
+al.}~\cite{HMN01} require reference counting to ensure memory safety,
+Cherem and Rugina's analysis~\cite{CR04} comes with no formal safety
+guarantees.
 %
-\cite{gpu14} uses regions to safely transfer data between the CPU and GPU
-in the context of Scheme. However, their setting only includes lexically-scoped
-regions for which Tofte and Talpin-style analysis~\cite{tofte97}
-suffices. In contrast, we provide first-class support for
-transferable regions with dynamic lifetimes.
+Holk \emph{et al.}~\cite{gpu14} use regions to safely transfer data
+between the CPU and GPU in the context of Scheme. However, their
+setting only includes lexically-scoped regions for which Tofte and
+Talpin-style analysis suffices. In contrast, we provide first-class
+support for transferable regions with dynamic lifetimes.
 % We require this generality in order to support streaming query operators, such as the

### 2018-03-20 paper/type-inference.tex
Ecoop18-related changes and the rebuttal.

@@
 
-Theorem~\ref{thm:constraint-generation-sc}
+Theorems~\ref{thm:constraint-generation-sc}
 and~\ref{thm:constraint-solver-sc} show that the second and third

### 2018-03-20 paper/type-system.tex
Ecoop18-related changes and the rebuttal.

@@
 instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
-\C{Pair} object contained in  $\pi$. Class fields are not allowed to
-be type polymorphic, in keeping with C\#, which constitutes the core
-of \name.
+\C{Pair} object contained in  $\pi$.
 
-Note that classes in \FB are independently parameterized over types
-and regions. While this design decision has a downside that it allows
-type variables to only denote the types of objects contained in a
-single region, it yields benefits that far outweigh the costs. In
-particular, it lets us support region-polymorphic higher-order
-functions as class fields.  This allows, for example, a generic class,
-whose type parameters are $a$ and $b$, to contain a region-polymorphic
-function of type
-$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
-\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
+Classes in \FB are independently parameterized over types and regions.
+While this design decision has a downside that it allows type
+variables to only denote the types of objects contained in a single
+region, it yields benefits that outweigh the costs. In particular, it
+lets us support region-polymorphic higher-order functions as class
+fields.  This allows, for example, a generic class, whose type
+parameters are $a$ and $b$, to contain a region-polymorphic function
+of type $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0
+\conj \rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
 \C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
@@
 and region parameterizations separate also simplifies the type system
-to the point that inference becomes practical
-(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
-type systems suggests, type inference is pivotal to the practical
-utility of sophisticated type systems.
+so that inference becomes practical (Sec.~\ref{sec:type-inference}).
+The need for polymorphism at the field level is also why \FB treats
+function closures specially, rather than as objects of type \C{Func}
+as in C\#.
 
@@
 In addition, the context for the expression typing judgment also
-includes (a). a type environment ($\env \in x \mapsto \tau$) that
-contains the type bindings for variables in scope, and (b). The region
+includes (a).~a type environment ($\env \in x \mapsto \tau$) that
+contains the type bindings for variables in scope, and (b).~the region
 ($\rgn$) that serves as the allocation context for the expression
@@
 constraints ($\phi$) imposed by the class on its region parameters.
-The later is enforced by checking the validity of $\phi$, with actual
+The latter is enforced by checking the validity of $\phi$, with actual
 region arguments substituted for formal region parameters, under the
@@
 
-The type rule for \C{new \RgnZ} expressions expects the \RgnZ class's
+The type rule for the \C{new \RgnZ} expression expects the \RgnZ class's
 constructor to be called with a nullary function that returns a value
@@
 
-The type rule for \C{letregion} expression requires that the static
+The type rule for the \C{letregion} expression requires that the static
 identifier introduced by the expression be unique under the current
@@
 
-The rule for \C{open} expression, unlike the rule for \C{letregion},
-does not introduce any outlives relationship between the newly opened
-region and any pre-existing region while checking the type of the
-expression ($e$) under \C{open}. This prevents new objects allocated
-inside the transferable region from storing references to those
-outside. The newly opened region becomes the allocation context for
-$e$, which is type checked under an environment ($\env$) extended with
-the type binding for the root object.
+The rule for the \C{open} expression, unlike the rule for
+\C{letregion}, does not introduce any outlives relationship between
+the newly opened region and any pre-existing region while checking the
+type of the expression ($e$) under \C{open}. This prevents new objects
+allocated inside the transferable region from storing references to
+those outside. The newly opened region becomes the allocation context
+for $e$, which is type checked under an environment ($\env$) extended
+with the type binding for the root object.
 
-The type rule for lambda expression typechecks the lambda-bound
+The type rule for the lambda expression typechecks the lambda-bound
 expression ($e$) under an extended type environment containing
@@
 environment is extended with a binding that binds the $\thisZ$ keyword
-to the type of the current class. Note that the the type of the
+to the type of the current class. Note that the type of the
 current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
@@
 component that introduces memory regions and performs run-time
-verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
+verification of their typestate. Fig.~\ref{fig:fb-syntax-extra} shows the
 additional language constructs of \FB that manifest only at run-time.
@@
 
-{\bf A note on region handlers}. \FB stores the region handlers in a
+{\bf Integrating regions with GC heap}. \FB stores the region handlers in a
 distinguished top region, which abstracts the GC heap. Type system
@@
 the region memory to identify references into the GC heap
-unfortunately defeats the purpose of regions. A solution is to disable
-collecting region handles as long as any static region is open. Since
-static regions are intended to store temporary objects while
-populating a dynamic region, their lifetimes are short enough to let
-region handles to be GC’d.
+unfortunately defeats the purpose of regions. The solution we adopt is
+to disable collecting region handles as long as any static region is
+open. Since static regions are intended to store temporary objects
+while populating a dynamic region, their lifetimes are short enough to
+let region handles to be GC’d.

### 2018-03-22 paper/ecoop18_rebuttal.txt
ECOOP18 Rebuttal

@@
 `free` statements.
+
+7. The immediate difference between Kilim and Broom seems to be the
+enforcement of linearity of references to transferable objects. Kilim
+relies on ownership and borrowing techniques (ala Rust) to ensure that
+transferable objects are not aliased. Broom on the other had lets
+transferable regions to be aliased freely, allowing idiomatic C# code
+(e.g., iterators) to work without modifications.

### 2018-05-29 paper/ecoop18_rebuttal.txt
Edits.

@@
 3. Our model does admit aliases (through the `let` expression). The
-operational semantics defines a a *store* (Σ) that maps locations to
+operational semantics defines a *store* (Σ) that maps locations to
 their typestates, and implicitly models a heap at the granularity of a

### 2018-05-29 paper/implementation.tex
Edits.

@@
 
-As mentioned earlier, our work is a continuation of the work reported in~\cite{Broom:HotOS},
-which provides users with a C\#-based implementation of transferable regions (the features
-described in Section~\ref{sec:overview}). This system has no type system and provides users
-with no safety guarantees. \cite{Broom:HotOS} presents evidence that realistic programs can
-be implemented using transferable regions and that this can yield significant performance gains.
-In particular, it reports speedups up to 34\% for typical big-data analytics jobs.
-We now describe our implementation and experience with the type system and type inference
-algorithm presented in the current paper.
+As mentioned earlier, our work is a continuation of the work reported
+in~\cite{Broom:HotOS}, which provides users with a C\#-based
+implementation of transferable regions (the features described in
+Section~\ref{sec:overview}). This system has no type system and
+provides users with no safety guarantees. \cite{Broom:HotOS} presents
+evidence that realistic programs can be implemented using transferable
+regions and that this can yield significant performance gains.  In
+particular, it reports speedups up to 34\% for typical big-data
+analytics jobs.  We now describe our implementation and experience
+with the type system and type inference algorithm presented in the
+current paper.
 
@@
 % Our implementation is called \namec.
-The input to \namec is a program in $\absof{\FB^+}$, an extended version of $\absof{\FB}$ that
-includes assignments, conditionals, loops, more primitive datatypes
-(\eg, integers), and a null value. 
+The input to \namec is a program in $\absof{\FB^+}$, an extended
+version of $\absof{\FB}$ that includes assignments, conditionals,
+loops, more primitive datatypes (\eg, integers), and a null value. 
 % If needed, a specification file containing region type annotations
@@
 
-We performed two kinds of experiments to evaluate our region type system and type
-inference.
-First, we implemented some 
-microbenchmarks ($\le$100 LOC) consisting of standard classes such
-as pairs, lists, list iterators, etc., in $\absof{\FB^+}$, and used
-our inference engine to infer their region types.
-These classes are region-oblivious. Hence, as long as they are well-typed
-as per the core type system, \namec must be able to automatically construct its
-region-type-annotated definition without fail. \namec was
-able to infer the expected region types for all these classes under 10ms.
-Fig~\ref{fig:rev} shows the region-type-annotated
-definition computed for the list reverse method. Observe that \namec
-was able to infer that the list and its data (of type \C{T}) can be
-allocated in different regions, as long as the latter outlives the
-former. This allows, for instance, a \C{preOrder} method to traverse a
-tree in a transferable region, and return a list of its nodes, where
-the list itself is allocated in the stack region.
+We performed two kinds of experiments to evaluate our region type
+system and type inference.  First, we implemented some microbenchmarks
+($\le$100 LOC) consisting of standard classes such as pairs, lists,
+list iterators, etc., in $\absof{\FB^+}$, and used our inference
+engine to infer their region types.  These classes are
+region-oblivious. Hence, as long as they are well-typed as per the
+core type system, \namec must be able to automatically construct its
+region-type-annotated definition without fail. \namec was able to
+infer the expected region types for all these classes under 10ms.
+Fig~\ref{fig:rev} shows the region-type-annotated definition computed
+for the list reverse method. Observe that \namec was able to infer
+that the list and its data (of type \C{T}) can be allocated in
+different regions, as long as the latter outlives the former. This
+allows, for instance, a \C{preOrder} method to traverse a tree in a
+transferable region, and return a list of its nodes, where the list
+itself is allocated in the stack region.
 
@@
 benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
-$\absof{\FB^+}$, and used \namec to verify their safety (The remaining
-two Naiad vertices are not interesting from the type system
-perspective).  The number of LOC performing operations on \C{Region}
+$\absof{\FB^+}$, and used \namec to verify their safety. The 2
+remaining benchmarks were left out because their region behavior (from
+the perspective of the type system) is subsumed by the included
+benchmarks. The number of LOC performing operations on \C{Region}
 objects relative to the total LOC is 8\% or under in the Naiad
@@
 
-Our experience with Naiad benchmarks shows the utility of our
-type inference/checking tool,
-particularly because it comes at no additional cost to the developer.
+Our experience with Naiad benchmarks shows the utility of our type
+inference/checking tool, particularly because it comes at no
+additional cost to the developer.
 %% If the developer is satisfied with

### 2018-05-29 paper/type-system.tex
Edits.

@@
 %
-% For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
-% component coming from ${\rho_1}$ and second from $\rho_2$, as long as
-% $\rho_1, \rho_2$ outlive $\rhoalloc$, as illustrated by the above definition.
+% For instance, a \C{Pair} object allocated in a region ${\rhoalloc}$
+% can have its first component coming from ${\rho_1}$ and second from
+% $\rho_2$, as long as $\rho_1, \rho_2$ outlive $\rhoalloc$, as
+% illustrated by the above definition.
 
@@
 In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
-$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
-and $\pi_1$ are live during the instantiation, and (b). $\pi_0
-\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
-reflexive). Observe that the region type of \C{p} conveys
-the fact that (a). it is allocated in region $\pi_1$, and (b). it
-holds references to objects allocated in region $\pi_0$ and $\pi_1$.
-In contrast, if we choose to allocate the \C{snd} object also in
-$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
-type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
-we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
+$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$ and
+$\pi_1$ are live during the instantiation, and (b). $\pi_0 \outlives
+\pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is reflexive).
+Observe that the region type of \C{p} conveys the fact that (a). it is
+allocated in region $\pi_1$, and (b). it holds references to objects
+allocated in region $\pi_0$ and $\pi_1$.  In contrast, if we choose to
+allocate the \C{snd} object also in $\pi_1$, then \C{p} would be
+contained in $\pi_1$, and its region type would be
+\C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which we abbreviate as
+\C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
 $B\inang{\tbar}@\pi$ as being equivalent to

### 2018-06-05 paper/abstract.tex
Edits.
Typos etc fixed.

@@
 There is an increasing interest in alternative memory management
-schemes that seek to the combine the convenience of garbage collection
+schemes that seek to combine the convenience of garbage collection
 and the performance of manual memory management in a single language
@@
 \emph{transferable regions}. Unsafe transferable regions have been
-previously used in dataflow computations to contain the latency due to
-unbounded GC pauses. Our approach shows that it is possible to restore
-safety without compromising on the benefits of transferable regions.
-We prove the type safety of \name in a formal framework that includes
-its C\#-inspired features, such as higher-order functions and
-generics. We complement our type system with a type inference
-algorithm, which eliminates the need for programmers to write region
-annotations on types. The inference algorithm has been proven sound
-and relatively complete. We describe a prototype implementation of the
-inference algorithm, and our experience of using it to enforce memory
-safety in dataflow programs.
+previously used to contain the latency due to unbounded GC pauses. Our
+approach shows that it is possible to restore safety without
+compromising on the benefits of transferable regions.  We prove the
+type safety of \name in a formal framework that includes its
+C\#-inspired features, such as higher-order functions and generics. We
+complement our type system with a type inference algorithm, which
+eliminates the need for programmers to write region annotations on
+types. The inference algorithm has been proven sound and relatively
+complete. We describe a prototype implementation of the inference
+algorithm, and our experience of using it to enforce memory safety in
+dataflow programs.
 % %

### 2018-06-05 paper/overview.tex
Typos etc fixed.

@@
 Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
-of Fig.~\ref{fig:motivating-eg} can be written in \name.
-The \C{onReceive} method receives its input message in a \emph{transferred}
-region (\ie, a \emph{closed} region whose ownership is transferred to the
-recipient).
-Line 7 creates a new region to store the output for time \C{t},
-initializing it to contain an empty list.
-Line 9 opens the input region to process it.
-Line 10 creates a stack region \C{R0}.
-Thus, the temporary objects created by the iteration in line 11,
-for example, will be allocated in this stack region that lives just
-long enough.
-We open the desired output region in line 12, so that the new output
-objects created by the invocation of \C{selector} in line 13
-are allocated in the output region.
-Finally, the input region is freed in line 19.
+of Fig.~\ref{fig:motivating-eg} can be written in \name. The
+\C{onReceive} method receives its input message in a
+\emph{transferred} region (\ie, a \emph{closed} region whose ownership
+is transferred to the recipient). On Line 7, we create a new region to
+store the output for time \C{t}, initializing it to contain an empty
+list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
+creating a new stack region \C{R0}. The temporary objects created by
+the iteration on line 10, for example, will be allocated in the stack
+region \C{Ro} that lives just long enough.  We open the desired output
+region on line 11, so that the new output objects created by the
+invocation of \C{selector} on line 12 are allocated in the output
+region.  Finally, the input region is freed on line 14.
 
-Note that this approach works correctly only if \C{selector}
-creates an output object that does not share any subobjects with its input
-object (which resides in the input region and will be freed at the end of
-the method).
-If desired, such subobjects must be cloned (to copy them from the input
-region to the output region).
-Fortunately, \name's region type system (\S~\ref{sec:type-system}) is
-capable of capturing such nuances in the type of \C{selector}
-and the type checker will ensure correctness.
-Furthermore, the type can be automatically inferred by \name's region type
-inference (\S~\ref{sec:type-inference}), which can perform the above
-non-trivial reasoning on behalf of the programmer.
+Note that this approach works correctly only if \C{selector} creates
+an output object that does not share any subobjects with its input
+object (which resides in the input region and will be freed at the end
+of the method).  If desired, such subobjects must be cloned (to copy
+them from the input region to the output region).  Fortunately,
+\name's region type system (\S~\ref{sec:type-system}) is capable of
+capturing such nuances in the type of \C{selector} and the type
+checker will ensure correctness.  Furthermore, the type can be
+automatically inferred by \name's region type inference
+(\S~\ref{sec:type-inference}), which can perform the above non-trivial
+reasoning on behalf of the programmer.

### 2018-06-05 paper/overview2.tex
Typos etc fixed.

@@
 % We now illustrate how the features of \name can be used to 
-Fig.~\ref{fig:motivating-eg-in-broom} shows how the motivating example
+Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
 of Fig.~\ref{fig:motivating-eg} can be written in \name. The
@@
 \emph{transferred} region (\ie, a \emph{closed} region whose ownership
-is transferred to the recipient).  Line 7 creates a new region to
+is transferred to the recipient). On Line 7, we create a new region to
 store the output for time \C{t}, initializing it to contain an empty
-list.  Line 8 opens the input region to process it.
-% \footnote{We omit \C{@R} annotation in \C{open} when we don't need \C{R}.}
-Line 9 creates a stack region \C{R0}.  Thus, the temporary objects created by
-the iteration in line 10, for example, will be allocated in this stack
-region that lives just long enough.  We open the desired output region
-in line 11, so that the new output objects created by the invocation
-of \C{selector} in line 12 are allocated in the output region.
-Finally, the input region is freed in line 14. The output region at
-\C{map[t]} stays as along as input messages with timestamp \C{t} keep
-arriving. When the timing message for \C{t} arrives, the \C{onNotify}
-method transfers the \C{outRgn} at \C{map[t]} to a downstream actor.
+list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
+creating a new stack region \C{R0}. The temporary objects created by
+the iteration on line 10, for example, will be allocated in the stack
+region \C{R0} that lives just long enough.  We open the desired output
+region on line 11, so that the new output objects created by the
+invocation of \C{selector} on line 12 are allocated in the output
+region.  Finally, the input region is freed on line 14. The output
+region at \C{map[t]} stays as along as input messages with timestamp
+\C{t} keep arriving. When the timing message for \C{t} arrives, the
+\C{onNotify} method transfers the \C{outRgn} at \C{map[t]} to a
+downstream actor.
 % After the transfer, \C{SelectVertex} is no longer allowed to access
@@
 Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
-\C{selector} (on Line 13) should not contain any references to the
+\C{selector} (on Line 12) should not contain any references to the
 input object, since the input region, where the object resides, will

### 2018-06-05 paper/type-system.tex
Typos etc fixed.

@@
 $B\inang{\tbar}@\pi$ as being equivalent to
-$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
-$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
+$B\inang{\tbar}\inang{\bar{\pi}}$. Region annotation on type $a$,
+where $a$ is a type variable, assumes the form $a@\pi$. If $a$ is
 instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
@@
 Classes in \FB are independently parameterized over types and regions.
-While this design decision has a downside that it allows type
+While this design decision has a downside in that it allows type
 variables to only denote the types of objects contained in a single
@@
 % $\rho^a$, the closure returns an object allocated in $\rho^a$.
+
+Note that mutable object fields are conspicuously absent from the \FB
+model (and also the FGJ model in general), but this isn't a major
+shortcoming considering that constructor application, which happens
+during a new object creation, includes assignments to the object
+fields (see FGJ~\cite{fgj}). Thus the type system is already obligated
+to handle unsafe assignments.
 
@@
 ($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
-manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
-while allocating a (static) region, and \C{open} reduces to \C{opened}
-while opening a (transferable) region. An \C{opened}
+manifestations of \C{letregion} and \C{open}; \C{letregion} reduces to
+\C{letd} while allocating a (static) region, and \C{open} reduces to
+\C{opened} while opening a (transferable) region. An \C{opened}
 expression is tagged with the typestate ($s$) of the transferable

### 2018-06-05 paper/abstract.tex
Format changes.

@@
 There is an increasing interest in alternative memory management
-schemes that seek to combine the convenience of garbage collection
-and the performance of manual memory management in a single language
+schemes that seek to combine the convenience of garbage collection and
+the performance of manual memory management in a single language
 framework.  Unfortunately, ensuring safety in presence of manual
 memory management remains as great a challenge as ever. In this paper,
-we present a C\#-like object-oriented language called \name that uses
-a combination of region type system and lightweight runtime checks to
-enforce safety in presence of user-managed memory regions called
-\emph{transferable regions}. Unsafe transferable regions have been
-previously used to contain the latency due to unbounded GC pauses. Our
-approach shows that it is possible to restore safety without
-compromising on the benefits of transferable regions.  We prove the
-type safety of \name in a formal framework that includes its
-C\#-inspired features, such as higher-order functions and generics. We
-complement our type system with a type inference algorithm, which
-eliminates the need for programmers to write region annotations on
-types. The inference algorithm has been proven sound and relatively
-complete. We describe a prototype implementation of the inference
-algorithm, and our experience of using it to enforce memory safety in
-dataflow programs.
+we present a C\#-like object-oriented language called {\sc
+Broom}\xspace that uses a combination of region type system and
+lightweight runtime checks to enforce safety in presence of
+user-managed memory regions called \emph{transferable regions}. Unsafe
+transferable regions have been previously used to contain the latency
+due to unbounded GC pauses. Our approach shows that it is possible to
+restore safety without compromising on the benefits of transferable
+regions.  We prove the type safety of {\sc Broom}\xspace in a formal
+framework that includes its C\#-inspired features, such as
+higher-order functions and generics. We complement our type system
+with a type inference algorithm, which eliminates the need for
+programmers to write region annotations on types. The inference
+algorithm has been proven sound and relatively complete. We describe a
+prototype implementation of the inference algorithm, and our
+experience of using it to enforce memory safety in dataflow programs.
 % %

### 2018-06-05 paper/broom.tex
Edits to appendix. CR04 added as related work.
Ecoop18-related changes and the rebuttal.
Format changes.

--- 2017-01-02 paper/broom.tex
@@
-\documentclass[a4paper,UKenglish]{lipics-v2016}
+\documentclass[a4paper,UKenglish]{lipics-v2018}
 %This is a template for producing LIPIcs articles. 
@@
 
-%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
-\author[1]{Gowthan Kaki}
-\author[2]{G. Ramalingam}
-\author[3]{Kapil Vaswani}
-\author[4]{Dimitrios Vytiniotis}
-\affil[1]{Purdue University}
-\affil[2]{Microsoft Research}
-\affil[3]{Microsoft Research}
-\affil[4]{Microsoft Research}
-\authorrunning{G. Kaki, G. Ramalingam, K. Vaswani, and D. Vytiniotis} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
+\author{Gowtham Kaki}
+       {Purdue University, USA\footnote{Work done during an internship at
+Microsoft Research, India.}}
+       {gkaki@purdue.edu}
+       {}
+       {}%mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{G. Ramalingam}
+       {Microsoft Research, India}
+       {grama@microsoft.com}
+       {}
+       {}%mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 
-\Copyright{Gowtham Kaki, G. Ramalingam, Kapil Vaswani, and Dimitrios Vytiniotis}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\authorrunning{G. Kaki and G. Ramalingam}%mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-% \subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html .
-E.g., cite as "F.1.1 Models of Computation".
-\keywords{Memory management, regions, type system, type inference}% mandatory: Please provide 1-5 keywords
+\Copyright{Gowtham Kaki and G. Ramalingam}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\subjclass{
+	\ccsdesc[500]{Software and its engineering~Allocation / deallocation strategies};\xspace
+	\ccsdesc[500]{Software and its engineering~Object oriented languages};\xspace
+	\ccsdesc[500]{Software and its engineering~Data flow languages};\xspace
+	\ccsdesc[300]{Software and its engineering~Software verification and validation};\xspace
+}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
+\keywords{Memory Safety, Formal Methods, Type System, Type Inference,
+Regions, Featherweight Java}% mandatory: Please provide 1-5 keywords
+
 % Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\acknowledgements{We are grateful to Kapil Vaswani, Dimitrios
+Vytiniotis, Michael Isard, and Steven Hand for their valuable inputs
+during the initial stages of this project. We would also like to thank
+the anonymous reviewers for their careful scrutiny and insightful
+comments that helped improve the quality of this paper.}%optional
+
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Acces}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{ECOOP 2017}
+\EventEditors{Todd Millstein}
+\EventNoEds{1}
+\EventLongTitle{32nd European Conference on Object-Oriented Programming (ECOOP 2018)}
+\EventShortTitle{ECOOP 2018}
 \EventAcronym{ECOOP}
-\EventYear{2017}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventYear{2018}
+\EventDate{July 16--21, 2018}
+\EventLocation{Amsterdam, Netherlands}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{109}
+\ArticleNo{11}
 % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \input{related}
+
+\bibliography{broom}
+
+\clearpage
 
@@
 
-\bibliography{broom}
-
 \end{document}

### 2018-06-11 paper/broom.tex
Camera-Ready first cut

@@
 
-\input{macros}
+%--------------------------------------------------------------------
+%Formatting Commands
+%--------------------------------------------------------------------
+\newcommand{\C}[1]{\code{#1}}
+\renewcommand{\k}[1]{\mathtt{#1}} % Code font in math mode.
+\newcommand{\code}[1]{\,{\tt #1}\,}
+\newcommand{\conj}{\wedge}
+\newcommand{\disj}{\vee}
+\newcommand{\tuplee}[1]{\langle #1 \rangle}
+\newcommand{\ALT}{~\mid~}
+\newcommand{\lamb}[2]{\lambda #1.\,#2}
+\newcommand{\spc}[0]{\quad}
+\newcommand{\redsto}[3]{#1 \vdash #2 \longrightarrow #3}
+\newcommand{\rredsto}[0]{\rightsquigarrow_r }
+\newcommand{\evalsto}[0]{\longrightarrow }
+\newcommand{\env}[0]{\Gamma}
+\newcommand{\tywf}[2]{#1\,\vdash\,#2 \; \texttt{ok}}
+\newcommand{\fgjtywf}[2]{#1\,\Vdash\,#2 \; \texttt{ok}}
+\newcommand{\isvalid}[2]{#1\,\vdash\,#2}
+\newcommand{\hastyp}[3]{#1\,\vdash\,#2:#3}
+\newcommand{\hasfgjtyp}[3]{#1\,\Vdash^{}\,#2\,:\,#3}
+\newcommand{\subtyp}[3]{#1\,\vdash\,#2 <: #3}
+\newcommand{\fgjsubtyp}[3]{#1\,\Vdash\,#2 <: #3}
+\newcommand{\msentails}[2]{#1\,\models\,#2}
+\newcommand{\thesemof}[1]{ \llbracket #1 \rrbracket}
+\newcommand{\subst}[2]{\lbrack #1/#2 \rbrack}
+\newcommand{\absof}[1]{\thesemof{#1}}
+\newcommand{\broom}{{\sc Broom}\xspace}
+\newcommand{\name}{{\sc Broom}\xspace}
+\newcommand{\namec}{{\sc Broomc}\xspace}
+\newcommand{\fbname}{{\sc Featherweight Broom}\xspace}
+\newcommand{\FB}{{\sc FB}\xspace}
+\newcommand{\csharp}{C\#~}
+\newcommand{\outlives}{\succeq}
+\newcommand{\underlives}{\preceq}
+\newcommand{\coloneqq}{::=}
+\newcommand{\M}[1]{\mathtt{#1}}
+\newcommand{\ObjZ}{\C{Object}}
+\newcommand{\RgnZ}{\C{Region}}
+\newcommand{\thisZ}{\C{this}}
+\newcommand{\superZ}{\C{super}}
+\newcommand{\unitZ}{\C{unit}}
+\newcommand{\FuncZ}{\C{Func}}
+\newcommand{\unitval}{\C{()}}
+\newcommand{\inang}[1]{\langle #1 \rangle}
+\newcommand{\rhoalloc}{\rho^a}
+\newcommand{\rhobar}{\bar{\rho}}
+\newcommand{\rhoallocm}{\rho^a_m}
+\newcommand{\rhobarm}{\bar{\rho_m}}
+\newcommand{\ralloc}{\pi^a}
+\newcommand{\rgn}{r}
+\newcommand{\rbar}{\bar{\rgn}}
+\newcommand{\tbar}{\bar{T}}
+\newcommand{\taubar}{\bar{\tau}}
+\newcommand{\xbar}{\bar{x}}
+\newcommand{\vbar}{\bar{v}}
+\newcommand{\ebar}{\bar{e}}
+\newcommand{\extends}{\triangleleft}
+\newcommand{\rextends}{\blacktriangleleft}
+\renewcommand{\bar}[1]{\overline{#1}}
+\newcommand{\angR}{\inang{\rhobar \,|\, \phi}}
+\newcommand{\angT}{\inang{\tbar}}
+\newcommand{\tyvar}{a}
+\newcommand{\tyvarb}{b}
+\newcommand{\angAlpha}{\inang{\bar{\tyvar} \extends \bar{\fgjN}}}
+\newcommand{\fgjN}{K}
+\newcommand{\fbN}{N}
+\newcommand{\headerOf}[1]{\C{class}\; #1\angAlpha\angR \extends \fbN}
+\newcommand{\substFn}{\mathcal{S}}
+% \newcommand{\regionSubstFn}{\mathcal{R}}
+% \newcommand{\predSubstFn}{\mathcal{P}}
+% \newcommand{\regionSubstFn}{\soln_V}
+% \newcommand{\predSubstFn}{\soln_P}
+\newcommand{\regionConstants}{R}
+\newcommand{\regionVars}{V}
+\newcommand{\predVars}{P}
+\newcommand{\constraintSet}{C}
+\newcommand{\regionDeltaMap}{{\Delta_R}}
+\newcommand{\predDeltaMap}{{\Delta_P}}
+\newcommand{\mang}{\inang{\bar{\rho_m} \,|\, \phi_m}}
+\newcommand{\rhoset}{\Sigma}
+\newcommand{\rhomap}{\Sigma}
+\newcommand{\rhoenv}{\Delta}
+\newcommand{\aenv}{\Theta}
+\newcommand{\phicx}{\Phi}
+\newcommand{\phictxt}{\phi_{cx}}
+\newcommand{\phicstr}{\phi_{cs}}
+\newcommand{\phisol}{\phi_{sol}}
+\newcommand{\subtypcx}{\rhoenv,\aenv,\phicx}
+\newcommand{\A}{{\mathcal{A}}}
+\newcommand{\exptycx}[2]{\A,#1,#2}
+\newcommand{\void}{\C{void}}
+\newcommand{\functy}[3]{\FuncZ\inang{#1}\inang{#2,#3}}
+\newcommand{\stmtsemcx}{\A,\env}
+\newcommand{\stmtsemcxp}{\A',\env'}
+\newcommand{\stmtsemcxpp}{\A'',\env''}
+\newcommand{\sredsto}{\Rightarrow}
+\newcommand{\stmtsem}[4]{#1 \, \vdash \, #2 \,/\, #3 \sredsto #4}
+\newcommand{\okin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
+\newcommand{\lambdaexp}[4]{\lambda@#1\inang{#2}(#3). #4}
+\newcommand{\Lambdaexp}[2]{\Lambda\inang{#1}.#2}
+\newcommand{\letexp}[3]{\C{let}\;#1\,=\,#2\;\C{in}\;#3}
+\newcommand{\letregion}[2]{\C{letregion}\;#1\;\C{in}\;#2}
+\newcommand{\open}[4]{\C{open}\;#1\;\C{as}\; #3@#2\;\C{in}\;#4}
+\newcommand{\opened}[3]{\C{opened}\;#1(#2)\;\C{in}\;#3}
+\newcommand{\letd}[2]{\C{letd}\;#1\;\C{in}\;#2}
+\newcommand{\unpackexp}[4]{\C{let}\;(#1,\,#2)\,=\,\C{unpack}\;#3\;\C{in}\;#4}
+\newcommand{\invalidexn}{\bot}
+\newcommand{\toprgn}{\pi_{\top}}
+\newcommand{\csolvestar}{{\sc CSolve*}\xspace}
+\newcommand{\elabExpr}{{\sf elabExpr}}
+\newcommand{\fields}{{\sf fields}}
+\newcommand{\mtype}{{\sf mtype}}
+\newcommand{\mbody}{{\sf mbody}}
+\newcommand{\bound}{{\sf bound}}
+\newcommand{\fresh}{{\sf fresh}}
+\newcommand{\typeOk}{{\sf typeOk}}
+\newcommand{\subtypeOk}{{\sf subtypeOk}}
+\newcommand{\templateTy}{{\sf templateTy}}
+\newcommand{\allocRgn}{{\sf allocRgn}}
+\newcommand{\shape}{{\sf shape}}
+\newcommand{\ctype}{{\sf ctype}}
+\newcommand{\override}{{\sf override}}
+%\newcommand{\smallskip}{\vspace*{0.1in}}
+\newcommand{\elabMeth}{{\sf elabMeth}}
+\newcommand{\frv}{{\sf frv}}
+\newcommand{\rv}{{\sf RVars}}
+\newcommand{\solve}{{\sf solve}}
+\newcommand{\valuee}{{\sf value}}
+\newcommand{\elabClass}{{\sf elabClass}}
+\newcommand{\elabCons}{{\sf elabCons}}
+\newcommand{\elabClassTable}{{\sf elabClassTable}}
+\newcommand{\CLOSED}{\mathtt{C}}
+\newcommand{\OPEN}{\mathtt{O}}
+\newcommand{\XFERRED}{{\color{blue}\times}}
+\newcommand{\LIVE}{{\color{ForestGreen}\blacksquare}}
+\newcommand{\SLIVE}{\LIVE}
+\newcommand{\TLIVE}{\LIVE}
+% \newcommand{\SLIVE}{{\color{ForestGreen}\widehat{\blacksquare}}}
+% \newcommand{\TLIVE}{{\color{ForestGreen}\overrightarrow{\blacksquare}}}
+%\newcommand{\FREE}{{\color{blue}\square}}
+% \newcommand{\USED}{{\color{blue}\overrightarrow{\blacksquare}}}
+\newcommand{\USED}{{\color{blue} \square}}
+\newcommand{\transfer}{{\sf transfer}}
+\newcommand{\CSP}{(C\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi}, \bar{\rhoenv})}
+\newcommand{\csp}{(c\lbrack \varphi,\rhobar \rbrack,\rhoenv_{\varphi},\bar{\rhoenv})}
+\newcommand{\Phicx}{\Phi_{cx}}
+\newcommand{\Phics}{\Phi_{cs}}
+\newcommand{\tynwf}[2]{#1\,\nvdash\,#2 \; \texttt{ok}}
+\newcommand{\csid}[1]{\lbrack \mathbf{c_{#1}} \rbrack:\;}
+\newcommand{\qqquad}{\quad\quad\quad}
+\newcommand{\emptyA}[0]{(\Delta,\cdot,\phicx)}
+\newcommand{\emptyASigp}[0]{(\Delta,\cdot,\phicx)}
+\newcommand{\RgnZT}[1]{\RgnZ\inang{T}\inang{#1}}
+\newcommand{\BZ}[0]{B\inang{\tbar}}
+\newcommand{\BZT}[1]{B\inang{\tbar}\inang{#1}}
+\newcommand{\redstoo}[3]{#2 \longrightarrow #3} %\Delta ignored
+\newcommand{\redstocup}[3]{\redstoo{\rhoenv \cup \{#1\}}{#2}{#3}}
+\newcommand{\loc}{\mathtt{l}}
+\newcommand{\locbar}{\overline{\loc}}
+\newcommand{\isLive}{{\sf isLive}}
+\newcommand{\mem}{\Sigma}
+\newcommand{\status}{s}
+\newcommand{\toploc}{\loc_{\top}}
+\newcommand{\consistent}{{\sf consistent}}
 
-%%%%%%%%%%%%%%%% OUR PRELIM STUFF ENDS %%%%%%%%%%%%%%%%%%%%
+%Rules
+%-----
+\newcommand{\rulelabel}[1]{\textrm{\sc {#1}}}
+\newcommand{\rulee}[1]{\textrm{\sc {#1}}}
+\newcommand{\ilrulelabel}[1]{{\sc #1}}
+\newcommand{\RULE}[2]{\frac{\begin{array}{c}#1\end{array}}
+                           {\begin{array}{c}#2\end{array}}}
 
-% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Math mode
+%-----------
+\newenvironment{nop}{}{}
+\newenvironment{smathpar}{
+\begin{nop}\small\begin{mathpar}}{
+\end{mathpar}\end{nop}\ignorespacesafterend}
+
+% Theorems Environment
+%----------------------
+% \newtheorem{theorem}{Theorem}[section]
+% \newtheorem{lemma}[theorem]{Lemma}
+% \newtheorem{proposition}[theorem]{Proposition}
+% \newtheorem{corollary}[theorem]{Corollary}
+
+% \newenvironment{proof}[1][Proof]{\begin{trivlist}
+% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+% \newenvironment{definition}[1][Definition]{\begin{trivlist}
+% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+% \newenvironment{example}[1][Example]{\begin{trivlist}
+% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+% \newenvironment{remark}[1][Remark]{\begin{trivlist}
+% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
+
+% \newcommand{\qed}{\nobreak \ifvmode \relax \else
+      % \ifdim\lastskip<1.5em \hskip-\lastskip
+      % \hskip1.5em plus0em minus0.5em \fi \nobreak
+      % \vrule height0.75em width0.5em depth0.25em\fi}
+
+\usepackage{listings}
+\usepackage{subcaption}
+\usepackage{stmaryrd}
+\usepackage[skip=5pt]{caption}
+
+\definecolor{Burgundy}{rgb}{0.5, 0.0, 0.13}
+\definecolor{Bittersweet}{rgb}{1.0, 0.44, 0.37}
+\definecolor{ForestGreen}{rgb}{0.13, 0.55, 0.13}
+\definecolor{Bubbles}{rgb}{0.91, 1.0, 1.0}
+\definecolor{LighterGray}{rgb}{0.93, 0.93, 0.93}
+
+\newcommand{\lstjava}{\lstset{ %
+language=Java, % choose the language of the code
+backgroundcolor=\color{LighterGray},   
+basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
+keywordstyle=\color{Burgundy}\bf,
+% numbers=left,                   % where to put the line-numbers
+numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
+stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
+numbersep=5pt,                  % how far the line-numbers are from the code
+showspaces=false,               % show spaces adding particular underscores
+showstringspaces=false,         % underline spaces within strings
+showtabs=false,                 % show tabs within strings adding particular underscores
+% frame=single,                   % adds a frame around the code
+tabsize=2,                      % sets default tabsize to 2 spaces
+captionpos=b,                   % sets the caption-position to bottom
+breaklines=true,                % sets automatic line breaking
+breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+commentstyle=\itshape\color{MidnightBlue},
+%escapeinside={\%*}{*)},         % if you want to add a comment within your code
+morekeywords={var, foreach, unit, letregion, let, in, as, openalloc, open, withroot, not, : , /\\}
+}}
+\lstnewenvironment{codejava}
+    { % \centering
+			\lstjava
+      \lstset{mathescape=true}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
+
+\lstnewenvironment{numcodejava}
+    { % \centering
+			\lstjava
+      \lstset{numbers=left,firstnumber=1}%
+      \csname lst@setfirstlabel\endcsname}
+    { %\centering
+      \csname lst@savefirstlabel\endcsname}
+
+%--------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%% OUR PRELIM STUFF ENDS %%%%%%%%%%%%%%%%%%%%%%
+%--------------------------------------------------------------------
+
 \title{Safe Transferable Regions}
@@
 during the initial stages of this project. We would also like to thank
-the anonymous reviewers for their careful scrutiny and insightful
-comments that helped improve the quality of this paper.}%optional
+multiple anonymous reviewers for their careful scrutiny and feedback at
+various stages of this work.}%optional
 
@@
 
+%--------------------------------------------------------------------
+%                           Abstract
+%--------------------------------------------------------------------
 \begin{abstract}
-\input{abstract}
+There is an increasing interest in alternative memory management
+schemes that seek to combine the convenience of garbage collection and
+the performance of manual memory management in a single language
+framework.  Unfortunately, ensuring safety in presence of manual
+memory management remains as great a challenge as ever. In this paper,
+we present a C\#-like object-oriented language called {\sc
+Broom}\xspace that uses a combination of region type system and
+lightweight runtime checks to enforce safety in presence of
+user-managed memory regions called \emph{transferable regions}. Unsafe
+transferable regions have been previously used to contain the latency
+due to unbounded GC pauses. Our approach shows that it is possible to
+restore safety without compromising on the benefits of transferable
+regions.  We prove the type safety of {\sc Broom}\xspace in a formal
+framework that includes its C\#-inspired features, such as
+higher-order functions and generics. We complement our type system
+with a type inference algorithm, which eliminates the need for
+programmers to write region annotations on types. The inference
+algorithm has been proven sound and relatively complete. We describe a
+prototype implementation of the inference algorithm, and our
+experience of using it to enforce memory safety in dataflow programs.
 \end{abstract}
 
-\input{introduction}
+%--------------------------------------------------------------------
+%                           Introduction
+%--------------------------------------------------------------------
+\newcommand{\TODO}[1]{\textbf{TODO: #1}} \newcommand{\eg}{\emph{e.g.}}
+\newcommand{\ie}{\emph{i.e.}}
+\newcommand{\mypara}[1]{\textbf{#1}}
 
-\input{overview2}
+\section{Introduction} \label{sec:introduction}
 
-\input{type-system}
+% This paper presents a \emph{safe} and \emph{efficient} region-based approach
+% to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.
 
+Computations performed by a network of concurrent communicating actors
+often involve data transfer between producers and consumers. Consider
+for example the \C{SELECT} query operator shown in
+Fig.~\ref{fig:motivating-eg}, which functions as an actor in a
+dataflow computation involving a network of other such query
+operators. \C{SELECT} receives a stream of input messages, each
+associated with a time window $t$, processed by method
+\texttt{onReceive}.  Each input message contains a list of inputs,
+each processed by applying a user-defined function to create a
+corresponding output.  Multiple messages with the same timestamp are
+permitted and messages with different timestamps may arrive out of
+order.  An invocation of method \texttt{OnNotify} indicates that no
+more input messages with a timestamp $t$ will be subsequently
+delivered.  At this point, the operator completes the processing for
+time window $t$ and sends a corresponding output message to its
+successor.
+
+\mypara{Performance.}
+When the transferred data structures are large, as is the case with many
+streaming big-data analysis systems, garbage collection overhead becomes
+significant~\cite{Broom:HotOS}.  Further, in a distributed
+dataflow system, the GC pause at one node can have a cascading adverse
+effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
+a GC pause at an upstream actor can block downstream actors that are waiting for
+messages.  However, much of the GC overhead results from the
+collector performing avoidable or unproductive work.  For the example
+in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
+\C{map}, although its objects cannot be collected until a
+suitable timing message arrives.
+There has been increasing interest in off-heap memory management for better performance
+in the context of several  systems and languages (e.g.,
+Spark~\cite{SPARK:2015}, Scala~\cite{ScalaOffHeap}, Rust).
+Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in
+the absence of language support.
+
+\begin{figure}[t!]
+\begin{numcodejava}
+class SelectVertex<TIn, TOut> {
+  Func<TIn, TOut> selector;
+  Dictionary<Time, List<TOut>> map;
+  ...
+  void onReceive(Time t, List<TIn> inList) {
+    if (!map.ContainsKey(t)) map[t] = new List<TOut>();
+    foreach (TIn input in inList) {
+      TOut output = selector(input);
+      map[t].add(output); } }
+  void onNotify(Time t) {
+     List<TOut> outList = map[t];
+     map.Remove(t);
+     transfer(successorId, t, outList); }
+}
+\end{numcodejava}
+\caption{\C{SELECT} dataflow operator}
+\label{fig:motivating-eg}
+\vspace*{-0.2in}
+\end{figure}
+
+\begin{figure}
+\centering
+\subcaptionbox {
+  Before transfer.
+  \label{fig:after-transfer}
+} [
+  0.35\columnwidth
+] {
+  \includegraphics[scale=0.36]{producer-heap}
+}
+\subcaptionbox {
+  After transfer.
+  \label{fig:before-transfer}
+} {
+  \includegraphics[scale=0.36]{producer-consumer-heap}
+}
+
+\caption{References in and out of a transferable region ($R$) become
+invalid after transfer.}
+\label{fig:transfer-violations}
+\end{figure}
+
+\mypara{Safety.}
+Several of these systems are designed to allow the producer and
+consumer to execute in the same address space or in different address
+spaces (as determined by the compiler and runtime).  We wish to ensure
+memory safety in the presence of such transferred data.  A transfer
+operation, with no additional checks, may cause memory safety
+violations, both at the producer of the data structure, and its
+(possibly remote) consumer. At the producer, any existing references
+into the transferred data structure become invalid post transfer. If
+the data structure contains references to objects outside (the
+transferred data), then such references become invalid in the context
+of the consumer. Both scenarios are depicted in
+Fig.~\ref{fig:transfer-violations}.  While references from within the
+transferable data to outside can be disallowed in the interest of
+safety, references into transferable data needs to be allowed before
+the data is transferred. Allowing such references is crucial, as any
+non-trivial program creates temporary references to the internal
+objects of a data-structure.
+
+\mypara{Safe Transferable Regions.}
+In this paper, we present a language-based approach to cleanly
+encapsulate transferable data and a safe and efficient implementation
+of such data based on regions. 
+%
+A region is a block of memory that is allocated
+and freed in one shot, often in constant time. A region may contain
+one or more contiguous range of memory locations, and individual
+objects may be dynamically allocated within the region over time,
+while they are deallocated en masse when the region is freed.  Thus, a
+region is a good fit for a transferable data-structure.  In
+Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
+time window $t$ (i.e., \C{map[t]}) can be a separate region that is
+allocated when the first message with timestamp $t$ arrives, and
+deallocated after \C{map[t]} is transferred in \C{onNotify}.
+
+The use of regions alleviates the performance concern related to
+garbage collection described earlier. (It also enables more efficient
+serialization/deserialization of data-structures across address spaces,
+which is also a significant performance bottleneck in such systems.)
+However, manual region memory management introduces memory safety
+issues such as dangling pointers.
+We present a single type system that simultaneously addresses
+this safety concern, as well as those described above related to
+the use of \emph{transferable data}.
+
+Safe region-based memory management using types was
+pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
+use only lexically scoped regions.  At runtime, the set of all regions
+(existing at a point in time) forms a stack. Thus, the lifetimes of
+all regions must be well-nested: it is not possible to have two
+regions whose lifetimes overlap, with neither one's lifetime contained
+within the other.  Unfortunately, the data structures in the above
+example do not satisfy this restriction (as the output messages for
+multiple time windows may be simultaneously live, without any
+containment relation between their lifetimes).  We refer to regions
+with lexically scoped lifetimes as \emph{stack regions} and to regions
+that do not have such a lexically scoped region as \emph{dynamic}
+regions.
+
+Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
+that can be safely transferred across address spaces. We refer to such dynamic
+regions as \emph{transferable} regions.
+%
+Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
+that combine linear types and regions to support dynamic regions.
+Unlike the prior work, we ensure memory safety in the presence of
+transferable regions through a \emph{combination of a static
+typing discipline and lightweight runtime checks in place of linear types}.
+
+\mypara{Language.}
+The cornerstone of this approach is an \C{open} lexical block for
+transferable regions, that ``opens'' a transferable region and
+guarantees that the region won't be transferred/freed while it is
+open. By nesting a Tofte-Talpin style \C{letregion} lexical block,
+that delimits the lifetime of a stack region, inside an \C{open}
+lexical block for a transferable region, we can guarantee that the
+transferable region will remain live as long as the stack region is
+live. We say that the former \emph{outlives} the latter, and any
+references from the stack region to the transferable region are
+therefore safe. This is particularly useful, since such stack regions
+serve as temporary working storage while working with the (opened)
+transferable regions.
+
+By controlling the outlives relationships between various
+regions, we only allow safe cross-region references, while prohibiting
+unsafe ones. In the above example, an outlives relationship
+\emph{from} the stack region \emph{to} the transferable region means
+that the references in that direction are allowed, but not the
+references in the opposite direction. In contrast, if an \C{open}
+block of a transferable region $R_0$ is nested inside an \C{open}
+block of another transferable region $R_1$, we do not establish any
+outlives relationships, thus declaring our intention to not allow any
+cross-region references between $R_0$ and $R_1$.  Finally, we observe
+that outlives relationships are established based on the lexical
+structure of the program, hence a static type system can enforce them
+effectively. By assigning region types to objects, which capture the
+regions such objects are allocated in, and by maintaining outlives
+relationships between various regions, we can statically decide the
+safety of all references in the program.
+
+\mypara{Type System.}
+Formally, our type system introduces \emph{region parameters} (for both
+classes and methods) and uses constrained parametric polymorphism
+over these parameters, where the constraints capture outlives constraints
+between the region parameters.
+The type system may be seen as a form of ownership type system~\cite{OwnershipSurvey},
+with a region being the owner of all objects allocated in that region.
+
+\mypara{Lightweight Runtime Checks.}
+Ensuring memory safety using this approach requires ensuring
+that the use of transferable regions satisfies certain temporal properties.
+Firstly, a transferable region should not be transferred/freed
+inside an open block of that region (i.e, while it is still open).
+Secondly, a transferred/freed region should not be opened. These are
+typestate invariants on the transferable region objects, which are
+hard to enforce statically in the presence of unrestricted
+aliasing. Techniques like linear types and unique pointers can be used
+to restrict aliasing, but the constraints they impose are often hard
+to program around. We therefore enforce typestate invariants at
+runtime via lightweight checks. In particular, we define an acceptable
+state transition discipline for transferable regions
+(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
+transition of a transferable region (\eg, from \emph{open} state to
+\emph{freed} state) is valid or not. The check is lightweight since it
+only involves checking a single tag that captures the current state.
+We believe that this is a reasonable choice since regions are
+coarse-grained objects manipulated infrequently, when compared to the
+fine-grained objects that are present inside these regions, for which
+safety is enforced statically.
+
+\mypara{Type Inference.}
+One of the key contributions of this paper is a type inference
+algorithm that eliminates the need for users to write region type
+annotations.  The users write programs in the underlying language that
+provides primitives for the users to manipulate regions (create, open,
+free, transfer) and to allocate objects in regions, but has no region
+type annotations.
+
+Our inference algorithm proceeds in three stages: in the first stage,
+we elaborate the program by introducing region parameters (for classes
+and methods); in the second stage we generate a set of constraints
+that must be satisfied for the program to type check; in the third
+stage, we solve the set of constraints, inferring the preconditions of
+all classes and methods (in terms of the expected outlives-constraints
+between the region parameters).  We show that the algorithm is sound,
+and the stages of constraint generation and solving are complete
+(i.e., the algorithm is complete relative to the first stage of
+elaboration).
+% (using an interprocedural fixed point computation).
+
+\mypara{Evaluation.}
+Our work was inspired by~\cite{Broom:HotOS}, which presents evidence
+that realistic programs can be written using transferable regions and
+that this can yield significant performance improvements.
+While~\cite{Broom:HotOS} addresses the engineering challenges in
+extending a managed runtime with transferable regions, it adopts an ad
+hoc approach insofar as language design is concerned, exposing the
+region functionality through an unsafe API, and in the process losing
+the safety guarantees. Our contribution is an alternative approach
+that is grounded in sound theory and restores the language safety
+guarantees. The utility of our approach is demonstrated by a prototype
+implementation of our type inference algorithm that was able to
+identify unsafe memory accesses among the benchmarks extracted
+from~\cite{Broom:HotOS}.
+
+\mypara{Contributions.}
+The paper makes the following contributions:
+
+\begin{itemize} 
+  \item We present \broom, a \csharp-like typed object-oriented
+  language that supports programmer-managed memory regions. \broom{}
+  extends its core language, which includes \emph{lambdas}
+  (higher-order functions) and \emph{generics} (parametric
+  polymorphism), with constructs to create, manage and free static and
+  transferable memory regions. Transferable regions are first-class
+  values in \broom.
+
+\item \broom{} is equipped with a region type system that statically
+  guarantees safety of all memory accesses in a well-typed program,
+  provided that certain typestate invariants on regions hold.  The
+  latter invariants are enforced via simple runtime checks.
+
+  \item We define an operational semantics for \broom, and a type
+  safety result that clearly defines and proves safety guarantees
+  described above.
+
+  \item We describe a region type inference algorithm for \broom that
+  (a). completely eliminates the need to annotate \broom programs with
+  region types, and (b). enables seamless interoperability between
+  region-aware \broom programs and legacy standard library code that
+  is region-oblivious. Our inference algorithm is based on a novel
+  constraint solver that performs abduction in a partial-order
+  constraint domain to infer weakest solutions to recursive
+  constraints.
+
+\item We establish the soundness and relative completeness of the type
+inference algorithm.
+
+  \item We describe an implementation of \broom frontend in OCaml,
+  along with case studies where the region type system was able to
+  identify unsafe memory accesses statically.
+
+\end{itemize}
+
+%--------------------------------------------------------------------
+%                             Overview
+%--------------------------------------------------------------------
+\newcommand{\COMMENT}[1]{}
+
+\section{An Informal Overview of \name} \label{sec:overview}
+
+\name enriches a simple object-oriented language (supporting
+parametric polymorphism and lambdas) with a set of region-specific
+constructs.  In this section, we present an informal overview of these
+region-specific constructs.
+
+\subsection{Using Regions in \name}
+\label{sec:alloc-ctxt}
+
+\mypara{Stack Regions} The ``\C{letregion R \{ S \}}'' construct
+creates a new stack region, with a static identifier \C{R}, whose
+scope is restricted to the statement \C{S}. The semantics of
+\C{letregion} is similar to Tofte and Talpin~\cite{tofte94}'s
+\C{letregion} expression: objects can be allocated by \C{S} in the
+newly created region while \C{R} is in scope, but the region and all
+objects allocated within it are freed at the end of \C{S}.
+
+\mypara{Object Allocation} The ``\C{new@R T()}'' construct creates
+a new object of type \C{T} in the region \C{R}. The specification of
+the allocation region \C{R} in this construct is optional.  At
+runtime, \name maintains a stack of \emph{active} regions, and we
+refer to the region at the top of the stack as the \emph{allocation
+context}. The statement \C{new T()} allocates the newly created object
+in the current allocation context.
+%
+This is important as it enables \name applications to use existing
+region-oblivious C\# libraries, as explained soon.
+
+\mypara{Transferable Regions} 
+Transferable regions are first class values of \name: they are objects
+of the class \C{Region}, they are created using the \C{new} keyword,
+and can be passed as arguments, stored in data structures, and
+returned from methods.  A transferable region is intended to
+encapsulate a single data-structure, consisting of a collection of
+objects with a distinguished root object of some type \C{T}, which we
+refer to as the region's \emph{root} object.  The class \C{Region} is
+parametric over the type \C{T} of this root object.
+
+The \C{Region} constructor takes as a parameter a function that
+constructs the root object: it creates a new region and invokes this
+function, with the new region as the allocation context, to create the
+root object of the region. The following code illustrates the
+creation of a transferable region, whose root is an object of type
+\C{A}.
+% (\C{r}) of type \C{T}:
+\begin{codejava} 
+  Region<A> rgn = new Region<A>(() => new A())
+\end{codejava} 
+In the above code, \C{rgn} is called the \emph{handler} to the newly
+created region, and is required to read the contents of the region, or
+change its state. The class \C{Region} offers two methods: a \C{free}
+method that deallocates the region (and all the objects allocated
+within it), and \C{transfer} method that transfers the region to a
+(possibly remote) consumer process.
+
+\mypara{Open and Closed Regions} A transferable region must be
+explicitly \emph{opened} using \name's \C{open} construct in order to
+either read or update or allocate objects in the region.
+Specifically, the construct ``\C{open rgn as v@R \{ S \}}'' does the
+following: (a). It opens the transferable region handled by \C{rgn}
+for allocation (i.e., makes it the current allocation context), (b).
+binds the identifier \C{R} to this open region, and (c). initializes
+the newly introduced local variable \C{v} to refer to the root object
+of the region.  The \C{@R} part of the statement is optional and may
+be omitted.  The \C{open} construct is intended to simplify the
+problem of ensuring memory safety, as will be explained soon.  We
+refer to a transferable region that has not been opened as a
+\emph{closed} region. A transferable region can only be transferred or
+freed when it is in closed state. The acceptable state transition
+discipline over the lifetime of a transferable region is described
+in Fig.~\ref{fig:region-fsm}. Enforcement of this discipline is done
+at runtime.
+
+Opening a region also makes it the current allocation context.
+Thus, given a  C\# library function \C{f} (that makes no use of \name's region constructs),
+opening a region \C{R} and invoking \C{f} has the effect that all objects created by this
+invocation are allocated in the region \C{R}.
+
+\mypara{Motivating Example}
+Fig~\ref{fig:motivating-eg-in-broom} shows how the motivating example
+of Fig.~\ref{fig:motivating-eg} can be written in \name. The
+\C{onReceive} method receives its input message in a
+\emph{transferred} region (\ie, a \emph{closed} region whose ownership
+is transferred to the recipient). On Line 7, we create a new region to
+store the output for time \C{t}, initializing it to contain an empty
+list. On Lines 8 and 9, we open the input region \C{inRgn} followed by
+creating a new stack region \C{R0}. The temporary objects created by
+the iteration on line 10, for example, will be allocated in the stack
+region \C{R0} that lives just long enough.  We open the desired output
+region on line 11, so that the new output objects created by the
+invocation of \C{selector} on line 12 are allocated in the output
+region.  Finally, the input region is freed on line 14. The output
+region at \C{map[t]} stays as along as input messages with timestamp
+\C{t} keep arriving. When the timing message for \C{t} arrives, the
+\C{onNotify} method transfers the \C{outRgn} at \C{map[t]} to a
+downstream actor.
+
+\begin{figure}[t!]
+\begin{numcodejava}
+class SelectVertex<TIn, TOut> {
+  Func<TIn, TOut> selector;
+  Dictionary<Time, Region<List<TOut>>> map;
+  ...
+  void onReceive(Time t,Region<List<TIn>> inRgn){
+    if (!map.ContainsKey(t))
+       map[t] = new Region<List<TOut>> (() => new List<TOut>());
+    open inRgn as inList {
+      letregion R0 {
+        foreach (TIn input in inList) {
+          open map[t] as outList {
+            TOut output = selector(input);
+            outList.add(output); } } } }
+    inRgn.free();
+  }
+  void onNotify(Time t) {
+     Region<List<TOut>> outRgn = map[t];
+     map.Remove(t);
+     outRgn.transfer(successorId);
+  }
+}
+\end{numcodejava}
+\caption{\C{SELECT} dataflow operator in \name}
+\label{fig:motivating-eg-in-broom}
+\vspace*{-0.15in}
+\end{figure}
+
+\begin{figure} 
+\includegraphics[scale=0.45]{region-fsm.png}
+\caption{The lifetime of a dynamic (transferable) region in \name}
+\label{fig:region-fsm} 
+\vspace*{-0.1in} 
+\end{figure}
+
+\begin{figure*}[t!]
+%
+\begin{smathpar}
+\renewcommand{\arraystretch}{1.2}
+\begin{array}{lclcl} 
+\multicolumn{5}{c}{
+  {\pi} \in \mathtt{Region\; identifiers} \qquad
+  {\rho} \in \mathtt{Region \; variables} \qquad
+% {\loc} \in \mathtt{Memory\; locations}
+  {\tyvar, \tyvarb} \in \mathtt{Type \; variables}}\\
+\multicolumn{5}{c}{
+% {\tyvar, \tyvarb} \in \mathtt{Type \; variables} \qquad
+  {m} \in \mathtt{Method \; names} \qquad
+  {x,y,f} \in \mathtt{Variables \; and \; fields} }\\
+cn & \in & \M{Class \; names} & \coloneqq & \ObjZ \ALT \RgnZ \ALT A \ALT B\\
+     \fgjN & \in & \M{FGJ \; class \; types} & \coloneqq & cn\inang{\tbar}\\
+T  & \in & \M{FGJ \; types} & \coloneqq & \tyvar \ALT  \fgjN \ALT \unitZ
+     \ALT \bar{T} \rightarrow T \\
+r  & \in & \M{Region\; annotations} & \coloneqq & \rho \ALT \pi \\
+% \status & \in & \mathtt{Region\; Typestate} & \coloneqq & \FREE \ALT
+%                   \USED \ALT \LIVE \ALT \XFERRED\\
+\fbN  & \in & \M{Annotated \; class \; types} & \coloneqq & 
+     cn\inang{\tbar}\inang{\rbar} \\
+\tau &\in& \M{types} & \coloneqq & T@\rgn  
+      \ALT \fbN \ALT \unitZ 
+      \ALT \inang{\rhobar \,|\, \phi}\bar{\tau}
+      \xrightarrow{\rgn} \tau \\
+C  & \in & \M{Class \; definitions} & \coloneqq & 
+     \C{class} \; cn\inang{\bar{\tyvar} \extends \bar{\fgjN}} 
+                    \inang{\rhobar \,|\, \phi}\extends \fbN 
+                    \{\bar{\tau} \; \bar{f};\; \bar{d}\}\;\\
+d  & \in & \M{Methods} & \coloneqq & 
+     \tau \; m\inang{\rhobar \,|\, \phi} (\taubar \; \xbar)
+     \{\C{return}\;e;\}\\
+\phi,\phicx &\in& \M{Region\;constraints} & \coloneqq & true 
+      \ALT \rgn \outlives \rgn \ALT \rgn = \rgn \ALT \phi \conj \phi\\
+e  & \in & \M{Expressions} & \coloneqq &
+     \unitval \ALT x \ALT e.f 
+     \ALT e.m\inang{\rbar}(\ebar) \ALT \C{new}\;\fbN(\ebar)
+ \\  & & & & 
+     \ALT \lambdaexp{\rgn}{\rhobar \,|\, \phi}  {\xbar:\taubar} {e}
+     \ALT e\inang{\rbar}(\bar{e})
+     \ALT \letexp{x}{e}{e}
+ \\  & & & & 
+     \ALT \letregion{\pi}{e} 
+      \ALT \open{e}{\pi}{y}{e}\\
+%  & & & & \ALT \letd{\loc}{e} \ALT \opened{\loc}{\status}{e}\\
+\end{array}
+\end{smathpar}
+
+\caption{\fbname: syntax}
+\label{fig:fb-syntax}
+\end{figure*}
+
+\mypara{Cloning} Note that in the example from
+Fig.~\ref{fig:motivating-eg-in-broom} the object returned by the
+\C{selector} (on Line 12) should not contain any references to the
+input object, since the input region, where the object resides, will
+be freed at the end of the method. If there is a need for the output
+object to point to subobjects of the input object, such subobjects
+must be cloned (to copy them from the input region to the output
+region).  Fortunately, \name's region type system
+(\S~\ref{sec:type-system}) is capable of capturing such nuances in the
+type of \C{selector} and the type checker will ensure correctness.
+Furthermore, the type can be automatically inferred by \name's region
+type inference (\S~\ref{sec:type-inference}), which can perform the
+above reasoning on behalf of the programmer.
+
+%--------------------------------------------------------------------
+%                             Type System
+%--------------------------------------------------------------------
+\section{\fbname}
+\label{sec:type-system}
+
+The purpose of \name's region type system is to enforce the key
+invariant required for memory safety, namely that an object $o_1$ in a
+region $R_1$ contains a reference to an object $o_2$ in $R_2$, only if
+$R_2$ is guaranteed to outlive $R_1$.  While the invariant is easily
+stated, enforcing it in the presence of first-class dynamic regions,
+parametric polymorphism (generics), and higher-order functions
+requires new reasoning principles that we formally develop in this
+section. We introduce \fbname (\FB), an explicitly typed core language
+(with region types) that incorporates the features introduced in the
+previous section. \fbname builds on the Featherweight Generic Java
+(FGJ)~\cite{fgj} formalism, and reuses notations and various
+definitions from~\cite{fgj}, such as the definition of type
+well-formedness for the core (region-free) language.  (The language
+used in Section~\ref{sec:overview} is essentially a version of FGJ
+without region types and with some syntactic sugar.)
+
+\subsection{Syntax}
+\label{sec:fb-syntax}
+
+Fig~\ref{fig:fb-syntax} describes the syntax of \FB.
+We refer to the class types of FGJ as \emph{core types}.
+%
+The following definition of \C{Pair} class in \FB illustrates some of
+the key elements of the formal language (the symbol $\extends$ should
+be read \emph{extends}, and the symbol $\outlives$ stands for
+\emph{outlives}):
+\begin{codejava}[mathescape=true]
+class Pair<a $\extends$ Object, b $\extends$ Object>
+          <$\rho_0, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
+          \rho_2 \outlives \rho_0$> $\extends$ Object<$\rho_0$> {
+  a@$\rho_1$ fst; 
+  b@$\rho_2$ snd;
+  a@$\rho_1$ getFst() { return this.fst; }
+}
+\end{codejava}
+Note that we elide showing constructors since they are uninteresting
+from the type system's standpoint; their behavior in \FB is same as in
+FGJ.
+
+A class in \FB is parametric over zero or more type variables (as in
+FGJ) as well as one or more region variables $\rho$.  We refer to the
+first region parameter ($\rho_0$ in the above example) as the
+\emph{allocation region} of the class: it serves to identify the
+region where an instance of the class is allocated\footnote{In
+general, $\rhobar$ denotes the sequence $\rho_1\rho_2...$ of region
+parameters of a class or a method. In some cases, we also represent
+region parameters as $\rho\rhobar$ to clearly distinguish between the
+allocation region parameter ($\rho$) and the rest.}.
+%
+An object in \FB can contain fields referring to objects allocated in
+regions ($\rhobar$) other than its own allocation region ($\rho$),
+provided that the former outlive the latter (i.e., $\rhobar \outlives
+\rho$). In such case, the definition of object's class needs to be
+parametric over allocation regions of its fields (i.e., their
+classes). Furthermore, the constraint that such regions must outlive
+the allocation region of the class needs to be made explicit in the
+definition, as the \C{Pair} class does in the above definition. We say
+that the \C{Pair} class exhibits \emph{constrained region
+polymorphism}.
+
+To construct objects of the \C{Pair} class, its type and region
+parameters need to be instantiated with core types ($T$) and region
+annotations\footnote{Region annotations ($\rgn$) include region
+variables ($\rho$) and region identifiers ($\pi$). Region identifiers
+are to region variables, as types ($T$) are to type variables ($a$)}
+($\rgn$), respectively. For example:
+\begin{codejava}
+
+letregion $\pi_0$ in
+  let snd = new Object<$\pi_0$>() in
+  letregion $\pi_1$ in
+    let fst = new Object<$\pi_1$>() in
+    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
+\end{codejava}
+%\vspace*{-0.1in}
+In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
+$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$ and
+$\pi_1$ are live during the instantiation, and (b). $\pi_0 \outlives
+\pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is reflexive).
+Observe that the region type of \C{p} conveys the fact that (a). it is
+allocated in region $\pi_1$, and (b). it holds references to objects
+allocated in region $\pi_0$ and $\pi_1$.  In contrast, if we choose to
+allocate the \C{snd} object also in $\pi_1$, then \C{p} would be
+contained in $\pi_1$, and its region type would be
+\C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which we abbreviate as
+\C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
+$B\inang{\tbar}@\pi$ as being equivalent to
+$B\inang{\tbar}\inang{\bar{\pi}}$. Region annotation on type $a$,
+where $a$ is a type variable, assumes the form $a@\pi$. If $a$ is
+instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
+\C{Pair} object contained in  $\pi$.
+
+Classes in \FB are independently parameterized over types and regions.
+While this design decision has a downside in that it allows type
+variables to only denote the types of objects contained in a single
+region, it yields benefits that outweigh the costs. In particular, it
+lets us support region-polymorphic higher-order functions as class
+fields.  This allows, for example, a generic class, whose type
+parameters are $a$ and $b$, to contain a region-polymorphic function
+of type $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0
+\conj \rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
+\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
+region-polymorphic higher-order fields are used frequently by the
+dataflow operators, which apply them in the context of various regions
+(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
+and region parameterizations separate also simplifies the type system
+so that inference becomes practical (Sec.~\ref{sec:type-inference}).
+The need for polymorphism at the field level is also why \FB treats
+function closures specially, rather than as objects of type \C{Func}
+as in C\#.
+
+Like classes, methods can also exhibit constrained region
+polymorphism.  A method definition in \FB is necessarily polymorphic
+over its allocation context (\S~\ref{sec:alloc-ctxt}), and optionally
+polymorphic with respect to the regions containing its arguments
+(i.e., a method has at least one region parameter). Region parameters,
+like those on classes, are qualified with constraints ($\phi$).  If a
+method is not intended to be polymorphic with respect to its
+allocation context (for example, if its allocation context needs to be
+same as the allocation region of its \emph{this} argument), then the
+required monomorphism can be captured as an equality constraint in
+$\phi$.
+
+\FB extends FGJ's expression language with a lambda expression and an
+application expression ($e\inang{\rbar}(\bar{e})$) to define and apply
+functions (lambdas). Functions, like methods, exhibit constrained
+region polymorphism, as evident in their arrow region type
+($\inang{\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn} \tau$).  A
+function, like a method, is necessarily polymorphic w.r.t its
+allocation context.  Since a function closure can escape the context
+in which it is created, it is important to keep track of the region in
+which it is created in order to avoid unsafe dereferences. The $\rgn$
+annotation above the arrow in the arrow type denotes the allocation
+region of the corresponding closure.
+
+Note that mutable object fields are conspicuously absent from the \FB
+model (and also the FGJ model in general), but this isn't a major
+shortcoming considering that constructor application, which happens
+during a new object creation, includes assignments to the object
+fields (see FGJ~\cite{fgj}). Thus the type system is already obligated
+to handle unsafe assignments.
+
+\subsection{Types and Well-formedness}
+
+Well-formedness and typing rules of \fbname establish the conditions
+under which a region type is considered well-formed, and an expression
+is considered to have a certain region type, respectively.
+Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
+rules\footnote{Full formal development can be found in the
+appendix}. The rules refer to a context ($\A$), which is a tuple of:
+\begin{itemize}
+  \item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
+  live,
+  \item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
+  variables to their \emph{bounds}, i.e., classes they are declared to
+  extend (this artifact is inherited from FGJ), and
+  \item A constraint formula ($\phicx$) that captures the outlives
+  constraints on regions in $\rhoenv$.
+\end{itemize}
+In addition, the context for the expression typing judgment also
+includes (a).~a type environment ($\env \in x \mapsto \tau$) that
+contains the type bindings for variables in scope, and (b).~the region
+($\rgn$) that serves as the allocation context for the expression
+being type checked. Like the judgments in FGJ~\cite{fgj}, all the
+judgments defined by the rules in Fig.~\ref{fig:fb-staticsem} are
+implicitly parameterized on a class table ($CT \in cn \mapsto D$) that
+maps class names to their definitions in \FB.
+
+\begin{figure*}[t!]
+
+%%%%%%%%%%%%%%%%%%%%%%% WF RULES %%%%%%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Type Well-formedness}  \; \fbox {\(\tywf{\A}{\tau} \)}\\
+%
+\begin{minipage}{2.55in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    CT(B) = \headerOf{B}\{...\}\\
+    \rbar \in \rhoenv \spc
+%   \mem(\rbar) = \overline{\LIVE} \spc
+    \fgjtywf{\aenv}{B\inang{\tbar}}\spc
+%   \substFn = [\rbar/\rhobar, \tbar/\bar{\tyvar}] \spc
+    \isvalid{\phicx}{[\rbar/\rhobar](\phi)}
+  }
+  {
+    \tywf{(\rhoenv,\aenv,\phicx)}{B\inang{\tbar}\inang{\rbar}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{1.25in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \fgjtywf{\aenv}{T}\spc
+%   \rgn \in \A.\rhoenv\\
+    \rgn \in \rhoenv\\
+    \fgjsubtyp{\aenv}{T}{\ObjZ}\spc
+  }
+  {
+    \tywf{(\rhoenv,\aenv,\phicx)}{T@\rgn}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{0.9in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  { 
+    \\
+    \A = (\subtypcx) \spc
+    \fgjtywf{\aenv}{T}
+  }
+  {
+    \tywf{\A}{\RgnZ\inang{T}\inang{\toprgn}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\bigskip
+%%%%%%%%%%%%%%%%%%%%%%%% TYPE RULES %%%%%%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Expression Typing}  \; \fbox
+  {\(\hastyp{\exptycx{\env}{\rgn}}{e}{\tau} \)}\\
+%
+\begin{minipage}{2.2in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\bar{e}}{\taubar} \\
+    \tywf{\A}{\fbN} \spc
+    \fields(\fbN) = \bar{f} : \taubar \\
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\C{new}\; \fbN(\bar{e})}{\fbN}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{2in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+%   We don't need the following condition because of the
+%   guarantee provided by the lambda type rule.
+%   \A = (\subtypcx) \spc
+%   \rgn \in \rhoenv \\
+    \\
+    \hastyp{\exptycx{\env}{\rgn}}{e} {\inang{\rho} \unitZ \xrightarrow{\rgn} T@\rho} \spc
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\C{new}\;
+    \RgnZ\inang{T}\inang{\toprgn}(e)} {\RgnZ\inang{T}\inang{\toprgn}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\begin{minipage}{2.2in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e}{\tau'}\\
+    \bar{f}:\taubar \,=\, \fields(\bound_{\A.\aenv}(\tau'))
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e.f_i}{\tau_i}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{2.9in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \A = (\subtypcx) \spc
+    \A' = (\rhoenv\cup\{\pi\}, \aenv, \phicx \conj \phi)\\
+    \pi \notin \rhoenv \spc
+%   \Delta = \{\rgn \,|\, \rgn \in \rhoenv\} \spc
+    \phi = \Delta \outlives \pi \spc
+    \hastyp{\A',\env,\pi}{e}{\tau} \spc
+    \tywf{\A}{\tau}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\letregion{\pi}{e}}{\tau}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\begin{minipage}{2.5in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \A = (\subtypcx) \spc
+    \pi \notin \rhoenv \\
+    \hastyp{\exptycx{\env}{\rgn}}{e_a}
+            {\RgnZ\inang{T}\inang{\toprgn}}\\
+%   \rgn \notin dom(\mem) \spc
+    \tywf{\A}{\tau} \spc
+    \A' = (\rhoenv\cup\{\pi\},\aenv,\phicx) \\
+    \env' =  \env[y\mapsto T@\pi]\spc
+    \hastyp{\A',\env',\pi}{e_b}{\tau} \spc
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\open{e_a}{\pi}{y}{e_b}}
+            {\tau}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{2.6in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \A = (\subtypcx) \spc
+    \rgn \in \rhoenv \spc
+    \rhoenv \outlives \rgn \\
+    \rho,\rhobar \notin \rhoenv \spc
+    \A' = (\rhoenv \cup \{\rho,\rhobar\}, \aenv, 
+          \phicx \conj \phi)\\
+    \tywf{\rhoenv \cup \{\rho,\rhobar\}}{\phi}\spc
+    \tywf{\A'}{\bar{\tau^1}}\\
+    \tywf{\A'}{\tau^2}\spc
+    \hastyp{\A',\env[\xbar \mapsto \bar{\tau^1}],\rho}{e}{\tau^2}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}
+           {\lambdaexp{\rgn}{\rho\rhobar \,|\, \phi}
+                      {\xbar:\bar{\tau^1}}{e}}
+           {\inang{\rho\rhobar \,|\, \phi}
+            \bar{\tau^1} \xrightarrow{\rgn} \tau^2}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\begin{minipage}{2.5in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \A = (\subtypcx) \spc
+    \hastyp{\exptycx{\env}{\rgn}}{e_0}{\tau} \\
+%   \rbar \in \A.\rhoenv \\
+    \mtype(m,\bound_{\aenv}(\tau)) = \inang{\rho\rhobar \,|\, 
+        \phi}\bar{\tau^1}\rightarrow{\tau^2} \\
+    \rgn,\rbar \in \rhoenv \spc
+    \tywf{\A}{\inang{\rho\rhobar \,|\,
+    \phi}\bar{\tau^1}\rightarrow{\tau^2}}\\
+    \hastyp{\exptycx{\env}{\rgn}}{\bar{e}}{[\rgn\rbar/\rho\rhobar]\,\bar{\tau^1}} \spc
+    \isvalid{\phicx}{[\rgn\rbar/\rho\rhobar]\,\phi}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e_0.m\inang{\rgn\rbar}(\bar{e})} 
+           {[\rgn\rbar/\rho\rhobar]\,\tau^2}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{2.3in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \\
+    \A = (\subtypcx) \spc
+    \rgn',\rgn,\rbar \in \rhoenv \\
+    \hastyp{\exptycx{\env}{\rgn}}{e}
+        {\inang{\rho\rhobar \,|\, \phi}
+            \bar{\tau^1} \xrightarrow{\rgn'} \tau^2}\\
+%   \substFn = \subst{\rbar}{\rhobar} \spc
+    \isvalid{\phicx}{\subst{\rgn\rbar}{\rho\rhobar}\phi} \spc
+    \hastyp{\exptycx{\env}{\rgn}}{\bar{e}}
+        {\subst{\rgn\rbar}{\rho\rhobar}\bar{\tau^1}}\spc
+%   \subtyp{\A}{\taubar}{\bar{\tau^1}}
+  }
+  {
+    \hastyp{\A,\env,\rgn}{e\inang{\rgn\rbar}(\bar{e})}
+           {\subst{\rgn\rbar}{\rho\rhobar}\tau^2}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\bigskip
+
+%%%%%%%%%%%%%%%%%%%%% METHOD WF %%%%%%%%%%%%%%%%%%%%%%%
+\textbf{Method Well-formedness}  \; \fbox
+  {\(d \; \mathtt{ok \; in} \; B \)}\\
+%
+\begin{minipage}{5.5in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+%   \rhoenv = \{\rhobar,\bar{\rho_m}\} \spc
+%   \aenv = [\bar{\tyvar} \mapsto \bar{\fgjN}] \spc
+%   \phicx = \phi \conj \phi_m \spc
+    \Delta = \{\rhobar,\rho_m, \bar{\rho_m}\} \spc
+    \A = (\Delta, [\bar{\tyvar} \mapsto \bar{\fgjN}], \phi_m)\spc
+    \tywf{\Delta}{\phi_m} \\
+    \tywf{\A}{\bar{\tau^1},\tau^2} \spc
+%   \tywf{\A}{\tau^2} \spc
+    CT(B) = \headerOf{B}\{...\}\\
+    \env = \cdot[\xbar\mapsto\bar{\tau^1}][\thisZ\mapsto
+            B\inang{\bar{\tyvar}}\inang{\rhobar}] \spc
+    \A \vdash \override(m,\fbN,\inang{\rho_m\rhobarm \,|\, \phi_m}\bar{\tau^1} 
+             \rightarrow \tau^2) \spc
+%   \stmtsem{\rhoallocm}{s}{\stmtsemcx}{\stmtsemcxp}\spc
+    \hastyp{\exptycx{\env}{\rho_m}}{e}{\tau^2} \spc
+%   \subtyp{\A}{\tau}{\tau^2}
+  }
+  {
+    \okin 
+        {\tau^2 \; m\inang{\rho_m\rhobarm \,|\, \phi_m}
+              (\bar{\tau^1}\;\xbar)\{\C{return}\,e;\}}
+        {B}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\caption{\fbname: select type rules}
+\label{fig:fb-staticsem}
+\vspace*{-0.1in}
+\end{figure*}
+
+\noindent The well-formedness judgment on region types
+($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
+judgments on core types. We use a double-piped turnstile ($\Vdash$)
+for judgments in FGJ~\cite{fgj}, and a simple turnstile ($\vdash$) for
+those in \FB. The class table ($\absof{CT}$) for FGJ judgments is
+derived from \FB's class table ($CT$) by erasing all region
+annotations on types, and region arguments in expressions
+($\absof{\cdot}$ denotes the region erasure operation).  The
+well-formedness rule for class types ($B\inang{\tbar}\inang{\rbar}$)
+is responsible for enforcing the safety property that prevents objects
+from containing unsafe references. It does so by insisting that
+regions $\rbar$ satisfy the constraints ($\phi$) imposed by the class
+on its region parameters.  The latter is enforced by checking the
+validity of $\phi$, with actual region arguments substituted for
+formal region parameters, under the conditions ($\phicx$) guaranteed
+by the context. The semantics of this sequent is straightforward, and
+follows directly from the properties of outlives and equality
+relations. For any well-formed core type $T$, $T@\rgn$ is a
+well-formed region type if $\rgn$ is a valid region.  The type
+$\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
+$\toprgn$ is a special immortal region that outlives every other live
+region. This arrangement allows \C{\RgnZ} handlers to be aliased and
+referenced freely from objects in various regions, regardless of their
+lifetimes. On the flip side, this also opens up the possibility of
+references between transferable regions, which become unsafe in
+context of the recipient's address space. Fortunately, such references
+are explicitly prohibited by the type rule of \C{\RgnZ} objects, as
+described below.
+
+The type rules distinguish between the \C{new} expressions that create
+objects of  the \RgnZ class, and \C{new} expressions that create
+objects of other classes. The rule for the latter relies on an
+auxiliary definition\footnote{All auxiliary definitions we use in this
+exposition originate from the FGJ calculus.} called $\fields$
+(undefined for $\RgnZ$ class) that returns the sequence of type
+bindings for fields (instance variables) of a given class type.  Like
+in FGJ, the names and types of a constructor's arguments in \FB are
+same as the names and types of its class's fields. The type rule for
+the field access expression ($e.f_i$) also uses $\fields$ and another
+definition called $\bound$, which returns the bound of a type variable
+($\bound$ is an identity function for concrete types).
+
+The type rule for the \C{new \RgnZ} expression expects the \RgnZ class's
+constructor to be called with a nullary function that returns a value
+in its allocation context.  This ensures that the value returned by
+the function stores no references to objects allocated elsewhere,
+including the top region ($\toprgn$), thus preventing cross-region
+references originating from transferable regions. The body of the
+function might however create new regions while execution, but this is
+not a problem as long as such regions, and objects allocated in them,
+don't find their way into the result of its evaluation.
+
+The type rule for the \C{letregion} expression requires that the static
+identifier introduced by the expression be unique under the current
+context (i.e., $\pi \notin \Delta$). This condition is needed in order
+to prevent the new region from incorrectly assuming existing outlives
+relationships on an eponymous region. The expression ($e$) under
+\C{letregion} is typechecked with the new region as its allocation
+context, under the assumption that the region is live ($\Delta \cup
+\{\pi\}$) and that it is outlived by all existing live regions
+($\Delta \outlives \pi$). The result of a \C{letregion} expression
+must have a type that is well-formed under a context not containing
+the new region. This ensures that the value obtained by evaluating a
+\C{letregion} expression contains no references to the temporary
+objects inside the region.
+
+The rule for the \C{open} expression, unlike the rule for
+\C{letregion}, does not introduce any outlives relationship between
+the newly opened region and any pre-existing region while checking the
+type of the expression ($e$) under \C{open}. This prevents new objects
+allocated inside the transferable region from storing references to
+those outside. The newly opened region becomes the allocation context
+for $e$, which is type checked under an environment ($\env$) extended
+with the type binding for the root object.
+
+The type rule for the lambda expression typechecks the lambda-bound
+expression ($e$) under an extended type environment containing
+bindings for function's arguments, assuming that region parameters are
+live, and that declared constraints over region parameters hold. The
+constraints ($\phi$) are required to be well-formed under $\Delta$
+extended with the function's region parameters ($\rho\rhobar$).
+Unlike a typical object, a function closure might capture references
+that may become unsafe if the closure escapes its allocation context.
+\FB prevents this scenario by requiring the function closure to always
+be allocated in the current allocation context ($\rgn$).
+
+The type rule for method application uses an auxiliary definition
+$\mtype$ that gives the type of a class method. Method application
+involves instantiation of method's region parameters, and the type
+rule requires the first region parameter to be instantiated with the
+current allocation context ($r$). The type rule for function
+application does similarly.  This requirement ensures the safety of
+closures as demonstrated by the following example.
+
+Consider a method $m$ with two region parameters - $\rho_0$ and
+$\rho_1$ (i.e., $m\inang{\rho_0,\rho_1}(\dots)$). Suppose the method
+immediately returns a function closure that contains a reference to
+(an object in) $\rho_1$. Since function closures are always allocated
+in the current allocation context, the closure is allocated in
+$\rho_0$, the allocation context for the method body (the rule for
+methods discussed below). Now, consider the following use of the
+method:
+\begin{codejava}
+letregion R0 in
+  let f = letregion R1 in m<R0,R1>()
+  in f()
+\end{codejava}
+Observe that the first region parameter of \C{m} is instantiated with
+\C{R0}, while the allocation context for the method call is \C{R1}.
+The function \C{f} returned by \C{m} stores a reference to \C{R1},
+which is unsafe when \C{f} is finally called. The type system
+fortunately disallows this scenario by enforcing certain restrictions
+during method and function applications as described above.
+
+The method well-formedness rule makes use of an auxiliary definition
+$\override$ that judges whether the current method is a valid
+overriding of any eponymous method from the super class. The type
+environment is extended with a binding that binds the $\thisZ$ keyword
+to the type of the current class. Note that the type of the
+current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
+region-parametric, thus admitting region-polymorphic recursion (i.e.,
+recursive call to $\C{m}$ can have different region arguments than
+$\C{m}$).
+
+\subsection{Operational Semantics and Type Safety}
+\label{sec:fb-opsem}
+
+\begin{figure*}[t!]
+%
+\begin{smathpar}
+\renewcommand{\arraystretch}{1.2}
+\begin{array}{lclcl} 
+{\loc} & \in & \mathtt{Memory\; locations} & & \\
+r  & \in & \M{Region\; annotations} & \coloneqq & \loc \ALT \dots \\
+\status & \in & \mathtt{Region\; Typestate} & \coloneqq & %\FREE \ALT
+                  \USED \ALT \LIVE \ALT \XFERRED\\
+e  & \in & \M{Expressions} & \coloneqq &  \letd{\loc}{e} \ALT 
+                        \opened{\loc}{\status}{e} \ALT \bot \ALT \dots \\
+\end{array}
+\end{smathpar}
+
+\caption{\fbname: extended syntax to accommodate run-time constructs}
+\label{fig:fb-syntax-extra}
+\end{figure*}
+
+The operational semantics of \FB describes a non-trivial runtime
+component that introduces memory regions and performs run-time
+verification of their typestate. Fig.~\ref{fig:fb-syntax-extra} shows the
+additional language constructs of \FB that manifest only at run-time.
+A location ($\loc$) abstracts all the memory locations associated with
+a region (i.e., each memory region is associated with a single
+location). A transferable region can be in one of three possible
+states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
+($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
+manifestations of \C{letregion} and \C{open}; \C{letregion} reduces to
+\C{letd} while allocating a (static) region, and \C{open} reduces to
+\C{opened} while opening a (transferable) region. An \C{opened}
+expression is tagged with the typestate ($s$) of the transferable
+region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
+transferable region can be \C{open}'d from either an open state or a
+closed state). Special value $\bot$ denotes an exception.
+
+Operational semantics of \FB defines a four-place small-step reduction
+relation of the form shown below:
+% \vspace*{-0.15in}
+\begin{smathpar}
+  \redstoo{\rhoenv}{(e,\mem)}{(e',\mem')}
+\end{smathpar}
+The typestate of regions is tracked by a finite map ($\mem$) from
+locations to typestates (Fig.~\ref{fig:region-fsm}). The reduction
+relation relates an expression ($e$) and a typestate map ($\mem$) to a
+reduced expression ($e'$) and an updated typestate map ($\mem'$). The
+semantics gets ``stuck'' if $e$ attempts to access an object whose
+allocation region is not present in $\rhoenv$, or if $e$ tries to
+\C{open} a transferable region that is not mapped to an appropriate
+typestate by $\mem$.  On the other hand, if $e$ attempts to commit an
+operation on a $\RgnZ$ object that is not sanctioned by the transition
+discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
+value ($\invalidexn$). An illustrative subset of operational semantics
+that formalize the intuitions described above is shown in
+Fig.~\ref{fig:fb-opsem}.  Fig.~\ref{fig:fb-opsem-1} of the appendix
+contains rest of the rules.
+
+\newcommand{\opsemrule}[3]{%
+\begin{minipage}{#1}\begin{smathpar}\begin{array}{c}%
+\renewcommand*{\arraystretch}{1.2}%
+\RULE {#2} {#3}%
+\end{array}\end{smathpar}\end{minipage}%
+}
+\newcommand{\lopsemrule}[4]{%
+\begin{minipage}{#1}\begin{smathpar}\begin{array}{c}%
+\renewcommand*{\arraystretch}{1.2}%
+[\rulelabel{#4}] \spc \RULE {#2} {#3}%
+\end{array}\end{smathpar}\end{minipage}%
+}
+%
+\newcommand{\anobjty}[0]{B\inang{\tbar}\inang{\ralloc\locbar}}
+\newcommand{\anobj}[0]{\C{new} \; \anobjty(\bar{v})}
+%
+
+\begin{figure*}[t!]
+
+%
+\fbox {\(\redstoo{\Delta}{(e,\mem)}{(e',\mem')}\)}\\
+
+\lopsemrule{3.2in}{
+%   \fresh(\rgn')\spc
+%   \mem(\loc) = \FREE \spc
+    \loc \not\in dom(\mem) \spc
+    \mem' = \mem[\loc \mapsto \SLIVE]
+%   \redstoo{\Delta}{(e,\mem)}{(e',\mem')}
+}{
+    \redstoo{\Delta}{(\letregion{\rgn}{e},\mem)}{(\letd{\loc}{[\loc/\rgn]e},\mem')}
+  }{LetRegionBegin}
+
+\lopsemrule{3in}{
+    \redstocup{\loc}{(e,\mem)}{(e',\mem')}
+}{
+    \redstoo{\Delta}{(\letd{\loc}{e},\mem)}{(\letd{\loc}{e'},\mem')}
+  }{LetRegion}
+
+\lopsemrule{3in}{
+%   \rgn \notin \rhoenv
+%   \mem' = \mem[\loc \mapsto \FREE]
+    \mem' = \mem[\loc \mapsto \XFERRED]
+}{
+    \redstoo{\Delta}{(\letd{\loc}{v},\mem)}{(v,\mem')}
+  }{LetRegionEnd}
+
+\lopsemrule{3.3in}{
+    \loc \not\in dom(\mem) \spc
+    \mem' = \mem[\loc \mapsto \USED]
+}{
+    \redstoo{\Delta}{(\C{new} \; \RgnZ\inang{T}\inang{\toprgn}
+                (v),\mem)}
+            {(\C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}
+                (v\inang{\loc}()),\mem')}
+	      }{NewRegion}
+
+\lopsemrule{3.3in}{
+%   \mem(\loc) = \USED \spc % Type system cannot guarantee this.
+    \redstocup{\loc}{(e,\mem[\loc \mapsto \LIVE]) }{(e',\mem')}
+}{
+    \redstoo{\Delta}{(\C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}
+                (e),\mem)}
+            {(\C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}
+                (e'),\mem')} %[\loc\mapsto\USED])}
+	      }{NewRegion}
+
+\lopsemrule{4in}{
+    v_a = \C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}(v) \spc
+    \mem(\loc) = \USED ~\texttt{or}~ \mem(\loc) = \TLIVE \spc
+%   \fresh(\rgn_1)\\
+%   \rgn_0 \notin \rhoenv \\
+    \mem' = \mem[\loc \mapsto \TLIVE]
+%   \mem'' = \mem'[\rgn_r \mapsto \mem(\rgn_r)]
+}{
+    \redstoo{\Delta}{(\open{v_a}{\rgn}{x}{e_b},\mem)} 
+            {(\opened{\loc}{\mem(\loc)}{[v/x][\loc/\rgn]e_b},\mem')}
+	  }{Open}
+
+\lopsemrule{2.7in}{
+    \redstocup{\loc}{(e,\mem)}{(e',\mem')}
+}{
+    \redstoo{\Delta}{(\opened{\loc}{\status}{e},\mem)}
+    {(\opened{\loc}{\status}{e'},\mem')}
+  }{Opened}
+
+\lopsemrule{2.7in}{
+    \mem' = \mem[\loc \mapsto \status]
+}{
+    \redstoo{\Delta}{(\opened{\loc}{\status}{v},\mem)} {(v,\mem')}
+  }{OpenEnd}
+
+\lopsemrule{3in}{
+    v_a = \C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}(v) \spc
+    \mem(\loc) \neq \USED ~\texttt{and}~ \mem(\loc) \neq \LIVE
+}{
+    \redstoo{\Delta}{(\open{v_a}{\rgn}{x}{e_b},\mem)} 
+            {\invalidexn}
+	  }{OpenTransferred}
+
+\lopsemrule{3.25in}{
+    \mem(\loc) = \USED \spc
+    \mem' = \mem[\loc \mapsto \XFERRED]
+}{
+    \redstoo{\Delta}{((\C{new}\;\RgnZ\inang{T}\inang{\toploc\loc}(v)).\transfer(\ldots),\mem)}
+            {(\unitval,\mem')}
+	  }{Transfer}
+
+\lopsemrule{2.5in}{
+    \mem(\loc) = \TLIVE \spc
+}{
+    \redstoo{\Delta}{((\C{new}\;\RgnZ\inang{T}\inang{\toploc\loc}(v)).\transfer(\dots),\mem)}
+            {\invalidexn}
+	  }{TransferOpened}
+
+\caption{\fbname: a select subset of operational semantics }
+\label{fig:fb-opsem}
+\end{figure*}
+
+To help state the type safety theorem, we define the syntactic class
+of (runtime) values:
+\begin{smathpar}
+\begin{array}{lclcl}
+v & \in & \mathtt{values} & \coloneqq & \C{new}\;
+    B\inang{\tbar}\inang{\overline{\loc}}(\bar{v}) \ALT
+\lambdaexp{\loc}{\rhobar\,|\,\phi}{\taubar \; \xbar}{e} \ALT
+\C{new}\; \RgnZT{\toploc\loc}(v)
+%  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
+\end{array}
+\end{smathpar}
+The first two forms are obtained by using locations for region
+annotations in $\C{new}$ and lambda expressions. The last form is the
+value that $\C{new}\;\RgnZ$ expressions gets evaluated to; the
+location $\toploc$ stands for the region $\toprgn$, and $\loc$ is the
+location of the newly allocated transferable region.  The following
+type safety theorem shows that a well-typed program will never attempt
+to dereference an ``invalid'' reference (a reference to an object in a
+region that has been transferred or freed):
+\begin{theorem}
+\emph{(\textbf{Type Safety})}
+\label{thm:fb-type-safety}
+$\forall e, \tau, \Delta, \mem$, such that $\consistent(\Delta,\mem)$
+and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,
+\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
+exception ($\redstoo{\Delta}{(e,\mem)}{\invalidexn}$), or there exists an
+$e'$ and a $\mem'$ such that $\consistent(\Delta,\mem')$ and
+$\redstoo{\Delta}{(e,\mem)}{(e',\mem')}$ and
+$\hastyp{\emptyASigp,\cdot}{e'}{\tau}$.
+\end{theorem}
+The relation $\consistent$ relates $\Delta$ and $\mem$ only if both
+make consistent assumptions about liveness of regions. Since $\Delta$
+is consistent with both $\mem$ and $\mem'$, the theorem also captures
+the key property of operational semantics that no live region is ever
+freed or transferred.
+
+{\bf Integrating regions with GC heap}. \FB stores the region handlers in a
+distinguished top region, which abstracts the GC heap. Type system
+prevents transferable regions from storing references into the GC
+heap. A static region however can store references to region handles,
+which need to be taken into account while GC'ing the heap. Traversing
+the region memory to identify references into the GC heap
+unfortunately defeats the purpose of regions. The solution we adopt is
+to disable collecting region handles as long as any static region is
+open. Since static regions are intended to store temporary objects
+while populating a dynamic region, their lifetimes are short enough to
+let region handles to be GC’d.
+
+%--------------------------------------------------------------------
+%                             Type Inference
+%--------------------------------------------------------------------
 \renewcommand{\rgn}{\pi}
-\input{type-inference}
-\input{misc}
+\newcommand{\deltaPC}{\Delta_P^C}
+
+\newcommand{\soln}{\eta}
+\newcommand{\solnR}{\soln_R}
+\newcommand{\solnP}{\soln_P}
+\newcommand{\thesoln}{\hat{\eta}}
+\newcommand{\thesolnR}{\thesoln_R}
+\newcommand{\thesolnP}{\thesoln_P}
+
+\newcommand{\saturate}[1]{{#1}^*}
+\newcommand{\myground}[1]{\saturate{#1}_g}
+\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
+\newcommand{\satC}{\saturate{C}}
+\newcommand{\groundC}{\myground{C}}
+\newcommand{\rhoC}{\text{WF}_R}
+\newcommand{\solveCon}[1]{\textsc{Solve}(#1)}
+
+\section{Type Inference}
+\label{sec:type-inference}
+
+\name's region type system imposes an annotation burden, and
+manually annotating C\# standard libraries with region types
+can be tedious. We now present our region type inference algorithm
+that eliminates the need to write region type annotations.
+% except on some higher-order functions.
+Formally, the type inference
+algorithm is an elaboration function from programs in $\absof{\FB}$
+(i.e., \FB without region types, but with \C{letregion} and \C{open}
+expressions, similar to the language introduced in
+\S~\ref{sec:overview}) to programs in \FB.
+For simplicity, we assume that each \C{letregion} and \C{open} construct
+in the input program introduces a distinct region identifier $\pi$.
+
+\mypara{Overview.}
+We now present a high-level outline of the type inference algorithm.
+The algorithm consists of the following steps:
+\begin{enumerate}
+ \item \emph{Region Parameterization}.
+   The first step elaborates the input program by introducing \emph{formal region parameters}
+   (for each class and method), and \emph{region variables} (representing yet undetermined
+   \emph{actual region parameters}). We also introduce for each class and method, a
+   \emph{predicate variable} ($\varphi$) to denote an undetermined set of outlives-constraints
+   over the region parameters of that class/method.
+
+ \item \emph{Constraint Generation}.
+   In the second step, we analyze the program to generate a set of constraints
+   (over the region identifiers and the predicate variables)
+   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).
+
+ \item \emph{Constraint Solving}.
+   We solve the generated set of constraints using our fixpoint constraint
+   solving algorithm, which reduces the constraint solving problem to
+   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
+   references, for example, a reference from a transferable region to a
+   stack region, then the constraints generated during the elaboration
+   are not satisfiable. In such a case, the solver fails to solve
+   the constraints.
+
+ \item If the solver succeeds, it returns a solution $\soln$ consisting of a
+   pair of substitution functions $\solnR$ and $\solnP$ for
+   free region and predicate variables, respectively, introduced in step 1.
+   We apply these substitutions to the elaborated program to produce the final program.
+\end{enumerate}
+
+We later on establish the soundness of the type inference algorithm,
+and the completeness of the constraint-generation and
+constraint-solving steps (steps 2 and 3 above).
+
+\subsection{Region Parameterization for Classes}
+\label{sec:fb-templatization}
+
+Region parameterization is an iterative process involving the
+following three steps, the first two of which are mutually dependent
+on each other.
+
+\emph{Introduction of Formal Region Parameters}.  For every class
+\C{C}, we identify a sequence of formal region parameters $\pi_0,
+\cdots \pi_n$ that \C{C} should be parametric over.
+
+\emph{Introduction of Actual Region Parameters}.  We then replace
+every instance of class \C{C} in the program by an instance
+$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots,
+\rho_n$ are fresh identifiers denoting actual region parameters.
+
+\emph{Predicate Variable Introduction}. For every class \C{C}, we
+introduce a fresh predicate variable $\varphi$, which represents the
+yet undetermined outlives-constraints between the formal region
+parameters of class \C{C}.
+
+We identify the region parameters of classes as follows.
+
+\emph{Non-Recursive Classes}.  The class \C{Object} is defined to have
+a single region parameter $\pi_0$ (the allocation region).  The region
+parameters for any other non-recursive class \C{C} is determined only
+after the region parameters of any class that \C{C} depends on have
+been determined: this includes the base-class \C{B} of \C{C} and the
+class (type) of any of its fields.  We replace every dependee type
+\C{T} in \C{C} by its instantiated type, using fresh region parameters
+as needed.  The sequence of region parameters for \C{C} is defined to
+be the sequence of region parameters for the base class \C{B}
+concatenated with the list of all  fresh region parameters introduced
+while instantiating the types of the fields in the class.  (The class
+inherits its allocation region from its base class. Note that if a
+class does not specify an explicit base class, it has an implicit base
+class \C{Object}.)
+
+This transformation is illustrated below, using a non-generic \C{Pair} class:
+
+\begin{tabular}{ccc}
+\begin{minipage}{0.33\linewidth}
+\begin{codejava}
+class Pair $\extends$ $\ObjZ$ {
+  $\ObjZ$ fst;
+  $\ObjZ$ snd;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow \; \;$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
+  $\ObjZ \langle \rho_1 \rangle$ fst;
+  $\ObjZ \langle \rho_2 \rangle$ snd;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
+
+\emph{Recursive Classes}.
+The region parameters for a recursive class is computed in
+a similar fashion, with the following difference: any recursive
+field is ignored while instantiating region parameters for the fields of
+the class, and the region parameters of the recursive class are computed
+as before. We then do parameter instantiation for all recursive fields,
+such that their region annotations (the actual region parameters) are
+exactly the same as the (formal) region parameters of the class.
+The following example illustrates this for a non-generic \C{List} class.
+The resulting class represents a linked list with spine in the region
+$\rho_0$ and data objects in the region $\rho_1$.
+
+\begin{tabular}{ccc}
+\begin{minipage}{0.33\linewidth}
+\begin{codejava}
+class List $\extends$ $\ObjZ$ {
+  $\ObjZ$ data;
+  List next;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow \; \;$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$ {
+  $\ObjZ \langle \rho_1 \rangle$ data;
+  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
+
+The above technique can be extended to mutually recursive classes in a
+straightforward manner, by simultaneously parameterizing them (and
+then instantiating them).
+
+\emph{Type-Parametric Classes}.  The type parameter \C{T} of a class
+\C{C} is instantiated as $\C{T}@\rho$ using a single region parameter
+$\rho$.  (This can be extended to use the bound specified for \C{T},
+if any.)
+
+\emph{Function Types}.  Since \FB{} is higher-order, fields of
+function type are allowed.We explain how the parameter instantiation
+step instantiates function types below, after discussing
+parameterization for methods.
+
+\subsection{Region Parameterization for Methods and Function Types}
+
+As the next step, we introduce region parameters for every method.  We
+do this by instantiating the types of all parameters and the return
+value (of the method) using fresh region identifiers (as explained
+previously), and then generalizing these region identifiers as formal
+region parameters of the method. In addition, a fresh region
+identifier is introduced to represent the allocation region.  We also
+introduce a fresh predicate variable $\varphi$ for every method, just
+as we did for each class.  Thus, the method
+
+\C{Object m (List x) \{...\} } \\
+is instantiated as
+
+\C{Object$\inang{\rho_3}$ m$\inang{\rho_0,\rho_1,\rho_2,\rho_3}$ (List$\inang{\rho_1,\rho_2}$ x) \{...\} }
+
+We then consider every method invocation in the program, and introduce
+fresh region variables representing the (yet unknown) actual region
+parameters for this particular invocation.
+%
+We similarly perform instantiation for every constructor invocation of
+the form \C{new@$\pi_0$ T($\ldots$)}, by instantiating the type \C{T}
+as before, turning it into \C{new T$\langle \pi_0, \rho_1, \cdots,
+\rho_n \rangle$($\ldots$)}, where $\rho_1, \cdots, \rho_n$ are fresh
+region variables. Note that the programmer typically specifies the
+region $\pi_0$ where the object is to be allocated. If the programmer
+does not specify this, it is allocated in the allocation-context
+region by default. Region variables are introduced only for the other
+region parameters.
+
+Function types (of fields and parameters) are instantiated just like
+methods above, (and the region where the closure is allocated is
+determined just as for other objects).  For example, the function type
+$\C{List} \rightarrow \C{Object}$ is instantiated as $\inang{\rho_0,
+\rho_1, \rho_2, \rho_3 \,|\, \varphi } \C{List} \inang{\rho_1,\rho_2}
+\xrightarrow{\pi} \C{Object}\inang{\rho_3}$.  Note that the newly
+introduced region identifiers are generalized as formal region
+parameters of the function type.  This is a (heuristic) choice made in
+the case of higher order functions.  Consider a higher order function
+\emph{f} with a function typed parameter $g$.  The fresh region
+identifiers introduced while instantiating the type of $g$ could be
+alternatively generalized as formal region parameters of $f$, but we
+choose to generalize them as formal region parameters of $g$.  We will
+discuss this aspect again later.
+
+\subsection{Constraint Generation}
+\label{sec:fb-constraintsem}
+
+The constraint generation algorithm mimics the static type checker,
+but accumulates constraints that must hold for the type checking to
+succeed.
+
+\mypara{Syntax of Constraints.}
+(See~Fig.~\ref{fig:constraint-syntax}.) The constraints are expressed
+using a set $\regionConstants$ of region constants, a set
+$\regionVars$ of region variables, and a set $\predVars$ of predicate
+variables.
+
+\begin{figure*}[!t]
+
+\[
+   \pi \in R \; \mathtt{ (Region \; constants) } \;\;
+   \nu \in V \; \mathtt{ (Region \; vars) } \;\;
+   \varphi \in P \; \mathtt{ (Predicate \; vars) } \;\;
+   \Delta \subseteq R
+\]
+\[
+\renewcommand{\arraystretch}{1.2}
+\setlength{\arraycolsep}{7pt}
+\begin{array}{rcl}
+
+\rho \in \mathtt{Region \; Identifiers} & \coloneqq & \pi \ALT \nu \\
+
+\phi \in \mathtt{Region \; Constraint} & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
+
+F \in \mathtt{Substitution} & \coloneqq & \cdot \ALT [\rho/\rho]F \\
+
+\ell \in \mathtt{Antecedent} & \coloneqq & \phi \ALT \varphi \ALT \varphi \conj \phi \\
+
+r \in \mathtt{Consequent} & \coloneqq & \phi \ALT F(\varphi) \\
+
+\mathtt{Constraint} & \coloneqq & \isvalid{\ell}{r} \ALT \nu \in \Delta \ALT \tywf{\Delta}{\varphi} \\
+
+\end{array}
+\]
+
+\caption{Syntax of constraints}
+\label{fig:constraint-syntax}
+% \vspace*{-0.15in}
+\end{figure*}
+
+Recall that a \emph{region constant} may be either (a) a \emph{formal
+region parameter} of a class or method, or (b) a \emph{static region
+identifier} introduced by a \C{letregion} construct, or (c) an
+\emph{open transferable region identifier} introduced by an \C{open}
+construct.  A \emph{region variable} is introduced to represent an
+unknown \emph{actual} region parameter of a method invocation or
+object allocation, and the constraint-solver, if successful, will bind
+each region variable to a region constant.
+
+A \emph{region-constraint} $\phi$ consists of a conjunction of
+\emph{outlives-constraints} of the form $\rho_1 \outlives \rho_2$.  A
+predicate variable $\varphi$ is introduced to represent, \eg, the
+unknown precondition of a method.  The constraint-solver will end up
+binding it to a \emph{region-constraint} $\phi$ over a set of fixed
+formal region parameters.  Our constraints also make uses of
+\emph{pending substitutions} $F$: A pending substitution serves to
+bind formal region parameters in $\varphi$ to the actual region
+parameters used in a particular context: E.g., in the validity
+constraint $\isvalid{\rgn_1 \outlives
+\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending
+substitution is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$.
+
+The constraints are primarily of the form $\isvalid{\varphi_i \conj
+\phictxt} {\phicstr}$ or $\isvalid{\varphi_i \conj \phictxt}
+{F_j(\varphi_j)}$.  Here, $\varphi_i$ is a predicate variable
+(representing the precondition of a method to be determined),
+$\phicstr$ is a region-constraint that is \emph{required} to hold at a
+particular program point (within the method), and $\phictxt$ is a
+region-constraint that is \emph{known} to hold at that program point.
+Constraints of the form $\isvalid{\varphi_i \conj \phictxt}
+{F_j(\varphi_j)}$ are generated by an invocation of a method with
+precondition $\varphi_j$.
+
+We use \emph{well-formedness constraints} of the form $\rho \in
+\rhoenv$ to restrict the domain of unification for a region variable
+($\rho$) to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of
+regions in scope, and  well-formedness constraints of form
+$\tywf{\rhoenv}{\varphi}$ to restrict the domain of a predicate
+variable ($\varphi$) to the set of all possible region-constraint
+formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots,
+\pi_n \}$) in scope.
+
+\mypara{Constraint Solution.}
+We define an \emph{assignment} $\soln$ to be a pair of functions
+$(\solnR,\solnP)$, where $\solnR$ is a map from $\regionVars$ to
+$\regionConstants$ and $\solnP$ is a map from $\predVars$ to a
+region-constraint formula.  Such an assignment is said to
+\emph{satisfy} a set of constraints $C$ if every sequent in $C$ is
+valid after the substitutions $\solnP$ and $ \solnR$.  We say that $C$
+is \emph{satisfiable} if it has a satisfying assignment.
+
+%% ************ MACROS *************
+
+\newcommand{\rulepart}[1]{%
+\begingroup%
+\renewcommand*{\arraystretch}{1.2}%
+\begin{array}{c} #1 \end{array}%
+\endgroup%
+}
+
+\newcommand{\SRULE}[2]{\frac{\rulepart{#1}}{\rulepart{#2}}}
+
+\newcommand{\beginrules}{%
+\renewcommand*{\arraystretch}{1.4}%
+\begin{smathpar}\begin{array}{lc}%
+}
+\newcommand{\myendrules}{\end{array}\end{smathpar}}
+
+\newcommand{\lgcrule}[3]{%
+[\rulelabel{#1}] & \SRULE {#2} {#3}%
+\\[0.5cm] % \\[0.3cm]%
+}
+
+\newcommand{\lgcfact}[2]{%
+[\rulelabel{#1}] & {#2} %
+\\[0.3cm] % \\[0.2cm]%
+}
+
+\newcommand{\nl}{\\}
+
+\newcommand{\subtypesym}{<:}
+\newcommand{\classok}[2]{\vdash \, {#1} \texttt{ ok} \, \lhd \, {#2}}
+\newcommand{\okinok}[3]{{\vdash\,#1 \; \texttt{ok} \; \texttt{in} \; #2 \, \lhd #3}}
+\newcommand{\typeok}[3]{{#1\,\vdash\,#2 \; \texttt{ok} \, \lhd #3}}
+\newcommand{\typeconsok}[3]{{#1\,\vdash\,#2 \; \texttt{ok} \, \lhd #3}}
+\newcommand{\exprok}[4]{{#1} \, \vdash \, {#2} : {#3} \, \lhd {#4}}
+\newcommand{\subtypeok}[4]{{#1} \, \vdash \, {#2}  \subtypesym {#3} \, \lhd {#4}} 
+\newcommand{\stdcontext}{\A,\env,r}
+
+%% ************ END MACROS *************
+
+\begin{figure*}[t!]
+
+%%%%%%%%%%% Header Box %%%%%%%%%%%
+  \textbf{Expression Typing Constraint Generation} \; \fbox{  \( \exprok{\stdcontext}{e}{\tau}{C} \)}
+\\
+
+\beginrules
+
+%%%%%%%%%%% NEW %%%%%%%%%%%
+  \lgcrule{New}
+  {
+    \exprok {\stdcontext} {\bar{e}} {\bar{\tau}} {C_1}
+    \spc
+    \typeok {\A} {\fbN} {C_2}
+    \spc
+    %% \spc C_3 = \{ \isvalid{\A.\phicx}{\allocRgn(\fbN)=\ralloc} \}
+    \fields(\fbN) = \bar{f} : \taubar
+% \spc \subtypeok {\A} {\bar{\tau'}} {\bar{\tau}} {C_3}
+  }
+  {
+    \exprok {\stdcontext}   {\C{new} \fbN(\bar{e})} {\fbN} {C_1 \cup C_2}
+  }
+
+%%%%%%%%%%% NEW-REGION %%%%%%%%%%%
+  \lgcrule{NewRegion}
+  {
+    % \typeok {\A.\aenv} {T} {C_1} \spc
+    % C_1 = \{ \rgn \in \A.\rhoenv \}
+   % \spc
+   \exprok {\stdcontext} {e} {\inang{\rho} \unitZ \xrightarrow{r} T@\rho} {C}
+    % \nl
+    % \exprok{(\{\rho\},\A.\aenv,true),{\cdot}}{e}{T@\rho}{C_2}
+  }
+  {
+    % \exprok {\stdcontext} {\C{new}\; \RgnZ\inang{T} \inang{\toprgn} (\lambdaexp{\rgn}{\rho}{}{e})}  {\RgnZ\inang{T} \inang{\toprgn}} {C_1 \cup C_2}
+    \exprok {\stdcontext} {\C{new}\; \RgnZ\inang{T} \inang{\toprgn} (e)}  {\RgnZ\inang{T} \inang{\toprgn}} {C}
+  }
+
+%%%%%%%%%%% LET-REGION %%%%%%%%%%%
+
+\lgcrule{LetRegion}{
+   \A = (\rhoenv,\aenv,\phicx)  \spc \rgn \notin \rhoenv \spc
+      \A' = (\rhoenv \cup \{\rgn\}, \aenv, \phicx \conj (\rhoenv \outlives \rgn))
+   \nl
+   \exprok{\A',\env,\rgn} {e} {\tau} {C_1} \spc
+      \typeok {\A} {\tau} {C_2}
+}{
+   \exprok{\stdcontext} {\letregion{\rgn}{e}} {\tau} {(C_1 \cup C_2)}
+}
+
+%%%%%%%%%%% FUNCTION INVOCATION %%%%%%%%%%%
+\lgcrule{FnApply}
+{
+  \A = (\rhoenv,\aenv,\phicx) \spc
+  C_1 = \{r \bar{r} \in \rhoenv\} \spc
+\exprok {\stdcontext} {e} {\inang{\rho\rhobar\,|\,\phi}\bar{\tau^1} \xrightarrow{r} \tau^2} {C_2}
+\nl
+C_3 = \{\isvalid{\phicx}{[r \bar{r} / \rho\rhobar]\phi}\} \spc
+\exprok {\stdcontext} {\bar{e}} {[r \bar{r} / \rho\bar{\rho}] \bar{\tau^1}} {C_4}
+% \spc \substFn = [\bar{\pi}/\rhobar]
+% \spc
+% \subtypeok {\A} {\bar{\tau'}} {\substFn(\bar{\tau})} {C_5}
+}{
+  \exprok {\stdcontext} {e\inang{r \bar{r}}(\bar{e})} {[r \bar{r}/ \rho\bar{\rho}] \tau^2} {\cup_{i=1}^4 C_i}
+}
+
+%%%%%%%%%%% OPEN-REGION %%%%%%%%%%%
+\lgcrule{Open}{
+   \A = (\rhoenv,\aenv,\phicx) \spc
+   \rgn \notin \rhoenv \spc
+   \exprok {\stdcontext} {e_a} {\RgnZ\inang{T}\inang{\toprgn}} {C_1}
+   \\
+   % (\A',\env') = ((\rhoenv \cup \{\rgn\},\aenv,\phicx),\env[y\mapsto T@\rgn] \spc
+   \exprok {(\rhoenv \cup \{\rgn\},\aenv,\phicx),\env[y\mapsto T@\rgn]} {e_b} {\tau} {C_2}
+\spc
+\typeok {\A} {\tau} {C_3}
+}{
+   \exprok {\stdcontext} {\open{e_a}{\rgn}{y}{e_b}} {\tau} {(C_1 \cup C_2 \cup C_3)}
+}
+
+\myendrules
+
+\caption{Constraint generation rules: Part 1}
+\label{fig:constraint-gen-0}
+\end{figure*}
+
+\mypara{Constraint Generation.} The constraint generation algorithm is
+a direct adaption of the type checker: each type checking judgment is
+modified to produce a set of constraints that must hold for the type
+checker to succeed.
+
+Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language
+constructs.  (The appendix contains the remaining constraint
+generation rules.) These rules use the same context as the
+corresponding typing judgment in Fig.~\ref{fig:fb-staticsem}, except
+that this is generalized to permit the use of region variables and
+predicate variables.  Symbols $\A$ and $\env$ retain their meaning,
+modulo this extension.  Thus, the component $\phicx$ of $\A$ will now
+be a \emph{symbolic expression} of the form $\varphi \conj \phi$,
+where $\varphi$ is the predicate variable representing the
+undetermined precondition of the method being analyzed.  The algorithm
+proceeds top-down, analyzes an expression $e$ in a context
+$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
+(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
+indicating that the expression will have a type $\tau$ provided the
+constraints $C$ hold.
+
+When we adapt a type-checking judgement rule to produce a
+corresponding constraint-generation rule, we treat the antecedent
+conditions of the type-checking rule in one of two ways. Many of these
+antecedent conditions are converted into constraints which are
+accumulated in the set $C$. However, some of the antecedent conditions
+are expected to be trivially satisfied and are expressed as antecedent
+conditions in the constraint-generation rule as well.  E.g., our
+frontend ensures that the region constants introduced by \C{open} and
+\C{letregion} constructs are unique (and the elaboration phase ensures
+this by alpha-renaming them as needed). Thus, the precondition $\rho
+\not\in \Delta$ in rules \textsc{LetRegion} and \textsc{Open} is
+expected to hold true at constraint-generation time (and does not
+produce any constraints).
+
+The rules for generating constraints from a method and class
+definition first build a context ($\A$) containing a set ($ \rhoenv$)
+denoting regions that are currently live, a map ($\aenv$) mapping type
+variables to their bounds, and a constraint formula ($\phicx$)
+capturing constraints over live region variables. We use predicate
+variables ($\varphi$ and $\varphi_m$) to capture constraints over
+variables in $\rhoenv$ that are yet to be inferred.
+
+Let $\consOf{q}$ denote the set of constraints generated from an
+elaborated program $q$.  The following theorem states that
+constraint-generation is sound and complete:
+\begin{theorem}
+\label{thm:constraint-generation-sc}
+  Let $C = \consOf{q}$.  An assignment $\soln$ (for the region and
+  predicate variables in $q$) satisfies $C$ iff $q[\soln]$ is
+  well-typed.
+\end{theorem}
+
+\subsection{The Constraint Solver}
+
+We refer to any validity constraint whose antecedent contains a
+predicate variable (\ie, is of the form $\varphi \conj \phictxt$) as
+an \emph{abduction constraint}.  Here,  $\phictxt$ is either
+\emph{true} (in which case, we call the constraint a trivial abduction
+constraint) or a conjunction of one or more outlives-constraints (in
+which case, we call the constraint a non-trivial abduction
+constraint).
+
+Non-trivial abduction constraints are a key challenge in solving the
+constraints.  We now describe some special properties of the set of
+constraints $C$ generated by our algorithm, which allow us to handle
+abduction constraints efficiently.
+
+For any predicate variable $\varphi$ used in $C$, $C$ has exactly one
+constraint of the form $\tywf{\Delta}{\varphi}$. We will refer to this
+$\Delta$ as $\deltaPC(\varphi)$.  (Our constraint-generation algorithm
+guarantees the preceding property.  Even otherwise, a set of
+constraints $\tywf{\Delta_i}{\varphi}$ can be replaced by the single
+equivalent constraint $\tywf{(\cap_i \Delta_i)}{\varphi}$.) We say
+that an abduction constraint is \emph{$C$-decomposable} if its
+antecedent is of the form $\varphi \conj \phictxt$ where $\varphi$ is
+a predicate variable and $\phictxt$ is a conjunction of zero or more
+outlives-constraints of the form $\pi_1 \outlives \pi_2$ satisfying
+the following conditions: (1) $\pi_2 \notin \deltaPC(\varphi)$.  (2)
+if $\pi_1 \in \deltaPC(\varphi)$, then for every $\pi_f \in
+\deltaPC(\varphi)$, $\pi_f \outlives \pi_2$ is a conjunct in
+$\phictxt$.  We will omit the reference to $C$ in the above notation
+if no confusion is likely.
+
+\begin{lemma}
+  \label{lemma:gc-is-decomposable}
+  Let $C = \consOf{q}$. Every abduction constraint in $C$ is $C$-decomposable.
+\end{lemma}
+
+\begin{lemma}
+\label{lemma:decomposition}
+  Consider any $C$-decomposable constraint $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+  where both $\pi_i$ and $\pi_j$ are region constants.
+  Let $\soln$ satisfy $C$.\\
+  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$:
+  $\soln$ satisfies $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+  iff
+  $\soln$ satisfies $\isvalid{\varphi}{\pi_i \outlives \pi_j}$.\\
+  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$:
+  $\soln$ satisfies $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
+  iff
+  $\soln$ satisfies $\isvalid{\phi}{\pi_i \outlives \pi_j}$
+\end{lemma}
+
+The above lemma shows how we can reduce a non-trivial abduction constraint to
+either a trivial abduction constraint or a non-abduction constraint,
+provided that the consequent is an outlives-constraint without region variables.
+
+\mypara{Constraint Solver.}
+The first step in our algorithm for solving a set of constraints $C$
+computes a set $\satC \supseteq C$ of constraints by iteratively applying
+the following rules until a fixed point is reached:
+
+\begin{enumerate}
+
+\item (Initialization) $\isvalid{\ell}{r} \in C \Rightarrow \isvalid{\ell}{r} \in \saturate{C}$
+
+\item (Transitivity)
+  \label{item:transitivity}
+\begin{enumerate}
+\item
+$\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
+$\isvalid{\ell \conj \phi}{\rho_2 \outlives \rho_3} \in \saturate{C}$
+$\Rightarrow$
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+\item
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
+$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$
+$\Rightarrow$
+$\isvalid{\ell \conj \phi}{\rho_1 \outlives \rho_3} \in \saturate{C}$
+\end{enumerate}
+
+\item (Substitution)
+  \label{item:subst}
+$\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
+$\isvalid{\varphi}{\phi} \in \saturate{C}$
+$\Rightarrow$ $\isvalid{\ell}{F(\phi)} \in \saturate{C}$
+
+\item (Abduction Decomposition)
+\label{item:context}
+\begin{enumerate}
+\item
+$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
+$\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$
+$\Rightarrow$
+$\isvalid{\varphi}{\pi_i \outlives \pi_j} \in \saturate{C} $
+
+\item 
+$\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j} \in \saturate{C}$,
+$\{ \pi_i, \pi_j \} \not\subseteq \predDeltaMap(\varphi)$,
+$\{ \pi_i, \pi_j \} \subseteq \regionConstants$
+$\Rightarrow$
+$\isvalid{\phictxt}{\pi_i \outlives \pi_j} \in \saturate{C} $
+\end{enumerate}
+
+\end{enumerate}
+
+We can show that every new constraint added by the above rules is
+implied by existing constraints:
+
+\begin{theorem}
+  \label{thm:closure}
+Let $C = \consOf{q}$.
+An assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
+\end{theorem}
+
+A key goal of this step is to identify the value every region variable
+must have in any solution of the set of constraints, as explained
+below.  Consider a set of constraints $D$. We say that a region
+variable $\rho$ \emph{occurs} in a \emph{context} $\ell$ (in $D$) if
+$D$ contains some constraint $\isvalid{\ell}{r}$ where $\rho$ occurs
+in $r$.  We say that a region variable $\rho$ is \emph{bound} to a
+region constant $\pi$ in a context $\ell$ if $\{ \isvalid{\ell}{\rho
+\outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
+We say that a region variable $\rho$ is \emph{bound} to a region
+constant $\pi$ (in $D$) if $\rho$ is bound to $\pi$ in every context
+$\ell$ in which it occurs.  We say that a region variable $\rho$ is
+\emph{somewhere-bound} to a region constant $\pi$ (in $D$) if $\rho$
+is bound to $\pi$ in some context $\ell$ in which it occurs.
+
+The set $\satC$ makes it easy to identify a solution $\thesoln$ (if
+one exists), as below.  For any predicate variable $\varphi$,
+$\thesolnP(\varphi)$ is defined to be $\conj \{ \pi_1 \outlives \pi_2
+\;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1
+\outlives \pi_2} \in \saturate{C} \}$.  For any region variable
+$\rho$, we define $\thesolnR(\rho)$ to be any element of the set $\{
+  \pi \in \regionConstants \:|\; \rho \text{ is somewhere-bound to }
+  \pi \text{ in } \saturate{C} \}$, if this set is non-empty. If this
+set is empty for any $\rho$, $\thesoln$ is undefined.
+
+The set $\satC$ also makes it easy to check if $C$ is satisfiable.
+Let $\myground{C}$ denote the subset of all ground constraints
+(\ie, constraints without any region variable or predicate variable)
+in $\saturate{C}$. Let $\rhoC$ denote the subset of all well-formedness
+constraints for region variables in $\satC$.
+Define $\solveCon{C}$ as below.
+
+\[
+\solveCon{C} = 
+\begin{cases}
+\textsc{Some} (\thesoln) & \textrm{if $\groundC$ is valid and $\thesoln$ is defined and satisfies $\rhoC$} \\
+\textsc{None} & \textrm{otherwise}
+\end{cases}
+\]
+
+\begin{theorem}
+\label{thm:constraint-solver-sc}
+  Let $C = \consOf{q}$.  (Soundness) If $\solveCon{C} =
+  \textsc{Some}(\soln)$, then $\soln$ satisfies $C$.  (Completeness)
+  If $\solveCon{C} = \textsc{None}$, then $C$ is unsatisfiable.
+\end{theorem}
+
+Checking the validity of a ground constraint is straightforward.  A
+ground constraint is of the form $\isvalid{\conj_{i \in I}
+\ell_i}{r}$, where each $\ell_i$ and $r$ is an outlives-constraint.
+Let $D$ denote $\{ \isvalid{}{\ell_i} \;|\; i \in I \}$.  The given
+constraint is valid iff $r$ belongs to $\saturate{D}$.
+
+\mypara{Algorithmic Aspects.}
+Note that checking the validity of a ground constraint can be realized
+using a simple graph reachability algorithm.
+%
+Given a set $S$ of outlives constraints, define the directed graph
+$G(S)=(V(S),E(S))$ as follows.  Every distinct region identifier
+$\rho$ in $S$ is represented by a vertex, which we will also refer to
+as $\rho$.  Every outlives constraint $\rho_1 \outlives \rho_2$ is
+represented by an edge from $\rho_1$ to $\rho_2$.
+%
+%
+It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$
+iff there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
+%
+Thus, a simple graph reachability algorithm can be used to check the
+validity of ground constraints.
+
+This idea generalizes.  Extending the simple graph reachability
+algorithm to incorporate the Substitution rule (in computing $\satC$)
+turns the problem into a context-free reachability problem in
+graphs~\cite{Reps:Reachability} (as usual for context-sensitive
+interprocedural analysis).
+
+Algorithms for context-free reachability can be adapted to incorporate
+the Abduction Decomposition step.  Alternatively, the iterative
+process described above is a standard fixed point computation and can
+be encoded using a set of Datalog rules, allowing us to compute the
+closure using any Datalog engine.
+% Further details omitted.
+
+\subsection{Soundness and Completeness: Discussion}
+
+Theorems~\ref{thm:constraint-generation-sc}
+and~\ref{thm:constraint-solver-sc} show that the second and third
+steps of the type-inference algorithm are sound. It is easy to verify
+the soundness of the first step (the elaboration phase): For any
+program $p \in \absof{\FB}$, the elaboration phase produces a $q$ such
+that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.
+%
+The soundness of the type inference algorithm follows.
+
+Theorem~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc} also
+establish the completeness of the constraint-generation and constraint-solving steps.
+%
+The only source of incompleteness in the type inference algorithm is
+the set of heuristic choices made during the first step,
+as explained below.
+
+ (1) We determine the set of region parameters for a recursive class
+using the heuristic that a recursive occurrence of the class has
+the same parameters, in the same order, as the class itself.
+This heuristic fails, for example, if the program uses a recursive list type whose
+elements alternatively come from two different regions. Such a program would require
+the following elaboration, which is beyond the scope of our approach:
+
+\begin{tabular}{ccc}
+\begin{minipage}{0.32\linewidth}
+\begin{codejava}
+class List $\extends$ $\ObjZ$
+{
+  $\ObjZ$ data;
+  List next;
+}
+\end{codejava}
+\end{minipage}
+&
+$\Rightarrow  \; \; \;$
+&
+\begin{minipage}{0.5\linewidth}
+\begin{codejava}
+class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
+{
+  $\ObjZ \langle \rho_1 \rangle$ data;
+  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
+}
+\end{codejava}
+\end{minipage}
+\end{tabular}
+
+  (2) Our technique for region parameterization also uses a heuristic in the case of
+higher order programs. The following examples illustrates that principal types may
+not exist for higher order functions.
+\begin{codejava}
+    unit apply ( T $\rightarrow$ unit f, T x, T y) { f(x); f(y); }
+\end{codejava}
+This method may be typed assuming either that \C{f} is polymorphic
+over the region that its parameter is allocated in (permitting \C{x}
+and \C{y} to be allocated in any regions), or by assuming that \C{x}
+and \C{y} are allocated in the same region $\rho_1$ that \C{f} expects
+its parameters to be allocated in. Neither type subsumes the other.
+Our algorithm heuristically chooses the first option, as it appears to
+be the more likely and useful candidate.
+
+If users provide partial region annotations, especially in situations
+(such as above) where elaboration makes a heuristic choice, the
+elaboration procedure can use the user-provided choices instead.  This
+can help the type-inference overcome these limitations.
+
+\subsection{Modularity Aspects of Type Inference}
+
+The type inference algorithm, as presented, traverses the entire
+program to generate the set of constraints, which are solved en masse,
+using an iterative fixed point computation. However, the type
+inference can be realized in a modular and compositional fashion,
+subject only to the restrictions imposed by recursion.
+
+In the elaboration phase, we can process a class \C{C} only after any
+class \C{B} that \C{C} depends on has been processed: class \C{C}
+depends on class \C{B} if \C{B} is either \C{C}'s base class or the
+type of any field of \C{C} depends on \C{B}. In effect, this means
+that any collection of mutually recursive classes must be processed
+together. Non-recursive dependences can be handled in a compositional
+fashion: if class \C{C} depends on \C{B} non-recursively, then the
+elaboration can be done for \C{B} first, and then \C{C} can be
+processed.
+
+The same idea applies to the constraint-solving phase as well.  Given
+a set of constraints, we say that a predicate variable $\varphi_1$
+\emph{directly-depends} on another predicate variable $\varphi_2$ if
+the set of constraints includes a constraint $\isvalid{\varphi_1 \conj
+\phictxt}{F(\varphi_2)}$.  We say that $\varphi_1$ \emph{depends} on
+$\varphi_2$ if $\varphi_1$ transitively depends on $\varphi_2$.  The
+constraint solver needs to process any collection of mutually
+dependent predicate variables together.  In effect, this requires the
+type inference to process any collection of mutually recursive methods
+together.  However, methods that are not mutually recursive can be
+processed separately.
+
 \renewcommand{\rgn}{r}
 
-\input{implementation}
+%--------------------------------------------------------------------
+%                             Implementation
+%--------------------------------------------------------------------
+\section{Implementation and Evaluation}
+\label{sec:implementation}
 
-\input{related}
+As mentioned earlier, our work is a continuation of the work reported
+in~\cite{Broom:HotOS}, which provides users with a C\#-based
+implementation of transferable regions (the features described in
+Section~\ref{sec:overview}). This system has no type system and
+provides users with no safety guarantees. \cite{Broom:HotOS} presents
+evidence that realistic programs can be implemented using transferable
+regions and that this can yield significant performance gains.  In
+particular, it reports speedups up to 34\% for typical big-data
+analytics jobs.  We now describe our implementation and experience
+with the type system and type inference algorithm presented in the
+current paper.
 
+We have implemented \namec, a prototype of \name compiler frontend,
+including its region type system and type inference, in 3k+ lines of
+OCaml.
+% Our implementation is called \namec.
+The input to \namec is a program in $\absof{\FB^+}$, an extended
+version of $\absof{\FB}$ that includes assignments, conditionals,
+loops, more primitive datatypes (\eg, integers), and a null value. 
+% If needed, a specification file containing region type annotations
+% for higher-order functions can also be provided. 
+Our implementation of region type inference and constraint solving
+closely follows the description given in
+Sec.~\ref{sec:type-inference}.
+
+We performed two kinds of experiments to evaluate our region type
+system and type inference.  First, we implemented some microbenchmarks
+($\le$100 LOC) consisting of standard classes such as pairs, lists,
+list iterators, etc., in $\absof{\FB^+}$, and used our inference
+engine to infer their region types.  These classes are
+region-oblivious. Hence, as long as they are well-typed as per the
+core type system, \namec must be able to automatically construct its
+region-type-annotated definition without fail. \namec was able to
+infer the expected region types for all these classes under 10ms.
+Fig~\ref{fig:rev} shows the region-type-annotated definition computed
+for the list reverse method. Observe that \namec was able to infer
+that the list and its data (of type \C{T}) can be allocated in
+different regions, as long as the latter outlives the former. This
+allows, for instance, a \C{preOrder} method to traverse a tree in a
+transferable region, and return a list of its nodes, where the list
+itself is allocated in the stack region.
+
+\begin{figure}
+\begin{codejava}
+class LinkedList<T><R5,R4 | R4$\outlives$R5> {
+  ListNode<T><R5,R4> head; ...
+  List<T><R17,R4> rev<R17,R4 | R4$\outlives$R17>(unit u) {
+    List<T><R17,R4> xs = new List<T><R17,R4>(this.head.val);
+    ListNode<T><R5,R4> cur = this.head.next;
+    while (!cur == Null) {
+      xs.add<R17>(cur.val)
+      cur = cur.next; }
+    return xs;
+  }
+\end{codejava}
+
+\caption{Region-annotated definition of \C{rev} computed by \namec}
+\label{fig:rev}
+\vspace*{-0.15in}
+\end{figure}
+
+Next, we translated 4 out of the 6 Naiad streaming query operator
+benchmarks (Naiad vertices) used in~\cite{Broom:HotOS} to
+$\absof{\FB^+}$, and used \namec to verify their safety. The 2
+remaining benchmarks were left out because their region behavior (from
+the perspective of the type system) is subsumed by the included
+benchmarks. The number of LOC performing operations on \C{Region}
+objects relative to the total LOC is 8\% or under in the Naiad
+benchmarks.  During the process, we found multiple instances of
+potential memory safety violations in the $\absof{FB^+}$ translation
+of all the 4 Naiad vertices, which we verified to be present in the
+original C\# implementation as well. The cause of all safety
+violations is the creation of a reference from the outgoing message (a
+transferable region) to the payload of the incoming message. For
+example, the implementation of \C{SelectVertex} contains the
+following:
+\begin{codejava}
+  if (this.selector(inMsg.payload[i])) {
+    outMsg.set(outputOffset, inMsg.payload[i]);
+    ...
+  } 
+\end{codejava}
+The \C{outMsg} is later transferred to a downstream actor, where the
+reference to \C{inMsg}'s payload becomes unsafe\footnote{This unsafe
+reference could have gone unnoticed during experiments
+in~\cite{Broom:HotOS} because their experimental setup included only
+one actor.}. We eliminated such unsafe references by creating a clone
+of \C{inMsg.payload[i]} in \C{outMsg}, and our compiler was
+subsequently able to certify the safety of all references.
+
+Our experience with Naiad benchmarks shows the utility of our type
+inference/checking tool, particularly because it comes at no
+additional cost to the developer.
+
+%--------------------------------------------------------------------
+%                             Related Work
+%--------------------------------------------------------------------
+\section{Related Work}
+\label{sec:related-work}
+
+Following Tofte and Talpin's seminal work
+in~\cite{tofte93,tofte94,tofte97}, static type systems for safe
+region-based memory management have been extensively studied in the
+context of various languages and problem
+settings~\cite{cyclone02,cyclone04,yates99,MIT03,DPJ09,HMN01,WW01,rust,gpu14}.
+Our work differs from the existing proposals in one or more of the
+following respects.
+
+\begin{enumerate}
+\item 
+   Our design choice focuses on ensuring memory-safety while giving programmers
+   control over  region management and allocation of objects in regions.
+   In contrast, some systems automate all aspects of memory management.
+   This is a convenience-performance trade-off.
+
+\item We support both lexically scoped (stack) regions and dynamic transferable
+regions (both programmer-managed).
+
+\item We exploit a combination of a simple static type discipline and lightweight
+  runtime checks to ensure memory safety.
+  % adopt a two-pronged approach to memory safety that relies on a
+In particular, our approach circumvents the need for restrictive
+static mechanisms (e.g., linear types and unique pointers) or
+expensive runtime mechanisms (e.g., garbage collection and reference
+counting) in order to guarantee safety.
+
+\item We present a full (interprocedural) type inference algorithm
+that eliminates the need to write region annotations on types.
+
+\item Our underlying language is an object-oriented programming language,
+equipped with higher-order functions and parameterized (generic) types.
+These language features necessitate some non-trivial choices in the design
+of the region-parametricity aspect of the language, which also have an
+impact on aspects such type inference.
+
+\end{enumerate}
+
+Tofte and Talpin's approach~\cite{tofte97} uses compiler-managed
+lexically scoped (stack) regions (as a replacement for GC).  Our type
+inference is analogous to theirs in some respects, while differing in
+others.  Their inference algorithm only generates equality
+constraints, solvable via unification.  Our type inference algorithm
+generates partial order outlives constraints.  Consequently, our
+constraint solving algorithm is more sophisticated, and is capable of
+inferring unknown outlives constraints over region arguments of
+polymorphic recursive functions.
+
+Walker and Watkins~\cite{WW01} extend lambda calculus with first-class
+regions with dynamic lifetimes, and impose linear typing to control
+accesses to regions.  Our open/close lexical block for transferable
+regions traces its origins to the \C{let!} expression in~\cite{WW01}
+and~\cite{wadler90}, which safely relaxes linear typing restrictions,
+allowing variables to be temporarily aliased.  We don't use linear
+typing (for references to regions), thus admit unrestricted aliasing,
+but use lightweight runtime checks for safety.  Moreover,
+~\cite{WW01}'s linear type system is insufficient to enforce the
+invariants needed to ensure safety under region transfers, such as the
+absence of references that escape a transferable region.
+
+Cyclone~\cite{cyclone02} equips C with programmer-managed stack
+regions, and a typing discipline that statically guarantees the safety
+of all pointer dereferences. Later
+proposals~\cite{cyclone04,cycloneSCP} extends Cyclone with dynamic
+regions. \name differs from Cyclone in its non-intrusiveness design
+principle, which requires its safety mechanisms to not intrude on the
+programming practices of C\#.  \name programmers, for example,
+shouldn't be forced to abandon iterators in favor of for-loops,
+annotate region types, or rewrite C\#'s standard libraries to use in
+\name. Cyclone requires C programmers to use new language constructs
+and abandon some standard programming idioms in the interest of
+preserving safety. For instance, Cyclone programmers are required to
+write region types for functions; the type inference is only
+intraprocedural. Ensuring safety in presence of dynamic regions
+requires using either unique pointers or reference-counted objects.
+Both approaches are intrusive. For example, unique pointers constrain,
+or in some cases forbid, the use of the familiar iterator pattern,
+which requires creation of aliases to objects in a collection. Some
+standard library functions, for example, those that use caching, may
+need to be rewritten.  Moreover, even with unique pointers, safety
+cannot be guaranteed statically; checks against \C{NULL} are needed at
+run-time to enforce safety. For ref-counted objects, Cyclone requires
+programmers to use special functions (\C{alias\_refptr} and
+\C{drop\_refptr}) to create and destroy aliases.  Reference count is
+affected only by these functions.  An alias going out of scope, for
+instance, does not decrement the ref-count. The requirement to use
+additional constructs to manage aliases makes reference counting
+more-or-less as intrusive as unique pointers.
+
+Our work differs from Cyclone also in terms of its technical
+contributions. While Cyclone equips C with a range of region
+constructs~\cite{cycloneSCP}, the semantics of (a significant subset
+of) such constructs, and the safety guarantees of the language are not
+formalized. In contrast, the (static and dynamic) semantics of Broom
+has been rigorously defined with respect to a well-understood formal
+system (FGJ). The safety guarantees have been formalized and proved.
+Similar contrast can be made of region type inference in both the
+languages.  Cyclone's type inference was only ever described as being
+similar to Tofte and Talpin's, and its effectiveness in presence of
+tracked pointers is not clear.  In contrast, a detailed type inference
+algorithm is one of our core contributions.
+
+Our region type system can also be thought of as a specialized
+ownership type system~\cite{OwnershipSurvey}, where each region is the
+owner of all objects allocated in the region.
+%
+An ownership type system for safe region-based memory management in
+real-time Java has been proposed by Boyapati \emph{et
+al.}~\cite{MIT03}.  Their language permits only lexically-scoped
+(stack) regions.  In contrast, we permit regions with dynamically
+determined lifetimes.  Our language also admits generics and
+higher-order functions.  We also establish type safety and transfer
+safety results that formalize the guarantees provided by our system.
+While Boyapati \emph{et al.}'s language is explicitly typed, our
+language comes equipped with full type inference.  However, several
+inference algorithms have been proposed in the context of other
+ownership type systems.  Our type inference algorithm is also novel
+compared to these existing ownership inference algorithms, which are
+based on, e.g., pointer analysis~\cite{HuangEtAl:ECOOP12} or boolean
+satisfiability~\cite{DietlEtAl:ECOOP11}.  (See Section 5.2 of Clark
+\emph{et al.}'s survey of ownership type
+systems~\cite{OwnershipSurvey} for a more comprehensive discussion of
+ownership inference algorithms.) Some distinguishing characteristics
+of our algorithm is that it is customized to our problem, does not use
+any pointer analysis algorithm (which can be a source of imprecision)
+or SAT solvers (which can be a source of inefficiency), and comes with
+relative completeness guarantees.
+
+Henglein \emph{et al.}~\cite{HMN01} propose a flow-sensitive approach
+for first-order programs to generalize Tofte and Talpin's approach to
+dynamic regions. Cherem and Rugina~\cite{CR04} describe a
+flow-insensitive and context-sensitive analysis that transforms Java
+programs to use (dynamic) regions.  However, neither of themr supports
+dynamic regions as first-class objects; they cannot be stored in data
+structures or passed to methods. Furthermore, while Henglein \emph{et
+al.}~\cite{HMN01} require reference counting to ensure memory safety,
+Cherem and Rugina's analysis~\cite{CR04} comes with no formal safety
+guarantees.
+%
+Holk \emph{et al.}~\cite{gpu14} use regions to safely transfer data
+between the CPU and GPU in the context of Scheme. However, their
+setting only includes lexically-scoped regions for which Tofte and
+Talpin-style analysis suffices. In contrast, we provide first-class
+support for transferable regions with dynamic lifetimes.
+
+%--------------------------------------------------------------------
+%                             Bibliography
+%--------------------------------------------------------------------
 \bibliography{broom}
@@
 \input{typeinference-proofs}
-\input{misc}
 

### 2018-06-11 paper/broom.tex
All graphics -> PDF
Formatting changes.

@@
 \documentclass[a4paper,UKenglish]{lipics-v2018}
+\nolinenumbers
 %This is a template for producing LIPIcs articles. 
@@
 \usepackage[multiple]{footmisc}
+\showboxdepth=5
+\showboxbreadth=5
 
@@
 
-% Theorems Environment
-%----------------------
-% \newtheorem{theorem}{Theorem}[section]
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{proposition}[theorem]{Proposition}
-% \newtheorem{corollary}[theorem]{Corollary}
-
-% \newenvironment{proof}[1][Proof]{\begin{trivlist}
-% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-% \newenvironment{definition}[1][Definition]{\begin{trivlist}
-% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-% \newenvironment{example}[1][Example]{\begin{trivlist}
-% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-% \newenvironment{remark}[1][Remark]{\begin{trivlist}
-% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
-
-% \newcommand{\qed}{\nobreak \ifvmode \relax \else
-      % \ifdim\lastskip<1.5em \hskip-\lastskip
-      % \hskip1.5em plus0em minus0.5em \fi \nobreak
-      % \vrule height0.75em width0.5em depth0.25em\fi}
-
 \usepackage{listings}
@@
 during the initial stages of this project. We would also like to thank
-multiple anonymous reviewers for their careful scrutiny and feedback at
-various stages of this work.}%optional
+multiple anonymous reviewers for their careful scrutiny and feedback
+on various versions of this work.}%optional
 
@@
 \begin{figure} 
-\includegraphics[scale=0.45]{region-fsm.png}
+\includegraphics[scale=0.45]{region-fsm}
 \caption{The lifetime of a dynamic (transferable) region in \name}
@@
 %
-\begin{minipage}{2.55in}
+\begin{minipage}{\textwidth}
 \begin{smathpar}
@@
   {
-    CT(B) = \headerOf{B}\{...\}\\
+    CT(B) = \headerOf{B}\{...\}\spc
     \rbar \in \rhoenv \spc
@@
 %
-\begin{minipage}{1.25in}
+
+\begin{minipage}{0.49\textwidth}
 \begin{smathpar}
@@
     \fgjtywf{\aenv}{T}\spc
-%   \rgn \in \A.\rhoenv\\
-    \rgn \in \rhoenv\\
+    \rgn \in \rhoenv\spc
     \fgjsubtyp{\aenv}{T}{\ObjZ}\spc
@@
 %
-\begin{minipage}{0.9in}
+\begin{minipage}{0.49\textwidth}
 \begin{smathpar}
@@
   { 
-    \\
-    \A = (\subtypcx) \spc
-    \fgjtywf{\aenv}{T}
+    \fgjtywf{\aenv}{T} \spc
+    \A = (\subtypcx)
   }
@@
 %
-\begin{minipage}{2.2in}
+\begin{minipage}{2in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2in}
+\begin{minipage}{3.2in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2.9in}
+\begin{minipage}{3.2in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2.5in}
+\begin{minipage}{2.4in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2.6in}
+\begin{minipage}{3in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2.5in}
+\begin{minipage}{2.6in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{2.3in}
+\begin{minipage}{2.5in}
 \begin{smathpar}
@@
 %
-\begin{minipage}{5.5in}
+\begin{minipage}{\textwidth}
 \begin{smathpar}
@@
 %   \tywf{\A}{\tau^2} \spc
-    CT(B) = \headerOf{B}\{...\}\\
+    CT(B) = \headerOf{B}\{...\}\spc
+    \hastyp{\exptycx{\env}{\rho_m}}{e}{\tau^2} \\
     \env = \cdot[\xbar\mapsto\bar{\tau^1}][\thisZ\mapsto
@@
 %   \stmtsem{\rhoallocm}{s}{\stmtsemcx}{\stmtsemcxp}\spc
-    \hastyp{\exptycx{\env}{\rho_m}}{e}{\tau^2} \spc
 %   \subtyp{\A}{\tau}{\tau^2}
@@
 }
+
 \newcommand{\lopsemrule}[4]{%
@@
 
-\lopsemrule{3.2in}{
+\lopsemrule{\textwidth}{
 %   \fresh(\rgn')\spc
@@
 
-\lopsemrule{3in}{
+\lopsemrule{\textwidth}{
     \redstocup{\loc}{(e,\mem)}{(e',\mem')}
@@
 
-\lopsemrule{3in}{
+\lopsemrule{\textwidth}{
 %   \rgn \notin \rhoenv
@@
 
-\lopsemrule{3.3in}{
+\lopsemrule{\textwidth}{
     \loc \not\in dom(\mem) \spc
@@
 
-\lopsemrule{3.3in}{
+\lopsemrule{\textwidth}{
 %   \mem(\loc) = \USED \spc % Type system cannot guarantee this.
@@
 
-\lopsemrule{4in}{
+\lopsemrule{\textwidth}{
     v_a = \C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}(v) \spc
@@
 
-\lopsemrule{2.7in}{
+\lopsemrule{\textwidth}{
     \redstocup{\loc}{(e,\mem)}{(e',\mem')}
@@
 
-\lopsemrule{2.7in}{
+\lopsemrule{\textwidth}{
     \mem' = \mem[\loc \mapsto \status]
@@
 
-\lopsemrule{3in}{
+\lopsemrule{\textwidth}{
     v_a = \C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}(v) \spc
@@
 
-\lopsemrule{3.25in}{
+\lopsemrule{\textwidth}{
     \mem(\loc) = \USED \spc
@@
 
-\lopsemrule{2.5in}{
+\lopsemrule{\textwidth}{
     \mem(\loc) = \TLIVE \spc
@@
   \label{lemma:gc-is-decomposable}
-  Let $C = \consOf{q}$. Every abduction constraint in $C$ is $C$-decomposable.
+  Every abduction constraint in $C$, where $C = \consOf{q}$, is $C$-decomposable.
 \end{lemma}

### 2018-06-11 paper/broom.tex
Appendix formatted. Proofs removed.

@@
 \usepackage[multiple]{footmisc}
-\showboxdepth=5
-\showboxbreadth=5
 
@@
 
-\caption{Constraint generation rules: Part 1}
+\caption{Constraint generation rules part 1}
 \label{fig:constraint-gen-0}
@@
 
-%%
-%% Appendix
-%%
+%--------------------------------------------------------------------
+%                              Appendix
+%--------------------------------------------------------------------
 
 \appendix
-
-\input{appendix}
-\input{typesystem-proofs}
-\input{fb-constraint-generation-extra}
-\renewcommand{\rgn}{\myownthrowexception}
-\input{typeinference-proofs}
-
-%%
-%% Bibliography
-%%
-
+\section{Appendix}
+
+\subsection{Full Static Semantics}
+
+\begin{figure*}[h!]
+%
+\begin{minipage}{2.6in}
+\begin{smathpar}
+\begin{array}{lcl}
+  \allocRgn(A\inang{\rgn\rbar}\inang{\tbar}) & = & \rgn\\
+  \allocRgn(\inang{\rhobar \,|\, \phi}\bar{\tau^1}
+      \xrightarrow{\rgn} \tau^2) & = & \rgn\\
+  \shape(A\inang{\rbar}\inang{\tbar}) & = & A\inang{\tbar}\\
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{2.6in}
+\begin{smathpar}
+\begin{array}{lcl}
+  \bound_{\aenv}(\tyvar@\rgn) & = & \aenv(\tyvar)@\rgn\\
+  \bound_{\aenv}(\fbN) & = & \fbN\\
+  \fields(\ObjZ\inang{\rgn}) & = & \bullet \\
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+% \begin{minipage}{\textwidth}
+% \begin{smathpar}
+% \begin{array}{lcl}
+%   \ctype(\ObjZ\inang{\rgn}) & = & \bullet \\
+%   \ctype(B\inang{\tbar}\inang{\rbar}) & = & 
+%     \fields(B\inang{\tbar}\inang{\rbar})\\
+% \end{array}
+% \end{smathpar}
+% \end{minipage}
+%
+
+%
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    CT(B) = \headerOf{B}\{\bar{\tau^f}\;\bar{f};\,...\}\spc
+    \substFn = [\rbar/\rhobar, \tbar/\bar{\tyvar}] \qquad 
+    \fields(\substFn(\fbN)) = \bar{g}:\bar{\tau^g}
+  }
+  {
+    \fields(B\inang{\tbar}\inang{\rbar}) \;=\;
+      \bar{g}:\bar{\tau^g},\,\bar{f}:\substFn(\bar{\tau^f})
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    CT(B) = \headerOf{B}\{\bar{\tau^f}\;\bar{f};\,\bar{d}\}\spc
+    m \notin \bar{d} \qquad 
+    \substFn = [\rbar/\rhobar, \tbar/\bar{\tyvar}]
+  }
+  {
+    \mtype (m,B\inang{\tbar}\inang{\rbar}) \;=\;
+    \mtype (m, \substFn(\fbN))
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    CT(B) = \headerOf{B}\{\bar{\tau^f}\;\bar{f};\,\bar{d}\}\spc
+    \tau^2 \; m\mang (\bar{\tau^1}\;\bar{x})\{...\} \in \bar{d} \spc
+    \substFn = [\rbar/\rhobar, \tbar/\bar{\tyvar}]
+  }
+  {
+    \mtype (m,B\inang{\tbar}\inang{\rbar}) \;=\;
+    \substFn(\mang\bar{\tau^1} \rightarrow \tau^2)
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+
+    \mtype(m,\fbN) = \inang{\bar{\rho_1}\,|\, \phi_1}\bar{\tau^{11}} 
+                      \rightarrow \tau^{12} \spc \texttt{implies} \\
+    \isvalid{\A.\phicx}
+                  {\phi_2 \Leftrightarrow \subst{\bar{\rho_2}}{\bar{\rho_1}}(\phi_1)}
+    \spc \texttt{and} \spc
+    \bar{\tau^{21}} = \subst{\bar{\rho_2}}{\bar{\rho_1}}(\bar{\tau^{11}}) \spc 
+    \texttt{and} \spc \subtyp{\A}{\tau^{22}} {\subst{\bar{\rho_2}}{\bar{\rho_1}}(\tau^{12})}
+%   \substFn = \subst{\bar{\rho_2}}{\bar{\rho_1}}
+%              \subst{\rhoalloc_2}{\rhoalloc_1}
+    %\substFn = [\rbar/\rhobar, \ralloc/\rhoalloc, \tbar/\bar{\tyvar}]
+  }
+  {
+    \A \vdash \override(m,\fbN,\inang{\bar{\rho_2}\,|\, \phi_1}
+              \bar{\tau^{21}} \rightarrow \tau^{22})
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\caption{\fbname: auxiliary definitions}
+\label{fig:fb-auxdef}
+\end{figure*}
+
+\begin{figure*}[!h]
+%
+\textbf{Subtyping}  \; \fbox
+  {\(\subtyp{\A}{\tau_1}{\tau_2}\)}\\
+%
+\begin{minipage}{1.8in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+  \subtyp{\A}{\tau}{\tau} \\
+  \subtyp{(\Delta,\aenv,\phicx)}{\tyvar @\rho}{\aenv(\tyvar) @\rho}\qquad
+% \subtyp{\A}{\RgnZ\inang{\rgn}}{\RgnZ\inang{\toprgn}}\qquad
+% \subtyp{\A}{\RgnZ\inang{\toprgn}}{\RgnZ\inang{\rgn}}
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{2.7in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \\
+    CT(B) = \headerOf{B}\{...\}
+  }
+  {
+    \subtyp{\A}{B\inang{\tbar}\inang{\rbar}}
+        {[\rbar/\rhobar, \tbar/\bar{\tyvar}](\fbN)}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\begin{minipage}{1.5in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \subtyp{\A}{\tau_1}{\tau_2}\\
+    \subtyp{\A}{\tau_2}{\tau_3}
+  }
+  {
+    \subtyp{\A}{\tau_1}{\tau_3}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+\begin{minipage}{2.75in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \isvalid{\A.\phicx}{\phi_1 \Rightarrow \phi_2} \\
+    \subtyp{\A}{\bar{\tau^{11}}}{\bar{\tau^{21}}} \spc
+    \subtyp{\A}{\tau^{22}}{\tau^{12}}
+  }
+  {
+    \subtyp{\A}
+      {\inang{\rhobar \,|\, \phi_2}\bar{\tau^{21}}
+          \xrightarrow{\rgn} \tau^{22}}
+      {\inang{\rhobar \,|\, \phi_1}\bar{\tau^{11}}
+          \xrightarrow{\rgn} \tau^{12}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+
+%
+
+\textbf{Type, and Type Constraint Well-formedness}  \; \fbox
+  {\(\tywf{\A}{\tau}, \spc 
+     \tywf{\rhoenv}{\phi}\)}\\
+%
+\begin{minipage}{1.95in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \rgn \in \rhoenv
+  }
+  {
+    \tywf{(\rhoenv,\aenv,\phicx)}{\ObjZ\inang{\rgn}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{1.5in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \rgn_0,\rgn_1 \in \rhoenv
+%   \mem(\rgn_0) = \LIVE \spc
+%   \mem(\rgn_1) = \LIVE
+  }
+  {
+    \tywf{\rhoenv}{\rgn_0 \outlives \rgn_1}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{1.8in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \tywf{\rhoenv}{\phi_0} \spc
+    \tywf{\rhoenv}{\phi_1}
+  }
+  {
+    \tywf{\rhoenv}{\phi_0 \wedge \phi_1}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+% 
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \rgn \in \rhoenv \spc
+    \rhobar \not\in \rhoenv\spc
+%   \rhobar \notin dom(\mem) \\
+%   \mem' = \mem[\rhobar \mapsto \overline{\LIVE}] \spc
+    \rhoenv' = \rhoenv \cup \{\rhobar\} \spc
+    \A' = (\rhoenv', \aenv, \phicx \conj \phi) \spc
+    \tywf{\rhoenv'}{\phi}\spc 
+    \tywf{\A'}{\bar{\tau^1}} \spc
+    \tywf{\A'}{\tau^2}
+  }
+  {
+    \tywf{(\rhoenv,\aenv,\phicx)}{\inang{\rhobar \,|\, \phi}
+              \bar{\tau^1} \xrightarrow{\rgn} \tau^2}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\textbf{Class Well-formedness}  \; \fbox
+  {\(B \; \mathtt{ok}\)}\\
+%
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \rhoenv = \{\rho,\rhobar\} \spc
+    \aenv = [\bar{\tyvar} \mapsto \bar{\fgjN}] \spc
+    \phicx = \phi \spc
+    \A = (\subtypcx)\spc
+%   \A = (\{\rho,\rhobar\},[\bar{\tyvar} \mapsto \bar{\fgjN}],\phi)\spc
+    \tywf{\rhoenv}{\phi} \spc
+    \fgjtywf{\aenv}{\bar{\fgjN}} \spc
+    \okin{\bar{d}}{B}
+    \\
+    \tywf{\A}{\fbN,\bar{\tau^f}} \spc
+    \shape(\fbN) \neq \RgnZ\inang{T}\spc
+%   \tywf{\A}{\bar{\tau^f}} \spc
+    \isvalid{\phicx}{\allocRgn(\bar{\tau^f}) \outlives \rho} \spc
+%   \ctype(\fbN) = \bar{\tau^{\fbN}} \spc
+%    k = B(\bar{\tau^{\fbN}}\;\xbar,\bar{\tau^f}\;\bar{y})
+%        \{\superZ(\xbar);\,\thisZ.\bar{f} = \bar{y};\} \spc
+    \allocRgn(\fbN) = \rho \spc
+  }
+  {
+    \C{class}\; B\angAlpha\inang{\rho\rhobar \,|\, \phi} \extends \fbN
+    \{\bar{\tau^f}\;\bar{f};\, \bar{d}\} \;{\texttt{ok}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+
+\caption{\fbname: subtyping and well-formedness rules (contiunuation
+of Fig.~\ref{fig:fb-staticsem})}
+\label{fig:fb-staticsem-1}
+\end{figure*}
+
+\begin{figure*}[!t]
+\textbf{Expression Typing}  \; \fbox {\(\hastyp{\exptycx{\env}{\rgn}}{e}{\tau}\)}\\
+
+%
+\begin{minipage}{0.4\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e_1}{\tau_1}\\
+    \hastyp{\exptycx{\env[x\mapsto\tau_1]}{\rgn}}{e_2}{\tau_2}\\
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\letexp{x}{e_1}{e_2}}{\tau_2}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+%
+\begin{minipage}{0.58\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+%   \A = (\subtypcx)\\
+%   \loc \in \rhoenv \spc
+%   \hastyp{\exptycx{\env}}{e}{\tau}
+    \A = (\subtypcx) \spc
+    \loc \notin \rhoenv \\
+    \A' = (\rhoenv\cup\{\loc\}, \aenv, \phicx \conj \Delta \outlives
+    \loc)\\
+    \hastyp{\A',\env,\loc}{e}{\tau} \spc
+    \tywf{\A}{\tau}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\letd{\loc}{e}}{\tau}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\begin{minipage}{\textwidth}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \A = (\subtypcx)\spc
+    \A' = (\rhoenv\cup\{\loc\}, \aenv,\phicx) \spc
+%   \mem(\loc) = \USED \spc
+    \tywf{\A'}{T@\loc} \spc
+%   \fgjN = \RgnZ\inang{T}\spc
+    \hastyp{\A',\env, \loc}{e}{T@\loc}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\C{new}\;
+    \RgnZT{\toploc\loc}(e)} {\RgnZT{\toprgn}}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+%
+\begin{minipage}{3.4in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+%   \A = (\subtypcx)\\
+%   \loc \in \rhoenv \spc
+%   \hastyp{\exptycx{\env}}{e}{\tau}
+    \A = (\subtypcx) \spc
+%   \loc \notin \rhoenv \\ % An open region can be opened again.
+%   \loc \notin dom(\mem) \spc
+    \A' = (\rhoenv\cup\{\loc\},\aenv,\phicx) \\
+    \loc \in \Delta ~\texttt{implies}~ s = \LIVE \spc
+    \hastyp{\A',\env,\loc}{e}{\tau} \spc
+    \tywf{\A}{\tau}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{\opened{\loc}{s}{e}}{\tau}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+% %
+% \begin{minipage}{2in}
+% \begin{smathpar}
+% \begin{array}{c}
+% \renewcommand*{\arraystretch}{1.2}
+% \RULE
+%   {
+%     \\
+%     \\
+%     \hastyp{\exptycx{\env}{\rgn}}{e}{\RgnZ\inang{T}\inang{\toprgn}}
+%   }
+%   {
+%     \hastyp{\exptycx{\env}{\rgn}}{e.\transfer(\ldots)}{\unitZ}
+%   }
+% \end{array}
+% \end{smathpar}
+% \end{minipage}
+%
+\begin{minipage}{1.2in}
+\begin{smathpar}
+\begin{array}{c}
+\renewcommand*{\arraystretch}{1.2}
+\RULE
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e}{\tau} \\
+    \subtyp{\A}{\tau}{\tau'}
+  }
+  {
+    \hastyp{\exptycx{\env}{\rgn}}{e}{\tau'}
+  }
+\end{array}
+\end{smathpar}
+\end{minipage}
+%
+
+\caption{\fbname: expression typing (contiunuation of Fig.~\ref{fig:fb-staticsem})}
+\label{fig:fb-staticsem-2}
+\end{figure*}
+
+% Figs.~\ref{fig:fb-staticsem-1} and~\ref{fig:fb-staticsem-2} show full
+% static semantics of \FB. Fig.~\ref{fig:fb-staticsem-1} contains subtyping
+% rules, and rules to check well-formedness of \FB types, type
+% constraints, methods, and class definitions. Method well-formedness
+% rule makes use of the expression typing judgment defined in
+% Fig.~\ref{fig:fb-staticsem-2}. Auxiliary definitions used in
+% Figs.~\ref{fig:fb-staticsem-1} and~\ref{fig:fb-staticsem-2} are
+% defined in Fig.~\ref{fig:fb-auxdef}. As described in
+% \S~\ref{sec:type-system}, all judgments are parameterized over the
+% class table ($CT$). The premise $CT(B) = \headerOf{B}\{...\}$
+% used in some of the rules means that the definition of class $B$ is
+% present in $CT$ and that the definition is well-formed.
+
+\subsection{Full Operational Semantics}
+
+% Figs.~\ref{fig:fb-opsem-1} and~\ref{fig:fb-opsem-2} show the
+% operational semantics of \fbname (Fig.~\ref{fig:fb-opsem-2} is also
+% contained in the main paper as Fig.~\ref{fig:fb-opsem}. It is included
+% here for the sake of completeness). 
+\begin{figure*}[h!]
+
+%
+\fbox {\(\redstoo{\Delta}{(e,\mem)}{(e',\mem')}\)}\\
+
+\lopsemrule{\textwidth}{
+    \redstoo{\Delta}{(e,\mem)}{(e',\mem')}
+}{
+    \redstoo{\Delta}{(E\lbrack e \rbrack, \mem)}
+            {(E\lbrack e' \rbrack, \mem')}
+	  }{EvalOrder}
+%
+\vspace*{-0.1in}
+%
+
+\lopsemrule{\textwidth}{
+    \redstoo{\Delta}{(e,\mem)}{\invalidexn}
+  }{
+    \redstoo{\Delta}{(E\lbrack e \rbrack, \mem)}
+            {\invalidexn}
+	  }{Exception}
+%
+
+\lopsemrule{\textwidth}{
+%   \allocRgn(A\inang{\rgn\locbar}\inang{\tbar}) & = & \rgn
+%   \allocRgn(\fbN) \in \rhoenv \spc
+    \mem(\loc) = \LIVE \spc
+    \fields(A\inang{\tbar}\inang{\loc\locbar}) = \bar{f}:\taubar
+}{
+    \redstoo{\Delta}{((\C{new} \; A\inang{\tbar}\inang{\loc\locbar}(\vbar)).f_i,\mem)}{(v_i,\mem)}
+}{FieldAccess}
+
+\lopsemrule{\textwidth}{
+%   \allocRgn(\fbN) \in \rhoenv \spc
+    \mem(\loc) = \LIVE \spc
+    \mbody(m,A\inang{\tbar}\inang{\loc\locbar}) = \rho\rhobar.\bar{x}.e 
+%   \redstoo{\Delta}{(, \mem)}{(e',\mem')}
+}{
+    \redstoo{\Delta}{((\C{new}\;A\inang{\tbar}\inang{\loc\locbar}(\bar{v})).m\inang{\loc'\overline{\loc'}}
+                      (\bar{v'}),\mem)}
+            {\\([\loc'\overline{\loc'}/\rho\rhobar][\bar{v'}/\xbar]
+                [\C{new} \; A\inang{\tbar}\inang{\loc\locbar}(\bar{v})/\thisZ]\,e,\mem)}
+	  }{MethodInv}
+
+\lopsemrule{\textwidth}{
+    v_a = \lambdaexp{\loc'}{\rho\rhobar}
+                        {\taubar \; \xbar}{e} \spc
+    \mem(\loc') = \LIVE  \spc 
+}{
+    \redstoo{\Delta}{(v_a\inang{\loc\locbar}(\bar{v}) ,\mem)}
+            {([\bar{v}/\xbar][\loc\locbar/\rho\rhobar]\,e,\mem)}
+	  }{FnApply}
+
+%
+% \lopsemrule{\textwidth}{
+% %   \rgn \notin \rhoenv \spc
+% %   \fresh(\rgn') \spc
+%     \redstoo{\Delta}{(e,\mem)}{\invalidexn}
+% }{
+%     \redstoo{\Delta}{(\letregion{\rgn}{e},\mem)}{\invalidexn}
+%   }{EXCEPTION}
+
+\lopsemrule{\textwidth}{
+%   \fgjN = \RgnZ\inang{T} \spc
+%   \rgn \in dom(\mem) \spc
+    \mem(\loc) = \USED \spc
+    \redstocup{\loc}{(e,\mem[\loc \mapsto \LIVE])}{\invalidexn}
+}{
+    \redstoo{\Delta}{(\C{new} \; \RgnZT{\toploc\loc} (e),\mem)}
+            {\invalidexn}
+	  }{Exception}
+
+\lopsemrule{\textwidth}{
+%   \fgjN = \RgnZ\inang{T} \spc
+%   \rgn \in dom(\mem) \spc
+}{
+    \redstoo{\Delta}{(\letexp{x}{v}{e},\mem)}
+            {([v/x]e,\mem)}
+	  }{LetExp}
+
+\textbf{Evaluation Context} \fbox {\(E\)}\\
+\begin{smathpar}
+\begin{array}{lcl}
+E & \coloneqq & \bullet \ALT (\bullet).f \ALT \bullet.m\inang{\locbar}(\ebar) \ALT
+      v.m\inang{\locbar}(...,\bullet,...) \ALT \C{new}\;
+      \fbN(...,\bullet,...) \\
+  &  & \ALT \C{new} \; \RgnZ\inang{T}\inang{\toprgn}(\bullet)
+       \ALT \bullet\inang{\locbar}(\ebar)\ALT
+       v\inang{\locbar}(...,\bullet,...)\\
+  &  & \ALT \letexp{x}{\bullet}{e} \ALT \open{\bullet}{\rgn}{y}{e} 
+%      \ALT \opened{\loc}{\status}{\bullet} \ALT \letd{\loc}{\bullet}
+%     The following should be forbidden. See NEW-REGION rule 2.
+%      \ALT \C{new} \; \RgnZ\inang{T}\inang{\toploc\loc}(\bullet)
+\end{array}
+\end{smathpar}
+
+\caption{\fbname: operational semantics (continuation of
+Fig.~\ref{fig:fb-opsem})}
+\label{fig:fb-opsem-1}
+\end{figure*}
+\clearpage
+
+\subsection{Full Constraint Generation Semantics}
+
+\newcommand{\localokin}[2]{#1 \; \texttt{ok} \; \texttt{in} \; #2}
+\newcommand{\exptycxFix}[1]{\A,#1,r}
+
+\begin{figure*}[!h]
+
+\beginrules
+
+%%%%%%%%%%% Header Box %%%%%%%%%%%
+\multicolumn{2}{l}{
+  \textbf{Expression Typing Constraint Generation} \; \fbox{  \( \exprok{\stdcontext}{e}{\tau}{C} \)}
+}
+\\[0.3cm]
+
+%%%%%%%%%%% () and x %%%%%%%%%%%
+\lgcfact{Unit}{\exprok{\stdcontext}{\unitval}{\unitZ}{\{\}}}
+
+\lgcfact{Var}{\exprok{\stdcontext}{x}{\env(\tau)}{\{\}}}
+
+% \begin{minipage}{1.2in}
+% \begin{smathpar}
+% \begin{array}{l}
+% \renewcommand*{\arraystretch}{1.2}
+% \exprok{\stdcontext}{\unitval}{\unitZ}{\{\}} \\
+% \exprok{\stdcontext}{x}{\env(\tau)}{\{\}}
+% \end{array}
+% \end{smathpar}
+% \end{minipage}
+
+%%%%%%%%%%% FIELD-ACCESS: e.f %%%%%%%%%%%
+\lgcrule{FieldAccess}
+  {
+    \exprok{\stdcontext}{e}{\tau'}{C} \spc
+    \bar{f}:\taubar \,=\, \fields(\bound_{\A.\aenv}(\tau'))
+  }
+  {
+    \exprok{\stdcontext}{e.f_i}{\tau_i}{C}
+  }
+
+%%%%%%%%%%% LET %%%%%%%%%%%
+  \lgcrule{Let}
+  {
+    \exprok{\stdcontext}{e_1}{\tau_1}{C_1} \spc
+    \exprok{\A,{\env[x\mapsto\tau_1]},r}{e_2}{\tau_2}{C_2} \\
+  }
+  {
+    \exprok{\stdcontext}{\letexp{x}{e_1}{e_2}}{\tau_2}{C_1 \cup C_2}
+  }
+
+%%%%%%%%%%% METHOD-INV %%%%%%%%%%%
+  \lgcrule{MethodInv}
+  {
+    \exprok {\stdcontext} {e_0} {\tau} {C_1} \spc C_2 = \{ r \rbar \in \A.\rhoenv \}
+    \nl
+    \mtype(m,\bound_{\A.\aenv}(\tau)) = \inang{\rho \rhobar \,|\, \phi}\bar{\tau^1}\rightarrow{\tau^2}
+    \nl
+%   \substFn = [\rbar/\rhobar] \\
+    \typeok {\A} {\inang{\rho\rhobar \,|\,\phi}\bar{\tau^1}\rightarrow{\tau^2}} {C_3}
+       \spc
+       \exprok {\stdcontext} {\bar{e}} {[r\rbar/\rho\rhobar](\bar{\tau^1})} {C_4}
+    \nl
+%   \subtyp{\A}{\bar{\tau'}}{\substFn(\bar{\tau^1})} \spc
+    C_5 = \{ \isvalid{\A.\phicx}{[r\rbar/\rho\rhobar](\phi)} \}
+  }
+  {
+    \exprok {\stdcontext} {e_0.m\inang{r \rbar}(\bar{e})} 
+       {[r \rbar / \rho \rhobar](\tau^2)} {C_1 \cup C_2 \cup C_3 \cup C_4 \cup C_5}
+  }
+
+%%%%%%%%%%% LAMBDA %%%%%%%%%%%
+
+\lgcrule{Lambda}
+  {
+    \rgn \in \A.\rhoenv \spc
+    \rho\rhobar \notin \A.\rhoenv
+    \spc
+%   \rhoenv' = \rhoenv \cup \{\rhoalloc,\rhobar\}\spc
+    \A' = (\A.\rhoenv \cup \{\rho\rhobar\}, \A.\aenv, 
+          \A.\phicx \conj \phi)
+    \\
+    \tywf{\A'.\rhoenv}{\phi}\spc
+    \typeok {\A'} {\bar{\tau^1}} {C_1} \spc
+    \typeok {\A'} {\tau^2} {C_2}
+    \\
+    \exprok {\A',\env[\xbar \mapsto \bar{\tau^1}],\rho} {e} {\tau^2} {C_3}
+    % \spc
+    % C_4 = \bigcup\limits_{\pi'\in \frv(e)\setminus\{\rhobar\}}\{\pi'\outlives\pi\}
+  }
+  {
+    \exprok {\stdcontext}
+           {\lambdaexp{\rgn}{\rho\rhobar \,|\, \phi} {\xbar:\bar{\tau^1}}{e}}
+           {\inang{\rho\rhobar \,|\, \phi} \bar{\tau^1} \xrightarrow{\rgn} \tau^2}
+	   {\cup_{i=1}^{4} C_i}
+  }
+
+%%%%%%%%%%% SUBTYPING %%%%%%%%%%%
+\lgcrule{SubTyping}{
+    \exprok {\exptycxFix{\env}} {e} {\tau} {C_1} \spc  \subtypeok {\A} {\tau} {\tau'} {C_2}
+}{
+    \exprok {\exptycxFix{\env}} {e} {\tau'} {C_1 \cup C_2}
+}
+
+%%%%%%%%%%% METHOD %%%%%%%%%%%
+\multicolumn{2}{l}{
+   \textbf{Method Well-formedness Constraint Generation}  \; \fbox{$\okinok{d}{B}{C}$}
+}
+\\[0.3cm]
+
+\lgcrule{Method}{
+CT(B) = \C{class}\; B \angAlpha \inang{\rhobar \,|\, \varphi} \extends \fbN
+   \{ \cdots \}
+   % \{ \bar{\tau^f}\,\xbar;\;\bar{d} \}
+\nl
+\A = (\rhoenv,\aenv,\phicx) = (\{\rhobar,\rho_m,\rhobarm\}, [\bar{\tyvar} \mapsto \bar{\fgjN}], \varphi_m) \spc\spc
+C_1 = \{ \tywf{\rhoenv}{\varphi_m} \}
+\nl
+\env = \cdot[\thisZ \mapsto B\inang{\bar{\tyvar}}\inang{\rhobar}][\xbar \mapsto \bar{\tau^1}]
+\spc
+\mtype(m,\fbN) = \inang{\rhobarm \,|\, \phi_m}\bar{\tau^1} \rightarrow \tau^2
+\nl
+\exprok {\A,\env,\rho_m}{e} {\tau^2} {C_2}
+\spc \typeok {\A} {\bar{\tau^1}} {C_3}
+\spc \typeok {\A} {\tau^2} {C_4}
+% \subtypeok {\A} {\tau'} {\tau} {C_3}
+}{
+\okinok {\tau^2 \; m\inang{\rho_m \rhobarm \,|\, \varphi_m} (\bar{\tau^1} \;  \xbar)\{\C{return} e;\}}
+   {B}
+   {(C_1 \cup C_2 \cup C_3 \cup C_4)}
+}
+
+%%%%%%%%%%% CLASS %%%%%%%%%%%
+\multicolumn{2}{l}{
+   \textbf{Class Well-formedness Constraint Generation}  \; \fbox{$\classok{B}{C}$}
+}
+\\[0.3cm]
+
+\lgcrule{Class}{
+  \A = (\rhoenv, \aenv, \phicx) = (\{\rho,\rhobar\}, [\bar{\tyvar} \mapsto \bar{\fgjN}],\varphi) \\
+C_1 = \{ \tywf{\rhoenv}{\varphi} \} \spc\spc
+\fgjtywf{\aenv}{\bar{\fgjN}} \spc\spc
+\typeok {\A} {\fbN} {C_2} \spc\spc
+\typeok {\A} {\bar{\tau^f}} {C_3} \\
+C_4 = \{\isvalid{\phicx}{\allocRgn(\bar{\tau^f}) \outlives \rho \conj \allocRgn(\fbN) = \rho}\} \\
+\okinok {\bar{d}} {B} {C_5}
+}{
+% \typeok {} {\hdOf{B}{\varphi}\{\bar{\tau^f}\,\xbar;\;\bar{d}\}} {\bigcup_{i=1}^5 C_i}
+\classok
+  {\C{class}\; B\angAlpha \inang{\rho \rhobar \,|\, \varphi} \extends \fbN
+      \{\bar{\tau^f}\,\xbar;\;\bar{d}\}}
+  {\bigcup_{i=1}^5 C_i}
+}
+
+\myendrules
+
+\caption{\fbname: Constraint generation rules part 2 (Continuation of
+Fig.~\ref{fig:constraint-gen-0})}
+\label{fig:constraint-gen-1}
+\end{figure*}
+\clearpage
+
+\begin{figure*}[htbp]
+
+\beginrules
+
+%%%%%%%%%%% Header Box %%%%%%%%%%%
+\multicolumn{2}{l}{
+  \textbf{Subtyping Constraint Generation} \;
+\fbox{  \( \subtypeok{\A}{\tau_1}{\tau_2}{C} \)}
+}
+\\[0.3cm]
+
+%%%%%%%%%%% SUBTYPING: REFLEXIVE %%%%%%%%%%%
+
+  \lgcfact{Reflexivity}{
+    \subtypeok{\A}{\tau}{\tau}{\{\}}
+  }
+
+%%%%%%%%%%% SUBTYPING: UNIFICATION %%%%%%%%%%%
+  \lgcfact{Unify}{
+    \subtypeok{\A}{\tau}{[\pi/\rho](\tau)}{ \{ \pi \outlives \rho, \rho \outlives \pi \} }
+  }
+
+%%%%%%%%%%% SUBTYPING: TRANSITIVITY %%%%%%%%%%%
+
+  \lgcrule{Transitivity}{
+    \subtypeok{\A}{\tau_1}{\tau_2}{C_1} \spc
+    \subtypeok{\A}{\tau_2}{\tau_3}{C_2}
+  }{
+    \subtypeok{\A}{\tau_1}{\tau_3}{C_1 \cup C_2}
+  }
+
+%%%%%%%%%%% FUNCTION SUBTYPING %%%%%%%%%%%
+  \lgcrule{FnSubtyping}
+  {
+    C_1 = \{ \isvalid{\A.\phicx}{\phi_1 \Rightarrow \phi_2} \}
+    \\
+    \subtypeok {\A} {\bar{\tau^{11}}} {\bar{\tau^{21}}} {C_2}
+    \spc
+    \subtypeok {\A} {\tau^{22}} {\tau^{12}} {C_3}
+  }
+  {
+    \subtypeok {\A}
+      {\inang{\rhobar \,|\, \phi_2}\bar{\tau^{21}} \xrightarrow{\rgn} \tau^{22}}
+      {\inang{\rhobar \,|\, \phi_1}\bar{\tau^{11}} \xrightarrow{\rgn} \tau^{12}}
+      {C_1 \cup C_2 \cup C_3}
+  }
+
+%%%%%%%%%%% Header Box %%%%%%%%%%%
+\multicolumn{2}{l}{
+  \textbf{Type Well-formedness Constraint Generation} \;
+  \fbox{  \( \typeok{\A}{\tau}{C} \)}
+}
+\\[0.3cm]
+
+%%%%%%%%%%% TYPE WELL-FORMEDNESS %%%%%%%%%%%
+
+%%%%%%%%%%% OBJECT TYPE %%%%%%%%%%%
+\lgcrule{ObjectType}
+  {
+    C = \{ \rgn \in \rhoenv \}
+  }
+  {
+    \typeok {(\rhoenv,\aenv,\phicx)} {\ObjZ\inang{\rgn}} {C}
+  }
+
+%%%%%%%%%%% CLASS TYPE %%%%%%%%%%%
+  \lgcrule{ClassType}
+  {
+    CT(B) = \headerOf{B}\{...\}
+    \spc
+    \fgjtywf{\aenv}{B\inang{\tbar}}
+    \\
+    C = \{ \rbar \in \rhoenv, \isvalid{\phicx}{[\rbar/\rhobar](\phi)} \}
+  }
+  {
+    \typeok {(\rhoenv,\aenv,\phicx)} {B\inang{\tbar}\inang{\rbar}} {C}
+  }
+
+%%%%%%%%%%% GENERIC TYPE PARAMETER %%%%%%%%%%%
+  \lgcrule{TypeParam}
+  {
+    \fgjtywf{\aenv}{T} \spc
+    \fgjsubtyp{\aenv}{T}{\ObjZ} \spc
+    \spc
+    C = \{ \rgn \in \rhoenv \}
+  }
+  {
+    \typeok {(\rhoenv,\aenv,\phicx)}{T@\rgn} {C}
+  }
+
+%%%%%%%%%%% FUNCTION TYPE %%%%%%%%%%%
+  \lgcrule{FnType}
+  {
+    C_1 = \{ \rgn \in \rhoenv \}
+    \\
+    \rhobar \notin \rhoenv \spc
+    \rhoenv' = \rhoenv \cup \{\rhobar\} \spc
+    \A' = (\rhoenv', \aenv, \phicx \conj \phi)
+    \\
+    \typeconsok{\rhoenv'}{\phi} {C_2} \spc 
+    \typeok{\A'}{\bar{\tau^1}} {C_3} \spc
+    \typeok{\A'}{\tau^2} {C_4}
+  }
+  {
+    \typeok{(\rhoenv,\aenv,\phicx)} {\inang{\rhobar \,|\, \phi} \bar{\tau^1} \xrightarrow{\rgn} \tau^2} 
+       {C_1 \cup C_2 \cup C_3 \cup C_4}
+  }
+
+%%%%%%%%%%% REGION TYPE %%%%%%%%%%%
+  \lgcrule{RegionType}
+  { 
+    \fgjtywf{\aenv}{T}
+  }
+  {
+    \typeok {(\rhoenv,\aenv,\phicx)} {\RgnZ\inang{T}\inang{\toprgn}} {\{\}}
+  }
+
+\myendrules
+
+\caption{\fbname: Constraint generation rules part 3 (Continuation of
+Fig.~\ref{fig:constraint-gen-1})}
+\label{fig:constraint-gen-2}
+\end{figure*}
+\vfill
+%\renewcommand{\rgn}{\myownthrowexception}
 \end{document}
